### 2021-05-18 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Initial commit with first version of data package

---
title: "Population density and basic reproductive number of COVID-19 across United States counties revisited"
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    corresponding: paezha@mcmaster.ca
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

author_summary: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

bibliography: bibliography.bib
output: rticles::plos_article
csl: plos.csl
---

```{r install-data-package, include = FALSE}
# Run only once if needed to install data package `sobiEquity`
if (!require("r0density", character.only = TRUE)) {
      devtools::install_github("https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed", subdir = "r0density")
  }
```

# Introduction

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

A list

- Item 1
- Item 2

Here are two sample references: @Feynman1963118 [@Dirac1953888].

# References {#references .unnumbered}

### 2021-05-18 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Initial commit with first version of data package

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended
% to minimize problems and delays during our production
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% Once your paper is accepted for publication,
% PLEASE REMOVE ALL TRACKED CHANGES in this file
% and leave only the final text of your manuscript.
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file.
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment.
%
% Please add line breaks to long display equations when possible in order to fit size of the column.
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace}
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
% \bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{forarray}
\usepackage{xstring}
\newcommand{\getIndex}[2]{
  \ForEach{,}{\IfEq{#1}{\thislevelitem}{\number\thislevelcount\ExitForEach}{}}{#2}
}

\setcounter{secnumdepth}{0}

\newcommand{\getAff}[1]{
  \getIndex{#1}{McMaster University}
}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Population density and basic reproductive number of
COVID-19 across United States counties
revisited} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Antonio Paez\textsuperscript{\getAff{McMaster
University}}\textsuperscript{*}\\
\bigskip
\textbf{\getAff{McMaster University}}School of Earth, Environment and
Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada\\
\bigskip
* Corresponding author: paezha@mcmaster.ca\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
pulvinar, volutpat risus.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
pulvinar, volutpat risus.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\emph{Text based on plos sample manuscript, see
\url{https://journals.plos.org/ploscompbiol/s/latex}}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
dui eu ante cursus gravida non sed sem. Phasellus mauris velit, dapibus
finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
pulvinar, volutpat risus.

A list

\begin{itemize}
\tightlist
\item
  Item 1
\item
  Item 2
\end{itemize}

Here are two sample references: {[}1,2{]}.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\hypertarget{ref-Feynman1963118}{}%
\CSLLeftMargin{1. }
\CSLRightInline{Feynman RP, Vernon Jr. FL. The theory of a general
quantum system interacting with a linear dissipative system. Annals of
Physics. 1963;24: 118--173.
doi:\href{https://doi.org/10.1016/0003-4916(63)90068-X}{10.1016/0003-4916(63)90068-X}}

\leavevmode\hypertarget{ref-Dirac1953888}{}%
\CSLLeftMargin{2. }
\CSLRightInline{Dirac PAM. The lorentz transformation and absolute time.
Physica. 1953;19: 888--896.
doi:\href{https://doi.org/10.1016/S0031-8914(53)80099-6}{10.1016/S0031-8914(53)80099-6}}

\end{CSLReferences}

\nolinenumbers

\end{document}

### 2021-05-18 r0density/raw-data-processing/00-Data-Preprocessing.Rmd
Initial commit with first version of data package

---
title: "00-Data-Preprocessing"
output: html_notebook
---

This [R Markdown](http://rmarkdown.rstudio.com) Notebook recovers the data used by Sy et al. ([2020](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0249271)) in their analysis of the basic reproductive number of COVID-19 and population density in US counties.

Sy et al. calculated the basic reproductive number and estimated models to study its relationship to population density (and some other controls). There are two issues with their approach:

1. Data censoring is ignored

2. Spatial autocorrelation is ignored

In this notebook I create a data package with the data necessary to reanalyze this issue.

Load packages:
```{r message=FALSE, warning=FALSE}
#library(censReg)
#library(Matrix)
#library(nlme)
#library(spatialprobit)
library(spdep)
library(tidycensus)
library(tidyverse)
library(units)
#library(tigris)
```

Load the data table prepared by Sy et al.:
```{r}
load("Sy_USDensityR0_Data.RData")
```

Notice that the table used by Sy et al. (county_R0_demog) includes only counties for which the basic reproductive number was calculated reliably (see pp. 2-3):

"We restricted our analyses to counties that met a certain threshold of cumulative case counts at the end of the exponential growth period, but including counties with less than 25 cases included daily incidence counts that were insufficient to calculate R0 and yielded computational errors. Our final analytic sample included counties with 25 cases or greater at the end of the exponential growth period."

Use `tidycensus` to retrieve the geometry of _all_ counties. The geometry is then used to calculate the population density (`vars <- tidycensus::load_variables()` can be used to explore the variable names):
```{r}
county_geo <- get_decennial(geography = "county", 
                            variables = "P001001", # Total population
                            year = 2010, 
                            geometry = TRUE, 
                            keep_geo_vars = TRUE) %>%
  rename(population = value) %>%
  select(-variable) %>%
  mutate(area = st_area(geometry) %>%
           set_units(km^2),
         density = population/(1000 * area)) # 1,000 people per square km
```

Also, retrieve other socio-economic and demographic variables (median household income and travel by private transportation):
```{r}
sed_variables <- get_acs(geography = "county", 
                            variables = c("B19013_001",
                                          "B08006_001",
                                          "B08006_002"), 
                            year = 2019, 
                            geometry = FALSE) %>%
  select(-moe) %>%
  pivot_wider(names_from = variable,
              values_from = estimate) %>%
  rename(hincome = B19013_001, 
         commuters = B08006_001,
         car = B08006_002) %>%
  transmute(GEOID,
            commuters,
            car,
            hincome = hincome/10000) # Median household income in tens of thousands
         #hincome_log = log(hincome),
         #private = car/commuters * 100) # Percentage of commuters who travel by car/van/truck

```

These are the variables collected.

2010 Census:

P001001: Total Population

2019 ACS:

B19013_001: Estimate!!Median household income in the past 12 months (in 2019 inflation-adjusted dollars) (MEDIAN HOUSEHOLD INCOME IN THE PAST 12 MONTHS IN 2019 INFLATION-ADJUSTED DOLLARS)
B08006_001: Estimate!!Total (SEX OF WORKERS BY MEANS OF TRANSPORTATION TO WORK)
B08006_002: Estimate!!Total:!!Car, truck, or van (SEX OF WORKERS BY MEANS OF TRANSPORTATION TO WORK)

Join the two tables:
```{r}
county_geo <- county_geo %>%
  left_join(sed_variables,
            by = "GEOID")
```

Now join Sy's data to the counties with geometry and other socio-economic variables (the equivalent variables are dropped from Sy et al.'s table, because they are missing all the counties without a basic reproductive number):
```{r}
county_geo <- county_geo %>%
  mutate(GEOID.num = as.numeric(GEOID)) %>%
  left_join(county_R0_demog %>%
              select(GEOID, state, R, R_death, sensitivitydensity),
            by = c("GEOID.num" = "GEOID")) %>%
  select(-GEOID.num, -CENSUSAREA, -COUNTYFP, -STATEFP, -starts_with("geometry"), geometry)
```

Save data table:
```{r}
save(county_geo, file = "county_geo.RData", compress = "xz")
```

Copy data file to data folder!

<!--

Convert missing R's to zeros and calculate expansion variables:
```{r}
county_geo_clean <- county_geo %>%
  mutate(R = replace_na(R, 0),
         density_2 = density^2) %>%
  drop_na(commuters) %>%
  st_as_sf()
```

Create list of neighbors:
```{r}
county_nb <- poly2nb(county_geo_clean)
```

Check list of neighbors:
```{r}
summary(county_nb)
```

Notice that there are 9 counties that are isolates. Which counties are these?
```{r}
county_geo_clean[c(69, 546, 547, 549, 1226, 1876, 2974, 3146, 3200),]
```

Find nearest neighbors by distance:
```{r}
county_nb_dist <- knearneigh(county_geo_clean %>% 
                               st_centroid()) %>%
  knn2nb()
```

Add nearest neighbor to isolates:
```{r}
county_nb[[69]] <- county_nb_dist[[69]]
county_nb[[546]] <- county_nb_dist[[546]]
county_nb[[547]] <- county_nb_dist[[547]]
county_nb[[549]] <- county_nb_dist[[549]]
county_nb[[1226]] <- county_nb_dist[[1226]]
county_nb[[1876]] <- county_nb_dist[[1876]]
county_nb[[2974]] <- county_nb_dist[[2974]]
county_nb[[3146]] <- county_nb_dist[[3146]]
county_nb[[3200]] <- county_nb_dist[[3200]]
```

Convert nearest neighbors to listw:
```{r}
nb_B <- nb2listw(county_nb, 
                 style="B", 
                 zero.policy = TRUE)
nb_B$style
```

Convert listw to sparse matrix for use in spatial tobit:
```{r}
B <- as(nb_B, "CsparseMatrix")
all(B == t(B))
```

## Replicate Sy et al.

Fit (mixed) linear models as in Sy et al.:
```{r}
table1model1 <- lme(R ~ density_log , 
                    random = ~ 1| state, 
                    data = county_geo_clean %>%
                      filter(R > 0))
summary(table1model1)
intervals(table1model1)
```

Linear mixed models of R0 and density + % private transportation
```{r}
table1model2 <- lme(R ~ density_log + private , 
                    random = ~ 1| state, 
                    data = county_geo_clean %>%
                      filter(R > 0))
summary(table1model2)
intervals(table1model2)
```

Linear mixed models of R0 and density + % private transportation + median income
```{r}
table1model3 <- lme(R ~ density_log + private + hincome, 
                    random = ~ 1| state, 
                    data = county_geo_clean %>%
                      filter(R > 0))
summary(table1model3)
intervals(table1model3)
```

Linear mixed models of R0 and density + % private transportation + median income + interaction of density and % private transportation
```{r}
table1model4 <- lme(R ~ density_log*private + hincome, 
                    random = ~ 1| state, 
                    data = county_R0_demog)
summary(table1model4)
intervals(table1model4)
```

## Fit tobit version of models

```{r}
table1model1 <- censReg(R ~ density_log, 
                        left = 0,
                    data = county_geo_clean)
summary(table1model1)
```

Linear mixed models of R0 and density + % private transportation
```{r}
table1model2 <- censReg(R ~ density_log + private, 
                    data = county_geo_clean)
summary(table1model2)
```

Linear mixed models of R0 and density + % private transportation + median income
```{r}
table1model3 <- censReg(R ~ density_log + private + hincome, 
                    data = county_geo_clean)
summary(table1model3)
```

Tobit models of R0 and density + % private transportation + median income + interaction of density and % private transportation
```{r}
table1model4 <- censReg(R ~ density_log*private + hincome, 
                    data = county_geo_clean)
summary(table1model4)

```

## Spatially autoregressive tobit

Fit spatially autoregressive tobit:
```{r}
# Fit SAR Tobit
fit_sartobit <- sartobit(R ~ density_log + private + hincome_log,
                         B,
                         ndraw = 1000,
                         burn.in = 200, 
                         showProgress = TRUE,
                         data = county_geo_clean,
                         computeMarginalEffects = TRUE)
summary(fit_sartobit)
impacts(fit_sartobit)
```

With interactions:
```{r}
# Fit SAR Tobit
fit_sartobit <- sartobit(R ~ density_log + density_log:private + hincome_log,
                         B,
                         ndraw = 1000,
                         burn.in = 200, 
                         showProgress = TRUE,
                         data = county_geo_clean)
summary(fit_sartobit)
impacts(fit_sartobit)
```
FIGURE 2 - POPULATION DENSITY THRESHOLD

Compare density of counties with oubreaks vs counties without outbreak
```{r}
county_outbreak_yn %>%
  group_by(included) %>%
  summarise(mean = median(Density, na.rm=TRUE), 
            lower_IQR = summary(Density)[2],
            upper_IQR=summary(Density)[5], 
            min=summary(Density)[1], 
            max=summary(Density)[6])

summary(county_outbreak_yn$Density)
wilcox.test(county_outbreak_yn$Density~county_outbreak_yn$included)
```

Boxplot of counties with oubreaks vs counties without outbreak
```{r}
county_outbreak_yn$included <- as.factor(county_outbreak_yn$included)

ggplot(county_outbreak_yn, aes(included, Density)) + 
  geom_boxplot(fill="gray92", width=0.5) + 
  ylim(0,200) + 
  #theme_minimal() + 
  ylab("Population Density (population/sq km)") + 
  xlab("United States Counties") + 
  theme_minimal(base_size=13)
```

TABLE 1 - LINEAR MIXED MODELS

Linear mixed models of R0 and density:
```{r}
table1model1 <- lme(R ~ density_log , 
                    random = ~ 1| state, 
                    data = county_geo_clean %>%
                      filter(R > 0))
summary(table1model1)
intervals(table1model1)
```

Linear mixed models of R0 and density + % private transportation
```{r}
table1model2 <- lme(R ~ density_log + private , 
                    random = ~ 1| state, 
                    data = county_geo_clean %>%
                      filter(R > 0))
summary(table1model2)
intervals(table1model2)
```

Linear mixed models of R0 and density + % private transportation + median income
```{r}
table1model3 <- lme(R ~ density_log + private + hincome, 
                    random = ~ 1| state, 
                    data = county_geo_clean %>%
                      filter(R > 0))
summary(table1model3)
intervals(table1model3)
```

Linear mixed models of R0 and density + % private transportation + median income + interaction of density and % private transportation
```{r}
table1model4 <- lme(R ~ density_log*private + hincome, 
                    random = ~ 1| state, 
                    data = county_R0_demog)
summary(table1model4)
intervals(table1model4)
```

## TABLE 2 - SENSITIVITY ANALYSIS

a. Deaths only - Subset counties with R0 from deaths only

```{r}
county_R0_demog_a <- subset(county_R0_demog, !is.na(county_R0_demog$R_death))
```

Linear mixed models of R0 (death) and density 
```{r}
table2model1_a <- lme(R_death ~ density_log , random = ~ 1| state, data = county_R0_demog_a)
summary(table2model1_a)
intervals(table2model1_a)
```

Linear mixed models of R0 (death) and density + % private transportation
```{r}
table2model2_a <- lme(R_death~ density_log + private , random = ~ 1| state, data = county_R0_demog_a)
summary(table2model2_a)
intervals(table2model2_a)
```

Linear mixed models of R0 (death) and density + % private transportation + median income
```{r}
table2model3_a <- lme(R_death~ density_log + private + hincome, random = ~ 1| state, data = county_R0_demog_a)
summary(table2model3_a)
intervals(table2model3_a)
```

Linear mixed models of R0 (death) and density + % private transportation + median income + interaction of density and % private transportation
```{r}
table2model4_a <- lme(R_death~ density_log*private + hincome, random = ~ 1| state, data = county_R0_demog_a)
summary(table2model4_a)
intervals(table2model4_a)
```

b. No counties adjacent to high density counties - within 15 mile radius, high density > 144.0492 population/sq km
```{r}
county_R0_demog_b <- county_R0_demog[county_R0_demog$sensitivitydensity,]
```

Linear mixed models of R0 and density (no counties adjacent to high density counties)
```{r}
table2model1_b <- lme(R ~ density_log , random = ~ 1| state, data = county_R0_demog_b)
summary(table2model1_b)
intervals(table2model1_b)
```

Linear mixed models of R0 and density + % private transportation (no counties adjacent to high density counties)
```{r}
table2model2_b <- lme(R ~ density_log + private , random = ~ 1| state, data = county_R0_demog_b)
summary(table2model2_b)
intervals(table2model2_b)
```

Linear mixed models of R0 and density + % private transportation + median income (no counties adjacent to high density counties)
```{r}
table2model3_b <- lme(R ~ density_log + private + hincome, random = ~ 1| state, data = county_R0_demog_b)
summary(table2model3_b)
intervals(table2model3_b)
```

Linear mixed models of R0 and density + % private transportation + median income + interaction of density and % private transportation (no counties adjacent to high density counties)
```{r}
table2model4_b <- lme(R ~ density_log*private + hincome , random = ~ 1| state, data = county_R0_demog_b)
summary(table2model4_b)
intervals(table2model4_b)
```

c. Removing high influence counties - Cook's D over 4/N for each model)

Linear mixed models of R0 and density (no outliers)
```{r}
sample_size<-nobs(table1model1)
cooksd1 <- CookD(table1model1)
influential <- as.numeric(names(cooksd1)[(cooksd1 > (4/sample_size))])
county_R0_demog_c <- county_R0_demog[-influential, ]
table1model1_c <- lme(R ~ density_log , random = ~ 1| state, data = county_R0_demog_c)
summary(table1model1_c)
intervals(table1model1_c)
```

Linear mixed models of R0 and density + % private transportation (no outliers)
```{r}
sample_size<-nobs(table1model2)
cooksd1 <- CookD(table1model2)
influential <- as.numeric(names(cooksd1)[(cooksd1 > (4/sample_size))])
county_R0_demog_c <- county_R0_demog[-influential, ]
table1model2_c <- lme(R ~ density_log + private , random = ~ 1| state, data = county_R0_demog_c)
summary(table1model2_c)
intervals(table1model2_c)
```

Linear mixed models of R0 and density + % private transportation + median income (no outliers)
```{r}
sample_size<-nobs(table1model3)
cooksd1 <- CookD(table1model3)
influential <- as.numeric(names(cooksd1)[(cooksd1 > (4/sample_size))])
county_R0_demog_c <- county_R0_demog[-influential, ]
table1model3_c <- lme(R ~ density_log + private + hincome, random = ~ 1| state, data = county_R0_demog_c)
summary(table1model3_c)
intervals(table1model3_c)
```

Linear mixed models of R0 and density + % private transportation + median income + interaction of density and % private transportation (no outliers)
```{r}
sample_size<-nobs(table1model4)
cooksd1 <- CookD(table1model4)
influential <- as.numeric(names(cooksd1)[(cooksd1 > (4/sample_size))])
county_R0_demog_c <- county_R0_demog[-influential, ]
table1model4_c <- lme(R ~ density_log*private + hincome, random = ~ 1| state, data = county_R0_demog_c)
summary(table1model4_c)
intervals(table1model4_c)
```

### 2021-05-18 r0density.Rcheck/00_pkg_src/r0density/R/r0density.R

@@ -1,0 +1,47 @@
+#' vaccHamilton
+#'
+#' vaccHamilton is a companion package for the paper "Population Density and Basic Reproductive Number of COVID-19 Revisited".
+#'
+#' The package includes the data used in the analysis.
+#'
+#' @author Antonio Paez, \email{paezha@@mcmaster.ca}
+#' @docType package
+#' @name r0density
+NULL
+
+#' Counties of the United States with relevant covariates.
+#'
+#' Boundaries of the counties in the United States.
+#'
+#' @format A simple features data frame with 3,221 rows and 18 variables:
+#' \describe{
+#'   \item{GEO_ID}{Unique identifier of the county}
+#'   \item{STATE}{Code of state}
+#'   \item{COUNTY}{Code of county}
+#'   \item{NAME.x}{Short name of county}
+#'   \item{LSAD}{Code of the Legal/Statistical Area Description}
+#'   \item{GEOID}{Unique identifier of the county}
+#'   \item{NAME.y}{Longer name of county}
+#'   \item{population}{Total population from the 2010 Census (variable P001001)}
+#'   \item{area}{Area of county in sq.km}
+#'   \item{density}{Population density in 1,000 people per sq.km}
+#'   \item{commuters}{ACS 2019 variable B08006_001: Estimate!!Total (SEX OF WORKERS BY MEANS OF TRANSPORTATION TO WORK)}
+#'   \item{car}{ACS 2019 variable B08006_002: Estimate!!Total:!!Car, truck, or van (SEX OF WORKERS BY MEANS OF TRANSPORTATION TO WORK)}
+#'   \item{hincome}{ACS 2019 variable B19013_001: Estimate!!Median household income in the past 12 months (in 2019 inflation-adjusted dollars) (MEDIAN HOUSEHOLD INCOME IN THE PAST 12 MONTHS IN 2019 INFLATION-ADJUSTED DOLLARS)}
+#'   \item{state}{Two letter code of state (from Sy et al.)}
+#'   \item{R}{Basic reproductive number (from Sy et al.)}
+#'   \item{R_death}{Basic reproductive number based on deaths (from Sy et al.)}
+#'   \item{sensitivitydensity}{Variable used by Sy et al. to indicate observations used in their sensitivity analysis}
+#'   \item{geometry}{Geometry of boundaries}
+#' }
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords covid-19
+#' @keywords Census
+#' @keywords United States
[…]

### 2021-05-18 r0density/R/r0density.R

@@ -1,0 +1,47 @@
+#' r0density
+#'
+#' r0density is a companion package for the paper "Population Density and Basic Reproductive Number of COVID-19 Revisited".
+#'
+#' The package includes the data used in the analysis.
+#'
+#' @author Antonio Paez, \email{paezha@@mcmaster.ca}
+#' @docType package
+#' @name r0density
+NULL
+
+#' Counties of the United States with relevant covariates.
+#'
+#' Boundaries of the counties in the United States.
+#'
+#' @format A simple features data frame with 3,221 rows and 18 variables:
+#' \describe{
+#'   \item{GEO_ID}{Unique identifier of the county}
+#'   \item{STATE}{Code of state}
+#'   \item{COUNTY}{Code of county}
+#'   \item{NAME.x}{Short name of county}
+#'   \item{LSAD}{Code of the Legal/Statistical Area Description}
+#'   \item{GEOID}{Unique identifier of the county}
+#'   \item{NAME.y}{Longer name of county}
+#'   \item{population}{Total population from the 2010 Census (variable P001001)}
+#'   \item{area}{Area of county in sq.km}
+#'   \item{density}{Population density in 1,000 people per sq.km}
+#'   \item{commuters}{ACS 2019 variable B08006_001: Estimate!!Total (SEX OF WORKERS BY MEANS OF TRANSPORTATION TO WORK)}
+#'   \item{car}{ACS 2019 variable B08006_002: Estimate!!Total:!!Car, truck, or van (SEX OF WORKERS BY MEANS OF TRANSPORTATION TO WORK)}
+#'   \item{hincome}{ACS 2019 variable B19013_001: Estimate!!Median household income in the past 12 months (in 2019 inflation-adjusted dollars) (MEDIAN HOUSEHOLD INCOME IN THE PAST 12 MONTHS IN 2019 INFLATION-ADJUSTED DOLLARS)}
+#'   \item{state}{Two letter code of state (from Sy et al.)}
+#'   \item{R}{Basic reproductive number (from Sy et al.)}
+#'   \item{R_death}{Basic reproductive number based on deaths (from Sy et al.)}
+#'   \item{sensitivitydensity}{Variable used by Sy et al. to indicate observations used in their sensitivity analysis}
+#'   \item{geometry}{Geometry of boundaries}
+#' }
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords covid-19
+#' @keywords Census
+#' @keywords United States
[…]

### 2021-05-19 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
First draft of introduction and background

@@
 ---
-title: "Population density and basic reproductive number of COVID-19 across United States counties revisited"
+title: "The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic"
 author:
@@
 abstract: |
-  Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+  The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research.
+Given the high stakes of the situation, it is essential that scientific activites, on which good policy depends, are as transparent and reproducible as possible.
+Reproducibility is key for the efficient operation of the self-correction mechanisms of science, which work to weed out errors and refine our understanding of social and physical phenomena.
+In this paper, the importance of reproducibility is illustrated for the case of the association between population density and the the spread of SARS-CoV-2.
+Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequantial changes in the conclusions, as the empirical example in this paper shows.
 
 author_summary: |
-  Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
 
@@
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(cache = TRUE)
+```
 
@@
 
+```{r load-packages, include=FALSE, cache=FALSE}
+library(censReg)
+library(Matrix)
+library(nlme)
+library(r0density) # Data package
+library(spatialprobit)
+library(spdep)
+library(tidycensus)
+library(tidyverse)
+library(units)
+```
+
+```{r invoke-data, include=FALSE}
+data(county_geo)
+```
+
+```{r prepare-data, include=FALSE}
+# Convert missing R's to zeros and calculate expansion variables: 
+county_geo_clean <- county_geo %>%
+  mutate(R = replace_na(R, 0),
+         density_log = log(drop_units(density)),
+         density_2 = drop_units(density)^2,
+         hincome_log = log(hincome),
+         private = car/commuters * 100) %>% # Percentage of commuters who travel by car/van/truck) %>%
+  drop_na(commuters) %>%
+  st_as_sf()
+```
+
+```{r list-of-geographical-neighbors, include=FALSE}
+# Create a list of neighbors based on the polygons; the default contiguity criterion is Queen:
+county_nb <- poly2nb(county_geo_clean)
+```
+
+```{r spatial-data-check, include=FALSE}
+# Check list of neighbors
+summary(county_nb)
+
+# Notice that there are 9 counties that are isolates. These are the counties that do not have neighbors based on contiguity
+county_geo_clean[c(69, 546, 547, 549, 1226, 1876, 2974, 3146, 3200),]
+
+# To rectify the lack of neighbors, find the first nearest neighbor using the centroids:
+county_nb_dist <- knearneigh(county_geo_clean %>% 
+                               st_centroid()) %>%
+  knn2nb()
+
+# Add nearest neighbor to isolates:
+county_nb[[69]] <- county_nb_dist[[69]]
+county_nb[[546]] <- county_nb_dist[[546]]
+county_nb[[547]] <- county_nb_dist[[547]]
+county_nb[[549]] <- county_nb_dist[[549]]
+county_nb[[1226]] <- county_nb_dist[[1226]]
+county_nb[[1876]] <- county_nb_dist[[1876]]
+county_nb[[2974]] <- county_nb_dist[[2974]]
+county_nb[[3146]] <- county_nb_dist[[3146]]
+county_nb[[3200]] <- county_nb_dist[[3200]]
+```
+
+```{r spatial-weights, include=FALSE}
+# Convert the nearest neighbor object to a listw object:
+nb_B <- nb2listw(county_nb, 
+                 style="B", 
+                 zero.policy = TRUE)
+
+# Convert listw to sparse matrix for use in spatial tobit:
+B <- as(nb_B, "CsparseMatrix")
+```
+
 # Introduction
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the global pandemic that followed in its wake, led to an explosive growth of research.
+According to Fraser et al.
+[-@Fraser2021evolving], over 125,000 COVID-19-related papers were released in the first ten months from the first confirmed case of the disease.
+Of these, more than 30,000 were shared in pre-print servers, the use of which also exploded in the past year [@Kwon2021swamped; @Vlasschaert2020proliferation; @Anazco2021publication].
 
-A list
+Given the heavy human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has even called this deluge of publications a "carnage of substandard research" [@Bramstedt2020carnage].
+Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [@Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty but somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
+As reported in the literature, only a woefully small proportion of published research was actually reproducible before the pandemic [@Iqbal2016reproducible; @Stodden2018empirical], a situation that does not appear to have changed substantially since [@Sumner2020reproducibility; @Gustot2020quality].
 
-- Item 1
-- Item 2
+The push for open data and software, along with more strenuous efforts towards open, reproducible research, is simply a continuation of long-standing scientific practices of independent verification.
+Despite the (at times disproportionate) attention that high profile scandals in science tend to elicit in the media, science as a collective endeavor is remarkable for being a self-correcting enterprise, one with built-in mechanisms and incentives to weed out erroneous ideas.
+Over the long term, facts tend to prevail in science.
+At stake is the shorter-term impacts that research may have in other spheres of economic and social life.
+The case of economists Reinhart and Rogoff comes to mind: by the time the inaccuracies in their research were uncovered [see @Herndon2014high], their claims about debt and economic growth had already been seized by policy-makers on both sides of the Atlantic to justify austerity policies in the aftermath of the Great Recession of 2007-2009^[Nobel Prize in Economics Paul Krugman noted that "Reinhart–Rogoff may have had more immediate influence on public debate than any previous paper in the history of economics" \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}].
+As later research has demonstrated, those policies cast a long shadow, and their sequels continued to be felt for years [@Basu2017ten].
 
-Here are two sample references: @Feynman1963118 [@Dirac1953888].
+In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how].
+The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, prompted a flurry of research into the associations between density and the spread of the pandemic.
+Some important questions hang on the results of these research efforts.
+For example, are lower density regions safer from the pandemic?
+Are de-densification policies warranted, at least in the short term?
+And in the longer term, will the risks of life in high density regions presage a flight from cities?
+Over the past year, numerous papers have sought to throw light into the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed.
+Further, to complicate matters, precious few of these studies appear to be sufficiently open to support independent verification.
+
+The objective of this paper is to illustrate the importance of reproducibility in research, particularly in the context of the flood of COVID-19 papers.
+To this end, a recent study by Sy et al.
+[@Sy2021population] is chosen as an example of reproducible research.
+The objective is not to malign the analysis of these researchers, but rather to demonstrate the value of openness to allow for independent verification and further analysis.
+Open data and open code mean that an independent researcher can, with only modest efforts, not only verify the findings reported, but also examine the same data from a perspective which may not have been available to the original researchers due to differences in disciplinary perspectives, methodological traditions, and/or training, among other possible factors.
+The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in research.
+This paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
+The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, is available in a public repository^[\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}].
+
+# Background
+
+The concern with population density and the spread of the virus during the COVID-19 pandemic was fueled, at least in part, by dramatic scenes seen in real-time around the world from large urban centers such as Wuhan, Milan, Madrid, and New York.
+In theory, there are good reasons to believe that higher density may have a positive association with the transmission of a contagious virus.
+It has long been known that the potential for inter-personal contact is greater in regions with higher density [see for example the research on urban fields and time-geography @Moore1970urban; @Moore1970some; @Farber2011running].
+Mathematically, models of exposure and contagion indicate that higher densities can catalyze the transmission of contagious diseases [@Rocklov2020high; @Li2018effect].
+Models such as these were likely at the root of messages, by some figures in positions of authority, that low density regions faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].
+
+As Rocklöv and Sjödin [@Rocklov2020high] note, however, mathematical models of contagion are valid at small-to-medium spatial scales (and presumably, small temporal scales too, such as time spent in restaurants, concert halls, cruises), and the results do not necessarily transfer to larger spatial units and different time scales.
+There are good reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers-however, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place.
+Nonetheless, the idea that high density correlates with high transmission is so intuitive that it is often taken for granted even at larger scales [e.g., @Cruz2020exploring; @Micallef2020first].
+At larger scales, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.
+
+A plausible behavioral adaptiation during a pandemic, especially one broadcast as widely and intensely as COVID-19, is risk compensation.
+Risk compensation is a process whereby people adjust their behavior in response to their _perception_ of risk [@Noland1995perceived; @Richens2000condoms; @Phillips2011risk].
+For example, it is possible that people who listened to the message of leaders saying that they were safe because of low density may not have taken adequate precautions.
+People in dense places who could more directly observe the impact of the pandemic may have become overly cautious.
+Both Paez et al.
+[-@Paez2020spatio] and Hamidi et al.
+[-@Hamidi2020density] posit this mechanism (i.e., greater compliance with social distancing in denser regions) to explain the results of their analyses.
+The evidence available does indeed show that there were important changes in behavior during the pandemic, at least with respect to mobility [@Jamal2020Changes; @Harris2021Changes; @Molloy2020Tracing]; furthermore, shelter in place orders may have had greater buy-in from the public in higher density regions [@Feyman2020effectiveness], and the behavior may have persisted beyond the duration of official social-distancing policies [@Praharaj2020Using].
+In addition, there is evidence that changes in mobility correlated with the trajectory of the pandemic [@Paez2020using; @Noland2021mobility].
+Given the potential for behavioral adaptation, the question of density becomes more nuanced: it is not just a matter of proximity, but also of human behavior, which is better studied using population-level data and models.
+
+In this respect, the literature to date remains inconclusive.
+
+On the one hand, there are studies that report positive associations between population density and various COVID-19-related outcomes.
+Bhadra [-@Bhadra2021impact], for example, reported a moderate positive correlation between the spread of COVID-19 and population density at the district level in India, however their analysis was bivariate and did not control for other variables, such as income.
+Similarly, Kadi and Khelfaoui [-@Kadi2020population] found a positive and significant correlation between number of cases and population density in cities in Algeria in a series of simple regression models (i.e., without other controls).
+A question in these relatively simple analyses is whether density is not a proxy for other factors.
+Other studies have included controls, such as Pequeno et al.
+[-@Pequeno2020air], a team that reported a positive association between density and cumulative counts of confirmed COVID-19 cases in state capitals in Brazil after controlling for covariates, including income, transport connectivity, and economic status.
+In a similar vein, Fielding-Miller et al.
+[-@Fielding2020social] reported a positive relationship between the absolute number of COVID-19 deaths and population density (rate) in rural counties in the US.
+Roy and Ghosh [-@Roy2020factors] used a battery of machine learning techniques to find discriminatory factors, and a positive and significant association between COVID-19 infection and death rates in US states.
+Wong and Li [-@Wong2020spreading] also found a positive and significant association between population density and number of confirmed COVID-19 cases in US counties, using both univariate and multivariate regressions with spatial effects.
+Most recently, Sy et al.
+[-@Sy2021population] reported that the basic reproductive number of COVID-19 in US counties tends to increase with population density, but at a decreasing rate at higher densities.
+
+On the gripping hand, a number of studies report non-significant or negative associations between population density and COVID-19 outcomes.
+This includes the research of Sun et al.
+[-@Sun2020impacts] who did not find evidence of significant correlation between population density and confirmed number of cases per day _in conditions of lockdown_ in China.
+This finding echoes the results of Paez et al.
+[-@Paez2020spatio], who in their study of provinces in Spain reported non-significant associations between population density and infection rates in the early days of the first wave of COVID-19, and negative significant associations in the later part of the first lockdown.
+Similarly, [-@Skorka2020macroecology] found zero or negative associations between population density and infection numbers/deaths by country.
+Fielding-Miller et al.
+[-@Fielding2020social] found a negative relationship between COVID-19 deaths and population density urban counties in the US.
+In their investigation of doubling time, White and Hébert-Dufresne [-@White2020state] identified a negative and significant correlation between population density and doubling time in US states.
+Likewise, [-@Khavarian2021high] fond a small negative (and significant) association between population density and COVID-19 morbidity in districts in Tehran.
+And two of the most complete studies in the US by Hamidi et al.
+[@Hamidi2020longitudinal; @Hamidi2020density] used an extensive set of controls to find negative and significant correlations between density at the level of counties in the US and COVID-19 cases and fatalities.
+
+As can be seen, these studies are implemented at different scales in different regions of the world.
+They also use a range of techniques, from correlation analysis, to multivariate regression, spatial regressions, and machine learning techniques.
+This is natural and to be expected: individual researchers have only limited time and expertise.
+This is why reproducibility is important.
+To pick an example (which will be further elaborated in the following sections), the study of Sy et al.
+[-@Sy2021population; hereafter SWN] would immediately grab the attention of a researcher with expertise in spatial modelling.
+Such an expert would likely ask some of the following questions: how were missing counties treated?
+Is it possible to spatially interpolate missing observations?
+What are the implications of the spatial sampling framework used in the analysis?
+Was there evidence of spatial autocorrelation in the residuals of the models?
+These are questions that in most cases would not occur to a researcher who has not been exposed to spatial statistics or spatial econometrics.
+Nonetheless, they are relevant and important.
+Fortunately, SWN give an example of a reasonably open, reproducible research product: their paper is accompanied by (most of) the data and (most of) the code used in the analysis.
+This means that an independent expert can, with only a moderate investment of time and effort, replicate the results in the paper, as well as ask additional questions.
+
+Alas, reproducibility is not necessarily the norm.
+
+There are various reasons why a project can fail to be reproducible.
+In some cases, there might be legitimate reasons to withhold the data, perhaps due to confidentiality and privacy reasons [e.g., @Lee2020human].
+But in many other cases the data are publicly available, as has been commonly the case with population-level COVID-19 information.
+Often the provenance of the data is documented, but the data themselves are not shared [e.g., @Amadu2021assessing; @Bhadra2021impact; @Cruz2020exploring; @Feng2020spread; @Fielding2020social; @Hamidi2020longitudinal; @Hamidi2020density; @Inbaraj2021seroprevalence; @Souris2020covid].
+As any researcher can attest, whether a graduate student or a seasoned scientist, collecting, organizing, and preparing data for a project can take a substantial amount of time.
+Pointing to the sources of data, even when these sources are public, is a small step towards reproducibility-but only a very small one.
+Faced with the prospect of having to recreate a data set from raw sources is probably sufficient to dissuade all but the most dedicated (or stubborn) researcher.
+This is true even if part of the data are shared [e.g., @Wong2020spreading].
+In other cases, data are shared, but the processes to document the preparation of the data are not fully documented [e.g., @Ahmad2020association; @Skorka2020macroecology].
+These processes matter, as shown by the errors in the spreadsheets of Reinhart and Rogoff [@Herndon2014high], and the data of biologist Jonathan Pruitt that led to an "avalanche" of paper retractions^[\url{https://doi.org/10.1038/d41586-020-00287-y}].
+Another situation is when papers share well-documented data, but fail to provide the code used in the analysis [e.g., @Noury2021how; @Pequeno2020air; @Wang2021transmission].
+Making code available only "on demand" [e.g., @Brandtner2021creatures] is an unnecessary barrier when most journals offer the facility to share supplemental materials online.
+Then there are those papers that strive towards reproducibility, sharing well-documented processes and data, as well as the code used in any analyses reported [e.g., @Paez2020spatio; @Feyman2020effectiveness; @Stephens2021impact; @White2020state].
+
+<!--
+@Ahmad2020association -> code is not shared, data are provided but data pre-processing is not documented
+@Amadu2021assessing -> code is not shared, data provenance is documented, data are not shared
+@Bhadra2021impact -> (code and data not available; only data provenance is documented)
+@Brandtner2021creatures -> code is available on demand, data are shared
+@Cruz2020exploring -> code is shared, sources of data are documented, data are not actually provided
+@Feng2020spread -> code not shared, data provenance is documented, data are not shared
+@Feyman2020effectiveness -> code and data are provided
+@Fielding2020social -> code is not shared, data provenance is documented, data are not actually provided
+@Hamidi2020longitudinal, @Hamidi2020density ->  (code and data not available)
+@Inbaraj2021seroprevalence -> code is not shared, data are not shared
+@Lee2020human -> code not provided, data sources are documented but data are not shared (privacy & confidentiality)
+@Noury2021how -> code is not shared, data are shared
+@Pequeno2020air -> (raw data supplied in suplemental files, code not available)
+@Roy2020factors -> code is not shared, data are provided and data processing is documented
+@Skorka2020macroecology -> code and data are shared but data processing is not documented, 
+@Souris2020covid -> code not provided, data sources are documented but data are not shared
+@Stephens2021impact -> code and data are shared
+@Wang2021transmission -> code is not shared, data are available
+@White2020state -> code and data are in repository
+@Wong2020spreading -> code not provided, data are shared but data preprocessing is not documented
+-->
+
+## Replicating SWN
+
+Fit (mixed) linear models as in Sy et al.:
+```{r incl}
+table1model1 <- lme(R ~ density_log , 
+                    random = ~ 1| state, 
+                    data = county_geo_clean %>%
+                      filter(R > 0))
+summary(table1model1)
+intervals(table1model1)
+```
+
+Linear mixed models of R0 and density + % private transportation
+```{r}
+table1model2 <- lme(R ~ density_log + private , 
+                    random = ~ 1| state, 
+                    data = county_geo_clean %>%
+                      filter(R > 0))
+summary(table1model2)
+intervals(table1model2)
+```
+
+Linear mixed models of R0 and density + % private transportation + median income
+```{r}
+table1model3 <- lme(R ~ density_log + private + hincome, 
+                    random = ~ 1| state, 
+                    data = county_geo_clean %>%
+                      filter(R > 0))
+summary(table1model3)
+intervals(table1model3)
+```
+
+Linear mixed models of R0 and density + % private transportation + median income + interaction of density and % private transportation
+```{r}
+table1model4 <- lme(R ~ density_log*private + hincome, 
+                    random = ~ 1| state, 
+                    data = county_geo_clean %>%
+                      filter(R > 0))
+summary(table1model4)
+intervals(table1model4)
+```
+
+# Some relevant questions
+
+Alas, despite decades' worth of developments in the field of geographical analysis, not all research presented to date has used proper methods for the study of COVID-19.
+While the answer to the question "do spatial effects really matter in regression analysis" was definitively answered in the positive at least 40 years ago, in practice many researchers continue to ignore the pitfalls of ignoring them.
+In this paper, I present a reanalysis of the data used by Sy et al.
+(2021) to study the correlations between the basic reproductive number of COVID-19 and population density in counties.
+I highlight two related issues: non-systematic sampling in space and spatial autocorrelation.
+The reanalysis is based on the use of tobit models to account for non-systematic sampling, and spatially autoregressive tobit models to account for spatial autocorrelation in the data generation process.
+The reanalysis highlights the importance of openness and reproducibility in COVID-19 research.
+Finally, the results provide a sobering example of the risks of not using appropriate methods in the analysis of geographical data.
+Furthermore,
+
+## Fit tobit version of models
+
+```{r}
+table1model1 <- censReg(R ~ density_log, 
+                        left = 0,
+                    data = county_geo_clean)
+summary(table1model1)
+```
+
+Linear mixed models of R0 and density + % private transportation
+```{r}
+table1model2 <- censReg(R ~ density_log + private, 
+                    data = county_geo_clean)
+summary(table1model2)
+```
+
+Linear mixed models of R0 and density + % private transportation + median income
+```{r}
+table1model3 <- censReg(R ~ density_log + private + hincome, 
+                    data = county_geo_clean)
+summary(table1model3)
+```
+
+Tobit models of R0 and density + % private transportation + median income + interaction of density and % private transportation
+```{r}
+table1model4 <- censReg(R ~ density_log*private + hincome, 
+                    data = county_geo_clean)
+summary(table1model4)
+
+```
+
+## Spatially autoregressive tobit
+
+Fit spatially autoregressive tobit:
+```{r}
+# Fit SAR Tobit
+fit_sartobit <- sartobit(R ~ density_log + private + hincome_log,
+                         B,
+                         ndraw = 1000,
+                         burn.in = 200, 
+                         showProgress = TRUE,
+                         data = county_geo_clean,
+                         computeMarginalEffects = TRUE)
+summary(fit_sartobit)
+impacts(fit_sartobit)
+```
+
+# Conclusion
+
+Words go here.
 

### 2021-05-19 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
First draft of introduction and background

@@
 
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+
 % Pandoc citation processing
@@
 {\Large
-\textbf\newline{Population density and basic reproductive number of
-COVID-19 across United States counties
-revisited} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{The importance of reproducibility in COVID-19 research:
+the case of population density and the spread of the
+pandemic} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 \section*{Abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
-porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
-dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
-velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
-finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
-diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
-ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
-pulvinar, volutpat risus.
+The emergence of the novel SARS-CoV-2 coronavirus and the global
+COVID-19 pandemic has led to explosive growth in scientific research.
+Given the high stakes of the situation, it is essential that scientific
+activites, on which good policy depends, are as transparent and
+reproducible as possible. Reproducibility is key for the efficient
+operation of the self-correction mechanisms of science, which work to
+weed out errors and refine our understanding of social and physical
+phenomena. In this paper, the importance of reproducibility is
+illustrated for the case of the association between population density
+and the the spread of SARS-CoV-2. Transparency and openness means that
+the same problem can, with relatively modest efforts, be examined by
+independent researchers who can verify findings, and bring to bear
+different perspectives, approaches, and methods---sometimes with
+consequantial changes in the conclusions, as the empirical example in
+this paper shows.
 
@@
 % Author Summary not valid for PLOS ONE submissions.
-\section*{Author summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
-porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
-dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
-velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
-finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
-diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
-ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
-pulvinar, volutpat risus.
 
@@
 % Use "Eq" instead of "Equation" for equation citations.
-\emph{Text based on plos sample manuscript, see
-\url{https://journals.plos.org/ploscompbiol/s/latex}}
-
 \hypertarget{introduction}{%
@@
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
-porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
-dui eu ante cursus gravida non sed sem. Phasellus mauris velit, dapibus
-finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
-diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
-ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
-pulvinar, volutpat risus.
+The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the
+global pandemic that followed in its wake, led to an explosive growth of
+research. According to Fraser et al. {[}1{]}, over 125,000
+COVID-19-related papers were released in the first ten months from the
+first confirmed case of the disease. Of these, more than 30,000 were
+shared in pre-print servers, the use of which also exploded in the past
+year {[}2--4{]}.
 
-A list
+Given the heavy human and economic cost of the pandemic, there has been
+a natural tension in the scientific community between the need to
+publish research results quickly and the imperative to maintain
+consistently high quality standards in scientific reporting; indeed, a
+call for maintaining the standards in published research has even called
+this deluge of publications a ``carnage of substandard research''
+{[}5{]}. Part of the challenge of maintaining quality standards in
+published research is that, despite an abundance of recommendations and
+guidelines {[}6--9{]}, in practice reproducibility has remained a lofty
+but somewhat aspirational goal {[}10,11{]}. As reported in the
+literature, only a woefully small proportion of published research was
+actually reproducible before the pandemic {[}12,13{]}, a situation that
+does not appear to have changed substantially since {[}14,15{]}.
 
-\begin{itemize}
-\tightlist
-\item
-  Item 1
-\item
-  Item 2
-\end{itemize}
+The push for open data and software, along with more strenuous efforts
+towards open, reproducible research, is simply a continuation of
+long-standing scientific practices of independent verification. Despite
+the (at times disproportionate) attention that high profile scandals in
+science tend to elicit in the media, science as a collective endeavor is
+remarkable for being a self-correcting enterprise, one with built-in
+mechanisms and incentives to weed out erroneous ideas. Over the long
+term, facts tend to prevail in science. At stake is the shorter-term
+impacts that research may have in other spheres of economic and social
+life. The case of economists Reinhart and Rogoff comes to mind: by the
+time the inaccuracies in their research were uncovered {[}see 16{]},
+their claims about debt and economic growth had already been seized by
+policy-makers on both sides of the Atlantic to justify austerity
+policies in the aftermath of the Great Recession of
+2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
+  ``Reinhart--Rogoff may have had more immediate influence on public
+  debate than any previous paper in the history of economics''
+  \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
+As later research has demonstrated, those policies cast a long shadow,
+and their sequels continued to be felt for years {[}17{]}.
 
-Here are two sample references: {[}1,2{]}.
+In the context of COVID-19, a topic that has grabbed the imagination of
+numerous thinkers has been the prospect of life in cities after the
+pandemic {[}18{]}. The fact that the worst of the pandemic was initially
+felt in dense population centers such as Wuhan, Milan, Madrid, and New
+York, prompted a flurry of research into the associations between
+density and the spread of the pandemic. Some important questions hang on
+the results of these research efforts. For example, are lower density
+regions safer from the pandemic? Are de-densification policies
+warranted, at least in the short term? And in the longer term, will the
+risks of life in high density regions presage a flight from cities? Over
+the past year, numerous papers have sought to throw light into the
+underlying issue of density and the pandemic; nonetheless the results,
+as will be detailed next, remain mixed. Further, to complicate matters,
+precious few of these studies appear to be sufficiently open to support
+independent verification.
+
+The objective of this paper is to illustrate the importance of
+reproducibility in research, particularly in the context of the flood of
+COVID-19 papers. To this end, a recent study by Sy et al. {[}19{]} is
+chosen as an example of reproducible research. The objective is not to
+malign the analysis of these researchers, but rather to demonstrate the
+value of openness to allow for independent verification and further
+analysis. Open data and open code mean that an independent researcher
+can, with only modest efforts, not only verify the findings reported,
+but also examine the same data from a perspective which may not have
+been available to the original researchers due to differences in
+disciplinary perspectives, methodological traditions, and/or training,
+among other possible factors. The example, which shows consequential
+changes in the conclusions reached by different analyses, should serve
+as a call to researchers to redouble their efforts to increase
+transparency and reproducibility in research. This paper, in addition,
+aims to show how data can be packaged in well-documented, shareable
+units, and code can be embedded into self-contained documents suitable
+for review and independent verification. The source for this paper is an
+\href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
+with the data package, is available in a public repository\footnote{\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}}.
+
+\hypertarget{background}{%
+\section{Background}\label{background}}
+
+The concern with population density and the spread of the virus during
+the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
+seen in real-time around the world from large urban centers such as
+Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
+believe that higher density may have a positive association with the
+transmission of a contagious virus. It has long been known that the
+potential for inter-personal contact is greater in regions with higher
+density {[}see for example the research on urban fields and
+time-geography 20,21,22{]}. Mathematically, models of exposure and
+contagion indicate that higher densities can catalyze the transmission
+of contagious diseases {[}23,24{]}. Models such as these were likely at
+the root of messages, by some figures in positions of authority, that
+low density regions faced lower risks from the pandemic\footnote{Governor
+  Kristi Noem of South Dakota, for example, claimed that sparse
+  population density allowed her state to face the pandemic without the
+  need for strict policy interventions
+  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.
+
+As Rocklöv and Sjödin {[}23{]} note, however, mathematical models of
+contagion are valid at small-to-medium spatial scales (and presumably,
+small temporal scales too, such as time spent in restaurants, concert
+halls, cruises), and the results do not necessarily transfer to larger
+spatial units and different time scales. There are good reasons for
+this: while in a restaurant, one can hardly avoid being in proximity to
+other customers-however, a person can choose to (or be forced to as a
+matter of policy) not go to a restaurant in the first place.
+Nonetheless, the idea that high density correlates with high
+transmission is so intuitive that it is often taken for granted even at
+larger scales {[}e.g., 25,26{]}. At larger scales, however, there exists
+the possibility of behavioral adaptations, which are difficult to
+capture in the mechanistic framework of differential equations {[}or can
+be missing in agent-based models, 27{]}; these adaptations, in fact, can
+be a key aspect of disease transmission.
+
+A plausible behavioral adaptiation during a pandemic, especially one
+broadcast as widely and intensely as COVID-19, is risk compensation.
+Risk compensation is a process whereby people adjust their behavior in
+response to their \emph{perception} of risk {[}28--30{]}. For example,
+it is possible that people who listened to the message of leaders saying
+that they were safe because of low density may not have taken adequate
+precautions. People in dense places who could more directly observe the
+impact of the pandemic may have become overly cautious. Both Paez et al.
+{[}31{]} and Hamidi et al. {[}32{]} posit this mechanism (i.e., greater
+compliance with social distancing in denser regions) to explain the
+results of their analyses. The evidence available does indeed show that
+there were important changes in behavior during the pandemic, at least
+with respect to mobility {[}33--35{]}; furthermore, shelter in place
+orders may have had greater buy-in from the public in higher density
+regions {[}36{]}, and the behavior may have persisted beyond the
+duration of official social-distancing policies {[}37{]}. In addition,
+there is evidence that changes in mobility correlated with the
+trajectory of the pandemic {[}38,39{]}. Given the potential for
+behavioral adaptation, the question of density becomes more nuanced: it
+is not just a matter of proximity, but also of human behavior, which is
+better studied using population-level data and models.
+
+In this respect, the literature to date remains inconclusive.
+
+On the one hand, there are studies that report positive associations
+between population density and various COVID-19-related outcomes. Bhadra
+{[}40{]}, for example, reported a moderate positive correlation between
+the spread of COVID-19 and population density at the district level in
+India, however their analysis was bivariate and did not control for
+other variables, such as income. Similarly, Kadi and Khelfaoui {[}41{]}
+found a positive and significant correlation between number of cases and
+population density in cities in Algeria in a series of simple regression
+models (i.e., without other controls). A question in these relatively
+simple analyses is whether density is not a proxy for other factors.
+Other studies have included controls, such as Pequeno et al. {[}42{]}, a
+team that reported a positive association between density and cumulative
+counts of confirmed COVID-19 cases in state capitals in Brazil after
+controlling for covariates, including income, transport connectivity,
+and economic status. In a similar vein, Fielding-Miller et al. {[}43{]}
+reported a positive relationship between the absolute number of COVID-19
+deaths and population density (rate) in rural counties in the US. Roy
+and Ghosh {[}44{]} used a battery of machine learning techniques to find
+discriminatory factors, and a positive and significant association
+between COVID-19 infection and death rates in US states. Wong and Li
+{[}45{]} also found a positive and significant association between
+population density and number of confirmed COVID-19 cases in US
+counties, using both univariate and multivariate regressions with
+spatial effects. Most recently, Sy et al. {[}19{]} reported that the
+basic reproductive number of COVID-19 in US counties tends to increase
+with population density, but at a decreasing rate at higher densities.
+
+On the gripping hand, a number of studies report non-significant or
+negative associations between population density and COVID-19 outcomes.
+This includes the research of Sun et al. {[}46{]} who did not find
+evidence of significant correlation between population density and
+confirmed number of cases per day \emph{in conditions of lockdown} in
+China. This finding echoes the results of Paez et al. {[}31{]}, who in
+their study of provinces in Spain reported non-significant associations
+between population density and infection rates in the early days of the
+first wave of COVID-19, and negative significant associations in the
+later part of the first lockdown. Similarly, {[}47{]} found zero or
+negative associations between population density and infection
+numbers/deaths by country. Fielding-Miller et al. {[}43{]} found a
+negative relationship between COVID-19 deaths and population density
+urban counties in the US. In their investigation of doubling time, White
+and Hébert-Dufresne {[}48{]} identified a negative and significant
+correlation between population density and doubling time in US states.
+Likewise, {[}49{]} fond a small negative (and significant) association
+between population density and COVID-19 morbidity in districts in
+Tehran. And two of the most complete studies in the US by Hamidi et al.
+{[}32,50{]} used an extensive set of controls to find negative and
+significant correlations between density at the level of counties in the
+US and COVID-19 cases and fatalities.
+
+As can be seen, these studies are implemented at different scales in
+different regions of the world. They also use a range of techniques,
+from correlation analysis, to multivariate regression, spatial
+regressions, and machine learning techniques. This is natural and to be
+expected: individual researchers have only limited time and expertise.
+This is why reproducibility is important. To pick an example (which will
+be further elaborated in the following sections), the study of Sy et
+al.~{[}{[}19{]}; hereafter SWN{]} would immediately grab the attention
+of a researcher with expertise in spatial modelling. Such an expert
+would likely ask some of the following questions: how were missing
+counties treated? Is it possible to spatially interpolate missing
+observations? What are the implications of the spatial sampling
+framework used in the analysis? Was there evidence of spatial
+autocorrelation in the residuals of the models? These are questions that
+in most cases would not occur to a researcher who has not been exposed
+to spatial statistics or spatial econometrics. Nonetheless, they are
+relevant and important. Fortunately, SWN give an example of a reasonably
+open, reproducible research product: their paper is accompanied by (most
+of) the data and (most of) the code used in the analysis. This means
+that an independent expert can, with only a moderate investment of time
+and effort, replicate the results in the paper, as well as ask
+additional questions.
+
+Alas, reproducibility is not necessarily the norm.
+
+There are various reasons why a project can fail to be reproducible. In
+some cases, there might be legitimate reasons to withhold the data,
+perhaps due to confidentiality and privacy reasons {[}e.g., 51{]}. But
+in many other cases the data are publicly available, as has been
+commonly the case with population-level COVID-19 information. Often the
+provenance of the data is documented, but the data themselves are not
+shared {[}25,32,40,43,50,e.g., 52,53--55{]}. As any researcher can
+attest, whether a graduate student or a seasoned scientist, collecting,
+organizing, and preparing data for a project can take a substantial
+amount of time. Pointing to the sources of data, even when these sources
+are public, is a small step towards reproducibility-but only a very
+small one. Faced with the prospect of having to recreate a data set from
+raw sources is probably sufficient to dissuade all but the most
+dedicated (or stubborn) researcher. This is true even if part of the
+data are shared {[}e.g., 45{]}. In other cases, data are shared, but the
+processes to document the preparation of the data are not fully
+documented {[}47,e.g., 56{]}. These processes matter, as shown by the
+errors in the spreadsheets of Reinhart and Rogoff {[}16{]}, and the data
+of biologist Jonathan Pruitt that led to an ``avalanche'' of paper
+retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
+Another situation is when papers share well-documented data, but fail to
+provide the code used in the analysis {[}42,e.g., 57,58{]}. Making code
+available only ``on demand'' {[}e.g., 59{]} is an unnecessary barrier
+when most journals offer the facility to share supplemental materials
+online. Then there are those papers that strive towards reproducibility,
+sharing well-documented processes and data, as well as the code used in
+any analyses reported {[}e.g., 31,36,48,60{]}.
+
+\hypertarget{replicating-swn}{%
+\subsection{Replicating SWN}\label{replicating-swn}}
+
+Fit (mixed) linear models as in Sy et al.:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{table1model1 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log , }
+                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
+                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
+                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
+\FunctionTok{summary}\NormalTok{(table1model1)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Linear mixed-effects model fit by REML
+##   Data: county_geo_clean %>% filter(R > 0) 
+##        AIC      BIC    logLik
+##   2383.621 2403.808 -1187.811
+## 
+## Random effects:
+##  Formula: ~1 | state
+##         (Intercept)  Residual
+## StdDev:    0.165504 0.6648325
+## 
+## Fixed effects:  R ~ density_log 
+##                 Value  Std.Error   DF  t-value p-value
+## (Intercept) 2.2740045 0.05468629 1099 41.58272       0
+## density_log 0.1621127 0.01485543 1099 10.91269       0
+##  Correlation: 
+##             (Intr)
+## density_log 0.797 
+## 
+## Standardized Within-Group Residuals:
+##        Min         Q1        Med         Q3        Max 
+## -2.3847932 -0.5972261 -0.1682299  0.3681864 16.2716371 
+## 
+## Number of Observations: 1151
+## Number of Groups: 51
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{intervals}\NormalTok{(table1model1)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Approximate 95% confidence intervals
+## 
+##  Fixed effects:
+##                 lower      est.     upper
+## (Intercept) 2.1667032 2.2740045 2.3813058
+## density_log 0.1329645 0.1621127 0.1912609
+## attr(,"label")
+## [1] "Fixed effects:"
+## 
+##  Random Effects:
+##   Level: state 
+##                     lower     est.     upper
+## sd((Intercept)) 0.1080424 0.165504 0.2535261
+## 
+##  Within-group standard error:
+##     lower      est.     upper 
+## 0.6375478 0.6648325 0.6932849
+\end{verbatim}
+
+Linear mixed models of R0 and density + \% private transportation
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{table1model2 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private , }
+                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
+                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
+                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
+\FunctionTok{summary}\NormalTok{(table1model2)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Linear mixed-effects model fit by REML
+##   Data: county_geo_clean %>% filter(R > 0) 
+##        AIC      BIC    logLik
+##   2385.478 2410.707 -1187.739
+## 
+## Random effects:
+##  Formula: ~1 | state
+##         (Intercept)  Residual
+## StdDev:   0.1358322 0.6649714
+## 
+## Fixed effects:  R ~ density_log + private 
+##                 Value Std.Error   DF   t-value p-value
+## (Intercept)  3.347070 0.3418910 1098  9.789874  0.0000
+## density_log  0.145258 0.0156188 1098  9.300177  0.0000
+## private     -0.012547 0.0039434 1098 -3.181760  0.0015
+##  Correlation: 
+##             (Intr) dnsty_
+## density_log -0.222       
+## private     -0.988  0.344
+## 
+## Standardized Within-Group Residuals:
+##        Min         Q1        Med         Q3        Max 
+## -2.7347592 -0.6005758 -0.1610880  0.3876853 16.2104980 
+## 
+## Number of Observations: 1151
+## Number of Groups: 51
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{intervals}\NormalTok{(table1model2)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Approximate 95% confidence intervals
+## 
+##  Fixed effects:
+##                   lower       est.        upper
+## (Intercept)  2.67623619  3.3470697  4.017903160
+## density_log  0.11461177  0.1452579  0.175904019
+## private     -0.02028416 -0.0125468 -0.004809444
+## attr(,"label")
+## [1] "Fixed effects:"
+## 
+##  Random Effects:
+##   Level: state 
+##                      lower      est.    upper
+## sd((Intercept)) 0.08059718 0.1358322 0.228921
+## 
+##  Within-group standard error:
+##     lower      est.     upper 
+## 0.6376445 0.6649714 0.6934695
+\end{verbatim}
+
+Linear mixed models of R0 and density + \% private transportation +
+median income
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{table1model3 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private }\SpecialCharTok{+}\NormalTok{ hincome, }
+                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
+                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
+                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
+\FunctionTok{summary}\NormalTok{(table1model3)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Linear mixed-effects model fit by REML
+##   Data: county_geo_clean %>% filter(R > 0) 
+##       AIC     BIC   logLik
+##   2393.98 2424.25 -1190.99
+## 
+## Random effects:
+##  Formula: ~1 | state
+##         (Intercept)  Residual
+## StdDev:   0.1373995 0.6651151
+## 
+## Fixed effects:  R ~ density_log + private + hincome 
+##                 Value Std.Error   DF   t-value p-value
+## (Intercept)  3.385550 0.3933810 1097  8.606289  0.0000
+## density_log  0.146826 0.0171534 1097  8.559585  0.0000
+## private     -0.012707 0.0040548 1097 -3.133786  0.0018
+## hincome     -0.003256 0.0151105 1097 -0.215493  0.8294
+##  Correlation: 
+##             (Intr) dnsty_ privat
+## density_log  0.027              
+## private     -0.950  0.211       
+## hincome     -0.493 -0.412  0.228
+## 
+## Standardized Within-Group Residuals:
+##        Min         Q1        Med         Q3        Max 
+## -2.7162603 -0.5984817 -0.1626793  0.3887931 16.2029477 
+## 
+## Number of Observations: 1151
+## Number of Groups: 51
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{intervals}\NormalTok{(table1model3)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Approximate 95% confidence intervals
+## 
+##  Fixed effects:
+##                   lower         est.       upper
+## (Intercept)  2.61368596  3.385550103  4.15741424
+## density_log  0.11316880  0.146825984  0.18048317
+## private     -0.02066315 -0.012707018 -0.00475089
+## hincome     -0.03290488 -0.003256195  0.02639249
+## attr(,"label")
+## [1] "Fixed effects:"
+## 
+##  Random Effects:
+##   Level: state 
+##                    lower      est.     upper
+## sd((Intercept)) 0.081461 0.1373995 0.2317504
+## 
+##  Within-group standard error:
+##     lower      est.     upper 
+## 0.6377582 0.6651151 0.6936455
+\end{verbatim}
+
+Linear mixed models of R0 and density + \% private transportation +
+median income + interaction of density and \% private transportation
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{table1model4 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log}\SpecialCharTok{*}\NormalTok{private }\SpecialCharTok{+}\NormalTok{ hincome, }
+                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
+                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
+                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
+\FunctionTok{summary}\NormalTok{(table1model4)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Linear mixed-effects model fit by REML
+##   Data: county_geo_clean %>% filter(R > 0) 
+##       AIC      BIC   logLik
+##   2407.04 2442.349 -1196.52
+## 
+## Random effects:
+##  Formula: ~1 | state
+##         (Intercept)  Residual
+## StdDev:   0.1378096 0.6653589
+## 
+## Fixed effects:  R ~ density_log * private + hincome 
+##                         Value Std.Error   DF   t-value p-value
+## (Intercept)          3.415859 0.4157861 1096  8.215424  0.0000
+## density_log          0.177645 0.1350641 1096  1.315266  0.1887
+## private             -0.013123 0.0044568 1096 -2.944502  0.0033
+## hincome             -0.002719 0.0153117 1096 -0.177607  0.8591
+## density_log:private -0.000354 0.0015414 1096 -0.229894  0.8182
+##  Correlation: 
+##                     (Intr) dnsty_ privat hincom
+## density_log          0.323                     
+## private             -0.952 -0.386              
+## hincome             -0.409  0.105  0.139       
+## density_log:private -0.323 -0.992  0.414 -0.158
+## 
+## Standardized Within-Group Residuals:
+##        Min         Q1        Med         Q3        Max 
+## -2.7371582 -0.5968733 -0.1638499  0.3883700 16.1980628 
+## 
+## Number of Observations: 1151
+## Number of Groups: 51
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{intervals}\NormalTok{(table1model4)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Approximate 95% confidence intervals
+## 
+##  Fixed effects:
+##                            lower          est.        upper
+## (Intercept)          2.600032192  3.4158588336  4.231685475
+## density_log         -0.087368297  0.1776452292  0.442658756
+## private             -0.021867960 -0.0131231030 -0.004378246
+## hincome             -0.032762972 -0.0027194668  0.027324038
+## density_log:private -0.003378809 -0.0003543603  0.002670088
+## attr(,"label")
+## [1] "Fixed effects:"
+## 
+##  Random Effects:
+##   Level: state 
+##                      lower      est.     upper
+## sd((Intercept)) 0.08169505 0.1378096 0.2324679
+## 
+##  Within-group standard error:
+##     lower      est.     upper 
+## 0.6379755 0.6653589 0.6939176
+\end{verbatim}
+
+\hypertarget{some-relevant-questions}{%
+\section{Some relevant questions}\label{some-relevant-questions}}
+
+Alas, despite decades' worth of developments in the field of
+geographical analysis, not all research presented to date has used
+proper methods for the study of COVID-19. While the answer to the
+question ``do spatial effects really matter in regression analysis'' was
+definitively answered in the positive at least 40 years ago, in practice
+many researchers continue to ignore the pitfalls of ignoring them. In
+this paper, I present a reanalysis of the data used by Sy et al.~(2021)
+to study the correlations between the basic reproductive number of
+COVID-19 and population density in counties. I highlight two related
+issues: non-systematic sampling in space and spatial autocorrelation.
+The reanalysis is based on the use of tobit models to account for
+non-systematic sampling, and spatially autoregressive tobit models to
+account for spatial autocorrelation in the data generation process. The
+reanalysis highlights the importance of openness and reproducibility in
+COVID-19 research. Finally, the results provide a sobering example of
+the risks of not using appropriate methods in the analysis of
+geographical data. Furthermore,
+
+\hypertarget{fit-tobit-version-of-models}{%
+\subsection{Fit tobit version of
+models}\label{fit-tobit-version-of-models}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{table1model1 }\OtherTok{\textless{}{-}} \FunctionTok{censReg}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log, }
+                        \AttributeTok{left =} \DecValTok{0}\NormalTok{,}
+                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean)}
+\FunctionTok{summary}\NormalTok{(table1model1)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## Call:
+## censReg(formula = R ~ density_log, left = 0, data = county_geo_clean)
+## 
+## Observations:
+##          Total  Left-censored     Uncensored Right-censored 
+##           3218           2067           1151              0 
+## 
+## Coefficients:
+##             Estimate Std. error t value Pr(> t)    
+## (Intercept)  2.33967    0.09232   25.34  <2e-16 ***
+## density_log  0.74323    0.02672   27.82  <2e-16 ***
+## logSigma     0.56894    0.02396   23.74  <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Newton-Raphson maximisation, 7 iterations
+## Return code 8: successive function values within relative tolerance limit (reltol)
+## Log-likelihood: -3246.5 on 3 Df
+\end{verbatim}
+
+Linear mixed models of R0 and density + \% private transportation
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{table1model2 }\OtherTok{\textless{}{-}} \FunctionTok{censReg}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private, }
+                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean)}
+\FunctionTok{summary}\NormalTok{(table1model2)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## Call:
+## censReg(formula = R ~ density_log + private, data = county_geo_clean)
+## 
+## Observations:
+##          Total  Left-censored     Uncensored Right-censored 
+##           3218           2067           1151              0 
+## 
+## Coefficients:
+##             Estimate Std. error t value Pr(> t)    
+## (Intercept) 0.825319   0.583014   1.416 0.15689    
+## density_log 0.754516   0.027546  27.391 < 2e-16 ***
+## private     0.017186   0.006548   2.624 0.00868 ** 
+## logSigma    0.570248   0.023977  23.783 < 2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Newton-Raphson maximisation, 7 iterations
+## Return code 8: successive function values within relative tolerance limit (reltol)
+## Log-likelihood: -3242.96 on 4 Df
+\end{verbatim}
+
+Linear mixed models of R0 and density + \% private transportation +
+median income
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{table1model3 }\OtherTok{\textless{}{-}} \FunctionTok{censReg}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private }\SpecialCharTok{+}\NormalTok{ hincome, }
+                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean)}
+\FunctionTok{summary}\NormalTok{(table1model3)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## Call:
+## censReg(formula = R ~ density_log + private + hincome, data = county_geo_clean)
+## 
+## Observations:
+##          Total  Left-censored     Uncensored Right-censored 
+##           3218           2067           1151              0 
+## 
+## Coefficients:
+##              Estimate Std. error t value  Pr(> t)    
+## (Intercept) -3.201908   0.680988  -4.702 2.58e-06 ***
+## density_log  0.662526   0.026876  24.651  < 2e-16 ***
+## private      0.040886   0.006844   5.974 2.32e-09 ***
+## hincome      0.297763   0.025726  11.574  < 2e-16 ***
+## logSigma     0.531694   0.023920  22.228  < 2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Newton-Raphson maximisation, 7 iterations
+## Return code 8: successive function values within relative tolerance limit (reltol)
+## Log-likelihood: -3174.96 on 5 Df
+\end{verbatim}
+
+Tobit models of R0 and density + \% private transportation + median
+income + interaction of density and \% private transportation
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{table1model4 }\OtherTok{\textless{}{-}} \FunctionTok{censReg}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log}\SpecialCharTok{*}\NormalTok{private }\SpecialCharTok{+}\NormalTok{ hincome, }
+                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean)}
+\FunctionTok{summary}\NormalTok{(table1model4)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## Call:
+## censReg(formula = R ~ density_log * private + hincome, data = county_geo_clean)
+## 
+## Observations:
+##          Total  Left-censored     Uncensored Right-censored 
+##           3218           2067           1151              0 
+## 
+## Coefficients:
+##                      Estimate Std. error t value  Pr(> t)    
+## (Intercept)         -3.687016   0.733410  -5.027 4.98e-07 ***
+## density_log          0.311841   0.174521   1.787   0.0740 .  
+## private              0.047454   0.007712   6.154 7.58e-10 ***
+## hincome              0.291536   0.025807  11.297  < 2e-16 ***
+## density_log:private  0.004097   0.002028   2.020   0.0434 *  
+## logSigma             0.530095   0.023924  22.158  < 2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Newton-Raphson maximisation, 7 iterations
+## Return code 8: successive function values within relative tolerance limit (reltol)
+## Log-likelihood: -3173.008 on 6 Df
+\end{verbatim}
+
+\hypertarget{spatially-autoregressive-tobit}{%
+\subsection{Spatially autoregressive
+tobit}\label{spatially-autoregressive-tobit}}
+
+Fit spatially autoregressive tobit:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# Fit SAR Tobit}
+\NormalTok{fit\_sartobit }\OtherTok{\textless{}{-}} \FunctionTok{sartobit}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private }\SpecialCharTok{+}\NormalTok{ hincome\_log,}
+\NormalTok{                         B,}
+                         \AttributeTok{ndraw =} \DecValTok{1000}\NormalTok{,}
+                         \AttributeTok{burn.in =} \DecValTok{200}\NormalTok{, }
+                         \AttributeTok{showProgress =} \ConstantTok{TRUE}\NormalTok{,}
+                         \AttributeTok{data =}\NormalTok{ county\_geo\_clean,}
+                         \AttributeTok{computeMarginalEffects =} \ConstantTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in model.matrix.default(mt, mf, contrasts): non-list contrasts argument
+## ignored
+\end{verbatim}
+
+\begin{verbatim}
+## Warning: Function SE_classic_setup moved to the spatialreg package
+\end{verbatim}
+
+\begin{verbatim}
+## Registered S3 methods overwritten by 'spatialreg':
+##   method                   from 
+##   residuals.stsls          spdep
+##   deviance.stsls           spdep
+##   coef.stsls               spdep
+##   print.stsls              spdep
+##   summary.stsls            spdep
+##   print.summary.stsls      spdep
+##   residuals.gmsar          spdep
+##   deviance.gmsar           spdep
+##   coef.gmsar               spdep
+##   fitted.gmsar             spdep
+##   print.gmsar              spdep
+##   summary.gmsar            spdep
+##   print.summary.gmsar      spdep
+##   print.lagmess            spdep
+##   summary.lagmess          spdep
+##   print.summary.lagmess    spdep
+##   residuals.lagmess        spdep
+##   deviance.lagmess         spdep
+##   coef.lagmess             spdep
+##   fitted.lagmess           spdep
+##   logLik.lagmess           spdep
+##   fitted.SFResult          spdep
+##   print.SFResult           spdep
+##   fitted.ME_res            spdep
+##   print.ME_res             spdep
+##   print.lagImpact          spdep
+##   plot.lagImpact           spdep
+##   summary.lagImpact        spdep
+##   HPDinterval.lagImpact    spdep
+##   print.summary.lagImpact  spdep
+##   print.sarlm              spdep
+##   summary.sarlm            spdep
+##   residuals.sarlm          spdep
+##   deviance.sarlm           spdep
+##   coef.sarlm               spdep
+##   vcov.sarlm               spdep
+##   fitted.sarlm             spdep
+##   logLik.sarlm             spdep
+##   anova.sarlm              spdep
+##   predict.sarlm            spdep
+##   print.summary.sarlm      spdep
+##   print.sarlm.pred         spdep
+##   as.data.frame.sarlm.pred spdep
+##   residuals.spautolm       spdep
+##   deviance.spautolm        spdep
+##   coef.spautolm            spdep
+##   fitted.spautolm          spdep
+##   print.spautolm           spdep
+##   summary.spautolm         spdep
+##   logLik.spautolm          spdep
+##   print.summary.spautolm   spdep
+##   print.WXImpact           spdep
+##   summary.WXImpact         spdep
+##   print.summary.WXImpact   spdep
+##   predict.SLX              spdep
+\end{verbatim}
+
+\begin{verbatim}
+##   |                                                                              |                                                                      |   0%  |                                                                              |                                                                      |   1%  |                                                                              |=                                                                     |   1%  |                                                                              |=                                                                     |   2%  |                                                                              |==                                                                    |   2%  |                                                                              |==                                                                    |   3%  |                                                                              |==                                                                    |   4%  |                                                                              |===                                                                   |   4%  |                                                                              |===                                                                   |   5%  |                                                                              |====                                                                  |   5%  |                                                                              |====                                                                  |   6%  |                                                                              |=====                                                                 |   6%  |                                                                              |=====                                                                 |   7%  |                                                                              |=====                                                                 |   8%  |                                                                              |======                                                                |   8%  |                                                                              |======                                                                |   9%  |                                                                              |=======                                                               |   9%  |                                                                              |=======                                                               |  10%  |                                                                              |=======                                                               |  11%  |                                                                              |========                                                              |  11%  |                                                                              |========                                                              |  12%  |                                                                              |=========                                                             |  12%  |                                                                              |=========                                                             |  13%  |                                                                              |=========                                                             |  14%  |                                                                              |==========                                                            |  14%  |                                                                              |==========                                                            |  15%  |                                                                              |===========                                                           |  15%  |                                                                              |===========                                                           |  16%  |                                                                              |============                                                          |  16%  |                                                                              |============                                                          |  17%  |                                                                              |============                                                          |  18%  |                                                                              |=============                                                         |  18%  |                                                                              |=============                                                         |  19%  |                                                                              |==============                                                        |  19%  |                                                                              |==============                                                        |  20%  |                                                                              |==============                                                        |  21%  |                                                                              |===============                                                       |  21%  |                                                                              |===============                                                       |  22%  |                                                                              |================                                                      |  22%  |                                                                              |================                                                      |  23%  |                                                                              |================                                                      |  24%  |                                                                              |=================                                                     |  24%  |                                                                              |=================                                                     |  25%  |                                                                              |==================                                                    |  25%  |                                                                              |==================                                                    |  26%  |                                                                              |===================                                                   |  26%  |                                                                              |===================                                                   |  27%  |                                                                              |===================                                                   |  28%  |                                                                              |====================                                                  |  28%  |                                                                              |====================                                                  |  29%  |                                                                              |=====================                                                 |  29%  |                                                                              |=====================                                                 |  30%  |                                                                              |=====================                                                 |  31%  |                                                                              |======================                                                |  31%  |                                                                              |======================                                                |  32%  |                                                                              |=======================                                               |  32%  |                                                                              |=======================                                               |  33%  |                                                                              |=======================                                               |  34%  |                                                                              |========================                                              |  34%  |                                                                              |========================                                              |  35%  |                                                                              |=========================                                             |  35%  |                                                                              |=========================                                             |  36%  |                                                                              |==========================                                            |  36%  |                                                                              |==========================                                            |  37%  |                                                                              |==========================                                            |  38%  |                                                                              |===========================                                           |  38%  |                                                                              |===========================                                           |  39%  |                                                                              |============================                                          |  39%  |                                                                              |============================                                          |  40%  |                                                                              |============================                                          |  41%  |                                                                              |=============================                                         |  41%  |                                                                              |=============================                                         |  42%  |                                                                              |==============================                                        |  42%  |                                                                              |==============================                                        |  43%  |                                                                              |==============================                                        |  44%  |                                                                              |===============================                                       |  44%  |                                                                              |===============================                                       |  45%  |                                                                              |================================                                      |  45%  |                                                                              |================================                                      |  46%  |                                                                              |=================================                                     |  46%  |                                                                              |=================================                                     |  47%  |                                                                              |=================================                                     |  48%  |                                                                              |==================================                                    |  48%  |                                                                              |==================================                                    |  49%  |                                                                              |===================================                                   |  49%  |                                                                              |===================================                                   |  50%  |                                                                              |===================================                                   |  51%  |                                                                              |====================================                                  |  51%  |                                                                              |====================================                                  |  52%  |                                                                              |=====================================                                 |  52%  |                                                                              |=====================================                                 |  53%  |                                                                              |=====================================                                 |  54%  |                                                                              |======================================                                |  54%  |                                                                              |======================================                                |  55%  |                                                                              |=======================================                               |  55%  |                                                                              |=======================================                               |  56%  |                                                                              |========================================                              |  56%  |                                                                              |========================================                              |  57%  |                                                                              |========================================                              |  58%  |                                                                              |=========================================                             |  58%  |                                                                              |=========================================                             |  59%  |                                                                              |==========================================                            |  59%  |                                                                              |==========================================                            |  60%  |                                                                              |==========================================                            |  61%  |                                                                              |===========================================                           |  61%  |                                                                              |===========================================                           |  62%  |                                                                              |============================================                          |  62%  |                                                                              |============================================                          |  63%  |                                                                              |============================================                          |  64%  |                                                                              |=============================================                         |  64%  |                                                                              |=============================================                         |  65%  |                                                                              |==============================================                        |  65%  |                                                                              |==============================================                        |  66%  |                                                                              |===============================================                       |  66%  |                                                                              |===============================================                       |  67%  |                                                                              |===============================================                       |  68%  |                                                                              |================================================                      |  68%  |                                                                              |================================================                      |  69%  |                                                                              |=================================================                     |  69%  |                                                                              |=================================================                     |  70%  |                                                                              |=================================================                     |  71%  |                                                                              |==================================================                    |  71%  |                                                                              |==================================================                    |  72%  |                                                                              |===================================================                   |  72%  |                                                                              |===================================================                   |  73%  |                                                                              |===================================================                   |  74%  |                                                                              |====================================================                  |  74%  |                                                                              |====================================================                  |  75%  |                                                                              |=====================================================                 |  75%  |                                                                              |=====================================================                 |  76%  |                                                                              |======================================================                |  76%  |                                                                              |======================================================                |  77%  |                                                                              |======================================================                |  78%  |                                                                              |=======================================================               |  78%  |                                                                              |=======================================================               |  79%  |                                                                              |========================================================              |  79%  |                                                                              |========================================================              |  80%  |                                                                              |========================================================              |  81%  |                                                                              |=========================================================             |  81%  |                                                                              |=========================================================             |  82%  |                                                                              |==========================================================            |  82%  |                                                                              |==========================================================            |  83%  |                                                                              |==========================================================            |  84%  |                                                                              |===========================================================           |  84%  |                                                                              |===========================================================           |  85%  |                                                                              |============================================================          |  85%  |                                                                              |============================================================          |  86%  |                                                                              |=============================================================         |  86%  |                                                                              |=============================================================         |  87%  |                                                                              |=============================================================         |  88%  |                                                                              |==============================================================        |  88%  |                                                                              |==============================================================        |  89%  |                                                                              |===============================================================       |  89%  |                                                                              |===============================================================       |  90%  |                                                                              |===============================================================       |  91%  |                                                                              |================================================================      |  91%  |                                                                              |================================================================      |  92%  |                                                                              |=================================================================     |  92%  |                                                                              |=================================================================     |  93%  |                                                                              |=================================================================     |  94%  |                                                                              |==================================================================    |  94%  |                                                                              |==================================================================    |  95%  |                                                                              |===================================================================   |  95%  |                                                                              |===================================================================   |  96%  |                                                                              |====================================================================  |  96%  |                                                                              |====================================================================  |  97%  |                                                                              |====================================================================  |  98%  |                                                                              |===================================================================== |  98%  |                                                                              |===================================================================== |  99%  |                                                                              |======================================================================|  99%  |                                                                              |======================================================================| 100%
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{summary}\NormalTok{(fit\_sartobit)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## ----MCMC spatial autoregressive Tobit model ----
+## Execution time  = 40.052 secs
+## 
+## N draws         =   1000, N omit (burn-in)=    200
+## N observations  =   3218, K covariates    =      4
+## # censored values =   2067, # observed values =   1151
+## Min rho         = -1.000, Max rho         =  1.000
+## --------------------------------------------------
+## 
+##              Estimate  Std. Dev   p-level t-value Pr(>|z|)    
+## (Intercept) -4.399260  0.737342  0.000000  -5.966 2.69e-09 ***
+## density_log  0.562608  0.026714  0.000000  21.061  < 2e-16 ***
+## private      0.039826  0.006874  0.000000   5.794 7.54e-09 ***
+## hincome_log  1.572387  0.138627  0.000000  11.343  < 2e-16 ***
+## sige         2.992428  0.173624  0.000000  17.235  < 2e-16 ***
+## rho          0.056389  0.004937  0.000000  11.421  < 2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{impacts}\NormalTok{(fit\_sartobit)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## --------Marginal Effects--------
+## 
+## (a) Direct effects
+##             lower_005 posterior_mean upper_095
+## density_log   0.21275        0.23102     0.248
+## private       0.01180        0.01635     0.021
+## hincome_log   0.55351        0.64566     0.744
+## 
+## (b) Indirect effects
+##             lower_005 posterior_mean upper_095
+## density_log  -0.29273       -0.27252    -0.251
+## private      -0.02504       -0.01929    -0.014
+## hincome_log  -0.88010       -0.76178    -0.651
+## 
+## (c) Total effects
+##             lower_005 posterior_mean upper_095
+## density_log -0.044493      -0.041495    -0.038
+## private     -0.003850      -0.002942    -0.002
+## hincome_log -0.136033      -0.116125    -0.097
+\end{verbatim}
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+Words go here.
 
@@
 \begin{CSLReferences}{0}{0}
-\leavevmode\hypertarget{ref-Feynman1963118}{}%
+\leavevmode\hypertarget{ref-Fraser2021evolving}{}%
 \CSLLeftMargin{1. }
-\CSLRightInline{Feynman RP, Vernon Jr. FL. The theory of a general
-quantum system interacting with a linear dissipative system. Annals of
-Physics. 1963;24: 118--173.
-doi:\href{https://doi.org/10.1016/0003-4916(63)90068-X}{10.1016/0003-4916(63)90068-X}}
+\CSLRightInline{Fraser N, Brierley L, Dey G, Polka JK, Pálfy M, Nanni F,
+et al. The evolving role of preprints in the dissemination of COVID-19
+research and their impact on the science communication landscape. PLOS
+Biology. 2021;19: e3000959.
+doi:\href{https://doi.org/10.1371/journal.pbio.3000959}{10.1371/journal.pbio.3000959}}
 
-\leavevmode\hypertarget{ref-Dirac1953888}{}%
+\leavevmode\hypertarget{ref-Kwon2021swamped}{}%
 \CSLLeftMargin{2. }
-\CSLRightInline{Dirac PAM. The lorentz transformation and absolute time.
-Physica. 1953;19: 888--896.
-doi:\href{https://doi.org/10.1016/S0031-8914(53)80099-6}{10.1016/S0031-8914(53)80099-6}}
+\CSLRightInline{Kwon D. How swamped preprint servers are blocking bad
+coronavirus research. Nature. 2020;581: 130--132. }
+
+\leavevmode\hypertarget{ref-Vlasschaert2020proliferation}{}%
+\CSLLeftMargin{3. }
+\CSLRightInline{Vlasschaert C, Topf JM, Hiremath S. Proliferation of
+papers and preprints during the coronavirus disease 2019 pandemic:
+Progress or problems with peer review? Advances in Chronic Kidney
+Disease. 2020;27: 418--426.
+doi:\href{https://doi.org/10.1053/j.ackd.2020.08.003}{10.1053/j.ackd.2020.08.003}}
+
+\leavevmode\hypertarget{ref-Anazco2021publication}{}%
+\CSLLeftMargin{4. }
+\CSLRightInline{Añazco D, Nicolalde B, Espinosa I, Camacho J, Mushtaq M,
+Gimenez J, et al. Publication rate and citation counts for preprints
+released during the COVID-19 pandemic: The good, the bad and the ugly.
+PeerJ. 2021;9: e10927.
+doi:\href{https://doi.org/10.7717/peerj.10927}{10.7717/peerj.10927}}
+
+\leavevmode\hypertarget{ref-Bramstedt2020carnage}{}%
+\CSLLeftMargin{5. }
+\CSLRightInline{Bramstedt KA. The carnage of substandard research during
+the COVID-19 pandemic: A call for quality. Journal of Medical Ethics.
+2020;46: 803--807.
+doi:\href{https://doi.org/10.1136/medethics-2020-106494}{10.1136/medethics-2020-106494}}
+
+\leavevmode\hypertarget{ref-Broggini2017reproducible}{}%
+\CSLLeftMargin{6. }
+\CSLRightInline{Broggini F, Dellinger J, Fomel S, Liu Y. Reproducible
+research: Geophysics papers of the future - introduction. Geophysics.
+2017;82.
+doi:\href{https://doi.org/10.1190/geo2017-0918-spseintro.1}{10.1190/geo2017-0918-spseintro.1}}
+
+\leavevmode\hypertarget{ref-Ince2012case}{}%
+\CSLLeftMargin{7. }
+\CSLRightInline{Ince DC, Hatton L, Graham-Cumming J. The case for open
+computer programs. Nature. 2012;482: 485--488.
+doi:\href{https://doi.org/10.1038/nature10836}{10.1038/nature10836}}
+
+\leavevmode\hypertarget{ref-Ioannidis2014increasing}{}%
+\CSLLeftMargin{8. }
+\CSLRightInline{Ioannidis JPA, Greenland S, Hlatky MA, Khoury MJ,
+Macleod MR, Moher D, et al. Increasing value and reducing waste in
+research design, conduct, and analysis. Lancet. 2014;383: 166--175.
+doi:\href{https://doi.org/10.1016/s0140-6736(13)62227-8}{10.1016/s0140-6736(13)62227-8}}
+
+\leavevmode\hypertarget{ref-Brunsdon2020opening}{}%
+\CSLLeftMargin{9. }
+\CSLRightInline{Brunsdon C, Comber A. Opening practice: Supporting
+reproducibility and critical spatial data science. Journal of
+Geographical Systems. 2020;
+doi:\href{https://doi.org/10.1007/s10109-020-00334-2}{10.1007/s10109-020-00334-2}}
+
+\leavevmode\hypertarget{ref-Konkol2019examination}{}%
+\CSLLeftMargin{10. }
+\CSLRightInline{Konkol M, Kray C. In-depth examination of spatiotemporal
+figures in open reproducible research. Cartography and Geographic
+Information Science. 2019;46: 412--427.
+doi:\href{https://doi.org/10.1080/15230406.2018.1512421}{10.1080/15230406.2018.1512421}}
+
+\leavevmode\hypertarget{ref-Konkol2019computational}{}%
+\CSLLeftMargin{11. }
+\CSLRightInline{Konkol M, Kray C, Pfeiffer M. Computational
+reproducibility in geoscientific papers: Insights from a series of
+studies with geoscientists and a reproduction study. International
+Journal of Geographical Information Science. 2019;33: 408--429.
+doi:\href{https://doi.org/10.1080/13658816.2018.1508687}{10.1080/13658816.2018.1508687}}
+
+\leavevmode\hypertarget{ref-Iqbal2016reproducible}{}%
+\CSLLeftMargin{12. }
+\CSLRightInline{Iqbal SA, Wallach JD, Khoury MJ, Schully SD, Ioannidis
+JPA. Reproducible research practices and transparency across the
+biomedical literature. Plos Biology. 2016;14.
+doi:\href{https://doi.org/10.1371/journal.pbio.1002333}{10.1371/journal.pbio.1002333}}
+
+\leavevmode\hypertarget{ref-Stodden2018empirical}{}%
+\CSLLeftMargin{13. }
+\CSLRightInline{Stodden V, Seiler J, Ma ZK. An empirical analysis of
+journal policy effectiveness for computational reproducibility.
+Proceedings of the National Academy of Sciences of the United States of
+America. 2018;115: 2584--2589.
+doi:\href{https://doi.org/10.1073/pnas.1708290115}{10.1073/pnas.1708290115}}
+
+\leavevmode\hypertarget{ref-Sumner2020reproducibility}{}%
+\CSLLeftMargin{14. }
+\CSLRightInline{Sumner J, Haynes L, Nathan S, Hudson-Vitale C, McIntosh
+LD. Reproducibility and reporting practices in COVID-19 preprint
+manuscripts. medRxiv. 2020; 2020.03.24.20042796.
+doi:\href{https://doi.org/10.1101/2020.03.24.20042796}{10.1101/2020.03.24.20042796}}
+
+\leavevmode\hypertarget{ref-Gustot2020quality}{}%
+\CSLLeftMargin{15. }
+\CSLRightInline{Gustot T. Quality and reproducibility during the
+COVID-19 pandemic. JHEP Rep. 2020;2: 100141.
+doi:\href{https://doi.org/10.1016/j.jhepr.2020.100141}{10.1016/j.jhepr.2020.100141}}
+
+\leavevmode\hypertarget{ref-Herndon2014high}{}%
+\CSLLeftMargin{16. }
+\CSLRightInline{Herndon T, Ash M, Pollin R. Does high public debt
+consistently stifle economic growth? A critique of reinhart and rogoff.
+Cambridge Journal of Economics. 2014;38: 257--279.
+doi:\href{https://doi.org/10.1093/cje/bet075}{10.1093/cje/bet075}}
+
+\leavevmode\hypertarget{ref-Basu2017ten}{}%
+\CSLLeftMargin{17. }
+\CSLRightInline{Basu S, Carney MA, Kenworthy NJ. Ten years after the
+financial crisis: The long reach of austerity and its global impacts on
+health. Social Science \& Medicine. 2017;187: 203--207.
+doi:\href{https://doi.org/10.1016/j.socscimed.2017.06.026}{10.1016/j.socscimed.2017.06.026}}
+
+\leavevmode\hypertarget{ref-Florida2020how}{}%
+\CSLLeftMargin{18. }
+\CSLRightInline{Florida R, Glaeser E, Sharif M, Bedi K, Campanella T,
+Chee C, et al. How life in our cities will look after the coronavirus
+pandemic. Foreign Policy. 2020;1. Available:
+\url{https://foreignpolicy.com/2020/05/01/future-of-cities-urban-life-after-coronavirus-pandemic/}}
+
+\leavevmode\hypertarget{ref-Sy2021population}{}%
+\CSLLeftMargin{19. }
+\CSLRightInline{Sy KTL, White LF, Nichols BE. Population density and
+basic reproductive number of COVID-19 across united states counties.
+PLOS ONE. 2021;16: e0249271.
+doi:\href{https://doi.org/10.1371/journal.pone.0249271}{10.1371/journal.pone.0249271}}
+
+\leavevmode\hypertarget{ref-Moore1970urban}{}%
+\CSLLeftMargin{20. }
+\CSLRightInline{Moore EG, Brown LA. Urban acquaintance fields: An
+evaluation of a spatial model. Environment and Planning. 1970;2:
+443--454. Available:
+\url{http://www.envplan.com/abstract.cgi?id=a020443}}
+
+\leavevmode\hypertarget{ref-Moore1970some}{}%
+\CSLLeftMargin{21. }
+\CSLRightInline{Moore EG. Some spatial properties of urban contact
+fields. Geographical Analysis. 1970;2: 376--386. }
+
+\leavevmode\hypertarget{ref-Farber2011running}{}%
+\CSLLeftMargin{22. }
+\CSLRightInline{Farber S, Páez A. Running to stay in place: The time-use
+implications of automobile oriented land-use and travel. Journal of
+Transport Geography. 2011;19: 782--793.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2010.09.008}{10.1016/j.jtrangeo.2010.09.008}}
+
+\leavevmode\hypertarget{ref-Rocklov2020high}{}%
+\CSLLeftMargin{23. }
+\CSLRightInline{Rocklöv J, Sjödin H. High population densities catalyse
+the spread of COVID-19. Journal of Travel Medicine. 2020;27.
+doi:\href{https://doi.org/10.1093/jtm/taaa038}{10.1093/jtm/taaa038}}
+
+\leavevmode\hypertarget{ref-Li2018effect}{}%
+\CSLLeftMargin{24. }
+\CSLRightInline{Li R, Richmond P, Roehner BM. Effect of population
+density on epidemics. Physica A: Statistical Mechanics and its
+Applications. 2018;510: 713--724.
+doi:\href{https://doi.org/10.1016/j.physa.2018.07.025}{10.1016/j.physa.2018.07.025}}
+
+\leavevmode\hypertarget{ref-Cruz2020exploring}{}%
+\CSLLeftMargin{25. }
+\CSLRightInline{Cruz CJP, Ganly R, Li Z, Gietel-Basten S. Exploring the
+young demographic profile of COVID-19 cases in hong kong: Evidence from
+migration and travel history data. PLOS ONE. 2020;15: e0235306.
+doi:\href{https://doi.org/10.1371/journal.pone.0235306}{10.1371/journal.pone.0235306}}
+
+\leavevmode\hypertarget{ref-Micallef2020first}{}%
+\CSLLeftMargin{26. }
+\CSLRightInline{Micallef S, Piscopo TV, Casha R, Borg D, Vella C, Zammit
+M-A, et al. The first wave of COVID-19 in malta; a national
+cross-sectional study. PLOS ONE. 2020;15: e0239389.
+doi:\href{https://doi.org/10.1371/journal.pone.0239389}{10.1371/journal.pone.0239389}}
+
+\leavevmode\hypertarget{ref-Gomez2021infekta}{}%
+\CSLLeftMargin{27. }
+\CSLRightInline{Gomez J, Prieto J, Leon E, Rodríguez A. INFEKTA---an
+agent-based model for transmission of infectious diseases: The COVID-19
+case in bogotá, colombia. PLOS ONE. 2021;16: e0245787.
+doi:\href{https://doi.org/10.1371/journal.pone.0245787}{10.1371/journal.pone.0245787}}
+
+\leavevmode\hypertarget{ref-Noland1995perceived}{}%
+\CSLLeftMargin{28. }
+\CSLRightInline{Noland RB. PERCEIVED RISK AND MODAL CHOICE - RISK
+COMPENSATION IN TRANSPORTATION SYSTEM. Accident Analysis and Prevention.
+1995;27: 503--521.
+doi:\href{https://doi.org/10.1016/0001-4575(94)00087-3}{10.1016/0001-4575(94)00087-3}}
+
+\leavevmode\hypertarget{ref-Richens2000condoms}{}%
+\CSLLeftMargin{29. }
+\CSLRightInline{Richens J, Imrie J, Copas A. Condoms and seat belts: The
+parallels and the lessons. Lancet. 2000;355: 400--403.
+doi:\href{https://doi.org/10.1016/s0140-6736(99)09109-6}{10.1016/s0140-6736(99)09109-6}}
+
+\leavevmode\hypertarget{ref-Phillips2011risk}{}%
+\CSLLeftMargin{30. }
+\CSLRightInline{Phillips RO, Fyhri A, Sagberg F. Risk compensation and
+bicycle helmets. Risk Analysis. 2011;31: 1187--1195.
+doi:\href{https://doi.org/10.1111/j.1539-6924.2011.01589.x}{10.1111/j.1539-6924.2011.01589.x}}
+
+\leavevmode\hypertarget{ref-Paez2020spatio}{}%
+\CSLLeftMargin{31. }
+\CSLRightInline{Paez A, Lopez FA, Menezes T, Cavalcanti R, Pitta MG da
+R. A spatio-temporal analysis of the environmental correlates of
+COVID-19 incidence in spain. Geographical Analysis. 2020;n/a.
+doi:\href{https://doi.org/10.1111/gean.12241}{10.1111/gean.12241}}
+
+\leavevmode\hypertarget{ref-Hamidi2020density}{}%
+\CSLLeftMargin{32. }
+\CSLRightInline{Hamidi S, Sabouri S, Ewing R. Does density aggravate the
+COVID-19 pandemic? Journal of the American Planning Association.
+2020;86: 495--509.
+doi:\href{https://doi.org/10.1080/01944363.2020.1777891}{10.1080/01944363.2020.1777891}}
+
+\leavevmode\hypertarget{ref-Jamal2020Changes}{}%
+\CSLLeftMargin{33. }
+\CSLRightInline{Jamal S, Paez A. Changes in trip-making frequency by
+mode during COVID-19. Findings. Network Design Lab; 2020;
+doi:\href{https://doi.org/10.32866/001c.17977}{10.32866/001c.17977}}
+
+\leavevmode\hypertarget{ref-Harris2021Changes}{}%
+\CSLLeftMargin{34. }
+\CSLRightInline{Harris MA, Branion-Calles M. Changes in commute mode
+attributed to COVID-19 risk in canadian national survey data. Findings.
+Network Design Lab; 2021;
+doi:\href{https://doi.org/10.32866/001c.19088}{10.32866/001c.19088}}
+
+\leavevmode\hypertarget{ref-Molloy2020Tracing}{}%
+\CSLLeftMargin{35. }
+\CSLRightInline{Molloy J, Tchervenkov C, Hintermann B, Axhausen KW.
+Tracing the sars-CoV-2 impact: The first month in switzerland. Findings.
+Network Design Lab; 2020;
+doi:\href{https://doi.org/10.32866/001c.12903}{10.32866/001c.12903}}
+
+\leavevmode\hypertarget{ref-Feyman2020effectiveness}{}%
+\CSLLeftMargin{36. }
+\CSLRightInline{Feyman Y, Bor J, Raifman J, Griffith KN. Effectiveness
+of COVID-19 shelter-in-place orders varied by state. PLOS ONE. 2020;15:
+e0245008.
+doi:\href{https://doi.org/10.1371/journal.pone.0245008}{10.1371/journal.pone.0245008}}
+
+\leavevmode\hypertarget{ref-Praharaj2020Using}{}%
+\CSLLeftMargin{37. }
+\CSLRightInline{Praharaj S, King D, Pettit C, Wentz E. Using aggregated
+mobility data to measure the effect of COVID-19 policies on mobility
+changes in sydney, london, phoenix, and pune. Findings. Network Design
+Lab; 2020;
+doi:\href{https://doi.org/10.32866/001c.17590}{10.32866/001c.17590}}
+
+\leavevmode\hypertarget{ref-Paez2020using}{}%
+\CSLLeftMargin{38. }
+\CSLRightInline{Paez A. Using google community mobility reports to
+investigate the incidence of COVID-19 in the united states. Findings.
+2020; doi:\url{https://doi.org/10.32866/001c.12976}}
+
+\leavevmode\hypertarget{ref-Noland2021mobility}{}%
+\CSLLeftMargin{39. }
+\CSLRightInline{Noland RB. Mobility and the effective reproduction rate
+of COVID-19. Journal of Transport \& Health. 2021;20: 101016.
+doi:\url{https://doi.org/10.1016/j.jth.2021.101016}}
+
+\leavevmode\hypertarget{ref-Bhadra2021impact}{}%
+\CSLLeftMargin{40. }
+\CSLRightInline{Bhadra A, Mukherjee A, Sarkar K. Impact of population
+density on covid-19 infected and mortality rate in india. Modeling Earth
+Systems and Environment. 2021;7: 623--629.
+doi:\href{https://doi.org/10.1007/s40808-020-00984-7}{10.1007/s40808-020-00984-7}}
+
+\leavevmode\hypertarget{ref-Kadi2020population}{}%
+\CSLLeftMargin{41. }
+\CSLRightInline{Kadi N, Khelfaoui M. Population density, a factor in the
+spread of COVID-19 in algeria: Statistic study. Bulletin of the National
+Research Centre. 2020;44.
+doi:\href{https://doi.org/10.1186/s42269-020-00393-x}{10.1186/s42269-020-00393-x}}
+
+\leavevmode\hypertarget{ref-Pequeno2020air}{}%
+\CSLLeftMargin{42. }
+\CSLRightInline{Pequeno P, Mendel B, Rosa C, Bosholn M, Souza JL,
+Baccaro F, et al. Air transportation, population density and temperature
+predict the spread of COVID-19 in brazil. PeerJ. 2020;8: e9322.
+doi:\href{https://doi.org/10.7717/peerj.9322}{10.7717/peerj.9322}}
+
+\leavevmode\hypertarget{ref-Fielding2020social}{}%
+\CSLLeftMargin{43. }
+\CSLRightInline{Fielding-Miller RK, Sundaram ME, Brouwer K. Social
+determinants of COVID-19 mortality at the county level. PLOS ONE.
+2020;15: e0240151.
+doi:\href{https://doi.org/10.1371/journal.pone.0240151}{10.1371/journal.pone.0240151}}
+
+\leavevmode\hypertarget{ref-Roy2020factors}{}%
+\CSLLeftMargin{44. }
+\CSLRightInline{Roy S, Ghosh P. Factors affecting COVID-19 infected and
+death rates inform lockdown-related policymaking. PLOS ONE. 2020;15:
+e0241165.
+doi:\href{https://doi.org/10.1371/journal.pone.0241165}{10.1371/journal.pone.0241165}}
+
+\leavevmode\hypertarget{ref-Wong2020spreading}{}%
+\CSLLeftMargin{45. }
+\CSLRightInline{Wong DWS, Li Y. Spreading of COVID-19: Density matters.
+PLOS ONE. 2020;15: e0242398.
+doi:\href{https://doi.org/10.1371/journal.pone.0242398}{10.1371/journal.pone.0242398}}
+
+\leavevmode\hypertarget{ref-Sun2020impacts}{}%
+\CSLLeftMargin{46. }
+\CSLRightInline{Sun Z, Zhang H, Yang Y, Wan H, Wang Y. Impacts of
+geographic factors and population density on the COVID-19 spreading
+under the lockdown policies of china. Science of The Total Environment.
+2020;746: 141347.
+doi:\href{https://doi.org/10.1016/j.scitotenv.2020.141347}{10.1016/j.scitotenv.2020.141347}}
+
+\leavevmode\hypertarget{ref-Skorka2020macroecology}{}%
+\CSLLeftMargin{47. }
+\CSLRightInline{Skórka P, Grzywacz B, Moroń D, Lenda M. The macroecology
+of the COVID-19 pandemic in the anthropocene. PLOS ONE. 2020;15:
+e0236856.
+doi:\href{https://doi.org/10.1371/journal.pone.0236856}{10.1371/journal.pone.0236856}}
+
+\leavevmode\hypertarget{ref-White2020state}{}%
+\CSLLeftMargin{48. }
+\CSLRightInline{White ER, Hébert-Dufresne L. State-level variation of
+initial COVID-19 dynamics in the united states. PLOS ONE. 2020;15:
+e0240648.
+doi:\href{https://doi.org/10.1371/journal.pone.0240648}{10.1371/journal.pone.0240648}}
+
+\leavevmode\hypertarget{ref-Khavarian2021high}{}%
+\CSLLeftMargin{49. }
+\CSLRightInline{Khavarian-Garmsir AR, Sharifi A, Moradpour N. Are
+high-density districts more vulnerable to the COVID-19 pandemic?
+Sustainable Cities and Society. 2021;70: 102911.
+doi:\href{https://doi.org/10.1016/j.scs.2021.102911}{10.1016/j.scs.2021.102911}}
+
+\leavevmode\hypertarget{ref-Hamidi2020longitudinal}{}%
+\CSLLeftMargin{50. }
+\CSLRightInline{Hamidi S, Ewing R, Sabouri S. Longitudinal analyses of
+the relationship between development density and the COVID-19 morbidity
+and mortality rates: Early evidence from 1,165 metropolitan counties in
+the united states. Health \& Place. 2020;64: 102378.
+doi:\href{https://doi.org/10.1016/j.healthplace.2020.102378}{10.1016/j.healthplace.2020.102378}}
+
+\leavevmode\hypertarget{ref-Lee2020human}{}%
+\CSLLeftMargin{51. }
+\CSLRightInline{Lee M, Zhao J, Sun Q, Pan Y, Zhou W, Xiong C, et al.
+Human mobility trends during the early stage of the COVID-19 pandemic in
+the united states. PLOS ONE. 2020;15: e0241468.
+doi:\href{https://doi.org/10.1371/journal.pone.0241468}{10.1371/journal.pone.0241468}}
+
+\leavevmode\hypertarget{ref-Amadu2021assessing}{}%
+\CSLLeftMargin{52. }
+\CSLRightInline{Amadu I, Ahinkorah BO, Afitiri A-R, Seidu A-A, Ameyaw
+EK, Hagan JE, et al. Assessing sub-regional-specific strengths of
+healthcare systems associated with COVID-19 prevalence, deaths and
+recoveries in africa. PLOS ONE. 2021;16: e0247274.
+doi:\href{https://doi.org/10.1371/journal.pone.0247274}{10.1371/journal.pone.0247274}}
+
+\leavevmode\hypertarget{ref-Feng2020spread}{}%
+\CSLLeftMargin{53. }
+\CSLRightInline{Feng Y, Li Q, Tong X, Wang R, Zhai S, Gao C, et al.
+Spatiotemporal spread pattern of the COVID-19 cases in china. PLOS ONE.
+2020;15: e0244351.
+doi:\href{https://doi.org/10.1371/journal.pone.0244351}{10.1371/journal.pone.0244351}}
+
+\leavevmode\hypertarget{ref-Inbaraj2021seroprevalence}{}%
+\CSLLeftMargin{54. }
+\CSLRightInline{Inbaraj LR, George CE, Chandrasingh S. Seroprevalence of
+COVID-19 infection in a rural district of south india: A
+population-based seroepidemiological study. PLOS ONE. 2021;16: e0249247.
+doi:\href{https://doi.org/10.1371/journal.pone.0249247}{10.1371/journal.pone.0249247}}
+
+\leavevmode\hypertarget{ref-Souris2020covid}{}%
+\CSLLeftMargin{55. }
+\CSLRightInline{Souris M, Gonzalez J-P. COVID-19: Spatial analysis of
+hospital case-fatality rate in france. PLOS ONE. 2020;15: e0243606.
+doi:\href{https://doi.org/10.1371/journal.pone.0243606}{10.1371/journal.pone.0243606}}
+
+\leavevmode\hypertarget{ref-Ahmad2020association}{}%
+\CSLLeftMargin{56. }
+\CSLRightInline{Ahmad K, Erqou S, Shah N, Nazir U, Morrison AR,
+Choudhary G, et al. Association of poor housing conditions with COVID-19
+incidence and mortality across US counties. PLOS ONE. 2020;15: e0241327.
+doi:\href{https://doi.org/10.1371/journal.pone.0241327}{10.1371/journal.pone.0241327}}
+
+\leavevmode\hypertarget{ref-Noury2021how}{}%
+\CSLLeftMargin{57. }
+\CSLRightInline{Noury A, François A, Gergaud O, Garel A. How does
+COVID-19 affect electoral participation? Evidence from the french
+municipal elections. PLOS ONE. 2021;16: e0247026.
+doi:\href{https://doi.org/10.1371/journal.pone.0247026}{10.1371/journal.pone.0247026}}
+
+\leavevmode\hypertarget{ref-Wang2021transmission}{}%
+\CSLLeftMargin{58. }
+\CSLRightInline{Wang F, Tan Z, Yu Z, Yao S, Guo C. Transmission and
+control pressure analysis of the COVID-19 epidemic situation using
+multisource spatio-temporal big data. PLOS ONE. 2021;16: e0249145.
+doi:\href{https://doi.org/10.1371/journal.pone.0249145}{10.1371/journal.pone.0249145}}
+
+\leavevmode\hypertarget{ref-Brandtner2021creatures}{}%
+\CSLLeftMargin{59. }
+\CSLRightInline{Brandtner C, Bettencourt LMA, Berman MG, Stier AJ.
+Creatures of the state? Metropolitan counties compensated for state
+inaction in initial u.s. Response to COVID-19 pandemic. PLOS ONE.
+2021;16: e0246249.
+doi:\href{https://doi.org/10.1371/journal.pone.0246249}{10.1371/journal.pone.0246249}}
+
+\leavevmode\hypertarget{ref-Stephens2021impact}{}%
+\CSLLeftMargin{60. }
+\CSLRightInline{Stephens KE, Chernyavskiy P, Bruns DR. Impact of
+altitude on COVID-19 infection and death in the united states: A
+modeling and observational study. PLOS ONE. 2021;16: e0245055.
+doi:\href{https://doi.org/10.1371/journal.pone.0245055}{10.1371/journal.pone.0245055}}
 

### 2021-05-19 README.Rmd
First draft of introduction and background

--- 2021-05-18 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
@@
 ---
-title: "Population density and basic reproductive number of COVID-19 across United States counties revisited"
-author:
-  - name: Antonio Paez
-    email: paezha@mcmaster.ca
-    affiliation: McMaster University
-    corresponding: paezha@mcmaster.ca
-address:
-  - code: McMaster University
-    address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
-abstract: |
-  Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
-
-author_summary: |
-  Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
-
-bibliography: bibliography.bib
-output: rticles::plos_article
-csl: plos.csl
+output:
+  github_document:
+    pandoc_args: --webtex
+always_allow_html: true
+bibliography: R0-Density-Reanalysis/bibliography.bib
 ---
 
-```{r install-data-package, include = FALSE}
-# Run only once if needed to install data package `sobiEquity`
-if (!require("r0density", character.only = TRUE)) {
-      devtools::install_github("https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed", subdir = "r0density")
-  }
+<!-- README.md is generated from README.Rmd. Please edit that file -->
+
+```{r, include = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>"
+)
 ```
 
-# Introduction
+# The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+<!-- badges: start -->
+<!-- badges: end -->
 
-A list
+Antonio Páez (McMaster University)
 
-- Item 1
-- Item 2
+Paper in preparation for submision to *PLoS ONE*
 
-Here are two sample references: @Feynman1963118 [@Dirac1953888].
+## Abstract
 
-# References {#references .unnumbered}
+The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research.
+Given the high stakes of the situation, it is essential that scientific activites, on which good policy depends, are as transparent and reproducible as possible.
+Reproducibility is key for the efficient operation of the self-correction mechanisms of science, which work to weed out errors and refine our understanding of social and physical phenomena.
+In this paper, the importance of reproducibility is illustrated for the case of the association between population density and the the spread of SARS-CoV-2.
+Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequantial changes in the conclusions, as the empirical example in this paper shows.
+
+## Keywords
+
+- COVID-19  
+- Reproducible research  
+- Population density  
+- Basic reproductive number
+
+## Introduction
+
+The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the global pandemic that followed in its wake, led to an explosive growth of research.
+According to Fraser et al.
+[-@Fraser2021evolving], over 125,000 COVID-19-related papers were released in the first ten months from the first confirmed case of the disease.
+Of these, more than 30,000 were shared in pre-print servers, the use of which also exploded in the past year [@Kwon2021swamped; @Vlasschaert2020proliferation; @Anazco2021publication].
+
+Given the heavy human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has even called this deluge of publications a "carnage of substandard research" [@Bramstedt2020carnage].
+Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [@Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty but somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
+As reported in the literature, only a woefully small proportion of published research was actually reproducible before the pandemic [@Iqbal2016reproducible; @Stodden2018empirical], a situation that does not appear to have changed substantially since [@Sumner2020reproducibility; @Gustot2020quality].
+
+The push for open data and software, along with more strenuous efforts towards open, reproducible research, is simply a continuation of long-standing scientific practices of independent verification.
+Despite the (at times disproportionate) attention that high profile scandals in science tend to elicit in the media, science as a collective endeavor is remarkable for being a self-correcting enterprise, one with built-in mechanisms and incentives to weed out erroneous ideas.
+Over the long term, facts tend to prevail in science.
+At stake is the shorter-term impacts that research may have in other spheres of economic and social life.
+The case of economists Reinhart and Rogoff comes to mind: by the time the inaccuracies in their research were uncovered [see @Herndon2014high], their claims about debt and economic growth had already been seized by policy-makers on both sides of the Atlantic to justify austerity policies in the aftermath of the Great Recession of 2007-2009^[Nobel Prize in Economics Paul Krugman noted that "Reinhart–Rogoff may have had more immediate influence on public debate than any previous paper in the history of economics" \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}].
+As later research has demonstrated, those policies cast a long shadow, and their sequels continued to be felt for years [@Basu2017ten].
+
+In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how].
+The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, prompted a flurry of research into the associations between density and the spread of the pandemic.
+Some important questions hang on the results of these research efforts.
+For example, are lower density regions safer from the pandemic?
+Are de-densification policies warranted, at least in the short term?
+And in the longer term, will the risks of life in high density regions presage a flight from cities?
+Over the past year, numerous papers have sought to throw light into the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed.
+Further, to complicate matters, precious few of these studies appear to be sufficiently open to support independent verification.
+
+The objective of this paper is to illustrate the importance of reproducibility in research, particularly in the context of the flood of COVID-19 papers.
+To this end, a recent study by Sy et al.
+[@Sy2021population] is chosen as an example of reproducible research.
+The objective is not to malign the analysis of these researchers, but rather to demonstrate the value of openness to allow for independent verification and further analysis.
+Open data and open code mean that an independent researcher can, with only modest efforts, not only verify the findings reported, but also examine the same data from a perspective which may not have been available to the original researchers due to differences in disciplinary perspectives, methodological traditions, and/or training, among other possible factors.
+The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in research.
+This paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
+The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, is available in a public repository^[\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}].
+
+## References

### 2021-05-19 r0density/R/r0density.R

@@ -1,6 +1,6 @@
 #' r0density
 #'
-#' r0density is a companion package for the paper "Population Density and Basic Reproductive Number of COVID-19 Revisited".
+#' r0density is a companion package for the paper "The importance of reproducibility in COVID-19 research: the case of population density and the basic reproductive number of COVID-19".
 #'
 #' The package includes the data used in the analysis.
 #'

### 2021-05-19 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Added reference to butress the point of risk perception

@@
 
-A plausible behavioral adaptiation during a pandemic, especially one broadcast as widely and intensely as COVID-19, is risk compensation.
+A plausible behavioral adaptation during a pandemic, especially one broadcast as widely and intensely as COVID-19, is risk compensation.
 Risk compensation is a process whereby people adjust their behavior in response to their _perception_ of risk [@Noland1995perceived; @Richens2000condoms; @Phillips2011risk].
-For example, it is possible that people who listened to the message of leaders saying that they were safe because of low density may not have taken adequate precautions.
+In the case of COVID-19, Chauhan et al.
+[@Chauhan2021covid] have found that perception of risks in the US varies between rural, suburban, and urban residents, with rural residents in general displaying less concerns about the virus.
+It is possible that people who listened to the message of leaders saying that they were safe because of low density may not have taken adequate precautions against the virus.
 People in dense places who could more directly observe the impact of the pandemic may have become overly cautious.

### 2021-05-19 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Added reference to butress the point of risk perception

@@
 
-A plausible behavioral adaptiation during a pandemic, especially one
+A plausible behavioral adaptation during a pandemic, especially one
 broadcast as widely and intensely as COVID-19, is risk compensation.
 Risk compensation is a process whereby people adjust their behavior in
-response to their \emph{perception} of risk {[}28--30{]}. For example,
-it is possible that people who listened to the message of leaders saying
-that they were safe because of low density may not have taken adequate
-precautions. People in dense places who could more directly observe the
-impact of the pandemic may have become overly cautious. Both Paez et al.
-{[}31{]} and Hamidi et al. {[}32{]} posit this mechanism (i.e., greater
-compliance with social distancing in denser regions) to explain the
-results of their analyses. The evidence available does indeed show that
-there were important changes in behavior during the pandemic, at least
-with respect to mobility {[}33--35{]}; furthermore, shelter in place
-orders may have had greater buy-in from the public in higher density
-regions {[}36{]}, and the behavior may have persisted beyond the
-duration of official social-distancing policies {[}37{]}. In addition,
-there is evidence that changes in mobility correlated with the
-trajectory of the pandemic {[}38,39{]}. Given the potential for
-behavioral adaptation, the question of density becomes more nuanced: it
-is not just a matter of proximity, but also of human behavior, which is
-better studied using population-level data and models.
+response to their \emph{perception} of risk {[}28--30{]}. In the case of
+COVID-19, Chauhan et al. {[}31{]} have found that perception of risks in
+the US varies between rural, suburban, and urban residents, with rural
+residents in general displaying less concerns about the virus. It is
+possible that people who listened to the message of leaders saying that
+they were safe because of low density may not have taken adequate
+precautions against the virus. People in dense places who could more
+directly observe the impact of the pandemic may have become overly
+cautious. Both Paez et al. {[}32{]} and Hamidi et al. {[}33{]} posit
+this mechanism (i.e., greater compliance with social distancing in
+denser regions) to explain the results of their analyses. The evidence
+available does indeed show that there were important changes in behavior
+during the pandemic, at least with respect to mobility {[}34--36{]};
+furthermore, shelter in place orders may have had greater buy-in from
+the public in higher density regions {[}37{]}, and the behavior may have
+persisted beyond the duration of official social-distancing policies
+{[}38{]}. In addition, there is evidence that changes in mobility
+correlated with the trajectory of the pandemic {[}39,40{]}. Given the
+potential for behavioral adaptation, the question of density becomes
+more nuanced: it is not just a matter of proximity, but also of human
+behavior, which is better studied using population-level data and
+models.
 
@@
 between population density and various COVID-19-related outcomes. Bhadra
-{[}40{]}, for example, reported a moderate positive correlation between
+{[}41{]}, for example, reported a moderate positive correlation between
 the spread of COVID-19 and population density at the district level in
 India, however their analysis was bivariate and did not control for
-other variables, such as income. Similarly, Kadi and Khelfaoui {[}41{]}
+other variables, such as income. Similarly, Kadi and Khelfaoui {[}42{]}
 found a positive and significant correlation between number of cases and
@@
 simple analyses is whether density is not a proxy for other factors.
-Other studies have included controls, such as Pequeno et al. {[}42{]}, a
+Other studies have included controls, such as Pequeno et al. {[}43{]}, a
 team that reported a positive association between density and cumulative
@@
 controlling for covariates, including income, transport connectivity,
-and economic status. In a similar vein, Fielding-Miller et al. {[}43{]}
+and economic status. In a similar vein, Fielding-Miller et al. {[}44{]}
 reported a positive relationship between the absolute number of COVID-19
 deaths and population density (rate) in rural counties in the US. Roy
-and Ghosh {[}44{]} used a battery of machine learning techniques to find
+and Ghosh {[}45{]} used a battery of machine learning techniques to find
 discriminatory factors, and a positive and significant association
 between COVID-19 infection and death rates in US states. Wong and Li
-{[}45{]} also found a positive and significant association between
+{[}46{]} also found a positive and significant association between
 population density and number of confirmed COVID-19 cases in US
@@
 negative associations between population density and COVID-19 outcomes.
-This includes the research of Sun et al. {[}46{]} who did not find
+This includes the research of Sun et al. {[}47{]} who did not find
 evidence of significant correlation between population density and
 confirmed number of cases per day \emph{in conditions of lockdown} in
-China. This finding echoes the results of Paez et al. {[}31{]}, who in
+China. This finding echoes the results of Paez et al. {[}32{]}, who in
 their study of provinces in Spain reported non-significant associations
@@
 first wave of COVID-19, and negative significant associations in the
-later part of the first lockdown. Similarly, {[}47{]} found zero or
+later part of the first lockdown. Similarly, {[}48{]} found zero or
 negative associations between population density and infection
-numbers/deaths by country. Fielding-Miller et al. {[}43{]} found a
+numbers/deaths by country. Fielding-Miller et al. {[}44{]} found a
 negative relationship between COVID-19 deaths and population density
 urban counties in the US. In their investigation of doubling time, White
-and Hébert-Dufresne {[}48{]} identified a negative and significant
+and Hébert-Dufresne {[}49{]} identified a negative and significant
 correlation between population density and doubling time in US states.
-Likewise, {[}49{]} fond a small negative (and significant) association
+Likewise, {[}50{]} fond a small negative (and significant) association
 between population density and COVID-19 morbidity in districts in
 Tehran. And two of the most complete studies in the US by Hamidi et al.
-{[}32,50{]} used an extensive set of controls to find negative and
+{[}33,51{]} used an extensive set of controls to find negative and
 significant correlations between density at the level of counties in the
@@
 some cases, there might be legitimate reasons to withhold the data,
-perhaps due to confidentiality and privacy reasons {[}e.g., 51{]}. But
+perhaps due to confidentiality and privacy reasons {[}e.g., 52{]}. But
 in many other cases the data are publicly available, as has been
@@
 provenance of the data is documented, but the data themselves are not
-shared {[}25,32,40,43,50,e.g., 52,53--55{]}. As any researcher can
+shared {[}25,33,41,44,51,e.g., 53,54--56{]}. As any researcher can
 attest, whether a graduate student or a seasoned scientist, collecting,
@@
 dedicated (or stubborn) researcher. This is true even if part of the
-data are shared {[}e.g., 45{]}. In other cases, data are shared, but the
+data are shared {[}e.g., 46{]}. In other cases, data are shared, but the
 processes to document the preparation of the data are not fully
-documented {[}47,e.g., 56{]}. These processes matter, as shown by the
+documented {[}48,e.g., 57{]}. These processes matter, as shown by the
 errors in the spreadsheets of Reinhart and Rogoff {[}16{]}, and the data
@@
 Another situation is when papers share well-documented data, but fail to
-provide the code used in the analysis {[}42,e.g., 57,58{]}. Making code
-available only ``on demand'' {[}e.g., 59{]} is an unnecessary barrier
+provide the code used in the analysis {[}43,e.g., 58,59{]}. Making code
+available only ``on demand'' {[}e.g., 60{]} is an unnecessary barrier
 when most journals offer the facility to share supplemental materials
@@
 sharing well-documented processes and data, as well as the code used in
-any analyses reported {[}e.g., 31,36,48,60{]}.
+any analyses reported {[}e.g., 32,37,49,61{]}.
 
@@
 
+\leavevmode\hypertarget{ref-Chauhan2021covid}{}%
+\CSLLeftMargin{31. }
+\CSLRightInline{Chauhan RS, Capasso da Silva D, Salon D, Shamshiripour
+A, Rahimi E, Sutradhar U, et al. COVID-19 related attitudes and risk
+perceptions across urban, rural, and suburban areas in the united
+states. Findings. Network Design Lab; 2021;
+doi:\href{https://doi.org/10.32866/001c.23714}{10.32866/001c.23714}}
+
 \leavevmode\hypertarget{ref-Paez2020spatio}{}%
-\CSLLeftMargin{31. }
+\CSLLeftMargin{32. }
 \CSLRightInline{Paez A, Lopez FA, Menezes T, Cavalcanti R, Pitta MG da
@@
 \leavevmode\hypertarget{ref-Hamidi2020density}{}%
-\CSLLeftMargin{32. }
+\CSLLeftMargin{33. }
 \CSLRightInline{Hamidi S, Sabouri S, Ewing R. Does density aggravate the
@@
 \leavevmode\hypertarget{ref-Jamal2020Changes}{}%
-\CSLLeftMargin{33. }
+\CSLLeftMargin{34. }
 \CSLRightInline{Jamal S, Paez A. Changes in trip-making frequency by
@@
 \leavevmode\hypertarget{ref-Harris2021Changes}{}%
-\CSLLeftMargin{34. }
+\CSLLeftMargin{35. }
 \CSLRightInline{Harris MA, Branion-Calles M. Changes in commute mode
@@
 \leavevmode\hypertarget{ref-Molloy2020Tracing}{}%
-\CSLLeftMargin{35. }
+\CSLLeftMargin{36. }
 \CSLRightInline{Molloy J, Tchervenkov C, Hintermann B, Axhausen KW.
@@
 \leavevmode\hypertarget{ref-Feyman2020effectiveness}{}%
-\CSLLeftMargin{36. }
+\CSLLeftMargin{37. }
 \CSLRightInline{Feyman Y, Bor J, Raifman J, Griffith KN. Effectiveness
@@
 \leavevmode\hypertarget{ref-Praharaj2020Using}{}%
-\CSLLeftMargin{37. }
+\CSLLeftMargin{38. }
 \CSLRightInline{Praharaj S, King D, Pettit C, Wentz E. Using aggregated
@@
 \leavevmode\hypertarget{ref-Paez2020using}{}%
-\CSLLeftMargin{38. }
+\CSLLeftMargin{39. }
 \CSLRightInline{Paez A. Using google community mobility reports to
@@
 \leavevmode\hypertarget{ref-Noland2021mobility}{}%
-\CSLLeftMargin{39. }
+\CSLLeftMargin{40. }
 \CSLRightInline{Noland RB. Mobility and the effective reproduction rate
@@
 \leavevmode\hypertarget{ref-Bhadra2021impact}{}%
-\CSLLeftMargin{40. }
+\CSLLeftMargin{41. }
 \CSLRightInline{Bhadra A, Mukherjee A, Sarkar K. Impact of population
@@
 \leavevmode\hypertarget{ref-Kadi2020population}{}%
-\CSLLeftMargin{41. }
+\CSLLeftMargin{42. }
 \CSLRightInline{Kadi N, Khelfaoui M. Population density, a factor in the
@@
 \leavevmode\hypertarget{ref-Pequeno2020air}{}%
-\CSLLeftMargin{42. }
+\CSLLeftMargin{43. }
 \CSLRightInline{Pequeno P, Mendel B, Rosa C, Bosholn M, Souza JL,
@@
 \leavevmode\hypertarget{ref-Fielding2020social}{}%
-\CSLLeftMargin{43. }
+\CSLLeftMargin{44. }
 \CSLRightInline{Fielding-Miller RK, Sundaram ME, Brouwer K. Social
@@
 \leavevmode\hypertarget{ref-Roy2020factors}{}%
-\CSLLeftMargin{44. }
+\CSLLeftMargin{45. }
 \CSLRightInline{Roy S, Ghosh P. Factors affecting COVID-19 infected and
@@
 \leavevmode\hypertarget{ref-Wong2020spreading}{}%
-\CSLLeftMargin{45. }
+\CSLLeftMargin{46. }
 \CSLRightInline{Wong DWS, Li Y. Spreading of COVID-19: Density matters.
@@
 \leavevmode\hypertarget{ref-Sun2020impacts}{}%
-\CSLLeftMargin{46. }
+\CSLLeftMargin{47. }
 \CSLRightInline{Sun Z, Zhang H, Yang Y, Wan H, Wang Y. Impacts of
@@
 \leavevmode\hypertarget{ref-Skorka2020macroecology}{}%
-\CSLLeftMargin{47. }
+\CSLLeftMargin{48. }
 \CSLRightInline{Skórka P, Grzywacz B, Moroń D, Lenda M. The macroecology
@@
 \leavevmode\hypertarget{ref-White2020state}{}%
-\CSLLeftMargin{48. }
+\CSLLeftMargin{49. }
 \CSLRightInline{White ER, Hébert-Dufresne L. State-level variation of
@@
 \leavevmode\hypertarget{ref-Khavarian2021high}{}%
-\CSLLeftMargin{49. }
+\CSLLeftMargin{50. }
 \CSLRightInline{Khavarian-Garmsir AR, Sharifi A, Moradpour N. Are
@@
 \leavevmode\hypertarget{ref-Hamidi2020longitudinal}{}%
-\CSLLeftMargin{50. }
+\CSLLeftMargin{51. }
 \CSLRightInline{Hamidi S, Ewing R, Sabouri S. Longitudinal analyses of
@@
 \leavevmode\hypertarget{ref-Lee2020human}{}%
-\CSLLeftMargin{51. }
+\CSLLeftMargin{52. }
 \CSLRightInline{Lee M, Zhao J, Sun Q, Pan Y, Zhou W, Xiong C, et al.
@@
 \leavevmode\hypertarget{ref-Amadu2021assessing}{}%
-\CSLLeftMargin{52. }
+\CSLLeftMargin{53. }
 \CSLRightInline{Amadu I, Ahinkorah BO, Afitiri A-R, Seidu A-A, Ameyaw
@@
 \leavevmode\hypertarget{ref-Feng2020spread}{}%
-\CSLLeftMargin{53. }
+\CSLLeftMargin{54. }
 \CSLRightInline{Feng Y, Li Q, Tong X, Wang R, Zhai S, Gao C, et al.
@@
 \leavevmode\hypertarget{ref-Inbaraj2021seroprevalence}{}%
-\CSLLeftMargin{54. }
+\CSLLeftMargin{55. }
 \CSLRightInline{Inbaraj LR, George CE, Chandrasingh S. Seroprevalence of
@@
 \leavevmode\hypertarget{ref-Souris2020covid}{}%
-\CSLLeftMargin{55. }
+\CSLLeftMargin{56. }
 \CSLRightInline{Souris M, Gonzalez J-P. COVID-19: Spatial analysis of
@@
 \leavevmode\hypertarget{ref-Ahmad2020association}{}%
-\CSLLeftMargin{56. }
+\CSLLeftMargin{57. }
 \CSLRightInline{Ahmad K, Erqou S, Shah N, Nazir U, Morrison AR,
@@
 \leavevmode\hypertarget{ref-Noury2021how}{}%
-\CSLLeftMargin{57. }
+\CSLLeftMargin{58. }
 \CSLRightInline{Noury A, François A, Gergaud O, Garel A. How does
@@
 \leavevmode\hypertarget{ref-Wang2021transmission}{}%
-\CSLLeftMargin{58. }
+\CSLLeftMargin{59. }
 \CSLRightInline{Wang F, Tan Z, Yu Z, Yao S, Guo C. Transmission and
@@
 \leavevmode\hypertarget{ref-Brandtner2021creatures}{}%
-\CSLLeftMargin{59. }
+\CSLLeftMargin{60. }
 \CSLRightInline{Brandtner C, Bettencourt LMA, Berman MG, Stier AJ.
@@
 \leavevmode\hypertarget{ref-Stephens2021impact}{}%
-\CSLLeftMargin{60. }
+\CSLLeftMargin{61. }
 \CSLRightInline{Stephens KE, Chernyavskiy P, Bruns DR. Impact of

### 2021-05-20 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Started working on replication of SWN

@@
 ```{r install-data-package, include = FALSE}
-# Run only once if needed to install data package `sobiEquity`
+# Run only once if needed to install data package `r0density`
 if (!require("r0density", character.only = TRUE)) {
@@
 ```{r load-packages, include=FALSE, cache=FALSE}
-library(censReg)
-library(Matrix)
-library(nlme)
+library(censReg) # Package to estimate tobit model
+library(kableExtra) # Package for creating tables
+library(Matrix) # Used to convert list of spatial weights to sparse matrix; needed for spatialprobit
+library(nlme) # Used to estimate 
 library(r0density) # Data package
-library(spatialprobit)
-library(spdep)
-library(tidycensus)
-library(tidyverse)
-library(units)
+library(sf) # Package to work with spatial information in simple features format
+library(spatialprobit) # Used to estimate spatially autoregressive tobit
+library(spdep) # Utilities to create spatial weights and work with spatial data
+library(tidycensus) # Package to retrieve census information 
+library(tidyverse) # Data carpentry and analysis tools
+library(units) # Package to work with units of measurement
 ```
@@
          hincome_log = log(hincome),
-         private = car/commuters * 100) %>% # Percentage of commuters who travel by car/van/truck) %>%
+         private = car/commuters * 100) %>% # Percentage of commuters who travel by car/van/truck (in 10%)
   drop_na(commuters) %>%
@@
 Given the heavy human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has even called this deluge of publications a "carnage of substandard research" [@Bramstedt2020carnage].
-Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [@Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty but somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
-As reported in the literature, only a woefully small proportion of published research was actually reproducible before the pandemic [@Iqbal2016reproducible; @Stodden2018empirical], a situation that does not appear to have changed substantially since [@Sumner2020reproducibility; @Gustot2020quality].
+Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [@Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty and somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
+As reported in the literature, only a woefully small proportion of published research was actually reproducible before the pandemic [@Iqbal2016reproducible; @Stodden2018empirical], and the situation does not appear to have changed substantially since [@Sumner2020reproducibility; @Gustot2020quality].
 
@@
 At stake is the shorter-term impacts that research may have in other spheres of economic and social life.
-The case of economists Reinhart and Rogoff comes to mind: by the time the inaccuracies in their research were uncovered [see @Herndon2014high], their claims about debt and economic growth had already been seized by policy-makers on both sides of the Atlantic to justify austerity policies in the aftermath of the Great Recession of 2007-2009^[Nobel Prize in Economics Paul Krugman noted that "Reinhart–Rogoff may have had more immediate influence on public debate than any previous paper in the history of economics" \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}].
+The case of economists Reinhart and Rogoff comes to mind: by the time the inaccuracies and errors in their research were uncovered [see @Herndon2014high], their claims about debt and economic growth had already been seized by policy-makers on both sides of the Atlantic to justify austerity policies in the aftermath of the Great Recession of 2007-2009^[Nobel Prize in Economics Paul Krugman noted that "Reinhart–Rogoff may have had more immediate influence on public debate than any previous paper in the history of economics" \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}].
 As later research has demonstrated, those policies cast a long shadow, and their sequels continued to be felt for years [@Basu2017ten].
@@
 In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how].
-The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, prompted a flurry of research into the associations between density and the spread of the pandemic.
+The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, brought a torrent of research into the associations between density and the spread of the pandemic.
 Some important questions hang on the results of these research efforts.
@@
 Mathematically, models of exposure and contagion indicate that higher densities can catalyze the transmission of contagious diseases [@Rocklov2020high; @Li2018effect].
-Models such as these were likely at the root of messages, by some figures in positions of authority, that low density regions faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].
+Models such as these were likely at the root of messages, by some figures in positions of authority, that regions with sparse population densities faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic down without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].
 
 As Rocklöv and Sjödin [@Rocklov2020high] note, however, mathematical models of contagion are valid at small-to-medium spatial scales (and presumably, small temporal scales too, such as time spent in restaurants, concert halls, cruises), and the results do not necessarily transfer to larger spatial units and different time scales.
-There are good reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers-however, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place.
+There are solid reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers-however, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place.
 Nonetheless, the idea that high density correlates with high transmission is so intuitive that it is often taken for granted even at larger scales [e.g., @Cruz2020exploring; @Micallef2020first].
-At larger scales, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.
+At larger scales, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, e.g., @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.
 
@@
 In the case of COVID-19, Chauhan et al.
-[@Chauhan2021covid] have found that perception of risks in the US varies between rural, suburban, and urban residents, with rural residents in general displaying less concerns about the virus.
+[@Chauhan2021covid] have found that perception of risks in the US varies between rural, suburban, and urban residents, with rural residents in general expressing less concern about the virus.
 It is possible that people who listened to the message of leaders saying that they were safe because of low density may not have taken adequate precautions against the virus.
@@
 [-@Hamidi2020density] posit this mechanism (i.e., greater compliance with social distancing in denser regions) to explain the results of their analyses.
-The evidence available does indeed show that there were important changes in behavior during the pandemic, at least with respect to mobility [@Jamal2020Changes; @Harris2021Changes; @Molloy2020Tracing]; furthermore, shelter in place orders may have had greater buy-in from the public in higher density regions [@Feyman2020effectiveness], and the behavior may have persisted beyond the duration of official social-distancing policies [@Praharaj2020Using].
+The evidence available does indeed show that there were important changes in behavior with respect to mobility during the pandemic [@Jamal2020Changes; @Harris2021Changes; @Molloy2020Tracing]; furthermore, shelter in place orders may have had greater buy-in from the public in higher density regions [@Feyman2020effectiveness], and the associated behavior may have persisted beyond the duration of official social-distancing policies [@Praharaj2020Using].
 In addition, there is evidence that changes in mobility correlated with the trajectory of the pandemic [@Paez2020using; @Noland2021mobility].
@@
 Wong and Li [-@Wong2020spreading] also found a positive and significant association between population density and number of confirmed COVID-19 cases in US counties, using both univariate and multivariate regressions with spatial effects.
-Most recently, Sy et al.
-[-@Sy2021population] reported that the basic reproductive number of COVID-19 in US counties tends to increase with population density, but at a decreasing rate at higher densities.
+More recently, Sy et al.
+[-@Sy2021population] reported that the basic reproductive number of COVID-19 in US counties tended to increase with population density, but at a decreasing rate at higher densities.
 
-On the gripping hand, a number of studies report non-significant or negative associations between population density and COVID-19 outcomes.
+On the flip side, a number of studies report non-significant or negative associations between population density and COVID-19 outcomes.
 This includes the research of Sun et al.
@@
 Fielding-Miller et al.
-[-@Fielding2020social] found a negative relationship between COVID-19 deaths and population density urban counties in the US.
-In their investigation of doubling time, White and Hébert-Dufresne [-@White2020state] identified a negative and significant correlation between population density and doubling time in US states.
+[-@Fielding2020social] contrast their finding about rural counties with a negative relationship between COVID-19 deaths and population density in urban counties in the US.
+For their part, in their investigation of doubling time, White and Hébert-Dufresne [-@White2020state] identified a negative and significant correlation between population density and doubling time in US states.
 Likewise, [-@Khavarian2021high] fond a small negative (and significant) association between population density and COVID-19 morbidity in districts in Tehran.
-And two of the most complete studies in the US by Hamidi et al.
-[@Hamidi2020longitudinal; @Hamidi2020density] used an extensive set of controls to find negative and significant correlations between density at the level of counties in the US and COVID-19 cases and fatalities.
+Finally, two of the most complete studies in the US [by Hamidi et al.; -@Hamidi2020longitudinal; -@Hamidi2020density] used an extensive set of controls to find negative and significant correlations between density and COVID-19 cases and fatalities at the level of counties in the US.
 
@@
 [-@Sy2021population; hereafter SWN] would immediately grab the attention of a researcher with expertise in spatial modelling.
-Such an expert would likely ask some of the following questions: how were missing counties treated?
-Is it possible to spatially interpolate missing observations?
+
+SWN investigated the basic reproductive number of COVID-19 in US counties, and its association with population density, median household income, and prevalence of private mobility.
+The basic reproductive number can only be calculated reliably with a minimum number of cases, and a large number of counties did not meet such threshold.
+As researchers do, SWN made certain seemingly reasonable modelling decision, for instance basing their analysis only on counties with valid observations.
+An experiences spatial modeler would likely ask some of the following questions on reading SWN's paper: how were missing counties treated?
+Is it possible to spatially interpolate the missing observations?
 What are the implications of the spatial sampling framework used in the analysis?
-Was there evidence of spatial autocorrelation in the residuals of the models?
+And, was there evidence of spatial autocorrelation in the residuals of the models?
 These are questions that in most cases would not occur to a researcher who has not been exposed to spatial statistics or spatial econometrics.
@@
 
-Alas, reproducibility is not necessarily the norm.
+Alas, reproducibility is not necessarily the norm in the relevant literature.
 
@@
 In some cases, there might be legitimate reasons to withhold the data, perhaps due to confidentiality and privacy reasons [e.g., @Lee2020human].
-But in many other cases the data are publicly available, as has been commonly the case with population-level COVID-19 information.
-Often the provenance of the data is documented, but the data themselves are not shared [e.g., @Amadu2021assessing; @Bhadra2021impact; @Cruz2020exploring; @Feng2020spread; @Fielding2020social; @Hamidi2020longitudinal; @Hamidi2020density; @Inbaraj2021seroprevalence; @Souris2020covid].
+But in many other cases the data are publicly available, which in fact has commonly been the case with population-level COVID-19 information.
+Typically the provenance of the data is documented, but in numerous studies the data themselves are not shared [@Amadu2021assessing; @Bhadra2021impact; @Cruz2020exploring; @Feng2020spread; @Fielding2020social; @Hamidi2020longitudinal; @Hamidi2020density; @Inbaraj2021seroprevalence; @Souris2020covid].
 As any researcher can attest, whether a graduate student or a seasoned scientist, collecting, organizing, and preparing data for a project can take a substantial amount of time.
 Pointing to the sources of data, even when these sources are public, is a small step towards reproducibility-but only a very small one.
-Faced with the prospect of having to recreate a data set from raw sources is probably sufficient to dissuade all but the most dedicated (or stubborn) researcher.
+Faced with the prospect of having to recreate a data set from raw sources is probably sufficient to dissuade all but the most dedicated (or stubborn) researcher from independent verification.
 This is true even if part of the data are shared [e.g., @Wong2020spreading].
-In other cases, data are shared, but the processes to document the preparation of the data are not fully documented [e.g., @Ahmad2020association; @Skorka2020macroecology].
-These processes matter, as shown by the errors in the spreadsheets of Reinhart and Rogoff [@Herndon2014high], and the data of biologist Jonathan Pruitt that led to an "avalanche" of paper retractions^[\url{https://doi.org/10.1038/d41586-020-00287-y}].
-Another situation is when papers share well-documented data, but fail to provide the code used in the analysis [e.g., @Noury2021how; @Pequeno2020air; @Wang2021transmission].
+In other cases, data are shared, but the processes followed in the preparation of the data are not fully documented [@Ahmad2020association; @Skorka2020macroecology].
+These processes matter, as shown by the errors in the spreadsheets of Reinhart and Rogoff [@Herndon2014high] and the data of biologist Jonathan Pruitt that led to an "avalanche" of paper retractions^[\url{https://doi.org/10.1038/d41586-020-00287-y}].
+Another situation is when papers share well-documented data, but fail to provide the code used in the analysis [@Noury2021how; @Pequeno2020air; @Wang2021transmission].
 Making code available only "on demand" [e.g., @Brandtner2021creatures] is an unnecessary barrier when most journals offer the facility to share supplemental materials online.
-Then there are those papers that strive towards reproducibility, sharing well-documented processes and data, as well as the code used in any analyses reported [e.g., @Paez2020spatio; @Feyman2020effectiveness; @Stephens2021impact; @White2020state].
+Then there are those papers that more closely comply with reproducibility standards, and share well-documented processes and data, as well as the code used in any analyses reported [@Paez2020spatio; @Feyman2020effectiveness; @Stephens2021impact; @White2020state; @Sy2021population].
 
@@
 
+In the following sections, the analysis of RWN is replicated, some relevant questions from the perspective of a spatial modeler are asked, and the data are reanalyzed.
+
 ## Replicating SWN
 
-Fit (mixed) linear models as in Sy et al.:
-```{r incl}
-table1model1 <- lme(R ~ density_log , 
+SWN examined the association between the basic reproductive number of COVID-19 and population density.
+The basic reproductive number $R_0$ is a summary measure of contact rates, probability of transmission of a pathogen, and duration of infectiousness.
+In rough terms, $R_0$ measures how many new infections each infections begets.
+Infectious disease outbreaks generally tend to die out when $R_0<1$, and to grow when $R_0>1$.
+Reliable calculation of $R_0$ requires a minimum number of cases to be able to assume that there is community transmission of the pathogen.
+Accordingly, SNW based their analysis only on counties that had at least 25 cases or more at the end of the exponential growth phase (see Fig.
+\ref{fig:R0-map}).
+Their final sample included `r prettyNum(nrow(county_geo %>% filter(!is.na(R))), big.mark = ",")` counties in the US, including in Alaska, Hawaii, Puerto Rico, and island territories.
+
+```{r R0-map, echo=FALSE, fig.cap="\\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
+# Choropleth map of basic reproductive number
+ggplot() + # Create a ggplot
+  # Choropleth map of basic reproductive number
+  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
+          aes(fill = R),
+          color = NA) +
+  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
+  scale_fill_viridis_c(name = "R0",
+                       direction = -1) +
+  # Overlay counties with missing values of the reproductive number in white
+  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
+                   R == 0), 
+          fill = "white",
+          color = NA) +
+  # Overlay county boundaries
+  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
+          fill = NA,
+          color = "lightgray") +
+  # Add caption
+  labs(caption = "Note: counties in white represent missing values of the basic reproductive number") +
+  # Set theme of plot
+  theme_minimal()
+```
+
+SWN used for their multivariate analysis mixed linear models.
+This is a reasonable modelling choice: $R_0$ is an interval-ratio variable that is suitably modeled using linear regression; further, as SWN note there is a likelihood that the process in not independent "among counties within each state, potentially due to variable resource allocation and differing health systems across states"  (p.
+3).
+A mixed linear model accounts for this by introducing random components (in the case of SNW, random intercepts at the state level).
+SWN estimated various models with different combinations of variables, including median household income and prevalence of travel by private transportation.
+These are sensible controls, given potential variations in behavior: people in more affluent counties may have greater opportunities to work from home, and use of private transportation reduces contact with strangers.
+Moreover, they also conducted various sensitivity analyses.
+After these efforts, SNW conclude that there is a positive association between the basic reproductive number and population density at the level of counties in the US.
+
+Table \ref{tab:swn-results} replicates Table 1 of SWN (p. 6).
+
+```{r replicate-swn, include=FALSE}
+# Fit (mixed) linear models as in Sy et al. These models are reported in table 1 of their paper
+
+# Model 1
+model1 <- lme(R ~ density_log , 
                     random = ~ 1| state, 
                     data = county_geo_clean %>%
-                      filter(R > 0))
-summary(table1model1)
-intervals(table1model1)
+                      filter(R > 0)) # Exclude observations with missing R
+#summary(table1model1)
+model1_intervals <- intervals(model1)
+
+# Model 2
+model2 <- lme(R ~ density_log + private , 
+                    random = ~ 1| state, 
+                    data = county_geo_clean %>%
+                      filter(R > 0)) # Exclude observations with missing R
+#summary(model2)
+model2_intervals <- intervals(model2)
+
+# Model 3
+model3 <- lme(R ~ density_log + private + hincome, 
+                    random = ~ 1| state, 
+                    data = county_geo_clean %>%
+                      filter(R > 0)) # Exclude observations with missing R
+#summary(model3)
+model3_intervals <- intervals(model3)
+
+# Model 4
+model4 <- lme(R ~ density_log*private + hincome, 
+                    random = ~ 1| state, 
+                    data = county_geo_clean %>%
+                      filter(R > 0)) # Exclude observations with missing R
+#summary(model4)
+model4_intervals <- intervals(model4)
 ```
 
-Linear mixed models of R0 and density + % private transportation
-```{r}
-table1model2 <- lme(R ~ density_log + private , 
-                    random = ~ 1| state, 
-                    data = county_geo_clean %>%
-                      filter(R > 0))
-summary(table1model2)
-intervals(table1model2)
+```{r prepare-swn-results-for-table, include=FALSE}
+# Join tables with results
+main_effects <- full_join(model1_intervals$fixed %>%
+            as.data.frame() %>%
+            rownames_to_column(),
+          model2_intervals$fixed %>%
+            as.data.frame() %>%
+            rownames_to_column(),
+          by = "rowname") %>%
+  full_join(model3_intervals$fixed %>%
+            as.data.frame() %>%
+            rownames_to_column(),
+          by = "rowname")
+
+random_component <- full_join(model1_intervals$reStruct$state %>%
+            as.data.frame() %>%
+            rownames_to_column(),
+          model2_intervals$reStruct$state %>%
+            as.data.frame() %>%
+            rownames_to_column(),
+          by = "rowname") %>%
+  full_join(model3_intervals$reStruct$state %>%
+            as.data.frame() %>%
+            rownames_to_column(),
+          by = "rowname")
+
+within_group_se <- full_join(model1_intervals$sigma %>%
+                               t() %>%
+            as.data.frame() %>%
+            rownames_to_column(),
+          model2_intervals$sigma %>%
+            t() %>%
+            as.data.frame() %>%
+            rownames_to_column(),
+          by = "rowname") %>%
+  full_join(model3_intervals$sigma %>%
+              t() %>%
+            as.data.frame() %>%
+            rownames_to_column(),
+          by = "rowname")
+
+results_swn <- rbind(main_effects,
+                     random_component,
+                     within_group_se)
+
+colnames(results_swn) <- c("Variable", "l1", "b1", "u1", "l2", "b2", "u2", "l3", "b3", "u3")
+
+results_swn <- results_swn %>%
+  transmute(Variable,
+            b1 = ifelse(is.na(b1),
+                        " ", 
+                        round(b1, 3)),
+            ci1 = ifelse(is.na(l1), 
+                         " ",
+                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")),
+            b2 = ifelse(is.na(b2),
+                        " ", 
+                        round(b2, 3)),
+            ci2 = ifelse(is.na(l2), 
+                         " ",
+                         paste0("[", round(l2, 3), ", ", round(u2, 3), "]")),
+            b3 = ifelse(is.na(b3),
+                        " ", 
+                        round(b3, 3)),
+            ci3 = ifelse(is.na(l3), 
+                         " ",
+                         paste0("[", round(l3, 3), ", ", round(u3, 3), "]")))
 ```
 
-Linear mixed models of R0 and density + % private transportation + median income
-```{r}
-table1model3 <- lme(R ~ density_log + private + hincome, 
-                    random = ~ 1| state, 
-                    data = county_geo_clean %>%
-                      filter(R > 0))
-summary(table1model3)
-intervals(table1model3)
-```
-
-Linear mixed models of R0 and density + % private transportation + median income + interaction of density and % private transportation
-```{r}
-table1model4 <- lme(R ~ density_log*private + hincome, 
-                    random = ~ 1| state, 
-                    data = county_geo_clean %>%
-                      filter(R > 0))
-summary(table1model4)
-intervals(table1model4)
+```{r tabulate-swn-results, echo=FALSE}
+results_swn %>%
+  mutate(Variable = c("Intercept", 
+                      "Log of population density",
+                      "Percent of private transportation", 
+                      "Median household income ($10,000)",
+                      "Standard deviation (Intercept)",
+                      "Within-group standard error")) %>%
+  kable("latex",
+        digits = 3,
+        booktabs = TRUE,
+        col.names = c("Variable",
+                      "beta",
+                      "95% CI",
+                      "beta",
+                      "95% CI",
+                      "beta",
+                      "95% CI"),
+        caption = "\\label{tab:swn-results}Replicating SWN: Models 1-3") %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  add_header_above(c(" ", "Model 1" = 2, "Model 2" = 2, "Model 3" = 2))
 ```

### 2021-05-20 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Started working on replication of SWN

@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
+
 \usepackage{forarray}
@@
 guidelines {[}6--9{]}, in practice reproducibility has remained a lofty
-but somewhat aspirational goal {[}10,11{]}. As reported in the
+and somewhat aspirational goal {[}10,11{]}. As reported in the
 literature, only a woefully small proportion of published research was
-actually reproducible before the pandemic {[}12,13{]}, a situation that
+actually reproducible before the pandemic {[}12,13{]}, and the situation
 does not appear to have changed substantially since {[}14,15{]}.
@@
 life. The case of economists Reinhart and Rogoff comes to mind: by the
-time the inaccuracies in their research were uncovered {[}see 16{]},
-their claims about debt and economic growth had already been seized by
-policy-makers on both sides of the Atlantic to justify austerity
-policies in the aftermath of the Great Recession of
+time the inaccuracies and errors in their research were uncovered {[}see
+16{]}, their claims about debt and economic growth had already been
+seized by policy-makers on both sides of the Atlantic to justify
+austerity policies in the aftermath of the Great Recession of
 2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
@@
 felt in dense population centers such as Wuhan, Milan, Madrid, and New
-York, prompted a flurry of research into the associations between
+York, brought a torrent of research into the associations between
 density and the spread of the pandemic. Some important questions hang on
@@
 the root of messages, by some figures in positions of authority, that
-low density regions faced lower risks from the pandemic\footnote{Governor
-  Kristi Noem of South Dakota, for example, claimed that sparse
-  population density allowed her state to face the pandemic without the
-  need for strict policy interventions
+regions with sparse population densities faced lower risks from the
+pandemic\footnote{Governor Kristi Noem of South Dakota, for example,
+  claimed that sparse population density allowed her state to face the
+  pandemic down without the need for strict policy interventions
   \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.
@@
 halls, cruises), and the results do not necessarily transfer to larger
-spatial units and different time scales. There are good reasons for
+spatial units and different time scales. There are solid reasons for
 this: while in a restaurant, one can hardly avoid being in proximity to
@@
 capture in the mechanistic framework of differential equations {[}or can
-be missing in agent-based models, 27{]}; these adaptations, in fact, can
-be a key aspect of disease transmission.
+be missing in agent-based models, e.g., 27{]}; these adaptations, in
+fact, can be a key aspect of disease transmission.
 
@@
 the US varies between rural, suburban, and urban residents, with rural
-residents in general displaying less concerns about the virus. It is
+residents in general expressing less concern about the virus. It is
 possible that people who listened to the message of leaders saying that
@@
 available does indeed show that there were important changes in behavior
-during the pandemic, at least with respect to mobility {[}34--36{]};
-furthermore, shelter in place orders may have had greater buy-in from
-the public in higher density regions {[}37{]}, and the behavior may have
+with respect to mobility during the pandemic {[}34--36{]}; furthermore,
+shelter in place orders may have had greater buy-in from the public in
+higher density regions {[}37{]}, and the associated behavior may have
 persisted beyond the duration of official social-distancing policies
@@
 counties, using both univariate and multivariate regressions with
-spatial effects. Most recently, Sy et al. {[}19{]} reported that the
-basic reproductive number of COVID-19 in US counties tends to increase
+spatial effects. More recently, Sy et al. {[}19{]} reported that the
+basic reproductive number of COVID-19 in US counties tended to increase
 with population density, but at a decreasing rate at higher densities.
 
-On the gripping hand, a number of studies report non-significant or
-negative associations between population density and COVID-19 outcomes.
-This includes the research of Sun et al. {[}47{]} who did not find
-evidence of significant correlation between population density and
-confirmed number of cases per day \emph{in conditions of lockdown} in
-China. This finding echoes the results of Paez et al. {[}32{]}, who in
-their study of provinces in Spain reported non-significant associations
-between population density and infection rates in the early days of the
-first wave of COVID-19, and negative significant associations in the
-later part of the first lockdown. Similarly, {[}48{]} found zero or
-negative associations between population density and infection
-numbers/deaths by country. Fielding-Miller et al. {[}44{]} found a
-negative relationship between COVID-19 deaths and population density
-urban counties in the US. In their investigation of doubling time, White
-and Hébert-Dufresne {[}49{]} identified a negative and significant
-correlation between population density and doubling time in US states.
-Likewise, {[}50{]} fond a small negative (and significant) association
-between population density and COVID-19 morbidity in districts in
-Tehran. And two of the most complete studies in the US by Hamidi et al.
-{[}33,51{]} used an extensive set of controls to find negative and
-significant correlations between density at the level of counties in the
-US and COVID-19 cases and fatalities.
+On the flip side, a number of studies report non-significant or negative
+associations between population density and COVID-19 outcomes. This
+includes the research of Sun et al. {[}47{]} who did not find evidence
+of significant correlation between population density and confirmed
+number of cases per day \emph{in conditions of lockdown} in China. This
+finding echoes the results of Paez et al. {[}32{]}, who in their study
+of provinces in Spain reported non-significant associations between
+population density and infection rates in the early days of the first
+wave of COVID-19, and negative significant associations in the later
+part of the first lockdown. Similarly, {[}48{]} found zero or negative
+associations between population density and infection numbers/deaths by
+country. Fielding-Miller et al. {[}44{]} contrast their finding about
+rural counties with a negative relationship between COVID-19 deaths and
+population density in urban counties in the US. For their part, in their
+investigation of doubling time, White and Hébert-Dufresne {[}49{]}
+identified a negative and significant correlation between population
+density and doubling time in US states. Likewise, {[}50{]} fond a small
+negative (and significant) association between population density and
+COVID-19 morbidity in districts in Tehran. Finally, two of the most
+complete studies in the US {[}by Hamidi et al.; {[}51{]}; {[}33{]}{]}
+used an extensive set of controls to find negative and significant
+correlations between density and COVID-19 cases and fatalities at the
+level of counties in the US.
 
@@
 al.~{[}{[}19{]}; hereafter SWN{]} would immediately grab the attention
-of a researcher with expertise in spatial modelling. Such an expert
-would likely ask some of the following questions: how were missing
-counties treated? Is it possible to spatially interpolate missing
+of a researcher with expertise in spatial modelling.
+
+SWN investigated the basic reproductive number of COVID-19 in US
+counties, and its association with population density, median household
+income, and prevalence of private mobility. The basic reproductive
+number can only be calculated reliably with a minimum number of cases,
+and a large number of counties did not meet such threshold. As
+researchers do, SWN made certain seemingly reasonable modelling
+decision, for instance basing their analysis only on counties with valid
+observations. An experiences spatial modeler would likely ask some of
+the following questions on reading SWN's paper: how were missing
+counties treated? Is it possible to spatially interpolate the missing
 observations? What are the implications of the spatial sampling
-framework used in the analysis? Was there evidence of spatial
+framework used in the analysis? And, was there evidence of spatial
 autocorrelation in the residuals of the models? These are questions that
@@
 
-Alas, reproducibility is not necessarily the norm.
+Alas, reproducibility is not necessarily the norm in the relevant
+literature.
 
@@
 perhaps due to confidentiality and privacy reasons {[}e.g., 52{]}. But
-in many other cases the data are publicly available, as has been
-commonly the case with population-level COVID-19 information. Often the
-provenance of the data is documented, but the data themselves are not
-shared {[}25,33,41,44,51,e.g., 53,54--56{]}. As any researcher can
-attest, whether a graduate student or a seasoned scientist, collecting,
-organizing, and preparing data for a project can take a substantial
-amount of time. Pointing to the sources of data, even when these sources
-are public, is a small step towards reproducibility-but only a very
-small one. Faced with the prospect of having to recreate a data set from
-raw sources is probably sufficient to dissuade all but the most
-dedicated (or stubborn) researcher. This is true even if part of the
-data are shared {[}e.g., 46{]}. In other cases, data are shared, but the
-processes to document the preparation of the data are not fully
-documented {[}48,e.g., 57{]}. These processes matter, as shown by the
-errors in the spreadsheets of Reinhart and Rogoff {[}16{]}, and the data
-of biologist Jonathan Pruitt that led to an ``avalanche'' of paper
+in many other cases the data are publicly available, which in fact has
+commonly been the case with population-level COVID-19 information.
+Typically the provenance of the data is documented, but in numerous
+studies the data themselves are not shared {[}25,33,41,44,51,53--56{]}.
+As any researcher can attest, whether a graduate student or a seasoned
+scientist, collecting, organizing, and preparing data for a project can
+take a substantial amount of time. Pointing to the sources of data, even
+when these sources are public, is a small step towards
+reproducibility-but only a very small one. Faced with the prospect of
+having to recreate a data set from raw sources is probably sufficient to
+dissuade all but the most dedicated (or stubborn) researcher from
+independent verification. This is true even if part of the data are
+shared {[}e.g., 46{]}. In other cases, data are shared, but the
+processes followed in the preparation of the data are not fully
+documented {[}48,57{]}. These processes matter, as shown by the errors
+in the spreadsheets of Reinhart and Rogoff {[}16{]} and the data of
+biologist Jonathan Pruitt that led to an ``avalanche'' of paper
 retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
 Another situation is when papers share well-documented data, but fail to
-provide the code used in the analysis {[}43,e.g., 58,59{]}. Making code
+provide the code used in the analysis {[}43,58,59{]}. Making code
 available only ``on demand'' {[}e.g., 60{]} is an unnecessary barrier
 when most journals offer the facility to share supplemental materials
-online. Then there are those papers that strive towards reproducibility,
-sharing well-documented processes and data, as well as the code used in
-any analyses reported {[}e.g., 32,37,49,61{]}.
+online. Then there are those papers that more closely comply with
+reproducibility standards, and share well-documented processes and data,
+as well as the code used in any analyses reported {[}19,32,37,49,61{]}.
+
+In the following sections, the analysis of RWN is replicated, some
+relevant questions from the perspective of a spatial modeler are asked,
+and the data are reanalyzed.
 
@@
 
-Fit (mixed) linear models as in Sy et al.:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{table1model1 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log , }
-                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
-                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
-                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(table1model1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Linear mixed-effects model fit by REML
-##   Data: county_geo_clean %>% filter(R > 0) 
-##        AIC      BIC    logLik
-##   2383.621 2403.808 -1187.811
-## 
-## Random effects:
-##  Formula: ~1 | state
-##         (Intercept)  Residual
-## StdDev:    0.165504 0.6648325
-## 
-## Fixed effects:  R ~ density_log 
-##                 Value  Std.Error   DF  t-value p-value
-## (Intercept) 2.2740045 0.05468629 1099 41.58272       0
-## density_log 0.1621127 0.01485543 1099 10.91269       0
-##  Correlation: 
-##             (Intr)
-## density_log 0.797 
-## 
-## Standardized Within-Group Residuals:
-##        Min         Q1        Med         Q3        Max 
-## -2.3847932 -0.5972261 -0.1682299  0.3681864 16.2716371 
-## 
-## Number of Observations: 1151
-## Number of Groups: 51
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{intervals}\NormalTok{(table1model1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Approximate 95% confidence intervals
-## 
-##  Fixed effects:
-##                 lower      est.     upper
-## (Intercept) 2.1667032 2.2740045 2.3813058
-## density_log 0.1329645 0.1621127 0.1912609
-## attr(,"label")
-## [1] "Fixed effects:"
-## 
-##  Random Effects:
-##   Level: state 
-##                     lower     est.     upper
-## sd((Intercept)) 0.1080424 0.165504 0.2535261
-## 
-##  Within-group standard error:
-##     lower      est.     upper 
-## 0.6375478 0.6648325 0.6932849
-\end{verbatim}
-
-Linear mixed models of R0 and density + \% private transportation
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{table1model2 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private , }
-                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
-                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
-                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(table1model2)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Linear mixed-effects model fit by REML
-##   Data: county_geo_clean %>% filter(R > 0) 
-##        AIC      BIC    logLik
-##   2385.478 2410.707 -1187.739
-## 
-## Random effects:
-##  Formula: ~1 | state
-##         (Intercept)  Residual
-## StdDev:   0.1358322 0.6649714
-## 
-## Fixed effects:  R ~ density_log + private 
-##                 Value Std.Error   DF   t-value p-value
-## (Intercept)  3.347070 0.3418910 1098  9.789874  0.0000
-## density_log  0.145258 0.0156188 1098  9.300177  0.0000
-## private     -0.012547 0.0039434 1098 -3.181760  0.0015
-##  Correlation: 
-##             (Intr) dnsty_
-## density_log -0.222       
-## private     -0.988  0.344
-## 
-## Standardized Within-Group Residuals:
-##        Min         Q1        Med         Q3        Max 
-## -2.7347592 -0.6005758 -0.1610880  0.3876853 16.2104980 
-## 
-## Number of Observations: 1151
-## Number of Groups: 51
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{intervals}\NormalTok{(table1model2)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Approximate 95% confidence intervals
-## 
-##  Fixed effects:
-##                   lower       est.        upper
-## (Intercept)  2.67623619  3.3470697  4.017903160
-## density_log  0.11461177  0.1452579  0.175904019
-## private     -0.02028416 -0.0125468 -0.004809444
-## attr(,"label")
-## [1] "Fixed effects:"
-## 
-##  Random Effects:
-##   Level: state 
-##                      lower      est.    upper
-## sd((Intercept)) 0.08059718 0.1358322 0.228921
-## 
-##  Within-group standard error:
-##     lower      est.     upper 
-## 0.6376445 0.6649714 0.6934695
-\end{verbatim}
-
-Linear mixed models of R0 and density + \% private transportation +
-median income
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{table1model3 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private }\SpecialCharTok{+}\NormalTok{ hincome, }
-                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
-                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
-                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(table1model3)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Linear mixed-effects model fit by REML
-##   Data: county_geo_clean %>% filter(R > 0) 
-##       AIC     BIC   logLik
-##   2393.98 2424.25 -1190.99
-## 
-## Random effects:
-##  Formula: ~1 | state
-##         (Intercept)  Residual
-## StdDev:   0.1373995 0.6651151
-## 
-## Fixed effects:  R ~ density_log + private + hincome 
-##                 Value Std.Error   DF   t-value p-value
-## (Intercept)  3.385550 0.3933810 1097  8.606289  0.0000
-## density_log  0.146826 0.0171534 1097  8.559585  0.0000
-## private     -0.012707 0.0040548 1097 -3.133786  0.0018
-## hincome     -0.003256 0.0151105 1097 -0.215493  0.8294
-##  Correlation: 
-##             (Intr) dnsty_ privat
-## density_log  0.027              
-## private     -0.950  0.211       
-## hincome     -0.493 -0.412  0.228
-## 
-## Standardized Within-Group Residuals:
-##        Min         Q1        Med         Q3        Max 
-## -2.7162603 -0.5984817 -0.1626793  0.3887931 16.2029477 
-## 
-## Number of Observations: 1151
-## Number of Groups: 51
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{intervals}\NormalTok{(table1model3)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Approximate 95% confidence intervals
-## 
-##  Fixed effects:
-##                   lower         est.       upper
-## (Intercept)  2.61368596  3.385550103  4.15741424
-## density_log  0.11316880  0.146825984  0.18048317
-## private     -0.02066315 -0.012707018 -0.00475089
-## hincome     -0.03290488 -0.003256195  0.02639249
-## attr(,"label")
-## [1] "Fixed effects:"
-## 
-##  Random Effects:
-##   Level: state 
-##                    lower      est.     upper
-## sd((Intercept)) 0.081461 0.1373995 0.2317504
-## 
-##  Within-group standard error:
-##     lower      est.     upper 
-## 0.6377582 0.6651151 0.6936455
-\end{verbatim}
-
-Linear mixed models of R0 and density + \% private transportation +
-median income + interaction of density and \% private transportation
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{table1model4 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log}\SpecialCharTok{*}\NormalTok{private }\SpecialCharTok{+}\NormalTok{ hincome, }
-                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
-                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
-                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(table1model4)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Linear mixed-effects model fit by REML
-##   Data: county_geo_clean %>% filter(R > 0) 
-##       AIC      BIC   logLik
-##   2407.04 2442.349 -1196.52
-## 
-## Random effects:
-##  Formula: ~1 | state
-##         (Intercept)  Residual
-## StdDev:   0.1378096 0.6653589
-## 
-## Fixed effects:  R ~ density_log * private + hincome 
-##                         Value Std.Error   DF   t-value p-value
-## (Intercept)          3.415859 0.4157861 1096  8.215424  0.0000
-## density_log          0.177645 0.1350641 1096  1.315266  0.1887
-## private             -0.013123 0.0044568 1096 -2.944502  0.0033
-## hincome             -0.002719 0.0153117 1096 -0.177607  0.8591
-## density_log:private -0.000354 0.0015414 1096 -0.229894  0.8182
-##  Correlation: 
-##                     (Intr) dnsty_ privat hincom
-## density_log          0.323                     
-## private             -0.952 -0.386              
-## hincome             -0.409  0.105  0.139       
-## density_log:private -0.323 -0.992  0.414 -0.158
-## 
-## Standardized Within-Group Residuals:
-##        Min         Q1        Med         Q3        Max 
-## -2.7371582 -0.5968733 -0.1638499  0.3883700 16.1980628 
-## 
-## Number of Observations: 1151
-## Number of Groups: 51
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{intervals}\NormalTok{(table1model4)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Approximate 95% confidence intervals
-## 
-##  Fixed effects:
-##                            lower          est.        upper
-## (Intercept)          2.600032192  3.4158588336  4.231685475
-## density_log         -0.087368297  0.1776452292  0.442658756
-## private             -0.021867960 -0.0131231030 -0.004378246
-## hincome             -0.032762972 -0.0027194668  0.027324038
-## density_log:private -0.003378809 -0.0003543603  0.002670088
-## attr(,"label")
-## [1] "Fixed effects:"
-## 
-##  Random Effects:
-##   Level: state 
-##                      lower      est.     upper
-## sd((Intercept)) 0.08169505 0.1378096 0.2324679
-## 
-##  Within-group standard error:
-##     lower      est.     upper 
-## 0.6379755 0.6653589 0.6939176
-\end{verbatim}
+SWN examined the association between the basic reproductive number of
+COVID-19 and population density. The basic reproductive number \(R_0\)
+is a summary measure of contact rates, probability of transmission of a
+pathogen, and duration of infectiousness. In rough terms, \(R_0\)
+measures how many new infections each infections begets. Infectious
+disease outbreaks generally tend to die out when \(R_0<1\), and to grow
+when \(R_0>1\). Reliable calculation of \(R_0\) requires a minimum
+number of cases to be able to assume that there is community
+transmission of the pathogen. Accordingly, SNW based their analysis only
+on counties that had at least 25 cases or more at the end of the
+exponential growth phase (see Fig. \ref{fig:R0-map}). Their final sample
+included 1,151 counties in the US, including in Alaska, Hawaii, Puerto
+Rico, and island territories.
+
+\begin{figure}
+\centering
+\includegraphics{R0-Density-Reanalysis_files/figure-latex/R0-map-1.pdf}
+\caption{\label{fig:R0-map}Basic reproductive rate in US counties
+(Alaska, Hawaii, Puerto Rico, and territories not shown).}
+\end{figure}
+
+SWN used for their multivariate analysis mixed linear models. This is a
+reasonable modelling choice: \(R_0\) is an interval-ratio variable that
+is suitably modeled using linear regression; further, as SWN note there
+is a likelihood that the process in not independent ``among counties
+within each state, potentially due to variable resource allocation and
+differing health systems across states'' (p.~3). A mixed linear model
+accounts for this by introducing random components (in the case of SNW,
+random intercepts at the state level). SWN estimated various models with
+different combinations of variables, including median household income
+and prevalence of travel by private transportation. These are sensible
+controls, given potential variations in behavior: people in more
+affluent counties may have greater opportunities to work from home, and
+use of private transportation reduces contact with strangers. Moreover,
+they also conducted various sensitivity analyses. After these efforts,
+SNW conclude that there is a positive association between the basic
+reproductive number and population density at the level of counties in
+the US.
+
+Table \ref{tab:swn-results} replicates Table 1 of SWN (p.~6).
+
+\begin{table}
+
+\caption{\label{tab:tabulate-swn-results}\label{tab:swn-results}Replicating SWN: Models 1-3}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllllrl}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Model 1} & \multicolumn{2}{c}{Model 2} & \multicolumn{2}{c}{Model 3} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
+Variable & beta & 95\% CI & beta & 95\% CI & beta & 95\% CI\\
+\midrule
+Intercept & 2.274 & [2.167, 2.381] & 3.347 & [2.676, 4.018] & 3.386 & [2.614, 4.157]\\
+Log of population density & 0.162 & [0.133, 0.191] & 0.145 & [0.115, 0.176] & 0.147 & [0.113, 0.18]\\
+Percent of private transportation &  &  & -0.013 & [-0.02, -0.005] & -0.013 & [-0.021, -0.005]\\
+Median household income (\$10,000) &  &  &  &  & -0.003 & [-0.033, 0.026]\\
+Standard deviation (Intercept) & 0.166 & [0.108, 0.254] & 0.136 & [0.081, 0.229] & 0.137 & [0.081, 0.232]\\
+\addlinespace
+Within-group standard error & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.694]\\
+\bottomrule
+\end{tabular}}
+\end{table}
 
@@
 ## ----MCMC spatial autoregressive Tobit model ----
-## Execution time  = 40.052 secs
+## Execution time  = 41.022 secs
 ## 

### 2021-05-20 r0density/raw-data-processing/00-Data-Preprocessing.Rmd
Recreated data package to include boundaries of the states

@@
 
+Retrieve the state boundaries for plotting:
+```{r}
+state_geo <- get_decennial(geography = "state", 
+                            variables = "P001001", # Total population
+                            year = 2010, 
+                            geometry = TRUE, 
+                            keep_geo_vars = TRUE) %>%
+  rename(population = value) %>%
+  select(-variable)
+```
+
 Now join Sy's data to the counties with geometry and other socio-economic variables (the equivalent variables are dropped from Sy et al.'s table, because they are missing all the counties without a basic reproductive number):
@@
 
+Prepare state boundaries for saving
+```{r}
+state_geo <- state_geo %>%
+  select(-CENSUSAREA, -starts_with("geometry"), geometry)
+```
+
 Save data table:
@@
 save(county_geo, file = "county_geo.RData", compress = "xz")
+save(state_geo, file = "state_geo.RData", compress = "xz")
 ```
@@
 Copy data file to data folder!
-
-<!--
-
-Convert missing R's to zeros and calculate expansion variables:
-```{r}
-county_geo_clean <- county_geo %>%
-  mutate(R = replace_na(R, 0),
-         density_2 = density^2) %>%
-  drop_na(commuters) %>%
-  st_as_sf()
-```
-
-Create list of neighbors:
-```{r}
-county_nb <- poly2nb(county_geo_clean)
-```
-
-Check list of neighbors:
-```{r}
-summary(county_nb)
-```
-
-Notice that there are 9 counties that are isolates. Which counties are these?
-```{r}
-county_geo_clean[c(69, 546, 547, 549, 1226, 1876, 2974, 3146, 3200),]
-```
-
-Find nearest neighbors by distance:
-```{r}
-county_nb_dist <- knearneigh(county_geo_clean %>% 
-                               st_centroid()) %>%
-  knn2nb()
-```
-
-Add nearest neighbor to isolates:
-```{r}
-county_nb[[69]] <- county_nb_dist[[69]]
-county_nb[[546]] <- county_nb_dist[[546]]
-county_nb[[547]] <- county_nb_dist[[547]]
-county_nb[[549]] <- county_nb_dist[[549]]
-county_nb[[1226]] <- county_nb_dist[[1226]]
-county_nb[[1876]] <- county_nb_dist[[1876]]
-county_nb[[2974]] <- county_nb_dist[[2974]]
-county_nb[[3146]] <- county_nb_dist[[3146]]
-county_nb[[3200]] <- county_nb_dist[[3200]]
-```
-
-Convert nearest neighbors to listw:
-```{r}
-nb_B <- nb2listw(county_nb, 
-                 style="B", 
-                 zero.policy = TRUE)
-nb_B$style
-```
-
-Convert listw to sparse matrix for use in spatial tobit:
-```{r}
-B <- as(nb_B, "CsparseMatrix")
-all(B == t(B))
-```
-
-## Replicate Sy et al.
-
-Fit (mixed) linear models as in Sy et al.:
-```{r}
-table1model1 <- lme(R ~ density_log , 
-                    random = ~ 1| state, 
-                    data = county_geo_clean %>%
-                      filter(R > 0))
-summary(table1model1)
-intervals(table1model1)
-```
-
-Linear mixed models of R0 and density + % private transportation
-```{r}
-table1model2 <- lme(R ~ density_log + private , 
-                    random = ~ 1| state, 
-                    data = county_geo_clean %>%
-                      filter(R > 0))
-summary(table1model2)
-intervals(table1model2)
-```
-
-Linear mixed models of R0 and density + % private transportation + median income
-```{r}
-table1model3 <- lme(R ~ density_log + private + hincome, 
-                    random = ~ 1| state, 
-                    data = county_geo_clean %>%
-                      filter(R > 0))
-summary(table1model3)
-intervals(table1model3)
-```
-
-Linear mixed models of R0 and density + % private transportation + median income + interaction of density and % private transportation
-```{r}
-table1model4 <- lme(R ~ density_log*private + hincome, 
-                    random = ~ 1| state, 
-                    data = county_R0_demog)
-summary(table1model4)
-intervals(table1model4)
-```
-
-## Fit tobit version of models
-
-```{r}
-table1model1 <- censReg(R ~ density_log, 
-                        left = 0,
-                    data = county_geo_clean)
-summary(table1model1)
-```
-
-Linear mixed models of R0 and density + % private transportation
-```{r}
-table1model2 <- censReg(R ~ density_log + private, 
-                    data = county_geo_clean)
-summary(table1model2)
-```
-
-Linear mixed models of R0 and density + % private transportation + median income
-```{r}
-table1model3 <- censReg(R ~ density_log + private + hincome, 
-                    data = county_geo_clean)
-summary(table1model3)
-```
-
-Tobit models of R0 and density + % private transportation + median income + interaction of density and % private transportation
-```{r}
-table1model4 <- censReg(R ~ density_log*private + hincome, 
-                    data = county_geo_clean)
-summary(table1model4)
-
-```
-
-## Spatially autoregressive tobit
-
-Fit spatially autoregressive tobit:
-```{r}
-# Fit SAR Tobit
-fit_sartobit <- sartobit(R ~ density_log + private + hincome_log,
-                         B,
-                         ndraw = 1000,
-                         burn.in = 200, 
-                         showProgress = TRUE,
-                         data = county_geo_clean,
-                         computeMarginalEffects = TRUE)
-summary(fit_sartobit)
-impacts(fit_sartobit)
-```
-
-With interactions:
-```{r}
-# Fit SAR Tobit
-fit_sartobit <- sartobit(R ~ density_log + density_log:private + hincome_log,
-                         B,
-                         ndraw = 1000,
-                         burn.in = 200, 
-                         showProgress = TRUE,
-                         data = county_geo_clean)
-summary(fit_sartobit)
-impacts(fit_sartobit)
-```
-FIGURE 2 - POPULATION DENSITY THRESHOLD
-
-Compare density of counties with oubreaks vs counties without outbreak
-```{r}
-county_outbreak_yn %>%
-  group_by(included) %>%
-  summarise(mean = median(Density, na.rm=TRUE), 
-            lower_IQR = summary(Density)[2],
-            upper_IQR=summary(Density)[5], 
-            min=summary(Density)[1], 
-            max=summary(Density)[6])
-
-summary(county_outbreak_yn$Density)
-wilcox.test(county_outbreak_yn$Density~county_outbreak_yn$included)
-```
-
-Boxplot of counties with oubreaks vs counties without outbreak
-```{r}
-county_outbreak_yn$included <- as.factor(county_outbreak_yn$included)
-
-ggplot(county_outbreak_yn, aes(included, Density)) + 
-  geom_boxplot(fill="gray92", width=0.5) + 
-  ylim(0,200) + 
-  #theme_minimal() + 
-  ylab("Population Density (population/sq km)") + 
-  xlab("United States Counties") + 
-  theme_minimal(base_size=13)
-```
-
-TABLE 1 - LINEAR MIXED MODELS
-
-Linear mixed models of R0 and density:
-```{r}
-table1model1 <- lme(R ~ density_log , 
-                    random = ~ 1| state, 
-                    data = county_geo_clean %>%
-                      filter(R > 0))
-summary(table1model1)
-intervals(table1model1)
-```
-
-Linear mixed models of R0 and density + % private transportation
-```{r}
-table1model2 <- lme(R ~ density_log + private , 
-                    random = ~ 1| state, 
-                    data = county_geo_clean %>%
-                      filter(R > 0))
-summary(table1model2)
-intervals(table1model2)
-```
-
-Linear mixed models of R0 and density + % private transportation + median income
-```{r}
-table1model3 <- lme(R ~ density_log + private + hincome, 
-                    random = ~ 1| state, 
-                    data = county_geo_clean %>%
-                      filter(R > 0))
-summary(table1model3)
-intervals(table1model3)
-```
-
-Linear mixed models of R0 and density + % private transportation + median income + interaction of density and % private transportation
-```{r}
-table1model4 <- lme(R ~ density_log*private + hincome, 
-                    random = ~ 1| state, 
-                    data = county_R0_demog)
-summary(table1model4)
-intervals(table1model4)
-```
-
-## TABLE 2 - SENSITIVITY ANALYSIS
-
-a. Deaths only - Subset counties with R0 from deaths only
-
-```{r}
-county_R0_demog_a <- subset(county_R0_demog, !is.na(county_R0_demog$R_death))
-```
-
-Linear mixed models of R0 (death) and density 
-```{r}
-table2model1_a <- lme(R_death ~ density_log , random = ~ 1| state, data = county_R0_demog_a)
-summary(table2model1_a)
-intervals(table2model1_a)
-```
-
-Linear mixed models of R0 (death) and density + % private transportation
-```{r}
-table2model2_a <- lme(R_death~ density_log + private , random = ~ 1| state, data = county_R0_demog_a)
-summary(table2model2_a)
-intervals(table2model2_a)
-```
-
-Linear mixed models of R0 (death) and density + % private transportation + median income
-```{r}
-table2model3_a <- lme(R_death~ density_log + private + hincome, random = ~ 1| state, data = county_R0_demog_a)
-summary(table2model3_a)
-intervals(table2model3_a)
-```
-
-Linear mixed models of R0 (death) and density + % private transportation + median income + interaction of density and % private transportation
-```{r}
-table2model4_a <- lme(R_death~ density_log*private + hincome, random = ~ 1| state, data = county_R0_demog_a)
-summary(table2model4_a)
-intervals(table2model4_a)
-```
-
-b. No counties adjacent to high density counties - within 15 mile radius, high density > 144.0492 population/sq km
-```{r}
-county_R0_demog_b <- county_R0_demog[county_R0_demog$sensitivitydensity,]
-```
-
-Linear mixed models of R0 and density (no counties adjacent to high density counties)
-```{r}
-table2model1_b <- lme(R ~ density_log , random = ~ 1| state, data = county_R0_demog_b)
-summary(table2model1_b)
-intervals(table2model1_b)
-```
-
-Linear mixed models of R0 and density + % private transportation (no counties adjacent to high density counties)
-```{r}
-table2model2_b <- lme(R ~ density_log + private , random = ~ 1| state, data = county_R0_demog_b)
-summary(table2model2_b)
-intervals(table2model2_b)
-```
-
-Linear mixed models of R0 and density + % private transportation + median income (no counties adjacent to high density counties)
-```{r}
-table2model3_b <- lme(R ~ density_log + private + hincome, random = ~ 1| state, data = county_R0_demog_b)
-summary(table2model3_b)
-intervals(table2model3_b)
-```
-
-Linear mixed models of R0 and density + % private transportation + median income + interaction of density and % private transportation (no counties adjacent to high density counties)
-```{r}
-table2model4_b <- lme(R ~ density_log*private + hincome , random = ~ 1| state, data = county_R0_demog_b)
-summary(table2model4_b)
-intervals(table2model4_b)
-```
-
-c. Removing high influence counties - Cook's D over 4/N for each model)
-
-Linear mixed models of R0 and density (no outliers)
-```{r}
-sample_size<-nobs(table1model1)
-cooksd1 <- CookD(table1model1)
-influential <- as.numeric(names(cooksd1)[(cooksd1 > (4/sample_size))])
-county_R0_demog_c <- county_R0_demog[-influential, ]
-table1model1_c <- lme(R ~ density_log , random = ~ 1| state, data = county_R0_demog_c)
-summary(table1model1_c)
-intervals(table1model1_c)
-```
-
-Linear mixed models of R0 and density + % private transportation (no outliers)
-```{r}
-sample_size<-nobs(table1model2)
-cooksd1 <- CookD(table1model2)
-influential <- as.numeric(names(cooksd1)[(cooksd1 > (4/sample_size))])
-county_R0_demog_c <- county_R0_demog[-influential, ]
-table1model2_c <- lme(R ~ density_log + private , random = ~ 1| state, data = county_R0_demog_c)
-summary(table1model2_c)
-intervals(table1model2_c)
-```
-
-Linear mixed models of R0 and density + % private transportation + median income (no outliers)
-```{r}
-sample_size<-nobs(table1model3)
-cooksd1 <- CookD(table1model3)
-influential <- as.numeric(names(cooksd1)[(cooksd1 > (4/sample_size))])
-county_R0_demog_c <- county_R0_demog[-influential, ]
-table1model3_c <- lme(R ~ density_log + private + hincome, random = ~ 1| state, data = county_R0_demog_c)
-summary(table1model3_c)
-intervals(table1model3_c)
-```
-
-Linear mixed models of R0 and density + % private transportation + median income + interaction of density and % private transportation (no outliers)
-```{r}
-sample_size<-nobs(table1model4)
-cooksd1 <- CookD(table1model4)
-influential <- as.numeric(names(cooksd1)[(cooksd1 > (4/sample_size))])
-county_R0_demog_c <- county_R0_demog[-influential, ]
-table1model4_c <- lme(R ~ density_log*private + hincome, random = ~ 1| state, data = county_R0_demog_c)
-summary(table1model4_c)
-intervals(table1model4_c)
-```

### 2021-05-20 r0density/R/r0density.R

@@ -45,3 +45,29 @@ NULL
 #' @source US Bureau of Census
 #' @source Sy KTL, White LF, Nichols BE (2021) Population density and basic reproductive numberof COVID-19 across United States counties. PLoSONE 16(4):e0249271
 "county_geo"
+
+#' States of the United States with population.
+#'
+#' Boundaries of the states in the United States.
+#'
+#' @format A simple features data frame with 52 rows and 8 variables:
+#' \describe{
+#'   \item{GEO_ID}{Unique identifier of the state}
+#'   \item{STATE}{Code of state}
+#'   \item{NAME.x}{Short name of state}
+#'   \item{LSAD}{Code of the Legal/Statistical Area Description}
+#'   \item{GEOID}{Unique identifier of the county}
+#'   \item{NAME.y}{Longer name of state}
+#'   \item{population}{Total population from the 2010 Census (variable P001001)}
+#'   \item{geometry}{Geometry of boundaries}
+#' }
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords covid-19
+#' @keywords Census
+#' @keywords United States
+#' @name state_geo
+#' @usage data(state_geo)
+#' @source US Bureau of Census
+"state_geo"

### 2021-05-20 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
First draft of background, started working on tobit models; recreated data package to include a common key for the state boundaries

@@
 library(units) # Package to work with units of measurement
+library(viridis) # Color palettes
 ```
@@
 data(county_geo)
+data(state_geo)
 ```
@@
 # Create a list of neighbors based on the polygons; the default contiguity criterion is Queen:
-county_nb <- poly2nb(county_geo_clean)
+# Counties
+county_nb <- county_geo_clean %>%
+  poly2nb()
+# States
+state_nb <- state_geo %>% 
+  filter(!STATE %in% c("02", "15", "72")) %>%
+  poly2nb()
 ```
@@
                  style="B", 
+                 zero.policy = TRUE)
+
+state_listw <- nb2listw(state_nb, 
+                 style="W", 
                  zero.policy = TRUE)
@@
 
-## Replicating SWN
+# Reproducing SWN
 
@@
 
-```{r R0-map, echo=FALSE, fig.cap="\\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
+```{r R0-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
 # Choropleth map of basic reproductive number
@@
   # Set theme of plot
-  theme_minimal()
+  theme_minimal() +
+  theme(legend.position = "bottom")
 ```
@@
 
-Table \ref{tab:swn-results} replicates Table 1 of SWN (p. 6).
+Table \ref{tab:swn-results} replicates the first three models of SWN (the fourth model did not have any significant variables; see Table 1 in SNW).
+It is possible to verify that the results match with the minor (and irrelevant) exception of the magnitude of the coefficient for travel by private transportation, which is due to a difference in the input (here the variable is one percent units, whereas in SWN it was ten percent units).
+The intercept in these models is a random variable, and the standard deviation is reported in the fourth row of Table \ref{tab:swn-results}.
+It is useful to map the random intercepts: as seen in Figure \ref{fig:random-terms-map}, other things being equal, counties in Texas tended to have somewhat lower values of $R_0$ (i.e., a negative random intercept), whereas counties in South Dakota tended to have higher values of $R_0$.
+The key of the analysis, that population density has a positive association with the basic reproductive number, is verified.
+But is it really?
 
@@
 model4_intervals <- intervals(model4)
+```
+
+```{r join-random-intercepts, include=FALSE}
+state_geo <- state_geo %>% 
+  left_join(model3$coefficients$random$state %>%
+              as.data.frame() %>%
+              rownames_to_column(var = "state"),
+            by = "state") %>%
+  rename(random_intercepts = `(Intercept)`)
 ```
@@
 
-# Some relevant questions
+```{r random-terms-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
+# Choropleth map of basic reproductive number
+ggplot() + 
+  geom_sf(data = state_geo %>%
+            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
+          aes(fill = random_intercepts), # Add the simple features object to the plot
+          color = "lightgray") +
+  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
+  scale_fill_viridis(name = "Random Intercepts",
+                     option = "turbo",
+                       direction = -1) +
+  # Set theme of plot
+  theme_minimal() +
+  theme(legend.position = "bottom")
+```
+
+# Expanding on SWN
+
+Thanks to the availability of code and data, it was possible to verify the results reported by SWN.
+As noted earlier, though, an experienced spatial modeller would have asked themselves about the implications of the spatial sampling procedure used by SWN.
+The decision to use a sample of counties with reliable basic reproductive numbers, although seemingly sensible, results in a non-random sampling scheme.
+Turning our attention back to Figure \ref{fig:R0-map}, we see that many counties without reliable values of $R_0$ are in more rural, less dense parts of the United States.
+The fact that $R_0$ could not be accurately computed in those counties does not mean that there was no transmission of the virus: it simply means that we do not know with precision whether that was the case.
+The low number of cases may be related to low population and/or low population density.
+This is intriguing, to say the least: by excluding cases in a non-random way such as this, we are in effect _truncating_ the sample.
+
+The issue with sample truncation is that parameter estimates can become unreliable.
 
@@
 
-```{r}
-table1model1 <- censReg(R ~ density_log, 
+```{r include=FALSE}
+#
+tobit1 <- censReg(R ~ density_log, 
                         left = 0,
                     data = county_geo_clean)
-summary(table1model1)
+
+#
+tobit2 <- censReg(R ~ density_log + private, 
+                    data = county_geo_clean)
+
+#
+tobit3 <- censReg(R ~ density_log + private + hincome, 
+                    data = county_geo_clean)
 ```
 
-Linear mixed models of R0 and density + % private transportation
-```{r}
-table1model2 <- censReg(R ~ density_log + private, 
-                    data = county_geo_clean)
-summary(table1model2)
+```{r include=FALSE}
+tobit1_me <- tobit1 %>%
+  margEff() %>%
+  summary() %>%
+  as.data.frame()
+
+colnames(tobit1_me) <- c("me", "se", "t", "p")
+
+tobit1_me <- tobit1_me %>%  rownames_to_column(var = "Variable") %>% 
+  transmute(Variable,
+            me,
+            ll = me - 1.96 * se, 
+            ul = me + 1.96 * se)
+
+tobit2_me <- tobit2 %>%
+  margEff() %>%
+  summary() %>%
+  as.data.frame()
+
+colnames(tobit2_me) <- c("me", "se", "t", "p")
+
+tobit2_me <- tobit2_me %>%  rownames_to_column(var = "Variable") %>% 
+  transmute(Variable,
+            me,
+            ll = me - 1.96 * se, 
+            ul = me + 1.96 * se)
+
+tobit3_me <- tobit3 %>%
+  margEff() %>%
+  summary() %>%
+  as.data.frame()
+
+colnames(tobit3_me) <- c("me", "se", "t", "p")
+
+tobit3_me <- tobit3_me %>%  rownames_to_column(var = "Variable") %>% 
+  transmute(Variable,
+            me,
+            ll = me - 1.96 * se, 
+            ul = me + 1.96 * se)
+
+results_me_tobit <- tobit1_me %>%
+  full_join(tobit2_me,
+            by = "Variable") %>%
+  full_join(tobit3_me,
+            by = "Variable")
+
+colnames(results_me_tobit) <- c("Variable",
+                                "b1",
+                                "l1",
+                                "u1",
+                                "b2",
+                                "l2",
+                                "u2",
+                                "b3",
+                                "l3",
+                                "u3")
+
+results_me_tobit <- results_me_tobit %>%
+  transmute(Variable,
+            b1 = ifelse(is.na(b1),
+                        " ", 
+                        round(b1, 3)),
+            ci1 = ifelse(is.na(l1), 
+                         " ",
+                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")),
+            b2 = ifelse(is.na(b2),
+                        " ", 
+                        round(b2, 3)),
+            ci2 = ifelse(is.na(l2), 
+                         " ",
+                         paste0("[", round(l2, 3), ", ", round(u2, 3), "]")),
+            b3 = ifelse(is.na(b3),
+                        " ", 
+                        round(b3, 3)),
+            ci3 = ifelse(is.na(l3), 
+                         " ",
+                         paste0("[", round(l3, 3), ", ", round(u3, 3), "]")))
 ```
 
-Linear mixed models of R0 and density + % private transportation + median income
-```{r}
-table1model3 <- censReg(R ~ density_log + private + hincome, 
-                    data = county_geo_clean)
-summary(table1model3)
-```
-
-Tobit models of R0 and density + % private transportation + median income + interaction of density and % private transportation
-```{r}
-table1model4 <- censReg(R ~ density_log*private + hincome, 
-                    data = county_geo_clean)
-summary(table1model4)
-
+```{r echo=FALSE}
+results_me_tobit %>%
+  mutate(Variable = c("Log of population density",
+                      "Percent of private transportation", 
+                      "Median household income ($10,000)")) %>%
+  kable("latex",
+        digits = 3,
+        booktabs = TRUE,
+        col.names = c("Variable",
+                      "m.e.",
+                      "95% CI",
+                      "m.e.",
+                      "95% CI",
+                      "m.e.",
+                      "95% CI"),
+        caption = "\\label{tab:tobit-results}Marginal effects of sample selection models") %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  add_header_above(c(" ", "Tobit Model 1" = 2, "Tobit Model 2" = 2, "Tobit Model 3" = 2))
 ```
@@
 Fit spatially autoregressive tobit:
-```{r}
+```{r include=FALSE}
 # Fit SAR Tobit

### 2021-05-20 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
First draft of background, started working on tobit models; recreated data package to include a common key for the state boundaries

@@
 \newcommand{\ipsum}{{\bf IPSUM}}
-
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 
@@
 
-\hypertarget{replicating-swn}{%
-\subsection{Replicating SWN}\label{replicating-swn}}
+\hypertarget{reproducing-swn}{%
+\section{Reproducing SWN}\label{reproducing-swn}}
 
@@
 \begin{figure}
-\centering
-\includegraphics{R0-Density-Reanalysis_files/figure-latex/R0-map-1.pdf}
-\caption{\label{fig:R0-map}Basic reproductive rate in US counties
-(Alaska, Hawaii, Puerto Rico, and territories not shown).}
+\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/R0-map-1} \caption{\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown).}\label{fig:R0-map}
 \end{figure}
@@
 
-Table \ref{tab:swn-results} replicates Table 1 of SWN (p.~6).
+Table \ref{tab:swn-results} replicates the first three models of SWN
+(the fourth model did not have any significant variables; see Table 1 in
+SNW). It is possible to verify that the results match with the minor
+(and irrelevant) exception of the magnitude of the coefficient for
+travel by private transportation, which is due to a difference in the
+input (here the variable is one percent units, whereas in SWN it was ten
+percent units). The intercept in these models is a random variable, and
+the standard deviation is reported in the fourth row of Table
+\ref{tab:swn-results}. It is useful to map the random intercepts: as
+seen in Figure \ref{fig:random-terms-map}, other things being equal,
+counties in Texas tended to have somewhat lower values of \(R_0\) (i.e.,
+a negative random intercept), whereas counties in South Dakota tended to
+have higher values of \(R_0\). The key of the analysis, that population
+density has a positive association with the basic reproductive number,
+is verified. But is it really?
 
@@
 
-\hypertarget{some-relevant-questions}{%
-\section{Some relevant questions}\label{some-relevant-questions}}
+\begin{figure}
+\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/random-terms-map-1} \caption{\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown).}\label{fig:random-terms-map}
+\end{figure}
+
+\hypertarget{expanding-on-swn}{%
+\section{Expanding on SWN}\label{expanding-on-swn}}
+
+Thanks to the availability of code and data, it was possible to verify
+the results reported by SWN. As noted earlier, though, an experienced
+spatial modeller would have asked themselves about the implications of
+the spatial sampling procedure used by SWN. The decision to use a sample
+of counties with reliable basic reproductive numbers, although seemingly
+sensible, results in a non-random sampling scheme. Turning our attention
+back to Figure \ref{fig:R0-map}, we see that many counties without
+reliable values of \(R_0\) are in more rural, less dense parts of the
+United States. The fact that \(R_0\) could not be accurately computed in
+those counties does not mean that there was no transmission of the
+virus: it simply means that we do not know with precision whether that
+was the case. The low number of cases may be related to low population
+and/or low population density. This is intriguing, to say the least: by
+excluding cases in a non-random way such as this, we are in effect
+\emph{truncating} the sample.
+
+The issue with sample truncation is that parameter estimates can become
+unreliable.
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{table1model1 }\OtherTok{\textless{}{-}} \FunctionTok{censReg}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log, }
-                        \AttributeTok{left =} \DecValTok{0}\NormalTok{,}
-                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean)}
-\FunctionTok{summary}\NormalTok{(table1model1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 
-## Call:
-## censReg(formula = R ~ density_log, left = 0, data = county_geo_clean)
-## 
-## Observations:
-##          Total  Left-censored     Uncensored Right-censored 
-##           3218           2067           1151              0 
-## 
-## Coefficients:
-##             Estimate Std. error t value Pr(> t)    
-## (Intercept)  2.33967    0.09232   25.34  <2e-16 ***
-## density_log  0.74323    0.02672   27.82  <2e-16 ***
-## logSigma     0.56894    0.02396   23.74  <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Newton-Raphson maximisation, 7 iterations
-## Return code 8: successive function values within relative tolerance limit (reltol)
-## Log-likelihood: -3246.5 on 3 Df
-\end{verbatim}
-
-Linear mixed models of R0 and density + \% private transportation
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{table1model2 }\OtherTok{\textless{}{-}} \FunctionTok{censReg}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private, }
-                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean)}
-\FunctionTok{summary}\NormalTok{(table1model2)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 
-## Call:
-## censReg(formula = R ~ density_log + private, data = county_geo_clean)
-## 
-## Observations:
-##          Total  Left-censored     Uncensored Right-censored 
-##           3218           2067           1151              0 
-## 
-## Coefficients:
-##             Estimate Std. error t value Pr(> t)    
-## (Intercept) 0.825319   0.583014   1.416 0.15689    
-## density_log 0.754516   0.027546  27.391 < 2e-16 ***
-## private     0.017186   0.006548   2.624 0.00868 ** 
-## logSigma    0.570248   0.023977  23.783 < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Newton-Raphson maximisation, 7 iterations
-## Return code 8: successive function values within relative tolerance limit (reltol)
-## Log-likelihood: -3242.96 on 4 Df
-\end{verbatim}
-
-Linear mixed models of R0 and density + \% private transportation +
-median income
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{table1model3 }\OtherTok{\textless{}{-}} \FunctionTok{censReg}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private }\SpecialCharTok{+}\NormalTok{ hincome, }
-                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean)}
-\FunctionTok{summary}\NormalTok{(table1model3)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 
-## Call:
-## censReg(formula = R ~ density_log + private + hincome, data = county_geo_clean)
-## 
-## Observations:
-##          Total  Left-censored     Uncensored Right-censored 
-##           3218           2067           1151              0 
-## 
-## Coefficients:
-##              Estimate Std. error t value  Pr(> t)    
-## (Intercept) -3.201908   0.680988  -4.702 2.58e-06 ***
-## density_log  0.662526   0.026876  24.651  < 2e-16 ***
-## private      0.040886   0.006844   5.974 2.32e-09 ***
-## hincome      0.297763   0.025726  11.574  < 2e-16 ***
-## logSigma     0.531694   0.023920  22.228  < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Newton-Raphson maximisation, 7 iterations
-## Return code 8: successive function values within relative tolerance limit (reltol)
-## Log-likelihood: -3174.96 on 5 Df
-\end{verbatim}
-
-Tobit models of R0 and density + \% private transportation + median
-income + interaction of density and \% private transportation
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{table1model4 }\OtherTok{\textless{}{-}} \FunctionTok{censReg}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log}\SpecialCharTok{*}\NormalTok{private }\SpecialCharTok{+}\NormalTok{ hincome, }
-                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean)}
-\FunctionTok{summary}\NormalTok{(table1model4)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 
-## Call:
-## censReg(formula = R ~ density_log * private + hincome, data = county_geo_clean)
-## 
-## Observations:
-##          Total  Left-censored     Uncensored Right-censored 
-##           3218           2067           1151              0 
-## 
-## Coefficients:
-##                      Estimate Std. error t value  Pr(> t)    
-## (Intercept)         -3.687016   0.733410  -5.027 4.98e-07 ***
-## density_log          0.311841   0.174521   1.787   0.0740 .  
-## private              0.047454   0.007712   6.154 7.58e-10 ***
-## hincome              0.291536   0.025807  11.297  < 2e-16 ***
-## density_log:private  0.004097   0.002028   2.020   0.0434 *  
-## logSigma             0.530095   0.023924  22.158  < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Newton-Raphson maximisation, 7 iterations
-## Return code 8: successive function values within relative tolerance limit (reltol)
-## Log-likelihood: -3173.008 on 6 Df
-\end{verbatim}
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-3}\label{tab:tobit-results}Marginal effects of sample selection models}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllllrl}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Tobit Model 1} & \multicolumn{2}{c}{Tobit Model 2} & \multicolumn{2}{c}{Tobit Model 3} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
+Variable & m.e. & 95\% CI & m.e. & 95\% CI & m.e. & 95\% CI\\
+\midrule
+Log of population density & 0.264 & [0.246, 0.281] & 0.264 & [0.247, 0.281] & 0.233 & [0.216, 0.25]\\
+Percent of private transportation &  &  & 0.006 & [0.002, 0.01] & 0.014 & [0.01, 0.019]\\
+Median household income (\$10,000) &  &  &  &  & 0.105 & [0.087, 0.122]\\
+\bottomrule
+\end{tabular}}
+\end{table}
 
@@
 Fit spatially autoregressive tobit:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Fit SAR Tobit}
-\NormalTok{fit\_sartobit }\OtherTok{\textless{}{-}} \FunctionTok{sartobit}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private }\SpecialCharTok{+}\NormalTok{ hincome\_log,}
-\NormalTok{                         B,}
-                         \AttributeTok{ndraw =} \DecValTok{1000}\NormalTok{,}
-                         \AttributeTok{burn.in =} \DecValTok{200}\NormalTok{, }
-                         \AttributeTok{showProgress =} \ConstantTok{TRUE}\NormalTok{,}
-                         \AttributeTok{data =}\NormalTok{ county\_geo\_clean,}
-                         \AttributeTok{computeMarginalEffects =} \ConstantTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in model.matrix.default(mt, mf, contrasts): non-list contrasts argument
-## ignored
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: Function SE_classic_setup moved to the spatialreg package
-\end{verbatim}
-
-\begin{verbatim}
-## Registered S3 methods overwritten by 'spatialreg':
-##   method                   from 
-##   residuals.stsls          spdep
-##   deviance.stsls           spdep
-##   coef.stsls               spdep
-##   print.stsls              spdep
-##   summary.stsls            spdep
-##   print.summary.stsls      spdep
-##   residuals.gmsar          spdep
-##   deviance.gmsar           spdep
-##   coef.gmsar               spdep
-##   fitted.gmsar             spdep
-##   print.gmsar              spdep
-##   summary.gmsar            spdep
-##   print.summary.gmsar      spdep
-##   print.lagmess            spdep
-##   summary.lagmess          spdep
-##   print.summary.lagmess    spdep
-##   residuals.lagmess        spdep
-##   deviance.lagmess         spdep
-##   coef.lagmess             spdep
-##   fitted.lagmess           spdep
-##   logLik.lagmess           spdep
-##   fitted.SFResult          spdep
-##   print.SFResult           spdep
-##   fitted.ME_res            spdep
-##   print.ME_res             spdep
-##   print.lagImpact          spdep
-##   plot.lagImpact           spdep
-##   summary.lagImpact        spdep
-##   HPDinterval.lagImpact    spdep
-##   print.summary.lagImpact  spdep
-##   print.sarlm              spdep
-##   summary.sarlm            spdep
-##   residuals.sarlm          spdep
-##   deviance.sarlm           spdep
-##   coef.sarlm               spdep
-##   vcov.sarlm               spdep
-##   fitted.sarlm             spdep
-##   logLik.sarlm             spdep
-##   anova.sarlm              spdep
-##   predict.sarlm            spdep
-##   print.summary.sarlm      spdep
-##   print.sarlm.pred         spdep
-##   as.data.frame.sarlm.pred spdep
-##   residuals.spautolm       spdep
-##   deviance.spautolm        spdep
-##   coef.spautolm            spdep
-##   fitted.spautolm          spdep
-##   print.spautolm           spdep
-##   summary.spautolm         spdep
-##   logLik.spautolm          spdep
-##   print.summary.spautolm   spdep
-##   print.WXImpact           spdep
-##   summary.WXImpact         spdep
-##   print.summary.WXImpact   spdep
-##   predict.SLX              spdep
-\end{verbatim}
-
-\begin{verbatim}
-##   |                                                                              |                                                                      |   0%  |                                                                              |                                                                      |   1%  |                                                                              |=                                                                     |   1%  |                                                                              |=                                                                     |   2%  |                                                                              |==                                                                    |   2%  |                                                                              |==                                                                    |   3%  |                                                                              |==                                                                    |   4%  |                                                                              |===                                                                   |   4%  |                                                                              |===                                                                   |   5%  |                                                                              |====                                                                  |   5%  |                                                                              |====                                                                  |   6%  |                                                                              |=====                                                                 |   6%  |                                                                              |=====                                                                 |   7%  |                                                                              |=====                                                                 |   8%  |                                                                              |======                                                                |   8%  |                                                                              |======                                                                |   9%  |                                                                              |=======                                                               |   9%  |                                                                              |=======                                                               |  10%  |                                                                              |=======                                                               |  11%  |                                                                              |========                                                              |  11%  |                                                                              |========                                                              |  12%  |                                                                              |=========                                                             |  12%  |                                                                              |=========                                                             |  13%  |                                                                              |=========                                                             |  14%  |                                                                              |==========                                                            |  14%  |                                                                              |==========                                                            |  15%  |                                                                              |===========                                                           |  15%  |                                                                              |===========                                                           |  16%  |                                                                              |============                                                          |  16%  |                                                                              |============                                                          |  17%  |                                                                              |============                                                          |  18%  |                                                                              |=============                                                         |  18%  |                                                                              |=============                                                         |  19%  |                                                                              |==============                                                        |  19%  |                                                                              |==============                                                        |  20%  |                                                                              |==============                                                        |  21%  |                                                                              |===============                                                       |  21%  |                                                                              |===============                                                       |  22%  |                                                                              |================                                                      |  22%  |                                                                              |================                                                      |  23%  |                                                                              |================                                                      |  24%  |                                                                              |=================                                                     |  24%  |                                                                              |=================                                                     |  25%  |                                                                              |==================                                                    |  25%  |                                                                              |==================                                                    |  26%  |                                                                              |===================                                                   |  26%  |                                                                              |===================                                                   |  27%  |                                                                              |===================                                                   |  28%  |                                                                              |====================                                                  |  28%  |                                                                              |====================                                                  |  29%  |                                                                              |=====================                                                 |  29%  |                                                                              |=====================                                                 |  30%  |                                                                              |=====================                                                 |  31%  |                                                                              |======================                                                |  31%  |                                                                              |======================                                                |  32%  |                                                                              |=======================                                               |  32%  |                                                                              |=======================                                               |  33%  |                                                                              |=======================                                               |  34%  |                                                                              |========================                                              |  34%  |                                                                              |========================                                              |  35%  |                                                                              |=========================                                             |  35%  |                                                                              |=========================                                             |  36%  |                                                                              |==========================                                            |  36%  |                                                                              |==========================                                            |  37%  |                                                                              |==========================                                            |  38%  |                                                                              |===========================                                           |  38%  |                                                                              |===========================                                           |  39%  |                                                                              |============================                                          |  39%  |                                                                              |============================                                          |  40%  |                                                                              |============================                                          |  41%  |                                                                              |=============================                                         |  41%  |                                                                              |=============================                                         |  42%  |                                                                              |==============================                                        |  42%  |                                                                              |==============================                                        |  43%  |                                                                              |==============================                                        |  44%  |                                                                              |===============================                                       |  44%  |                                                                              |===============================                                       |  45%  |                                                                              |================================                                      |  45%  |                                                                              |================================                                      |  46%  |                                                                              |=================================                                     |  46%  |                                                                              |=================================                                     |  47%  |                                                                              |=================================                                     |  48%  |                                                                              |==================================                                    |  48%  |                                                                              |==================================                                    |  49%  |                                                                              |===================================                                   |  49%  |                                                                              |===================================                                   |  50%  |                                                                              |===================================                                   |  51%  |                                                                              |====================================                                  |  51%  |                                                                              |====================================                                  |  52%  |                                                                              |=====================================                                 |  52%  |                                                                              |=====================================                                 |  53%  |                                                                              |=====================================                                 |  54%  |                                                                              |======================================                                |  54%  |                                                                              |======================================                                |  55%  |                                                                              |=======================================                               |  55%  |                                                                              |=======================================                               |  56%  |                                                                              |========================================                              |  56%  |                                                                              |========================================                              |  57%  |                                                                              |========================================                              |  58%  |                                                                              |=========================================                             |  58%  |                                                                              |=========================================                             |  59%  |                                                                              |==========================================                            |  59%  |                                                                              |==========================================                            |  60%  |                                                                              |==========================================                            |  61%  |                                                                              |===========================================                           |  61%  |                                                                              |===========================================                           |  62%  |                                                                              |============================================                          |  62%  |                                                                              |============================================                          |  63%  |                                                                              |============================================                          |  64%  |                                                                              |=============================================                         |  64%  |                                                                              |=============================================                         |  65%  |                                                                              |==============================================                        |  65%  |                                                                              |==============================================                        |  66%  |                                                                              |===============================================                       |  66%  |                                                                              |===============================================                       |  67%  |                                                                              |===============================================                       |  68%  |                                                                              |================================================                      |  68%  |                                                                              |================================================                      |  69%  |                                                                              |=================================================                     |  69%  |                                                                              |=================================================                     |  70%  |                                                                              |=================================================                     |  71%  |                                                                              |==================================================                    |  71%  |                                                                              |==================================================                    |  72%  |                                                                              |===================================================                   |  72%  |                                                                              |===================================================                   |  73%  |                                                                              |===================================================                   |  74%  |                                                                              |====================================================                  |  74%  |                                                                              |====================================================                  |  75%  |                                                                              |=====================================================                 |  75%  |                                                                              |=====================================================                 |  76%  |                                                                              |======================================================                |  76%  |                                                                              |======================================================                |  77%  |                                                                              |======================================================                |  78%  |                                                                              |=======================================================               |  78%  |                                                                              |=======================================================               |  79%  |                                                                              |========================================================              |  79%  |                                                                              |========================================================              |  80%  |                                                                              |========================================================              |  81%  |                                                                              |=========================================================             |  81%  |                                                                              |=========================================================             |  82%  |                                                                              |==========================================================            |  82%  |                                                                              |==========================================================            |  83%  |                                                                              |==========================================================            |  84%  |                                                                              |===========================================================           |  84%  |                                                                              |===========================================================           |  85%  |                                                                              |============================================================          |  85%  |                                                                              |============================================================          |  86%  |                                                                              |=============================================================         |  86%  |                                                                              |=============================================================         |  87%  |                                                                              |=============================================================         |  88%  |                                                                              |==============================================================        |  88%  |                                                                              |==============================================================        |  89%  |                                                                              |===============================================================       |  89%  |                                                                              |===============================================================       |  90%  |                                                                              |===============================================================       |  91%  |                                                                              |================================================================      |  91%  |                                                                              |================================================================      |  92%  |                                                                              |=================================================================     |  92%  |                                                                              |=================================================================     |  93%  |                                                                              |=================================================================     |  94%  |                                                                              |==================================================================    |  94%  |                                                                              |==================================================================    |  95%  |                                                                              |===================================================================   |  95%  |                                                                              |===================================================================   |  96%  |                                                                              |====================================================================  |  96%  |                                                                              |====================================================================  |  97%  |                                                                              |====================================================================  |  98%  |                                                                              |===================================================================== |  98%  |                                                                              |===================================================================== |  99%  |                                                                              |======================================================================|  99%  |                                                                              |======================================================================| 100%
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{summary}\NormalTok{(fit\_sartobit)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## ----MCMC spatial autoregressive Tobit model ----
-## Execution time  = 41.022 secs
-## 
-## N draws         =   1000, N omit (burn-in)=    200
-## N observations  =   3218, K covariates    =      4
-## # censored values =   2067, # observed values =   1151
-## Min rho         = -1.000, Max rho         =  1.000
-## --------------------------------------------------
-## 
-##              Estimate  Std. Dev   p-level t-value Pr(>|z|)    
-## (Intercept) -4.399260  0.737342  0.000000  -5.966 2.69e-09 ***
-## density_log  0.562608  0.026714  0.000000  21.061  < 2e-16 ***
-## private      0.039826  0.006874  0.000000   5.794 7.54e-09 ***
-## hincome_log  1.572387  0.138627  0.000000  11.343  < 2e-16 ***
-## sige         2.992428  0.173624  0.000000  17.235  < 2e-16 ***
-## rho          0.056389  0.004937  0.000000  11.421  < 2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{impacts}\NormalTok{(fit\_sartobit)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## --------Marginal Effects--------
-## 
-## (a) Direct effects
-##             lower_005 posterior_mean upper_095
-## density_log   0.21275        0.23102     0.248
-## private       0.01180        0.01635     0.021
-## hincome_log   0.55351        0.64566     0.744
-## 
-## (b) Indirect effects
-##             lower_005 posterior_mean upper_095
-## density_log  -0.29273       -0.27252    -0.251
-## private      -0.02504       -0.01929    -0.014
-## hincome_log  -0.88010       -0.76178    -0.651
-## 
-## (c) Total effects
-##             lower_005 posterior_mean upper_095
-## density_log -0.044493      -0.041495    -0.038
-## private     -0.003850      -0.002942    -0.002
-## hincome_log -0.136033      -0.116125    -0.097
-\end{verbatim}
 

### 2021-05-20 r0density/raw-data-processing/00-Data-Preprocessing.Rmd
First draft of background, started working on tobit models; recreated data package to include a common key for the state boundaries

@@
 
-Prepare state boundaries for saving
+Prepare state boundaries for saving:
 ```{r}
+# Get state number code and two letter code from county table:
+state_key <- county_geo %>% 
+  st_drop_geometry() %>%
+  drop_na(state) %>% 
+  distinct(STATE, 
+           .keep_all = TRUE) %>% 
+  select(STATE, state)
+
+# Join state two letter code to table:
+state_geo <- state_geo %>%
+  left_join(state_key,
+            by = "STATE")
+
 state_geo <- state_geo %>%

### 2021-05-20 r0density/R/r0density.R

@@ -59,6 +59,7 @@ NULL
 #'   \item{GEOID}{Unique identifier of the county}
 #'   \item{NAME.y}{Longer name of state}
 #'   \item{population}{Total population from the 2010 Census (variable P001001)}
+#'   \item{state}{Two letter code of state (from Sy et al.)}
 #'   \item{geometry}{Geometry of boundaries}
 #' }
 #'

### 2021-05-21 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Revised data package to include boundaries of urbanized areas

@@
 As noted earlier, though, an experienced spatial modeller would have asked themselves about the implications of the spatial sampling procedure used by SWN.
-The decision to use a sample of counties with reliable basic reproductive numbers, although seemingly sensible, results in a non-random sampling scheme.
-Turning our attention back to Figure \ref{fig:R0-map}, we see that many counties without reliable values of $R_0$ are in more rural, less dense parts of the United States.
+The decision to use a sample of counties with reliable basic reproductive numbers, although seemingly sensible, results in a non-random spatial sampling scheme.
+Turning our attention back to Figure \ref{fig:R0-map}, we form the impression that many counties without reliable values of $R_0$ are in more rural, less dense parts of the United States: this is .
 The fact that $R_0$ could not be accurately computed in those counties does not mean that there was no transmission of the virus: it simply means that we do not know with precision whether that was the case.

### 2021-05-21 r0density/raw-data-processing/00-Data-Preprocessing.Rmd
Revised data package to include boundaries of urbanized areas

@@
 
+Retrieve metropolitan/micropolitan boundaries for plotting:
+```{r}
+urban_geo <- get_acs(geography = "urban area", 
+                            variables = "B01003_001", # Total population
+                            year = 2019, 
+                            geometry = TRUE, 
+                            keep_geo_vars = TRUE) %>%
+  rename(population = estimate) %>%
+  select(-c(variable, moe))
+```
+
 Now join Sy's data to the counties with geometry and other socio-economic variables (the equivalent variables are dropped from Sy et al.'s table, because they are missing all the counties without a basic reproductive number):
@@
 
+Prepare urban boundaries for saving:
+```{r}
+# Get state number code and two letter code from state table:
+urban_geo <- urban_geo %>%
+  st_join(state_geo %>%
+            select(STATE, state))
+```
+
 Save data table:
@@
 save(county_geo, file = "county_geo.RData", compress = "xz")
+save(urban_geo, file = "urban_geo.RData", compress = "xz")
 save(state_geo, file = "state_geo.RData", compress = "xz")

### 2021-05-21 r0density.Rcheck/00_pkg_src/r0density/R/r0density.R

@@ -1,6 +1,6 @@
-#' vaccHamilton
+#' r0density
 #'
-#' vaccHamilton is a companion package for the paper "Population Density and Basic Reproductive Number of COVID-19 Revisited".
+#' r0density is a companion package for the paper "The importance of reproducibility in COVID-19 research: the case of population density and the basic reproductive number of COVID-19".
 #'
 #' The package includes the data used in the analysis.
 #'
@@ -45,3 +45,62 @@ NULL
 #' @source US Bureau of Census
 #' @source Sy KTL, White LF, Nichols BE (2021) Population density and basic reproductive numberof COVID-19 across United States counties. PLoSONE 16(4):e0249271
 "county_geo"
+
+#' States of the United States with population.
+#'
+#' Boundaries of the states in the United States.
+#'
+#' @format A simple features data frame with 52 rows and 8 variables:
+#' \describe{
+#'   \item{GEO_ID}{Unique identifier of the state}
+#'   \item{STATE}{Code of state}
+#'   \item{NAME.x}{Short name of state}
+#'   \item{LSAD}{Code of the Legal/Statistical Area Description}
+#'   \item{GEOID}{Unique identifier of the county}
+#'   \item{NAME.y}{Longer name of state}
+#'   \item{population}{Total population from the 2010 Census (variable P001001)}
+#'   \item{state}{Two letter code of state (from Sy et al.)}
+#'   \item{geometry}{Geometry of boundaries}
+#' }
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords covid-19
+#' @keywords Census
+#' @keywords United States
+#' @name state_geo
+#' @usage data(state_geo)
+#' @source US Bureau of Census
+"state_geo"
+
+#' Urban areas of the United States with population.
+#'
+#' Boundaries of urban areas in the United States.
+#'
+#' @format A simple features data frame with 3,779 rows and 13 variables:
+#' \describe{
+#'   \item{UACE10}{2010 Census urban area code}
+#'   \item{AFFGEOID10}{Long geographic identifier}
+#'   \item{GEOID10}{Geographic identifier}
+#'   \item{NAME10}{Short name of area}
+#'   \item{LSAD10}{Code of the Legal/Statistical Area Description}
[…]

### 2021-05-21 r0density/R/r0density.R

@@ -72,3 +72,35 @@ NULL
 #' @usage data(state_geo)
 #' @source US Bureau of Census
 "state_geo"
+
+#' Urban areas of the United States with population.
+#'
+#' Boundaries of urban areas in the United States.
+#'
+#' @format A simple features data frame with 3,779 rows and 13 variables:
+#' \describe{
+#'   \item{UACE10}{2010 Census urban area code}
+#'   \item{AFFGEOID10}{Long geographic identifier}
+#'   \item{GEOID10}{Geographic identifier}
+#'   \item{NAME10}{Short name of area}
+#'   \item{LSAD10}{Code of the Legal/Statistical Area Description}
+#'   \item{UATYP10}{2010 Census urban area type: C: Cluster with more than 2,500 but less than 50,000 population; U: Urbanized area with more than 50,000 population}
+#'   \item{ALAND10}{2010 Census land area}
+#'   \item{AWATER10}{2010 Census water area}
+#'   \item{NAME}{Name of area}
+#'   \item{population}{Total population from the 2010 Census (variable P001001)}
+#'   \item{STATE}{Code of state}
+#'   \item{state}{Two letter code of state (from Sy et al.)}
+#'   \item{geometry}{Geometry of boundaries}
+#' }
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords covid-19
+#' @keywords Census
+#' @keywords United States
+#' @name urban_geo
+#' @usage data(urban_geo)
+#' @source US Bureau of Census
+"urban_geo"
+

### 2021-05-21 R0-Density-Reanalysis/Cover-Letter-v1.Rmd
First version for submission to PLoS ONE

---
title: "Cover Letter"
output:
  pdf_document: default
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

To the Editor:

Please find attached the paper titled "The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic". The paper illustrates the importance of reproducibility in research in the particular context of COVID-19. It takes a recent paper from the literature (Sy et al., 2021, published in PLoS ONE) and demonstrates how openly sharing data and code can accelerate research.

This paper makes the following contributions to the literature:

1. It reviews the emerging literature on the relationship between population density and the spread of COVID-19.

2. It shows that reproducibility in this literature remains an elusive goal.

3. It reanalyzes the data of Sy et al., and demonstrates that a different technical approach can lead to substantially different conclusions from those reached by the original analysis of Sy et al.

4. The reanalysis and reappraisal of the conclusions in a matter of weeks demonstrates the importance of reproducible research.

4. Finally, the paper provides an example of a reproducible research project, with extensively documented workflows, from raw data processing, to modelling, and post-modelling outputs. The source document is an R Markdown file, a self-contained document with the text and _all_ the analysis. These materials are publicly available in a repository, as indicated in the paper.

I trust that you will find that the paper is technically correct, and suitable for publication.

Sincerely,

Antonio Paez

### 2021-05-21 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
First version for submission to PLoS ONE

--- 2021-05-20 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
@@
 library(nlme) # Used to estimate 
+library(patchwork)
 library(r0density) # Data package
+library(sampleSelection) # Package to estimate sampling selection models
 library(sf) # Package to work with spatial information in simple features format
@@
 data(county_geo)
+data(state_geo)
+data(urba_geo)
 ```
@@
   st_as_sf()
+
+# Boundary of continental US excluding Alaska
+us_geo <- st_union(state_geo %>% filter(!STATE %in% c("02", "15", "72")))
 ```
@@
 
+state_listw <- nb2listw(state_nb, 
+                        style="W", 
+                        zero.policy = TRUE)
+
 # Convert listw to sparse matrix for use in spatial tobit:
@@
 
-The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the global pandemic that followed in its wake, led to an explosive growth of research.
+The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the global pandemic that followed in its wake, led to an explosive growth of research around the globe.
 According to Fraser et al.
@@
 For example, are lower density regions safer from the pandemic?
-Are de-densification policies warranted, at least in the short term?
+Are de-densification policies warranted, even if just in the short term?
 And in the longer term, will the risks of life in high density regions presage a flight from cities?
@@
 
-# Background
+# Background: the intuitive relationship between density and spread of contagious diseases
 
@@
 Mathematically, models of exposure and contagion indicate that higher densities can catalyze the transmission of contagious diseases [@Rocklov2020high; @Li2018effect].
-Models such as these were likely at the root of messages, by some figures in positions of authority, that regions with sparse population densities faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic down without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].
+The idea is intuitive and likely at the root of messages, by some figures in positions of authority, that regions with sparse population densities faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic down without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].
 
@@
 There are solid reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers-however, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place.
-Nonetheless, the idea that high density correlates with high transmission is so intuitive that it is often taken for granted even at larger scales [e.g., @Cruz2020exploring; @Micallef2020first].
+Nonetheless, the idea that high density correlates with high transmission is so seemingly reasonable that it is often taken for granted even at larger scales [e.g., @Cruz2020exploring; @Micallef2020first].
 At larger scales, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, e.g., @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.
@@
 
-In this respect, the literature to date remains inconclusive.
+# Background: but what does the literature say?
+
+When it comes to population density and the spread of COVID-19, the international literature to date remains inconclusive.
 
@@
 This is why reproducibility is important.
-To pick an example (which will be further elaborated in the following sections), the study of Sy et al.
-[-@Sy2021population; hereafter SWN] would immediately grab the attention of a researcher with expertise in spatial modelling.
+To pick an example (which will be further elaborated in later sections of this paper), the study of Sy et al.
+[-@Sy2021population; hereafter SWN] would immediately grab the attention of a researcher with a somewhat different toolbox.
+
+# Reproducibility of research
 
 SWN investigated the basic reproductive number of COVID-19 in US counties, and its association with population density, median household income, and prevalence of private mobility.
-The basic reproductive number can only be calculated reliably with a minimum number of cases, and a large number of counties did not meet such threshold.
-As researchers do, SWN made certain seemingly reasonable modelling decision, for instance basing their analysis only on counties with valid observations.
-An experiences spatial modeler would likely ask some of the following questions on reading SWN's paper: how were missing counties treated?
+For their multivariate analysis, SWN used mixed linear models.
+This is a reasonable modelling choice: $R_0$ is an interval-ratio variable that is suitably modeled using linear regression; further, as SWN note there is a likelihood that the process in not independent "among counties within each state, potentially due to variable resource allocation and differing health systems across states"  (p.
+3).
+A mixed linear model accounts for this by introducing random components (in the case of SWN, random intercepts at the state level).
+SWN estimated various models with different combinations of variables, including median household income and prevalence of travel by private transportation.
+These are sensible controls, given potential variations in behavior: people in more affluent counties may have greater opportunities to work from home, and use of private transportation reduces contact with strangers.
+Moreover, they also conducted various sensitivity analyses.
+After these efforts, SWN conclude that there is a positive association between the basic reproductive number and population density at the level of counties in the US.
+
+One salient aspect of the analysis in SWN is that the basic reproductive number can only be calculated reliably with a minimum number of cases, and a large number of counties did not meet such threshold.
+As researchers do, SWN made modelling decisions, in this case basing their analysis only on counties with valid observations.
+A modeler with expertise in different methods would likely ask some of the following questions on reading SWN's paper: how were missing counties treated?
+What are the implications of the spatial sampling framework used in the analysis?
 Is it possible to spatially interpolate the missing observations?
-What are the implications of the spatial sampling framework used in the analysis?
 And, was there evidence of spatial autocorrelation in the residuals of the models?
-These are questions that in most cases would not occur to a researcher who has not been exposed to spatial statistics or spatial econometrics.
-Nonetheless, they are relevant and important.
-Fortunately, SWN give an example of a reasonably open, reproducible research product: their paper is accompanied by (most of) the data and (most of) the code used in the analysis.
-This means that an independent expert can, with only a moderate investment of time and effort, replicate the results in the paper, as well as ask additional questions.
+These questions are relevant and their implications important.
+Fortunately, SWN are an example of a reasonably open, reproducible research product: their paper is accompanied by (most of) the data and (most of) the code used in the analysis.
+This means that an independent expert can, with only a moderate investment of time and effort, reproduce the results in the paper, as well as ask additional questions.
 
@@
 
-In the following sections, the analysis of RWN is replicated, some relevant questions from the perspective of a spatial modeler are asked, and the data are reanalyzed.
+In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of a spatial modeler are asked, and the data are reanalyzed.
 
-## Replicating SWN
+# Reproducing SWN
 
@@
 Reliable calculation of $R_0$ requires a minimum number of cases to be able to assume that there is community transmission of the pathogen.
-Accordingly, SNW based their analysis only on counties that had at least 25 cases or more at the end of the exponential growth phase (see Fig.
+Accordingly, SWN based their analysis only on counties that had at least 25 cases or more at the end of the exponential growth phase (see Fig.
 \ref{fig:R0-map}).
@@
   # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
-  scale_fill_viridis_c(name = "R0",
+  scale_fill_viridis_c(name = expression(R[0]),
                        direction = -1) +
@@
           color = "lightgray") +
+  # Add boundary of continental US
+  geom_sf(data = us_geo,
+          fill = NA) + 
   # Add caption
@@
   # Set theme of plot
-  theme_minimal()
+  theme_minimal() +
+  theme(legend.position = "bottom")
 ```
 
-SWN used for their multivariate analysis mixed linear models.
-This is a reasonable modelling choice: $R_0$ is an interval-ratio variable that is suitably modeled using linear regression; further, as SWN note there is a likelihood that the process in not independent "among counties within each state, potentially due to variable resource allocation and differing health systems across states"  (p.
-3).
-A mixed linear model accounts for this by introducing random components (in the case of SNW, random intercepts at the state level).
-SWN estimated various models with different combinations of variables, including median household income and prevalence of travel by private transportation.
-These are sensible controls, given potential variations in behavior: people in more affluent counties may have greater opportunities to work from home, and use of private transportation reduces contact with strangers.
-Moreover, they also conducted various sensitivity analyses.
-After these efforts, SNW conclude that there is a positive association between the basic reproductive number and population density at the level of counties in the US.
+Table \ref{tab:swn-results} reproduces the first three models of SWN (the fourth model did not have any significant variables; see Table 1 in SWN).
+It is possible to verify that the results match, with only the minor (and irrelevant) exception of the magnitude of the coefficient for travel by private transportation, which is due to a difference in the input (here the variable is one percent units, whereas in SWN it was ten percent units).
+The mixed linear model gives random intercepts (i.e., the intercept is a random variable), and the standard deviation is reported in the fourth row of Table \ref{tab:swn-results}.
+It is useful to map the random intercepts: as seen in Figure \ref{fig:random-terms-map}, other things being equal, counties in Texas tend to have somewhat lower values of $R_0$ (i.e., a negative random intercept), whereas counties in South Dakota tend to have higher values of $R_0$.
+The key of the analysis, after extensive sensitivity analysis, is a robust finding that population density has a positive association with the basic reproductive number.
+But does it?
 
-Table \ref{tab:swn-results} replicates Table 1 of SWN (p. 6).
-
-```{r replicate-swn, include=FALSE}
+```{r reproduce-swn, include=FALSE}
 # Fit (mixed) linear models as in Sy et al. These models are reported in table 1 of their paper
@@
                       "95% CI"),
-        caption = "\\label{tab:swn-results}Replicating SWN: Models 1-3") %>%
+        caption = "\\label{tab:swn-results}Reproducing SWN: Models 1-3") %>%
   kable_styling(latex_options = c("scale_down")) %>%
@@
 
-# Some relevant questions
+```{r random-terms-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
+# Choropleth map of basic reproductive number
+ggplot() + 
+  geom_sf(data = state_geo %>%
+            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
+          aes(fill = random_intercepts), # Add the simple features object to the plot
+          color = "lightgray") +
+  # Add boundary of continental US
+  geom_sf(data = us_geo,
+          fill = NA) + 
+  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
+  scale_fill_viridis(name = "Random Intercepts",
+                     option = "turbo",
+                     direction = -1) +
+  # Set theme of plot
+  theme_minimal() +
+  theme(legend.position = "bottom")
+```
 
-Alas, despite decades' worth of developments in the field of geographical analysis, not all research presented to date has used proper methods for the study of COVID-19.
-While the answer to the question "do spatial effects really matter in regression analysis" was definitively answered in the positive at least 40 years ago, in practice many researchers continue to ignore the pitfalls of ignoring them.
-In this paper, I present a reanalysis of the data used by Sy et al.
-(2021) to study the correlations between the basic reproductive number of COVID-19 and population density in counties.
-I highlight two related issues: non-systematic sampling in space and spatial autocorrelation.
-The reanalysis is based on the use of tobit models to account for non-systematic sampling, and spatially autoregressive tobit models to account for spatial autocorrelation in the data generation process.
-The reanalysis highlights the importance of openness and reproducibility in COVID-19 research.
-Finally, the results provide a sobering example of the risks of not using appropriate methods in the analysis of geographical data.
-Furthermore,
+# Expanding on SWN
+
+The preceding section shows that thanks to the availability of code and data, it is possible to verify the results reported by SWN.
+As noted earlier, though, an independent researcher might have wondered about the implications of the spatial sampling procedure used by SWN.
+The decision to use a sample of counties with reliable basic reproductive numbers, although apparently sensible, results in a non-random spatial sampling scheme.
+Turning our attention back to Figure \ref{fig:R0-map}, we form the impression that many counties without reliable values of $R_0$ are in more rural, less dense parts of the United States.
+This impression is reinforced when we overlay the boundaries of urban areas with population greater than 50,000 on the counties with valid values of $R_0$ (see Figure \ref{fig:urban-areas-map}).
+The fact that $R_0$ could not be accurately computed in many counties without large urban areas does not mean that there was no transmission of the virus: it simply means that we do not know with precision whether that was the case.
+The low number of cases may be related to low population and/or low population density.
+This is intriguing, to say the least: by excluding cases based on the ability to calculate $R_0$ we are potentially _censoring_ the sample in a non-random way.
+
+```{r urban-areas-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
+# Choropleth map of basic reproductive number
+ggplot() + # Create a ggplot
+  # Choropleth map of basic reproductive number
+  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
+          aes(fill = R),
+          color = NA) +
+  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
+  scale_fill_viridis_c(name = "R0",
+                       direction = -1) +
+  # Overlay counties with missing values of the reproductive number in white
+  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
+                   R == 0), 
+          fill = "white",
+          color = NA) +
+  # Overlay boundaries of urban areas
+  geom_sf(data = urban_geo %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
+                   UATYP10 == "U"), # Include only urbanized areas with population > 50,000
+          fill = NA,
+          color = "red") +
+  # Add boundary of continental US
+  geom_sf(data = us_geo,
+          fill = NA) + 
+  # Add caption
+  labs(caption = "Note: boundaris of urbanized areas with population > 50,000 are shown in red") +
+  # Set theme of plot
+  theme_minimal() +
+  theme(legend.position = "bottom")
+```
+
+A problematic issue with non-random sample selection is that parameter estimates can become unreliable, and numerous techniques have been developed over time to address this.
+A model useful for sample selection problems is Heckman's selection model [see @Maddala1983limited].
+The selection model is in fact a system of two equations, as follows:
+
+$$
+\begin{array}{c}
+y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
+y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
+\end{array}
+$$
+\noindent where $y_i^{S*}$ is a latent variable for the sample selection process, and $y_i^{O*}$ is the latent outcome.
+Vectors $x_i^S$ and $x_i^O$ are explanatory variables (with the possibility that $x_i^S = x_i^S$).
+Both equations include random terms (i.e., $\epsilon_i^S$ and $\epsilon_i^O$)  The first equation is designed to model the _probability_ of sampling, and the second equation the outcome of interest (say $R_0$).
+The random terms are jointly distributed and correlated with parameter $\rho$.
+
+What the analyst observes is the following:
+$$
+y_i^S =
+\begin{cases}
+0 & \text{if } y_i^{S*} < 0\\
+1 & \text{otherwise}
+\end{cases}
+$$
+\noindent and:
+$$
+y_i^O =
+\begin{cases}
+0 & \text{if } y_i^{S} = 0\\
+y_i^{O*} & \text{otherwise}
+\end{cases}
+$$
+
+In other words, the outcome of interest is observed _only_ for certain cases ($y_i^S=1$, i.e., for sampled observations).
+The probability of sampling depends on $x_i^S$.
+For the cases observed, the outcome $y_i^O$ depends on $x_i^O$.
+
+A sample selection model is estimated using the same selection of variables as SWN Model 3.
+This is Sample Selection Model 1 in Table \ref{tab:selection-results}.
+The first thing to notice about this model is that the sample selection process and the outcome are not independent ($\rho\ne0$ with 5% of confidence).
+The selection equation indicates that the probability of a county to be in the sample increases with population density (but at a decreasing rate due to the log-transformation), when travel by private modes is more prevalent, and as median household income in the county is higher.
+This is in line with the impression left by Figure \ref{fig:urban-areas-map} that counties with reliable values of $R_0$ tended to be those with larger urban centers.
+Once that the selection probabilities are accounted for in the model, several things happen with the outcomes model.
+First, the coefficient for population density is still positive, but the magnitude changes: in effect, it appears that the effect of density is more pronounced than what SWN Model 3 indicated.
+The coefficient for percent of private transportation changes signs.
+And the coefficient for median household income is now significant.
+
+The second model in Table \ref{tab:selection-results} (Selection Model 2) changes the way the variables are entered into the model.
+The log-transformation of density in SWN and Selection Model 1 assumes that the association between density and $R_0$ is monotonically increasing (if the sign of the coefficient is positive) or decreasing (if the sign of the coefficient is negative).
+There are some indications that the relationship may actually not be monotonical.
+For example, Paez et al.
+[-@Paez2020spatio] found a positive (if non-significant) relationship between density and incidence of COVID-19 in the provinces of Spain at the beginning of the pandemic.
+This changed to a negative (and significant) relationship during the lockdown.
+In the case of the US, Fielding-Miller et al.
+[-@Fielding2020social] found that the association between COVID-19 deaths and population density was positive in rural counties, but negative in urban counties.
+A variable transformation that allows for non-monotonic changes in the relationship is the square of the density.
+
+As seen in the table, Selection Model 2 replaces the log-transformation of population density with a quadratic expansion.
+The results of this analysis indicate that with this variable transformation, the selection and outcome processes are still not independent ($\rho\ne0$ with 5% of confidence).
+But a few interesting things emerge.
+When we examine the outcomes model, we see that the quadratic expansion has a positive coefficient for the first order term, but a negative coefficient for the second order term.
+This indicates that $R_0$ initially tends to increase with higher density, but only up to a point, after which the negative second term (which grows more rapidly due to the square), becomes increasingly dominant.
+Secondly, the sign of the coefficient for travel by private transportation becomes negative again.
+This, of course, makes more sense than the positive sign of Selection Model 1: if people tend to travel in private transportation, the potential for contact should be lower instead of higher.
+And finally median household income is no longer significant.
+
+```{r selection-variable, include=FALSE}
+# Create selection variable
+county_geo_clean <- county_geo_clean %>%
+  mutate(Ys = ifelse(R>0, TRUE, FALSE))
+```
+
+```{r estimate-selection-models, include=FALSE}
+# Selection model using SWN variables
+selection1 <- selection(Ys ~ density_log + private + hincome, R ~ density_log + private + hincome, data = county_geo_clean)
+#summary(selection1)
+
+# Selection model with different variable specification
+selection2 <- selection(Ys ~ density + density_2 + private + hincome, R ~ density + density_2 + private + hincome, data = county_geo_clean)
+#summary(selection2)
+
+# Checking SWN with different variable specification
+model3.2 <- lme(R ~ density + density_2 + private + hincome, 
+                random = ~ 1| state, 
+                data = county_geo_clean %>%
+                  filter(R > 0))
+#summary(model3.2)
+```
+
+```{r prepare-selection-results, include=FALSE}
+selection1_me <- selection1 %>%
+  summary()
+
+selection1_me <- selection1_me$estimate %>%
+  as.data.frame()
+
+colnames(selection1_me) <- c("b", "se", "t", "p")
+
+selection1_me <- selection1_me %>%  
+  rownames_to_column(var = "Variable") %>% 
+  transmute(Variable,
+            b,
+            ll = b - 1.96 * se, 
+            ul = b + 1.96 * se)
+
+selection2_me <- selection2 %>%
+  summary()
+
+selection2_me <- selection2_me$estimate %>%
+  as.data.frame()
+
+colnames(selection2_me) <- c("b", "se", "t", "p")
+
+selection2_me <- selection2_me %>%  
+  rownames_to_column(var = "Variable") %>% 
+  transmute(Variable,
+            b,
+            ll = b - 1.96 * se, 
+            ul = b + 1.96 * se)
+
+results_selection <- selection1_me %>%
+  full_join(selection2_me,
+            by = "Variable")
+
+colnames(results_selection) <- c("Variable",
+                                 "b1",
+                                 "l1",
+                                 "u1",
+                                 "b2",
+                                 "l2",
+                                 "u2")
+
+results_selection <- results_selection %>%
+  transmute(Order = c(1, 2, 5, 6, 7, 8, 11, 12, 13, 14, 3, 4, 9, 10),
+            Variable,
+            b1 = ifelse(is.na(b1),
+                        " ", 
+                        round(b1, 3)),
+            ci1 = ifelse(is.na(l1), 
+                         " ",
+                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")),
+            b2 = ifelse(is.na(b2),
+                        " ", 
+                        round(b2, 3)),
+            ci2 = ifelse(is.na(l2), 
+                         " ",
+                         paste0("[", round(l2, 3), ", ", round(u2, 3), "]"))) %>%
+  arrange(Order) %>%
+  select(-Order)
+```
+
+```{r tabulate-sample-selection-results, echo=FALSE}
+results_selection  %>%
+  mutate(Variable = c("Intercept",
+                      "Log of population density",
+                      "Density (1,000 per sq.km)",
+                      "Density squared",
+                      "Percent of private transportation", 
+                      "Median household income (10,000)",
+                      "Intercept",
+                      "Log of population density",
+                      "Density (1,000 per sq.km)",
+                      "Density squared",
+                      "Percent of private transportation", 
+                      "Median household income (\\$10,000)",
+                      "$\\sigma$",
+                      "$\\rho$")) %>%
+  kable("latex",
+        digits = 3,
+        booktabs = TRUE,
+        escape = FALSE,
+        col.names = c("Variable",
+                      "$\\beta$", 
+                      "95\\% CI",
+                      "$\\beta$",
+                      "95\\% CI"),
+        align = c("lcccc"),
+        caption = "\\label{tab:selection-results}Estimation results of sample selection models") %>%
+  kable_styling(latex_options = c("scale_down")) %>% 
+  pack_rows(group_label = "Sample Selection Model", 1, 6) %>%
+  pack_rows(group_label = "Outcome Model", 7, 12) %>%
+  add_header_above(c(" ", "Selection Model 1" = 2, "Selection Model 2" = 2))
+```
+
+<!--
+If x is the density:
+$$
+\frac{dR}{dx} = 0.758 - 2 * 0.132 * density 
+$$
+
+Then, the maximum of this curve is when the density is:
+$$
+density = \frac{0.758}{2 * 0.132}
+$$
+-->
+
+How relevant is the difference between these different model specifications?
+Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 1 and Selection Model 2.
+The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1.
+The conclusion is that at higher densities, $R_0$ is _always_ higher.
+The right panel, in contrast, shows that, according to Selection Model 2, $R_0$ is zero when density is zero (as expected), and then it tends to increase at higher densities.
+This continues until a density of approximately 2.9 (1,000 people per sq.km).
+At higher densities than that $R_0$ begins to decline, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
+
+Thus, other things being equal, the effect of density in a county like Charlottesville in Virginia	 (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
+In contrast, the effect of density on $R_0$ in a county like Arlington in Virginia (density ~3,093 people per sq.km) is _stronger_ than either of the previous two examples.
+Lastly, the density of counties like San Francisco in California, or Queens and Bronx in NY, which are among the densest in the US, contributes even less to $R_0$ than even the most rural counties in the country.
+
+It is worth at this point to recall Cressie's dictum about modelling: "[w]hat is one person's mean structure could be another person's correlation structure" [@Cressie1989geostatistics, p.
+201].
+There are almost always multiple ways to approach a modelling situation.
+In the present case, we would argue that spatial sampling is an important aspect of the modeling process, but one that perhaps required different technical skills than those available to SWN.
+There is nothing wrong with that.
+What matters is that, by adopting relatively high reproducibility standards, these researchers made a valuable and honest contribution to the collective enterprise of seeking knowledge.
+Their effort, and subsequent efforts to validate and expand on their work, can potentially contribute to provide clarity to ongoing conversations about the relevance of density and the spread of COVID-19.
+
+In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [as put by Wong and Li, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading].
+At the same time, this also stands in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with in a longitudinal setting [@Hamidi2020longitudinal.
+In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al.
+[-@Fielding2020social] might be onto something with respect to the difference between rural and urban counties.
+More generally, in population-level studies, density is indicative of proximity, no doubt about that, but also for adaptive behavior.
+And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.
+
+```{r comparison-results, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2."}
+# The maximum in-sample density is 6.255358
+# county_geo_clean %>% st_drop_geometry() %>% filter(R > 0) %>% summary()
+
+example_df <- data.frame(density = seq(0, 6.3, 0.01)) %>% 
+  mutate(density_log = log(density),
+         density_2 = density^2, 
+         b_swn = model3$coefficients$fixed["density_log"] * density_log,
+         b_selection = selection2$estimate[7] * density + selection2$estimate[8] * density_2)
+
+# Plot of coefficient of density to R_0 according to SWN Model 3
+density_swn <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_swn)) +
+  ggtitle("SWN Model 3") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+# Plot of coefficient of density to R_0  
+density_selection <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_selection)) +
+  ggtitle("Selection Model 2") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+density_swn + density_selection
+```
+
+```{r counties-by-density, include=FALSE}
+# the relationship between density and R_0 become negative in Selection Model 2 at about 5.74
+
+# Values of b_selection for a density similar a Philadelphia 
+example_df %>% filter(density > 4.10 & density < 4.14)
+# Densities with a similar b_selection
+example_df %>% filter(b_selection > 0.87 & b_selection < 0.89)
+
+# Counties with densities with similar response to Philadelphia
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 1.59 & density < 1.64) %>%
+  select(NAME.y, density)
+
+# Counties with densities associated with the highest response to density
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 2.70 & density < 3.10) %>%
+  select(NAME.y, density)
+
+# Highly dense counties
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 5.74) %>%
+  select(NAME.y)
+
+```
+
+<!--
+
+## Experiments with tobit and spatially autoregressive tobit
 
@@
 
-```{r}
-table1model1 <- censReg(R ~ density_log, 
+```{r include=FALSE}
+#
+tobit1 <- censReg(R ~ density_log, 
                   left = 0,
                   data = county_geo_clean)
-summary(table1model1)
-```
 
-Linear mixed models of R0 and density + % private transportation
-```{r}
-table1model2 <- censReg(R ~ density_log + private, 
+#
+tobit2 <- censReg(R ~ density_log + private, 
                   data = county_geo_clean)
-summary(table1model2)
-```
 
-Linear mixed models of R0 and density + % private transportation + median income
-```{r}
-table1model3 <- censReg(R ~ density_log + private + hincome, 
+#
+tobit3 <- censReg(R ~ density_log + private + hincome, 
                   data = county_geo_clean)
-summary(table1model3)
-```
-
-Tobit models of R0 and density + % private transportation + median income + interaction of density and % private transportation
-```{r}
-table1model4 <- censReg(R ~ density_log*private + hincome, 
-                    data = county_geo_clean)
-summary(table1model4)
-
 ```
@@
 
+-->
+
 # Conclusion
 
-Words go here.
+The tension between the need to publish research potentially useful in dealing with a global pandemic, and a "carnage of substandard research" [@Bramstedt2020carnage], highlights the importance of efforts to maintain the quality of scientific outputs during COVID-19.
+An important part of quality control is the ability of independent researchers to verify and examine the results of materials published in the literature.
+As previous research illustrates, reproducibility in scientific research remains an important but elusive goal [e.g., @Iqbal2016reproducible; @Stodden2018empirical; @Sumner2020reproducibility; @Gustot2020quality].
+This idea is reinforced by the review conducted for this paper in the context of research about population density and the spread of COVID-19.
+
+Taking one recent example from the literature [Sy et al., @Sy2021population; SWN], the present paper illustrates the importance of good reproducibility practices.
+Sharing data and code can catalyze research, by allowing independent verification of findings, as well as additional research.
+After verifying the results of SWN, experiments with sample selection models and variations in the definition of model inputs, lead to an important reappraisal of the conclusion that high density is associated with greater spread of the virus.
+Instead, the possibility of a non-monotonical relationship between population density and contagion is raised.
+
+In the spirit of openness, this paper is prepared as an R Markdown document, an a companion data package is provided.
+The data package contains the relevant documentation of the data, and all data pre-processing is fully documented.
+Hopefully this, and similar reproducible papers, will continue to encourage others to adopt reproducible standards in their research.
 

### 2021-05-21 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
First version for submission to PLoS ONE

@@
 global pandemic that followed in its wake, led to an explosive growth of
-research. According to Fraser et al. {[}1{]}, over 125,000
-COVID-19-related papers were released in the first ten months from the
-first confirmed case of the disease. Of these, more than 30,000 were
-shared in pre-print servers, the use of which also exploded in the past
-year {[}2--4{]}.
+research around the globe. According to Fraser et al. {[}1{]}, over
+125,000 COVID-19-related papers were released in the first ten months
+from the first confirmed case of the disease. Of these, more than 30,000
+were shared in pre-print servers, the use of which also exploded in the
+past year {[}2--4{]}.
 
@@
 regions safer from the pandemic? Are de-densification policies
-warranted, at least in the short term? And in the longer term, will the
-risks of life in high density regions presage a flight from cities? Over
-the past year, numerous papers have sought to throw light into the
+warranted, even if just in the short term? And in the longer term, will
+the risks of life in high density regions presage a flight from cities?
+Over the past year, numerous papers have sought to throw light into the
 underlying issue of density and the pandemic; nonetheless the results,
@@
 
-\hypertarget{background}{%
-\section{Background}\label{background}}
+\hypertarget{background-the-intuitive-relationship-between-density-and-spread-of-contagious-diseases}{%
+\section{Background: the intuitive relationship between density and
+spread of contagious
+diseases}\label{background-the-intuitive-relationship-between-density-and-spread-of-contagious-diseases}}
 
@@
 contagion indicate that higher densities can catalyze the transmission
-of contagious diseases {[}23,24{]}. Models such as these were likely at
+of contagious diseases {[}23,24{]}. The idea is intuitive and likely at
 the root of messages, by some figures in positions of authority, that
@@
 Nonetheless, the idea that high density correlates with high
-transmission is so intuitive that it is often taken for granted even at
-larger scales {[}e.g., 25,26{]}. At larger scales, however, there exists
-the possibility of behavioral adaptations, which are difficult to
-capture in the mechanistic framework of differential equations {[}or can
-be missing in agent-based models, e.g., 27{]}; these adaptations, in
-fact, can be a key aspect of disease transmission.
+transmission is so seemingly reasonable that it is often taken for
+granted even at larger scales {[}e.g., 25,26{]}. At larger scales,
+however, there exists the possibility of behavioral adaptations, which
+are difficult to capture in the mechanistic framework of differential
+equations {[}or can be missing in agent-based models, e.g., 27{]}; these
+adaptations, in fact, can be a key aspect of disease transmission.
 
@@
 
-In this respect, the literature to date remains inconclusive.
+\hypertarget{background-but-what-does-the-literature-say}{%
+\section{Background: but what does the literature
+say?}\label{background-but-what-does-the-literature-say}}
+
+When it comes to population density and the spread of COVID-19, the
+international literature to date remains inconclusive.
 
@@
 This is why reproducibility is important. To pick an example (which will
-be further elaborated in the following sections), the study of Sy et
-al.~{[}{[}19{]}; hereafter SWN{]} would immediately grab the attention
-of a researcher with expertise in spatial modelling.
+be further elaborated in later sections of this paper), the study of Sy
+et al.~{[}{[}19{]}; hereafter SWN{]} would immediately grab the
+attention of a researcher with a somewhat different toolbox.
+
+\hypertarget{reproducibility-of-research}{%
+\section{Reproducibility of
+research}\label{reproducibility-of-research}}
 
@@
 counties, and its association with population density, median household
-income, and prevalence of private mobility. The basic reproductive
+income, and prevalence of private mobility. For their multivariate
+analysis, SWN used mixed linear models. This is a reasonable modelling
+choice: \(R_0\) is an interval-ratio variable that is suitably modeled
+using linear regression; further, as SWN note there is a likelihood that
+the process in not independent ``among counties within each state,
+potentially due to variable resource allocation and differing health
+systems across states'' (p.~3). A mixed linear model accounts for this
+by introducing random components (in the case of SWN, random intercepts
+at the state level). SWN estimated various models with different
+combinations of variables, including median household income and
+prevalence of travel by private transportation. These are sensible
+controls, given potential variations in behavior: people in more
+affluent counties may have greater opportunities to work from home, and
+use of private transportation reduces contact with strangers. Moreover,
+they also conducted various sensitivity analyses. After these efforts,
+SWN conclude that there is a positive association between the basic
+reproductive number and population density at the level of counties in
+the US.
+
+One salient aspect of the analysis in SWN is that the basic reproductive
 number can only be calculated reliably with a minimum number of cases,
 and a large number of counties did not meet such threshold. As
-researchers do, SWN made certain seemingly reasonable modelling
-decision, for instance basing their analysis only on counties with valid
-observations. An experiences spatial modeler would likely ask some of
-the following questions on reading SWN's paper: how were missing
-counties treated? Is it possible to spatially interpolate the missing
-observations? What are the implications of the spatial sampling
-framework used in the analysis? And, was there evidence of spatial
-autocorrelation in the residuals of the models? These are questions that
-in most cases would not occur to a researcher who has not been exposed
-to spatial statistics or spatial econometrics. Nonetheless, they are
-relevant and important. Fortunately, SWN give an example of a reasonably
+researchers do, SWN made modelling decisions, in this case basing their
+analysis only on counties with valid observations. A modeler with
+expertise in different methods would likely ask some of the following
+questions on reading SWN's paper: how were missing counties treated?
+What are the implications of the spatial sampling framework used in the
+analysis? Is it possible to spatially interpolate the missing
+observations? And, was there evidence of spatial autocorrelation in the
+residuals of the models? These questions are relevant and their
+implications important. Fortunately, SWN are an example of a reasonably
 open, reproducible research product: their paper is accompanied by (most
@@
 that an independent expert can, with only a moderate investment of time
-and effort, replicate the results in the paper, as well as ask
+and effort, reproduce the results in the paper, as well as ask
 additional questions.
@@
 
-In the following sections, the analysis of RWN is replicated, some
+In the following sections, the analysis of RWN is reproduced, some
 relevant questions from the perspective of a spatial modeler are asked,
@@
 number of cases to be able to assume that there is community
-transmission of the pathogen. Accordingly, SNW based their analysis only
+transmission of the pathogen. Accordingly, SWN based their analysis only
 on counties that had at least 25 cases or more at the end of the
@@
 
-SWN used for their multivariate analysis mixed linear models. This is a
-reasonable modelling choice: \(R_0\) is an interval-ratio variable that
-is suitably modeled using linear regression; further, as SWN note there
-is a likelihood that the process in not independent ``among counties
-within each state, potentially due to variable resource allocation and
-differing health systems across states'' (p.~3). A mixed linear model
-accounts for this by introducing random components (in the case of SNW,
-random intercepts at the state level). SWN estimated various models with
-different combinations of variables, including median household income
-and prevalence of travel by private transportation. These are sensible
-controls, given potential variations in behavior: people in more
-affluent counties may have greater opportunities to work from home, and
-use of private transportation reduces contact with strangers. Moreover,
-they also conducted various sensitivity analyses. After these efforts,
-SNW conclude that there is a positive association between the basic
-reproductive number and population density at the level of counties in
-the US.
-
-Table \ref{tab:swn-results} replicates the first three models of SWN
+Table \ref{tab:swn-results} reproduces the first three models of SWN
 (the fourth model did not have any significant variables; see Table 1 in
-SNW). It is possible to verify that the results match with the minor
-(and irrelevant) exception of the magnitude of the coefficient for
+SWN). It is possible to verify that the results match, with only the
+minor (and irrelevant) exception of the magnitude of the coefficient for
 travel by private transportation, which is due to a difference in the
 input (here the variable is one percent units, whereas in SWN it was ten
-percent units). The intercept in these models is a random variable, and
-the standard deviation is reported in the fourth row of Table
-\ref{tab:swn-results}. It is useful to map the random intercepts: as
-seen in Figure \ref{fig:random-terms-map}, other things being equal,
-counties in Texas tended to have somewhat lower values of \(R_0\) (i.e.,
-a negative random intercept), whereas counties in South Dakota tended to
-have higher values of \(R_0\). The key of the analysis, that population
-density has a positive association with the basic reproductive number,
-is verified. But is it really?
+percent units). The mixed linear model gives random intercepts (i.e.,
+the intercept is a random variable), and the standard deviation is
+reported in the fourth row of Table \ref{tab:swn-results}. It is useful
+to map the random intercepts: as seen in Figure
+\ref{fig:random-terms-map}, other things being equal, counties in Texas
+tend to have somewhat lower values of \(R_0\) (i.e., a negative random
+intercept), whereas counties in South Dakota tend to have higher values
+of \(R_0\). The key of the analysis, after extensive sensitivity
+analysis, is a robust finding that population density has a positive
+association with the basic reproductive number. But does it?
 
@@
 
-\caption{\label{tab:tabulate-swn-results}\label{tab:swn-results}Replicating SWN: Models 1-3}
+\caption{\label{tab:tabulate-swn-results}\label{tab:swn-results}Reproducing SWN: Models 1-3}
 \centering
@@
 
-Thanks to the availability of code and data, it was possible to verify
-the results reported by SWN. As noted earlier, though, an experienced
-spatial modeller would have asked themselves about the implications of
-the spatial sampling procedure used by SWN. The decision to use a sample
-of counties with reliable basic reproductive numbers, although seemingly
-sensible, results in a non-random sampling scheme. Turning our attention
-back to Figure \ref{fig:R0-map}, we see that many counties without
-reliable values of \(R_0\) are in more rural, less dense parts of the
-United States. The fact that \(R_0\) could not be accurately computed in
-those counties does not mean that there was no transmission of the
-virus: it simply means that we do not know with precision whether that
-was the case. The low number of cases may be related to low population
-and/or low population density. This is intriguing, to say the least: by
-excluding cases in a non-random way such as this, we are in effect
-\emph{truncating} the sample.
-
-The issue with sample truncation is that parameter estimates can become
-unreliable.
-
-Alas, despite decades' worth of developments in the field of
-geographical analysis, not all research presented to date has used
-proper methods for the study of COVID-19. While the answer to the
-question ``do spatial effects really matter in regression analysis'' was
-definitively answered in the positive at least 40 years ago, in practice
-many researchers continue to ignore the pitfalls of ignoring them. In
-this paper, I present a reanalysis of the data used by Sy et al.~(2021)
-to study the correlations between the basic reproductive number of
-COVID-19 and population density in counties. I highlight two related
-issues: non-systematic sampling in space and spatial autocorrelation.
-The reanalysis is based on the use of tobit models to account for
-non-systematic sampling, and spatially autoregressive tobit models to
-account for spatial autocorrelation in the data generation process. The
-reanalysis highlights the importance of openness and reproducibility in
-COVID-19 research. Finally, the results provide a sobering example of
-the risks of not using appropriate methods in the analysis of
-geographical data. Furthermore,
-
-\hypertarget{fit-tobit-version-of-models}{%
-\subsection{Fit tobit version of
-models}\label{fit-tobit-version-of-models}}
+The preceding section shows that thanks to the availability of code and
+data, it is possible to verify the results reported by SWN. As noted
+earlier, though, an independent researcher might have wondered about the
+implications of the spatial sampling procedure used by SWN. The decision
+to use a sample of counties with reliable basic reproductive numbers,
+although apparently sensible, results in a non-random spatial sampling
+scheme. Turning our attention back to Figure \ref{fig:R0-map}, we form
+the impression that many counties without reliable values of \(R_0\) are
+in more rural, less dense parts of the United States. This impression is
+reinforced when we overlay the boundaries of urban areas with population
+greater than 50,000 on the counties with valid values of \(R_0\) (see
+Figure \ref{fig:urban-areas-map}). The fact that \(R_0\) could not be
+accurately computed in many counties without large urban areas does not
+mean that there was no transmission of the virus: it simply means that
+we do not know with precision whether that was the case. The low number
+of cases may be related to low population and/or low population density.
+This is intriguing, to say the least: by excluding cases based on the
+ability to calculate \(R_0\) we are potentially \emph{censoring} the
+sample in a non-random way.
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/urban-areas-map-1} \caption{\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown).}\label{fig:urban-areas-map}
+\end{figure}
+
+A problematic issue with non-random sample selection is that parameter
+estimates can become unreliable, and numerous techniques have been
+developed over time to address this. A model useful for sample selection
+problems is Heckman's selection model {[}see 62{]}. The selection model
+is in fact a system of two equations, as follows:
+
+\[
+\begin{array}{c}
+y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
+y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
+\end{array}
+\] \noindent where \(y_i^{S*}\) is a latent variable for the sample
+selection process, and \(y_i^{O*}\) is the latent outcome. Vectors
+\(x_i^S\) and \(x_i^O\) are explanatory variables (with the possibility
+that \(x_i^S = x_i^S\)). Both equations include random terms (i.e.,
+\(\epsilon_i^S\) and \(\epsilon_i^O\)) The first equation is designed to
+model the \emph{probability} of sampling, and the second equation the
+outcome of interest (say \(R_0\)). The random terms are jointly
+distributed and correlated with parameter \(\rho\).
+
+What the analyst observes is the following: \[
+y_i^S =
+\begin{cases}
+0 & \text{if } y_i^{S*} < 0\\
+1 & \text{otherwise}
+\end{cases}
+\] \noindent and: \[
+y_i^O =
+\begin{cases}
+0 & \text{if } y_i^{S} = 0\\
+y_i^{O*} & \text{otherwise}
+\end{cases}
+\]
+
+In other words, the outcome of interest is observed \emph{only} for
+certain cases (\(y_i^S=1\), i.e., for sampled observations). The
+probability of sampling depends on \(x_i^S\). For the cases observed,
+the outcome \(y_i^O\) depends on \(x_i^O\).
+
+A sample selection model is estimated using the same selection of
+variables as SWN Model 3. This is Sample Selection Model 1 in Table
+\ref{tab:selection-results}. The first thing to notice about this model
+is that the sample selection process and the outcome are not independent
+(\(\rho\ne0\) with 5\% of confidence). The selection equation indicates
+that the probability of a county to be in the sample increases with
+population density (but at a decreasing rate due to the
+log-transformation), when travel by private modes is more prevalent, and
+as median household income in the county is higher. This is in line with
+the impression left by Figure \ref{fig:urban-areas-map} that counties
+with reliable values of \(R_0\) tended to be those with larger urban
+centers. Once that the selection probabilities are accounted for in the
+model, several things happen with the outcomes model. First, the
+coefficient for population density is still positive, but the magnitude
+changes: in effect, it appears that the effect of density is more
+pronounced than what SWN Model 3 indicated. The coefficient for percent
+of private transportation changes signs. And the coefficient for median
+household income is now significant.
+
+The second model in Table \ref{tab:selection-results} (Selection Model
+2) changes the way the variables are entered into the model. The
+log-transformation of density in SWN and Selection Model 1 assumes that
+the association between density and \(R_0\) is monotonically increasing
+(if the sign of the coefficient is positive) or decreasing (if the sign
+of the coefficient is negative). There are some indications that the
+relationship may actually not be monotonical. For example, Paez et al.
+{[}32{]} found a positive (if non-significant) relationship between
+density and incidence of COVID-19 in the provinces of Spain at the
+beginning of the pandemic. This changed to a negative (and significant)
+relationship during the lockdown. In the case of the US, Fielding-Miller
+et al. {[}44{]} found that the association between COVID-19 deaths and
+population density was positive in rural counties, but negative in urban
+counties. A variable transformation that allows for non-monotonic
+changes in the relationship is the square of the density.
+
+As seen in the table, Selection Model 2 replaces the log-transformation
+of population density with a quadratic expansion. The results of this
+analysis indicate that with this variable transformation, the selection
+and outcome processes are still not independent (\(\rho\ne0\) with 5\%
+of confidence). But a few interesting things emerge. When we examine the
+outcomes model, we see that the quadratic expansion has a positive
+coefficient for the first order term, but a negative coefficient for the
+second order term. This indicates that \(R_0\) initially tends to
+increase with higher density, but only up to a point, after which the
+negative second term (which grows more rapidly due to the square),
+becomes increasingly dominant. Secondly, the sign of the coefficient for
+travel by private transportation becomes negative again. This, of
+course, makes more sense than the positive sign of Selection Model 1: if
+people tend to travel in private transportation, the potential for
+contact should be lower instead of higher. And finally median household
+income is no longer significant.
 
@@
 
-\caption{\label{tab:unnamed-chunk-3}\label{tab:tobit-results}Marginal effects of sample selection models}
+\caption{\label{tab:tabulate-sample-selection-results}\label{tab:selection-results}Estimation results of sample selection models}
 \centering
 \resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllllrl}
+\begin{tabular}[t]{lcccc}
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Tobit Model 1} & \multicolumn{2}{c}{Tobit Model 2} & \multicolumn{2}{c}{Tobit Model 3} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
-Variable & m.e. & 95\% CI & m.e. & 95\% CI & m.e. & 95\% CI\\
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Selection Model 1} & \multicolumn{2}{c}{Selection Model 2} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+Variable & $\beta$ & 95\% CI & $\beta$ & 95\% CI\\
 \midrule
-Log of population density & 0.264 & [0.246, 0.281] & 0.264 & [0.247, 0.281] & 0.233 & [0.216, 0.25]\\
-Percent of private transportation &  &  & 0.006 & [0.002, 0.01] & 0.014 & [0.01, 0.019]\\
-Median household income (\$10,000) &  &  &  &  & 0.105 & [0.087, 0.122]\\
+\addlinespace[0.3em]
+\multicolumn{5}{l}{\textbf{Sample Selection Model}}\\
+\hspace{1em}Intercept & -2.237 & [-3.109, -1.365] & -7.339 & [-8.381, -6.297]\\
+\hspace{1em}Log of population density & 0.385 & [0.352, 0.418] &  & \\
+\hspace{1em}Density (1,000 per sq.km) &  &  & 2.484 & [2.13, 2.838]\\
+\hspace{1em}Density squared &  &  & -0.387 & [-0.473, -0.3]\\
+\hspace{1em}Percent of private transportation & 0.025 & [0.016, 0.034] & 0.057 & [0.046, 0.067]\\
+\hspace{1em}Median household income (10,000) & 0.202 & [0.168, 0.235] & 0.32 & [0.283, 0.357]\\
+\addlinespace[0.3em]
+\multicolumn{5}{l}{\textbf{Outcome Model}}\\
+\hspace{1em}Intercept & 0.605 & [-0.257, 1.466] & 2.784 & [1.652, 3.915]\\
+\hspace{1em}Log of population density & 0.39 & [0.354, 0.426] &  & \\
+\hspace{1em}Density (1,000 per sq.km) &  &  & 0.758 & [0.509, 1.008]\\
+\hspace{1em}Density squared &  &  & -0.132 & [-0.187, -0.077]\\
+\hspace{1em}Percent of private transportation & 0.01 & [0.001, 0.018] & -0.011 & [-0.021, -0.001]\\
+\hspace{1em}Median household income (\$10,000) & 0.126 & [0.094, 0.159] & 0.002 & [-0.033, 0.037]\\
+$\sigma$ & 0.954 & [0.904, 1.003] & 0.684 & [0.652, 0.716]\\
+$\rho$ & 0.971 & [0.961, 0.98] & -0.199 & [-0.377, -0.022]\\
 \bottomrule
@@
 
-\hypertarget{spatially-autoregressive-tobit}{%
-\subsection{Spatially autoregressive
-tobit}\label{spatially-autoregressive-tobit}}
-
-Fit spatially autoregressive tobit:
+How relevant is the difference between these different model
+specifications? Figure \ref{fig:comparison-results} shows the
+relationship between density and \(R_0\) implied by SWN Model 1 and
+Selection Model 2. The left panel of the figure shows the non-linear but
+monotonic relationship implied by SWN Model 1. The conclusion is that at
+higher densities, \(R_0\) is \emph{always} higher. The right panel, in
+contrast, shows that, according to Selection Model 2, \(R_0\) is zero
+when density is zero (as expected), and then it tends to increase at
+higher densities. This continues until a density of approximately 2.9
+(1,000 people per sq.km). At higher densities than that \(R_0\) begins
+to decline, and the relationship becomes negative at densities higher
+than approximately 5.7 (1,000 people per sq.km).
+
+Thus, other things being equal, the effect of density in a county like
+Charlottesville in Virginia (density \textasciitilde1,639 people per
+sq.km) is roughly the same as that in a county like Philadelphia
+(density \textasciitilde4,127 people per sq.km). In contrast, the effect
+of density on \(R_0\) in a county like Arlington in Virginia (density
+\textasciitilde3,093 people per sq.km) is \emph{stronger} than either of
+the previous two examples. Lastly, the density of counties like San
+Francisco in California, or Queens and Bronx in NY, which are among the
+densest in the US, contributes even less to \(R_0\) than even the most
+rural counties in the country.
+
+It is worth at this point to recall Cressie's dictum about modelling:
+``{[}w{]}hat is one person's mean structure could be another person's
+correlation structure'' {[}63{]}. There are almost always multiple ways
+to approach a modelling situation. In the present case, we would argue
+that spatial sampling is an important aspect of the modeling process,
+but one that perhaps required different technical skills than those
+available to SWN. There is nothing wrong with that. What matters is
+that, by adopting relatively high reproducibility standards, these
+researchers made a valuable and honest contribution to the collective
+enterprise of seeking knowledge. Their effort, and subsequent efforts to
+validate and expand on their work, can potentially contribute to provide
+clarity to ongoing conversations about the relevance of density and the
+spread of COVID-19.
+
+In particular, it is noteworthy that a sample selection model with a
+different variable transformation does not lend support to the thesis
+that higher density is \emph{always} associated with a greater risk of
+spread of the virus {[}as put by Wong and Li, ``\,`Density is destiny'
+is probably an overstatement''; {[}46{]}{]}. At the same time, this also
+stands in contrast to the findings of Hamidi et al., who found that
+higher density was either not significantly associated with the rate of
+the virus in a cross-sectional study {[}33{]}, or was negatively
+associated with in a longitudinal setting {[}{[}51{]}. In this sense,
+the conclusion that density does not aggravate the pandemic may have
+been somewhat premature; instead, reanalysis of the data of SWN suggests
+that Fielding-Miller et al. {[}44{]} might be onto something with
+respect to the difference between rural and urban counties. More
+generally, in population-level studies, density is indicative of
+proximity, no doubt about that, but also for adaptive behavior. And it
+is possible that the determining factor during COVID-19, at least in the
+US, has been variations in perceptions of the risks associated with
+contagion {[}31{]}, and subsequent compensations in behavior in more and
+less dense regions.
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/comparison-results-1} \caption{\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2.}\label{fig:comparison-results}
+\end{figure}
 
@@
 
-Words go here.
+The tension between the need to publish research potentially useful in
+dealing with a global pandemic, and a ``carnage of substandard
+research'' {[}5{]}, highlights the importance of efforts to maintain the
+quality of scientific outputs during COVID-19. An important part of
+quality control is the ability of independent researchers to verify and
+examine the results of materials published in the literature. As
+previous research illustrates, reproducibility in scientific research
+remains an important but elusive goal {[}e.g., 12,13--15{]}. This idea
+is reinforced by the review conducted for this paper in the context of
+research about population density and the spread of COVID-19.
+
+Taking one recent example from the literature {[}Sy et al., {[}19{]};
+SWN{]}, the present paper illustrates the importance of good
+reproducibility practices. Sharing data and code can catalyze research,
+by allowing independent verification of findings, as well as additional
+research. After verifying the results of SWN, experiments with sample
+selection models and variations in the definition of model inputs, lead
+to an important reappraisal of the conclusion that high density is
+associated with greater spread of the virus. Instead, the possibility of
+a non-monotonical relationship between population density and contagion
+is raised.
+
+In the spirit of openness, this paper is prepared as an R Markdown
+document, an a companion data package is provided. The data package
+contains the relevant documentation of the data, and all data
+pre-processing is fully documented. Hopefully this, and similar
+reproducible papers, will continue to encourage others to adopt
+reproducible standards in their research.
 
@@
 
+\leavevmode\hypertarget{ref-Maddala1983limited}{}%
+\CSLLeftMargin{62. }
+\CSLRightInline{Maddala GS. Limited-dependent and qualitative variables
+in econometrics. Cambridge: Cambridge University Press; 1983. }
+
+\leavevmode\hypertarget{ref-Cressie1989geostatistics}{}%
+\CSLLeftMargin{63. }
+\CSLRightInline{Cressie N. Geostatistics. The American Statistician.
+1989;43: 197.
+doi:\href{https://doi.org/10.2307/2685361}{10.2307/2685361}}
+
 \end{CSLReferences}

### 2021-05-21 README.Rmd
Knitted README.Rmd with first submission version

@@
 
-Paper in preparation for submision to *PLoS ONE*
+Paper submitted to *PLoS ONE*
 
@@
 
-## Introduction
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(cache = TRUE)
+```
 
-The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the global pandemic that followed in its wake, led to an explosive growth of research.
+```{r install-data-package, include = FALSE}
+# Run only once if needed to install data package `r0density`
+if (!require("r0density", character.only = TRUE)) {
+  devtools::install_github("https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed", subdir = "r0density")
+}
+```
+
+```{r load-packages, include=FALSE, cache=FALSE}
+library(censReg) # Package to estimate tobit model
+library(gridExtra)
+library(kableExtra) # Package for creating tables
+library(Matrix) # Used to convert list of spatial weights to sparse matrix; needed for spatialprobit
+library(nlme) # Used to estimate 
+library(patchwork)
+library(r0density) # Data package
+library(sampleSelection) # Package to estimate sampling selection models
+library(sf) # Package to work with spatial information in simple features format
+library(spatialprobit) # Used to estimate spatially autoregressive tobit
+library(spdep) # Utilities to create spatial weights and work with spatial data
+library(tidycensus) # Package to retrieve census information 
+library(tidyverse) # Data carpentry and analysis tools
+library(units) # Package to work with units of measurement
+library(viridis) # Color palettes
+```
+
+```{r invoke-data, include=FALSE}
+data(county_geo)
+data(state_geo)
+data(urba_geo)
+```
+
+```{r prepare-data, include=FALSE}
+# Convert missing R's to zeros and calculate expansion variables: 
+county_geo_clean <- county_geo %>%
+  mutate(R = replace_na(R, 0),
+         density_log = log(drop_units(density)),
+         density_2 = drop_units(density)^2,
+         hincome_log = log(hincome),
+         private = car/commuters * 100) %>% # Percentage of commuters who travel by car/van/truck (in 10%)
+  drop_na(commuters) %>%
+  st_as_sf()
+
+# Boundary of continental US excluding Alaska
+us_geo <- st_union(state_geo %>% filter(!STATE %in% c("02", "15", "72")))
+```
+
+```{r list-of-geographical-neighbors, include=FALSE}
+# Create a list of neighbors based on the polygons; the default contiguity criterion is Queen:
+# Counties
+county_nb <- county_geo_clean %>%
+  poly2nb()
+# States
+state_nb <- state_geo %>% 
+  filter(!STATE %in% c("02", "15", "72")) %>%
+  poly2nb()
+```
+
+```{r spatial-data-check, include=FALSE}
+# Check list of neighbors
+summary(county_nb)
+
+# Notice that there are 9 counties that are isolates. These are the counties that do not have neighbors based on contiguity
+county_geo_clean[c(69, 546, 547, 549, 1226, 1876, 2974, 3146, 3200),]
+
+# To rectify the lack of neighbors, find the first nearest neighbor using the centroids:
+county_nb_dist <- knearneigh(county_geo_clean %>% 
+                               st_centroid()) %>%
+  knn2nb()
+
+# Add nearest neighbor to isolates:
+county_nb[[69]] <- county_nb_dist[[69]]
+county_nb[[546]] <- county_nb_dist[[546]]
+county_nb[[547]] <- county_nb_dist[[547]]
+county_nb[[549]] <- county_nb_dist[[549]]
+county_nb[[1226]] <- county_nb_dist[[1226]]
+county_nb[[1876]] <- county_nb_dist[[1876]]
+county_nb[[2974]] <- county_nb_dist[[2974]]
+county_nb[[3146]] <- county_nb_dist[[3146]]
+county_nb[[3200]] <- county_nb_dist[[3200]]
+```
+
+```{r spatial-weights, include=FALSE}
+# Convert the nearest neighbor object to a listw object:
+nb_B <- nb2listw(county_nb, 
+                 style="B", 
+                 zero.policy = TRUE)
+
+state_listw <- nb2listw(state_nb, 
+                        style="W", 
+                        zero.policy = TRUE)
+
+# Convert listw to sparse matrix for use in spatial tobit:
+B <- as(nb_B, "CsparseMatrix")
+```
+
+# Introduction
+
+The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the global pandemic that followed in its wake, led to an explosive growth of research around the globe.
 According to Fraser et al.
@@
 In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how].
-The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, prompted a flurry of research into the associations between density and the spread of the pandemic.
+The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, brought a torrent of research into the associations between density and the spread of the pandemic.
 Some important questions hang on the results of these research efforts.
 For example, are lower density regions safer from the pandemic?
-Are de-densification policies warranted, at least in the short term?
+Are de-densification policies warranted, even if just in the short term?
 And in the longer term, will the risks of life in high density regions presage a flight from cities?
@@
 
-## References
+# Background: the intuitive relationship between density and spread of contagious diseases
+
+The concern with population density and the spread of the virus during the COVID-19 pandemic was fueled, at least in part, by dramatic scenes seen in real-time around the world from large urban centers such as Wuhan, Milan, Madrid, and New York.
+In theory, there are good reasons to believe that higher density may have a positive association with the transmission of a contagious virus.
+It has long been known that the potential for inter-personal contact is greater in regions with higher density [see for example the research on urban fields and time-geography @Moore1970urban; @Moore1970some; @Farber2011running].
+Mathematically, models of exposure and contagion indicate that higher densities can catalyze the transmission of contagious diseases [@Rocklov2020high; @Li2018effect].
+The idea is intuitive and likely at the root of messages, by some figures in positions of authority, that regions with sparse population densities faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic down without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].
+
+As Rocklöv and Sjödin [@Rocklov2020high] note, however, mathematical models of contagion are valid at small-to-medium spatial scales (and presumably, small temporal scales too, such as time spent in restaurants, concert halls, cruises), and the results do not necessarily transfer to larger spatial units and different time scales.
+There are solid reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers-however, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place.
+Nonetheless, the idea that high density correlates with high transmission is so seemingly reasonable that it is often taken for granted even at larger scales [e.g., @Cruz2020exploring; @Micallef2020first].
+At larger scales, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, e.g., @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.
+
+A plausible behavioral adaptation during a pandemic, especially one broadcast as widely and intensely as COVID-19, is risk compensation.
+Risk compensation is a process whereby people adjust their behavior in response to their _perception_ of risk [@Noland1995perceived; @Richens2000condoms; @Phillips2011risk].
+In the case of COVID-19, Chauhan et al.
+[@Chauhan2021covid] have found that perception of risks in the US varies between rural, suburban, and urban residents, with rural residents in general expressing less concern about the virus.
+It is possible that people who listened to the message of leaders saying that they were safe because of low density may not have taken adequate precautions against the virus.
+People in dense places who could more directly observe the impact of the pandemic may have become overly cautious.
+Both Paez et al.
+[-@Paez2020spatio] and Hamidi et al.
+[-@Hamidi2020density] posit this mechanism (i.e., greater compliance with social distancing in denser regions) to explain the results of their analyses.
+The evidence available does indeed show that there were important changes in behavior with respect to mobility during the pandemic [@Jamal2020Changes; @Harris2021Changes; @Molloy2020Tracing]; furthermore, shelter in place orders may have had greater buy-in from the public in higher density regions [@Feyman2020effectiveness], and the associated behavior may have persisted beyond the duration of official social-distancing policies [@Praharaj2020Using].
+In addition, there is evidence that changes in mobility correlated with the trajectory of the pandemic [@Paez2020using; @Noland2021mobility].
+Given the potential for behavioral adaptation, the question of density becomes more nuanced: it is not just a matter of proximity, but also of human behavior, which is better studied using population-level data and models.
+
+# Background: but what does the literature say?
+
+When it comes to population density and the spread of COVID-19, the international literature to date remains inconclusive.
+
+On the one hand, there are studies that report positive associations between population density and various COVID-19-related outcomes.
+Bhadra [-@Bhadra2021impact], for example, reported a moderate positive correlation between the spread of COVID-19 and population density at the district level in India, however their analysis was bivariate and did not control for other variables, such as income.
+Similarly, Kadi and Khelfaoui [-@Kadi2020population] found a positive and significant correlation between number of cases and population density in cities in Algeria in a series of simple regression models (i.e., without other controls).
+A question in these relatively simple analyses is whether density is not a proxy for other factors.
+Other studies have included controls, such as Pequeno et al.
+[-@Pequeno2020air], a team that reported a positive association between density and cumulative counts of confirmed COVID-19 cases in state capitals in Brazil after controlling for covariates, including income, transport connectivity, and economic status.
+In a similar vein, Fielding-Miller et al.
+[-@Fielding2020social] reported a positive relationship between the absolute number of COVID-19 deaths and population density (rate) in rural counties in the US.
+Roy and Ghosh [-@Roy2020factors] used a battery of machine learning techniques to find discriminatory factors, and a positive and significant association between COVID-19 infection and death rates in US states.
+Wong and Li [-@Wong2020spreading] also found a positive and significant association between population density and number of confirmed COVID-19 cases in US counties, using both univariate and multivariate regressions with spatial effects.
+More recently, Sy et al.
+[-@Sy2021population] reported that the basic reproductive number of COVID-19 in US counties tended to increase with population density, but at a decreasing rate at higher densities.
+
+On the flip side, a number of studies report non-significant or negative associations between population density and COVID-19 outcomes.
+This includes the research of Sun et al.
+[-@Sun2020impacts] who did not find evidence of significant correlation between population density and confirmed number of cases per day _in conditions of lockdown_ in China.
+This finding echoes the results of Paez et al.
+[-@Paez2020spatio], who in their study of provinces in Spain reported non-significant associations between population density and infection rates in the early days of the first wave of COVID-19, and negative significant associations in the later part of the first lockdown.
+Similarly, [-@Skorka2020macroecology] found zero or negative associations between population density and infection numbers/deaths by country.
+Fielding-Miller et al.
+[-@Fielding2020social] contrast their finding about rural counties with a negative relationship between COVID-19 deaths and population density in urban counties in the US.
+For their part, in their investigation of doubling time, White and Hébert-Dufresne [-@White2020state] identified a negative and significant correlation between population density and doubling time in US states.
+Likewise, [-@Khavarian2021high] fond a small negative (and significant) association between population density and COVID-19 morbidity in districts in Tehran.
+Finally, two of the most complete studies in the US [by Hamidi et al.; -@Hamidi2020longitudinal; -@Hamidi2020density] used an extensive set of controls to find negative and significant correlations between density and COVID-19 cases and fatalities at the level of counties in the US.
+
+As can be seen, these studies are implemented at different scales in different regions of the world.
+They also use a range of techniques, from correlation analysis, to multivariate regression, spatial regressions, and machine learning techniques.
+This is natural and to be expected: individual researchers have only limited time and expertise.
+This is why reproducibility is important.
+To pick an example (which will be further elaborated in later sections of this paper), the study of Sy et al.
+[-@Sy2021population; hereafter SWN] would immediately grab the attention of a researcher with a somewhat different toolbox.
+
+# Reproducibility of research
+
+SWN investigated the basic reproductive number of COVID-19 in US counties, and its association with population density, median household income, and prevalence of private mobility.
+For their multivariate analysis, SWN used mixed linear models.
+This is a reasonable modelling choice: $R_0$ is an interval-ratio variable that is suitably modeled using linear regression; further, as SWN note there is a likelihood that the process in not independent "among counties within each state, potentially due to variable resource allocation and differing health systems across states"  (p.
+3).
+A mixed linear model accounts for this by introducing random components (in the case of SWN, random intercepts at the state level).
+SWN estimated various models with different combinations of variables, including median household income and prevalence of travel by private transportation.
+These are sensible controls, given potential variations in behavior: people in more affluent counties may have greater opportunities to work from home, and use of private transportation reduces contact with strangers.
+Moreover, they also conducted various sensitivity analyses.
+After these efforts, SWN conclude that there is a positive association between the basic reproductive number and population density at the level of counties in the US.
+
+One salient aspect of the analysis in SWN is that the basic reproductive number can only be calculated reliably with a minimum number of cases, and a large number of counties did not meet such threshold.
+As researchers do, SWN made modelling decisions, in this case basing their analysis only on counties with valid observations.
+A modeler with expertise in different methods would likely ask some of the following questions on reading SWN's paper: how were missing counties treated?
+What are the implications of the spatial sampling framework used in the analysis?
+Is it possible to spatially interpolate the missing observations?
+And, was there evidence of spatial autocorrelation in the residuals of the models?
+These questions are relevant and their implications important.
+Fortunately, SWN are an example of a reasonably open, reproducible research product: their paper is accompanied by (most of) the data and (most of) the code used in the analysis.
+This means that an independent expert can, with only a moderate investment of time and effort, reproduce the results in the paper, as well as ask additional questions.
+
+Alas, reproducibility is not necessarily the norm in the relevant literature.
+
+There are various reasons why a project can fail to be reproducible.
+In some cases, there might be legitimate reasons to withhold the data, perhaps due to confidentiality and privacy reasons [e.g., @Lee2020human].
+But in many other cases the data are publicly available, which in fact has commonly been the case with population-level COVID-19 information.
+Typically the provenance of the data is documented, but in numerous studies the data themselves are not shared [@Amadu2021assessing; @Bhadra2021impact; @Cruz2020exploring; @Feng2020spread; @Fielding2020social; @Hamidi2020longitudinal; @Hamidi2020density; @Inbaraj2021seroprevalence; @Souris2020covid].
+As any researcher can attest, whether a graduate student or a seasoned scientist, collecting, organizing, and preparing data for a project can take a substantial amount of time.
+Pointing to the sources of data, even when these sources are public, is a small step towards reproducibility-but only a very small one.
+Faced with the prospect of having to recreate a data set from raw sources is probably sufficient to dissuade all but the most dedicated (or stubborn) researcher from independent verification.
+This is true even if part of the data are shared [e.g., @Wong2020spreading].
+In other cases, data are shared, but the processes followed in the preparation of the data are not fully documented [@Ahmad2020association; @Skorka2020macroecology].
+These processes matter, as shown by the errors in the spreadsheets of Reinhart and Rogoff [@Herndon2014high] and the data of biologist Jonathan Pruitt that led to an "avalanche" of paper retractions^[\url{https://doi.org/10.1038/d41586-020-00287-y}].
+Another situation is when papers share well-documented data, but fail to provide the code used in the analysis [@Noury2021how; @Pequeno2020air; @Wang2021transmission].
+Making code available only "on demand" [e.g., @Brandtner2021creatures] is an unnecessary barrier when most journals offer the facility to share supplemental materials online.
+Then there are those papers that more closely comply with reproducibility standards, and share well-documented processes and data, as well as the code used in any analyses reported [@Paez2020spatio; @Feyman2020effectiveness; @Stephens2021impact; @White2020state; @Sy2021population].
+
+<!--
+@Ahmad2020association -> code is not shared, data are provided but data pre-processing is not documented
+@Amadu2021assessing -> code is not shared, data provenance is documented, data are not shared
+@Bhadra2021impact -> (code and data not available; only data provenance is documented)
+@Brandtner2021creatures -> code is available on demand, data are shared
+@Cruz2020exploring -> code is shared, sources of data are documented, data are not actually provided
+@Feng2020spread -> code not shared, data provenance is documented, data are not shared
+@Feyman2020effectiveness -> code and data are provided
+@Fielding2020social -> code is not shared, data provenance is documented, data are not actually provided
+@Hamidi2020longitudinal, @Hamidi2020density ->  (code and data not available)
+@Inbaraj2021seroprevalence -> code is not shared, data are not shared
+@Lee2020human -> code not provided, data sources are documented but data are not shared (privacy & confidentiality)
+@Noury2021how -> code is not shared, data are shared
+@Pequeno2020air -> (raw data supplied in suplemental files, code not available)
+@Roy2020factors -> code is not shared, data are provided and data processing is documented
+@Skorka2020macroecology -> code and data are shared but data processing is not documented, 
+@Souris2020covid -> code not provided, data sources are documented but data are not shared
+@Stephens2021impact -> code and data are shared
+@Wang2021transmission -> code is not shared, data are available
+@White2020state -> code and data are in repository
+@Wong2020spreading -> code not provided, data are shared but data preprocessing is not documented
+-->
+
+In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of a spatial modeler are asked, and the data are reanalyzed.
+
+# Reproducing SWN
+
+SWN examined the association between the basic reproductive number of COVID-19 and population density.
+The basic reproductive number $R_0$ is a summary measure of contact rates, probability of transmission of a pathogen, and duration of infectiousness.
+In rough terms, $R_0$ measures how many new infections each infections begets.
+Infectious disease outbreaks generally tend to die out when $R_0<1$, and to grow when $R_0>1$.
+Reliable calculation of $R_0$ requires a minimum number of cases to be able to assume that there is community transmission of the pathogen.
+Accordingly, SWN based their analysis only on counties that had at least 25 cases or more at the end of the exponential growth phase (see Fig.
+\ref{fig:R0-map}).
+Their final sample included `r prettyNum(nrow(county_geo %>% filter(!is.na(R))), big.mark = ",")` counties in the US, including in Alaska, Hawaii, Puerto Rico, and island territories.
+
+```{r R0-map, echo=FALSE, fig.cap="\\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
+# Choropleth map of basic reproductive number
+ggplot() + # Create a ggplot
+  # Choropleth map of basic reproductive number
+  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
+          aes(fill = R),
+          color = NA) +
+  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
+  scale_fill_viridis_c(name = expression(R[0]),
+                       direction = -1) +
+  # Overlay counties with missing values of the reproductive number in white
+  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
+                   R == 0), 
+          fill = "white",
+          color = NA) +
+  # Overlay county boundaries
+  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
+          fill = NA,
+          color = "lightgray") +
+  # Add boundary of continental US
+  geom_sf(data = us_geo,
+          fill = NA) + 
+  # Add caption
+  labs(caption = "Note: counties in white represent missing values of the basic reproductive number") +
+  # Set theme of plot
+  theme_minimal() +
+  theme(legend.position = "bottom")
+```
+
+Table \ref{tab:swn-results} reproduces the first three models of SWN (the fourth model did not have any significant variables; see Table 1 in SWN).
+It is possible to verify that the results match, with only the minor (and irrelevant) exception of the magnitude of the coefficient for travel by private transportation, which is due to a difference in the input (here the variable is one percent units, whereas in SWN it was ten percent units).
+The mixed linear model gives random intercepts (i.e., the intercept is a random variable), and the standard deviation is reported in the fourth row of Table \ref{tab:swn-results}.
+It is useful to map the random intercepts: as seen in Figure \ref{fig:random-terms-map}, other things being equal, counties in Texas tend to have somewhat lower values of $R_0$ (i.e., a negative random intercept), whereas counties in South Dakota tend to have higher values of $R_0$.
+The key of the analysis, after extensive sensitivity analysis, is a robust finding that population density has a positive association with the basic reproductive number.
+But does it?
+
+```{r reproduce-swn, include=FALSE}
+# Fit (mixed) linear models as in Sy et al. These models are reported in table 1 of their paper
+
+# Model 1
+model1 <- lme(R ~ density_log , 
+              random = ~ 1| state, 
+              data = county_geo_clean %>%
+                filter(R > 0)) # Exclude observations with missing R
+#summary(table1model1)
+model1_intervals <- intervals(model1)
+
+# Model 2
+model2 <- lme(R ~ density_log + private , 
+              random = ~ 1| state, 
+              data = county_geo_clean %>%
+                filter(R > 0)) # Exclude observations with missing R
+#summary(model2)
+model2_intervals <- intervals(model2)
+
+# Model 3
+model3 <- lme(R ~ density_log + private + hincome, 
+              random = ~ 1| state, 
+              data = county_geo_clean %>%
+                filter(R > 0)) # Exclude observations with missing R
+#summary(model3)
+model3_intervals <- intervals(model3)
+
+# Model 4
+model4 <- lme(R ~ density_log*private + hincome, 
+              random = ~ 1| state, 
+              data = county_geo_clean %>%
+                filter(R > 0)) # Exclude observations with missing R
+#summary(model4)
+model4_intervals <- intervals(model4)
+```
+
+```{r join-random-intercepts, include=FALSE}
+state_geo <- state_geo %>% 
+  left_join(model3$coefficients$random$state %>%
+              as.data.frame() %>%
+              rownames_to_column(var = "state"),
+            by = "state") %>%
+  rename(random_intercepts = `(Intercept)`)
+```
+
+```{r prepare-swn-results-for-table, include=FALSE}
+# Join tables with results
+main_effects <- full_join(model1_intervals$fixed %>%
+                            as.data.frame() %>%
+                            rownames_to_column(),
+                          model2_intervals$fixed %>%
+                            as.data.frame() %>%
+                            rownames_to_column(),
+                          by = "rowname") %>%
+  full_join(model3_intervals$fixed %>%
+              as.data.frame() %>%
+              rownames_to_column(),
+            by = "rowname")
+
+random_component <- full_join(model1_intervals$reStruct$state %>%
+                                as.data.frame() %>%
+                                rownames_to_column(),
+                              model2_intervals$reStruct$state %>%
+                                as.data.frame() %>%
+                                rownames_to_column(),
+                              by = "rowname") %>%
+  full_join(model3_intervals$reStruct$state %>%
+              as.data.frame() %>%
+              rownames_to_column(),
+            by = "rowname")
+
+within_group_se <- full_join(model1_intervals$sigma %>%
+                               t() %>%
+                               as.data.frame() %>%
+                               rownames_to_column(),
+                             model2_intervals$sigma %>%
+                               t() %>%
+                               as.data.frame() %>%
+                               rownames_to_column(),
+                             by = "rowname") %>%
+  full_join(model3_intervals$sigma %>%
+              t() %>%
+              as.data.frame() %>%
+              rownames_to_column(),
+            by = "rowname")
+
+results_swn <- rbind(main_effects,
+                     random_component,
+                     within_group_se)
+
+colnames(results_swn) <- c("Variable", "l1", "b1", "u1", "l2", "b2", "u2", "l3", "b3", "u3")
+
+results_swn <- results_swn %>%
+  transmute(Variable,
+            b1 = ifelse(is.na(b1),
+                        " ", 
+                        round(b1, 3)),
+            ci1 = ifelse(is.na(l1), 
+                         " ",
+                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")),
+            b2 = ifelse(is.na(b2),
+                        " ", 
+                        round(b2, 3)),
+            ci2 = ifelse(is.na(l2), 
+                         " ",
+                         paste0("[", round(l2, 3), ", ", round(u2, 3), "]")),
+            b3 = ifelse(is.na(b3),
+                        " ", 
+                        round(b3, 3)),
+            ci3 = ifelse(is.na(l3), 
+                         " ",
+                         paste0("[", round(l3, 3), ", ", round(u3, 3), "]")))
+```
+
+```{r tabulate-swn-results, echo=FALSE}
+results_swn %>%
+  mutate(Variable = c("Intercept", 
+                      "Log of population density",
+                      "Percent of private transportation", 
+                      "Median household income ($10,000)",
+                      "Standard deviation (Intercept)",
+                      "Within-group standard error")) %>%
+  kable("html",
+        digits = 3,
+        booktabs = TRUE,
+        col.names = c("Variable",
+                      "beta",
+                      "95% CI",
+                      "beta",
+                      "95% CI",
+                      "beta",
+                      "95% CI"),
+        caption = "\\label{tab:swn-results}Reproducing SWN: Models 1-3") %>%
+  kable_styling(latex_options = c("scale_down")) %>%
+  add_header_above(c(" ", "Model 1" = 2, "Model 2" = 2, "Model 3" = 2))
+```
+
+```{r random-terms-map, echo=FALSE, fig.cap="\\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
+# Choropleth map of basic reproductive number
+ggplot() + 
+  geom_sf(data = state_geo %>%
+            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
+          aes(fill = random_intercepts), # Add the simple features object to the plot
+          color = "lightgray") +
+  # Add boundary of continental US
+  geom_sf(data = us_geo,
+          fill = NA) + 
+  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
+  scale_fill_viridis(name = "Random Intercepts",
+                     option = "turbo",
+                     direction = -1) +
+  # Set theme of plot
+  theme_minimal() +
+  theme(legend.position = "bottom")
+```
+
+# Expanding on SWN
+
+The preceding section shows that thanks to the availability of code and data, it is possible to verify the results reported by SWN.
+As noted earlier, though, an independent researcher might have wondered about the implications of the spatial sampling procedure used by SWN.
+The decision to use a sample of counties with reliable basic reproductive numbers, although apparently sensible, results in a non-random spatial sampling scheme.
+Turning our attention back to Figure \ref{fig:R0-map}, we form the impression that many counties without reliable values of $R_0$ are in more rural, less dense parts of the United States.
+This impression is reinforced when we overlay the boundaries of urban areas with population greater than 50,000 on the counties with valid values of $R_0$ (see Figure \ref{fig:urban-areas-map}).
+The fact that $R_0$ could not be accurately computed in many counties without large urban areas does not mean that there was no transmission of the virus: it simply means that we do not know with precision whether that was the case.
+The low number of cases may be related to low population and/or low population density.
+This is intriguing, to say the least: by excluding cases based on the ability to calculate $R_0$ we are potentially _censoring_ the sample in a non-random way.
+
+```{r urban-areas-map, echo=FALSE, fig.cap="\\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
+# Choropleth map of basic reproductive number
+ggplot() + # Create a ggplot
+  # Choropleth map of basic reproductive number
+  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
+          aes(fill = R),
+          color = NA) +
+  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
+  scale_fill_viridis_c(name = "R0",
+                       direction = -1) +
+  # Overlay counties with missing values of the reproductive number in white
+  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
+                   R == 0), 
+          fill = "white",
+          color = NA) +
+  # Overlay boundaries of urban areas
+  geom_sf(data = urban_geo %>% # Add the simple features object to the plot
+            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
+                   UATYP10 == "U"), # Include only urbanized areas with population > 50,000
+          fill = NA,
+          color = "red") +
+  # Add boundary of continental US
+  geom_sf(data = us_geo,
+          fill = NA) + 
+  # Add caption
+  labs(caption = "Note: boundaris of urbanized areas with population > 50,000 are shown in red") +
+  # Set theme of plot
+  theme_minimal() +
+  theme(legend.position = "bottom")
+```
+
+A problematic issue with non-random sample selection is that parameter estimates can become unreliable, and numerous techniques have been developed over time to address this.
+A model useful for sample selection problems is Heckman's selection model [see @Maddala1983limited].
+The selection model is in fact a system of two equations, as follows:
+
+$$
+\begin{array}{c}
+y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
+y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
+\end{array}
+$$
+\noindent where $y_i^{S*}$ is a latent variable for the sample selection process, and $y_i^{O*}$ is the latent outcome.
+Vectors $x_i^S$ and $x_i^O$ are explanatory variables (with the possibility that $x_i^S = x_i^S$).
+Both equations include random terms (i.e., $\epsilon_i^S$ and $\epsilon_i^O$)  The first equation is designed to model the _probability_ of sampling, and the second equation the outcome of interest (say $R_0$).
+The random terms are jointly distributed and correlated with parameter $\rho$.
+
+What the analyst observes is the following:
+$$
+y_i^S =
+\begin{cases}
+0 & \text{if } y_i^{S*} < 0\\
+1 & \text{otherwise}
+\end{cases}
+$$
+\noindent and:
+$$
+y_i^O =
+\begin{cases}
+0 & \text{if } y_i^{S} = 0\\
+y_i^{O*} & \text{otherwise}
+\end{cases}
+$$
+
+In other words, the outcome of interest is observed _only_ for certain cases ($y_i^S=1$, i.e., for sampled observations).
+The probability of sampling depends on $x_i^S$.
+For the cases observed, the outcome $y_i^O$ depends on $x_i^O$.
+
+A sample selection model is estimated using the same selection of variables as SWN Model 3.
+This is Sample Selection Model 1 in Table \ref{tab:selection-results}.
+The first thing to notice about this model is that the sample selection process and the outcome are not independent ($\rho\ne0$ with 5% of confidence).
+The selection equation indicates that the probability of a county to be in the sample increases with population density (but at a decreasing rate due to the log-transformation), when travel by private modes is more prevalent, and as median household income in the county is higher.
+This is in line with the impression left by Figure \ref{fig:urban-areas-map} that counties with reliable values of $R_0$ tended to be those with larger urban centers.
+Once that the selection probabilities are accounted for in the model, several things happen with the outcomes model.
+First, the coefficient for population density is still positive, but the magnitude changes: in effect, it appears that the effect of density is more pronounced than what SWN Model 3 indicated.
+The coefficient for percent of private transportation changes signs.
+And the coefficient for median household income is now significant.
+
+The second model in Table \ref{tab:selection-results} (Selection Model 2) changes the way the variables are entered into the model.
+The log-transformation of density in SWN and Selection Model 1 assumes that the association between density and $R_0$ is monotonically increasing (if the sign of the coefficient is positive) or decreasing (if the sign of the coefficient is negative).
+There are some indications that the relationship may actually not be monotonical.
+For example, Paez et al.
+[-@Paez2020spatio] found a positive (if non-significant) relationship between density and incidence of COVID-19 in the provinces of Spain at the beginning of the pandemic.
+This changed to a negative (and significant) relationship during the lockdown.
+In the case of the US, Fielding-Miller et al.
+[-@Fielding2020social] found that the association between COVID-19 deaths and population density was positive in rural counties, but negative in urban counties.
+A variable transformation that allows for non-monotonic changes in the relationship is the square of the density.
+
+As seen in the table, Selection Model 2 replaces the log-transformation of population density with a quadratic expansion.
+The results of this analysis indicate that with this variable transformation, the selection and outcome processes are still not independent ($\rho\ne0$ with 5% of confidence).
+But a few interesting things emerge.
+When we examine the outcomes model, we see that the quadratic expansion has a positive coefficient for the first order term, but a negative coefficient for the second order term.
+This indicates that $R_0$ initially tends to increase with higher density, but only up to a point, after which the negative second term (which grows more rapidly due to the square), becomes increasingly dominant.
+Secondly, the sign of the coefficient for travel by private transportation becomes negative again.
+This, of course, makes more sense than the positive sign of Selection Model 1: if people tend to travel in private transportation, the potential for contact should be lower instead of higher.
+And finally median household income is no longer significant.
+
+```{r selection-variable, include=FALSE}
+# Create selection variable
+county_geo_clean <- county_geo_clean %>%
+  mutate(Ys = ifelse(R>0, TRUE, FALSE))
+```
+
+```{r estimate-selection-models, include=FALSE}
+# Selection model using SWN variables
+selection1 <- selection(Ys ~ density_log + private + hincome, R ~ density_log + private + hincome, data = county_geo_clean)
+#summary(selection1)
+
+# Selection model with different variable specification
+selection2 <- selection(Ys ~ density + density_2 + private + hincome, R ~ density + density_2 + private + hincome, data = county_geo_clean)
+#summary(selection2)
+
+# Checking SWN with different variable specification
+model3.2 <- lme(R ~ density + density_2 + private + hincome, 
+                random = ~ 1| state, 
+                data = county_geo_clean %>%
+                  filter(R > 0))
+#summary(model3.2)
+```
+
+```{r prepare-selection-results, include=FALSE}
+selection1_me <- selection1 %>%
+  summary()
+
+selection1_me <- selection1_me$estimate %>%
+  as.data.frame()
+
+colnames(selection1_me) <- c("b", "se", "t", "p")
+
+selection1_me <- selection1_me %>%  
+  rownames_to_column(var = "Variable") %>% 
+  transmute(Variable,
+            b,
+            ll = b - 1.96 * se, 
+            ul = b + 1.96 * se)
+
+selection2_me <- selection2 %>%
+  summary()
+
+selection2_me <- selection2_me$estimate %>%
+  as.data.frame()
+
+colnames(selection2_me) <- c("b", "se", "t", "p")
+
+selection2_me <- selection2_me %>%  
+  rownames_to_column(var = "Variable") %>% 
+  transmute(Variable,
+            b,
+            ll = b - 1.96 * se, 
+            ul = b + 1.96 * se)
+
+results_selection <- selection1_me %>%
+  full_join(selection2_me,
+            by = "Variable")
+
+colnames(results_selection) <- c("Variable",
+                                 "b1",
+                                 "l1",
+                                 "u1",
+                                 "b2",
+                                 "l2",
+                                 "u2")
+
+results_selection <- results_selection %>%
+  transmute(Order = c(1, 2, 5, 6, 7, 8, 11, 12, 13, 14, 3, 4, 9, 10),
+            Variable,
+            b1 = ifelse(is.na(b1),
+                        " ", 
+                        round(b1, 3)),
+            ci1 = ifelse(is.na(l1), 
+                         " ",
+                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")),
+            b2 = ifelse(is.na(b2),
+                        " ", 
+                        round(b2, 3)),
+            ci2 = ifelse(is.na(l2), 
+                         " ",
+                         paste0("[", round(l2, 3), ", ", round(u2, 3), "]"))) %>%
+  arrange(Order) %>%
+  select(-Order)
+```
+
+```{r tabulate-sample-selection-results, echo=FALSE}
+results_selection  %>%
+  mutate(Variable = c("Intercept",
+                      "Log of population density",
+                      "Density (1,000 per sq.km)",
+                      "Density squared",
+                      "Percent of private transportation", 
+                      "Median household income (10,000)",
+                      "Intercept",
+                      "Log of population density",
+                      "Density (1,000 per sq.km)",
+                      "Density squared",
+                      "Percent of private transportation", 
+                      "Median household income (\\$10,000)",
+                      "$\\sigma$",
+                      "$\\rho$")) %>%
+  kable("html",
+        digits = 3,
+        booktabs = TRUE,
+        escape = FALSE,
+        col.names = c("Variable",
+                      "$\\beta$", 
+                      "95\\% CI",
+                      "$\\beta$",
+                      "95\\% CI"),
+        align = c("lcccc"),
+        caption = "\\label{tab:selection-results}Estimation results of sample selection models") %>%
+  kable_styling(latex_options = c("scale_down")) %>% 
+  pack_rows(group_label = "Sample Selection Model", 1, 6) %>%
+  pack_rows(group_label = "Outcome Model", 7, 12) %>%
+  add_header_above(c(" ", "Selection Model 1" = 2, "Selection Model 2" = 2))
+```
+
+<!--
+If x is the density:
+$$
+\frac{dR}{dx} = 0.758 - 2 * 0.132 * density 
+$$
+
+Then, the maximum of this curve is when the density is:
+$$
+density = \frac{0.758}{2 * 0.132}
+$$
+-->
+
+How relevant is the difference between these different model specifications?
+Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 1 and Selection Model 2.
+The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1.
+The conclusion is that at higher densities, $R_0$ is _always_ higher.
+The right panel, in contrast, shows that, according to Selection Model 2, $R_0$ is zero when density is zero (as expected), and then it tends to increase at higher densities.
+This continues until a density of approximately 2.9 (1,000 people per sq.km).
+At higher densities than that $R_0$ begins to decline, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
+
+Thus, other things being equal, the effect of density in a county like Charlottesville in Virginia	 (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
+In contrast, the effect of density on $R_0$ in a county like Arlington in Virginia (density ~3,093 people per sq.km) is _stronger_ than either of the previous two examples.
+Lastly, the density of counties like San Francisco in California, or Queens and Bronx in NY, which are among the densest in the US, contributes even less to $R_0$ than even the most rural counties in the country.
+
+It is worth at this point to recall Cressie's dictum about modelling: "[w]hat is one person's mean structure could be another person's correlation structure" [@Cressie1989geostatistics, p.
+201].
+There are almost always multiple ways to approach a modelling situation.
+In the present case, we would argue that spatial sampling is an important aspect of the modeling process, but one that perhaps required different technical skills than those available to SWN.
+There is nothing wrong with that.
+What matters is that, by adopting relatively high reproducibility standards, these researchers made a valuable and honest contribution to the collective enterprise of seeking knowledge.
+Their effort, and subsequent efforts to validate and expand on their work, can potentially contribute to provide clarity to ongoing conversations about the relevance of density and the spread of COVID-19.
+
+In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [as put by Wong and Li, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading].
+At the same time, this also stands in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with in a longitudinal setting [@Hamidi2020longitudinal.
+In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al.
+[-@Fielding2020social] might be onto something with respect to the difference between rural and urban counties.
+More generally, in population-level studies, density is indicative of proximity, no doubt about that, but also for adaptive behavior.
+And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.
+
+```{r comparison-results, echo=FALSE, fig.cap="\\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2."}
+# The maximum in-sample density is 6.255358
+# county_geo_clean %>% st_drop_geometry() %>% filter(R > 0) %>% summary()
+
+example_df <- data.frame(density = seq(0, 6.3, 0.01)) %>% 
+  mutate(density_log = log(density),
+         density_2 = density^2, 
+         b_swn = model3$coefficients$fixed["density_log"] * density_log,
+         b_selection = selection2$estimate[7] * density + selection2$estimate[8] * density_2)
+
+# Plot of coefficient of density to R_0 according to SWN Model 3
+density_swn <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_swn)) +
+  ggtitle("SWN Model 3") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+# Plot of coefficient of density to R_0  
+density_selection <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_selection)) +
+  ggtitle("Selection Model 2") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+grid.arrange(density_swn, density_selection, nrow = 1)
+```
+
+```{r counties-by-density, include=FALSE}
+# the relationship between density and R_0 become negative in Selection Model 2 at about 5.74
+
+# Values of b_selection for a density similar a Philadelphia 
+example_df %>% filter(density > 4.10 & density < 4.14)
+# Densities with a similar b_selection
+example_df %>% filter(b_selection > 0.87 & b_selection < 0.89)
+
+# Counties with densities with similar response to Philadelphia
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 1.59 & density < 1.64) %>%
+  select(NAME.y, density)
+
+# Counties with densities associated with the highest response to density
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 2.70 & density < 3.10) %>%
+  select(NAME.y, density)
+
+# Highly dense counties
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 5.74) %>%
+  select(NAME.y)
+
+```
+
+<!--
+
+## Experiments with tobit and spatially autoregressive tobit
+
+## Fit tobit version of models
+
+```{r include=FALSE}
+#
+tobit1 <- censReg(R ~ density_log, 
+                  left = 0,
+                  data = county_geo_clean)
+
+#
+tobit2 <- censReg(R ~ density_log + private, 
+                  data = county_geo_clean)
+
+#
+tobit3 <- censReg(R ~ density_log + private + hincome, 
+                  data = county_geo_clean)
+```
+
+## Spatially autoregressive tobit
+
+Fit spatially autoregressive tobit:
+```{r include=FALSE}
+# Fit SAR Tobit
+fit_sartobit <- sartobit(R ~ density_log + private + hincome_log,
+                         B,
+                         ndraw = 1000,
+                         burn.in = 200, 
+                         showProgress = TRUE,
+                         data = county_geo_clean,
+                         computeMarginalEffects = TRUE)
+summary(fit_sartobit)
+impacts(fit_sartobit)
+```
+
+-->
+
+# Conclusion
+
+The tension between the need to publish research potentially useful in dealing with a global pandemic, and a "carnage of substandard research" [@Bramstedt2020carnage], highlights the importance of efforts to maintain the quality of scientific outputs during COVID-19.
+An important part of quality control is the ability of independent researchers to verify and examine the results of materials published in the literature.
+As previous research illustrates, reproducibility in scientific research remains an important but elusive goal [e.g., @Iqbal2016reproducible; @Stodden2018empirical; @Sumner2020reproducibility; @Gustot2020quality].
+This idea is reinforced by the review conducted for this paper in the context of research about population density and the spread of COVID-19.
+
+Taking one recent example from the literature [Sy et al., @Sy2021population; SWN], the present paper illustrates the importance of good reproducibility practices.
+Sharing data and code can catalyze research, by allowing independent verification of findings, as well as additional research.
+After verifying the results of SWN, experiments with sample selection models and variations in the definition of model inputs, lead to an important reappraisal of the conclusion that high density is associated with greater spread of the virus.
+Instead, the possibility of a non-monotonical relationship between population density and contagion is raised.
+
+In the spirit of openness, this paper is prepared as an R Markdown document, an a companion data package is provided.
+The data package contains the relevant documentation of the data, and all data pre-processing is fully documented.
+Hopefully this, and similar reproducible papers, will continue to encourage others to adopt reproducible standards in their research.
+
+# References {#references .unnumbered}

### 2021-05-25 R0-Density-Reanalysis/Cover-Letter-v1.Rmd
Paper reformatted for submission to Science of the Total Environment after PLoS ONE desk rejection

@@
 The paper illustrates the importance of reproducibility in research in the particular context of COVID-19.
-It takes a recent paper from the literature (Sy et al., 2021, published in PLoS ONE) and demonstrates how openly sharing data and code can accelerate research.
+Research on population density and the spread of COVID-19 has been published in Science of the Total Environment (see Sharifi and Khavarian-Garmsir, 2020; and Sun et al., 2020).
+
+The present submission includes some original research, since in it I reanalyze data from a recent paper (Sy et al., 2021, published in PLoS ONE) and show that after taking into account sample selection the conclusions are drastically different.
+I think, however, that this paper is better classified as a "Discussion", since the reanalysis is the motivation for a broader discussion about reproducibility in COVID-19 research, and an example of how openly sharing data and code can accelerate research.
 

### 2021-05-25 R0-Density-Reanalysis/R0-Density-Reanalysis-Title-Page.Rmd
Paper reformatted for submission to Science of the Total Environment after PLoS ONE desk rejection

---
title: "The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic"
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research. Given the high stakes of the situation, it is essential that scientific activities, on which good policy depends, are as transparent and reproducible as possible. Reproducibility is key for the efficient operation of the self-correction mechanisms of science, which work to weed out errors and refine our understanding of social and physical phenomena. In this paper, the importance of reproducibility is illustrated for the case of the association between population density and the the spread of SARS-CoV-2. Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example in this paper shows.

journal: "Science of the Total Environment"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
    - \usepackage{setspace}\doublespacing
---

### 2021-05-25 R0-Density-Reanalysis/R0-Density-Reanalysis-Title-Page.tex
Paper reformatted for submission to Science of the Total Environment after PLoS ONE desk rejection

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Science of the Total Environment} % Sets Journal name

\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc citation processing

% Pandoc header
\usepackage{setspace}\doublespacing

\begin{document}
\begin{frontmatter}

  \title{The importance of reproducibility in COVID-19 research: the
case of population density and the spread of the pandemic}
    \author[McMaster University]{Antonio Paez\corref{1}}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment and
Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  The emergence of the novel SARS-CoV-2 coronavirus and the global
  COVID-19 pandemic has led to explosive growth in scientific research.
  Given the high stakes of the situation, it is essential that
  scientific activities, on which good policy depends, are as
  transparent and reproducible as possible. Reproducibility is key for
  the efficient operation of the self-correction mechanisms of science,
  which work to weed out errors and refine our understanding of social
  and physical phenomena. In this paper, the importance of
  reproducibility is illustrated for the case of the association between
  population density and the the spread of SARS-CoV-2. Transparency and
  openness means that the same problem can, with relatively modest
  efforts, be examined by independent researchers who can verify
  findings, and bring to bear different perspectives, approaches, and
  methods---sometimes with consequential changes in the conclusions, as
  the empirical example in this paper shows.
  \end{abstract}

 \end{frontmatter}

\end{document}

### 2021-05-25 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Paper reformatted for submission to Science of the Total Environment after PLoS ONE desk rejection

@@
     affiliation: McMaster University
-    corresponding: paezha@mcmaster.ca
+    footnote: 1
 address:
@@
     address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
+footnote:
+  - code: 1
+    text: "Corresponding Author"
 abstract: |
   The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research.
-Given the high stakes of the situation, it is essential that scientific activites, on which good policy depends, are as transparent and reproducible as possible.
+Given the high stakes of the situation, it is essential that scientific activities, on which good policy depends, are as transparent and reproducible as possible.
 Reproducibility is key for the efficient operation of the self-correction mechanisms of science, which work to weed out errors and refine our understanding of social and physical phenomena.
 In this paper, the importance of reproducibility is illustrated for the case of the association between population density and the the spread of SARS-CoV-2.
-Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequantial changes in the conclusions, as the empirical example in this paper shows.
+Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example in this paper shows.
 
-author_summary: |
+journal: "Science of the Total Environment"
+date: "`r Sys.Date()`"
+bibliography: bibliography.bib
+linenumbers: true
+#numbersections: true
+csl: elsevier-harvard.csl
+output: rticles::elsevier_article
+header-includes:
+    - \usepackage{setspace}\doublespacing
+---
 
-bibliography: bibliography.bib
-output: rticles::plos_article
-csl: plos.csl
----
+\newpage
 
@@
 
-In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how].
+In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how]; the implications are the topic of ongoing research [@Sharifi2020covid].
 The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, brought a torrent of research into the associations between density and the spread of the pandemic.
@@
 To this end, a recent study by Sy et al.
-[@Sy2021population] is chosen as an example of reproducible research.
+[-@Sy2021population] is chosen as an example of reproducible research.
 The objective is not to malign the analysis of these researchers, but rather to demonstrate the value of openness to allow for independent verification and further analysis.
@@
 In theory, there are good reasons to believe that higher density may have a positive association with the transmission of a contagious virus.
-It has long been known that the potential for inter-personal contact is greater in regions with higher density [see for example the research on urban fields and time-geography @Moore1970urban; @Moore1970some; @Farber2011running].
+It has long been known that the potential for inter-personal contact is greater in regions with higher density [see for example the research on urban fields and time-geography, including @Farber2011running; @Moore1970urban; @Moore1970some].
 Mathematically, models of exposure and contagion indicate that higher densities can catalyze the transmission of contagious diseases [@Rocklov2020high; @Li2018effect].
@@
 Is it possible to spatially interpolate the missing observations?
-And, was there evidence of spatial autocorrelation in the residuals of the models?
 These questions are relevant and their implications important.
@@
 
-In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of a spatial modeler are asked, and the data are reanalyzed.
+In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of an independent researcher are asked, and the data are reanalyzed.
 

### 2021-05-25 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Paper reformatted for submission to Science of the Total Environment after PLoS ONE desk rejection

@@
-% Template for PLoS
-% Version 3.5 March 2018
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended
-% to minimize problems and delays during our production
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% Once your paper is accepted for publication,
-% PLEASE REMOVE ALL TRACKED CHANGES in this file
-% and leave only the final text of your manuscript.
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file.
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-%
-% Do not include text that is not math in the math environment.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column.
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-% \usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-%\usepackage{setspace}
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-% \bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-%\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-
-%% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
+\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
+%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
+\usepackage[hyphens]{url}
+
+  \journal{Science of the Total Environment} % Sets Journal name
+
+\usepackage{lineno} % add
+  \linenumbers % turns line numbering on
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{graphicx}
+\usepackage{booktabs} % book-quality tables
+%%%%%%%%%%%%%%%% end my additions to header
+
+\usepackage[T1]{fontenc}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \usepackage{fontspec}
+  \ifxetex
+    \usepackage{xltxtra,xunicode}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use microtype if available
+\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+\bibliographystyle{elsarticle-harv}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic},
+            colorlinks=false,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+
+\setcounter{secnumdepth}{0}
+% Pandoc toggle for numbering sections (defaults to be off)
+\setcounter{secnumdepth}{0}
 
@@
 
+% Pandoc header
+\usepackage{setspace}\doublespacing
 \usepackage{booktabs}
@@
 
-\usepackage{forarray}
-\usepackage{xstring}
-\newcommand{\getIndex}[2]{
-  \ForEach{,}{\IfEq{#1}{\thislevelitem}{\number\thislevelcount\ExitForEach}{}}{#2}
-}
-
-\setcounter{secnumdepth}{0}
-
-\newcommand{\getAff}[1]{
-  \getIndex{#1}{McMaster University}
-}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
 \begin{document}
-\vspace*{0.2in}
-
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{The importance of reproducibility in COVID-19 research:
-the case of population density and the spread of the
-pandemic} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Antonio Paez\textsuperscript{\getAff{McMaster
-University}}\textsuperscript{*}\\
-\bigskip
-\textbf{\getAff{McMaster University}}School of Earth, Environment and
-Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada\\
-\bigskip
-* Corresponding author: paezha@mcmaster.ca\\
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
+\begin{frontmatter}
+
+  \title{The importance of reproducibility in COVID-19 research: the
+case of population density and the spread of the pandemic}
+    \author[McMaster University]{Antonio Paez\corref{1}}
+   \ead{paezha@mcmaster.ca} 
+      \address[McMaster University]{School of Earth, Environment and
+Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada}
+      \cortext[1]{Corresponding Author}
+
+  \begin{abstract}
   The emergence of the novel SARS-CoV-2 coronavirus and the global
   COVID-19 pandemic has led to explosive growth in scientific research.
-Given the high stakes of the situation, it is essential that scientific
-activites, on which good policy depends, are as transparent and
-reproducible as possible. Reproducibility is key for the efficient
-operation of the self-correction mechanisms of science, which work to
-weed out errors and refine our understanding of social and physical
-phenomena. In this paper, the importance of reproducibility is
-illustrated for the case of the association between population density
-and the the spread of SARS-CoV-2. Transparency and openness means that
-the same problem can, with relatively modest efforts, be examined by
-independent researchers who can verify findings, and bring to bear
-different perspectives, approaches, and methods---sometimes with
-consequantial changes in the conclusions, as the empirical example in
-this paper shows.
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step.
-% Author Summary not valid for PLOS ONE submissions.
-
-\linenumbers
-
-% Use "Eq" instead of "Equation" for equation citations.
+  Given the high stakes of the situation, it is essential that
+  scientific activities, on which good policy depends, are as
+  transparent and reproducible as possible. Reproducibility is key for
+  the efficient operation of the self-correction mechanisms of science,
+  which work to weed out errors and refine our understanding of social
+  and physical phenomena. In this paper, the importance of
+  reproducibility is illustrated for the case of the association between
+  population density and the the spread of SARS-CoV-2. Transparency and
+  openness means that the same problem can, with relatively modest
+  efforts, be examined by independent researchers who can verify
+  findings, and bring to bear different perspectives, approaches, and
+  methods---sometimes with consequential changes in the conclusions, as
+  the empirical example in this paper shows.
+  \end{abstract}
+
+ \end{frontmatter}
+
+\newpage
+
 \hypertarget{introduction}{%
@@
 global pandemic that followed in its wake, led to an explosive growth of
-research around the globe. According to Fraser et al. {[}1{]}, over
+research around the globe. According to Fraser et al. (2021), over
 125,000 COVID-19-related papers were released in the first ten months
@@
 were shared in pre-print servers, the use of which also exploded in the
-past year {[}2--4{]}.
+past year (Añazco et al., 2021; Kwon, 2020; Vlasschaert et al., 2020).
 
@@
 this deluge of publications a ``carnage of substandard research''
-{[}5{]}. Part of the challenge of maintaining quality standards in
-published research is that, despite an abundance of recommendations and
-guidelines {[}6--9{]}, in practice reproducibility has remained a lofty
-and somewhat aspirational goal {[}10,11{]}. As reported in the
+(Bramstedt, 2020). Part of the challenge of maintaining quality
+standards in published research is that, despite an abundance of
+recommendations and guidelines (Broggini et al., 2017; Brunsdon and
+Comber, 2020; Ince et al., 2012; Ioannidis et al., 2014), in practice
+reproducibility has remained a lofty and somewhat aspirational goal
+(Konkol et al., 2019; Konkol and Kray, 2019). As reported in the
 literature, only a woefully small proportion of published research was
-actually reproducible before the pandemic {[}12,13{]}, and the situation
-does not appear to have changed substantially since {[}14,15{]}.
+actually reproducible before the pandemic (Iqbal et al., 2016; Stodden
+et al., 2018), and the situation does not appear to have changed
+substantially since (Gustot, 2020; Sumner et al., 2020).
 
@@
 life. The case of economists Reinhart and Rogoff comes to mind: by the
-time the inaccuracies and errors in their research were uncovered {[}see
-16{]}, their claims about debt and economic growth had already been
-seized by policy-makers on both sides of the Atlantic to justify
-austerity policies in the aftermath of the Great Recession of
+time the inaccuracies and errors in their research were uncovered (see
+Herndon et al., 2014), their claims about debt and economic growth had
+already been seized by policy-makers on both sides of the Atlantic to
+justify austerity policies in the aftermath of the Great Recession of
 2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
@@
 As later research has demonstrated, those policies cast a long shadow,
-and their sequels continued to be felt for years {[}17{]}.
+and their sequels continued to be felt for years (Basu et al., 2017).
 
@@
 numerous thinkers has been the prospect of life in cities after the
-pandemic {[}18{]}. The fact that the worst of the pandemic was initially
-felt in dense population centers such as Wuhan, Milan, Madrid, and New
-York, brought a torrent of research into the associations between
-density and the spread of the pandemic. Some important questions hang on
-the results of these research efforts. For example, are lower density
-regions safer from the pandemic? Are de-densification policies
-warranted, even if just in the short term? And in the longer term, will
-the risks of life in high density regions presage a flight from cities?
-Over the past year, numerous papers have sought to throw light into the
-underlying issue of density and the pandemic; nonetheless the results,
-as will be detailed next, remain mixed. Further, to complicate matters,
-precious few of these studies appear to be sufficiently open to support
-independent verification.
+pandemic (Florida et al., 2020); the implications are the topic of
+ongoing research (Sharifi and Khavarian-Garmsir, 2020). The fact that
+the worst of the pandemic was initially felt in dense population centers
+such as Wuhan, Milan, Madrid, and New York, brought a torrent of
+research into the associations between density and the spread of the
+pandemic. Some important questions hang on the results of these research
+efforts. For example, are lower density regions safer from the pandemic?
+Are de-densification policies warranted, even if just in the short term?
+And in the longer term, will the risks of life in high density regions
+presage a flight from cities? Over the past year, numerous papers have
+sought to throw light into the underlying issue of density and the
+pandemic; nonetheless the results, as will be detailed next, remain
+mixed. Further, to complicate matters, precious few of these studies
+appear to be sufficiently open to support independent verification.
 
@@
 reproducibility in research, particularly in the context of the flood of
-COVID-19 papers. To this end, a recent study by Sy et al. {[}19{]} is
+COVID-19 papers. To this end, a recent study by Sy et al. (2021) is
 chosen as an example of reproducible research. The objective is not to
@@
 potential for inter-personal contact is greater in regions with higher
-density {[}see for example the research on urban fields and
-time-geography 20,21,22{]}. Mathematically, models of exposure and
-contagion indicate that higher densities can catalyze the transmission
-of contagious diseases {[}23,24{]}. The idea is intuitive and likely at
-the root of messages, by some figures in positions of authority, that
-regions with sparse population densities faced lower risks from the
-pandemic\footnote{Governor Kristi Noem of South Dakota, for example,
-  claimed that sparse population density allowed her state to face the
-  pandemic down without the need for strict policy interventions
+density (see for example the research on urban fields and
+time-geography, including Farber and Páez, 2011; Moore, 1970; Moore and
+Brown, 1970). Mathematically, models of exposure and contagion indicate
+that higher densities can catalyze the transmission of contagious
+diseases (Li et al., 2018; Rocklöv and Sjödin, 2020). The idea is
+intuitive and likely at the root of messages, by some figures in
+positions of authority, that regions with sparse population densities
+faced lower risks from the pandemic\footnote{Governor Kristi Noem of
+  South Dakota, for example, claimed that sparse population density
+  allowed her state to face the pandemic down without the need for
+  strict policy interventions
   \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.
 
-As Rocklöv and Sjödin {[}23{]} note, however, mathematical models of
-contagion are valid at small-to-medium spatial scales (and presumably,
-small temporal scales too, such as time spent in restaurants, concert
-halls, cruises), and the results do not necessarily transfer to larger
-spatial units and different time scales. There are solid reasons for
-this: while in a restaurant, one can hardly avoid being in proximity to
-other customers-however, a person can choose to (or be forced to as a
-matter of policy) not go to a restaurant in the first place.
-Nonetheless, the idea that high density correlates with high
-transmission is so seemingly reasonable that it is often taken for
-granted even at larger scales {[}e.g., 25,26{]}. At larger scales,
-however, there exists the possibility of behavioral adaptations, which
-are difficult to capture in the mechanistic framework of differential
-equations {[}or can be missing in agent-based models, e.g., 27{]}; these
-adaptations, in fact, can be a key aspect of disease transmission.
+As Rocklöv and Sjödin (Rocklöv and Sjödin, 2020) note, however,
+mathematical models of contagion are valid at small-to-medium spatial
+scales (and presumably, small temporal scales too, such as time spent in
+restaurants, concert halls, cruises), and the results do not necessarily
+transfer to larger spatial units and different time scales. There are
+solid reasons for this: while in a restaurant, one can hardly avoid
+being in proximity to other customers-however, a person can choose to
+(or be forced to as a matter of policy) not go to a restaurant in the
+first place. Nonetheless, the idea that high density correlates with
+high transmission is so seemingly reasonable that it is often taken for
+granted even at larger scales (e.g., Cruz et al., 2020; Micallef et al.,
+2020). At larger scales, however, there exists the possibility of
+behavioral adaptations, which are difficult to capture in the
+mechanistic framework of differential equations (or can be missing in
+agent-based models, e.g., Gomez et al., 2021); these adaptations, in
+fact, can be a key aspect of disease transmission.
 
@@
 Risk compensation is a process whereby people adjust their behavior in
-response to their \emph{perception} of risk {[}28--30{]}. In the case of
-COVID-19, Chauhan et al. {[}31{]} have found that perception of risks in
-the US varies between rural, suburban, and urban residents, with rural
+response to their \emph{perception} of risk (Noland, 1995; Phillips et
+al., 2011; Richens et al., 2000). In the case of COVID-19, Chauhan et
+al. (Chauhan et al., 2021) have found that perception of risks in the US
+varies between rural, suburban, and urban residents, with rural
 residents in general expressing less concern about the virus. It is
@@
 directly observe the impact of the pandemic may have become overly
-cautious. Both Paez et al. {[}32{]} and Hamidi et al. {[}33{]} posit
-this mechanism (i.e., greater compliance with social distancing in
-denser regions) to explain the results of their analyses. The evidence
+cautious. Both Paez et al. (2020) and Hamidi et al. (2020b) posit this
+mechanism (i.e., greater compliance with social distancing in denser
+regions) to explain the results of their analyses. The evidence
 available does indeed show that there were important changes in behavior
-with respect to mobility during the pandemic {[}34--36{]}; furthermore,
-shelter in place orders may have had greater buy-in from the public in
-higher density regions {[}37{]}, and the associated behavior may have
-persisted beyond the duration of official social-distancing policies
-{[}38{]}. In addition, there is evidence that changes in mobility
-correlated with the trajectory of the pandemic {[}39,40{]}. Given the
-potential for behavioral adaptation, the question of density becomes
-more nuanced: it is not just a matter of proximity, but also of human
-behavior, which is better studied using population-level data and
-models.
+with respect to mobility during the pandemic (Harris and Branion-Calles,
+2021; Jamal and Paez, 2020; Molloy et al., 2020); furthermore, shelter
+in place orders may have had greater buy-in from the public in higher
+density regions (Feyman et al., 2020), and the associated behavior may
+have persisted beyond the duration of official social-distancing
+policies (Praharaj et al., 2020). In addition, there is evidence that
+changes in mobility correlated with the trajectory of the pandemic
+(Noland, 2021; Paez, 2020). Given the potential for behavioral
+adaptation, the question of density becomes more nuanced: it is not just
+a matter of proximity, but also of human behavior, which is better
+studied using population-level data and models.
 
@@
 between population density and various COVID-19-related outcomes. Bhadra
-{[}41{]}, for example, reported a moderate positive correlation between
+(2021), for example, reported a moderate positive correlation between
 the spread of COVID-19 and population density at the district level in
 India, however their analysis was bivariate and did not control for
-other variables, such as income. Similarly, Kadi and Khelfaoui {[}42{]}
+other variables, such as income. Similarly, Kadi and Khelfaoui (2020)
 found a positive and significant correlation between number of cases and
@@
 simple analyses is whether density is not a proxy for other factors.
-Other studies have included controls, such as Pequeno et al. {[}43{]}, a
+Other studies have included controls, such as Pequeno et al. (2020), a
 team that reported a positive association between density and cumulative
@@
 controlling for covariates, including income, transport connectivity,
-and economic status. In a similar vein, Fielding-Miller et al. {[}44{]}
+and economic status. In a similar vein, Fielding-Miller et al. (2020)
 reported a positive relationship between the absolute number of COVID-19
 deaths and population density (rate) in rural counties in the US. Roy
-and Ghosh {[}45{]} used a battery of machine learning techniques to find
+and Ghosh (2020) used a battery of machine learning techniques to find
 discriminatory factors, and a positive and significant association
 between COVID-19 infection and death rates in US states. Wong and Li
-{[}46{]} also found a positive and significant association between
+(2020) also found a positive and significant association between
 population density and number of confirmed COVID-19 cases in US
 counties, using both univariate and multivariate regressions with
-spatial effects. More recently, Sy et al. {[}19{]} reported that the
-basic reproductive number of COVID-19 in US counties tended to increase
-with population density, but at a decreasing rate at higher densities.
+spatial effects. More recently, Sy et al. (2021) reported that the basic
+reproductive number of COVID-19 in US counties tended to increase with
+population density, but at a decreasing rate at higher densities.
 
@@
 associations between population density and COVID-19 outcomes. This
-includes the research of Sun et al. {[}47{]} who did not find evidence
-of significant correlation between population density and confirmed
-number of cases per day \emph{in conditions of lockdown} in China. This
-finding echoes the results of Paez et al. {[}32{]}, who in their study
-of provinces in Spain reported non-significant associations between
+includes the research of Sun et al. (2020) who did not find evidence of
+significant correlation between population density and confirmed number
+of cases per day \emph{in conditions of lockdown} in China. This finding
+echoes the results of Paez et al. (2020), who in their study of
+provinces in Spain reported non-significant associations between
 population density and infection rates in the early days of the first
 wave of COVID-19, and negative significant associations in the later
-part of the first lockdown. Similarly, {[}48{]} found zero or negative
+part of the first lockdown. Similarly, (2020) found zero or negative
 associations between population density and infection numbers/deaths by
-country. Fielding-Miller et al. {[}44{]} contrast their finding about
+country. Fielding-Miller et al. (2020) contrast their finding about
 rural counties with a negative relationship between COVID-19 deaths and
 population density in urban counties in the US. For their part, in their
-investigation of doubling time, White and Hébert-Dufresne {[}49{]}
+investigation of doubling time, White and Hébert-Dufresne (2020)
 identified a negative and significant correlation between population
-density and doubling time in US states. Likewise, {[}50{]} fond a small
+density and doubling time in US states. Likewise, (2021) fond a small
 negative (and significant) association between population density and
 COVID-19 morbidity in districts in Tehran. Finally, two of the most
-complete studies in the US {[}by Hamidi et al.; {[}51{]}; {[}33{]}{]}
-used an extensive set of controls to find negative and significant
+complete studies in the US {[}by Hamidi et al.; (2020a); (2020b){]} used
+an extensive set of controls to find negative and significant
 correlations between density and COVID-19 cases and fatalities at the
@@
 be further elaborated in later sections of this paper), the study of Sy
-et al.~{[}{[}19{]}; hereafter SWN{]} would immediately grab the
-attention of a researcher with a somewhat different toolbox.
+et al.~{[}(2021); hereafter SWN{]} would immediately grab the attention
+of a researcher with a somewhat different toolbox.
 
@@
 analysis? Is it possible to spatially interpolate the missing
-observations? And, was there evidence of spatial autocorrelation in the
-residuals of the models? These questions are relevant and their
-implications important. Fortunately, SWN are an example of a reasonably
-open, reproducible research product: their paper is accompanied by (most
-of) the data and (most of) the code used in the analysis. This means
-that an independent expert can, with only a moderate investment of time
-and effort, reproduce the results in the paper, as well as ask
-additional questions.
+observations? These questions are relevant and their implications
+important. Fortunately, SWN are an example of a reasonably open,
+reproducible research product: their paper is accompanied by (most of)
+the data and (most of) the code used in the analysis. This means that an
+independent expert can, with only a moderate investment of time and
+effort, reproduce the results in the paper, as well as ask additional
+questions.
 
@@
 some cases, there might be legitimate reasons to withhold the data,
-perhaps due to confidentiality and privacy reasons {[}e.g., 52{]}. But
-in many other cases the data are publicly available, which in fact has
-commonly been the case with population-level COVID-19 information.
-Typically the provenance of the data is documented, but in numerous
-studies the data themselves are not shared {[}25,33,41,44,51,53--56{]}.
-As any researcher can attest, whether a graduate student or a seasoned
-scientist, collecting, organizing, and preparing data for a project can
-take a substantial amount of time. Pointing to the sources of data, even
-when these sources are public, is a small step towards
-reproducibility-but only a very small one. Faced with the prospect of
-having to recreate a data set from raw sources is probably sufficient to
-dissuade all but the most dedicated (or stubborn) researcher from
-independent verification. This is true even if part of the data are
-shared {[}e.g., 46{]}. In other cases, data are shared, but the
-processes followed in the preparation of the data are not fully
-documented {[}48,57{]}. These processes matter, as shown by the errors
-in the spreadsheets of Reinhart and Rogoff {[}16{]} and the data of
-biologist Jonathan Pruitt that led to an ``avalanche'' of paper
+perhaps due to confidentiality and privacy reasons (e.g., Lee et al.,
+2020). But in many other cases the data are publicly available, which in
+fact has commonly been the case with population-level COVID-19
+information. Typically the provenance of the data is documented, but in
+numerous studies the data themselves are not shared (Amadu et al., 2021;
+Bhadra et al., 2021; Cruz et al., 2020; Feng et al., 2020;
+Fielding-Miller et al., 2020; Hamidi et al., 2020a, 2020b; Inbaraj et
+al., 2021; Souris and Gonzalez, 2020). As any researcher can attest,
+whether a graduate student or a seasoned scientist, collecting,
+organizing, and preparing data for a project can take a substantial
+amount of time. Pointing to the sources of data, even when these sources
+are public, is a small step towards reproducibility-but only a very
+small one. Faced with the prospect of having to recreate a data set from
+raw sources is probably sufficient to dissuade all but the most
+dedicated (or stubborn) researcher from independent verification. This
+is true even if part of the data are shared (e.g., Wong and Li, 2020).
+In other cases, data are shared, but the processes followed in the
+preparation of the data are not fully documented (Ahmad et al., 2020;
+Skórka et al., 2020). These processes matter, as shown by the errors in
+the spreadsheets of Reinhart and Rogoff (Herndon et al., 2014) and the
+data of biologist Jonathan Pruitt that led to an ``avalanche'' of paper
 retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
 Another situation is when papers share well-documented data, but fail to
-provide the code used in the analysis {[}43,58,59{]}. Making code
-available only ``on demand'' {[}e.g., 60{]} is an unnecessary barrier
-when most journals offer the facility to share supplemental materials
-online. Then there are those papers that more closely comply with
-reproducibility standards, and share well-documented processes and data,
-as well as the code used in any analyses reported {[}19,32,37,49,61{]}.
+provide the code used in the analysis (Noury et al., 2021; Pequeno et
+al., 2020; Wang et al., 2021). Making code available only ``on demand''
+(e.g., Brandtner et al., 2021) is an unnecessary barrier when most
+journals offer the facility to share supplemental materials online. Then
+there are those papers that more closely comply with reproducibility
+standards, and share well-documented processes and data, as well as the
+code used in any analyses reported (Feyman et al., 2020; Paez et al.,
+2020; Stephens et al., 2021; Sy et al., 2021; White and Hébert-Dufresne,
+2020).
 
 In the following sections, the analysis of RWN is reproduced, some
-relevant questions from the perspective of a spatial modeler are asked,
-and the data are reanalyzed.
+relevant questions from the perspective of an independent researcher are
+asked, and the data are reanalyzed.
 
@@
 developed over time to address this. A model useful for sample selection
-problems is Heckman's selection model {[}see 62{]}. The selection model
-is in fact a system of two equations, as follows:
+problems is Heckman's selection model (see Maddala, 1983). The selection
+model is in fact a system of two equations, as follows:
 
@@
 relationship may actually not be monotonical. For example, Paez et al.
-{[}32{]} found a positive (if non-significant) relationship between
+(2020) found a positive (if non-significant) relationship between
 density and incidence of COVID-19 in the provinces of Spain at the
@@
 relationship during the lockdown. In the case of the US, Fielding-Miller
-et al. {[}44{]} found that the association between COVID-19 deaths and
+et al. (2020) found that the association between COVID-19 deaths and
 population density was positive in rural counties, but negative in urban
@@
 ``{[}w{]}hat is one person's mean structure could be another person's
-correlation structure'' {[}63{]}. There are almost always multiple ways
-to approach a modelling situation. In the present case, we would argue
-that spatial sampling is an important aspect of the modeling process,
-but one that perhaps required different technical skills than those
-available to SWN. There is nothing wrong with that. What matters is
-that, by adopting relatively high reproducibility standards, these
+correlation structure'' (Cressie, 1989, p. 201). There are almost always
+multiple ways to approach a modelling situation. In the present case, we
+would argue that spatial sampling is an important aspect of the modeling
+process, but one that perhaps required different technical skills than
+those available to SWN. There is nothing wrong with that. What matters
+is that, by adopting relatively high reproducibility standards, these
 researchers made a valuable and honest contribution to the collective
@@
 spread of the virus {[}as put by Wong and Li, ``\,`Density is destiny'
-is probably an overstatement''; {[}46{]}{]}. At the same time, this also
+is probably an overstatement''; (2020){]}. At the same time, this also
 stands in contrast to the findings of Hamidi et al., who found that
 higher density was either not significantly associated with the rate of
-the virus in a cross-sectional study {[}33{]}, or was negatively
-associated with in a longitudinal setting {[}{[}51{]}. In this sense,
-the conclusion that density does not aggravate the pandemic may have
-been somewhat premature; instead, reanalysis of the data of SWN suggests
-that Fielding-Miller et al. {[}44{]} might be onto something with
-respect to the difference between rural and urban counties. More
-generally, in population-level studies, density is indicative of
-proximity, no doubt about that, but also for adaptive behavior. And it
-is possible that the determining factor during COVID-19, at least in the
-US, has been variations in perceptions of the risks associated with
-contagion {[}31{]}, and subsequent compensations in behavior in more and
-less dense regions.
+the virus in a cross-sectional study (Hamidi et al., 2020b), or was
+negatively associated with in a longitudinal setting {[}Hamidi et al.
+(2020a). In this sense, the conclusion that density does not aggravate
+the pandemic may have been somewhat premature; instead, reanalysis of
+the data of SWN suggests that Fielding-Miller et al. (2020) might be
+onto something with respect to the difference between rural and urban
+counties. More generally, in population-level studies, density is
+indicative of proximity, no doubt about that, but also for adaptive
+behavior. And it is possible that the determining factor during
+COVID-19, at least in the US, has been variations in perceptions of the
+risks associated with contagion (Chauhan et al., 2021), and subsequent
+compensations in behavior in more and less dense regions.
 
@@
 dealing with a global pandemic, and a ``carnage of substandard
-research'' {[}5{]}, highlights the importance of efforts to maintain the
-quality of scientific outputs during COVID-19. An important part of
-quality control is the ability of independent researchers to verify and
-examine the results of materials published in the literature. As
-previous research illustrates, reproducibility in scientific research
-remains an important but elusive goal {[}e.g., 12,13--15{]}. This idea
-is reinforced by the review conducted for this paper in the context of
+research'' (Bramstedt, 2020), highlights the importance of efforts to
+maintain the quality of scientific outputs during COVID-19. An important
+part of quality control is the ability of independent researchers to
+verify and examine the results of materials published in the literature.
+As previous research illustrates, reproducibility in scientific research
+remains an important but elusive goal (Gustot, 2020; e.g., Iqbal et al.,
+2016; Stodden et al., 2018; Sumner et al., 2020). This idea is
+reinforced by the review conducted for this paper in the context of
 research about population density and the spread of COVID-19.
 
-Taking one recent example from the literature {[}Sy et al., {[}19{]};
-SWN{]}, the present paper illustrates the importance of good
+Taking one recent example from the literature {[}Sy et al., Sy et al.
+(2021); SWN{]}, the present paper illustrates the importance of good
 reproducibility practices. Sharing data and code can catalyze research,
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-Ahmad2020association}{}%
+Ahmad, K., Erqou, S., Shah, N., Nazir, U., Morrison, A.R., Choudhary,
+G., Wu, W.-C., 2020. Association of poor housing conditions with
+COVID-19 incidence and mortality across US counties. PLOS ONE 15,
+e0241327.
+doi:\href{https://doi.org/10.1371/journal.pone.0241327}{10.1371/journal.pone.0241327}
+
+\leavevmode\hypertarget{ref-Amadu2021assessing}{}%
+Amadu, I., Ahinkorah, B.O., Afitiri, A.-R., Seidu, A.-A., Ameyaw, E.K.,
+Hagan, J.E., Duku, E., Aram, S.A., 2021. Assessing sub-regional-specific
+strengths of healthcare systems associated with COVID-19 prevalence,
+deaths and recoveries in africa. PLOS ONE 16, e0247274.
+doi:\href{https://doi.org/10.1371/journal.pone.0247274}{10.1371/journal.pone.0247274}
+
+\leavevmode\hypertarget{ref-Anazco2021publication}{}%
+Añazco, D., Nicolalde, B., Espinosa, I., Camacho, J., Mushtaq, M.,
+Gimenez, J., Teran, E., 2021. Publication rate and citation counts for
+preprints released during the COVID-19 pandemic: The good, the bad and
+the ugly. PeerJ 9, e10927.
+doi:\href{https://doi.org/10.7717/peerj.10927}{10.7717/peerj.10927}
+
+\leavevmode\hypertarget{ref-Basu2017ten}{}%
+Basu, S., Carney, M.A., Kenworthy, N.J., 2017. Ten years after the
+financial crisis: The long reach of austerity and its global impacts on
+health. Social Science \& Medicine 187, 203--207.
+doi:\href{https://doi.org/10.1016/j.socscimed.2017.06.026}{10.1016/j.socscimed.2017.06.026}
+
+\leavevmode\hypertarget{ref-Bhadra2021impact}{}%
+Bhadra, A., Mukherjee, A., Sarkar, K., 2021. Impact of population
+density on covid-19 infected and mortality rate in india. Modeling Earth
+Systems and Environment 7, 623--629.
+doi:\href{https://doi.org/10.1007/s40808-020-00984-7}{10.1007/s40808-020-00984-7}
+
+\leavevmode\hypertarget{ref-Bramstedt2020carnage}{}%
+Bramstedt, K.A., 2020. The carnage of substandard research during the
+COVID-19 pandemic: A call for quality. Journal of Medical Ethics 46,
+803--807.
+doi:\href{https://doi.org/10.1136/medethics-2020-106494}{10.1136/medethics-2020-106494}
+
+\leavevmode\hypertarget{ref-Brandtner2021creatures}{}%
+Brandtner, C., Bettencourt, L.M.A., Berman, M.G., Stier, A.J., 2021.
+Creatures of the state? Metropolitan counties compensated for state
+inaction in initial u.s. Response to COVID-19 pandemic. PLOS ONE 16,
+e0246249.
+doi:\href{https://doi.org/10.1371/journal.pone.0246249}{10.1371/journal.pone.0246249}
+
+\leavevmode\hypertarget{ref-Broggini2017reproducible}{}%
+Broggini, F., Dellinger, J., Fomel, S., Liu, Y., 2017. Reproducible
+research: Geophysics papers of the future - introduction. Geophysics 82.
+doi:\href{https://doi.org/10.1190/geo2017-0918-spseintro.1}{10.1190/geo2017-0918-spseintro.1}
+
+\leavevmode\hypertarget{ref-Brunsdon2020opening}{}%
+Brunsdon, C., Comber, A., 2020. Opening practice: Supporting
+reproducibility and critical spatial data science. Journal of
+Geographical Systems.
+doi:\href{https://doi.org/10.1007/s10109-020-00334-2}{10.1007/s10109-020-00334-2}
+
+\leavevmode\hypertarget{ref-Chauhan2021covid}{}%
+Chauhan, R.S., Capasso da Silva, D., Salon, D., Shamshiripour, A.,
+Rahimi, E., Sutradhar, U., Khoeini, S., Mohammadian, A.(Kouros).,
+Derrible, S., Pendyala, R., 2021. COVID-19 related attitudes and risk
+perceptions across urban, rural, and suburban areas in the united
+states. Findings.
+doi:\href{https://doi.org/10.32866/001c.23714}{10.32866/001c.23714}
+
+\leavevmode\hypertarget{ref-Cressie1989geostatistics}{}%
+Cressie, N., 1989. Geostatistics. The American Statistician 43, 197.
+doi:\href{https://doi.org/10.2307/2685361}{10.2307/2685361}
+
+\leavevmode\hypertarget{ref-Cruz2020exploring}{}%
+Cruz, C.J.P., Ganly, R., Li, Z., Gietel-Basten, S., 2020. Exploring the
+young demographic profile of COVID-19 cases in hong kong: Evidence from
+migration and travel history data. PLOS ONE 15, e0235306.
+doi:\href{https://doi.org/10.1371/journal.pone.0235306}{10.1371/journal.pone.0235306}
+
+\leavevmode\hypertarget{ref-Farber2011running}{}%
+Farber, S., Páez, A., 2011. Running to stay in place: The time-use
+implications of automobile oriented land-use and travel. Journal of
+Transport Geography 19, 782--793.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2010.09.008}{10.1016/j.jtrangeo.2010.09.008}
+
+\leavevmode\hypertarget{ref-Feng2020spread}{}%
+Feng, Y., Li, Q., Tong, X., Wang, R., Zhai, S., Gao, C., Lei, Z., Chen,
+S., Zhou, Y., Wang, J., Yan, X., Xie, H., Chen, P., Liu, S., Xv, X.,
+Liu, S., Jin, Y., Wang, C., Hong, Z., Luan, K., Wei, C., Xu, J., Jiang,
+H., Xiao, C., Guo, Y., 2020. Spatiotemporal spread pattern of the
+COVID-19 cases in china. PLOS ONE 15, e0244351.
+doi:\href{https://doi.org/10.1371/journal.pone.0244351}{10.1371/journal.pone.0244351}
+
+\leavevmode\hypertarget{ref-Feyman2020effectiveness}{}%
+Feyman, Y., Bor, J., Raifman, J., Griffith, K.N., 2020. Effectiveness of
+COVID-19 shelter-in-place orders varied by state. PLOS ONE 15, e0245008.
+doi:\href{https://doi.org/10.1371/journal.pone.0245008}{10.1371/journal.pone.0245008}
+
+\leavevmode\hypertarget{ref-Fielding2020social}{}%
+Fielding-Miller, R.K., Sundaram, M.E., Brouwer, K., 2020. Social
+determinants of COVID-19 mortality at the county level. PLOS ONE 15,
+e0240151.
+doi:\href{https://doi.org/10.1371/journal.pone.0240151}{10.1371/journal.pone.0240151}
+
+\leavevmode\hypertarget{ref-Florida2020how}{}%
+Florida, R., Glaeser, E., Sharif, M., Bedi, K., Campanella, T., Chee,
+C., Doctoroff, D., Katz, B., Katz, R., Kotkin, J., 2020. How life in our
+cities will look after the coronavirus pandemic. Foreign Policy 1.
+
 \leavevmode\hypertarget{ref-Fraser2021evolving}{}%
-\CSLLeftMargin{1. }
-\CSLRightInline{Fraser N, Brierley L, Dey G, Polka JK, Pálfy M, Nanni F,
-et al. The evolving role of preprints in the dissemination of COVID-19
-research and their impact on the science communication landscape. PLOS
-Biology. 2021;19: e3000959.
-doi:\href{https://doi.org/10.1371/journal.pbio.3000959}{10.1371/journal.pbio.3000959}}
+Fraser, N., Brierley, L., Dey, G., Polka, J.K., Pálfy, M., Nanni, F.,
+Coates, J.A., 2021. The evolving role of preprints in the dissemination
+of COVID-19 research and their impact on the science communication
+landscape. PLOS Biology 19, e3000959.
+doi:\href{https://doi.org/10.1371/journal.pbio.3000959}{10.1371/journal.pbio.3000959}
+
+\leavevmode\hypertarget{ref-Gomez2021infekta}{}%
+Gomez, J., Prieto, J., Leon, E., Rodríguez, A., 2021. INFEKTA---an
+agent-based model for transmission of infectious diseases: The COVID-19
+case in bogotá, colombia. PLOS ONE 16, e0245787.
+doi:\href{https://doi.org/10.1371/journal.pone.0245787}{10.1371/journal.pone.0245787}
+
+\leavevmode\hypertarget{ref-Gustot2020quality}{}%
+Gustot, T., 2020. Quality and reproducibility during the COVID-19
+pandemic. JHEP Rep 2, 100141.
+doi:\href{https://doi.org/10.1016/j.jhepr.2020.100141}{10.1016/j.jhepr.2020.100141}
+
+\leavevmode\hypertarget{ref-Hamidi2020longitudinal}{}%
+Hamidi, S., Ewing, R., Sabouri, S., 2020a. Longitudinal analyses of the
+relationship between development density and the COVID-19 morbidity and
+mortality rates: Early evidence from 1,165 metropolitan counties in the
+united states. Health \& Place 64, 102378.
+doi:\href{https://doi.org/10.1016/j.healthplace.2020.102378}{10.1016/j.healthplace.2020.102378}
+
+\leavevmode\hypertarget{ref-Hamidi2020density}{}%
+Hamidi, S., Sabouri, S., Ewing, R., 2020b. Does density aggravate the
+COVID-19 pandemic? Journal of the American Planning Association 86,
+495--509.
+doi:\href{https://doi.org/10.1080/01944363.2020.1777891}{10.1080/01944363.2020.1777891}
+
+\leavevmode\hypertarget{ref-Harris2021Changes}{}%
+Harris, M.A., Branion-Calles, M., 2021. Changes in commute mode
+attributed to COVID-19 risk in canadian national survey data. Findings.
+doi:\href{https://doi.org/10.32866/001c.19088}{10.32866/001c.19088}
+
+\leavevmode\hypertarget{ref-Herndon2014high}{}%
+Herndon, T., Ash, M., Pollin, R., 2014. Does high public debt
+consistently stifle economic growth? A critique of reinhart and rogoff.
+Cambridge Journal of Economics 38, 257--279.
+doi:\href{https://doi.org/10.1093/cje/bet075}{10.1093/cje/bet075}
+
+\leavevmode\hypertarget{ref-Inbaraj2021seroprevalence}{}%
+Inbaraj, L.R., George, C.E., Chandrasingh, S., 2021. Seroprevalence of
+COVID-19 infection in a rural district of south india: A
+population-based seroepidemiological study. PLOS ONE 16, e0249247.
+doi:\href{https://doi.org/10.1371/journal.pone.0249247}{10.1371/journal.pone.0249247}
+
+\leavevmode\hypertarget{ref-Ince2012case}{}%
+Ince, D.C., Hatton, L., Graham-Cumming, J., 2012. The case for open
+computer programs. Nature 482, 485--488.
+doi:\href{https://doi.org/10.1038/nature10836}{10.1038/nature10836}
+
+\leavevmode\hypertarget{ref-Ioannidis2014increasing}{}%
+Ioannidis, J.P.A., Greenland, S., Hlatky, M.A., Khoury, M.J., Macleod,
+M.R., Moher, D., Schulz, K.F., Tibshirani, R., 2014. Increasing value
+and reducing waste in research design, conduct, and analysis. Lancet
+383, 166--175.
+doi:\href{https://doi.org/10.1016/s0140-6736(13)62227-8}{10.1016/s0140-6736(13)62227-8}
+
+\leavevmode\hypertarget{ref-Iqbal2016reproducible}{}%
+Iqbal, S.A., Wallach, J.D., Khoury, M.J., Schully, S.D., Ioannidis,
+J.P.A., 2016. Reproducible research practices and transparency across
+the biomedical literature. Plos Biology 14.
+doi:\href{https://doi.org/10.1371/journal.pbio.1002333}{10.1371/journal.pbio.1002333}
+
+\leavevmode\hypertarget{ref-Jamal2020Changes}{}%
+Jamal, S., Paez, A., 2020. Changes in trip-making frequency by mode
+during COVID-19. Findings.
+doi:\href{https://doi.org/10.32866/001c.17977}{10.32866/001c.17977}
+
+\leavevmode\hypertarget{ref-Kadi2020population}{}%
+Kadi, N., Khelfaoui, M., 2020. Population density, a factor in the
+spread of COVID-19 in algeria: Statistic study. Bulletin of the National
+Research Centre 44.
+doi:\href{https://doi.org/10.1186/s42269-020-00393-x}{10.1186/s42269-020-00393-x}
+
+\leavevmode\hypertarget{ref-Khavarian2021high}{}%
+Khavarian-Garmsir, A.R., Sharifi, A., Moradpour, N., 2021. Are
+high-density districts more vulnerable to the COVID-19 pandemic?
+Sustainable Cities and Society 70, 102911.
+doi:\href{https://doi.org/10.1016/j.scs.2021.102911}{10.1016/j.scs.2021.102911}
+
+\leavevmode\hypertarget{ref-Konkol2019examination}{}%
+Konkol, M., Kray, C., 2019. In-depth examination of spatiotemporal
+figures in open reproducible research. Cartography and Geographic
+Information Science 46, 412--427.
+doi:\href{https://doi.org/10.1080/15230406.2018.1512421}{10.1080/15230406.2018.1512421}
+
+\leavevmode\hypertarget{ref-Konkol2019computational}{}%
+Konkol, M., Kray, C., Pfeiffer, M., 2019. Computational reproducibility
+in geoscientific papers: Insights from a series of studies with
+geoscientists and a reproduction study. International Journal of
+Geographical Information Science 33, 408--429.
+doi:\href{https://doi.org/10.1080/13658816.2018.1508687}{10.1080/13658816.2018.1508687}
 
 \leavevmode\hypertarget{ref-Kwon2021swamped}{}%
-\CSLLeftMargin{2. }
-\CSLRightInline{Kwon D. How swamped preprint servers are blocking bad
-coronavirus research. Nature. 2020;581: 130--132. }
-
-\leavevmode\hypertarget{ref-Vlasschaert2020proliferation}{}%
-\CSLLeftMargin{3. }
-\CSLRightInline{Vlasschaert C, Topf JM, Hiremath S. Proliferation of
-papers and preprints during the coronavirus disease 2019 pandemic:
-Progress or problems with peer review? Advances in Chronic Kidney
-Disease. 2020;27: 418--426.
-doi:\href{https://doi.org/10.1053/j.ackd.2020.08.003}{10.1053/j.ackd.2020.08.003}}
-
-\leavevmode\hypertarget{ref-Anazco2021publication}{}%
-\CSLLeftMargin{4. }
-\CSLRightInline{Añazco D, Nicolalde B, Espinosa I, Camacho J, Mushtaq M,
-Gimenez J, et al. Publication rate and citation counts for preprints
-released during the COVID-19 pandemic: The good, the bad and the ugly.
-PeerJ. 2021;9: e10927.
-doi:\href{https://doi.org/10.7717/peerj.10927}{10.7717/peerj.10927}}
-
-\leavevmode\hypertarget{ref-Bramstedt2020carnage}{}%
-\CSLLeftMargin{5. }
-\CSLRightInline{Bramstedt KA. The carnage of substandard research during
-the COVID-19 pandemic: A call for quality. Journal of Medical Ethics.
-2020;46: 803--807.
-doi:\href{https://doi.org/10.1136/medethics-2020-106494}{10.1136/medethics-2020-106494}}
-
-\leavevmode\hypertarget{ref-Broggini2017reproducible}{}%
-\CSLLeftMargin{6. }
-\CSLRightInline{Broggini F, Dellinger J, Fomel S, Liu Y. Reproducible
-research: Geophysics papers of the future - introduction. Geophysics.
-2017;82.
-doi:\href{https://doi.org/10.1190/geo2017-0918-spseintro.1}{10.1190/geo2017-0918-spseintro.1}}
-
-\leavevmode\hypertarget{ref-Ince2012case}{}%
-\CSLLeftMargin{7. }
-\CSLRightInline{Ince DC, Hatton L, Graham-Cumming J. The case for open
-computer programs. Nature. 2012;482: 485--488.
-doi:\href{https://doi.org/10.1038/nature10836}{10.1038/nature10836}}
-
-\leavevmode\hypertarget{ref-Ioannidis2014increasing}{}%
-\CSLLeftMargin{8. }
-\CSLRightInline{Ioannidis JPA, Greenland S, Hlatky MA, Khoury MJ,
-Macleod MR, Moher D, et al. Increasing value and reducing waste in
-research design, conduct, and analysis. Lancet. 2014;383: 166--175.
-doi:\href{https://doi.org/10.1016/s0140-6736(13)62227-8}{10.1016/s0140-6736(13)62227-8}}
-
-\leavevmode\hypertarget{ref-Brunsdon2020opening}{}%
-\CSLLeftMargin{9. }
-\CSLRightInline{Brunsdon C, Comber A. Opening practice: Supporting
-reproducibility and critical spatial data science. Journal of
-Geographical Systems. 2020;
-doi:\href{https://doi.org/10.1007/s10109-020-00334-2}{10.1007/s10109-020-00334-2}}
-
-\leavevmode\hypertarget{ref-Konkol2019examination}{}%
-\CSLLeftMargin{10. }
-\CSLRightInline{Konkol M, Kray C. In-depth examination of spatiotemporal
-figures in open reproducible research. Cartography and Geographic
-Information Science. 2019;46: 412--427.
-doi:\href{https://doi.org/10.1080/15230406.2018.1512421}{10.1080/15230406.2018.1512421}}
-
-\leavevmode\hypertarget{ref-Konkol2019computational}{}%
-\CSLLeftMargin{11. }
-\CSLRightInline{Konkol M, Kray C, Pfeiffer M. Computational
-reproducibility in geoscientific papers: Insights from a series of
-studies with geoscientists and a reproduction study. International
-Journal of Geographical Information Science. 2019;33: 408--429.
-doi:\href{https://doi.org/10.1080/13658816.2018.1508687}{10.1080/13658816.2018.1508687}}
-
-\leavevmode\hypertarget{ref-Iqbal2016reproducible}{}%
-\CSLLeftMargin{12. }
-\CSLRightInline{Iqbal SA, Wallach JD, Khoury MJ, Schully SD, Ioannidis
-JPA. Reproducible research practices and transparency across the
-biomedical literature. Plos Biology. 2016;14.
-doi:\href{https://doi.org/10.1371/journal.pbio.1002333}{10.1371/journal.pbio.1002333}}
+Kwon, D., 2020. How swamped preprint servers are blocking bad
+coronavirus research. Nature 581, 130--132.
+
+\leavevmode\hypertarget{ref-Lee2020human}{}%
+Lee, M., Zhao, J., Sun, Q., Pan, Y., Zhou, W., Xiong, C., Zhang, L.,
+2020. Human mobility trends during the early stage of the COVID-19
+pandemic in the united states. PLOS ONE 15, e0241468.
+doi:\href{https://doi.org/10.1371/journal.pone.0241468}{10.1371/journal.pone.0241468}
+
+\leavevmode\hypertarget{ref-Li2018effect}{}%
+Li, R., Richmond, P., Roehner, B.M., 2018. Effect of population density
+on epidemics. Physica A: Statistical Mechanics and its Applications 510,
+713--724.
+doi:\href{https://doi.org/10.1016/j.physa.2018.07.025}{10.1016/j.physa.2018.07.025}
+
+\leavevmode\hypertarget{ref-Maddala1983limited}{}%
+Maddala, G.S., 1983. Limited-dependent and qualitative variables in
+econometrics. Cambridge University Press, Cambridge.
+
+\leavevmode\hypertarget{ref-Micallef2020first}{}%
+Micallef, S., Piscopo, T.V., Casha, R., Borg, D., Vella, C., Zammit,
+M.-A., Borg, J., Mallia, D., Farrugia, J., Vella, S.M., Xerri, T.,
+Portelli, A., Fenech, M., Fsadni, C., Mallia Azzopardi, C., 2020. The
+first wave of COVID-19 in malta; a national cross-sectional study. PLOS
+ONE 15, e0239389.
+doi:\href{https://doi.org/10.1371/journal.pone.0239389}{10.1371/journal.pone.0239389}
+
+\leavevmode\hypertarget{ref-Molloy2020Tracing}{}%
+Molloy, J., Tchervenkov, C., Hintermann, B., Axhausen, K.W., 2020.
+Tracing the sars-CoV-2 impact: The first month in switzerland. Findings.
+doi:\href{https://doi.org/10.32866/001c.12903}{10.32866/001c.12903}
+
+\leavevmode\hypertarget{ref-Moore1970some}{}%
+Moore, E.G., 1970. Some spatial properties of urban contact fields.
+Geographical Analysis 2, 376--386.
+
+\leavevmode\hypertarget{ref-Moore1970urban}{}%
+Moore, E.G., Brown, L.A., 1970. Urban acquaintance fields: An evaluation
+of a spatial model. Environment and Planning 2, 443--454.
+
+\leavevmode\hypertarget{ref-Noland1995perceived}{}%
+Noland, R.B., 1995. PERCEIVED RISK AND MODAL CHOICE - RISK COMPENSATION
+IN TRANSPORTATION SYSTEM. Accident Analysis and Prevention 27, 503--521.
+doi:\href{https://doi.org/10.1016/0001-4575(94)00087-3}{10.1016/0001-4575(94)00087-3}
+
+\leavevmode\hypertarget{ref-Noland2021mobility}{}%
+Noland, R.B., 2021. Mobility and the effective reproduction rate of
+COVID-19. Journal of Transport \& Health 20, 101016.
+doi:\url{https://doi.org/10.1016/j.jth.2021.101016}
+
+\leavevmode\hypertarget{ref-Noury2021how}{}%
+Noury, A., François, A., Gergaud, O., Garel, A., 2021. How does COVID-19
+affect electoral participation? Evidence from the french municipal
+elections. PLOS ONE 16, e0247026.
+doi:\href{https://doi.org/10.1371/journal.pone.0247026}{10.1371/journal.pone.0247026}
+
+\leavevmode\hypertarget{ref-Paez2020using}{}%
+Paez, A., 2020. Using google community mobility reports to investigate
+the incidence of COVID-19 in the united states. Findings.
+doi:\url{https://doi.org/10.32866/001c.12976}
+
+\leavevmode\hypertarget{ref-Paez2020spatio}{}%
+Paez, A., Lopez, F.A., Menezes, T., Cavalcanti, R., Pitta, M.G. da R.,
+2020. A spatio-temporal analysis of the environmental correlates of
+COVID-19 incidence in spain. Geographical Analysis n/a.
+doi:\href{https://doi.org/10.1111/gean.12241}{10.1111/gean.12241}
+
+\leavevmode\hypertarget{ref-Pequeno2020air}{}%
+Pequeno, P., Mendel, B., Rosa, C., Bosholn, M., Souza, J.L., Baccaro,
+F., Barbosa, R., Magnusson, W., 2020. Air transportation, population
+density and temperature predict the spread of COVID-19 in brazil. PeerJ
+8, e9322.
+doi:\href{https://doi.org/10.7717/peerj.9322}{10.7717/peerj.9322}
+
+\leavevmode\hypertarget{ref-Phillips2011risk}{}%
+Phillips, R.O., Fyhri, A., Sagberg, F., 2011. Risk compensation and
+bicycle helmets. Risk Analysis 31, 1187--1195.
+doi:\href{https://doi.org/10.1111/j.1539-6924.2011.01589.x}{10.1111/j.1539-6924.2011.01589.x}
+
+\leavevmode\hypertarget{ref-Praharaj2020Using}{}%
+Praharaj, S., King, D., Pettit, C., Wentz, E., 2020. Using aggregated
+mobility data to measure the effect of COVID-19 policies on mobility
+changes in sydney, london, phoenix, and pune. Findings.
+doi:\href{https://doi.org/10.32866/001c.17590}{10.32866/001c.17590}
+
+\leavevmode\hypertarget{ref-Richens2000condoms}{}%
+Richens, J., Imrie, J., Copas, A., 2000. Condoms and seat belts: The
+parallels and the lessons. Lancet 355, 400--403.
+doi:\href{https://doi.org/10.1016/s0140-6736(99)09109-6}{10.1016/s0140-6736(99)09109-6}
+
+\leavevmode\hypertarget{ref-Rocklov2020high}{}%
+Rocklöv, J., Sjödin, H., 2020. High population densities catalyse the
+spread of COVID-19. Journal of Travel Medicine 27.
+doi:\href{https://doi.org/10.1093/jtm/taaa038}{10.1093/jtm/taaa038}
+
+\leavevmode\hypertarget{ref-Roy2020factors}{}%
+Roy, S., Ghosh, P., 2020. Factors affecting COVID-19 infected and death
+rates inform lockdown-related policymaking. PLOS ONE 15, e0241165.
+doi:\href{https://doi.org/10.1371/journal.pone.0241165}{10.1371/journal.pone.0241165}
+
+\leavevmode\hypertarget{ref-Sharifi2020covid}{}%
+Sharifi, A., Khavarian-Garmsir, A.R., 2020. The COVID-19 pandemic:
+Impacts on cities and major lessons for urban planning, design, and
+management. Science of The Total Environment 749, 142391.
+doi:\url{https://doi.org/10.1016/j.scitotenv.2020.142391}
+
+\leavevmode\hypertarget{ref-Skorka2020macroecology}{}%
+Skórka, P., Grzywacz, B., Moroń, D., Lenda, M., 2020. The macroecology
+of the COVID-19 pandemic in the anthropocene. PLOS ONE 15, e0236856.
+doi:\href{https://doi.org/10.1371/journal.pone.0236856}{10.1371/journal.pone.0236856}
+
+\leavevmode\hypertarget{ref-Souris2020covid}{}%
+Souris, M., Gonzalez, J.-P., 2020. COVID-19: Spatial analysis of
+hospital case-fatality rate in france. PLOS ONE 15, e0243606.
+doi:\href{https://doi.org/10.1371/journal.pone.0243606}{10.1371/journal.pone.0243606}
+
+\leavevmode\hypertarget{ref-Stephens2021impact}{}%
+Stephens, K.E., Chernyavskiy, P., Bruns, D.R., 2021. Impact of altitude
+on COVID-19 infection and death in the united states: A modeling and
+observational study. PLOS ONE 16, e0245055.
+doi:\href{https://doi.org/10.1371/journal.pone.0245055}{10.1371/journal.pone.0245055}
 
 \leavevmode\hypertarget{ref-Stodden2018empirical}{}%
-\CSLLeftMargin{13. }
-\CSLRightInline{Stodden V, Seiler J, Ma ZK. An empirical analysis of
+Stodden, V., Seiler, J., Ma, Z.K., 2018. An empirical analysis of
 journal policy effectiveness for computational reproducibility.
 Proceedings of the National Academy of Sciences of the United States of
-America. 2018;115: 2584--2589.
-doi:\href{https://doi.org/10.1073/pnas.1708290115}{10.1073/pnas.1708290115}}
+America 115, 2584--2589.
+doi:\href{https://doi.org/10.1073/pnas.1708290115}{10.1073/pnas.1708290115}
 
 \leavevmode\hypertarget{ref-Sumner2020reproducibility}{}%
-\CSLLeftMargin{14. }
-\CSLRightInline{Sumner J, Haynes L, Nathan S, Hudson-Vitale C, McIntosh
-LD. Reproducibility and reporting practices in COVID-19 preprint
-manuscripts. medRxiv. 2020; 2020.03.24.20042796.
-doi:\href{https://doi.org/10.1101/2020.03.24.20042796}{10.1101/2020.03.24.20042796}}
-
-\leavevmode\hypertarget{ref-Gustot2020quality}{}%
-\CSLLeftMargin{15. }
-\CSLRightInline{Gustot T. Quality and reproducibility during the
-COVID-19 pandemic. JHEP Rep. 2020;2: 100141.
-doi:\href{https://doi.org/10.1016/j.jhepr.2020.100141}{10.1016/j.jhepr.2020.100141}}
-
-\leavevmode\hypertarget{ref-Herndon2014high}{}%
-\CSLLeftMargin{16. }
-\CSLRightInline{Herndon T, Ash M, Pollin R. Does high public debt
-consistently stifle economic growth? A critique of reinhart and rogoff.
-Cambridge Journal of Economics. 2014;38: 257--279.
-doi:\href{https://doi.org/10.1093/cje/bet075}{10.1093/cje/bet075}}
-
-\leavevmode\hypertarget{ref-Basu2017ten}{}%
-\CSLLeftMargin{17. }
-\CSLRightInline{Basu S, Carney MA, Kenworthy NJ. Ten years after the
-financial crisis: The long reach of austerity and its global impacts on
-health. Social Science \& Medicine. 2017;187: 203--207.
-doi:\href{https://doi.org/10.1016/j.socscimed.2017.06.026}{10.1016/j.socscimed.2017.06.026}}
-
-\leavevmode\hypertarget{ref-Florida2020how}{}%
-\CSLLeftMargin{18. }
-\CSLRightInline{Florida R, Glaeser E, Sharif M, Bedi K, Campanella T,
-Chee C, et al. How life in our cities will look after the coronavirus
-pandemic. Foreign Policy. 2020;1. Available:
-\url{https://foreignpolicy.com/2020/05/01/future-of-cities-urban-life-after-coronavirus-pandemic/}}
+Sumner, J., Haynes, L., Nathan, S., Hudson-Vitale, C., McIntosh, L.D.,
+2020. Reproducibility and reporting practices in COVID-19 preprint
+manuscripts. medRxiv 2020.03.24.20042796.
+doi:\href{https://doi.org/10.1101/2020.03.24.20042796}{10.1101/2020.03.24.20042796}
+
+\leavevmode\hypertarget{ref-Sun2020impacts}{}%
+Sun, Z., Zhang, H., Yang, Y., Wan, H., Wang, Y., 2020. Impacts of
+geographic factors and population density on the COVID-19 spreading
+under the lockdown policies of china. Science of The Total Environment
+746, 141347.
+doi:\href{https://doi.org/10.1016/j.scitotenv.2020.141347}{10.1016/j.scitotenv.2020.141347}
 
 \leavevmode\hypertarget{ref-Sy2021population}{}%
-\CSLLeftMargin{19. }
-\CSLRightInline{Sy KTL, White LF, Nichols BE. Population density and
+Sy, K.T.L., White, L.F., Nichols, B.E., 2021. Population density and
 basic reproductive number of COVID-19 across united states counties.
-PLOS ONE. 2021;16: e0249271.
-doi:\href{https://doi.org/10.1371/journal.pone.0249271}{10.1371/journal.pone.0249271}}
-
-\leavevmode\hypertarget{ref-Moore1970urban}{}%
-\CSLLeftMargin{20. }
-\CSLRightInline{Moore EG, Brown LA. Urban acquaintance fields: An
-evaluation of a spatial model. Environment and Planning. 1970;2:
-443--454. Available:
-\url{http://www.envplan.com/abstract.cgi?id=a020443}}
-
-\leavevmode\hypertarget{ref-Moore1970some}{}%
-\CSLLeftMargin{21. }
-\CSLRightInline{Moore EG. Some spatial properties of urban contact
-fields. Geographical Analysis. 1970;2: 376--386. }
-
-\leavevmode\hypertarget{ref-Farber2011running}{}%
-\CSLLeftMargin{22. }
-\CSLRightInline{Farber S, Páez A. Running to stay in place: The time-use
-implications of automobile oriented land-use and travel. Journal of
-Transport Geography. 2011;19: 782--793.
-doi:\href{https://doi.org/10.1016/j.jtrangeo.2010.09.008}{10.1016/j.jtrangeo.2010.09.008}}
-
-\leavevmode\hypertarget{ref-Rocklov2020high}{}%
-\CSLLeftMargin{23. }
-\CSLRightInline{Rocklöv J, Sjödin H. High population densities catalyse
-the spread of COVID-19. Journal of Travel Medicine. 2020;27.
-doi:\href{https://doi.org/10.1093/jtm/taaa038}{10.1093/jtm/taaa038}}
-
-\leavevmode\hypertarget{ref-Li2018effect}{}%
-\CSLLeftMargin{24. }
-\CSLRightInline{Li R, Richmond P, Roehner BM. Effect of population
-density on epidemics. Physica A: Statistical Mechanics and its
-Applications. 2018;510: 713--724.
-doi:\href{https://doi.org/10.1016/j.physa.2018.07.025}{10.1016/j.physa.2018.07.025}}
-
-\leavevmode\hypertarget{ref-Cruz2020exploring}{}%
-\CSLLeftMargin{25. }
-\CSLRightInline{Cruz CJP, Ganly R, Li Z, Gietel-Basten S. Exploring the
-young demographic profile of COVID-19 cases in hong kong: Evidence from
-migration and travel history data. PLOS ONE. 2020;15: e0235306.
-doi:\href{https://doi.org/10.1371/journal.pone.0235306}{10.1371/journal.pone.0235306}}
-
-\leavevmode\hypertarget{ref-Micallef2020first}{}%
-\CSLLeftMargin{26. }
-\CSLRightInline{Micallef S, Piscopo TV, Casha R, Borg D, Vella C, Zammit
-M-A, et al. The first wave of COVID-19 in malta; a national
-cross-sectional study. PLOS ONE. 2020;15: e0239389.
-doi:\href{https://doi.org/10.1371/journal.pone.0239389}{10.1371/journal.pone.0239389}}
-
-\leavevmode\hypertarget{ref-Gomez2021infekta}{}%
-\CSLLeftMargin{27. }
-\CSLRightInline{Gomez J, Prieto J, Leon E, Rodríguez A. INFEKTA---an
-agent-based model for transmission of infectious diseases: The COVID-19
-case in bogotá, colombia. PLOS ONE. 2021;16: e0245787.
-doi:\href{https://doi.org/10.1371/journal.pone.0245787}{10.1371/journal.pone.0245787}}
-
-\leavevmode\hypertarget{ref-Noland1995perceived}{}%
-\CSLLeftMargin{28. }
-\CSLRightInline{Noland RB. PERCEIVED RISK AND MODAL CHOICE - RISK
-COMPENSATION IN TRANSPORTATION SYSTEM. Accident Analysis and Prevention.
-1995;27: 503--521.
-doi:\href{https://doi.org/10.1016/0001-4575(94)00087-3}{10.1016/0001-4575(94)00087-3}}
-
-\leavevmode\hypertarget{ref-Richens2000condoms}{}%
-\CSLLeftMargin{29. }
-\CSLRightInline{Richens J, Imrie J, Copas A. Condoms and seat belts: The
-parallels and the lessons. Lancet. 2000;355: 400--403.
-doi:\href{https://doi.org/10.1016/s0140-6736(99)09109-6}{10.1016/s0140-6736(99)09109-6}}
-
-\leavevmode\hypertarget{ref-Phillips2011risk}{}%
-\CSLLeftMargin{30. }
-\CSLRightInline{Phillips RO, Fyhri A, Sagberg F. Risk compensation and
-bicycle helmets. Risk Analysis. 2011;31: 1187--1195.
-doi:\href{https://doi.org/10.1111/j.1539-6924.2011.01589.x}{10.1111/j.1539-6924.2011.01589.x}}
-
-\leavevmode\hypertarget{ref-Chauhan2021covid}{}%
-\CSLLeftMargin{31. }
-\CSLRightInline{Chauhan RS, Capasso da Silva D, Salon D, Shamshiripour
-A, Rahimi E, Sutradhar U, et al. COVID-19 related attitudes and risk
-perceptions across urban, rural, and suburban areas in the united
-states. Findings. Network Design Lab; 2021;
-doi:\href{https://doi.org/10.32866/001c.23714}{10.32866/001c.23714}}
-
-\leavevmode\hypertarget{ref-Paez2020spatio}{}%
-\CSLLeftMargin{32. }
-\CSLRightInline{Paez A, Lopez FA, Menezes T, Cavalcanti R, Pitta MG da
-R. A spatio-temporal analysis of the environmental correlates of
-COVID-19 incidence in spain. Geographical Analysis. 2020;n/a.
-doi:\href{https://doi.org/10.1111/gean.12241}{10.1111/gean.12241}}
-
-\leavevmode\hypertarget{ref-Hamidi2020density}{}%
-\CSLLeftMargin{33. }
-\CSLRightInline{Hamidi S, Sabouri S, Ewing R. Does density aggravate the
-COVID-19 pandemic? Journal of the American Planning Association.
-2020;86: 495--509.
-doi:\href{https://doi.org/10.1080/01944363.2020.1777891}{10.1080/01944363.2020.1777891}}
-
-\leavevmode\hypertarget{ref-Jamal2020Changes}{}%
-\CSLLeftMargin{34. }
-\CSLRightInline{Jamal S, Paez A. Changes in trip-making frequency by
-mode during COVID-19. Findings. Network Design Lab; 2020;
-doi:\href{https://doi.org/10.32866/001c.17977}{10.32866/001c.17977}}
-
-\leavevmode\hypertarget{ref-Harris2021Changes}{}%
-\CSLLeftMargin{35. }
-\CSLRightInline{Harris MA, Branion-Calles M. Changes in commute mode
-attributed to COVID-19 risk in canadian national survey data. Findings.
-Network Design Lab; 2021;
-doi:\href{https://doi.org/10.32866/001c.19088}{10.32866/001c.19088}}
-
-\leavevmode\hypertarget{ref-Molloy2020Tracing}{}%
-\CSLLeftMargin{36. }
-\CSLRightInline{Molloy J, Tchervenkov C, Hintermann B, Axhausen KW.
-Tracing the sars-CoV-2 impact: The first month in switzerland. Findings.
-Network Design Lab; 2020;
-doi:\href{https://doi.org/10.32866/001c.12903}{10.32866/001c.12903}}
-
-\leavevmode\hypertarget{ref-Feyman2020effectiveness}{}%
-\CSLLeftMargin{37. }
-\CSLRightInline{Feyman Y, Bor J, Raifman J, Griffith KN. Effectiveness
-of COVID-19 shelter-in-place orders varied by state. PLOS ONE. 2020;15:
-e0245008.
-doi:\href{https://doi.org/10.1371/journal.pone.0245008}{10.1371/journal.pone.0245008}}
-
-\leavevmode\hypertarget{ref-Praharaj2020Using}{}%
-\CSLLeftMargin{38. }
-\CSLRightInline{Praharaj S, King D, Pettit C, Wentz E. Using aggregated
-mobility data to measure the effect of COVID-19 policies on mobility
-changes in sydney, london, phoenix, and pune. Findings. Network Design
-Lab; 2020;
-doi:\href{https://doi.org/10.32866/001c.17590}{10.32866/001c.17590}}
-
-\leavevmode\hypertarget{ref-Paez2020using}{}%
-\CSLLeftMargin{39. }
-\CSLRightInline{Paez A. Using google community mobility reports to
-investigate the incidence of COVID-19 in the united states. Findings.
-2020; doi:\url{https://doi.org/10.32866/001c.12976}}
-
-\leavevmode\hypertarget{ref-Noland2021mobility}{}%
-\CSLLeftMargin{40. }
-\CSLRightInline{Noland RB. Mobility and the effective reproduction rate
-of COVID-19. Journal of Transport \& Health. 2021;20: 101016.
-doi:\url{https://doi.org/10.1016/j.jth.2021.101016}}
-
-\leavevmode\hypertarget{ref-Bhadra2021impact}{}%
-\CSLLeftMargin{41. }
-\CSLRightInline{Bhadra A, Mukherjee A, Sarkar K. Impact of population
-density on covid-19 infected and mortality rate in india. Modeling Earth
-Systems and Environment. 2021;7: 623--629.
-doi:\href{https://doi.org/10.1007/s40808-020-00984-7}{10.1007/s40808-020-00984-7}}
-
-\leavevmode\hypertarget{ref-Kadi2020population}{}%
-\CSLLeftMargin{42. }
-\CSLRightInline{Kadi N, Khelfaoui M. Population density, a factor in the
-spread of COVID-19 in algeria: Statistic study. Bulletin of the National
-Research Centre. 2020;44.
-doi:\href{https://doi.org/10.1186/s42269-020-00393-x}{10.1186/s42269-020-00393-x}}
-
-\leavevmode\hypertarget{ref-Pequeno2020air}{}%
-\CSLLeftMargin{43. }
-\CSLRightInline{Pequeno P, Mendel B, Rosa C, Bosholn M, Souza JL,
-Baccaro F, et al. Air transportation, population density and temperature
-predict the spread of COVID-19 in brazil. PeerJ. 2020;8: e9322.
-doi:\href{https://doi.org/10.7717/peerj.9322}{10.7717/peerj.9322}}
-
-\leavevmode\hypertarget{ref-Fielding2020social}{}%
-\CSLLeftMargin{44. }
-\CSLRightInline{Fielding-Miller RK, Sundaram ME, Brouwer K. Social
-determinants of COVID-19 mortality at the county level. PLOS ONE.
-2020;15: e0240151.
-doi:\href{https://doi.org/10.1371/journal.pone.0240151}{10.1371/journal.pone.0240151}}
-
-\leavevmode\hypertarget{ref-Roy2020factors}{}%
-\CSLLeftMargin{45. }
-\CSLRightInline{Roy S, Ghosh P. Factors affecting COVID-19 infected and
-death rates inform lockdown-related policymaking. PLOS ONE. 2020;15:
-e0241165.
-doi:\href{https://doi.org/10.1371/journal.pone.0241165}{10.1371/journal.pone.0241165}}
+PLOS ONE 16, e0249271.
+doi:\href{https://doi.org/10.1371/journal.pone.0249271}{10.1371/journal.pone.0249271}
+
+\leavevmode\hypertarget{ref-Vlasschaert2020proliferation}{}%
+Vlasschaert, C., Topf, J.M., Hiremath, S., 2020. Proliferation of papers
+and preprints during the coronavirus disease 2019 pandemic: Progress or
+problems with peer review? Advances in Chronic Kidney Disease 27,
+418--426.
+doi:\href{https://doi.org/10.1053/j.ackd.2020.08.003}{10.1053/j.ackd.2020.08.003}
+
+\leavevmode\hypertarget{ref-Wang2021transmission}{}%
+Wang, F., Tan, Z., Yu, Z., Yao, S., Guo, C., 2021. Transmission and
+control pressure analysis of the COVID-19 epidemic situation using
+multisource spatio-temporal big data. PLOS ONE 16, e0249145.
+doi:\href{https://doi.org/10.1371/journal.pone.0249145}{10.1371/journal.pone.0249145}
+
+\leavevmode\hypertarget{ref-White2020state}{}%
+White, E.R., Hébert-Dufresne, L., 2020. State-level variation of initial
+COVID-19 dynamics in the united states. PLOS ONE 15, e0240648.
+doi:\href{https://doi.org/10.1371/journal.pone.0240648}{10.1371/journal.pone.0240648}
 
 \leavevmode\hypertarget{ref-Wong2020spreading}{}%
-\CSLLeftMargin{46. }
-\CSLRightInline{Wong DWS, Li Y. Spreading of COVID-19: Density matters.
-PLOS ONE. 2020;15: e0242398.
-doi:\href{https://doi.org/10.1371/journal.pone.0242398}{10.1371/journal.pone.0242398}}
-
-\leavevmode\hypertarget{ref-Sun2020impacts}{}%
-\CSLLeftMargin{47. }
-\CSLRightInline{Sun Z, Zhang H, Yang Y, Wan H, Wang Y. Impacts of
-geographic factors and population density on the COVID-19 spreading
-under the lockdown policies of china. Science of The Total Environment.
-2020;746: 141347.
-doi:\href{https://doi.org/10.1016/j.scitotenv.2020.141347}{10.1016/j.scitotenv.2020.141347}}
-
-\leavevmode\hypertarget{ref-Skorka2020macroecology}{}%
-\CSLLeftMargin{48. }
-\CSLRightInline{Skórka P, Grzywacz B, Moroń D, Lenda M. The macroecology
-of the COVID-19 pandemic in the anthropocene. PLOS ONE. 2020;15:
-e0236856.
-doi:\href{https://doi.org/10.1371/journal.pone.0236856}{10.1371/journal.pone.0236856}}
-
-\leavevmode\hypertarget{ref-White2020state}{}%
-\CSLLeftMargin{49. }
-\CSLRightInline{White ER, Hébert-Dufresne L. State-level variation of
-initial COVID-19 dynamics in the united states. PLOS ONE. 2020;15:
-e0240648.
-doi:\href{https://doi.org/10.1371/journal.pone.0240648}{10.1371/journal.pone.0240648}}
-
-\leavevmode\hypertarget{ref-Khavarian2021high}{}%
-\CSLLeftMargin{50. }
-\CSLRightInline{Khavarian-Garmsir AR, Sharifi A, Moradpour N. Are
-high-density districts more vulnerable to the COVID-19 pandemic?
-Sustainable Cities and Society. 2021;70: 102911.
-doi:\href{https://doi.org/10.1016/j.scs.2021.102911}{10.1016/j.scs.2021.102911}}
-
-\leavevmode\hypertarget{ref-Hamidi2020longitudinal}{}%
-\CSLLeftMargin{51. }
-\CSLRightInline{Hamidi S, Ewing R, Sabouri S. Longitudinal analyses of
-the relationship between development density and the COVID-19 morbidity
-and mortality rates: Early evidence from 1,165 metropolitan counties in
-the united states. Health \& Place. 2020;64: 102378.
-doi:\href{https://doi.org/10.1016/j.healthplace.2020.102378}{10.1016/j.healthplace.2020.102378}}
-
-\leavevmode\hypertarget{ref-Lee2020human}{}%
-\CSLLeftMargin{52. }
-\CSLRightInline{Lee M, Zhao J, Sun Q, Pan Y, Zhou W, Xiong C, et al.
-Human mobility trends during the early stage of the COVID-19 pandemic in
-the united states. PLOS ONE. 2020;15: e0241468.
-doi:\href{https://doi.org/10.1371/journal.pone.0241468}{10.1371/journal.pone.0241468}}
-
-\leavevmode\hypertarget{ref-Amadu2021assessing}{}%
-\CSLLeftMargin{53. }
-\CSLRightInline{Amadu I, Ahinkorah BO, Afitiri A-R, Seidu A-A, Ameyaw
-EK, Hagan JE, et al. Assessing sub-regional-specific strengths of
-healthcare systems associated with COVID-19 prevalence, deaths and
-recoveries in africa. PLOS ONE. 2021;16: e0247274.
-doi:\href{https://doi.org/10.1371/journal.pone.0247274}{10.1371/journal.pone.0247274}}
-
-\leavevmode\hypertarget{ref-Feng2020spread}{}%
-\CSLLeftMargin{54. }
-\CSLRightInline{Feng Y, Li Q, Tong X, Wang R, Zhai S, Gao C, et al.
-Spatiotemporal spread pattern of the COVID-19 cases in china. PLOS ONE.
-2020;15: e0244351.
-doi:\href{https://doi.org/10.1371/journal.pone.0244351}{10.1371/journal.pone.0244351}}
-
-\leavevmode\hypertarget{ref-Inbaraj2021seroprevalence}{}%
-\CSLLeftMargin{55. }
-\CSLRightInline{Inbaraj LR, George CE, Chandrasingh S. Seroprevalence of
-COVID-19 infection in a rural district of south india: A
-population-based seroepidemiological study. PLOS ONE. 2021;16: e0249247.
-doi:\href{https://doi.org/10.1371/journal.pone.0249247}{10.1371/journal.pone.0249247}}
-
-\leavevmode\hypertarget{ref-Souris2020covid}{}%
-\CSLLeftMargin{56. }
-\CSLRightInline{Souris M, Gonzalez J-P. COVID-19: Spatial analysis of
-hospital case-fatality rate in france. PLOS ONE. 2020;15: e0243606.
-doi:\href{https://doi.org/10.1371/journal.pone.0243606}{10.1371/journal.pone.0243606}}
-
-\leavevmode\hypertarget{ref-Ahmad2020association}{}%
-\CSLLeftMargin{57. }
-\CSLRightInline{Ahmad K, Erqou S, Shah N, Nazir U, Morrison AR,
-Choudhary G, et al. Association of poor housing conditions with COVID-19
-incidence and mortality across US counties. PLOS ONE. 2020;15: e0241327.
-doi:\href{https://doi.org/10.1371/journal.pone.0241327}{10.1371/journal.pone.0241327}}
-
-\leavevmode\hypertarget{ref-Noury2021how}{}%
-\CSLLeftMargin{58. }
-\CSLRightInline{Noury A, François A, Gergaud O, Garel A. How does
-COVID-19 affect electoral participation? Evidence from the french
-municipal elections. PLOS ONE. 2021;16: e0247026.
-doi:\href{https://doi.org/10.1371/journal.pone.0247026}{10.1371/journal.pone.0247026}}
-
-\leavevmode\hypertarget{ref-Wang2021transmission}{}%
-\CSLLeftMargin{59. }
-\CSLRightInline{Wang F, Tan Z, Yu Z, Yao S, Guo C. Transmission and
-control pressure analysis of the COVID-19 epidemic situation using
-multisource spatio-temporal big data. PLOS ONE. 2021;16: e0249145.
-doi:\href{https://doi.org/10.1371/journal.pone.0249145}{10.1371/journal.pone.0249145}}
-
-\leavevmode\hypertarget{ref-Brandtner2021creatures}{}%
-\CSLLeftMargin{60. }
-\CSLRightInline{Brandtner C, Bettencourt LMA, Berman MG, Stier AJ.
-Creatures of the state? Metropolitan counties compensated for state
-inaction in initial u.s. Response to COVID-19 pandemic. PLOS ONE.
-2021;16: e0246249.
-doi:\href{https://doi.org/10.1371/journal.pone.0246249}{10.1371/journal.pone.0246249}}
-
-\leavevmode\hypertarget{ref-Stephens2021impact}{}%
-\CSLLeftMargin{61. }
-\CSLRightInline{Stephens KE, Chernyavskiy P, Bruns DR. Impact of
-altitude on COVID-19 infection and death in the united states: A
-modeling and observational study. PLOS ONE. 2021;16: e0245055.
-doi:\href{https://doi.org/10.1371/journal.pone.0245055}{10.1371/journal.pone.0245055}}
-
-\leavevmode\hypertarget{ref-Maddala1983limited}{}%
-\CSLLeftMargin{62. }
-\CSLRightInline{Maddala GS. Limited-dependent and qualitative variables
-in econometrics. Cambridge: Cambridge University Press; 1983. }
-
-\leavevmode\hypertarget{ref-Cressie1989geostatistics}{}%
-\CSLLeftMargin{63. }
-\CSLRightInline{Cressie N. Geostatistics. The American Statistician.
-1989;43: 197.
-doi:\href{https://doi.org/10.2307/2685361}{10.2307/2685361}}
+Wong, D.W.S., Li, Y., 2020. Spreading of COVID-19: Density matters. PLOS
+ONE 15, e0242398.
+doi:\href{https://doi.org/10.1371/journal.pone.0242398}{10.1371/journal.pone.0242398}
 
@@
 
-\nolinenumbers
-
 \end{document}

### 2021-05-25 README.Rmd
Paper reformatted for submission to Science of the Total Environment after PLoS ONE desk rejection

@@
 
-Paper submitted to *PLoS ONE*
+Paper submitted to *Science of the Total Environment*
 
@@
 The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research.
-Given the high stakes of the situation, it is essential that scientific activites, on which good policy depends, are as transparent and reproducible as possible.
+Given the high stakes of the situation, it is essential that scientific activities, on which good policy depends, are as transparent and reproducible as possible.
 Reproducibility is key for the efficient operation of the self-correction mechanisms of science, which work to weed out errors and refine our understanding of social and physical phenomena.
 In this paper, the importance of reproducibility is illustrated for the case of the association between population density and the the spread of SARS-CoV-2.
-Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequantial changes in the conclusions, as the empirical example in this paper shows.
+Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example in this paper shows.
 

### 2021-05-26 R0-Density-Reanalysis/Cover-Letter-v1.Rmd
Paper prepared for submission to Landscape and Urban Planning

--- 2021-05-21 R0-Density-Reanalysis/Cover-Letter-v1.Rmd
@@
 
-Please find attached the paper titled "The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic".
-The paper illustrates the importance of reproducibility in research in the particular context of COVID-19.
-It takes a recent paper from the literature (Sy et al., 2021, published in PLoS ONE) and demonstrates how openly sharing data and code can accelerate research.
+Please find attached the paper titled "Population density and the spread of the COVID-19 pandemic: a reproducible research example".
+The question of population density has been of interest during the pandemic.
+On the one hand, mathematical models of contagion suggest that high density can catalyze transmission; on the other, the empirical literature remains decidedly mixed.
+In addition, an important aspect of the flood of research during the pandemic has been the need to maintain the quality of research, for which reproducibility is key.
+This paper presents an investigation of population density and the spread of COVID-19, and by doing so illustrates the importance of reproducible research.
 
@@
 The source document is an R Markdown file, a self-contained document with the text and _all_ the analysis.
-These materials are publicly available in a repository, as indicated in the paper.
+These materials are publicly available in a repository.
 

### 2021-05-26 R0-Density-Reanalysis/Highlights.Rmd
Paper prepared for submission to Landscape and Urban Planning

---
title: "Highlights"
output:
  pdf_document: default
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

- The emerging literature on the relationship between population density and the spread of COVID-19 is reviewed.

- It is shown that reproducibility in this literature remains an elusive goal.

- Analysis of population density and basic reproductive number shows that density is not destiny.

- The reanalysis and reappraisal of the conclusions demonstrates the importance of reproducible research.

- The paper is an example of a reproducible research project

### 2021-05-26 R0-Density-Reanalysis/R0-Density-Reanalysis-Abstract.Rmd
Paper prepared for submission to Landscape and Urban Planning

---
title: "Population density and the spread of the COVID-19 pandemic: a reproducible research example"
abstract: |
  The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research. Of interest has been the associations between population density and the spread of the pandemic. In this paper, population density and the basic reproductive number of SARS-CoV-2 are examined in an example of reproducible research. Given the high stakes of the situation, it is essential that scientific activities, on which good policy depends, are as transparent and reproducible as possible. Reproducibility is key for the efficient operation of the self-correction mechanisms of science. Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example of the spread of COVID-19 in the US shows.

journal: "Landscape and Urban Planning"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
    - \usepackage{setspace}\doublespacing
---

### 2021-05-26 R0-Density-Reanalysis/R0-Density-Reanalysis-Abstract.tex
Paper prepared for submission to Landscape and Urban Planning

--- 2021-05-25 R0-Density-Reanalysis/R0-Density-Reanalysis-Title-Page.tex
@@
 
-  \journal{Science of the Total Environment} % Sets Journal name
+  \journal{Landscape and Urban Planning} % Sets Journal name
 
@@
             pdfauthor={},
-            pdftitle={The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic},
+            pdftitle={Population density and the spread of the COVID-19 pandemic: a reproducible research example},
             colorlinks=false,
@@
 
-  \title{The importance of reproducibility in COVID-19 research: the
-case of population density and the spread of the pandemic}
-    \author[McMaster University]{Antonio Paez\corref{1}}
-   \ead{paezha@mcmaster.ca} 
-      \address[McMaster University]{School of Earth, Environment and
-Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada}
-      \cortext[1]{Corresponding Author}
+  \title{Population density and the spread of the COVID-19 pandemic: a
+reproducible research example}
 
@@
   COVID-19 pandemic has led to explosive growth in scientific research.
-  Given the high stakes of the situation, it is essential that
-  scientific activities, on which good policy depends, are as
-  transparent and reproducible as possible. Reproducibility is key for
-  the efficient operation of the self-correction mechanisms of science,
-  which work to weed out errors and refine our understanding of social
-  and physical phenomena. In this paper, the importance of
-  reproducibility is illustrated for the case of the association between
-  population density and the the spread of SARS-CoV-2. Transparency and
-  openness means that the same problem can, with relatively modest
-  efforts, be examined by independent researchers who can verify
-  findings, and bring to bear different perspectives, approaches, and
-  methods---sometimes with consequential changes in the conclusions, as
-  the empirical example in this paper shows.
+  Of interest has been the associations between population density and
+  the spread of the pandemic. In this paper, population density and the
+  basic reproductive number of SARS-CoV-2 are examined in an example of
+  reproducible research. Given the high stakes of the situation, it is
+  essential that scientific activities, on which good policy depends,
+  are as transparent and reproducible as possible. Reproducibility is
+  key for the efficient operation of the self-correction mechanisms of
+  science. Transparency and openness means that the same problem can,
+  with relatively modest efforts, be examined by independent researchers
+  who can verify findings, and bring to bear different perspectives,
+  approaches, and methods---sometimes with consequential changes in the
+  conclusions, as the empirical example of the spread of COVID-19 in the
+  US shows.
   \end{abstract}

### 2021-05-26 R0-Density-Reanalysis/R0-Density-Reanalysis-Title-Page.Rmd
Paper prepared for submission to Landscape and Urban Planning

@@
 ---
-title: "The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic"
+title: "Population density and the spread of the COVID-19 pandemic: a reproducible research example"
 author:
@@
   The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research.
+Of interest has been the associations between population density and the spread of the pandemic.
+In this paper, population density and the basic reproductive number of SARS-CoV-2 are examined in an example of reproducible research.
 Given the high stakes of the situation, it is essential that scientific activities, on which good policy depends, are as transparent and reproducible as possible.
-Reproducibility is key for the efficient operation of the self-correction mechanisms of science, which work to weed out errors and refine our understanding of social and physical phenomena.
-In this paper, the importance of reproducibility is illustrated for the case of the association between population density and the the spread of SARS-CoV-2.
-Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example in this paper shows.
+Reproducibility is key for the efficient operation of the self-correction mechanisms of science.
+Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example of the spread of COVID-19 in the US shows.
 
-journal: "Science of the Total Environment"
+journal: "Landscape and Urban Planning"
 date: "`r Sys.Date()`"

### 2021-05-26 R0-Density-Reanalysis/R0-Density-Reanalysis-Title-Page.tex
Paper prepared for submission to Landscape and Urban Planning

@@
 
-  \title{The importance of reproducibility in COVID-19 research: the
-case of population density and the spread of the pandemic}
+  \title{Population density and the spread of the COVID-19 pandemic: a
+reproducible research example}
     \author[McMaster University]{Antonio Paez\corref{1}}

### 2021-05-26 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Paper prepared for submission to Landscape and Urban Planning

@@
 ---
-title: "The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic"
+title: "Population density and the spread of the COVID-19 pandemic: a reproducible research example"
 author:
-  - name: Antonio Paez
-    email: paezha@mcmaster.ca
-    affiliation: McMaster University
+  - name: Author
+    email: author@institution.edu
+    affiliation: University
     footnote: 1
 address:
-  - code: McMaster University
-    address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
+  - code: University
+    address: Department, Street, City, State ZIP
 footnote:
@@
 
-```{r install-data-package, include = FALSE}
-# Run only once if needed to install data package `r0density`
-if (!require("r0density", character.only = TRUE)) {
-  devtools::install_github("https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed", subdir = "r0density")
-}
-```
+<!--
+Install data package as appropriate. The data package is file `r0density_0.1.0.tar.gz`
+-->
 
@@
 
-Given the heavy human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has even called this deluge of publications a "carnage of substandard research" [@Bramstedt2020carnage].
+Given the heavy human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has termed this deluge of publications a "carnage of substandard research" [@Bramstedt2020carnage].
 Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [@Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty and somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
@@
 And in the longer term, will the risks of life in high density regions presage a flight from cities?
+What are the implications of the pandemic for future urban planning and practice?
 Over the past year, numerous papers have sought to throw light into the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed.
@@
 
-The objective of this paper is to illustrate the importance of reproducibility in research, particularly in the context of the flood of COVID-19 papers.
+The objective of this paper is to illustrate the importance of reproducibility in research in the context of the flood of COVID-19 papers.
 To this end, a recent study by Sy et al.
@@
 This paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
-The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, is available in a public repository^[\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}].
+The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, will be available in a public repository^[For peer-review purposes, the data package and code are currently in an anonymous Drive folder: \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}].
 
@@
 [-@Hamidi2020density] posit this mechanism (i.e., greater compliance with social distancing in denser regions) to explain the results of their analyses.
-The evidence available does indeed show that there were important changes in behavior with respect to mobility during the pandemic [@Jamal2020Changes; @Harris2021Changes; @Molloy2020Tracing]; furthermore, shelter in place orders may have had greater buy-in from the public in higher density regions [@Feyman2020effectiveness], and the associated behavior may have persisted beyond the duration of official social-distancing policies [@Praharaj2020Using].
+The evidence available does indeed show that there were important changes in behavior with respect to mobility during the pandemic [@Jamal2020Changes; @Harris2021Changes; @Molloy2020Tracing]; furthermore, shelter in place orders may have had greater buy-in from the public in higher density regions [@Feyman2020effectiveness; @Hamidi2021compact], and the associated behavior may have persisted beyond the duration of official social-distancing policies [@Praharaj2020Using].
 In addition, there is evidence that changes in mobility correlated with the trajectory of the pandemic [@Paez2020using; @Noland2021mobility].

### 2021-05-26 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Paper prepared for submission to Landscape and Urban Planning

@@
 
-  \title{The importance of reproducibility in COVID-19 research: the
-case of population density and the spread of the pandemic}
-    \author[McMaster University]{Antonio Paez\corref{1}}
-   \ead{paezha@mcmaster.ca} 
-      \address[McMaster University]{School of Earth, Environment and
-Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada}
+  \title{Population density and the spread of the COVID-19 pandemic: a
+reproducible research example}
+    \author[University]{Author\corref{1}}
+   \ead{author@institution.edu} 
+      \address[University]{Department, Street, City, State ZIP}
       \cortext[1]{Corresponding Author}
@@
 consistently high quality standards in scientific reporting; indeed, a
-call for maintaining the standards in published research has even called
-this deluge of publications a ``carnage of substandard research''
-(Bramstedt, 2020). Part of the challenge of maintaining quality
-standards in published research is that, despite an abundance of
-recommendations and guidelines (Broggini et al., 2017; Brunsdon and
-Comber, 2020; Ince et al., 2012; Ioannidis et al., 2014), in practice
-reproducibility has remained a lofty and somewhat aspirational goal
-(Konkol et al., 2019; Konkol and Kray, 2019). As reported in the
-literature, only a woefully small proportion of published research was
-actually reproducible before the pandemic (Iqbal et al., 2016; Stodden
-et al., 2018), and the situation does not appear to have changed
-substantially since (Gustot, 2020; Sumner et al., 2020).
+call for maintaining the standards in published research has termed this
+deluge of publications a ``carnage of substandard research'' (Bramstedt,
+2020). Part of the challenge of maintaining quality standards in
+published research is that, despite an abundance of recommendations and
+guidelines (Broggini et al., 2017; Brunsdon and Comber, 2020; Ince et
+al., 2012; Ioannidis et al., 2014), in practice reproducibility has
+remained a lofty and somewhat aspirational goal (Konkol et al., 2019;
+Konkol and Kray, 2019). As reported in the literature, only a woefully
+small proportion of published research was actually reproducible before
+the pandemic (Iqbal et al., 2016; Stodden et al., 2018), and the
+situation does not appear to have changed substantially since (Gustot,
+2020; Sumner et al., 2020).
 
@@
 And in the longer term, will the risks of life in high density regions
-presage a flight from cities? Over the past year, numerous papers have
-sought to throw light into the underlying issue of density and the
-pandemic; nonetheless the results, as will be detailed next, remain
-mixed. Further, to complicate matters, precious few of these studies
-appear to be sufficiently open to support independent verification.
+presage a flight from cities? What are the implications of the pandemic
+for future urban planning and practice? Over the past year, numerous
+papers have sought to throw light into the underlying issue of density
+and the pandemic; nonetheless the results, as will be detailed next,
+remain mixed. Further, to complicate matters, precious few of these
+studies appear to be sufficiently open to support independent
+verification.
 
 The objective of this paper is to illustrate the importance of
-reproducibility in research, particularly in the context of the flood of
-COVID-19 papers. To this end, a recent study by Sy et al. (2021) is
-chosen as an example of reproducible research. The objective is not to
-malign the analysis of these researchers, but rather to demonstrate the
-value of openness to allow for independent verification and further
-analysis. Open data and open code mean that an independent researcher
-can, with only modest efforts, not only verify the findings reported,
-but also examine the same data from a perspective which may not have
-been available to the original researchers due to differences in
-disciplinary perspectives, methodological traditions, and/or training,
-among other possible factors. The example, which shows consequential
-changes in the conclusions reached by different analyses, should serve
-as a call to researchers to redouble their efforts to increase
-transparency and reproducibility in research. This paper, in addition,
-aims to show how data can be packaged in well-documented, shareable
-units, and code can be embedded into self-contained documents suitable
-for review and independent verification. The source for this paper is an
+reproducibility in research in the context of the flood of COVID-19
+papers. To this end, a recent study by Sy et al. (2021) is chosen as an
+example of reproducible research. The objective is not to malign the
+analysis of these researchers, but rather to demonstrate the value of
+openness to allow for independent verification and further analysis.
+Open data and open code mean that an independent researcher can, with
+only modest efforts, not only verify the findings reported, but also
+examine the same data from a perspective which may not have been
+available to the original researchers due to differences in disciplinary
+perspectives, methodological traditions, and/or training, among other
+possible factors. The example, which shows consequential changes in the
+conclusions reached by different analyses, should serve as a call to
+researchers to redouble their efforts to increase transparency and
+reproducibility in research. This paper, in addition, aims to show how
+data can be packaged in well-documented, shareable units, and code can
+be embedded into self-contained documents suitable for review and
+independent verification. The source for this paper is an
 \href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
-with the data package, is available in a public repository\footnote{\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}}.
+with the data package, will be available in a public
+repository\footnote{For peer-review purposes, the data package and code
+  are currently in an anonymous Drive folder:
+  \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}}.
 
@@
 in place orders may have had greater buy-in from the public in higher
-density regions (Feyman et al., 2020), and the associated behavior may
-have persisted beyond the duration of official social-distancing
-policies (Praharaj et al., 2020). In addition, there is evidence that
-changes in mobility correlated with the trajectory of the pandemic
-(Noland, 2021; Paez, 2020). Given the potential for behavioral
-adaptation, the question of density becomes more nuanced: it is not just
-a matter of proximity, but also of human behavior, which is better
-studied using population-level data and models.
+density regions (Feyman et al., 2020; Hamidi and Zandiatashbar, 2021),
+and the associated behavior may have persisted beyond the duration of
+official social-distancing policies (Praharaj et al., 2020). In
+addition, there is evidence that changes in mobility correlated with the
+trajectory of the pandemic (Noland, 2021; Paez, 2020). Given the
+potential for behavioral adaptation, the question of density becomes
+more nuanced: it is not just a matter of proximity, but also of human
+behavior, which is better studied using population-level data and
+models.
 
@@
 
+\leavevmode\hypertarget{ref-Hamidi2021compact}{}%
+Hamidi, S., Zandiatashbar, A., 2021. Compact development and adherence
+to stay-at-home order during the COVID-19 pandemic: A longitudinal
+investigation in the united states. Landscape and Urban Planning 205,
+103952. doi:\url{https://doi.org/10.1016/j.landurbplan.2020.103952}
+
 \leavevmode\hypertarget{ref-Harris2021Changes}{}%

### 2021-05-26 README.Rmd
Paper prepared for submission to Landscape and Urban Planning

@@
 
-# The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic
+# Population density and the spread of the COVID-19 pandemic: a reproducible research example
 
@@
 
-Paper submitted to *Science of the Total Environment*
+Paper submitted to *Landscape and Urban Planning*
 
@@
 The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research.
+Of interest has been the associations between population density and the spread of the pandemic.
+In this paper, population density and the basic reproductive number of SARS-CoV-2 are examined in an example of reproducible research.
 Given the high stakes of the situation, it is essential that scientific activities, on which good policy depends, are as transparent and reproducible as possible.
-Reproducibility is key for the efficient operation of the self-correction mechanisms of science, which work to weed out errors and refine our understanding of social and physical phenomena.
-In this paper, the importance of reproducibility is illustrated for the case of the association between population density and the the spread of SARS-CoV-2.
-Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example in this paper shows.
+Reproducibility is key for the efficient operation of the self-correction mechanisms of science.
+Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example of the spread of COVID-19 in the US shows.
 

### 2021-06-07 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Reknitted for possible submission to GA

@@
 ---
-title: "Population density and the spread of the COVID-19 pandemic: a reproducible research example"
+title: "The importance of reproducibility in geographical research: the case of population density and the spread of COVID-19"
 author:
@@
 
-journal: "Landscape and Urban Planning"
+journal: "Geographical Analysis"
 date: "`r Sys.Date()`"

### 2021-06-07 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Reknitted for possible submission to GA

@@
 
-  \journal{Landscape and Urban Planning} % Sets Journal name
+  \journal{Geographical Analysis} % Sets Journal name
 
@@
             pdfauthor={},
-            pdftitle={Population density and the spread of the COVID-19 pandemic: a reproducible research example},
+            pdftitle={The importance of reproducibility in geographical research: the case of population density and the spread of COVID-19},
             colorlinks=false,
@@
 
-  \title{Population density and the spread of the COVID-19 pandemic: a
-reproducible research example}
+  \title{The importance of reproducibility in geographical research: the
+case of population density and the spread of COVID-19}
     \author[University]{Author\corref{1}}

### 2021-06-07 README.Rmd
Reknitted for possible submission to GA

@@
 
-Paper submitted to *Landscape and Urban Planning*
+Paper submitted to [~~PLoS ONE~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/PLOS ONE Decision_ PONE-D-21-16874.pdf) [~~Science of the Total Environment~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Your Submission STOTEN-D-21-12077.pdf) [~~Landscape and Urban Planning~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Decision on submission to Landscape and Urban Planning.pdf)
 
@@
 library(censReg) # Package to estimate tobit model
-library(gridExtra)
+library(gridExtra) # PAckage used to arrange graphical objects
 library(kableExtra) # Package for creating tables
 library(Matrix) # Used to convert list of spatial weights to sparse matrix; needed for spatialprobit
-library(nlme) # Used to estimate 
-library(patchwork)
+library(nlme) # Used to estimate hierarchical models
+library(patchwork) # Package for arranging ggplot objects
 library(r0density) # Data package

### 2021-06-09 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Experiments with spatial filter in the outcome model of the Heckman selection model

@@
 ```{r load-packages, include=FALSE, cache=FALSE}
+library(adespatial) # To calculate Moran eigenvector maps
 library(censReg) # Package to estimate tobit model
@@
 
+county_listw <- nb2listw(county_nb, 
+                        style="W", 
+                        zero.policy = TRUE)
+
 state_listw <- nb2listw(state_nb, 
@@
 B <- as(nb_B, "CsparseMatrix")
+```
+
+```{r moran-eigenvector-maps, include=FALSE}
+county_mem <- mem(county_listw)
 ```
@@
 
-How relevant is the difference between these different model specifications?
-Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 1 and Selection Model 2.
-The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1.
-The conclusion is that at higher densities, $R_0$ is _always_ higher.
-The right panel, in contrast, shows that, according to Selection Model 2, $R_0$ is zero when density is zero (as expected), and then it tends to increase at higher densities.
-This continues until a density of approximately 2.9 (1,000 people per sq.km).
-At higher densities than that $R_0$ begins to decline, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
+# Proceed with caution: Spatial effects ahead!
 
-Thus, other things being equal, the effect of density in a county like Charlottesville in Virginia	 (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
-In contrast, the effect of density on $R_0$ in a county like Arlington in Virginia (density ~3,093 people per sq.km) is _stronger_ than either of the previous two examples.
-Lastly, the density of counties like San Francisco in California, or Queens and Bronx in NY, which are among the densest in the US, contributes even less to $R_0$ than even the most rural counties in the country.
+Words go here.
 
-It is worth at this point to recall Cressie's dictum about modelling: "[w]hat is one person's mean structure could be another person's correlation structure" [@Cressie1989geostatistics, p.
-201].
-There are almost always multiple ways to approach a modelling situation.
-In the present case, we would argue that spatial sampling is an important aspect of the modeling process, but one that perhaps required different technical skills than those available to SWN.
-There is nothing wrong with that.
-What matters is that, by adopting relatively high reproducibility standards, these researchers made a valuable and honest contribution to the collective enterprise of seeking knowledge.
-Their effort, and subsequent efforts to validate and expand on their work, can potentially contribute to provide clarity to ongoing conversations about the relevance of density and the spread of COVID-19.
+```{r selection-model-with-spatial-filter, include=FALSE}
 
-In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [as put by Wong and Li, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading].
-At the same time, this also stands in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with in a longitudinal setting [@Hamidi2020longitudinal.
-In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al.
-[-@Fielding2020social] might be onto something with respect to the difference between rural and urban counties.
-More generally, in population-level studies, density is indicative of proximity, no doubt about that, but also for adaptive behavior.
-And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.
+# Refit model selection2 to begin 
+selection3 <- selection(Ys ~ density + density_2 + private + hincome, R ~ density + density_2 + private + hincome, data = county_geo_clean)
 
-```{r comparison-results, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2."}
-# The maximum in-sample density is 6.255358
-# county_geo_clean %>% st_drop_geometry() %>% filter(R > 0) %>% summary()
+# Extract the residuals of the model. These residuals have NAs where the outcome was not observed
+res_selection3 <- residuals(selection3)
 
-example_df <- data.frame(density = seq(0, 6.3, 0.01)) %>% 
-  mutate(density_log = log(density),
-         density_2 = density^2, 
-         b_swn = model3$coefficients$fixed["density_log"] * density_log,
-         b_selection = selection2$estimate[7] * density + selection2$estimate[8] * density_2)
+#Initialize a spatial filter:
+SF <- numeric(length = length(res_selection3))
 
-# Plot of coefficient of density to R_0 according to SWN Model 3
-density_swn <- ggplot(data = example_df) + 
-  geom_line(aes(x = density, 
-                y = b_swn)) +
-  ggtitle("SWN Model 3") +
-  ylab(expression(paste("E[", R[0], "]"))) +
-  theme_minimal()
+#Initialize tolerance, counter, and index:
+tol <- 0;
 
-# Plot of coefficient of density to R_0  
-density_selection <- ggplot(data = example_df) + 
-  geom_line(aes(x = density, 
-                y = b_selection)) +
-  ggtitle("Selection Model 2") +
-  ylab(expression(paste("E[", R[0], "]"))) +
-  theme_minimal()
+while(tol < 0.25){
+  #Initialize counter and index:
+  count <- 0;
+  SF_INDEX <- 0
 
-density_swn + density_selection
+  #Obtain filter to simulate the residuals
+  for(i in 1:length(county_mem)){
+    count <- count + 1
+    junkSF <- data.frame(SF = SF, V = county_mem[, count])
+    #remove columns with all zeros
+    junkSF <- Filter(function(x)!all(x == 0), junkSF)
+    #estimate model
+    junkmod <- lm(res_selection3 ~ .,
+                  data = cbind(res_selection3, junkSF))
+    junkmod_summary <- summary(junkmod) # Summary of model; needed to extract the pval of the filter
+    pvals<- junkmod_summary$coefficients[nrow(junkmod_summary$coefficients), 4]
+    #pvals <- get_pvalue(junkmod)
+    junkb <- coef(junkmod)
+    if(pvals <= 0.10){
+      SF_INDEX[count] <- count
+      SF <- as.matrix(junkSF) %*% junkb[(0+2):length(junkb)]
+      junkmod <- lm(res_selection3 ~ ., 
+                    data = data.frame(res_selection3, SF))
+    }
+  }
+  #SF_INDEX <- SF_INDEX[!is.na(SF_INDEX)]
+  Filter1 <- SF # Spatial filter for residuals
+  res_filter_cor <- cor(res_selection3, Filter1, use = "complete.obs")
+  Filter_mt <- moran.test(Filter1, county_listw)
+  tol <- Filter_mt$p.value
+  if(tol < 0.25){
+    selection3 <- selection(Ys ~ density + density_2 + private + hincome, 
+                            R ~ density + density_2 + private + hincome + Filter1, 
+                            data = county_geo_clean)
+  }  
+}
 ```
 
-```{r counties-by-density, include=FALSE}
-# the relationship between density and R_0 become negative in Selection Model 2 at about 5.74
+```{r prepare-selection3-results, include=FALSE}
+selection3_me <- selection3 %>%
+  summary()
 
-# Values of b_selection for a density similar a Philadelphia 
-example_df %>% filter(density > 4.10 & density < 4.14)
-# Densities with a similar b_selection
-example_df %>% filter(b_selection > 0.87 & b_selection < 0.89)
+selection3_me <- selection3_me$estimate %>%
+  as.data.frame()
 
-# Counties with densities with similar response to Philadelphia
-county_geo_clean %>% 
-  mutate(density = drop_units(density)) %>%
-  filter(density > 1.59 & density < 1.64) %>%
-  select(NAME.y, density)
+colnames(selection3_me) <- c("b", "se", "t", "p")
 
-# Counties with densities associated with the highest response to density
-county_geo_clean %>% 
-  mutate(density = drop_units(density)) %>%
-  filter(density > 2.70 & density < 3.10) %>%
-  select(NAME.y, density)
+selection3_me <- selection3_me %>%  
+  rownames_to_column(var = "Variable") %>% 
+  transmute(Variable,
+            b,
+            ll = b - 1.96 * se, 
+            ul = b + 1.96 * se)
 
-# Highly dense counties
-county_geo_clean %>% 
-  mutate(density = drop_units(density)) %>%
-  filter(density > 5.74) %>%
-  select(NAME.y)
+results_selection <- selection1_me %>%
+  full_join(selection2_me,
+            by = "Variable")
 
+colnames(selection3_me) <- c("Variable",
+                                 "b1",
+                                 "l1",
+                                 "u1")
+
+selection3_me <- selection3_me %>%
+  transmute(Variable,
+            b1 = ifelse(is.na(b1),
+                        " ", 
+                        round(b1, 3)),
+            ci1 = ifelse(is.na(l1), 
+                         " ",
+                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")))
+#%>%
+#  arrange(Order) %>%
+#  select(-Order)
+```
+
+```{r tabulate-sample-selection3-results, echo=FALSE}
+selection3_me  %>%
+  mutate(Variable = c("Intercept",
+                      "Density (1,000 per sq.km)",
+                      "Density squared",
+                      "Percent of private transportation", 
+                      "Median household income (10,000)",
+                      "Intercept",
+                      "Density (1,000 per sq.km)",
+                      "Density squared",
+                      "Percent of private transportation", 
+                      "Median household income (\\$10,000)",
+                      "Spatial filter",
+                      "$\\sigma$",
+                      "$\\rho$")) %>%
+  kable("latex",
+        digits = 3,
+        booktabs = TRUE,
+        escape = FALSE,
+        col.names = c("Variable",
+                      "$\\beta$", 
+                      "95\\% CI"),
+        align = c("lcccc"),
+        caption = "\\label{tab:selection-results}Estimation results of sample selection model with spatial filter") %>%
+  kable_styling(latex_options = c("scale_down")) %>% 
+  pack_rows(group_label = "Sample Selection Model", 1, 5) %>%
+  pack_rows(group_label = "Outcome Model", 6, 11) %>%
+  add_header_above(c(" ", "Selection Model 3" = 2))
 ```
@@
 
+# Discussion
+
+How relevant are the difference between the different model specifications presented above?
+Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 3, Selection Model 2, and Selection Model 3.
+The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1.
+The conclusion is that at higher densities, $R_0$ is _always_ higher.
+The right panel, in contrast, shows that, according to Selection Model 2, $R_0$ is zero when density is zero (as expected), and then it tends to increase at higher densities.
+This continues until a density of approximately 2.9 (1,000 people per sq.km).
+At higher densities than that $R_0$ begins to decline, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
+
+Thus, other things being equal, the effect of density in a county like Charlottesville in Virginia	 (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
+In contrast, the effect of density on $R_0$ in a county like Arlington in Virginia (density ~3,093 people per sq.km) is _stronger_ than either of the previous two examples.
+Lastly, the density of counties like San Francisco in California, or Queens and Bronx in NY, which are among the densest in the US, contributes even less to $R_0$ than even the most rural counties in the country.
+
+It is worth at this point to recall Cressie's dictum about modelling: "[w]hat is one person's mean structure could be another person's correlation structure" [@Cressie1989geostatistics, p.
+201].
+There are almost always multiple ways to approach a modelling situation.
+In the present case, we would argue that spatial sampling is an important aspect of the modeling process, but one that perhaps required different technical skills than those available to SWN.
+There is nothing wrong with that.
+What matters is that, by adopting relatively high reproducibility standards, these researchers made a valuable and honest contribution to the collective enterprise of seeking knowledge.
+Their effort, and subsequent efforts to validate and expand on their work, can potentially contribute to provide clarity to ongoing conversations about the relevance of density and the spread of COVID-19.
+
+In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [as put by Wong and Li, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading].
+At the same time, this also stands in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with in a longitudinal setting [@Hamidi2020longitudinal.
+In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al.
+[-@Fielding2020social] might be onto something with respect to the difference between rural and urban counties.
+More generally, in population-level studies, density is indicative of proximity, no doubt about that, but also for adaptive behavior.
+And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.
+
+```{r comparison-results, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2."}
+# The maximum in-sample density is 6.255358
+# county_geo_clean %>% st_drop_geometry() %>% filter(R > 0) %>% summary()
+
+example_df <- data.frame(density = seq(0, 6.3, 0.01)) %>% 
+  mutate(density_log = log(density),
+         density_2 = density^2, 
+         b_swn = model3$coefficients$fixed["density_log"] * density_log,
+         b_selection = selection2$estimate[7] * density + selection2$estimate[8] * density_2,
+         b_selection3 = selection3$estimate[7] * density + selection3$estimate[8] * density_2)
+
+# Plot of coefficient of density to R_0 according to SWN Model 3
+density_swn <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_swn)) +
+  ggtitle("SWN Model 3") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+# Plot of coefficient of density to R_0  
+density_selection <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_selection)) +
+  ggtitle("Selection Model 2") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+# Plot of coefficient of density to R_0  
+density_selection3 <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_selection3)) +
+  ggtitle("Selection Model 3") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+density_swn + density_selection + density_selection3
+```
+
+```{r counties-by-density, include=FALSE}
+# the relationship between density and R_0 become negative in Selection Model 2 at about 5.74
+
+# Values of b_selection for a density similar a Philadelphia 
+example_df %>% filter(density > 4.10 & density < 4.14)
+# Densities with a similar b_selection
+example_df %>% filter(b_selection > 0.87 & b_selection < 0.89)
+
+# Counties with densities with similar response to Philadelphia
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 1.59 & density < 1.64) %>%
+  select(NAME.y, density)
+
+# Counties with densities associated with the highest response to density
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 2.70 & density < 3.10) %>%
+  select(NAME.y, density)
+
+# Highly dense counties
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 5.74) %>%
+  select(NAME.y)
+
+```
+
 # Conclusion

### 2021-06-09 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Experiments with spatial filter in the outcome model of the Heckman selection model

@@
 
-How relevant is the difference between these different model
-specifications? Figure \ref{fig:comparison-results} shows the
-relationship between density and \(R_0\) implied by SWN Model 1 and
-Selection Model 2. The left panel of the figure shows the non-linear but
-monotonic relationship implied by SWN Model 1. The conclusion is that at
-higher densities, \(R_0\) is \emph{always} higher. The right panel, in
-contrast, shows that, according to Selection Model 2, \(R_0\) is zero
-when density is zero (as expected), and then it tends to increase at
-higher densities. This continues until a density of approximately 2.9
-(1,000 people per sq.km). At higher densities than that \(R_0\) begins
-to decline, and the relationship becomes negative at densities higher
-than approximately 5.7 (1,000 people per sq.km).
+\hypertarget{proceed-with-caution-spatial-effects-ahead}{%
+\section{Proceed with caution: Spatial effects
+ahead!}\label{proceed-with-caution-spatial-effects-ahead}}
+
+Words go here.
+
+\begin{table}
+
+\caption{\label{tab:tabulate-sample-selection3-results}\label{tab:selection-results}Estimation results of sample selection model with spatial filter}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lcc}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Selection Model 3} \\
+\cmidrule(l{3pt}r{3pt}){2-3}
+Variable & $\beta$ & 95\% CI\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{3}{l}{\textbf{Sample Selection Model}}\\
+\hspace{1em}Intercept & -7.304 & [-8.346, -6.262]\\
+\hspace{1em}Density (1,000 per sq.km) & 2.445 & [2.089, 2.802]\\
+\hspace{1em}Density squared & -0.380 & [-0.468, -0.292]\\
+\hspace{1em}Percent of private transportation & 0.056 & [0.046, 0.067]\\
+\hspace{1em}Median household income (10,000) & 0.318 & [0.28, 0.356]\\
+\addlinespace[0.3em]
+\multicolumn{3}{l}{\textbf{Outcome Model}}\\
+\hspace{1em}Intercept & 2.563 & [2.497, 2.629]\\
+\hspace{1em}Density (1,000 per sq.km) & 0.760 & [0.746, 0.774]\\
+\hspace{1em}Density squared & -0.133 & [-0.135, -0.13]\\
+\hspace{1em}Percent of private transportation & -0.011 & [-0.012, -0.011]\\
+\hspace{1em}Median household income (\$10,000) & 0.002 & [-0.001, 0.004]\\
+\hspace{1em}Spatial filter & 1.000 & [0.998, 1.001]\\
+$\sigma$ & 0.017 & [0.015, 0.019]\\
+$\rho$ & -0.304 & [-0.957, 0.349]\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+How relevant are the difference between the different model
+specifications presented above? Figure \ref{fig:comparison-results}
+shows the relationship between density and \(R_0\) implied by SWN Model
+3, Selection Model 2, and Selection Model 3. The left panel of the
+figure shows the non-linear but monotonic relationship implied by SWN
+Model 1. The conclusion is that at higher densities, \(R_0\) is
+\emph{always} higher. The right panel, in contrast, shows that,
+according to Selection Model 2, \(R_0\) is zero when density is zero (as
+expected), and then it tends to increase at higher densities. This
+continues until a density of approximately 2.9 (1,000 people per sq.km).
+At higher densities than that \(R_0\) begins to decline, and the
+relationship becomes negative at densities higher than approximately 5.7
+(1,000 people per sq.km).
 

### 2021-06-10 R0-Density-Reanalysis/Cover-Letter-v1.Rmd
Submission to Geographical Analysis

@@
 
-To the Editor:
+Dear Rachel:
 
-Please find attached the paper titled "Population density and the spread of the COVID-19 pandemic: a reproducible research example".
+Please find attached the paper titled "Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis".
 The question of population density has been of interest during the pandemic.
@@
 The source document is an R Markdown file, a self-contained document with the text and _all_ the analysis.
-These materials are publicly available in a repository.
+These materials will be publicly available in a repository.
+For review purposes, they are currently in an anonymous Drive folder, and also supplied as supplementary files for review purposes.
 
-I trust that you will find that the paper is technically correct, and suitable for publication.
+I trust that you will find that the paper is technically correct, and suitable for review.
 
@@
 
-Antonio Paez
+Antonio

### 2021-06-10 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Submission to Geographical Analysis

@@
 ---
-title: "The importance of reproducibility in geographical research: the case of population density and the spread of COVID-19"
+title: "Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis"
 author:
@@
   The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research.
-Of interest has been the associations between population density and the spread of the pandemic.
-In this paper, population density and the basic reproductive number of SARS-CoV-2 are examined in an example of reproducible research.
-Given the high stakes of the situation, it is essential that scientific activities, on which good policy depends, are as transparent and reproducible as possible.
-Reproducibility is key for the efficient operation of the self-correction mechanisms of science.
-Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example of the spread of COVID-19 in the US shows.
+Alas, much of the research in the literature lacks conditions to be reproducible, and recent publications on the association between population density and the basic reproductive number of SARS-CoV-2 are no exception.
+Relatively few papers share code and data sufficiently, which hinders not only verification but additional experimentation.
+In this paper, an example of reproducible research shows the potential of spatial analysis for epidemiology research during COVID-19.
+Transparency and openness means that independent researchers can, with relatively modest efforts, verify findings and use different approaches as appropriate.
+Given the high stakes of the situation, it is essential that scientific findings, on which good policy depends, are as robust as possible; as the empirical example shows, reproducibility is one of the keys to ensure this.
 
@@
 header-includes:
-    - \usepackage{setspace}\doublespacing
+    - \usepackage{setspace}\onehalfspacing
 ---
@@
 
-Given the heavy human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has termed this deluge of publications a "carnage of substandard research" [@Bramstedt2020carnage].
-Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [@Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty and somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
+Given the ruinous human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has termed the deluge of COVID-19 publications a "carnage of substandard research" [@Bramstedt2020carnage].
+Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [e.g., @Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty and somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
 As reported in the literature, only a woefully small proportion of published research was actually reproducible before the pandemic [@Iqbal2016reproducible; @Stodden2018empirical], and the situation does not appear to have changed substantially since [@Sumner2020reproducibility; @Gustot2020quality].
@@
 
-In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how]; the implications are the topic of ongoing research [@Sharifi2020covid].
-The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, brought a torrent of research into the associations between density and the spread of the pandemic.
-Some important questions hang on the results of these research efforts.
+In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how]; the implications of the pandemic for urban planning, design, and management are the topic of ongoing research [@Sharifi2020covid].
+The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent of research into the associations between density and the spread of the pandemic.
+he answers to some important questions hang on the results of these research efforts.
 For example, are lower density regions safer from the pandemic?
@@
 What are the implications of the pandemic for future urban planning and practice?
-Over the past year, numerous papers have sought to throw light into the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed.
+Over the past year, numerous papers have sought to throw light on the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed.
 Further, to complicate matters, precious few of these studies appear to be sufficiently open to support independent verification.
@@
 The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in research.
-This paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
+The present paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
 The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, will be available in a public repository^[For peer-review purposes, the data package and code are currently in an anonymous Drive folder: \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}].
@@
 As Rocklöv and Sjödin [@Rocklov2020high] note, however, mathematical models of contagion are valid at small-to-medium spatial scales (and presumably, small temporal scales too, such as time spent in restaurants, concert halls, cruises), and the results do not necessarily transfer to larger spatial units and different time scales.
-There are solid reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers-however, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place.
-Nonetheless, the idea that high density correlates with high transmission is so seemingly reasonable that it is often taken for granted even at larger scales [e.g., @Cruz2020exploring; @Micallef2020first].
+There are solid reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers.
+On the other hand, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place.
+Nonetheless, the idea that high density correlates with high transmission is so seemingly sensible that it is often taken for granted even at larger scales [e.g., @Cruz2020exploring; @Micallef2020first].
 At larger scales, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, e.g., @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.
@@
 [@Chauhan2021covid] have found that perception of risks in the US varies between rural, suburban, and urban residents, with rural residents in general expressing less concern about the virus.
-It is possible that people who listened to the message of leaders saying that they were safe because of low density may not have taken adequate precautions against the virus.
-People in dense places who could more directly observe the impact of the pandemic may have become overly cautious.
+It is possible that people who listened to the message of leaders saying that they were safe from the virus because of low density may not have taken adequate precautions.
+Conversely, people in dense places who could more directly observe the impact of the pandemic may have become overly cautious.
 Both Paez et al.
@@
 For their part, in their investigation of doubling time, White and Hébert-Dufresne [-@White2020state] identified a negative and significant correlation between population density and doubling time in US states.
-Likewise, [-@Khavarian2021high] fond a small negative (and significant) association between population density and COVID-19 morbidity in districts in Tehran.
-Finally, two of the most complete studies in the US [by Hamidi et al.; -@Hamidi2020longitudinal; -@Hamidi2020density] used an extensive set of controls to find negative and significant correlations between density and COVID-19 cases and fatalities at the level of counties in the US.
+Likewise, [-@Khavarian2021high] found a small negative (and significant) association between population density and COVID-19 morbidity in districts in Tehran.
+Finally, two of the most complete studies in the US [by Hamidi et al.
+-@Hamidi2020longitudinal; and -@Hamidi2020density] used an extensive set of controls to find negative and significant correlations between density and COVID-19 cases and fatalities at the level of counties in the US.
 
@@
 To pick an example (which will be further elaborated in later sections of this paper), the study of Sy et al.
-[-@Sy2021population; hereafter SWN] would immediately grab the attention of a researcher with a somewhat different toolbox.
+[-@Sy2021population; hereafter SWN] would immediately grab the attention of a researcher with expertise in spatial analysis.
 
@@
 3).
-A mixed linear model accounts for this by introducing random components (in the case of SWN, random intercepts at the state level).
+A mixed linear model accounts for this by introducing random components; in the case of SWN, these are random intercepts at the state level.
 SWN estimated various models with different combinations of variables, including median household income and prevalence of travel by private transportation.
@@
 Moreover, they also conducted various sensitivity analyses.
-After these efforts, SWN conclude that there is a positive association between the basic reproductive number and population density at the level of counties in the US.
+After these efforts, SWN concluded that there is a positive association between the basic reproductive number and population density at the level of counties in the US.
 
@@
 As researchers do, SWN made modelling decisions, in this case basing their analysis only on counties with valid observations.
-A modeler with expertise in different methods would likely ask some of the following questions on reading SWN's paper: how were missing counties treated?
+A modeler with expertise in spatial analysis would likely ask some of the following questions on reading SWN's paper: how were missing counties treated?
 What are the implications of the spatial sampling framework used in the analysis?
 Is it possible to spatially interpolate the missing observations?
+Was there spatial residual autocorrelation in the models, or was the use of mixed models sufficient to capture spatial dependencies?
 These questions are relevant and their implications important.
@@
 Then there are those papers that more closely comply with reproducibility standards, and share well-documented processes and data, as well as the code used in any analyses reported [@Paez2020spatio; @Feyman2020effectiveness; @Stephens2021impact; @White2020state; @Sy2021population].
+Even in this case, the pressure to publish "new findings" instead of replication studies can act as a deterrent, perhaps particularly for younger researchers^[The present paper was desk rejected by three journals that had previously published research on population density and the spread of COVID-19; in one case, the paper was too opinionated for the journal, in the other two cases, the paper was not a "good fit" despite dealing with a nearly identical issue as papers previously published in said journals.
+This does not inspire much confidence in the commitment of journals to reproducibility in research.].
 
@@
 
-In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of an independent researcher are asked, and the data are reanalyzed.
+In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of an independent researcher with expertise in spatial analysis are asked, and the data are reanalyzed.
 
@@
 Table \ref{tab:swn-results} reproduces the first three models of SWN (the fourth model did not have any significant variables; see Table 1 in SWN).
-It is possible to verify that the results match, with only the minor (and irrelevant) exception of the magnitude of the coefficient for travel by private transportation, which is due to a difference in the input (here the variable is one percent units, whereas in SWN it was ten percent units).
+It is possible to verify that the results match, with only the minor (and irrelevant) exception of the magnitude of the coefficient for travel by private transportation, which is due to a difference in the input (here the variable is changed to one percent units, instead of the ten percent units used by SWN).
 The mixed linear model gives random intercepts (i.e., the intercept is a random variable), and the standard deviation is reported in the fourth row of Table \ref{tab:swn-results}.
@@
 This impression is reinforced when we overlay the boundaries of urban areas with population greater than 50,000 on the counties with valid values of $R_0$ (see Figure \ref{fig:urban-areas-map}).
-The fact that $R_0$ could not be accurately computed in many counties without large urban areas does not mean that there was no transmission of the virus: it simply means that we do not know with precision whether that was the case.
+The fact that $R_0$ could not be accurately computed in many counties without large urban areas does not mean that there was no transmission of the virus: it simply means that we do not know with sufficient precision to what extent that was the case.
 The low number of cases may be related to low population and/or low population density.
-This is intriguing, to say the least: by excluding cases based on the ability to calculate $R_0$ we are potentially _censoring_ the sample in a non-random way.
+This is intriguing, to say the least: by excluding cases based on the ability to calculate $R_0$ we are potentially _selecting_ the sample in a non-random way.
 
@@
 
-A problematic issue with non-random sample selection is that parameter estimates can become unreliable, and numerous techniques have been developed over time to address this.
+A problematic issue with non-random sample selection is that parameter estimates can become unreliable, and numerous techniques have been developed to address this.
 A model useful for sample selection problems is Heckman's selection model [see @Maddala1983limited].
 The selection model is in fact a system of two equations, as follows:
-
 $$
@@
 Vectors $x_i^S$ and $x_i^O$ are explanatory variables (with the possibility that $x_i^S = x_i^S$).
-Both equations include random terms (i.e., $\epsilon_i^S$ and $\epsilon_i^O$)  The first equation is designed to model the _probability_ of sampling, and the second equation the outcome of interest (say $R_0$).
+Both equations include random terms (i.e., $\epsilon_i^S$ and $\epsilon_i^O$).
+The first equation is designed to model the _probability_ of sampling, and the second equation the outcome of interest (say $R_0$).
 The random terms are jointly distributed and correlated with parameter $\rho$.
@@
 This is Sample Selection Model 1 in Table \ref{tab:selection-results}.
-The first thing to notice about this model is that the sample selection process and the outcome are not independent ($\rho\ne0$ with 5% of confidence).
+The first thing to notice about this model is that the sample selection process and the outcome are correlated ($\rho\ne0$ with 5% of confidence).
 The selection equation indicates that the probability of a county to be in the sample increases with population density (but at a decreasing rate due to the log-transformation), when travel by private modes is more prevalent, and as median household income in the county is higher.
-This is in line with the impression left by Figure \ref{fig:urban-areas-map} that counties with reliable values of $R_0$ tended to be those with larger urban centers.
+This is in line with the impression made by Figure \ref{fig:urban-areas-map} that counties with reliable values of $R_0$ tended to be those with larger urban centers.
 Once that the selection probabilities are accounted for in the model, several things happen with the outcomes model.
@@
 As seen in the table, Selection Model 2 replaces the log-transformation of population density with a quadratic expansion.
-The results of this analysis indicate that with this variable transformation, the selection and outcome processes are still not independent ($\rho\ne0$ with 5% of confidence).
-But a few interesting things emerge.
+The results of this analysis indicate that with this variable transformation, the selection and outcome processes are still correlated ($\rho\ne0$ with 5% of confidence).
+But a few other interesting things emerge.
 When we examine the outcomes model, we see that the quadratic expansion has a positive coefficient for the first order term, but a negative coefficient for the second order term.
-This indicates that $R_0$ initially tends to increase with higher density, but only up to a point, after which the negative second term (which grows more rapidly due to the square), becomes increasingly dominant.
+This indicates that $R_0$ initially tends to increase as density grows, but only up to a point, after which the negative second term (which grows more rapidly due to the square), becomes increasingly dominant.
 Secondly, the sign of the coefficient for travel by private transportation becomes negative again.
 This, of course, makes more sense than the positive sign of Selection Model 1: if people tend to travel in private transportation, the potential for contact should be lower instead of higher.
-And finally median household income is no longer significant.
+And finally median household income is no longer significant, similar to SWN Model 3.
 
@@
 
-# Proceed with caution: Spatial effects ahead!
-
-Words go here.
+# Proceed with caution: spatial effects ahead
+
+The results of the selection models, in particular Selection Model 2, make us reassess the original conclusion that density has a positive association with the basic reproductive number of COVID-19.
+A spatial analyst might still wonder about spatial residual autocorrelation.
+A challenge here is that spatial models tend to be technically more demanding, and although spatial models for qualitative variables exist, a spatial implementation of the sample selection model does not appear to exist.
+It might be argued that a reproducible research project can also allow a researcher to be more adventurous with their modeling decisions: since data and code are shared, other researchers can promptly and with relative ease poke the methods and see if they appear to be sound.
+
+In the present case, it appears that an application of spatial filtering [see @Getis2002comparative; @Paez2019using; @Griffith2004spatial] can help.
+Spatial filtering provides an elegant solution to regression problems that may have difficulties handling the spatial structures of spatial statistical and econometric models [@Griffith2000linear].
+A key issue in the present example is the fact that there are numerous missing observations, which prevents the calculation of autocorrelation statistics, let alone the estimation of models with spatial components.
+
+The following is an unorthodox, but potentially effective use of filters in a sample selection model:
+
+1. Estimate a sample selection model and retrieve the residuals of the outcome. This will be a vector with missing values for locations that were not sampled.
+
+2. Fit a spatial filter to the residuals. This is done by regressing the estimated residuals of the _observed_ data on the corresponding values of the Moran eigenvectors.
+
+3. The resulting filter will correlate highly with the known residuals, and will provide information in non-sampled locations that is consistent with the spatial pattern of the known residuals.
+
+4. Test the filter for spatial autocorrelation:
+
+    4.1 If significant spatial autocorrelation is detected, this would be indicative of residual spatial pattern.
+Introduce the filter as a covariate in the outcome model of the sample selection model and return to step 1.
+
+    4.2 If no significant spatial autocorrelation is detected, this would be indicative of random residual pattern. Stop.
+
+This procedure is implemented using a stopping criterion whereby the search for the filter only stops when the p-value of Moran's Coefficient of the filter fitted to the residuals is greater than 0.25, which was chosen as a sufficiently conservative value for testing for autocorrelation.
+The correlation of the known residuals with the corresponding elements of the filter is consistently high (the correlation coefficient typically is greater than 0.9).
+The results of implementing this procedure appear in Table \ref{tab:selection3-results} as Selection Model 3.
+The results are consistent with Selection Model 2, with two intriguing differences: 1) the variance of Sample Model 3 is smaller; and 2) the sample and outcome processes are no longer correlated (the confidence interval of $\rho$ includes zero).
+It appears that by capturing the spatial pattern of the residuals, which is likely strongly determined by the non-random sampling framework, the outcome model is not only substantially more precise, but also appears to be independent from the selection process.
 
@@
         align = c("lcccc"),
-        caption = "\\label{tab:selection-results}Estimation results of sample selection model with spatial filter") %>%
-  kable_styling(latex_options = c("scale_down")) %>% 
+        caption = "\\label{tab:selection3-results}Estimation results of sample selection model with spatial filter") %>%
+  kable_styling(font_size = 8, latex_options = "hold_position") %>% 
   pack_rows(group_label = "Sample Selection Model", 1, 5) %>%
@@
 
-How relevant are the difference between the different model specifications presented above?
+How relevant are the differences between the various model specifications presented above?
 Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 3, Selection Model 2, and Selection Model 3.
@@
 The conclusion is that at higher densities, $R_0$ is _always_ higher.
-The right panel, in contrast, shows that, according to Selection Model 2, $R_0$ is zero when density is zero (as expected), and then it tends to increase at higher densities.
+The two panels on the right, in contrast, shows that Selection Model 2 and Selection Model 3 coincide that $R_0$ tends to increase as density grows.
 This continues until a density of approximately 2.9 (1,000 people per sq.km).
-At higher densities than that $R_0$ begins to decline, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
-
-Thus, other things being equal, the effect of density in a county like Charlottesville in Virginia	 (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
+At higher densities than that the relationship between density and $R_0$ begins to weaken, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
+
+To put this into context, other things being equal, the effect of density in a county like Charlottesville in Virginia (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
 In contrast, the effect of density on $R_0$ in a county like Arlington in Virginia (density ~3,093 people per sq.km) is _stronger_ than either of the previous two examples.
@@
 In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [as put by Wong and Li, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading].
-At the same time, this also stands in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with in a longitudinal setting [@Hamidi2020longitudinal.
+At the same time, this also stands in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with it in a longitudinal setting [@Hamidi2020longitudinal.
 In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al.
 [-@Fielding2020social] might be onto something with respect to the difference between rural and urban counties.
-More generally, in population-level studies, density is indicative of proximity, no doubt about that, but also for adaptive behavior.
+More generally, there is no doubt that in population-level studies density is indicative of proximity, but it also potentially is a proxy for adaptive behavior.
 And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.
@@
 
-The tension between the need to publish research potentially useful in dealing with a global pandemic, and a "carnage of substandard research" [@Bramstedt2020carnage], highlights the importance of efforts to maintain the quality of scientific outputs during COVID-19.
+The tension between the need to publish research potentially useful in dealing with a global pandemic, and a potential "carnage of substandard research" [@Bramstedt2020carnage], highlights the importance of efforts to maintain the quality of scientific outputs during COVID-19.
 An important part of quality control is the ability of independent researchers to verify and examine the results of materials published in the literature.

### 2021-06-10 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Submission to Geographical Analysis

@@
             pdfauthor={},
-            pdftitle={The importance of reproducibility in geographical research: the case of population density and the spread of COVID-19},
+            pdftitle={Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis},
             colorlinks=false,
@@
 % Pandoc header
-\usepackage{setspace}\doublespacing
+\usepackage{setspace}\onehalfspacing
 \usepackage{booktabs}
@@
 
-  \title{The importance of reproducibility in geographical research: the
-case of population density and the spread of COVID-19}
+  \title{Reproducibility of research during COVID-19: examining the case
+of population density and the basic reproductive rate from the
+perspective of spatial analysis}
     \author[University]{Author\corref{1}}
@@
   COVID-19 pandemic has led to explosive growth in scientific research.
-  Of interest has been the associations between population density and
-  the spread of the pandemic. In this paper, population density and the
-  basic reproductive number of SARS-CoV-2 are examined in an example of
-  reproducible research. Given the high stakes of the situation, it is
-  essential that scientific activities, on which good policy depends,
-  are as transparent and reproducible as possible. Reproducibility is
-  key for the efficient operation of the self-correction mechanisms of
-  science. Transparency and openness means that the same problem can,
-  with relatively modest efforts, be examined by independent researchers
-  who can verify findings, and bring to bear different perspectives,
-  approaches, and methods---sometimes with consequential changes in the
-  conclusions, as the empirical example of the spread of COVID-19 in the
-  US shows.
+  Alas, much of the research in the literature lacks conditions to be
+  reproducible, and recent publications on the association between
+  population density and the basic reproductive number of SARS-CoV-2 are
+  no exception. Relatively few papers share code and data sufficiently,
+  which hinders not only verification but additional experimentation. In
+  this paper, an example of reproducible research shows the potential of
+  spatial analysis for epidemiology research during COVID-19.
+  Transparency and openness means that independent researchers can, with
+  relatively modest efforts, verify findings and use different
+  approaches as appropriate. Given the high stakes of the situation, it
+  is essential that scientific findings, on which good policy depends,
+  are as robust as possible; as the empirical example shows,
+  reproducibility is one of the keys to ensure this.
   \end{abstract}
@@
 
-Given the heavy human and economic cost of the pandemic, there has been
-a natural tension in the scientific community between the need to
+Given the ruinous human and economic cost of the pandemic, there has
+been a natural tension in the scientific community between the need to
 publish research results quickly and the imperative to maintain
 consistently high quality standards in scientific reporting; indeed, a
-call for maintaining the standards in published research has termed this
-deluge of publications a ``carnage of substandard research'' (Bramstedt,
-2020). Part of the challenge of maintaining quality standards in
-published research is that, despite an abundance of recommendations and
-guidelines (Broggini et al., 2017; Brunsdon and Comber, 2020; Ince et
-al., 2012; Ioannidis et al., 2014), in practice reproducibility has
-remained a lofty and somewhat aspirational goal (Konkol et al., 2019;
-Konkol and Kray, 2019). As reported in the literature, only a woefully
-small proportion of published research was actually reproducible before
-the pandemic (Iqbal et al., 2016; Stodden et al., 2018), and the
-situation does not appear to have changed substantially since (Gustot,
-2020; Sumner et al., 2020).
+call for maintaining the standards in published research has termed the
+deluge of COVID-19 publications a ``carnage of substandard research''
+(Bramstedt, 2020). Part of the challenge of maintaining quality
+standards in published research is that, despite an abundance of
+recommendations and guidelines (e.g., Broggini et al., 2017; Brunsdon
+and Comber, 2020; Ince et al., 2012; Ioannidis et al., 2014), in
+practice reproducibility has remained a lofty and somewhat aspirational
+goal (Konkol et al., 2019; Konkol and Kray, 2019). As reported in the
+literature, only a woefully small proportion of published research was
+actually reproducible before the pandemic (Iqbal et al., 2016; Stodden
+et al., 2018), and the situation does not appear to have changed
+substantially since (Gustot, 2020; Sumner et al., 2020).
 
@@
 numerous thinkers has been the prospect of life in cities after the
-pandemic (Florida et al., 2020); the implications are the topic of
-ongoing research (Sharifi and Khavarian-Garmsir, 2020). The fact that
-the worst of the pandemic was initially felt in dense population centers
-such as Wuhan, Milan, Madrid, and New York, brought a torrent of
-research into the associations between density and the spread of the
-pandemic. Some important questions hang on the results of these research
+pandemic (Florida et al., 2020); the implications of the pandemic for
+urban planning, design, and management are the topic of ongoing research
+(Sharifi and Khavarian-Garmsir, 2020). The fact that the worst of the
+pandemic was initially felt in dense population centers such as Wuhan,
+Milan, Madrid, and New York, unleashed a torrent of research into the
+associations between density and the spread of the pandemic. he answers
+to some important questions hang on the results of these research
 efforts. For example, are lower density regions safer from the pandemic?
@@
 for future urban planning and practice? Over the past year, numerous
-papers have sought to throw light into the underlying issue of density
-and the pandemic; nonetheless the results, as will be detailed next,
-remain mixed. Further, to complicate matters, precious few of these
-studies appear to be sufficiently open to support independent
-verification.
+papers have sought to throw light on the underlying issue of density and
+the pandemic; nonetheless the results, as will be detailed next, remain
+mixed. Further, to complicate matters, precious few of these studies
+appear to be sufficiently open to support independent verification.
 
@@
 researchers to redouble their efforts to increase transparency and
-reproducibility in research. This paper, in addition, aims to show how
-data can be packaged in well-documented, shareable units, and code can
-be embedded into self-contained documents suitable for review and
-independent verification. The source for this paper is an
+reproducibility in research. The present paper, in addition, aims to
+show how data can be packaged in well-documented, shareable units, and
+code can be embedded into self-contained documents suitable for review
+and independent verification. The source for this paper is an
 \href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
@@
 solid reasons for this: while in a restaurant, one can hardly avoid
-being in proximity to other customers-however, a person can choose to
-(or be forced to as a matter of policy) not go to a restaurant in the
-first place. Nonetheless, the idea that high density correlates with
-high transmission is so seemingly reasonable that it is often taken for
-granted even at larger scales (e.g., Cruz et al., 2020; Micallef et al.,
-2020). At larger scales, however, there exists the possibility of
+being in proximity to other customers. On the other hand, a person can
+choose to (or be forced to as a matter of policy) not go to a restaurant
+in the first place. Nonetheless, the idea that high density correlates
+with high transmission is so seemingly sensible that it is often taken
+for granted even at larger scales (e.g., Cruz et al., 2020; Micallef et
+al., 2020). At larger scales, however, there exists the possibility of
 behavioral adaptations, which are difficult to capture in the
@@
 possible that people who listened to the message of leaders saying that
-they were safe because of low density may not have taken adequate
-precautions against the virus. People in dense places who could more
+they were safe from the virus because of low density may not have taken
+adequate precautions. Conversely, people in dense places who could more
 directly observe the impact of the pandemic may have become overly
@@
 identified a negative and significant correlation between population
-density and doubling time in US states. Likewise, (2021) fond a small
+density and doubling time in US states. Likewise, (2021) found a small
 negative (and significant) association between population density and
 COVID-19 morbidity in districts in Tehran. Finally, two of the most
-complete studies in the US {[}by Hamidi et al.; (2020a); (2020b){]} used
-an extensive set of controls to find negative and significant
-correlations between density and COVID-19 cases and fatalities at the
-level of counties in the US.
+complete studies in the US (by Hamidi et al. 2020a, and 2020b) used an
+extensive set of controls to find negative and significant correlations
+between density and COVID-19 cases and fatalities at the level of
+counties in the US.
 
@@
 et al.~{[}(2021); hereafter SWN{]} would immediately grab the attention
-of a researcher with a somewhat different toolbox.
+of a researcher with expertise in spatial analysis.
 
@@
 systems across states'' (p.~3). A mixed linear model accounts for this
-by introducing random components (in the case of SWN, random intercepts
-at the state level). SWN estimated various models with different
-combinations of variables, including median household income and
-prevalence of travel by private transportation. These are sensible
+by introducing random components; in the case of SWN, these are random
+intercepts at the state level. SWN estimated various models with
+different combinations of variables, including median household income
+and prevalence of travel by private transportation. These are sensible
 controls, given potential variations in behavior: people in more
@@
 they also conducted various sensitivity analyses. After these efforts,
-SWN conclude that there is a positive association between the basic
+SWN concluded that there is a positive association between the basic
 reproductive number and population density at the level of counties in
@@
 analysis only on counties with valid observations. A modeler with
-expertise in different methods would likely ask some of the following
+expertise in spatial analysis would likely ask some of the following
 questions on reading SWN's paper: how were missing counties treated?
@@
 analysis? Is it possible to spatially interpolate the missing
-observations? These questions are relevant and their implications
+observations? Was there spatial residual autocorrelation in the models,
+or was the use of mixed models sufficient to capture spatial
+dependencies? These questions are relevant and their implications
 important. Fortunately, SWN are an example of a reasonably open,
@@
 2020; Stephens et al., 2021; Sy et al., 2021; White and Hébert-Dufresne,
-2020).
+2020). Even in this case, the pressure to publish ``new findings''
+instead of replication studies can act as a deterrent, perhaps
+particularly for younger researchers\footnote{The present paper was desk
+  rejected by three journals that had previously published research on
+  population density and the spread of COVID-19; in one case, the paper
+  was too opinionated for the journal, in the other two cases, the paper
+  was not a ``good fit'' despite dealing with a nearly identical issue
+  as papers previously published in said journals. This does not inspire
+  much confidence in the commitment of journals to reproducibility in
+  research.}.
 
 In the following sections, the analysis of RWN is reproduced, some
-relevant questions from the perspective of an independent researcher are
-asked, and the data are reanalyzed.
+relevant questions from the perspective of an independent researcher
+with expertise in spatial analysis are asked, and the data are
+reanalyzed.
 
@@
 travel by private transportation, which is due to a difference in the
-input (here the variable is one percent units, whereas in SWN it was ten
-percent units). The mixed linear model gives random intercepts (i.e.,
-the intercept is a random variable), and the standard deviation is
-reported in the fourth row of Table \ref{tab:swn-results}. It is useful
-to map the random intercepts: as seen in Figure
+input (here the variable is changed to one percent units, instead of the
+ten percent units used by SWN). The mixed linear model gives random
+intercepts (i.e., the intercept is a random variable), and the standard
+deviation is reported in the fourth row of Table \ref{tab:swn-results}.
+It is useful to map the random intercepts: as seen in Figure
 \ref{fig:random-terms-map}, other things being equal, counties in Texas
@@
 mean that there was no transmission of the virus: it simply means that
-we do not know with precision whether that was the case. The low number
-of cases may be related to low population and/or low population density.
-This is intriguing, to say the least: by excluding cases based on the
-ability to calculate \(R_0\) we are potentially \emph{censoring} the
-sample in a non-random way.
+we do not know with sufficient precision to what extent that was the
+case. The low number of cases may be related to low population and/or
+low population density. This is intriguing, to say the least: by
+excluding cases based on the ability to calculate \(R_0\) we are
+potentially \emph{selecting} the sample in a non-random way.
 
@@
 estimates can become unreliable, and numerous techniques have been
-developed over time to address this. A model useful for sample selection
-problems is Heckman's selection model (see Maddala, 1983). The selection
-model is in fact a system of two equations, as follows:
-
-\[
+developed to address this. A model useful for sample selection problems
+is Heckman's selection model (see Maddala, 1983). The selection model is
+in fact a system of two equations, as follows: \[
 \begin{array}{c}
@@
 that \(x_i^S = x_i^S\)). Both equations include random terms (i.e.,
-\(\epsilon_i^S\) and \(\epsilon_i^O\)) The first equation is designed to
-model the \emph{probability} of sampling, and the second equation the
+\(\epsilon_i^S\) and \(\epsilon_i^O\)). The first equation is designed
+to model the \emph{probability} of sampling, and the second equation the
 outcome of interest (say \(R_0\)). The random terms are jointly
@@
 \ref{tab:selection-results}. The first thing to notice about this model
-is that the sample selection process and the outcome are not independent
+is that the sample selection process and the outcome are correlated
 (\(\rho\ne0\) with 5\% of confidence). The selection equation indicates
@@
 as median household income in the county is higher. This is in line with
-the impression left by Figure \ref{fig:urban-areas-map} that counties
+the impression made by Figure \ref{fig:urban-areas-map} that counties
 with reliable values of \(R_0\) tended to be those with larger urban
@@
 analysis indicate that with this variable transformation, the selection
-and outcome processes are still not independent (\(\rho\ne0\) with 5\%
-of confidence). But a few interesting things emerge. When we examine the
-outcomes model, we see that the quadratic expansion has a positive
+and outcome processes are still correlated (\(\rho\ne0\) with 5\% of
+confidence). But a few other interesting things emerge. When we examine
+the outcomes model, we see that the quadratic expansion has a positive
 coefficient for the first order term, but a negative coefficient for the
 second order term. This indicates that \(R_0\) initially tends to
-increase with higher density, but only up to a point, after which the
+increase as density grows, but only up to a point, after which the
 negative second term (which grows more rapidly due to the square),
@@
 contact should be lower instead of higher. And finally median household
-income is no longer significant.
+income is no longer significant, similar to SWN Model 3.
 
@@
 \hypertarget{proceed-with-caution-spatial-effects-ahead}{%
-\section{Proceed with caution: Spatial effects
-ahead!}\label{proceed-with-caution-spatial-effects-ahead}}
-
-Words go here.
-
-\begin{table}
-
-\caption{\label{tab:tabulate-sample-selection3-results}\label{tab:selection-results}Estimation results of sample selection model with spatial filter}
+\section{Proceed with caution: spatial effects
+ahead}\label{proceed-with-caution-spatial-effects-ahead}}
+
+The results of the selection models, in particular Selection Model 2,
+make us reassess the original conclusion that density has a positive
+association with the basic reproductive number of COVID-19. A spatial
+analyst might still wonder about spatial residual autocorrelation. A
+challenge here is that spatial models tend to be technically more
+demanding, and although spatial models for qualitative variables exist,
+a spatial implementation of the sample selection model does not appear
+to exist. It might be argued that a reproducible research project can
+also allow a researcher to be more adventurous with their modeling
+decisions: since data and code are shared, other researchers can
+promptly and with relative ease poke the methods and see if they appear
+to be sound.
+
+In the present case, it appears that an application of spatial filtering
+(see Getis and Griffith, 2002; Griffith, 2004; Paez, 2019) can help.
+Spatial filtering provides an elegant solution to regression problems
+that may have difficulties handling the spatial structures of spatial
+statistical and econometric models (Griffith, 2000). A key issue in the
+present example is the fact that there are numerous missing
+observations, which prevents the calculation of autocorrelation
+statistics, let alone the estimation of models with spatial components.
+
+The following is an unorthodox, but potentially effective use of filters
+in a sample selection model:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Estimate a sample selection model and retrieve the residuals of the
+  outcome. This will be a vector with missing values for locations that
+  were not sampled.
+\item
+  Fit a spatial filter to the residuals. This is done by regressing the
+  estimated residuals of the \emph{observed} data on the corresponding
+  values of the Moran eigenvectors.
+\item
+  The resulting filter will correlate highly with the known residuals,
+  and will provide information in non-sampled locations that is
+  consistent with the spatial pattern of the known residuals.
+\item
+  Test the filter for spatial autocorrelation:
+
+  4.1 If significant spatial autocorrelation is detected, this would be
+  indicative of residual spatial pattern. Introduce the filter as a
+  covariate in the outcome model of the sample selection model and
+  return to step 1.
+
+  4.2 If no significant spatial autocorrelation is detected, this would
+  be indicative of random residual pattern. Stop.
+\end{enumerate}
+
+This procedure is implemented using a stopping criterion whereby the
+search for the filter only stops when the p-value of Moran's Coefficient
+of the filter fitted to the residuals is greater than 0.25, which was
+chosen as a sufficiently conservative value for testing for
+autocorrelation. The correlation of the known residuals with the
+corresponding elements of the filter is consistently high (the
+correlation coefficient typically is greater than 0.9). The results of
+implementing this procedure appear in Table \ref{tab:selection3-results}
+as Selection Model 3. The results are consistent with Selection Model 2,
+with two intriguing differences: 1) the variance of Sample Model 3 is
+smaller; and 2) the sample and outcome processes are no longer
+correlated (the confidence interval of \(\rho\) includes zero). It
+appears that by capturing the spatial pattern of the residuals, which is
+likely strongly determined by the non-random sampling framework, the
+outcome model is not only substantially more precise, but also appears
+to be independent from the selection process.
+
+\begin{table}[!h]
+
+\caption{\label{tab:tabulate-sample-selection3-results}\label{tab:selection3-results}Estimation results of sample selection model with spatial filter}
 \centering
-\resizebox{\linewidth}{!}{
+\fontsize{8}{10}\selectfont
 \begin{tabular}[t]{lcc}
@@
 \bottomrule
-\end{tabular}}
+\end{tabular}
 \end{table}
@@
 
-How relevant are the difference between the different model
+How relevant are the differences between the various model
 specifications presented above? Figure \ref{fig:comparison-results}
@@
 Model 1. The conclusion is that at higher densities, \(R_0\) is
-\emph{always} higher. The right panel, in contrast, shows that,
-according to Selection Model 2, \(R_0\) is zero when density is zero (as
-expected), and then it tends to increase at higher densities. This
-continues until a density of approximately 2.9 (1,000 people per sq.km).
-At higher densities than that \(R_0\) begins to decline, and the
-relationship becomes negative at densities higher than approximately 5.7
-(1,000 people per sq.km).
-
-Thus, other things being equal, the effect of density in a county like
-Charlottesville in Virginia (density \textasciitilde1,639 people per
-sq.km) is roughly the same as that in a county like Philadelphia
-(density \textasciitilde4,127 people per sq.km). In contrast, the effect
-of density on \(R_0\) in a county like Arlington in Virginia (density
-\textasciitilde3,093 people per sq.km) is \emph{stronger} than either of
-the previous two examples. Lastly, the density of counties like San
-Francisco in California, or Queens and Bronx in NY, which are among the
-densest in the US, contributes even less to \(R_0\) than even the most
-rural counties in the country.
+\emph{always} higher. The two panels on the right, in contrast, shows
+that Selection Model 2 and Selection Model 3 coincide that \(R_0\) tends
+to increase as density grows. This continues until a density of
+approximately 2.9 (1,000 people per sq.km). At higher densities than
+that the relationship between density and \(R_0\) begins to weaken, and
+the relationship becomes negative at densities higher than approximately
+5.7 (1,000 people per sq.km).
+
+To put this into context, other things being equal, the effect of
+density in a county like Charlottesville in Virginia (density
+\textasciitilde1,639 people per sq.km) is roughly the same as that in a
+county like Philadelphia (density \textasciitilde4,127 people per
+sq.km). In contrast, the effect of density on \(R_0\) in a county like
+Arlington in Virginia (density \textasciitilde3,093 people per sq.km) is
+\emph{stronger} than either of the previous two examples. Lastly, the
+density of counties like San Francisco in California, or Queens and
+Bronx in NY, which are among the densest in the US, contributes even
+less to \(R_0\) than even the most rural counties in the country.
 
@@
 the virus in a cross-sectional study (Hamidi et al., 2020b), or was
-negatively associated with in a longitudinal setting {[}Hamidi et al.
+negatively associated with it in a longitudinal setting {[}Hamidi et al.
 (2020a). In this sense, the conclusion that density does not aggravate
@@
 onto something with respect to the difference between rural and urban
-counties. More generally, in population-level studies, density is
-indicative of proximity, no doubt about that, but also for adaptive
-behavior. And it is possible that the determining factor during
-COVID-19, at least in the US, has been variations in perceptions of the
-risks associated with contagion (Chauhan et al., 2021), and subsequent
-compensations in behavior in more and less dense regions.
+counties. More generally, there is no doubt that in population-level
+studies density is indicative of proximity, but it also potentially is a
+proxy for adaptive behavior. And it is possible that the determining
+factor during COVID-19, at least in the US, has been variations in
+perceptions of the risks associated with contagion (Chauhan et al.,
+2021), and subsequent compensations in behavior in more and less dense
+regions.
 
@@
 The tension between the need to publish research potentially useful in
-dealing with a global pandemic, and a ``carnage of substandard
+dealing with a global pandemic, and a potential ``carnage of substandard
 research'' (Bramstedt, 2020), highlights the importance of efforts to
@@
 
+\leavevmode\hypertarget{ref-Getis2002comparative}{}%
+Getis, A., Griffith, D.A., 2002. Comparative spatial filtering in
+regression analysis. Geographical Analysis 34, 130--140.
+
 \leavevmode\hypertarget{ref-Gomez2021infekta}{}%
@@
 doi:\href{https://doi.org/10.1371/journal.pone.0245787}{10.1371/journal.pone.0245787}
+
+\leavevmode\hypertarget{ref-Griffith2000linear}{}%
+Griffith, D.A., 2000. A linear regression solution to the spatial
+autocorrelation problem. Journal of Geographical Systems 2, 141--156.
+
+\leavevmode\hypertarget{ref-Griffith2004spatial}{}%
+Griffith, D.A., 2004. A spatial filtering specification for the
+autologistic model. Environment and Planning A 36, 1791--1811.
 
@@
 
+\leavevmode\hypertarget{ref-Paez2019using}{}%
+Paez, A., 2019. Using spatial filters and exploratory data analysis to
+enhance regression models of spatial data. Geographical Analysis 51,
+314--338.
+doi:\href{https://doi.org/10.1111/gean.12180}{10.1111/gean.12180}
+
 \leavevmode\hypertarget{ref-Paez2020using}{}%

### 2021-06-10 README.Rmd
Reknitted README.md
Submission to Geographical Analysis

--- 2021-05-26 README.Rmd
@@
 
-# Population density and the spread of the COVID-19 pandemic: a reproducible research example
+# Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis
 
@@
 
-Paper submitted to *Landscape and Urban Planning*
+Paper submitted to [~~PLoS ONE~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/PLOS ONE Decision_ PONE-D-21-16874.pdf) [~~Science of the Total Environment~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Your Submission STOTEN-D-21-12077.pdf) [~~Landscape and Urban Planning~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Decision on submission to Landscape and Urban Planning.pdf) Geographical Analysis
 
@@
 The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research.
-Of interest has been the associations between population density and the spread of the pandemic.
-In this paper, population density and the basic reproductive number of SARS-CoV-2 are examined in an example of reproducible research.
-Given the high stakes of the situation, it is essential that scientific activities, on which good policy depends, are as transparent and reproducible as possible.
-Reproducibility is key for the efficient operation of the self-correction mechanisms of science.
-Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example of the spread of COVID-19 in the US shows.
+Alas, much of the research in the literature lacks conditions to be reproducible, and recent publications on the association between population density and the basic reproductive number of SARS-CoV-2 are no exception.
+Relatively few papers share code and data sufficiently, which hinders not only verification but additional experimentation.
+In this paper, an example of reproducible research shows the potential of spatial analysis for epidemiology research during COVID-19.
+Transparency and openness means that independent researchers can, with relatively modest efforts, verify findings and use different approaches as appropriate.
+Given the high stakes of the situation, it is essential that scientific findings, on which good policy depends, are as robust as possible; as the empirical example shows, reproducibility is one of the keys to ensure this.
 
@@
 
-```{r install-data-package, include = FALSE}
-# Run only once if needed to install data package `r0density`
-if (!require("r0density", character.only = TRUE)) {
-  devtools::install_github("https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed", subdir = "r0density")
-}
-```
+<!--
+Install data package as appropriate. The data package is file `r0density_0.1.0.tar.gz`
+-->
 
 ```{r load-packages, include=FALSE, cache=FALSE}
+library(adespatial) # To calculate Moran eigenvector maps
 library(censReg) # Package to estimate tobit model
-library(gridExtra)
 library(kableExtra) # Package for creating tables
@@
 
-Given the heavy human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has even called this deluge of publications a "carnage of substandard research" [@Bramstedt2020carnage].
-Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [@Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty and somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
+Given the ruinous human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has termed the deluge of COVID-19 publications a "carnage of substandard research" [@Bramstedt2020carnage].
+Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [e.g., @Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty and somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
 As reported in the literature, only a woefully small proportion of published research was actually reproducible before the pandemic [@Iqbal2016reproducible; @Stodden2018empirical], and the situation does not appear to have changed substantially since [@Sumner2020reproducibility; @Gustot2020quality].
@@
 
-In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how].
-The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, brought a torrent of research into the associations between density and the spread of the pandemic.
-Some important questions hang on the results of these research efforts.
+In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how]; the implications of the pandemic for urban planning, design, and management are the topic of ongoing research [@Sharifi2020covid].
+The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent of research into the associations between density and the spread of the pandemic.
+he answers to some important questions hang on the results of these research efforts.
 For example, are lower density regions safer from the pandemic?
@@
 And in the longer term, will the risks of life in high density regions presage a flight from cities?
-Over the past year, numerous papers have sought to throw light into the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed.
+What are the implications of the pandemic for future urban planning and practice?
+Over the past year, numerous papers have sought to throw light on the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed.
 Further, to complicate matters, precious few of these studies appear to be sufficiently open to support independent verification.
 
-The objective of this paper is to illustrate the importance of reproducibility in research, particularly in the context of the flood of COVID-19 papers.
+The objective of this paper is to illustrate the importance of reproducibility in research in the context of the flood of COVID-19 papers.
 To this end, a recent study by Sy et al.
-[@Sy2021population] is chosen as an example of reproducible research.
+[-@Sy2021population] is chosen as an example of reproducible research.
 The objective is not to malign the analysis of these researchers, but rather to demonstrate the value of openness to allow for independent verification and further analysis.
@@
 The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in research.
-This paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
-The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, is available in a public repository^[\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}].
+The present paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
+The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, will be available in a public repository^[For peer-review purposes, the data package and code are currently in an anonymous Drive folder: \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}].
 
@@
 As researchers do, SWN made modelling decisions, in this case basing their analysis only on counties with valid observations.
-A modeler with expertise in different methods would likely ask some of the following questions on reading SWN's paper: how were missing counties treated?
+A modeler with expertise in spatial analysis would likely ask some of the following questions on reading SWN's paper: how were missing counties treated?
 What are the implications of the spatial sampling framework used in the analysis?
 Is it possible to spatially interpolate the missing observations?
-And, was there evidence of spatial autocorrelation in the residuals of the models?
+Was there spatial residual autocorrelation in the models, or was the use of mixed models sufficient to capture spatial dependencies?
 These questions are relevant and their implications important.
@@
 
-In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of a spatial modeler are asked, and the data are reanalyzed.
+In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of an independent researcher with expertise in spatial analysis are asked, and the data are reanalyzed.
 
@@
                       "Within-group standard error")) %>%
-  kable("html",
+  kable("latex",
         digits = 3,
@@
 
-```{r random-terms-map, echo=FALSE, fig.cap="\\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
+```{r random-terms-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
 # Choropleth map of basic reproductive number
@@
 This impression is reinforced when we overlay the boundaries of urban areas with population greater than 50,000 on the counties with valid values of $R_0$ (see Figure \ref{fig:urban-areas-map}).
-The fact that $R_0$ could not be accurately computed in many counties without large urban areas does not mean that there was no transmission of the virus: it simply means that we do not know with precision whether that was the case.
+The fact that $R_0$ could not be accurately computed in many counties without large urban areas does not mean that there was no transmission of the virus: it simply means that we do not know with sufficient precision to what extent that was the case.
 The low number of cases may be related to low population and/or low population density.
-This is intriguing, to say the least: by excluding cases based on the ability to calculate $R_0$ we are potentially _censoring_ the sample in a non-random way.
+This is intriguing, to say the least: by excluding cases based on the ability to calculate $R_0$ we are potentially _selecting_ the sample in a non-random way.
 
-```{r urban-areas-map, echo=FALSE, fig.cap="\\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
+```{r urban-areas-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
 # Choropleth map of basic reproductive number
@@
                       "$\\rho$")) %>%
-  kable("html",
+  kable("latex",
         digits = 3,
@@
 
-How relevant is the difference between these different model specifications?
-Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 1 and Selection Model 2.
-The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1.
-The conclusion is that at higher densities, $R_0$ is _always_ higher.
-The right panel, in contrast, shows that, according to Selection Model 2, $R_0$ is zero when density is zero (as expected), and then it tends to increase at higher densities.
-This continues until a density of approximately 2.9 (1,000 people per sq.km).
-At higher densities than that $R_0$ begins to decline, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
+# Proceed with caution: spatial effects ahead
 
-Thus, other things being equal, the effect of density in a county like Charlottesville in Virginia	 (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
-In contrast, the effect of density on $R_0$ in a county like Arlington in Virginia (density ~3,093 people per sq.km) is _stronger_ than either of the previous two examples.
-Lastly, the density of counties like San Francisco in California, or Queens and Bronx in NY, which are among the densest in the US, contributes even less to $R_0$ than even the most rural counties in the country.
+The results of the selection models, in particular Selection Model 2, make us reassess the original conclusion that density has a positive association with the basic reproductive number of COVID-19.
+A spatial analyst might still wonder about spatial residual autocorrelation.
+A challenge here is that spatial models tend to be technically more demanding, and although spatial models for qualitative variables exist, a spatial implementation of the sample selection model does not appear to exist.
+It might be argued that a reproducible research project can also allow a researcher to be more adventurous with their modeling decisions: since data and code are shared, other researchers can promptly and with relative ease poke the methods and see if they appear to be sound.
 
-It is worth at this point to recall Cressie's dictum about modelling: "[w]hat is one person's mean structure could be another person's correlation structure" [@Cressie1989geostatistics, p.
-201].
-There are almost always multiple ways to approach a modelling situation.
-In the present case, we would argue that spatial sampling is an important aspect of the modeling process, but one that perhaps required different technical skills than those available to SWN.
-There is nothing wrong with that.
-What matters is that, by adopting relatively high reproducibility standards, these researchers made a valuable and honest contribution to the collective enterprise of seeking knowledge.
-Their effort, and subsequent efforts to validate and expand on their work, can potentially contribute to provide clarity to ongoing conversations about the relevance of density and the spread of COVID-19.
+In the present case, it appears that an application of spatial filtering [see @Getis2002comparative; @Paez2019using; @Griffith2004spatial] can help.
+Spatial filtering provides an elegant solution to regression problems that may have difficulties handling the spatial structures of spatial statistical and econometric models [@Griffith2000linear].
+A key issue in the present example is the fact that there are numerous missing observations, which prevents the calculation of autocorrelation statistics, let alone the estimation of models with spatial components.
 
-In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [as put by Wong and Li, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading].
-At the same time, this also stands in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with in a longitudinal setting [@Hamidi2020longitudinal.
-In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al.
-[-@Fielding2020social] might be onto something with respect to the difference between rural and urban counties.
-More generally, in population-level studies, density is indicative of proximity, no doubt about that, but also for adaptive behavior.
-And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.
+The following is an unorthodox, but potentially effective use of filters in a sample selection model:
 
-```{r comparison-results, echo=FALSE, fig.cap="\\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2."}
-# The maximum in-sample density is 6.255358
-# county_geo_clean %>% st_drop_geometry() %>% filter(R > 0) %>% summary()
+1. Estimate a sample selection model and retrieve the residuals of the outcome. This will be a vector with missing values for locations that were not sampled.
 
-example_df <- data.frame(density = seq(0, 6.3, 0.01)) %>% 
-  mutate(density_log = log(density),
-         density_2 = density^2, 
-         b_swn = model3$coefficients$fixed["density_log"] * density_log,
-         b_selection = selection2$estimate[7] * density + selection2$estimate[8] * density_2)
+2. Fit a spatial filter to the residuals. This is done by regressing the estimated residuals of the _observed_ data on the corresponding values of the Moran eigenvectors.
 
-# Plot of coefficient of density to R_0 according to SWN Model 3
-density_swn <- ggplot(data = example_df) + 
-  geom_line(aes(x = density, 
-                y = b_swn)) +
-  ggtitle("SWN Model 3") +
-  ylab(expression(paste("E[", R[0], "]"))) +
-  theme_minimal()
+3. The resulting filter will correlate highly with the known residuals, and will provide information in non-sampled locations that is consistent with the spatial pattern of the known residuals.
 
-# Plot of coefficient of density to R_0  
-density_selection <- ggplot(data = example_df) + 
-  geom_line(aes(x = density, 
-                y = b_selection)) +
-  ggtitle("Selection Model 2") +
-  ylab(expression(paste("E[", R[0], "]"))) +
-  theme_minimal()
+4. Test the filter for spatial autocorrelation:
 
-grid.arrange(density_swn, density_selection, nrow = 1)
+    4.1 If significant spatial autocorrelation is detected, this would be indicative of residual spatial pattern.
+Introduce the filter as a covariate in the outcome model of the sample selection model and return to step 1.
+
+    4.2 If no significant spatial autocorrelation is detected, this would be indicative of random residual pattern. Stop.
+
+This procedure is implemented using a stopping criterion whereby the search for the filter only stops when the p-value of Moran's Coefficient of the filter fitted to the residuals is greater than 0.25, which was chosen as a sufficiently conservative value for testing for autocorrelation.
+The correlation of the known residuals with the corresponding elements of the filter is consistently high (the correlation coefficient typically is greater than 0.9).
+The results of implementing this procedure appear in Table \ref{tab:selection3-results} as Selection Model 3.
+The results are consistent with Selection Model 2, with two intriguing differences: 1) the variance of Sample Model 3 is smaller; and 2) the sample and outcome processes are no longer correlated (the confidence interval of $\rho$ includes zero).
+It appears that by capturing the spatial pattern of the residuals, which is likely strongly determined by the non-random sampling framework, the outcome model is not only substantially more precise, but also appears to be independent from the selection process.
+
+```{r selection-model-with-spatial-filter, include=FALSE}
+
+# Refit model selection2 to begin 
+selection3 <- selection(Ys ~ density + density_2 + private + hincome, R ~ density + density_2 + private + hincome, data = county_geo_clean)
+
+# Extract the residuals of the model. These residuals have NAs where the outcome was not observed
+res_selection3 <- residuals(selection3)
+
+#Initialize a spatial filter:
+SF <- numeric(length = length(res_selection3))
+
+#Initialize tolerance, counter, and index:
+tol <- 0;
+
+while(tol < 0.25){
+  #Initialize counter and index:
+  count <- 0;
+  SF_INDEX <- 0
+
+  #Obtain filter to simulate the residuals
+  for(i in 1:length(county_mem)){
+    count <- count + 1
+    junkSF <- data.frame(SF = SF, V = county_mem[, count])
+    #remove columns with all zeros
+    junkSF <- Filter(function(x)!all(x == 0), junkSF)
+    #estimate model
+    junkmod <- lm(res_selection3 ~ .,
+                  data = cbind(res_selection3, junkSF))
+    junkmod_summary <- summary(junkmod) # Summary of model; needed to extract the pval of the filter
+    pvals<- junkmod_summary$coefficients[nrow(junkmod_summary$coefficients), 4]
+    #pvals <- get_pvalue(junkmod)
+    junkb <- coef(junkmod)
+    if(pvals <= 0.10){
+      SF_INDEX[count] <- count
+      SF <- as.matrix(junkSF) %*% junkb[(0+2):length(junkb)]
+      junkmod <- lm(res_selection3 ~ ., 
+                    data = data.frame(res_selection3, SF))
+    }
+  }
+  #SF_INDEX <- SF_INDEX[!is.na(SF_INDEX)]
+  Filter1 <- SF # Spatial filter for residuals
+  res_filter_cor <- cor(res_selection3, Filter1, use = "complete.obs")
+  Filter_mt <- moran.test(Filter1, county_listw)
+  tol <- Filter_mt$p.value
+  if(tol < 0.25){
+    selection3 <- selection(Ys ~ density + density_2 + private + hincome, 
+                            R ~ density + density_2 + private + hincome + Filter1, 
+                            data = county_geo_clean)
+  }  
+}
 ```
 
-```{r counties-by-density, include=FALSE}
-# the relationship between density and R_0 become negative in Selection Model 2 at about 5.74
+```{r prepare-selection3-results, include=FALSE}
+selection3_me <- selection3 %>%
+  summary()
 
-# Values of b_selection for a density similar a Philadelphia 
-example_df %>% filter(density > 4.10 & density < 4.14)
-# Densities with a similar b_selection
-example_df %>% filter(b_selection > 0.87 & b_selection < 0.89)
+selection3_me <- selection3_me$estimate %>%
+  as.data.frame()
 
-# Counties with densities with similar response to Philadelphia
-county_geo_clean %>% 
-  mutate(density = drop_units(density)) %>%
-  filter(density > 1.59 & density < 1.64) %>%
-  select(NAME.y, density)
+colnames(selection3_me) <- c("b", "se", "t", "p")
 
-# Counties with densities associated with the highest response to density
-county_geo_clean %>% 
-  mutate(density = drop_units(density)) %>%
-  filter(density > 2.70 & density < 3.10) %>%
-  select(NAME.y, density)
+selection3_me <- selection3_me %>%  
+  rownames_to_column(var = "Variable") %>% 
+  transmute(Variable,
+            b,
+            ll = b - 1.96 * se, 
+            ul = b + 1.96 * se)
 
-# Highly dense counties
-county_geo_clean %>% 
-  mutate(density = drop_units(density)) %>%
-  filter(density > 5.74) %>%
-  select(NAME.y)
+results_selection <- selection1_me %>%
+  full_join(selection2_me,
+            by = "Variable")
 
+colnames(selection3_me) <- c("Variable",
+                                 "b1",
+                                 "l1",
+                                 "u1")
+
+selection3_me <- selection3_me %>%
+  transmute(Variable,
+            b1 = ifelse(is.na(b1),
+                        " ", 
+                        round(b1, 3)),
+            ci1 = ifelse(is.na(l1), 
+                         " ",
+                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")))
+#%>%
+#  arrange(Order) %>%
+#  select(-Order)
+```
+
+```{r tabulate-sample-selection3-results, echo=FALSE}
+selection3_me  %>%
+  mutate(Variable = c("Intercept",
+                      "Density (1,000 per sq.km)",
+                      "Density squared",
+                      "Percent of private transportation", 
+                      "Median household income (10,000)",
+                      "Intercept",
+                      "Density (1,000 per sq.km)",
+                      "Density squared",
+                      "Percent of private transportation", 
+                      "Median household income (\\$10,000)",
+                      "Spatial filter",
+                      "$\\sigma$",
+                      "$\\rho$")) %>%
+  kable("latex",
+        digits = 3,
+        booktabs = TRUE,
+        escape = FALSE,
+        col.names = c("Variable",
+                      "$\\beta$", 
+                      "95\\% CI"),
+        align = c("lcccc"),
+        caption = "\\label{tab:selection3-results}Estimation results of sample selection model with spatial filter") %>%
+  kable_styling(font_size = 8, latex_options = "hold_position") %>% 
+  pack_rows(group_label = "Sample Selection Model", 1, 5) %>%
+  pack_rows(group_label = "Outcome Model", 6, 11) %>%
+  add_header_above(c(" ", "Selection Model 3" = 2))
 ```
@@
 
+# Discussion
+
+How relevant are the differences between the various model specifications presented above?
+Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 3, Selection Model 2, and Selection Model 3.
+The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1.
+The conclusion is that at higher densities, $R_0$ is _always_ higher.
+The two panels on the right, in contrast, shows that Selection Model 2 and Selection Model 3 coincide that $R_0$ tends to increase as density grows.
+This continues until a density of approximately 2.9 (1,000 people per sq.km).
+At higher densities than that the relationship between density and $R_0$ begins to weaken, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
+
+To put this into context, other things being equal, the effect of density in a county like Charlottesville in Virginia (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
+In contrast, the effect of density on $R_0$ in a county like Arlington in Virginia (density ~3,093 people per sq.km) is _stronger_ than either of the previous two examples.
+Lastly, the density of counties like San Francisco in California, or Queens and Bronx in NY, which are among the densest in the US, contributes even less to $R_0$ than even the most rural counties in the country.
+
+It is worth at this point to recall Cressie's dictum about modelling: "[w]hat is one person's mean structure could be another person's correlation structure" [@Cressie1989geostatistics, p.
+201].
+There are almost always multiple ways to approach a modelling situation.
+In the present case, we would argue that spatial sampling is an important aspect of the modeling process, but one that perhaps required different technical skills than those available to SWN.
+There is nothing wrong with that.
+What matters is that, by adopting relatively high reproducibility standards, these researchers made a valuable and honest contribution to the collective enterprise of seeking knowledge.
+Their effort, and subsequent efforts to validate and expand on their work, can potentially contribute to provide clarity to ongoing conversations about the relevance of density and the spread of COVID-19.
+
+In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [as put by Wong and Li, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading].
+At the same time, this also stands in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with it in a longitudinal setting [@Hamidi2020longitudinal.
+In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al.
+[-@Fielding2020social] might be onto something with respect to the difference between rural and urban counties.
+More generally, there is no doubt that in population-level studies density is indicative of proximity, but it also potentially is a proxy for adaptive behavior.
+And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.
+
+```{r comparison-results, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2."}
+# The maximum in-sample density is 6.255358
+# county_geo_clean %>% st_drop_geometry() %>% filter(R > 0) %>% summary()
+
+example_df <- data.frame(density = seq(0, 6.3, 0.01)) %>% 
+  mutate(density_log = log(density),
+         density_2 = density^2, 
+         b_swn = model3$coefficients$fixed["density_log"] * density_log,
+         b_selection = selection2$estimate[7] * density + selection2$estimate[8] * density_2,
+         b_selection3 = selection3$estimate[7] * density + selection3$estimate[8] * density_2)
+
+# Plot of coefficient of density to R_0 according to SWN Model 3
+density_swn <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_swn)) +
+  ggtitle("SWN Model 3") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+# Plot of coefficient of density to R_0  
+density_selection <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_selection)) +
+  ggtitle("Selection Model 2") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+# Plot of coefficient of density to R_0  
+density_selection3 <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_selection3)) +
+  ggtitle("Selection Model 3") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+density_swn + density_selection + density_selection3
+```
+
+```{r counties-by-density, include=FALSE}
+# the relationship between density and R_0 become negative in Selection Model 2 at about 5.74
+
+# Values of b_selection for a density similar a Philadelphia 
+example_df %>% filter(density > 4.10 & density < 4.14)
+# Densities with a similar b_selection
+example_df %>% filter(b_selection > 0.87 & b_selection < 0.89)
+
+# Counties with densities with similar response to Philadelphia
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 1.59 & density < 1.64) %>%
+  select(NAME.y, density)
+
+# Counties with densities associated with the highest response to density
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 2.70 & density < 3.10) %>%
+  select(NAME.y, density)
+
+# Highly dense counties
+county_geo_clean %>% 
+  mutate(density = drop_units(density)) %>%
+  filter(density > 5.74) %>%
+  select(NAME.y)
+
+```
+
 # Conclusion
 
-The tension between the need to publish research potentially useful in dealing with a global pandemic, and a "carnage of substandard research" [@Bramstedt2020carnage], highlights the importance of efforts to maintain the quality of scientific outputs during COVID-19.
+The tension between the need to publish research potentially useful in dealing with a global pandemic, and a potential "carnage of substandard research" [@Bramstedt2020carnage], highlights the importance of efforts to maintain the quality of scientific outputs during COVID-19.
 An important part of quality control is the ability of independent researchers to verify and examine the results of materials published in the literature.

### 2021-09-16 R0-Density-Reanalysis/Response-to-Reviewers-v2.Rmd
Revisions after decision by Geographical Analysis

---
title: "Response to Reviewers"
output:
  pdf_document: default
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: bibliography.bib
nocite: |
  @Gordon1989gasoline, @Newman1989gasoline
---

Dear Editor:

I would like to thank you and the two anonymous reviewers for the attention paid to my submission entitled "Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis" (GEAN-06-21-053).

The reviewers are generally positive about the contents of the paper, and there are no comments requiring additional analytical work. The most serious comment, from Reviewer 2, concerns the _tone_ and _presentation_ of the paper. You ask that I revise the paper to ensure that I am fair to a junior researcher (Reese Sy). This is a reasonable request. In this letter I respond to the thoughtful feedback I received and how I revised the paper in response.

I would like to assure the reviewers and you that my intention in using their work as an example was to constructively engage with Sy, White, and Nichols (SWN). In fact, even before submission to GA I had already taken steps to ensure that I was not acting like a complete jackass and a bully. I hope that my responses will persuade you of this.

Sincerely,

The author

<!--
Note to self: check this on how to create a document with the differences: https://tex.stackexchange.com/questions/165543/using-latexdiff-in-miktex
-->

\newpage

## Reviewer 1

The study aims to demonstrate the value of reproducible research, which involves making code and data publicly available. A strength of the manuscript is the expansion of relevant concerns with the SWN model, from the perspective of spatial analysis, and extensive and detailed justification for every model specification choice in the reanalysis. The inclusion of a RMarkdown file also guides the reader in understanding the models used in the reanalysis.

I have no major comments that can improve the study, as I believe the study is analytically sound and well-written. The only minor comment I have is the length of the background of the association of density and R0, which could be condensed further. Overall, the study is a good example of a manuscript that demonstrates the value of reproducible research, and would be an outstanding contribution to the scientific literature.

> \color{blue}{Many thanks for your comments and positive assessment of the paper. I lightly edited the background section for conciseness, while trying to keep the paper as self-contained as possible.}

\newpage

## Reviewer 2

The stated goal of this article is to promote reproducible research. This is done by reviewing another paper (SWN) that followed the standards advocated by this paper.

> \color{blue}{Thank you for this comment. In fact, I reviewed at least 22 papers for this research. Of these, SWN was almost alone in being reproducible. Furthermore, of those that were reproducible, SWN was the only one that I thought I could engage with in an interesting way. The papers that I reviewed, along with notes about their reproducibility, are in the table below.}

Paper | Reproducibility Notes
---   | ---
@Ahmad2020association | Code not shared, data are provided but data pre-processing is not documented
@Amadu2021assessing | Code not shared, data provenance is documented, data are not shared
@Bhadra2021impact | Code and data not available; only data provenance is documented
@Brandtner2021creatures | Code is available on demand, data are shared
@Cruz2020exploring | Code is shared, sources of data are documented, data are not actually provided
@Feng2020spread | Code not shared, data provenance is documented, data are not shared
@Feyman2020effectiveness | Code and data are provided
@Fielding2020social | Code not shared, data provenance is documented, data are not actually provided
@Hamidi2020longitudinal |  Code and data not available
@Hamidi2020density |  Code and data not available
@Inbaraj2021seroprevalence | Code not shared, data are not shared
@Lee2020human | Code not provided, data sources are documented but data are not shared (privacy & confidentiality)
@Noury2021how | Code not shared, data are shared
@Pequeno2020air | Raw data supplied in supplemental files, code not available
@Roy2020factors | Code not shared, data are provided and data processing is documented
@Skorka2020macroecology | Code and data are shared but data processing is not documented, 
@Souris2020covid | Code not provided, data sources are documented but data are not shared
@Stephens2021impact | Code and data are shared
@Sy2021population | Code and data are shared
@Wang2021transmission | Code not shared, data are available
@White2020state | Code and data are in repository
@Wong2020spreading | Code not provided, data are shared but data pre-processing is not documented

My summary of the paper: 1) It is partly a new "COVID-19 paper" because it provides a new way to understand the population density hypothesis. 2) It is partly a "response paper" because it provides in-depth criticism of a recent paper. 3) It is partly an "advocacy paper" because it is clearly promoting open science. I will address these in order.

> \color{blue}{Thank you for this insightful summary, which I think captures quite well the different missions of the paper.}

1) The new models shown look sound and are an improvement over other models in this field. The density hypothesis is often discussed, so it is important to continuously improve our understanding of it.

> \color{blue}{Thank you for your positive assessment of the technical aspects of the paper.}

2) Since this will not be published in PLOS, the original authors will not have the opportunity to respond. It does not feel fair. The authors of the GA paper might argue, “this is not a response paper.” SWN seems to be a strange choice for a larger example. The first author is a graduate student. No one should be immune from scientific criticism (the selection bias criticism is valid), but choosing a generally good graduate-led paper (if I correctly understand the GA authors) feels wrong. If the GA authors chose the papers led by well-established authors such as Paez or Wong, I would not feel so queasy.

> \color{blue}{Thank you for the thoughtful response to this aspect of my paper. I would like to begin my response by noting that PLoS ONE desk-rejected my paper, which deprived me of an opportunity to engage in scientific discourse with SWN (apparently PLoS ONE does not think that science should be discursive). Even worse, by forcing me to take my paper elsewhere, PLoS ONE deprived SWN of the opportunity to respond to my paper in their journal of choice. This is, in my view, a very weird aspect of PLoS ONE's editorial policy. Secondly, I did not choose SWN for the example to pick on them, but rather because they were, as I noted above, almost the only example of reproducible research that I found in my research. I chose SWN partly of necessity, and partly as a homage, not because I particularly wished to criticize them.}

> \color{blue}{More generally, I fully agree with you and the editor that criticism should be professional, in particular when directed at the work of junior researchers. Graduate students should be nurtured, not crushed. I tried to be respectful of Sy, White, and Nichols, which is why I carefully checked myself to avoid any and all arguments that could be construed as ad hominem. I tried to engage with the science, not with them personally. I also made an effort to recognize that I, just like everyone else, have strengths and limitations as researcher; I don't pretend to have the last work on density and the spread of COVID, and recognize that others may be better equipped to analyze these data more competently than me.}

> \color{blue}{Incidentally, your comment brought to mind memories of Gordon and Richardson's (1989) response to Newman and Kenworthy's (1989) paper on Gasoline Consumption and Cities. In their response, Gordon and Richardson concluded by suggesting that "[p]erhaps Newman and Kenworthy would be well advised to seek out another planet, preferably unpopulated, where they can build their compact cities from scratch with solar-powered transit." I read this many years ago as a graduate student, and to the date I still find it shockingly unprofessional. I must admit that I lack the viciousness and ferocity of Gordon and Richardson. Rather, my aim is to be a good scientist, a decent professional, but if at all possible an even better human being. For this reason I took steps, from the moment I completed the first draft of the paper, to ensure that I was operating within the parameters of my moral compass:}

> \color{blue}{1. I submitted the paper to PLoS ONE so that, if accepted, it would be a companion to the original SWN paper. As I noted above, PLoS ONE issued a desk rejection within two days: apparently their editorial policy precludes "commentary"; this leads me to believe that a response from SWN would also not have been acceptable in PLoS ONE. Only after being rejected there, did I try a different journal for the paper.}

> \color{blue}{2. Immediately after I pushed the first version of the paper to my repository, thereby making it public, I communicated with Reese Sy, Laura White, and Brooke Nichols. This was my e-mail to them on May 21, 2021:}.

> \color{blue}{\textit{Dear Ms. Sy, Dr. White, and Dr. Nichols, I would like to express my appreciation for your recent work on population density and the spread of COVID-19. I came across your paper while doing a search for a review assignment, and it caught my attention given some of my earlier work on COVID-19. Your paper is a relatively rare example of reproducible research, and it allowed me to relatively quickly replicate your results. I submitted a paper to PLoS ONE that builds on your analyses, and I thought that it would be just courteous to share it with you. Please see: *repository*. The README page needs some tidying up (for some reason a figure is not displayed). The source document is an R Markdown document which, along with the pdf you can find in this folder: R0-Density-Reanalysis. I also packaged the data for ease of distribution, and the workflow for data preparation is documented in this folder: raw-data-processing). If you have any questions or comments about this, I would be happy to hear from you.}}

>\color{blue}{As you can see from the salutation, I was well-aware that Ms. Sy is a graduate student. This was the response from Prof. White to my email:}

> \color{blue}{\textit{Dear Dr. X, Thanks for sharing this. I enjoyed reading your paper and find your results really interesting. I am happy to see you addressing the selection bias issues and considering nonlinearities in the association between density and R0. As you noted, these were issues that we chose not to address, but I am happy to see this picked up on. I agree with your premise the making our research reproducible only leads to better science. Hopefully we all get a bit better at this going forward. Best regards, Laura}}

> \color{blue}{This was followed by a response from Prof. Nichols:}

> \color{blue}{\textit{Dear X, I would like to echo Laura’s words. I really enjoyed reading your work- and this is the type of thing we should do more often in science! Cheers, Brooke}}

> \color{blue}{Last but not least, this was the response from Ms. Sy:}

> \color{blue}{\textit{Dear X, I agree with Laura and Brooke - thank you for sharing this with us. I completely agree with what you wrote about reproducible research as vital to examine the same data from a variety of different perspectives. This strengthens scientific discourse! Have a wonderful day. Best, Reese}}

> \color{blue}{Why did I ask SWN to comment on a first version of the paper? First, I strongly felt it was the courteous and decent thing to do. And secondly, I think that science is better and more fun when it is done in a spirit of cooperation instead of antagonism. Their responses do not give a sense that they felt the paper was unfair to any of them as a team, let alone to Ms. Sy as an individual. Quite the opposite. I was very happy (not to mention relieved!) when their responses were as supportive as they were.}

> \color{blue}{3. In addition, when submitting the paper to PLoS ONE (and after that to Science of the Total Environment, Landscape and Urban Planning, and Geographical Analysis), I always recommended that at least one of the authors of SWN be invited to review the paper.}

> \color{blue}{It seems clear to me that PLoS ONE is not interested in responses to research, only in research. But does PLoS ONE's editorial policy really mean that SWN cannot respond? If fairness to SWN is important, the Editor of GA could easily invite them to respond to my paper in GA! Since PLoS ONE does not seem to care about this conversation, why shouldn't we take it to continue it elsewhere?}

3) A well-documented paper containing data and code is easier to replicate and improve than a paper without these items. This seems obvious. The paper possibly indicates to journal editors that response papers are more likely to be submitted under open science. It possibly forms a cautionary tale that authors need to be extra diligent if they make their work easily replicable. However, using a case study method, as the GA paper does, seems to hurt the original paper/authors more than promote open science.

> \color{blue}{I agree with you that this seems obvious. The reality is that, as I noted above, the examples of replicable research are still few and far between, and in the whole literature of density and COVID, SWN was in the minority. I think my message is only partially directed to editors: for example, PLoS ONE already has a data and code policy - and yet, it is so seldom followed, let alone enforced, as to be next to useless. My message, instead, is mainly directed to researchers: we need to do much better than the very lowly expectations of journals. I already explained that I did not choose SWN to pick on them, but rather because they were one of the precious few papers that were reproducible. My communications with them do not give the impression that they were hurt by this paper, but rather felt it was an encouraging product of open science. And, if in doubt, GA could always invite them to respond.}

I want to think the GA paper comes from a place that respects the SWN paper, but the implementation of the GA paper does not give that impression. However, I do like the improved COVID-19 density model.

>\color{blue}{I am sorry that you were left with the impression that I was not respectful of SWN. I have no reason to think that they themselves felt disrespected. In my revisions I have been extra-careful to ensure that I engage with the material in a respectful and professional way.}

Other suggestions

Schweinsberg et al. (2021) supports the view that different researchers reach different conclusions (https://doi.org/10.1016/j.obhdp.2021.02.003).

>\color{blue}{Thank you for suggesting this paper. I have included it in the discussion and references.}

Pages 4,5. I think the authors have swapped the meanings of "large" and "small" scales.

>\color{blue}{Thank you for pointing this out. The text has been revised to avoid confusion.}

The authors did not mention the time period studied.

>\color{blue}{SWN obtained publicly available daily COVID-19 cases among United States counties from the New York Times and for each county assumed that the exponential growth period began one week prior to the second daily increase in cases. We assumed that the period of exponential growth lasted approximately 18 days, as previous research have shown the COVID-19 exponential period to be around 20–24 days in New York City.}

I assume that the data for the GA paper comes from the Johns Hopkins database. This data is sometimes revised, so this particular data set is an example of citations being insufficient to achieve exact replicability.

>\color{blue}{The data were obtained from the New York Times. Since the New York Times uses GitHub for their data repository, data versioning is available. See: https://github.com/nytimes/covid-19-data.}

Please define the basic reproductive number for non-epidemiologist readers earlier in the paper.

>\color{blue}{Done.}

The first part of the discussion section contains additional results. The discussion seems to start on page 18, line 41.

>\color{blue}{I reorganized the discussion following this comment.}

These opinions on the SWN paper and its authors seem unnecessary.
Page 7, line 30. "reasonable modeling choice"
Page 7, line 40. "sensible controls"
Page 8, lines 34,35. "whether a graduate student or a seasoned scientist"
Page 18, lines 46-47. "required different technical skills than those available to SWN"

>\color{blue}{I appreciate your attention to these statements, which have been revised to make them more a matter of factly and less opinionated. Some statements were deleted altogether since they were unnecessary.}

Page 8, line 19. It seems that the authors call themselves "expert"s. This seems unnecessary.

>\color{blue}{This was more of a general statement rather than an exlusive reference to me, and has been changed to "independent researcher" to avoid implying expertise.}

Page 9, lines 49-52. There are many reasons to delete this footnote.

>\color{blue}{It is a fact that the paper was desk rejected by the journals for the reasons stated. The fact that one of them was PLoS ONE helps to explain why publishing this paper elsewhere is not unfair to SWN. I did delete the editorial comment to the effect that "[t]his does not inspire much confidence in the commitment of journals to reproducibility in research."}

Citation typos: page 6, line 44; page 6, line 51; page 7, line 9; page 7, line 20; page 8, line 51

>\color{blue}{Errors corrected.}

Typos
Page 9, line 20. "RWN" should be "SWN"
Page 11, line 21. "fourth" should be "fifth"

>\color{blue}{Typos corrected.}

Mapping
Figure 1: A narrower border on counties would better highlight the fill color.
Figure 2: Rainbow colors are difficult to interpret. A divergent color band would be better.
Figure 3: A single color could display counties with data. This will link the map to the binary variable in the model.

>\color{blue}{Changes in the cartography include narrower borders for the boundaries and the use of different color palettes. I used palettes from the `scico` package. It provides perceptually uniform and colourblind safe palettes.}

>\color{blue}{I wish to thank you again for the care paid to reviewing this submission. I hope that you will find my responses to your concerns to be fair and sufficient.}

## References

### 2021-09-16 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Revisions after decision by Geographical Analysis

@@
 abstract: |
-  The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research.
+  The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic in 2019 led to explosive growth in scientific research.
 Alas, much of the research in the literature lacks conditions to be reproducible, and recent publications on the association between population density and the basic reproductive number of SARS-CoV-2 are no exception.
@@
 In this paper, an example of reproducible research shows the potential of spatial analysis for epidemiology research during COVID-19.
-Transparency and openness means that independent researchers can, with relatively modest efforts, verify findings and use different approaches as appropriate.
+Transparency and openness means that independent researchers can, with only modest efforts, verify findings and use different approaches as appropriate.
 Given the high stakes of the situation, it is essential that scientific findings, on which good policy depends, are as robust as possible; as the empirical example shows, reproducibility is one of the keys to ensure this.
@@
 ```{r load-packages, include=FALSE, cache=FALSE}
-library(adespatial) # To calculate Moran eigenvector maps
+library(adespatial) # Used to calculate Moran eigenvector maps
 library(censReg) # Package to estimate tobit model
@@
 library(nlme) # Used to estimate 
-library(patchwork)
+library(patchwork) # Package to organize ggplot2 plots
 library(r0density) # Data package
 library(sampleSelection) # Package to estimate sampling selection models
+library(scico) # Package with color palettes
 library(sf) # Package to work with spatial information in simple features format
@@
 library(units) # Package to work with units of measurement
-library(viridis) # Color palettes
+#library(viridis) # Color palettes
 ```
@@
 
-Given the ruinous human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has termed the deluge of COVID-19 publications a "carnage of substandard research" [@Bramstedt2020carnage].
+Given the ruinous human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research termed the deluge of COVID-19 publications a "carnage of substandard research" [@Bramstedt2020carnage].
 Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [e.g., @Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty and somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
@@
 
-The push for open data and software, along with more strenuous efforts towards open, reproducible research, is simply a continuation of long-standing scientific practices of independent verification.
+The push for open software and data [e.g., @Bivand2020progress; @Arribas2021open], along with more strenuous efforts towards open, reproducible research, is simply a continuation of long-standing scientific practices of independent verification.
 Despite the (at times disproportionate) attention that high profile scandals in science tend to elicit in the media, science as a collective endeavor is remarkable for being a self-correcting enterprise, one with built-in mechanisms and incentives to weed out erroneous ideas.
@@
 
-In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how]; the implications of the pandemic for urban planning, design, and management are the topic of ongoing research [@Sharifi2020covid].
+In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [e.g., @Florida2020how]; as a result, the implications of the pandemic for urban planning, design, and management are the topic of ongoing research [e.g., @Sharifi2020covid].
 The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent of research into the associations between density and the spread of the pandemic.
-he answers to some important questions hang on the results of these research efforts.
+The answers to some important questions hang on the results of these research efforts.
 For example, are lower density regions safer from the pandemic?
 Are de-densification policies warranted, even if just in the short term?
-And in the longer term, will the risks of life in high density regions presage a flight from cities?
-What are the implications of the pandemic for future urban planning and practice?
+In the longer term, will the risks of life in high density regions presage a flight from cities?
+And, what are the implications of the pandemic for future urban planning and practice?
 Over the past year, numerous papers have sought to throw light on the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed.
@@
 The objective of this paper is to illustrate the importance of reproducibility in research in the context of the flood of COVID-19 papers.
-To this end, a recent study by Sy et al.
-[-@Sy2021population] is chosen as an example of reproducible research.
-The objective is not to malign the analysis of these researchers, but rather to demonstrate the value of openness to allow for independent verification and further analysis.
-Open data and open code mean that an independent researcher can, with only modest efforts, not only verify the findings reported, but also examine the same data from a perspective which may not have been available to the original researchers due to differences in disciplinary perspectives, methodological traditions, and/or training, among other possible factors.
-The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in research.
-The present paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
+For this, I focus on a recent study by Sy et al.
+[-@Sy2021population] that examined the correlation between the basic reproductive number of COVID-19, $R_0$, and population density.
+The basic reproductive number is a summary measure of contact rates, probability of transmission of a pathogen, and duration of infectiousness.
+In rough terms, it measures how many new infections each infections begets.
+The paper of Sy et al.
+[-@Sy2021population] was selected for being, in the literature examined, almost alone in supporting reproducible research.
+Accordingly, I wish to be clear that my objective in singling their work for discussion is not to malign their efforts, but rather to demonstrate how open and reproducible research efforts can greatly help to accelerate discovery.
+More concretely, open data and open code mean that an independent researcher can, with only modest efforts, not only verify the findings reported, but also examine the same data from a perspective which may not have been available to the original researchers due to differences in disciplinary perspectives, methodological traditions, and/or training, among other possible factors.
+The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in their research.
+In this spirit, the present paper also aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
 The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, will be available in a public repository^[For peer-review purposes, the data package and code are currently in an anonymous Drive folder: \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}].
@@
 The concern with population density and the spread of the virus during the COVID-19 pandemic was fueled, at least in part, by dramatic scenes seen in real-time around the world from large urban centers such as Wuhan, Milan, Madrid, and New York.
-In theory, there are good reasons to believe that higher density may have a positive association with the transmission of a contagious virus.
+In theory, there are good reasons to believe that higher density could have a positive association with the transmission of a contagious virus.
 It has long been known that the potential for inter-personal contact is greater in regions with higher density [see for example the research on urban fields and time-geography, including @Farber2011running; @Moore1970urban; @Moore1970some].
@@
 
-As Rocklöv and Sjödin [@Rocklov2020high] note, however, mathematical models of contagion are valid at small-to-medium spatial scales (and presumably, small temporal scales too, such as time spent in restaurants, concert halls, cruises), and the results do not necessarily transfer to larger spatial units and different time scales.
+As Rocklöv and Sjödin [@Rocklov2020high] note, however, mathematical models of contagion are valid at small-to-medium spaces (and presumably, smaller time intervals too, such as time spent in restaurants, concert halls, cruises), and the results do not necessarily transfer to larger spatial units and longer time periods.
 There are solid reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers.
 On the other hand, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place.
-Nonetheless, the idea that high density correlates with high transmission is so seemingly sensible that it is often taken for granted even at larger scales [e.g., @Cruz2020exploring; @Micallef2020first].
-At larger scales, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, e.g., @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.
+Nonetheless, the idea that high density correlates with high transmission is so seemingly sensible that it is often taken for granted even at the scale of large spaces [e.g., @Cruz2020exploring; @Micallef2020first].
+In such conditions, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, e.g., @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.
 
@@
 [-@Paez2020spatio], who in their study of provinces in Spain reported non-significant associations between population density and infection rates in the early days of the first wave of COVID-19, and negative significant associations in the later part of the first lockdown.
-Similarly, [-@Skorka2020macroecology] found zero or negative associations between population density and infection numbers/deaths by country.
+Similarly, @Skorka2020macroecology found zero or negative associations between population density and infection numbers/deaths by country.
 Fielding-Miller et al.
@@
 For their part, in their investigation of doubling time, White and Hébert-Dufresne [-@White2020state] identified a negative and significant correlation between population density and doubling time in US states.
-Likewise, [-@Khavarian2021high] found a small negative (and significant) association between population density and COVID-19 morbidity in districts in Tehran.
-Finally, two of the most complete studies in the US [by Hamidi et al.
--@Hamidi2020longitudinal; and -@Hamidi2020density] used an extensive set of controls to find negative and significant correlations between density and COVID-19 cases and fatalities at the level of counties in the US.
+Likewise, @Khavarian2021high found a small negative (and significant) association between population density and COVID-19 morbidity in districts in Tehran.
+Finally, two of the most complete studies in the US, by @Hamidi2020longitudinal and @Hamidi2020density, used an extensive set of controls to find negative and significant correlations between density and COVID-19 cases and fatalities at the level of counties in the US.
 
@@
 To pick an example (which will be further elaborated in later sections of this paper), the study of Sy et al.
-[-@Sy2021population; hereafter SWN] would immediately grab the attention of a researcher with expertise in spatial analysis.
+[-@Sy2021population], hereafter referred to as SWN, would immediately grab the attention of a researcher with expertise in spatial analysis.
 
@@
 For their multivariate analysis, SWN used mixed linear models.
-This is a reasonable modelling choice: $R_0$ is an interval-ratio variable that is suitably modeled using linear regression; further, as SWN note there is a likelihood that the process in not independent "among counties within each state, potentially due to variable resource allocation and differing health systems across states"  (p.
+This is an appropriate modelling choice: $R_0$ is an interval-ratio variable that is suitably modeled using linear regression; further, as SWN note there is a likelihood that the process in not independent "among counties within each state, potentially due to variable resource allocation and differing health systems across states"  (p.
 3).
@@
 SWN estimated various models with different combinations of variables, including median household income and prevalence of travel by private transportation.
-These are sensible controls, given potential variations in behavior: people in more affluent counties may have greater opportunities to work from home, and use of private transportation reduces contact with strangers.
+These controls help to account for potential variations in behavior: people in more affluent counties may have greater opportunities to work from home, and use of private transportation reduces contact with strangers.
 Moreover, they also conducted various sensitivity analyses.
@@
 Fortunately, SWN are an example of a reasonably open, reproducible research product: their paper is accompanied by (most of) the data and (most of) the code used in the analysis.
-This means that an independent expert can, with only a moderate investment of time and effort, reproduce the results in the paper, as well as ask additional questions.
+This means that an independent researcher can, with only a moderate investment of time and effort, reproduce the results in the paper, as well as ask additional questions.
 
@@
 Typically the provenance of the data is documented, but in numerous studies the data themselves are not shared [@Amadu2021assessing; @Bhadra2021impact; @Cruz2020exploring; @Feng2020spread; @Fielding2020social; @Hamidi2020longitudinal; @Hamidi2020density; @Inbaraj2021seroprevalence; @Souris2020covid].
-As any researcher can attest, whether a graduate student or a seasoned scientist, collecting, organizing, and preparing data for a project can take a substantial amount of time.
+As any researcher can attest, collecting, organizing, and preparing data for a project can take a substantial amount of time.
 Pointing to the sources of data, even when these sources are public, is a small step towards reproducibility-but only a very small one.
@@
 In other cases, data are shared, but the processes followed in the preparation of the data are not fully documented [@Ahmad2020association; @Skorka2020macroecology].
-These processes matter, as shown by the errors in the spreadsheets of Reinhart and Rogoff [@Herndon2014high] and the data of biologist Jonathan Pruitt that led to an "avalanche" of paper retractions^[\url{https://doi.org/10.1038/d41586-020-00287-y}].
+These processes matter, as shown by the errors in the spreadsheets of Reinhart and Rogoff [see @Herndon2014high for the discovery of these errors], as well as by the data of biologist Jonathan Pruitt that led to an "avalanche" of paper retractions [see @Viglione2020avalanche].
 Another situation is when papers share well-documented data, but fail to provide the code used in the analysis [@Noury2021how; @Pequeno2020air; @Wang2021transmission].
@@
 Then there are those papers that more closely comply with reproducibility standards, and share well-documented processes and data, as well as the code used in any analyses reported [@Paez2020spatio; @Feyman2020effectiveness; @Stephens2021impact; @White2020state; @Sy2021population].
-Even in this case, the pressure to publish "new findings" instead of replication studies can act as a deterrent, perhaps particularly for younger researchers^[The present paper was desk rejected by three journals that had previously published research on population density and the spread of COVID-19; in one case, the paper was too opinionated for the journal, in the other two cases, the paper was not a "good fit" despite dealing with a nearly identical issue as papers previously published in said journals.
-This does not inspire much confidence in the commitment of journals to reproducibility in research.].
+Even in this case, the pressure to publish "new findings" instead of replication studies can act as a deterrent, perhaps particularly for younger researchers^[The present paper was desk rejected by three journals that had previously published research on population density and the spread of COVID-19; in one case, the paper was too opinionated for the journal, in the other two cases, the paper was not a "good fit" despite dealing with a nearly identical issue as papers previously published in said journals.].
 
@@
 
-In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of an independent researcher with expertise in spatial analysis are asked, and the data are reanalyzed.
+In the following sections, the analysis of SWN is reproduced, some relevant questions from the perspective of an independent researcher with expertise in spatial analysis are asked, and the data are reanalyzed.
 
@@
 Their final sample included `r prettyNum(nrow(county_geo %>% filter(!is.na(R))), big.mark = ",")` counties in the US, including in Alaska, Hawaii, Puerto Rico, and island territories.
+SWN used COVID-19 data collected by the New York Times and made available (with versioning) in a GitHub repository^[https://github.com/nytimes/covid-19-data].
+For each county, SWN assumed that the exponential growth period began one week prior to the second daily increase in cases, and assumed that the period of exponential growth lasted approximately 18 days.
 
@@
           color = NA) +
-  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
-  scale_fill_viridis_c(name = expression(R[0]),
-                       direction = -1) +
+  # Use viridis color palette: scico color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
+  scale_fill_scico(name = expression(R[0]),
+                       direction = 1,
+                   palette = "lajolla") +
   # Overlay counties with missing values of the reproductive number in white
@@
             filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
+          size = 0.05,
           fill = NA,
-          color = "lightgray") +
+          color = "gray60") +
   # Add boundary of continental US
@@
 It is possible to verify that the results match, with only the minor (and irrelevant) exception of the magnitude of the coefficient for travel by private transportation, which is due to a difference in the input (here the variable is changed to one percent units, instead of the ten percent units used by SWN).
-The mixed linear model gives random intercepts (i.e., the intercept is a random variable), and the standard deviation is reported in the fourth row of Table \ref{tab:swn-results}.
+The mixed linear model gives random intercepts (i.e., the intercept is a random variable), and the standard deviation is reported in the fifth row of Table \ref{tab:swn-results}.
 It is useful to map the random intercepts: as seen in Figure \ref{fig:random-terms-map}, other things being equal, counties in Texas tend to have somewhat lower values of $R_0$ (i.e., a negative random intercept), whereas counties in South Dakota tend to have higher values of $R_0$.
@@
           fill = NA) + 
-  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
-  scale_fill_viridis(name = "Random Intercepts",
-                     option = "turbo",
-                     direction = -1) +
+  # Use viridis color palette: scico color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
+  scale_fill_scico(name = "Random Intercepts",
+                   direction = -1,
+                   palette = "vikO") +
   # Set theme of plot
@@
           color = NA) +
-  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
-  scale_fill_viridis_c(name = "R0",
-                       direction = -1) +
+  # Use viridis color palette: scico color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
+  scale_fill_scico(name = "R0",
+                   palette = "lajolla") +
   # Overlay counties with missing values of the reproductive number in white
@@
           fill = NA,
-          color = "red") +
+          color = "blue",
+          size = 0.05) +
   # Add boundary of continental US
@@
   # Add caption
-  labs(caption = "Note: boundaris of urbanized areas with population > 50,000 are shown in red") +
+  labs(caption = "Note: boundaries of urbanized areas with population > 50,000 are shown in blue") +
   # Set theme of plot
@@
 
+Clearly, the various models display some intriguing differences; but how relevant are said differences from a more substantive standpoint?
+Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 3, Selection Model 2, and Selection Model 3.
+The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1.
+The conclusion is that at higher densities, $R_0$ is _always_ higher.
+The two panels on the right, in contrast, shows that Selection Model 2 and Selection Model 3 coincide that $R_0$ tends to increase as density grows.
+This continues until a density of approximately 2.9 (1,000 people per sq.km).
+At higher densities than that the relationship between density and $R_0$ begins to weaken, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
+
+To put this into context, other things being equal, the effect of density in a county like Charlottesville in Virginia (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
+In contrast, the effect of density on $R_0$ in a county like Arlington in Virginia (density ~3,093 people per sq.km) is _stronger_ than either of the previous two examples.
+Lastly, the density of counties like San Francisco in California, or Queens and Bronx in NY, which are among the densest in the US, contributes even less to $R_0$ than even the most rural counties in the country.
+
+```{r comparison-results, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2."}
+# The maximum in-sample density is 6.255358
+# county_geo_clean %>% st_drop_geometry() %>% filter(R > 0) %>% summary()
+
+example_df <- data.frame(density = seq(0, 6.3, 0.01)) %>% 
+  mutate(density_log = log(density),
+         density_2 = density^2, 
+         b_swn = model3$coefficients$fixed["density_log"] * density_log,
+         b_selection = selection2$estimate[7] * density + selection2$estimate[8] * density_2,
+         b_selection3 = selection3$estimate[7] * density + selection3$estimate[8] * density_2)
+
+# Plot of coefficient of density to R_0 according to SWN Model 3
+density_swn <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_swn)) +
+  ggtitle("SWN Model 3") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+# Plot of coefficient of density to R_0  
+density_selection <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_selection)) +
+  ggtitle("Selection Model 2") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+# Plot of coefficient of density to R_0  
+density_selection3 <- ggplot(data = example_df) + 
+  geom_line(aes(x = density, 
+                y = b_selection3)) +
+  ggtitle("Selection Model 3") +
+  ylab(expression(paste("E[", R[0], "]"))) +
+  theme_minimal()
+
+density_swn + density_selection + density_selection3
+```
+
 <!--
@@
 
-How relevant are the differences between the various model specifications presented above?
-Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 3, Selection Model 2, and Selection Model 3.
-The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1.
-The conclusion is that at higher densities, $R_0$ is _always_ higher.
-The two panels on the right, in contrast, shows that Selection Model 2 and Selection Model 3 coincide that $R_0$ tends to increase as density grows.
-This continues until a density of approximately 2.9 (1,000 people per sq.km).
-At higher densities than that the relationship between density and $R_0$ begins to weaken, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
-
-To put this into context, other things being equal, the effect of density in a county like Charlottesville in Virginia (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
-In contrast, the effect of density on $R_0$ in a county like Arlington in Virginia (density ~3,093 people per sq.km) is _stronger_ than either of the previous two examples.
-Lastly, the density of counties like San Francisco in California, or Queens and Bronx in NY, which are among the densest in the US, contributes even less to $R_0$ than even the most rural counties in the country.
-
 It is worth at this point to recall Cressie's dictum about modelling: "[w]hat is one person's mean structure could be another person's correlation structure" [@Cressie1989geostatistics, p.
 201].
-There are almost always multiple ways to approach a modelling situation.
-In the present case, we would argue that spatial sampling is an important aspect of the modeling process, but one that perhaps required different technical skills than those available to SWN.
-There is nothing wrong with that.
-What matters is that, by adopting relatively high reproducibility standards, these researchers made a valuable and honest contribution to the collective enterprise of seeking knowledge.
+There are almost always multiple ways to approach a modelling situation, as lively illustrated by a recent paper that reports the results of a crowdsourced modelling experiment [@Schweinsberg2021same].
+In the present case, we would argue that spatial sampling is an important aspect of the modeling process.
+Importantly, by adopting high reproducibility standards, SWN made a valuable contribution to the collective enterprise of seeking knowledge.
 Their effort, and subsequent efforts to validate and expand on their work, can potentially contribute to provide clarity to ongoing conversations about the relevance of density and the spread of COVID-19.
 
-In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [as put by Wong and Li, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading].
-At the same time, this also stands in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with it in a longitudinal setting [@Hamidi2020longitudinal.
+In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [in Wong and Li's words, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading].
+At the same time, the results presented here also stand in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with it in a longitudinal setting [@Hamidi2020longitudinal.
 In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al.
@@
 And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.
-
-```{r comparison-results, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2."}
-# The maximum in-sample density is 6.255358
-# county_geo_clean %>% st_drop_geometry() %>% filter(R > 0) %>% summary()
-
-example_df <- data.frame(density = seq(0, 6.3, 0.01)) %>% 
-  mutate(density_log = log(density),
-         density_2 = density^2, 
-         b_swn = model3$coefficients$fixed["density_log"] * density_log,
-         b_selection = selection2$estimate[7] * density + selection2$estimate[8] * density_2,
-         b_selection3 = selection3$estimate[7] * density + selection3$estimate[8] * density_2)
-
-# Plot of coefficient of density to R_0 according to SWN Model 3
-density_swn <- ggplot(data = example_df) + 
-  geom_line(aes(x = density, 
-                y = b_swn)) +
-  ggtitle("SWN Model 3") +
-  ylab(expression(paste("E[", R[0], "]"))) +
-  theme_minimal()
-
-# Plot of coefficient of density to R_0  
-density_selection <- ggplot(data = example_df) + 
-  geom_line(aes(x = density, 
-                y = b_selection)) +
-  ggtitle("Selection Model 2") +
-  ylab(expression(paste("E[", R[0], "]"))) +
-  theme_minimal()
-
-# Plot of coefficient of density to R_0  
-density_selection3 <- ggplot(data = example_df) + 
-  geom_line(aes(x = density, 
-                y = b_selection3)) +
-  ggtitle("Selection Model 3") +
-  ylab(expression(paste("E[", R[0], "]"))) +
-  theme_minimal()
-
-density_swn + density_selection + density_selection3
-```
 
@@
 Instead, the possibility of a non-monotonical relationship between population density and contagion is raised.
-
-In the spirit of openness, this paper is prepared as an R Markdown document, an a companion data package is provided.
-The data package contains the relevant documentation of the data, and all data pre-processing is fully documented.
-Hopefully this, and similar reproducible papers, will continue to encourage others to adopt reproducible standards in their research.
+I do not claim that the analysis presented here is the last word on the topic of density and the spread of COVID-19, and there is always the possibility that someone else will be better equipped to analyze these data with greater competence.
+By opening up the analysis, documenting the way data were pre-processed, and by sharing analysis ready data, my hope would be that others will be able to discover the limitations of my own analysis and improve on it, as appropriate.
+
+More generally, my hope is that the research of Sy et al.
+[-@Sy2021population], the present paper, and similar reproducible publications, will continue to encourage others to adopt higher reproducibility standards in their research.
 

### 2021-09-16 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Revisions after decision by Geographical Analysis

@@
 % Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
 \newlength{\csllabelwidth}
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
 % for Pandoc 2.8 to 2.10.1
@@
 % For Pandoc 2.11+
-\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
  {% don't indent paragraphs
@@
  {}
-\usepackage{calc} % for calculating minipage widths
+\usepackage{calc}
 \newcommand{\CSLBlock}[1]{#1\hfill\break}
 \newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
   The emergence of the novel SARS-CoV-2 coronavirus and the global
-  COVID-19 pandemic has led to explosive growth in scientific research.
-  Alas, much of the research in the literature lacks conditions to be
-  reproducible, and recent publications on the association between
-  population density and the basic reproductive number of SARS-CoV-2 are
-  no exception. Relatively few papers share code and data sufficiently,
-  which hinders not only verification but additional experimentation. In
-  this paper, an example of reproducible research shows the potential of
-  spatial analysis for epidemiology research during COVID-19.
-  Transparency and openness means that independent researchers can, with
-  relatively modest efforts, verify findings and use different
-  approaches as appropriate. Given the high stakes of the situation, it
-  is essential that scientific findings, on which good policy depends,
-  are as robust as possible; as the empirical example shows,
-  reproducibility is one of the keys to ensure this.
+  COVID-19 pandemic in 2019 led to explosive growth in scientific
+  research. Alas, much of the research in the literature lacks
+  conditions to be reproducible, and recent publications on the
+  association between population density and the basic reproductive
+  number of SARS-CoV-2 are no exception. Relatively few papers share
+  code and data sufficiently, which hinders not only verification but
+  additional experimentation. In this paper, an example of reproducible
+  research shows the potential of spatial analysis for epidemiology
+  research during COVID-19. Transparency and openness means that
+  independent researchers can, with only modest efforts, verify findings
+  and use different approaches as appropriate. Given the high stakes of
+  the situation, it is essential that scientific findings, on which good
+  policy depends, are as robust as possible; as the empirical example
+  shows, reproducibility is one of the keys to ensure this.
   \end{abstract}
@@
 consistently high quality standards in scientific reporting; indeed, a
-call for maintaining the standards in published research has termed the
+call for maintaining the standards in published research termed the
 deluge of COVID-19 publications a ``carnage of substandard research''
@@
 
-The push for open data and software, along with more strenuous efforts
-towards open, reproducible research, is simply a continuation of
-long-standing scientific practices of independent verification. Despite
-the (at times disproportionate) attention that high profile scandals in
-science tend to elicit in the media, science as a collective endeavor is
-remarkable for being a self-correcting enterprise, one with built-in
-mechanisms and incentives to weed out erroneous ideas. Over the long
-term, facts tend to prevail in science. At stake is the shorter-term
-impacts that research may have in other spheres of economic and social
-life. The case of economists Reinhart and Rogoff comes to mind: by the
-time the inaccuracies and errors in their research were uncovered (see
-Herndon et al., 2014), their claims about debt and economic growth had
-already been seized by policy-makers on both sides of the Atlantic to
-justify austerity policies in the aftermath of the Great Recession of
+The push for open software and data (Arribas-Bel et al., 2021; e.g.,
+Bivand, 2020), along with more strenuous efforts towards open,
+reproducible research, is simply a continuation of long-standing
+scientific practices of independent verification. Despite the (at times
+disproportionate) attention that high profile scandals in science tend
+to elicit in the media, science as a collective endeavor is remarkable
+for being a self-correcting enterprise, one with built-in mechanisms and
+incentives to weed out erroneous ideas. Over the long term, facts tend
+to prevail in science. At stake is the shorter-term impacts that
+research may have in other spheres of economic and social life. The case
+of economists Reinhart and Rogoff comes to mind: by the time the
+inaccuracies and errors in their research were uncovered (see Herndon et
+al., 2014), their claims about debt and economic growth had already been
+seized by policy-makers on both sides of the Atlantic to justify
+austerity policies in the aftermath of the Great Recession of
 2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
@@
 numerous thinkers has been the prospect of life in cities after the
-pandemic (Florida et al., 2020); the implications of the pandemic for
-urban planning, design, and management are the topic of ongoing research
-(Sharifi and Khavarian-Garmsir, 2020). The fact that the worst of the
-pandemic was initially felt in dense population centers such as Wuhan,
-Milan, Madrid, and New York, unleashed a torrent of research into the
-associations between density and the spread of the pandemic. he answers
-to some important questions hang on the results of these research
-efforts. For example, are lower density regions safer from the pandemic?
-Are de-densification policies warranted, even if just in the short term?
-And in the longer term, will the risks of life in high density regions
-presage a flight from cities? What are the implications of the pandemic
-for future urban planning and practice? Over the past year, numerous
-papers have sought to throw light on the underlying issue of density and
-the pandemic; nonetheless the results, as will be detailed next, remain
-mixed. Further, to complicate matters, precious few of these studies
-appear to be sufficiently open to support independent verification.
+pandemic (e.g., Florida et al., 2020); as a result, the implications of
+the pandemic for urban planning, design, and management are the topic of
+ongoing research (e.g., Sharifi and Khavarian-Garmsir, 2020). The fact
+that the worst of the pandemic was initially felt in dense population
+centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent
+of research into the associations between density and the spread of the
+pandemic. The answers to some important questions hang on the results of
+these research efforts. For example, are lower density regions safer
+from the pandemic? Are de-densification policies warranted, even if just
+in the short term? In the longer term, will the risks of life in high
+density regions presage a flight from cities? And, what are the
+implications of the pandemic for future urban planning and practice?
+Over the past year, numerous papers have sought to throw light on the
+underlying issue of density and the pandemic; nonetheless the results,
+as will be detailed next, remain mixed. Further, to complicate matters,
+precious few of these studies appear to be sufficiently open to support
+independent verification.
 
@@
 reproducibility in research in the context of the flood of COVID-19
-papers. To this end, a recent study by Sy et al. (2021) is chosen as an
-example of reproducible research. The objective is not to malign the
-analysis of these researchers, but rather to demonstrate the value of
-openness to allow for independent verification and further analysis.
-Open data and open code mean that an independent researcher can, with
-only modest efforts, not only verify the findings reported, but also
-examine the same data from a perspective which may not have been
-available to the original researchers due to differences in disciplinary
-perspectives, methodological traditions, and/or training, among other
-possible factors. The example, which shows consequential changes in the
-conclusions reached by different analyses, should serve as a call to
-researchers to redouble their efforts to increase transparency and
-reproducibility in research. The present paper, in addition, aims to
-show how data can be packaged in well-documented, shareable units, and
-code can be embedded into self-contained documents suitable for review
-and independent verification. The source for this paper is an
+papers. For this, I focus on a recent study by Sy et al. (2021) that
+examined the correlation between the basic reproductive number of
+COVID-19, \(R_0\), and population density. The basic reproductive number
+is a summary measure of contact rates, probability of transmission of a
+pathogen, and duration of infectiousness. In rough terms, it measures
+how many new infections each infections begets. The paper of Sy et al.
+(2021) was selected for being, in the literature examined, almost alone
+in supporting reproducible research. Accordingly, I wish to be clear
+that my objective in singling their work for discussion is not to malign
+their efforts, but rather to demonstrate how open and reproducible
+research efforts can greatly help to accelerate discovery. More
+concretely, open data and open code mean that an independent researcher
+can, with only modest efforts, not only verify the findings reported,
+but also examine the same data from a perspective which may not have
+been available to the original researchers due to differences in
+disciplinary perspectives, methodological traditions, and/or training,
+among other possible factors. The example, which shows consequential
+changes in the conclusions reached by different analyses, should serve
+as a call to researchers to redouble their efforts to increase
+transparency and reproducibility in their research. In this spirit, the
+present paper also aims to show how data can be packaged in
+well-documented, shareable units, and code can be embedded into
+self-contained documents suitable for review and independent
+verification. The source for this paper is an
 \href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
@@
 Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
-believe that higher density may have a positive association with the
+believe that higher density could have a positive association with the
 transmission of a contagious virus. It has long been known that the
@@
 As Rocklöv and Sjödin (Rocklöv and Sjödin, 2020) note, however,
-mathematical models of contagion are valid at small-to-medium spatial
-scales (and presumably, small temporal scales too, such as time spent in
+mathematical models of contagion are valid at small-to-medium spaces
+(and presumably, smaller time intervals too, such as time spent in
 restaurants, concert halls, cruises), and the results do not necessarily
-transfer to larger spatial units and different time scales. There are
+transfer to larger spatial units and longer time periods. There are
 solid reasons for this: while in a restaurant, one can hardly avoid
@@
 with high transmission is so seemingly sensible that it is often taken
-for granted even at larger scales (e.g., Cruz et al., 2020; Micallef et
-al., 2020). At larger scales, however, there exists the possibility of
-behavioral adaptations, which are difficult to capture in the
-mechanistic framework of differential equations (or can be missing in
-agent-based models, e.g., Gomez et al., 2021); these adaptations, in
+for granted even at the scale of large spaces (e.g., Cruz et al., 2020;
+Micallef et al., 2020). In such conditions, however, there exists the
+possibility of behavioral adaptations, which are difficult to capture in
+the mechanistic framework of differential equations (or can be missing
+in agent-based models, e.g., Gomez et al., 2021); these adaptations, in
 fact, can be a key aspect of disease transmission.
@@
 wave of COVID-19, and negative significant associations in the later
-part of the first lockdown. Similarly, (2020) found zero or negative
-associations between population density and infection numbers/deaths by
-country. Fielding-Miller et al. (2020) contrast their finding about
-rural counties with a negative relationship between COVID-19 deaths and
-population density in urban counties in the US. For their part, in their
-investigation of doubling time, White and Hébert-Dufresne (2020)
-identified a negative and significant correlation between population
-density and doubling time in US states. Likewise, (2021) found a small
-negative (and significant) association between population density and
-COVID-19 morbidity in districts in Tehran. Finally, two of the most
-complete studies in the US (by Hamidi et al. 2020a, and 2020b) used an
+part of the first lockdown. Similarly, Skórka et al. (2020) found zero
+or negative associations between population density and infection
+numbers/deaths by country. Fielding-Miller et al. (2020) contrast their
+finding about rural counties with a negative relationship between
+COVID-19 deaths and population density in urban counties in the US. For
+their part, in their investigation of doubling time, White and
+Hébert-Dufresne (2020) identified a negative and significant correlation
+between population density and doubling time in US states. Likewise,
+Khavarian-Garmsir et al. (2021) found a small negative (and significant)
+association between population density and COVID-19 morbidity in
+districts in Tehran. Finally, two of the most complete studies in the
+US, by Hamidi et al. (2020a) and Hamidi et al. (2020b), used an
 extensive set of controls to find negative and significant correlations
@@
 be further elaborated in later sections of this paper), the study of Sy
-et al.~{[}(2021); hereafter SWN{]} would immediately grab the attention
-of a researcher with expertise in spatial analysis.
+et al. (2021), hereafter referred to as SWN, would immediately grab the
+attention of a researcher with expertise in spatial analysis.
 
@@
 income, and prevalence of private mobility. For their multivariate
-analysis, SWN used mixed linear models. This is a reasonable modelling
+analysis, SWN used mixed linear models. This is an appropriate modelling
 choice: \(R_0\) is an interval-ratio variable that is suitably modeled
@@
 different combinations of variables, including median household income
-and prevalence of travel by private transportation. These are sensible
-controls, given potential variations in behavior: people in more
-affluent counties may have greater opportunities to work from home, and
-use of private transportation reduces contact with strangers. Moreover,
-they also conducted various sensitivity analyses. After these efforts,
-SWN concluded that there is a positive association between the basic
+and prevalence of travel by private transportation. These controls help
+to account for potential variations in behavior: people in more affluent
+counties may have greater opportunities to work from home, and use of
+private transportation reduces contact with strangers. Moreover, they
+also conducted various sensitivity analyses. After these efforts, SWN
+concluded that there is a positive association between the basic
 reproductive number and population density at the level of counties in
@@
 the data and (most of) the code used in the analysis. This means that an
-independent expert can, with only a moderate investment of time and
+independent researcher can, with only a moderate investment of time and
 effort, reproduce the results in the paper, as well as ask additional
@@
 al., 2021; Souris and Gonzalez, 2020). As any researcher can attest,
-whether a graduate student or a seasoned scientist, collecting,
-organizing, and preparing data for a project can take a substantial
-amount of time. Pointing to the sources of data, even when these sources
-are public, is a small step towards reproducibility-but only a very
-small one. Faced with the prospect of having to recreate a data set from
-raw sources is probably sufficient to dissuade all but the most
-dedicated (or stubborn) researcher from independent verification. This
-is true even if part of the data are shared (e.g., Wong and Li, 2020).
-In other cases, data are shared, but the processes followed in the
-preparation of the data are not fully documented (Ahmad et al., 2020;
-Skórka et al., 2020). These processes matter, as shown by the errors in
-the spreadsheets of Reinhart and Rogoff (Herndon et al., 2014) and the
-data of biologist Jonathan Pruitt that led to an ``avalanche'' of paper
-retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
-Another situation is when papers share well-documented data, but fail to
-provide the code used in the analysis (Noury et al., 2021; Pequeno et
-al., 2020; Wang et al., 2021). Making code available only ``on demand''
-(e.g., Brandtner et al., 2021) is an unnecessary barrier when most
-journals offer the facility to share supplemental materials online. Then
-there are those papers that more closely comply with reproducibility
-standards, and share well-documented processes and data, as well as the
-code used in any analyses reported (Feyman et al., 2020; Paez et al.,
-2020; Stephens et al., 2021; Sy et al., 2021; White and Hébert-Dufresne,
-2020). Even in this case, the pressure to publish ``new findings''
-instead of replication studies can act as a deterrent, perhaps
-particularly for younger researchers\footnote{The present paper was desk
-  rejected by three journals that had previously published research on
-  population density and the spread of COVID-19; in one case, the paper
-  was too opinionated for the journal, in the other two cases, the paper
-  was not a ``good fit'' despite dealing with a nearly identical issue
-  as papers previously published in said journals. This does not inspire
-  much confidence in the commitment of journals to reproducibility in
-  research.}.
-
-In the following sections, the analysis of RWN is reproduced, some
+collecting, organizing, and preparing data for a project can take a
+substantial amount of time. Pointing to the sources of data, even when
+these sources are public, is a small step towards reproducibility-but
+only a very small one. Faced with the prospect of having to recreate a
+data set from raw sources is probably sufficient to dissuade all but the
+most dedicated (or stubborn) researcher from independent verification.
+This is true even if part of the data are shared (e.g., Wong and Li,
+2020). In other cases, data are shared, but the processes followed in
+the preparation of the data are not fully documented (Ahmad et al.,
+2020; Skórka et al., 2020). These processes matter, as shown by the
+errors in the spreadsheets of Reinhart and Rogoff (see Herndon et al.,
+2014 for the discovery of these errors), as well as by the data of
+biologist Jonathan Pruitt that led to an ``avalanche'' of paper
+retractions (see Viglione, 2020). Another situation is when papers share
+well-documented data, but fail to provide the code used in the analysis
+(Noury et al., 2021; Pequeno et al., 2020; Wang et al., 2021). Making
+code available only ``on demand'' (e.g., Brandtner et al., 2021) is an
+unnecessary barrier when most journals offer the facility to share
+supplemental materials online. Then there are those papers that more
+closely comply with reproducibility standards, and share well-documented
+processes and data, as well as the code used in any analyses reported
+(Feyman et al., 2020; Paez et al., 2020; Stephens et al., 2021; Sy et
+al., 2021; White and Hébert-Dufresne, 2020). Even in this case, the
+pressure to publish ``new findings'' instead of replication studies can
+act as a deterrent, perhaps particularly for younger
+researchers\footnote{The present paper was desk rejected by three
+  journals that had previously published research on population density
+  and the spread of COVID-19; in one case, the paper was too opinionated
+  for the journal, in the other two cases, the paper was not a ``good
+  fit'' despite dealing with a nearly identical issue as papers
+  previously published in said journals.}.
+
+In the following sections, the analysis of SWN is reproduced, some
 relevant questions from the perspective of an independent researcher
@@
 included 1,151 counties in the US, including in Alaska, Hawaii, Puerto
-Rico, and island territories.
+Rico, and island territories. SWN used COVID-19 data collected by the
+New York Times and made available (with versioning) in a GitHub
+repository\footnote{https://github.com/nytimes/covid-19-data}. For each
+county, SWN assumed that the exponential growth period began one week
+prior to the second daily increase in cases, and assumed that the period
+of exponential growth lasted approximately 18 days.
 
@@
 intercepts (i.e., the intercept is a random variable), and the standard
-deviation is reported in the fourth row of Table \ref{tab:swn-results}.
+deviation is reported in the fifth row of Table \ref{tab:swn-results}.
 It is useful to map the random intercepts: as seen in Figure
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-How relevant are the differences between the various model
-specifications presented above? Figure \ref{fig:comparison-results}
-shows the relationship between density and \(R_0\) implied by SWN Model
-3, Selection Model 2, and Selection Model 3. The left panel of the
-figure shows the non-linear but monotonic relationship implied by SWN
-Model 1. The conclusion is that at higher densities, \(R_0\) is
-\emph{always} higher. The two panels on the right, in contrast, shows
-that Selection Model 2 and Selection Model 3 coincide that \(R_0\) tends
-to increase as density grows. This continues until a density of
-approximately 2.9 (1,000 people per sq.km). At higher densities than
-that the relationship between density and \(R_0\) begins to weaken, and
-the relationship becomes negative at densities higher than approximately
-5.7 (1,000 people per sq.km).
+Clearly, the various models display some intriguing differences; but how
+relevant are said differences from a more substantive standpoint? Figure
+\ref{fig:comparison-results} shows the relationship between density and
+\(R_0\) implied by SWN Model 3, Selection Model 2, and Selection Model
+3. The left panel of the figure shows the non-linear but monotonic
+relationship implied by SWN Model 1. The conclusion is that at higher
+densities, \(R_0\) is \emph{always} higher. The two panels on the right,
+in contrast, shows that Selection Model 2 and Selection Model 3 coincide
+that \(R_0\) tends to increase as density grows. This continues until a
+density of approximately 2.9 (1,000 people per sq.km). At higher
+densities than that the relationship between density and \(R_0\) begins
+to weaken, and the relationship becomes negative at densities higher
+than approximately 5.7 (1,000 people per sq.km).
 
@@
 
+\begin{figure}
+\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/comparison-results-1} \caption{\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2.}\label{fig:comparison-results}
+\end{figure}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
 It is worth at this point to recall Cressie's dictum about modelling:
@@
 correlation structure'' (Cressie, 1989, p. 201). There are almost always
-multiple ways to approach a modelling situation. In the present case, we
-would argue that spatial sampling is an important aspect of the modeling
-process, but one that perhaps required different technical skills than
-those available to SWN. There is nothing wrong with that. What matters
-is that, by adopting relatively high reproducibility standards, these
-researchers made a valuable and honest contribution to the collective
-enterprise of seeking knowledge. Their effort, and subsequent efforts to
-validate and expand on their work, can potentially contribute to provide
-clarity to ongoing conversations about the relevance of density and the
-spread of COVID-19.
+multiple ways to approach a modelling situation, as lively illustrated
+by a recent paper that reports the results of a crowdsourced modelling
+experiment (Schweinsberg et al., 2021). In the present case, we would
+argue that spatial sampling is an important aspect of the modeling
+process. Importantly, by adopting high reproducibility standards, SWN
+made a valuable contribution to the collective enterprise of seeking
+knowledge. Their effort, and subsequent efforts to validate and expand
+on their work, can potentially contribute to provide clarity to ongoing
+conversations about the relevance of density and the spread of COVID-19.
 
@@
 that higher density is \emph{always} associated with a greater risk of
-spread of the virus {[}as put by Wong and Li, ``\,`Density is destiny'
-is probably an overstatement''; (2020){]}. At the same time, this also
-stands in contrast to the findings of Hamidi et al., who found that
-higher density was either not significantly associated with the rate of
-the virus in a cross-sectional study (Hamidi et al., 2020b), or was
-negatively associated with it in a longitudinal setting {[}Hamidi et al.
-(2020a). In this sense, the conclusion that density does not aggravate
-the pandemic may have been somewhat premature; instead, reanalysis of
-the data of SWN suggests that Fielding-Miller et al. (2020) might be
-onto something with respect to the difference between rural and urban
-counties. More generally, there is no doubt that in population-level
-studies density is indicative of proximity, but it also potentially is a
-proxy for adaptive behavior. And it is possible that the determining
-factor during COVID-19, at least in the US, has been variations in
-perceptions of the risks associated with contagion (Chauhan et al.,
-2021), and subsequent compensations in behavior in more and less dense
-regions.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/comparison-results-1} \caption{\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2.}\label{fig:comparison-results}
-\end{figure}
+spread of the virus {[}in Wong and Li's words, ``\,`Density is destiny'
+is probably an overstatement''; (2020){]}. At the same time, the results
+presented here also stand in contrast to the findings of Hamidi et al.,
+who found that higher density was either not significantly associated
+with the rate of the virus in a cross-sectional study (Hamidi et al.,
+2020b), or was negatively associated with it in a longitudinal setting
+{[}Hamidi et al. (2020a). In this sense, the conclusion that density
+does not aggravate the pandemic may have been somewhat premature;
+instead, reanalysis of the data of SWN suggests that Fielding-Miller et
+al. (2020) might be onto something with respect to the difference
+between rural and urban counties. More generally, there is no doubt that
+in population-level studies density is indicative of proximity, but it
+also potentially is a proxy for adaptive behavior. And it is possible
+that the determining factor during COVID-19, at least in the US, has
+been variations in perceptions of the risks associated with contagion
+(Chauhan et al., 2021), and subsequent compensations in behavior in more
+and less dense regions.
 
@@
 a non-monotonical relationship between population density and contagion
-is raised.
-
-In the spirit of openness, this paper is prepared as an R Markdown
-document, an a companion data package is provided. The data package
-contains the relevant documentation of the data, and all data
-pre-processing is fully documented. Hopefully this, and similar
-reproducible papers, will continue to encourage others to adopt
-reproducible standards in their research.
+is raised. I do not claim that the analysis presented here is the last
+word on the topic of density and the spread of COVID-19, and there is
+always the possibility that someone else will be better equipped to
+analyze these data with greater competence. By opening up the analysis,
+documenting the way data were pre-processed, and by sharing analysis
+ready data, my hope would be that others will be able to discover the
+limitations of my own analysis and improve on it, as appropriate.
+
+More generally, my hope is that the research of Sy et al. (2021), the
+present paper, and similar reproducible publications, will continue to
+encourage others to adopt higher reproducibility standards in their
+research.
 
@@
 
+\leavevmode\hypertarget{ref-Arribas2021open}{}%
+Arribas-Bel, D., Green, M., Rowe, F., Singleton, A., 2021. Open data
+products: A framework for creating valuable analysis-ready data. Journal
+of Geographical Systems.
+
 \leavevmode\hypertarget{ref-Basu2017ten}{}%
@@
 doi:\href{https://doi.org/10.1007/s40808-020-00984-7}{10.1007/s40808-020-00984-7}
+
+\leavevmode\hypertarget{ref-Bivand2020progress}{}%
+Bivand, R.S., 2020. Progress in the r ecosystem for representing and
+handling spatial data. Journal of Geographical Systems.
+doi:\href{https://doi.org/10.1007/s10109-020-00336-0}{10.1007/s10109-020-00336-0}
 
@@
 
+\leavevmode\hypertarget{ref-Schweinsberg2021same}{}%
+Schweinsberg, M., Feldman, M., Staub, N., van den Akker, O.R., van Aert,
+R.C.M., van Assen, M.A.L.M., Liu, Y., Althoff, T., Heer, J., Kale, A.,
+Mohamed, Z., Amireh, H., Venkatesh Prasad, V., Bernstein, A., Robinson,
+E., Snellman, K., Amy Sommer, S., Otner, S.M.G., Robinson, D., Madan,
+N., Silberzahn, R., Goldstein, P., Tierney, W., Murase, T., Mandl, B.,
+Viganola, D., Strobl, C., Schaumans, C.B.C., Kelchtermans, S., Naseeb,
+C., Mason Garrison, S., Yarkoni, T., Richard Chan, C.S., Adie, P.,
+Alaburda, P., Albers, C., Alspaugh, S., Alstott, J., Nelson, A.A., Ariño
+de la Rubia, E., Arzi, A., Bahník, Š., Baik, J., Winther Balling, L.,
+Banker, S., AA Baranger, D., Barr, D.J., Barros-Rivera, B., Bauer, M.,
+Blaise, E., Boelen, L., Bohle Carbonell, K., Briers, R.A., Burkhard, O.,
+Canela, M.-A., Castrillo, L., Catlett, T., Chen, O., Clark, M., Cohn,
+B., Coppock, A., Cugueró-Escofet, N., Curran, P.G., Cyrus-Lai, W., Dai,
+D., Valentino Dalla Riva, G., Danielsson, H., F. S. M. Russo, R. de, de
+Silva, N., Derungs, C., Dondelinger, F., Duarte de Souza, C., Tyson
+Dube, B., Dubova, M., Mark Dunn, B., Adriaan Edelsbrunner, P., Finley,
+S., Fox, N., Gnambs, T., Gong, Y., Grand, E., Greenawalt, B., Han, D.,
+Hanel, P.H.P., Hong, A.B., Hood, D., Hsueh, J., Huang, L., Hui, K.N.,
+Hultman, K.A., Javaid, A., Ji Jiang, L., Jong, J., Kamdar, J., Kane, D.,
+Kappler, G., Kaszubowski, E., Kavanagh, C.M., Khabsa, M., Kleinberg, B.,
+Kouros, J., Krause, H., Krypotos, A.-M., Lavbič, D., Ling Lee, R.,
+Leffel, T., Yang Lim, W., Liverani, S., Loh, B., Lønsmann, D., Wei Low,
+J., Lu, A., MacDonald, K., Madan, C.R., Hjorth Madsen, L., Maimone, C.,
+Mangold, A., Marshall, A., Ester Matskewich, H., Mavon, K., McLain,
+K.L., McNamara, A.A., McNeill, M., Mertens, U., Miller, D., Moore, B.,
+Moore, A., Nantz, E., Nasrullah, Z., Nejkovic, V., Nell, C.S., Arthur
+Nelson, A., Nilsonne, G., Nolan, R., O'Brien, C.E., O'Neill, P., O'Shea,
+K., Olita, T., Otterbacher, J., Palsetia, D., Pereira, B., Pozdniakov,
+I., Protzko, J., Reyt, J.-N., Riddle, T., (Akmal) Ridhwan Omar Ali, A.,
+Ropovik, I., Rosenberg, J.M., Rothen, S., Schulte-Mecklenbeck, M.,
+Sharma, N., Shotwell, G., Skarzynski, M., Stedden, W., Stodden, V.,
+Stoffel, M.A., Stoltzman, S., Subbaiah, S., Tatman, R., Thibodeau, P.H.,
+Tomkins, S., Valdivia, A., Druijff-van de Woestijne, G.B., Viana, L.,
+Villesèche, F., Duncan Wadsworth, W., Wanders, F., Watts, K., Wells,
+J.D., Whelpley, C.E., Won, A., Wu, L., Yip, A., Youngflesh, C., Yu,
+J.-C., Zandian, A., Zhang, L., Zibman, C., Luis Uhlmann, E., 2021. Same
+data, different conclusions: Radical dispersion in empirical results
+when independent analysts operationalize and test the same hypothesis.
+Organizational Behavior and Human Decision Processes 165, 228--249.
+doi:\url{https://doi.org/10.1016/j.obhdp.2021.02.003}
+
 \leavevmode\hypertarget{ref-Sharifi2020covid}{}%
@@
 
+\leavevmode\hypertarget{ref-Viglione2020avalanche}{}%
+Viglione, G., 2020. 'Avalanche'of spider-paper retractions shakes
+behavioural-ecology community. Nature 578, 199--201.
+
 \leavevmode\hypertarget{ref-Vlasschaert2020proliferation}{}%

### 2021-09-16 README.Rmd
Revisions after decision by Geographical Analysis

@@
 
-The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research.
+The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic in 2019 led to explosive growth in scientific research.
 Alas, much of the research in the literature lacks conditions to be reproducible, and recent publications on the association between population density and the basic reproductive number of SARS-CoV-2 are no exception.
@@
                       "Within-group standard error")) %>%
-  kable("latex",
+  kable("html",
         digits = 3,
@@
                       "$\\rho$")) %>%
-  kable("latex",
+  kable("html",
         digits = 3,

### 2021-09-30 README.Rmd
Paper accepted in Geographical Analysis

@@
 
-Paper submitted to [~~PLoS ONE~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/PLOS ONE Decision_ PONE-D-21-16874.pdf) [~~Science of the Total Environment~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Your Submission STOTEN-D-21-12077.pdf) [~~Landscape and Urban Planning~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Decision on submission to Landscape and Urban Planning.pdf) Geographical Analysis
+Paper desk-rejected by [~~PLoS ONE~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/PLOS ONE Decision_ PONE-D-21-16874.pdf) [~~Science of the Total Environment~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Your Submission STOTEN-D-21-12077.pdf) [~~Landscape and Urban Planning~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Decision on submission to Landscape and Urban Planning.pdf)
+
+Forthcoming in Geographical Analysis
 

### 2021-11-01 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Knitted the paper before sending to source files publisher for production

@@
 date: "`r Sys.Date()`"
-bibliography: bibliography.bib
+bibliography: [bibliography.bib, packages.bib]
 linenumbers: true
@@
 
-<!--
-Install data package as appropriate. The data package is file `r0density_0.1.0.tar.gz`
--->
+```{r install-data-package, include=FALSE}
+if(!require(r0density)){
+    remotes::install_github("paezha/Reproductive-Rate-and-Density-US-Reanalyzed", 
+                         subdir = "r0density")
+    library(r0density)
+}
+```
 
@@
 
+```{r generate-package-bibliography, include=FALSE}
+knitr::write_bib(c(.packages(), 
+                   "knitr", 
+                   "rmarkdown"), 
+                 file = "packages.bib")
+```
+
 ```{r invoke-data, include=FALSE}
@@
 In this spirit, the present paper also aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
-The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, will be available in a public repository^[For peer-review purposes, the data package and code are currently in an anonymous Drive folder: \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}].
+The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, are available in a public repository^[\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}].
 
@@
 
+# Acknowledgments {#acknwledgments .unnumbered}
+
+The analysis reported in this paper was conducted in the `R` computing statistical language [@R-base].
+The source document is an Rmarkdown document [@rmarkdown2018; @rmarkdown2020] processed using `knitr` [@knitr2015; @knitr2014].
+The following packages were used in the analysis, and I wish to acknowledge their creators for their generous efforts: `adespatial` [@R-adespatial], `censReg` [@R-censReg], `dplyr` [@R-dplyr], `forcats` [@R-forcats], `ggplot2` [@ggplot22016], `gmm` [@gmm2010], `kableExtra` [@R-kableExtra], `Matrix` [@R-Matrix], `maxLik` [@maxLik2011], `miscTools` [@R-miscTools], `mvtnorm` [@mvtnorm2009], `nlme` [@R-nlme], `patchwork` [@R-patchwork], `purrr` [@R-purrr], `readr` [@R-readr], `sampleSelection` [@sampleSelection2008], `sandwich` [@sandwich2020; @sandwich2004; @sandwich2006], `scico` [@R-scico], `sf` [@sf2018], `sp` [@sp2005; @sp2013], `spatialprobit` [@R-spatialprobit], `spData` [@R-spData], `spdep` [@spdep2018; @spdep2013], `stringr` [R-stringr], `tibble` [@R-tibble], `tidycensus` [@R-tidycensus], `tidyr` [@R-tidyr], `tidyverse` [@tidyverse2019], `tmvtnorm` [@R-tmvtnorm], `units` [@units2016].
+This research was not supported by Canada's Research Councils.
+
 # References {#references .unnumbered}

### 2021-11-01 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Knitted the paper before sending to source files publisher for production

@@
 \href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
-with the data package, will be available in a public
-repository\footnote{For peer-review purposes, the data package and code
-  are currently in an anonymous Drive folder:
-  \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}}.
+with the data package, are available in a public repository\footnote{\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}}.
 
@@
 
+\hypertarget{acknwledgments}{%
+\section*{Acknowledgments}\label{acknwledgments}}
+\addcontentsline{toc}{section}{Acknowledgments}
+
+The analysis reported in this paper was conducted in the \texttt{R}
+computing statistical language (R Core Team, 2021). The source document
+is an Rmarkdown document (Xie et al., 2020, 2018) processed using
+\texttt{knitr} (Xie, 2015, 2014). The following packages were used in
+the analysis, and I wish to acknowledge their creators for their
+generous efforts: \texttt{adespatial} (Dray et al., 2021),
+\texttt{censReg} (Henningsen, 2020), \texttt{dplyr} (Wickham et al.,
+2021), \texttt{forcats} (Wickham, 2021a), \texttt{ggplot2} (Wickham,
+2016), \texttt{gmm} (Chaussé, 2010), \texttt{kableExtra} (Zhu, 2021),
+\texttt{Matrix} (Bates and Maechler, 2021), \texttt{maxLik} (Henningsen
+and Toomet, 2011), \texttt{miscTools} (Henningsen and Toomet, 2019),
+\texttt{mvtnorm} (Genz and Bretz, 2009), \texttt{nlme} (Pinheiro et al.,
+2021), \texttt{patchwork} (Pedersen, 2020), \texttt{purrr} (Henry and
+Wickham, 2020), \texttt{readr} (Wickham and Hester, 2021),
+\texttt{sampleSelection} (Toomet and Henningsen, 2008),
+\texttt{sandwich} (Zeileis et al., 2020; Zeileis, 2006, 2004),
+\texttt{scico} (Pedersen and Crameri, 2020), \texttt{sf} (Pebesma,
+2018), \texttt{sp} (Bivand et al., 2013a; Pebesma and Bivand, 2005),
+\texttt{spatialprobit} (Wilhelm and de Matos, 2015), \texttt{spData}
+(Bivand et al., 2021), \texttt{spdep} (Bivand et al., 2013b; Bivand and
+Wong, 2018), \texttt{stringr} {[}R-stringr{]}, \texttt{tibble} (Müller
+and Wickham, 2021), \texttt{tidycensus} (Walker and Herman, 2021),
+\texttt{tidyr} (Wickham, 2021b), \texttt{tidyverse} (Wickham et al.,
+2019), \texttt{tmvtnorm} (Manjunath B G
+\textless bgmanjunath@gmail.com\textgreater, 2015), \texttt{units}
+(Pebesma et al., 2016). This research was not supported by Canada's
+Research Councils.
+
 \hypertarget{references}{%
@@
 
+\leavevmode\hypertarget{ref-R-Matrix}{}%
+Bates, D., Maechler, M., 2021. Matrix: Sparse and dense matrix classes
+and methods.
+
 \leavevmode\hypertarget{ref-Bhadra2021impact}{}%
@@
 
+\leavevmode\hypertarget{ref-R-spData}{}%
+Bivand, R., Nowosad, J., Lovelace, R., 2021. spData: Datasets for
+spatial analysis.
+
 \leavevmode\hypertarget{ref-Bivand2020progress}{}%
@@
 doi:\href{https://doi.org/10.1007/s10109-020-00336-0}{10.1007/s10109-020-00336-0}
+
+\leavevmode\hypertarget{ref-sp2013}{}%
+Bivand, R.S., Pebesma, E., Gomez-Rubio, V., 2013a. Applied spatial data
+analysis with {R}, second edition. Springer, NY.
+
+\leavevmode\hypertarget{ref-spdep2013}{}%
+Bivand, R.S., Pebesma, E., Gomez-Rubio, V., 2013b. Applied spatial data
+analysis with {R}, second edition. Springer, NY.
+
+\leavevmode\hypertarget{ref-spdep2018}{}%
+Bivand, R., Wong, D.W.S., 2018. Comparing implementations of global and
+local indicators of spatial association. TEST 27, 716--748.
 
@@
 
+\leavevmode\hypertarget{ref-gmm2010}{}%
+Chaussé, P., 2010. Computing generalized method of moments and
+generalized empirical likelihood with {R}. Journal of Statistical
+Software 34, 1--35.
+
 \leavevmode\hypertarget{ref-Cressie1989geostatistics}{}%
@@
 doi:\href{https://doi.org/10.1371/journal.pone.0235306}{10.1371/journal.pone.0235306}
+
+\leavevmode\hypertarget{ref-R-adespatial}{}%
+Dray, S., Bauman, D., Blanchet, G., Borcard, D., Clappe, S., Guenard,
+G., Jombart, T., Larocque, G., Legendre, P., Madi, N., Wagner, H.H.,
+2021. Adespatial: Multivariate multiscale spatial analysis.
 
@@
 
+\leavevmode\hypertarget{ref-mvtnorm2009}{}%
+Genz, A., Bretz, F., 2009. Computation of multivariate normal and t
+probabilities, Lecture notes in statistics. Springer-Verlag, Heidelberg.
+
 \leavevmode\hypertarget{ref-Getis2002comparative}{}%
@@
 
+\leavevmode\hypertarget{ref-R-censReg}{}%
+Henningsen, A., 2020. censReg: Censored regression (tobit) models.
+
+\leavevmode\hypertarget{ref-maxLik2011}{}%
+Henningsen, A., Toomet, O., 2011. maxLik: A package for maximum
+likelihood estimation in {R}. Computational Statistics 26, 443--458.
+doi:\href{https://doi.org/10.1007/s00180-010-0217-1}{10.1007/s00180-010-0217-1}
+
+\leavevmode\hypertarget{ref-R-miscTools}{}%
+Henningsen, A., Toomet, O., 2019. miscTools: Miscellaneous tools and
+utilities.
+
+\leavevmode\hypertarget{ref-R-purrr}{}%
+Henry, L., Wickham, H., 2020. Purrr: Functional programming tools.
+
 \leavevmode\hypertarget{ref-Herndon2014high}{}%
@@
 econometrics. Cambridge University Press, Cambridge.
+
+\leavevmode\hypertarget{ref-R-tmvtnorm}{}%
+Manjunath B G \textless bgmanjunath@gmail.com\textgreater, S.W. with
+contributions from, 2015. Tmvtnorm: Truncated multivariate normal and
+student t distribution.
 
@@
 
+\leavevmode\hypertarget{ref-R-tibble}{}%
+Müller, K., Wickham, H., 2021. Tibble: Simple data frames.
+
 \leavevmode\hypertarget{ref-Noland1995perceived}{}%
@@
 doi:\href{https://doi.org/10.1111/gean.12241}{10.1111/gean.12241}
+
+\leavevmode\hypertarget{ref-sf2018}{}%
+Pebesma, E., 2018. {Simple Features for R: Standardized Support for
+Spatial Vector Data}. {The R Journal} 10, 439--446.
+doi:\href{https://doi.org/10.32614/RJ-2018-009}{10.32614/RJ-2018-009}
+
+\leavevmode\hypertarget{ref-sp2005}{}%
+Pebesma, E.J., Bivand, R.S., 2005. Classes and methods for spatial data
+in {R}. R News 5, 9--13.
+
+\leavevmode\hypertarget{ref-units2016}{}%
+Pebesma, E., Mailund, T., Hiebert, J., 2016. Measurement units in {R}. R
+Journal 8, 486--494.
+doi:\href{https://doi.org/10.32614/RJ-2016-061}{10.32614/RJ-2016-061}
+
+\leavevmode\hypertarget{ref-R-patchwork}{}%
+Pedersen, T.L., 2020. Patchwork: The composer of plots.
+
+\leavevmode\hypertarget{ref-R-scico}{}%
+Pedersen, T.L., Crameri, F., 2020. Scico: Colour palettes based on the
+scientific colour-maps.
 
@@
 
+\leavevmode\hypertarget{ref-R-nlme}{}%
+Pinheiro, J., Bates, D., R-core, 2021. Nlme: Linear and nonlinear mixed
+effects models.
+
 \leavevmode\hypertarget{ref-Praharaj2020Using}{}%
@@
 doi:\href{https://doi.org/10.32866/001c.17590}{10.32866/001c.17590}
+
+\leavevmode\hypertarget{ref-R-base}{}%
+R Core Team, 2021. R: A language and environment for statistical
+computing. R Foundation for Statistical Computing, Vienna, Austria.
 
@@
 
+\leavevmode\hypertarget{ref-sampleSelection2008}{}%
+Toomet, O., Henningsen, A., 2008. Sample selection models in {R}:
+Package {sampleSelection}. Journal of Statistical Software 27.
+
 \leavevmode\hypertarget{ref-Viglione2020avalanche}{}%
@@
 
+\leavevmode\hypertarget{ref-R-tidycensus}{}%
+Walker, K., Herman, M., 2021. Tidycensus: Load US census boundary and
+attribute data as tidyverse and sf-ready data frames.
+
 \leavevmode\hypertarget{ref-Wang2021transmission}{}%
@@
 
+\leavevmode\hypertarget{ref-ggplot22016}{}%
+Wickham, H., 2016. ggplot2: Elegant graphics for data analysis.
+Springer-Verlag New York.
+
+\leavevmode\hypertarget{ref-R-forcats}{}%
+Wickham, H., 2021a. Forcats: Tools for working with categorical
+variables (factors).
+
+\leavevmode\hypertarget{ref-R-tidyr}{}%
+Wickham, H., 2021b. Tidyr: Tidy messy data.
+
+\leavevmode\hypertarget{ref-tidyverse2019}{}%
+Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L.D., François,
+R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen,
+T.L., Miller, E., Bache, S.M., Müller, K., Ooms, J., Robinson, D.,
+Seidel, D.P., Spinu, V., Takahashi, K., Vaughan, D., Wilke, C., Woo, K.,
+Yutani, H., 2019. Welcome to the {tidyverse}. Journal of Open Source
+Software 4, 1686.
+doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}
+
+\leavevmode\hypertarget{ref-R-dplyr}{}%
+Wickham, H., François, R., Henry, L., Müller, K., 2021. Dplyr: A grammar
+of data manipulation.
+
+\leavevmode\hypertarget{ref-R-readr}{}%
+Wickham, H., Hester, J., 2021. Readr: Read rectangular text data.
+
+\leavevmode\hypertarget{ref-R-spatialprobit}{}%
+Wilhelm, S., de Matos, M.G., 2015. Spatialprobit: Spatial probit models.
+
 \leavevmode\hypertarget{ref-Wong2020spreading}{}%
@@
 
+\leavevmode\hypertarget{ref-knitr2014}{}%
+Xie, Y., 2014. Knitr: A comprehensive tool for reproducible research in
+{R}, in: Stodden, V., Leisch, F., Peng, R.D. (Eds.), Implementing
+Reproducible Computational Research. Chapman; Hall/CRC.
+
+\leavevmode\hypertarget{ref-knitr2015}{}%
+Xie, Y., 2015. Dynamic documents with {R} and knitr, 2nd ed. Chapman;
+Hall/CRC, Boca Raton, Florida.
+
+\leavevmode\hypertarget{ref-rmarkdown2018}{}%
+Xie, Y., Allaire, J.J., Grolemund, G., 2018. R markdown: The definitive
+guide. Chapman; Hall/CRC, Boca Raton, Florida.
+
+\leavevmode\hypertarget{ref-rmarkdown2020}{}%
+Xie, Y., Dervieux, C., Riederer, E., 2020. R markdown cookbook. Chapman;
+Hall/CRC, Boca Raton, Florida.
+
+\leavevmode\hypertarget{ref-sandwich2004}{}%
+Zeileis, A., 2004. Econometric computing with {HC} and {HAC} covariance
+matrix estimators. Journal of Statistical Software 11, 1--17.
+doi:\href{https://doi.org/10.18637/jss.v011.i10}{10.18637/jss.v011.i10}
+
+\leavevmode\hypertarget{ref-sandwich2006}{}%
+Zeileis, A., 2006. Object-oriented computation of sandwich estimators.
+Journal of Statistical Software 16, 1--16.
+doi:\href{https://doi.org/10.18637/jss.v016.i09}{10.18637/jss.v016.i09}
+
+\leavevmode\hypertarget{ref-sandwich2020}{}%
+Zeileis, A., Köll, S., Graham, N., 2020. Various versatile variances: An
+object-oriented implementation of clustered covariances in {R}. Journal
+of Statistical Software 95, 1--36.
+doi:\href{https://doi.org/10.18637/jss.v095.i01}{10.18637/jss.v095.i01}
+
+\leavevmode\hypertarget{ref-R-kableExtra}{}%
+Zhu, H., 2021. kableExtra: Construct complex table with kable and pipe
+syntax.
+
 \end{CSLReferences}

### 2021-11-01 README.Rmd
Knitted the paper before sending to source files publisher for production

@@
 always_allow_html: true
-bibliography: R0-Density-Reanalysis/bibliography.bib
+bibliography: [R0-Density-Reanalysis/bibliography.bib, R0-Density-Reanalysis/packages.bib]
 ---
@@
 
+```{r generate-package-bibliography, include=FALSE}
+knitr::write_bib(c(.packages(), 
+                   "knitr", 
+                   "rmarkdown"), 
+                 file = "R0-Density-Reanalysis/packages.bib")
+```
+
 ```{r invoke-data, include=FALSE}
@@
 In this spirit, the present paper also aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
-The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, will be available in a public repository^[For peer-review purposes, the data package and code are currently in an anonymous Drive folder: \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}].
+The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, are available in a public repository^[\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}]
 

### 2021-11-18 R0-Density-Reanalysis/Editorial-Materials-and-Correspondence/R0-Density-Reanalysis-Abstract.Rmd
Reorganized the repository after paper's publication online at Geographical Analysis

--- 2021-09-16 README.Rmd
@@
 ---
-output:
-  github_document:
-    pandoc_args: --webtex
-always_allow_html: true
-bibliography: R0-Density-Reanalysis/bibliography.bib
----
-
-<!-- README.md is generated from README.Rmd. Please edit that file -->
-
-```{r, include = FALSE}
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>"
-)
-```
-
-# Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis
-
-<!-- badges: start -->
-<!-- badges: end -->
-
-Antonio Páez (McMaster University)
-
-Paper submitted to [~~PLoS ONE~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/PLOS ONE Decision_ PONE-D-21-16874.pdf) [~~Science of the Total Environment~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Your Submission STOTEN-D-21-12077.pdf) [~~Landscape and Urban Planning~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Decision on submission to Landscape and Urban Planning.pdf) Geographical Analysis
-
-## Abstract
-
+title: "Population density and the spread of the COVID-19 pandemic: a reproducible research example"
+abstract: |
   The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic in 2019 led to explosive growth in scientific research.
@@
 
-## Keywords
-
-- COVID-19  
-- Reproducible research  
-- Population density  
-- Basic reproductive number
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(cache = TRUE)
-```
-
-<!--
-Install data package as appropriate. The data package is file `r0density_0.1.0.tar.gz`
--->
-
-```{r load-packages, include=FALSE, cache=FALSE}
-library(adespatial) # Used to calculate Moran eigenvector maps
-library(censReg) # Package to estimate tobit model
-library(kableExtra) # Package for creating tables
-library(Matrix) # Used to convert list of spatial weights to sparse matrix; needed for spatialprobit
-library(nlme) # Used to estimate 
-library(patchwork) # Package to organize ggplot2 plots
-library(r0density) # Data package
-library(sampleSelection) # Package to estimate sampling selection models
-library(scico) # Package with color palettes
-library(sf) # Package to work with spatial information in simple features format
-library(spatialprobit) # Used to estimate spatially autoregressive tobit
-library(spdep) # Utilities to create spatial weights and work with spatial data
-library(tidycensus) # Package to retrieve census information 
-library(tidyverse) # Data carpentry and analysis tools
-library(units) # Package to work with units of measurement
-#library(viridis) # Color palettes
-```
-
-```{r invoke-data, include=FALSE}
-data(county_geo)
-data(state_geo)
-data(urba_geo)
-```
-
-```{r prepare-data, include=FALSE}
-# Convert missing R's to zeros and calculate expansion variables: 
-county_geo_clean <- county_geo %>%
-  mutate(R = replace_na(R, 0),
-         density_log = log(drop_units(density)),
-         density_2 = drop_units(density)^2,
-         hincome_log = log(hincome),
-         private = car/commuters * 100) %>% # Percentage of commuters who travel by car/van/truck (in 10%)
-  drop_na(commuters) %>%
-  st_as_sf()
-
-# Boundary of continental US excluding Alaska
-us_geo <- st_union(state_geo %>% filter(!STATE %in% c("02", "15", "72")))
-```
-
-```{r list-of-geographical-neighbors, include=FALSE}
-# Create a list of neighbors based on the polygons; the default contiguity criterion is Queen:
-# Counties
-county_nb <- county_geo_clean %>%
-  poly2nb()
-# States
-state_nb <- state_geo %>% 
-  filter(!STATE %in% c("02", "15", "72")) %>%
-  poly2nb()
-```
-
-```{r spatial-data-check, include=FALSE}
-# Check list of neighbors
-summary(county_nb)
-
-# Notice that there are 9 counties that are isolates. These are the counties that do not have neighbors based on contiguity
-county_geo_clean[c(69, 546, 547, 549, 1226, 1876, 2974, 3146, 3200),]
-
-# To rectify the lack of neighbors, find the first nearest neighbor using the centroids:
-county_nb_dist <- knearneigh(county_geo_clean %>% 
-                               st_centroid()) %>%
-  knn2nb()
-
-# Add nearest neighbor to isolates:
-county_nb[[69]] <- county_nb_dist[[69]]
-county_nb[[546]] <- county_nb_dist[[546]]
-county_nb[[547]] <- county_nb_dist[[547]]
-county_nb[[549]] <- county_nb_dist[[549]]
-county_nb[[1226]] <- county_nb_dist[[1226]]
-county_nb[[1876]] <- county_nb_dist[[1876]]
-county_nb[[2974]] <- county_nb_dist[[2974]]
-county_nb[[3146]] <- county_nb_dist[[3146]]
-county_nb[[3200]] <- county_nb_dist[[3200]]
-```
-
-```{r spatial-weights, include=FALSE}
-# Convert the nearest neighbor object to a listw object:
-nb_B <- nb2listw(county_nb, 
-                 style="B", 
-                 zero.policy = TRUE)
-
-county_listw <- nb2listw(county_nb, 
-                        style="W", 
-                        zero.policy = TRUE)
-
-state_listw <- nb2listw(state_nb, 
-                        style="W", 
-                        zero.policy = TRUE)
-
-# Convert listw to sparse matrix for use in spatial tobit:
-B <- as(nb_B, "CsparseMatrix")
-```
-
-```{r moran-eigenvector-maps, include=FALSE}
-county_mem <- mem(county_listw)
-```
-
-# Introduction
-
-The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the global pandemic that followed in its wake, led to an explosive growth of research around the globe.
-According to Fraser et al.
-[-@Fraser2021evolving], over 125,000 COVID-19-related papers were released in the first ten months from the first confirmed case of the disease.
-Of these, more than 30,000 were shared in pre-print servers, the use of which also exploded in the past year [@Kwon2021swamped; @Vlasschaert2020proliferation; @Anazco2021publication].
-
-Given the ruinous human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research termed the deluge of COVID-19 publications a "carnage of substandard research" [@Bramstedt2020carnage].
-Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [e.g., @Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty and somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational].
-As reported in the literature, only a woefully small proportion of published research was actually reproducible before the pandemic [@Iqbal2016reproducible; @Stodden2018empirical], and the situation does not appear to have changed substantially since [@Sumner2020reproducibility; @Gustot2020quality].
-
-The push for open software and data [e.g., @Bivand2020progress; @Arribas2021open], along with more strenuous efforts towards open, reproducible research, is simply a continuation of long-standing scientific practices of independent verification.
-Despite the (at times disproportionate) attention that high profile scandals in science tend to elicit in the media, science as a collective endeavor is remarkable for being a self-correcting enterprise, one with built-in mechanisms and incentives to weed out erroneous ideas.
-Over the long term, facts tend to prevail in science.
-At stake is the shorter-term impacts that research may have in other spheres of economic and social life.
-The case of economists Reinhart and Rogoff comes to mind: by the time the inaccuracies and errors in their research were uncovered [see @Herndon2014high], their claims about debt and economic growth had already been seized by policy-makers on both sides of the Atlantic to justify austerity policies in the aftermath of the Great Recession of 2007-2009^[Nobel Prize in Economics Paul Krugman noted that "Reinhart–Rogoff may have had more immediate influence on public debate than any previous paper in the history of economics" \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}].
-As later research has demonstrated, those policies cast a long shadow, and their sequels continued to be felt for years [@Basu2017ten].
-
-In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [e.g., @Florida2020how]; as a result, the implications of the pandemic for urban planning, design, and management are the topic of ongoing research [e.g., @Sharifi2020covid].
-The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent of research into the associations between density and the spread of the pandemic.
-The answers to some important questions hang on the results of these research efforts.
-For example, are lower density regions safer from the pandemic?
-Are de-densification policies warranted, even if just in the short term?
-In the longer term, will the risks of life in high density regions presage a flight from cities?
-And, what are the implications of the pandemic for future urban planning and practice?
-Over the past year, numerous papers have sought to throw light on the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed.
-Further, to complicate matters, precious few of these studies appear to be sufficiently open to support independent verification.
-
-The objective of this paper is to illustrate the importance of reproducibility in research in the context of the flood of COVID-19 papers.
-For this, I focus on a recent study by Sy et al.
-[-@Sy2021population] that examined the correlation between the basic reproductive number of COVID-19, $R_0$, and population density.
-The basic reproductive number is a summary measure of contact rates, probability of transmission of a pathogen, and duration of infectiousness.
-In rough terms, it measures how many new infections each infections begets.
-The paper of Sy et al.
-[-@Sy2021population] was selected for being, in the literature examined, almost alone in supporting reproducible research.
-Accordingly, I wish to be clear that my objective in singling their work for discussion is not to malign their efforts, but rather to demonstrate how open and reproducible research efforts can greatly help to accelerate discovery.
-More concretely, open data and open code mean that an independent researcher can, with only modest efforts, not only verify the findings reported, but also examine the same data from a perspective which may not have been available to the original researchers due to differences in disciplinary perspectives, methodological traditions, and/or training, among other possible factors.
-The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in their research.
-In this spirit, the present paper also aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification.
-The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, will be available in a public repository^[For peer-review purposes, the data package and code are currently in an anonymous Drive folder: \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}].
-
-# Background: the intuitive relationship between density and spread of contagious diseases
-
-The concern with population density and the spread of the virus during the COVID-19 pandemic was fueled, at least in part, by dramatic scenes seen in real-time around the world from large urban centers such as Wuhan, Milan, Madrid, and New York.
-In theory, there are good reasons to believe that higher density could have a positive association with the transmission of a contagious virus.
-It has long been known that the potential for inter-personal contact is greater in regions with higher density [see for example the research on urban fields and time-geography, including @Farber2011running; @Moore1970urban; @Moore1970some].
-Mathematically, models of exposure and contagion indicate that higher densities can catalyze the transmission of contagious diseases [@Rocklov2020high; @Li2018effect].
-The idea is intuitive and likely at the root of messages, by some figures in positions of authority, that regions with sparse population densities faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic down without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].
-
-As Rocklöv and Sjödin [@Rocklov2020high] note, however, mathematical models of contagion are valid at small-to-medium spaces (and presumably, smaller time intervals too, such as time spent in restaurants, concert halls, cruises), and the results do not necessarily transfer to larger spatial units and longer time periods.
-There are solid reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers.
-On the other hand, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place.
-Nonetheless, the idea that high density correlates with high transmission is so seemingly sensible that it is often taken for granted even at the scale of large spaces [e.g., @Cruz2020exploring; @Micallef2020first].
-In such conditions, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, e.g., @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.
-
-A plausible behavioral adaptation during a pandemic, especially one broadcast as widely and intensely as COVID-19, is risk compensation.
-Risk compensation is a process whereby people adjust their behavior in response to their _perception_ of risk [@Noland1995perceived; @Richens2000condoms; @Phillips2011risk].
-In the case of COVID-19, Chauhan et al.
-[@Chauhan2021covid] have found that perception of risks in the US varies between rural, suburban, and urban residents, with rural residents in general expressing less concern about the virus.
-It is possible that people who listened to the message of leaders saying that they were safe from the virus because of low density may not have taken adequate precautions.
-Conversely, people in dense places who could more directly observe the impact of the pandemic may have become overly cautious.
-Both Paez et al.
-[-@Paez2020spatio] and Hamidi et al.
-[-@Hamidi2020density] posit this mechanism (i.e., greater compliance with social distancing in denser regions) to explain the results of their analyses.
-The evidence available does indeed show that there were important changes in behavior with respect to mobility during the pandemic [@Jamal2020Changes; @Harris2021Changes; @Molloy2020Tracing]; furthermore, shelter in place orders may have had greater buy-in from the public in higher density regions [@Feyman2020effectiveness; @Hamidi2021compact], and the associated behavior may have persisted beyond the duration of official social-distancing policies [@Praharaj2020Using].
-In addition, there is evidence that changes in mobility correlated with the trajectory of the pandemic [@Paez2020using; @Noland2021mobility].
-Given the potential for behavioral adaptation, the question of density becomes more nuanced: it is not just a matter of proximity, but also of human behavior, which is better studied using population-level data and models.
-
-# Background: but what does the literature say?
-
-When it comes to population density and the spread of COVID-19, the international literature to date remains inconclusive.
-
-On the one hand, there are studies that report positive associations between population density and various COVID-19-related outcomes.
-Bhadra [-@Bhadra2021impact], for example, reported a moderate positive correlation between the spread of COVID-19 and population density at the district level in India, however their analysis was bivariate and did not control for other variables, such as income.
-Similarly, Kadi and Khelfaoui [-@Kadi2020population] found a positive and significant correlation between number of cases and population density in cities in Algeria in a series of simple regression models (i.e., without other controls).
-A question in these relatively simple analyses is whether density is not a proxy for other factors.
-Other studies have included controls, such as Pequeno et al.
-[-@Pequeno2020air], a team that reported a positive association between density and cumulative counts of confirmed COVID-19 cases in state capitals in Brazil after controlling for covariates, including income, transport connectivity, and economic status.
-In a similar vein, Fielding-Miller et al.
-[-@Fielding2020social] reported a positive relationship between the absolute number of COVID-19 deaths and population density (rate) in rural counties in the US.
-Roy and Ghosh [-@Roy2020factors] used a battery of machine learning techniques to find discriminatory factors, and a positive and significant association between COVID-19 infection and death rates in US states.
-Wong and Li [-@Wong2020spreading] also found a positive and significant association between population density and number of confirmed COVID-19 cases in US counties, using both univariate and multivariate regressions with spatial effects.
-More recently, Sy et al.
-[-@Sy2021population] reported that the basic reproductive number of COVID-19 in US counties tended to increase with population density, but at a decreasing rate at higher densities.
-
-On the flip side, a number of studies report non-significant or negative associations between population density and COVID-19 outcomes.
-This includes the research of Sun et al.
-[-@Sun2020impacts] who did not find evidence of significant correlation between population density and confirmed number of cases per day _in conditions of lockdown_ in China.
-This finding echoes the results of Paez et al.
-[-@Paez2020spatio], who in their study of provinces in Spain reported non-significant associations between population density and infection rates in the early days of the first wave of COVID-19, and negative significant associations in the later part of the first lockdown.
-Similarly, @Skorka2020macroecology found zero or negative associations between population density and infection numbers/deaths by country.
-Fielding-Miller et al.
-[-@Fielding2020social] contrast their finding about rural counties with a negative relationship between COVID-19 deaths and population density in urban counties in the US.
-For their part, in their investigation of doubling time, White and Hébert-Dufresne [-@White2020state] identified a negative and significant correlation between population density and doubling time in US states.
-Likewise, @Khavarian2021high found a small negative (and significant) association between population density and COVID-19 morbidity in districts in Tehran.
-Finally, two of the most complete studies in the US, by @Hamidi2020longitudinal and @Hamidi2020density, used an extensive set of controls to find negative and significant correlations between density and COVID-19 cases and fatalities at the level of counties in the US.
-
-As can be seen, these studies are implemented at different scales in different regions of the world.
-They also use a range of techniques, from correlation analysis, to multivariate regression, spatial regressions, and machine learning techniques.
-This is natural and to be expected: individual researchers have only limited time and expertise.
-This is why reproducibility is important.
-To pick an example (which will be further elaborated in later sections of this paper), the study of Sy et al.
-[-@Sy2021population], hereafter referred to as SWN, would immediately grab the attention of a researcher with expertise in spatial analysis.
-
-# Reproducibility of research
-
-SWN investigated the basic reproductive number of COVID-19 in US counties, and its association with population density, median household income, and prevalence of private mobility.
-For their multivariate analysis, SWN used mixed linear models.
-This is an appropriate modelling choice: $R_0$ is an interval-ratio variable that is suitably modeled using linear regression; further, as SWN note there is a likelihood that the process in not independent "among counties within each state, potentially due to variable resource allocation and differing health systems across states"  (p.
-3).
-A mixed linear model accounts for this by introducing random components; in the case of SWN, these are random intercepts at the state level.
-SWN estimated various models with different combinations of variables, including median household income and prevalence of travel by private transportation.
-These controls help to account for potential variations in behavior: people in more affluent counties may have greater opportunities to work from home, and use of private transportation reduces contact with strangers.
-Moreover, they also conducted various sensitivity analyses.
-After these efforts, SWN concluded that there is a positive association between the basic reproductive number and population density at the level of counties in the US.
-
-One salient aspect of the analysis in SWN is that the basic reproductive number can only be calculated reliably with a minimum number of cases, and a large number of counties did not meet such threshold.
-As researchers do, SWN made modelling decisions, in this case basing their analysis only on counties with valid observations.
-A modeler with expertise in spatial analysis would likely ask some of the following questions on reading SWN's paper: how were missing counties treated?
-What are the implications of the spatial sampling framework used in the analysis?
-Is it possible to spatially interpolate the missing observations?
-Was there spatial residual autocorrelation in the models, or was the use of mixed models sufficient to capture spatial dependencies?
-These questions are relevant and their implications important.
-Fortunately, SWN are an example of a reasonably open, reproducible research product: their paper is accompanied by (most of) the data and (most of) the code used in the analysis.
-This means that an independent researcher can, with only a moderate investment of time and effort, reproduce the results in the paper, as well as ask additional questions.
-
-Alas, reproducibility is not necessarily the norm in the relevant literature.
-
-There are various reasons why a project can fail to be reproducible.
-In some cases, there might be legitimate reasons to withhold the data, perhaps due to confidentiality and privacy reasons [e.g., @Lee2020human].
-But in many other cases the data are publicly available, which in fact has commonly been the case with population-level COVID-19 information.
-Typically the provenance of the data is documented, but in numerous studies the data themselves are not shared [@Amadu2021assessing; @Bhadra2021impact; @Cruz2020exploring; @Feng2020spread; @Fielding2020social; @Hamidi2020longitudinal; @Hamidi2020density; @Inbaraj2021seroprevalence; @Souris2020covid].
-As any researcher can attest, collecting, organizing, and preparing data for a project can take a substantial amount of time.
-Pointing to the sources of data, even when these sources are public, is a small step towards reproducibility-but only a very small one.
-Faced with the prospect of having to recreate a data set from raw sources is probably sufficient to dissuade all but the most dedicated (or stubborn) researcher from independent verification.
-This is true even if part of the data are shared [e.g., @Wong2020spreading].
-In other cases, data are shared, but the processes followed in the preparation of the data are not fully documented [@Ahmad2020association; @Skorka2020macroecology].
-These processes matter, as shown by the errors in the spreadsheets of Reinhart and Rogoff [see @Herndon2014high for the discovery of these errors], as well as by the data of biologist Jonathan Pruitt that led to an "avalanche" of paper retractions [see @Viglione2020avalanche].
-Another situation is when papers share well-documented data, but fail to provide the code used in the analysis [@Noury2021how; @Pequeno2020air; @Wang2021transmission].
-Making code available only "on demand" [e.g., @Brandtner2021creatures] is an unnecessary barrier when most journals offer the facility to share supplemental materials online.
-Then there are those papers that more closely comply with reproducibility standards, and share well-documented processes and data, as well as the code used in any analyses reported [@Paez2020spatio; @Feyman2020effectiveness; @Stephens2021impact; @White2020state; @Sy2021population].
-Even in this case, the pressure to publish "new findings" instead of replication studies can act as a deterrent, perhaps particularly for younger researchers^[The present paper was desk rejected by three journals that had previously published research on population density and the spread of COVID-19; in one case, the paper was too opinionated for the journal, in the other two cases, the paper was not a "good fit" despite dealing with a nearly identical issue as papers previously published in said journals.].
-
-<!--
-@Ahmad2020association -> code is not shared, data are provided but data pre-processing is not documented
-@Amadu2021assessing -> code is not shared, data provenance is documented, data are not shared
-@Bhadra2021impact -> (code and data not available; only data provenance is documented)
-@Brandtner2021creatures -> code is available on demand, data are shared
-@Cruz2020exploring -> code is shared, sources of data are documented, data are not actually provided
-@Feng2020spread -> code not shared, data provenance is documented, data are not shared
-@Feyman2020effectiveness -> code and data are provided
-@Fielding2020social -> code is not shared, data provenance is documented, data are not actually provided
-@Hamidi2020longitudinal, @Hamidi2020density ->  (code and data not available)
-@Inbaraj2021seroprevalence -> code is not shared, data are not shared
-@Lee2020human -> code not provided, data sources are documented but data are not shared (privacy & confidentiality)
-@Noury2021how -> code is not shared, data are shared
-@Pequeno2020air -> (raw data supplied in suplemental files, code not available)
-@Roy2020factors -> code is not shared, data are provided and data processing is documented
-@Skorka2020macroecology -> code and data are shared but data processing is not documented, 
-@Souris2020covid -> code not provided, data sources are documented but data are not shared
-@Stephens2021impact -> code and data are shared
-@Wang2021transmission -> code is not shared, data are available
-@White2020state -> code and data are in repository
-@Wong2020spreading -> code not provided, data are shared but data preprocessing is not documented
--->
-
-In the following sections, the analysis of SWN is reproduced, some relevant questions from the perspective of an independent researcher with expertise in spatial analysis are asked, and the data are reanalyzed.
-
-# Reproducing SWN
-
-SWN examined the association between the basic reproductive number of COVID-19 and population density.
-The basic reproductive number $R_0$ is a summary measure of contact rates, probability of transmission of a pathogen, and duration of infectiousness.
-In rough terms, $R_0$ measures how many new infections each infections begets.
-Infectious disease outbreaks generally tend to die out when $R_0<1$, and to grow when $R_0>1$.
-Reliable calculation of $R_0$ requires a minimum number of cases to be able to assume that there is community transmission of the pathogen.
-Accordingly, SWN based their analysis only on counties that had at least 25 cases or more at the end of the exponential growth phase (see Fig.
-\ref{fig:R0-map}).
-Their final sample included `r prettyNum(nrow(county_geo %>% filter(!is.na(R))), big.mark = ",")` counties in the US, including in Alaska, Hawaii, Puerto Rico, and island territories.
-SWN used COVID-19 data collected by the New York Times and made available (with versioning) in a GitHub repository^[https://github.com/nytimes/covid-19-data].
-For each county, SWN assumed that the exponential growth period began one week prior to the second daily increase in cases, and assumed that the period of exponential growth lasted approximately 18 days.
-
-```{r R0-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
-# Choropleth map of basic reproductive number
-ggplot() + # Create a ggplot
-  # Choropleth map of basic reproductive number
-  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
-            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
-          aes(fill = R),
-          color = NA) +
-  # Use viridis color palette: scico color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
-  scale_fill_scico(name = expression(R[0]),
-                       direction = 1,
-                   palette = "lajolla") +
-  # Overlay counties with missing values of the reproductive number in white
-  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
-            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
-                   R == 0), 
-          fill = "white",
-          color = NA) +
-  # Overlay county boundaries
-  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
-            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
-          size = 0.05,
-          fill = NA,
-          color = "gray60") +
-  # Add boundary of continental US
-  geom_sf(data = us_geo,
-          fill = NA) + 
-  # Add caption
-  labs(caption = "Note: counties in white represent missing values of the basic reproductive number") +
-  # Set theme of plot
-  theme_minimal() +
-  theme(legend.position = "bottom")
-```
-
-Table \ref{tab:swn-results} reproduces the first three models of SWN (the fourth model did not have any significant variables; see Table 1 in SWN).
-It is possible to verify that the results match, with only the minor (and irrelevant) exception of the magnitude of the coefficient for travel by private transportation, which is due to a difference in the input (here the variable is changed to one percent units, instead of the ten percent units used by SWN).
-The mixed linear model gives random intercepts (i.e., the intercept is a random variable), and the standard deviation is reported in the fifth row of Table \ref{tab:swn-results}.
-It is useful to map the random intercepts: as seen in Figure \ref{fig:random-terms-map}, other things being equal, counties in Texas tend to have somewhat lower values of $R_0$ (i.e., a negative random intercept), whereas counties in South Dakota tend to have higher values of $R_0$.
-The key of the analysis, after extensive sensitivity analysis, is a robust finding that population density has a positive association with the basic reproductive number.
-But does it?
-
-```{r reproduce-swn, include=FALSE}
-# Fit (mixed) linear models as in Sy et al. These models are reported in table 1 of their paper
-
-# Model 1
-model1 <- lme(R ~ density_log , 
-              random = ~ 1| state, 
-              data = county_geo_clean %>%
-                filter(R > 0)) # Exclude observations with missing R
-#summary(table1model1)
-model1_intervals <- intervals(model1)
-
-# Model 2
-model2 <- lme(R ~ density_log + private , 
-              random = ~ 1| state, 
-              data = county_geo_clean %>%
-                filter(R > 0)) # Exclude observations with missing R
-#summary(model2)
-model2_intervals <- intervals(model2)
-
-# Model 3
-model3 <- lme(R ~ density_log + private + hincome, 
-              random = ~ 1| state, 
-              data = county_geo_clean %>%
-                filter(R > 0)) # Exclude observations with missing R
-#summary(model3)
-model3_intervals <- intervals(model3)
-
-# Model 4
-model4 <- lme(R ~ density_log*private + hincome, 
-              random = ~ 1| state, 
-              data = county_geo_clean %>%
-                filter(R > 0)) # Exclude observations with missing R
-#summary(model4)
-model4_intervals <- intervals(model4)
-```
-
-```{r join-random-intercepts, include=FALSE}
-state_geo <- state_geo %>% 
-  left_join(model3$coefficients$random$state %>%
-              as.data.frame() %>%
-              rownames_to_column(var = "state"),
-            by = "state") %>%
-  rename(random_intercepts = `(Intercept)`)
-```
-
-```{r prepare-swn-results-for-table, include=FALSE}
-# Join tables with results
-main_effects <- full_join(model1_intervals$fixed %>%
-                            as.data.frame() %>%
-                            rownames_to_column(),
-                          model2_intervals$fixed %>%
-                            as.data.frame() %>%
-                            rownames_to_column(),
-                          by = "rowname") %>%
-  full_join(model3_intervals$fixed %>%
-              as.data.frame() %>%
-              rownames_to_column(),
-            by = "rowname")
-
-random_component <- full_join(model1_intervals$reStruct$state %>%
-                                as.data.frame() %>%
-                                rownames_to_column(),
-                              model2_intervals$reStruct$state %>%
-                                as.data.frame() %>%
-                                rownames_to_column(),
-                              by = "rowname") %>%
-  full_join(model3_intervals$reStruct$state %>%
-              as.data.frame() %>%
-              rownames_to_column(),
-            by = "rowname")
-
-within_group_se <- full_join(model1_intervals$sigma %>%
-                               t() %>%
-                               as.data.frame() %>%
-                               rownames_to_column(),
-                             model2_intervals$sigma %>%
-                               t() %>%
-                               as.data.frame() %>%
-                               rownames_to_column(),
-                             by = "rowname") %>%
-  full_join(model3_intervals$sigma %>%
-              t() %>%
-              as.data.frame() %>%
-              rownames_to_column(),
-            by = "rowname")
-
-results_swn <- rbind(main_effects,
-                     random_component,
-                     within_group_se)
-
-colnames(results_swn) <- c("Variable", "l1", "b1", "u1", "l2", "b2", "u2", "l3", "b3", "u3")
-
-results_swn <- results_swn %>%
-  transmute(Variable,
-            b1 = ifelse(is.na(b1),
-                        " ", 
-                        round(b1, 3)),
-            ci1 = ifelse(is.na(l1), 
-                         " ",
-                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")),
-            b2 = ifelse(is.na(b2),
-                        " ", 
-                        round(b2, 3)),
-            ci2 = ifelse(is.na(l2), 
-                         " ",
-                         paste0("[", round(l2, 3), ", ", round(u2, 3), "]")),
-            b3 = ifelse(is.na(b3),
-                        " ", 
-                        round(b3, 3)),
-            ci3 = ifelse(is.na(l3), 
-                         " ",
-                         paste0("[", round(l3, 3), ", ", round(u3, 3), "]")))
-```
-
-```{r tabulate-swn-results, echo=FALSE}
-results_swn %>%
-  mutate(Variable = c("Intercept", 
-                      "Log of population density",
-                      "Percent of private transportation", 
-                      "Median household income ($10,000)",
-                      "Standard deviation (Intercept)",
-                      "Within-group standard error")) %>%
-  kable("html",
-        digits = 3,
-        booktabs = TRUE,
-        col.names = c("Variable",
-                      "beta",
-                      "95% CI",
-                      "beta",
-                      "95% CI",
-                      "beta",
-                      "95% CI"),
-        caption = "\\label{tab:swn-results}Reproducing SWN: Models 1-3") %>%
-  kable_styling(latex_options = c("scale_down")) %>%
-  add_header_above(c(" ", "Model 1" = 2, "Model 2" = 2, "Model 3" = 2))
-```
-
-```{r random-terms-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
-# Choropleth map of basic reproductive number
-ggplot() + 
-  geom_sf(data = state_geo %>%
-            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
-          aes(fill = random_intercepts), # Add the simple features object to the plot
-          color = "lightgray") +
-  # Add boundary of continental US
-  geom_sf(data = us_geo,
-          fill = NA) + 
-  # Use viridis color palette: scico color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
-  scale_fill_scico(name = "Random Intercepts",
-                   direction = -1,
-                   palette = "vikO") +
-  # Set theme of plot
-  theme_minimal() +
-  theme(legend.position = "bottom")
-```
-
-# Expanding on SWN
-
-The preceding section shows that thanks to the availability of code and data, it is possible to verify the results reported by SWN.
-As noted earlier, though, an independent researcher might have wondered about the implications of the spatial sampling procedure used by SWN.
-The decision to use a sample of counties with reliable basic reproductive numbers, although apparently sensible, results in a non-random spatial sampling scheme.
-Turning our attention back to Figure \ref{fig:R0-map}, we form the impression that many counties without reliable values of $R_0$ are in more rural, less dense parts of the United States.
-This impression is reinforced when we overlay the boundaries of urban areas with population greater than 50,000 on the counties with valid values of $R_0$ (see Figure \ref{fig:urban-areas-map}).
-The fact that $R_0$ could not be accurately computed in many counties without large urban areas does not mean that there was no transmission of the virus: it simply means that we do not know with sufficient precision to what extent that was the case.
-The low number of cases may be related to low population and/or low population density.
-This is intriguing, to say the least: by excluding cases based on the ability to calculate $R_0$ we are potentially _selecting_ the sample in a non-random way.
-
-```{r urban-areas-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
-# Choropleth map of basic reproductive number
-ggplot() + # Create a ggplot
-  # Choropleth map of basic reproductive number
-  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
-            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
-          aes(fill = R),
-          color = NA) +
-  # Use viridis color palette: scico color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
-  scale_fill_scico(name = "R0",
-                   palette = "lajolla") +
-  # Overlay counties with missing values of the reproductive number in white
-  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
-            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
-                   R == 0), 
-          fill = "white",
-          color = NA) +
-  # Overlay boundaries of urban areas
-  geom_sf(data = urban_geo %>% # Add the simple features object to the plot
-            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
-                   UATYP10 == "U"), # Include only urbanized areas with population > 50,000
-          fill = NA,
-          color = "blue",
-          size = 0.05) +
-  # Add boundary of continental US
-  geom_sf(data = us_geo,
-          fill = NA) + 
-  # Add caption
-  labs(caption = "Note: boundaries of urbanized areas with population > 50,000 are shown in blue") +
-  # Set theme of plot
-  theme_minimal() +
-  theme(legend.position = "bottom")
-```
-
-A problematic issue with non-random sample selection is that parameter estimates can become unreliable, and numerous techniques have been developed to address this.
-A model useful for sample selection problems is Heckman's selection model [see @Maddala1983limited].
-The selection model is in fact a system of two equations, as follows:
-$$
-\begin{array}{c}
-y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
-y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
-\end{array}
-$$
-\noindent where $y_i^{S*}$ is a latent variable for the sample selection process, and $y_i^{O*}$ is the latent outcome.
-Vectors $x_i^S$ and $x_i^O$ are explanatory variables (with the possibility that $x_i^S = x_i^S$).
-Both equations include random terms (i.e., $\epsilon_i^S$ and $\epsilon_i^O$).
-The first equation is designed to model the _probability_ of sampling, and the second equation the outcome of interest (say $R_0$).
-The random terms are jointly distributed and correlated with parameter $\rho$.
-
-What the analyst observes is the following:
-$$
-y_i^S =
-\begin{cases}
-0 & \text{if } y_i^{S*} < 0\\
-1 & \text{otherwise}
-\end{cases}
-$$
-\noindent and:
-$$
-y_i^O =
-\begin{cases}
-0 & \text{if } y_i^{S} = 0\\
-y_i^{O*} & \text{otherwise}
-\end{cases}
-$$
-
-In other words, the outcome of interest is observed _only_ for certain cases ($y_i^S=1$, i.e., for sampled observations).
-The probability of sampling depends on $x_i^S$.
-For the cases observed, the outcome $y_i^O$ depends on $x_i^O$.
-
-A sample selection model is estimated using the same selection of variables as SWN Model 3.
-This is Sample Selection Model 1 in Table \ref{tab:selection-results}.
-The first thing to notice about this model is that the sample selection process and the outcome are correlated ($\rho\ne0$ with 5% of confidence).
-The selection equation indicates that the probability of a county to be in the sample increases with population density (but at a decreasing rate due to the log-transformation), when travel by private modes is more prevalent, and as median household income in the county is higher.
-This is in line with the impression made by Figure \ref{fig:urban-areas-map} that counties with reliable values of $R_0$ tended to be those with larger urban centers.
-Once that the selection probabilities are accounted for in the model, several things happen with the outcomes model.
-First, the coefficient for population density is still positive, but the magnitude changes: in effect, it appears that the effect of density is more pronounced than what SWN Model 3 indicated.
-The coefficient for percent of private transportation changes signs.
-And the coefficient for median household income is now significant.
-
-The second model in Table \ref{tab:selection-results} (Selection Model 2) changes the way the variables are entered into the model.
-The log-transformation of density in SWN and Selection Model 1 assumes that the association between density and $R_0$ is monotonically increasing (if the sign of the coefficient is positive) or decreasing (if the sign of the coefficient is negative).
-There are some indications that the relationship may actually not be monotonical.
-For example, Paez et al.
-[-@Paez2020spatio] found a positive (if non-significant) relationship between density and incidence of COVID-19 in the provinces of Spain at the beginning of the pandemic.
-This changed to a negative (and significant) relationship during the lockdown.
-In the case of the US, Fielding-Miller et al.
-[-@Fielding2020social] found that the association between COVID-19 deaths and population density was positive in rural counties, but negative in urban counties.
-A variable transformation that allows for non-monotonic changes in the relationship is the square of the density.
-
-As seen in the table, Selection Model 2 replaces the log-transformation of population density with a quadratic expansion.
-The results of this analysis indicate that with this variable transformation, the selection and outcome processes are still correlated ($\rho\ne0$ with 5% of confidence).
-But a few other interesting things emerge.
-When we examine the outcomes model, we see that the quadratic expansion has a positive coefficient for the first order term, but a negative coefficient for the second order term.
-This indicates that $R_0$ initially tends to increase as density grows, but only up to a point, after which the negative second term (which grows more rapidly due to the square), becomes increasingly dominant.
-Secondly, the sign of the coefficient for travel by private transportation becomes negative again.
-This, of course, makes more sense than the positive sign of Selection Model 1: if people tend to travel in private transportation, the potential for contact should be lower instead of higher.
-And finally median household income is no longer significant, similar to SWN Model 3.
-
-```{r selection-variable, include=FALSE}
-# Create selection variable
-county_geo_clean <- county_geo_clean %>%
-  mutate(Ys = ifelse(R>0, TRUE, FALSE))
-```
-
-```{r estimate-selection-models, include=FALSE}
-# Selection model using SWN variables
-selection1 <- selection(Ys ~ density_log + private + hincome, R ~ density_log + private + hincome, data = county_geo_clean)
-#summary(selection1)
-
-# Selection model with different variable specification
-selection2 <- selection(Ys ~ density + density_2 + private + hincome, R ~ density + density_2 + private + hincome, data = county_geo_clean)
-#summary(selection2)
-
-# Checking SWN with different variable specification
-model3.2 <- lme(R ~ density + density_2 + private + hincome, 
-                random = ~ 1| state, 
-                data = county_geo_clean %>%
-                  filter(R > 0))
-#summary(model3.2)
-```
-
-```{r prepare-selection-results, include=FALSE}
-selection1_me <- selection1 %>%
-  summary()
-
-selection1_me <- selection1_me$estimate %>%
-  as.data.frame()
-
-colnames(selection1_me) <- c("b", "se", "t", "p")
-
-selection1_me <- selection1_me %>%  
-  rownames_to_column(var = "Variable") %>% 
-  transmute(Variable,
-            b,
-            ll = b - 1.96 * se, 
-            ul = b + 1.96 * se)
-
-selection2_me <- selection2 %>%
-  summary()
-
-selection2_me <- selection2_me$estimate %>%
-  as.data.frame()
-
-colnames(selection2_me) <- c("b", "se", "t", "p")
-
-selection2_me <- selection2_me %>%  
-  rownames_to_column(var = "Variable") %>% 
-  transmute(Variable,
-            b,
-            ll = b - 1.96 * se, 
-            ul = b + 1.96 * se)
-
-results_selection <- selection1_me %>%
-  full_join(selection2_me,
-            by = "Variable")
-
-colnames(results_selection) <- c("Variable",
-                                 "b1",
-                                 "l1",
-                                 "u1",
-                                 "b2",
-                                 "l2",
-                                 "u2")
-
-results_selection <- results_selection %>%
-  transmute(Order = c(1, 2, 5, 6, 7, 8, 11, 12, 13, 14, 3, 4, 9, 10),
-            Variable,
-            b1 = ifelse(is.na(b1),
-                        " ", 
-                        round(b1, 3)),
-            ci1 = ifelse(is.na(l1), 
-                         " ",
-                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")),
-            b2 = ifelse(is.na(b2),
-                        " ", 
-                        round(b2, 3)),
-            ci2 = ifelse(is.na(l2), 
-                         " ",
-                         paste0("[", round(l2, 3), ", ", round(u2, 3), "]"))) %>%
-  arrange(Order) %>%
-  select(-Order)
-```
-
-```{r tabulate-sample-selection-results, echo=FALSE}
-results_selection  %>%
-  mutate(Variable = c("Intercept",
-                      "Log of population density",
-                      "Density (1,000 per sq.km)",
-                      "Density squared",
-                      "Percent of private transportation", 
-                      "Median household income (10,000)",
-                      "Intercept",
-                      "Log of population density",
-                      "Density (1,000 per sq.km)",
-                      "Density squared",
-                      "Percent of private transportation", 
-                      "Median household income (\\$10,000)",
-                      "$\\sigma$",
-                      "$\\rho$")) %>%
-  kable("html",
-        digits = 3,
-        booktabs = TRUE,
-        escape = FALSE,
-        col.names = c("Variable",
-                      "$\\beta$", 
-                      "95\\% CI",
-                      "$\\beta$",
-                      "95\\% CI"),
-        align = c("lcccc"),
-        caption = "\\label{tab:selection-results}Estimation results of sample selection models") %>%
-  kable_styling(latex_options = c("scale_down")) %>% 
-  pack_rows(group_label = "Sample Selection Model", 1, 6) %>%
-  pack_rows(group_label = "Outcome Model", 7, 12) %>%
-  add_header_above(c(" ", "Selection Model 1" = 2, "Selection Model 2" = 2))
-```
-
-<!--
-If x is the density:
-$$
-\frac{dR}{dx} = 0.758 - 2 * 0.132 * density 
-$$
-
-Then, the maximum of this curve is when the density is:
-$$
-density = \frac{0.758}{2 * 0.132}
-$$
--->
-
-# Proceed with caution: spatial effects ahead
-
-The results of the selection models, in particular Selection Model 2, make us reassess the original conclusion that density has a positive association with the basic reproductive number of COVID-19.
-A spatial analyst might still wonder about spatial residual autocorrelation.
-A challenge here is that spatial models tend to be technically more demanding, and although spatial models for qualitative variables exist, a spatial implementation of the sample selection model does not appear to exist.
-It might be argued that a reproducible research project can also allow a researcher to be more adventurous with their modeling decisions: since data and code are shared, other researchers can promptly and with relative ease poke the methods and see if they appear to be sound.
-
-In the present case, it appears that an application of spatial filtering [see @Getis2002comparative; @Paez2019using; @Griffith2004spatial] can help.
-Spatial filtering provides an elegant solution to regression problems that may have difficulties handling the spatial structures of spatial statistical and econometric models [@Griffith2000linear].
-A key issue in the present example is the fact that there are numerous missing observations, which prevents the calculation of autocorrelation statistics, let alone the estimation of models with spatial components.
-
-The following is an unorthodox, but potentially effective use of filters in a sample selection model:
-
-1. Estimate a sample selection model and retrieve the residuals of the outcome. This will be a vector with missing values for locations that were not sampled.
-
-2. Fit a spatial filter to the residuals. This is done by regressing the estimated residuals of the _observed_ data on the corresponding values of the Moran eigenvectors.
-
-3. The resulting filter will correlate highly with the known residuals, and will provide information in non-sampled locations that is consistent with the spatial pattern of the known residuals.
-
-4. Test the filter for spatial autocorrelation:
-
-    4.1 If significant spatial autocorrelation is detected, this would be indicative of residual spatial pattern.
-Introduce the filter as a covariate in the outcome model of the sample selection model and return to step 1.
-
-    4.2 If no significant spatial autocorrelation is detected, this would be indicative of random residual pattern. Stop.
-
-This procedure is implemented using a stopping criterion whereby the search for the filter only stops when the p-value of Moran's Coefficient of the filter fitted to the residuals is greater than 0.25, which was chosen as a sufficiently conservative value for testing for autocorrelation.
-The correlation of the known residuals with the corresponding elements of the filter is consistently high (the correlation coefficient typically is greater than 0.9).
-The results of implementing this procedure appear in Table \ref{tab:selection3-results} as Selection Model 3.
-The results are consistent with Selection Model 2, with two intriguing differences: 1) the variance of Sample Model 3 is smaller; and 2) the sample and outcome processes are no longer correlated (the confidence interval of $\rho$ includes zero).
-It appears that by capturing the spatial pattern of the residuals, which is likely strongly determined by the non-random sampling framework, the outcome model is not only substantially more precise, but also appears to be independent from the selection process.
-
-```{r selection-model-with-spatial-filter, include=FALSE}
-
-# Refit model selection2 to begin 
-selection3 <- selection(Ys ~ density + density_2 + private + hincome, R ~ density + density_2 + private + hincome, data = county_geo_clean)
-
-# Extract the residuals of the model. These residuals have NAs where the outcome was not observed
-res_selection3 <- residuals(selection3)
-
-#Initialize a spatial filter:
-SF <- numeric(length = length(res_selection3))
-
-#Initialize tolerance, counter, and index:
-tol <- 0;
-
-while(tol < 0.25){
-  #Initialize counter and index:
-  count <- 0;
-  SF_INDEX <- 0
-
-  #Obtain filter to simulate the residuals
-  for(i in 1:length(county_mem)){
-    count <- count + 1
-    junkSF <- data.frame(SF = SF, V = county_mem[, count])
-    #remove columns with all zeros
-    junkSF <- Filter(function(x)!all(x == 0), junkSF)
-    #estimate model
-    junkmod <- lm(res_selection3 ~ .,
-                  data = cbind(res_selection3, junkSF))
-    junkmod_summary <- summary(junkmod) # Summary of model; needed to extract the pval of the filter
-    pvals<- junkmod_summary$coefficients[nrow(junkmod_summary$coefficients), 4]
-    #pvals <- get_pvalue(junkmod)
-    junkb <- coef(junkmod)
-    if(pvals <= 0.10){
-      SF_INDEX[count] <- count
-      SF <- as.matrix(junkSF) %*% junkb[(0+2):length(junkb)]
-      junkmod <- lm(res_selection3 ~ ., 
-                    data = data.frame(res_selection3, SF))
-    }
-  }
-  #SF_INDEX <- SF_INDEX[!is.na(SF_INDEX)]
-  Filter1 <- SF # Spatial filter for residuals
-  res_filter_cor <- cor(res_selection3, Filter1, use = "complete.obs")
-  Filter_mt <- moran.test(Filter1, county_listw)
-  tol <- Filter_mt$p.value
-  if(tol < 0.25){
-    selection3 <- selection(Ys ~ density + density_2 + private + hincome, 
-                            R ~ density + density_2 + private + hincome + Filter1, 
-                            data = county_geo_clean)
-  }  
-}
-```
-
-```{r prepare-selection3-results, include=FALSE}
-selection3_me <- selection3 %>%
-  summary()
-
-selection3_me <- selection3_me$estimate %>%
-  as.data.frame()
-
-colnames(selection3_me) <- c("b", "se", "t", "p")
-
-selection3_me <- selection3_me %>%  
-  rownames_to_column(var = "Variable") %>% 
-  transmute(Variable,
-            b,
-            ll = b - 1.96 * se, 
-            ul = b + 1.96 * se)
-
-results_selection <- selection1_me %>%
-  full_join(selection2_me,
-            by = "Variable")
-
-colnames(selection3_me) <- c("Variable",
-                                 "b1",
-                                 "l1",
-                                 "u1")
-
-selection3_me <- selection3_me %>%
-  transmute(Variable,
-            b1 = ifelse(is.na(b1),
-                        " ", 
-                        round(b1, 3)),
-            ci1 = ifelse(is.na(l1), 
-                         " ",
-                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")))
-#%>%
-#  arrange(Order) %>%
-#  select(-Order)
-```
-
-```{r tabulate-sample-selection3-results, echo=FALSE}
-selection3_me  %>%
-  mutate(Variable = c("Intercept",
-                      "Density (1,000 per sq.km)",
-                      "Density squared",
-                      "Percent of private transportation", 
-                      "Median household income (10,000)",
-                      "Intercept",
-                      "Density (1,000 per sq.km)",
-                      "Density squared",
-                      "Percent of private transportation", 
-                      "Median household income (\\$10,000)",
-                      "Spatial filter",
-                      "$\\sigma$",
-                      "$\\rho$")) %>%
-  kable("html",
-        digits = 3,
-        booktabs = TRUE,
-        escape = FALSE,
-        col.names = c("Variable",
-                      "$\\beta$", 
-                      "95\\% CI"),
-        align = c("lcccc"),
-        caption = "\\label{tab:selection3-results}Estimation results of sample selection model with spatial filter") %>%
-  kable_styling(font_size = 8, latex_options = "hold_position") %>% 
-  pack_rows(group_label = "Sample Selection Model", 1, 5) %>%
-  pack_rows(group_label = "Outcome Model", 6, 11) %>%
-  add_header_above(c(" ", "Selection Model 3" = 2))
-```
-
-Clearly, the various models display some intriguing differences; but how relevant are said differences from a more substantive standpoint?
-Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 3, Selection Model 2, and Selection Model 3.
-The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1.
-The conclusion is that at higher densities, $R_0$ is _always_ higher.
-The two panels on the right, in contrast, shows that Selection Model 2 and Selection Model 3 coincide that $R_0$ tends to increase as density grows.
-This continues until a density of approximately 2.9 (1,000 people per sq.km).
-At higher densities than that the relationship between density and $R_0$ begins to weaken, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).
-
-To put this into context, other things being equal, the effect of density in a county like Charlottesville in Virginia (density ~1,639 people per sq.km) is roughly the same as that in a county like Philadelphia (density ~4,127 people per sq.km).
-In contrast, the effect of density on $R_0$ in a county like Arlington in Virginia (density ~3,093 people per sq.km) is _stronger_ than either of the previous two examples.
-Lastly, the density of counties like San Francisco in California, or Queens and Bronx in NY, which are among the densest in the US, contributes even less to $R_0$ than even the most rural counties in the country.
-
-```{r comparison-results, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2."}
-# The maximum in-sample density is 6.255358
-# county_geo_clean %>% st_drop_geometry() %>% filter(R > 0) %>% summary()
-
-example_df <- data.frame(density = seq(0, 6.3, 0.01)) %>% 
-  mutate(density_log = log(density),
-         density_2 = density^2, 
-         b_swn = model3$coefficients$fixed["density_log"] * density_log,
-         b_selection = selection2$estimate[7] * density + selection2$estimate[8] * density_2,
-         b_selection3 = selection3$estimate[7] * density + selection3$estimate[8] * density_2)
-
-# Plot of coefficient of density to R_0 according to SWN Model 3
-density_swn <- ggplot(data = example_df) + 
-  geom_line(aes(x = density, 
-                y = b_swn)) +
-  ggtitle("SWN Model 3") +
-  ylab(expression(paste("E[", R[0], "]"))) +
-  theme_minimal()
-
-# Plot of coefficient of density to R_0  
-density_selection <- ggplot(data = example_df) + 
-  geom_line(aes(x = density, 
-                y = b_selection)) +
-  ggtitle("Selection Model 2") +
-  ylab(expression(paste("E[", R[0], "]"))) +
-  theme_minimal()
-
-# Plot of coefficient of density to R_0  
-density_selection3 <- ggplot(data = example_df) + 
-  geom_line(aes(x = density, 
-                y = b_selection3)) +
-  ggtitle("Selection Model 3") +
-  ylab(expression(paste("E[", R[0], "]"))) +
-  theme_minimal()
-
-density_swn + density_selection + density_selection3
-```
-
-<!--
-
-## Experiments with tobit and spatially autoregressive tobit
-
-## Fit tobit version of models
-
-```{r include=FALSE}
-#
-tobit1 <- censReg(R ~ density_log, 
-                  left = 0,
-                  data = county_geo_clean)
-
-#
-tobit2 <- censReg(R ~ density_log + private, 
-                  data = county_geo_clean)
-
-#
-tobit3 <- censReg(R ~ density_log + private + hincome, 
-                  data = county_geo_clean)
-```
-
-## Spatially autoregressive tobit
-
-Fit spatially autoregressive tobit:
-```{r include=FALSE}
-# Fit SAR Tobit
-fit_sartobit <- sartobit(R ~ density_log + private + hincome_log,
-                         B,
-                         ndraw = 1000,
-                         burn.in = 200, 
-                         showProgress = TRUE,
-                         data = county_geo_clean,
-                         computeMarginalEffects = TRUE)
-summary(fit_sartobit)
-impacts(fit_sartobit)
-```
-
--->
-
-# Discussion
-
-It is worth at this point to recall Cressie's dictum about modelling: "[w]hat is one person's mean structure could be another person's correlation structure" [@Cressie1989geostatistics, p.
-201].
-There are almost always multiple ways to approach a modelling situation, as lively illustrated by a recent paper that reports the results of a crowdsourced modelling experiment [@Schweinsberg2021same].
-In the present case, we would argue that spatial sampling is an important aspect of the modeling process.
-Importantly, by adopting high reproducibility standards, SWN made a valuable contribution to the collective enterprise of seeking knowledge.
-Their effort, and subsequent efforts to validate and expand on their work, can potentially contribute to provide clarity to ongoing conversations about the relevance of density and the spread of COVID-19.
-
-In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [in Wong and Li's words, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading].
-At the same time, the results presented here also stand in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with it in a longitudinal setting [@Hamidi2020longitudinal.
-In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al.
-[-@Fielding2020social] might be onto something with respect to the difference between rural and urban counties.
-More generally, there is no doubt that in population-level studies density is indicative of proximity, but it also potentially is a proxy for adaptive behavior.
-And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.
-
-```{r counties-by-density, include=FALSE}
-# the relationship between density and R_0 become negative in Selection Model 2 at about 5.74
-
-# Values of b_selection for a density similar a Philadelphia 
-example_df %>% filter(density > 4.10 & density < 4.14)
-# Densities with a similar b_selection
-example_df %>% filter(b_selection > 0.87 & b_selection < 0.89)
-
-# Counties with densities with similar response to Philadelphia
-county_geo_clean %>% 
-  mutate(density = drop_units(density)) %>%
-  filter(density > 1.59 & density < 1.64) %>%
-  select(NAME.y, density)
-
-# Counties with densities associated with the highest response to density
-county_geo_clean %>% 
-  mutate(density = drop_units(density)) %>%
-  filter(density > 2.70 & density < 3.10) %>%
-  select(NAME.y, density)
-
-# Highly dense counties
-county_geo_clean %>% 
-  mutate(density = drop_units(density)) %>%
-  filter(density > 5.74) %>%
-  select(NAME.y)
-
-```
-
-# Conclusion
-
-The tension between the need to publish research potentially useful in dealing with a global pandemic, and a potential "carnage of substandard research" [@Bramstedt2020carnage], highlights the importance of efforts to maintain the quality of scientific outputs during COVID-19.
-An important part of quality control is the ability of independent researchers to verify and examine the results of materials published in the literature.
-As previous research illustrates, reproducibility in scientific research remains an important but elusive goal [e.g., @Iqbal2016reproducible; @Stodden2018empirical; @Sumner2020reproducibility; @Gustot2020quality].
-This idea is reinforced by the review conducted for this paper in the context of research about population density and the spread of COVID-19.
-
-Taking one recent example from the literature [Sy et al., @Sy2021population; SWN], the present paper illustrates the importance of good reproducibility practices.
-Sharing data and code can catalyze research, by allowing independent verification of findings, as well as additional research.
-After verifying the results of SWN, experiments with sample selection models and variations in the definition of model inputs, lead to an important reappraisal of the conclusion that high density is associated with greater spread of the virus.
-Instead, the possibility of a non-monotonical relationship between population density and contagion is raised.
-I do not claim that the analysis presented here is the last word on the topic of density and the spread of COVID-19, and there is always the possibility that someone else will be better equipped to analyze these data with greater competence.
-By opening up the analysis, documenting the way data were pre-processed, and by sharing analysis ready data, my hope would be that others will be able to discover the limitations of my own analysis and improve on it, as appropriate.
-
-More generally, my hope is that the research of Sy et al.
-[-@Sy2021population], the present paper, and similar reproducible publications, will continue to encourage others to adopt higher reproducibility standards in their research.
-
-# References {#references .unnumbered}
+journal: "Landscape and Urban Planning"
+date: "`r Sys.Date()`"
+bibliography: bibliography.bib
+linenumbers: true
+#numbersections: true
+csl: elsevier-harvard.csl
+output: rticles::elsevier_article
+header-includes:
+    - \usepackage{setspace}\doublespacing
+---

### 2021-11-18 R0-Density-Reanalysis/Editorial-Materials-and-Correspondence/R0-Density-Reanalysis-Title-Page.Rmd
Reorganized the repository after paper's publication online at Geographical Analysis

--- 2021-11-18 R0-Density-Reanalysis/Editorial-Materials-and-Correspondence/R0-Density-Reanalysis-Abstract.Rmd
@@
 ---
-title: "Population density and the spread of the COVID-19 pandemic: a reproducible research example"
+title: "Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis"
+author:
+  - name: Antonio Paez
+    email: paezha@mcmaster.ca
+    affiliation: McMaster University
+    footnote: 1
+address:
+  - code: McMaster University
+    address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
+footnote:
+  - code: 1
+    text: "Corresponding Author"
 abstract: |

### 2021-11-18 R0-Density-Reanalysis/Editorial-Materials-and-Correspondence/R0-Density-Reanalysis-Title-Page.tex
Reorganized the repository after paper's publication online at Geographical Analysis

--- 2021-06-10 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
@@
 % Pandoc citation processing
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc} % for calculating minipage widths
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \usepackage{setspace}\onehalfspacing
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 
@@
 perspective of spatial analysis}
-    \author[University]{Author\corref{1}}
-   \ead{author@institution.edu} 
-      \address[University]{Department, Street, City, State ZIP}
+    \author[McMaster University]{Antonio Paez\corref{1}}
+   \ead{paezha@mcmaster.ca} 
+      \address[McMaster University]{School of Earth, Environment and
+Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada}
       \cortext[1]{Corresponding Author}
@@
 
-\newpage
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the
-global pandemic that followed in its wake, led to an explosive growth of
-research around the globe. According to Fraser et al. (2021), over
-125,000 COVID-19-related papers were released in the first ten months
-from the first confirmed case of the disease. Of these, more than 30,000
-were shared in pre-print servers, the use of which also exploded in the
-past year (Añazco et al., 2021; Kwon, 2020; Vlasschaert et al., 2020).
-
-Given the ruinous human and economic cost of the pandemic, there has
-been a natural tension in the scientific community between the need to
-publish research results quickly and the imperative to maintain
-consistently high quality standards in scientific reporting; indeed, a
-call for maintaining the standards in published research has termed the
-deluge of COVID-19 publications a ``carnage of substandard research''
-(Bramstedt, 2020). Part of the challenge of maintaining quality
-standards in published research is that, despite an abundance of
-recommendations and guidelines (e.g., Broggini et al., 2017; Brunsdon
-and Comber, 2020; Ince et al., 2012; Ioannidis et al., 2014), in
-practice reproducibility has remained a lofty and somewhat aspirational
-goal (Konkol et al., 2019; Konkol and Kray, 2019). As reported in the
-literature, only a woefully small proportion of published research was
-actually reproducible before the pandemic (Iqbal et al., 2016; Stodden
-et al., 2018), and the situation does not appear to have changed
-substantially since (Gustot, 2020; Sumner et al., 2020).
-
-The push for open data and software, along with more strenuous efforts
-towards open, reproducible research, is simply a continuation of
-long-standing scientific practices of independent verification. Despite
-the (at times disproportionate) attention that high profile scandals in
-science tend to elicit in the media, science as a collective endeavor is
-remarkable for being a self-correcting enterprise, one with built-in
-mechanisms and incentives to weed out erroneous ideas. Over the long
-term, facts tend to prevail in science. At stake is the shorter-term
-impacts that research may have in other spheres of economic and social
-life. The case of economists Reinhart and Rogoff comes to mind: by the
-time the inaccuracies and errors in their research were uncovered (see
-Herndon et al., 2014), their claims about debt and economic growth had
-already been seized by policy-makers on both sides of the Atlantic to
-justify austerity policies in the aftermath of the Great Recession of
-2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
-  ``Reinhart--Rogoff may have had more immediate influence on public
-  debate than any previous paper in the history of economics''
-  \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
-As later research has demonstrated, those policies cast a long shadow,
-and their sequels continued to be felt for years (Basu et al., 2017).
-
-In the context of COVID-19, a topic that has grabbed the imagination of
-numerous thinkers has been the prospect of life in cities after the
-pandemic (Florida et al., 2020); the implications of the pandemic for
-urban planning, design, and management are the topic of ongoing research
-(Sharifi and Khavarian-Garmsir, 2020). The fact that the worst of the
-pandemic was initially felt in dense population centers such as Wuhan,
-Milan, Madrid, and New York, unleashed a torrent of research into the
-associations between density and the spread of the pandemic. he answers
-to some important questions hang on the results of these research
-efforts. For example, are lower density regions safer from the pandemic?
-Are de-densification policies warranted, even if just in the short term?
-And in the longer term, will the risks of life in high density regions
-presage a flight from cities? What are the implications of the pandemic
-for future urban planning and practice? Over the past year, numerous
-papers have sought to throw light on the underlying issue of density and
-the pandemic; nonetheless the results, as will be detailed next, remain
-mixed. Further, to complicate matters, precious few of these studies
-appear to be sufficiently open to support independent verification.
-
-The objective of this paper is to illustrate the importance of
-reproducibility in research in the context of the flood of COVID-19
-papers. To this end, a recent study by Sy et al. (2021) is chosen as an
-example of reproducible research. The objective is not to malign the
-analysis of these researchers, but rather to demonstrate the value of
-openness to allow for independent verification and further analysis.
-Open data and open code mean that an independent researcher can, with
-only modest efforts, not only verify the findings reported, but also
-examine the same data from a perspective which may not have been
-available to the original researchers due to differences in disciplinary
-perspectives, methodological traditions, and/or training, among other
-possible factors. The example, which shows consequential changes in the
-conclusions reached by different analyses, should serve as a call to
-researchers to redouble their efforts to increase transparency and
-reproducibility in research. The present paper, in addition, aims to
-show how data can be packaged in well-documented, shareable units, and
-code can be embedded into self-contained documents suitable for review
-and independent verification. The source for this paper is an
-\href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
-with the data package, will be available in a public
-repository\footnote{For peer-review purposes, the data package and code
-  are currently in an anonymous Drive folder:
-  \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}}.
-
-\hypertarget{background-the-intuitive-relationship-between-density-and-spread-of-contagious-diseases}{%
-\section{Background: the intuitive relationship between density and
-spread of contagious
-diseases}\label{background-the-intuitive-relationship-between-density-and-spread-of-contagious-diseases}}
-
-The concern with population density and the spread of the virus during
-the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
-seen in real-time around the world from large urban centers such as
-Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
-believe that higher density may have a positive association with the
-transmission of a contagious virus. It has long been known that the
-potential for inter-personal contact is greater in regions with higher
-density (see for example the research on urban fields and
-time-geography, including Farber and Páez, 2011; Moore, 1970; Moore and
-Brown, 1970). Mathematically, models of exposure and contagion indicate
-that higher densities can catalyze the transmission of contagious
-diseases (Li et al., 2018; Rocklöv and Sjödin, 2020). The idea is
-intuitive and likely at the root of messages, by some figures in
-positions of authority, that regions with sparse population densities
-faced lower risks from the pandemic\footnote{Governor Kristi Noem of
-  South Dakota, for example, claimed that sparse population density
-  allowed her state to face the pandemic down without the need for
-  strict policy interventions
-  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.
-
-As Rocklöv and Sjödin (Rocklöv and Sjödin, 2020) note, however,
-mathematical models of contagion are valid at small-to-medium spatial
-scales (and presumably, small temporal scales too, such as time spent in
-restaurants, concert halls, cruises), and the results do not necessarily
-transfer to larger spatial units and different time scales. There are
-solid reasons for this: while in a restaurant, one can hardly avoid
-being in proximity to other customers. On the other hand, a person can
-choose to (or be forced to as a matter of policy) not go to a restaurant
-in the first place. Nonetheless, the idea that high density correlates
-with high transmission is so seemingly sensible that it is often taken
-for granted even at larger scales (e.g., Cruz et al., 2020; Micallef et
-al., 2020). At larger scales, however, there exists the possibility of
-behavioral adaptations, which are difficult to capture in the
-mechanistic framework of differential equations (or can be missing in
-agent-based models, e.g., Gomez et al., 2021); these adaptations, in
-fact, can be a key aspect of disease transmission.
-
-A plausible behavioral adaptation during a pandemic, especially one
-broadcast as widely and intensely as COVID-19, is risk compensation.
-Risk compensation is a process whereby people adjust their behavior in
-response to their \emph{perception} of risk (Noland, 1995; Phillips et
-al., 2011; Richens et al., 2000). In the case of COVID-19, Chauhan et
-al. (Chauhan et al., 2021) have found that perception of risks in the US
-varies between rural, suburban, and urban residents, with rural
-residents in general expressing less concern about the virus. It is
-possible that people who listened to the message of leaders saying that
-they were safe from the virus because of low density may not have taken
-adequate precautions. Conversely, people in dense places who could more
-directly observe the impact of the pandemic may have become overly
-cautious. Both Paez et al. (2020) and Hamidi et al. (2020b) posit this
-mechanism (i.e., greater compliance with social distancing in denser
-regions) to explain the results of their analyses. The evidence
-available does indeed show that there were important changes in behavior
-with respect to mobility during the pandemic (Harris and Branion-Calles,
-2021; Jamal and Paez, 2020; Molloy et al., 2020); furthermore, shelter
-in place orders may have had greater buy-in from the public in higher
-density regions (Feyman et al., 2020; Hamidi and Zandiatashbar, 2021),
-and the associated behavior may have persisted beyond the duration of
-official social-distancing policies (Praharaj et al., 2020). In
-addition, there is evidence that changes in mobility correlated with the
-trajectory of the pandemic (Noland, 2021; Paez, 2020). Given the
-potential for behavioral adaptation, the question of density becomes
-more nuanced: it is not just a matter of proximity, but also of human
-behavior, which is better studied using population-level data and
-models.
-
-\hypertarget{background-but-what-does-the-literature-say}{%
-\section{Background: but what does the literature
-say?}\label{background-but-what-does-the-literature-say}}
-
-When it comes to population density and the spread of COVID-19, the
-international literature to date remains inconclusive.
-
-On the one hand, there are studies that report positive associations
-between population density and various COVID-19-related outcomes. Bhadra
-(2021), for example, reported a moderate positive correlation between
-the spread of COVID-19 and population density at the district level in
-India, however their analysis was bivariate and did not control for
-other variables, such as income. Similarly, Kadi and Khelfaoui (2020)
-found a positive and significant correlation between number of cases and
-population density in cities in Algeria in a series of simple regression
-models (i.e., without other controls). A question in these relatively
-simple analyses is whether density is not a proxy for other factors.
-Other studies have included controls, such as Pequeno et al. (2020), a
-team that reported a positive association between density and cumulative
-counts of confirmed COVID-19 cases in state capitals in Brazil after
-controlling for covariates, including income, transport connectivity,
-and economic status. In a similar vein, Fielding-Miller et al. (2020)
-reported a positive relationship between the absolute number of COVID-19
-deaths and population density (rate) in rural counties in the US. Roy
-and Ghosh (2020) used a battery of machine learning techniques to find
-discriminatory factors, and a positive and significant association
-between COVID-19 infection and death rates in US states. Wong and Li
-(2020) also found a positive and significant association between
-population density and number of confirmed COVID-19 cases in US
-counties, using both univariate and multivariate regressions with
-spatial effects. More recently, Sy et al. (2021) reported that the basic
-reproductive number of COVID-19 in US counties tended to increase with
-population density, but at a decreasing rate at higher densities.
-
-On the flip side, a number of studies report non-significant or negative
-associations between population density and COVID-19 outcomes. This
-includes the research of Sun et al. (2020) who did not find evidence of
-significant correlation between population density and confirmed number
-of cases per day \emph{in conditions of lockdown} in China. This finding
-echoes the results of Paez et al. (2020), who in their study of
-provinces in Spain reported non-significant associations between
-population density and infection rates in the early days of the first
-wave of COVID-19, and negative significant associations in the later
-part of the first lockdown. Similarly, (2020) found zero or negative
-associations between population density and infection numbers/deaths by
-country. Fielding-Miller et al. (2020) contrast their finding about
-rural counties with a negative relationship between COVID-19 deaths and
-population density in urban counties in the US. For their part, in their
-investigation of doubling time, White and Hébert-Dufresne (2020)
-identified a negative and significant correlation between population
-density and doubling time in US states. Likewise, (2021) found a small
-negative (and significant) association between population density and
-COVID-19 morbidity in districts in Tehran. Finally, two of the most
-complete studies in the US (by Hamidi et al. 2020a, and 2020b) used an
-extensive set of controls to find negative and significant correlations
-between density and COVID-19 cases and fatalities at the level of
-counties in the US.
-
-As can be seen, these studies are implemented at different scales in
-different regions of the world. They also use a range of techniques,
-from correlation analysis, to multivariate regression, spatial
-regressions, and machine learning techniques. This is natural and to be
-expected: individual researchers have only limited time and expertise.
-This is why reproducibility is important. To pick an example (which will
-be further elaborated in later sections of this paper), the study of Sy
-et al.~{[}(2021); hereafter SWN{]} would immediately grab the attention
-of a researcher with expertise in spatial analysis.
-
-\hypertarget{reproducibility-of-research}{%
-\section{Reproducibility of
-research}\label{reproducibility-of-research}}
-
-SWN investigated the basic reproductive number of COVID-19 in US
-counties, and its association with population density, median household
-income, and prevalence of private mobility. For their multivariate
-analysis, SWN used mixed linear models. This is a reasonable modelling
-choice: \(R_0\) is an interval-ratio variable that is suitably modeled
-using linear regression; further, as SWN note there is a likelihood that
-the process in not independent ``among counties within each state,
-potentially due to variable resource allocation and differing health
-systems across states'' (p.~3). A mixed linear model accounts for this
-by introducing random components; in the case of SWN, these are random
-intercepts at the state level. SWN estimated various models with
-different combinations of variables, including median household income
-and prevalence of travel by private transportation. These are sensible
-controls, given potential variations in behavior: people in more
-affluent counties may have greater opportunities to work from home, and
-use of private transportation reduces contact with strangers. Moreover,
-they also conducted various sensitivity analyses. After these efforts,
-SWN concluded that there is a positive association between the basic
-reproductive number and population density at the level of counties in
-the US.
-
-One salient aspect of the analysis in SWN is that the basic reproductive
-number can only be calculated reliably with a minimum number of cases,
-and a large number of counties did not meet such threshold. As
-researchers do, SWN made modelling decisions, in this case basing their
-analysis only on counties with valid observations. A modeler with
-expertise in spatial analysis would likely ask some of the following
-questions on reading SWN's paper: how were missing counties treated?
-What are the implications of the spatial sampling framework used in the
-analysis? Is it possible to spatially interpolate the missing
-observations? Was there spatial residual autocorrelation in the models,
-or was the use of mixed models sufficient to capture spatial
-dependencies? These questions are relevant and their implications
-important. Fortunately, SWN are an example of a reasonably open,
-reproducible research product: their paper is accompanied by (most of)
-the data and (most of) the code used in the analysis. This means that an
-independent expert can, with only a moderate investment of time and
-effort, reproduce the results in the paper, as well as ask additional
-questions.
-
-Alas, reproducibility is not necessarily the norm in the relevant
-literature.
-
-There are various reasons why a project can fail to be reproducible. In
-some cases, there might be legitimate reasons to withhold the data,
-perhaps due to confidentiality and privacy reasons (e.g., Lee et al.,
-2020). But in many other cases the data are publicly available, which in
-fact has commonly been the case with population-level COVID-19
-information. Typically the provenance of the data is documented, but in
-numerous studies the data themselves are not shared (Amadu et al., 2021;
-Bhadra et al., 2021; Cruz et al., 2020; Feng et al., 2020;
-Fielding-Miller et al., 2020; Hamidi et al., 2020a, 2020b; Inbaraj et
-al., 2021; Souris and Gonzalez, 2020). As any researcher can attest,
-whether a graduate student or a seasoned scientist, collecting,
-organizing, and preparing data for a project can take a substantial
-amount of time. Pointing to the sources of data, even when these sources
-are public, is a small step towards reproducibility-but only a very
-small one. Faced with the prospect of having to recreate a data set from
-raw sources is probably sufficient to dissuade all but the most
-dedicated (or stubborn) researcher from independent verification. This
-is true even if part of the data are shared (e.g., Wong and Li, 2020).
-In other cases, data are shared, but the processes followed in the
-preparation of the data are not fully documented (Ahmad et al., 2020;
-Skórka et al., 2020). These processes matter, as shown by the errors in
-the spreadsheets of Reinhart and Rogoff (Herndon et al., 2014) and the
-data of biologist Jonathan Pruitt that led to an ``avalanche'' of paper
-retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
-Another situation is when papers share well-documented data, but fail to
-provide the code used in the analysis (Noury et al., 2021; Pequeno et
-al., 2020; Wang et al., 2021). Making code available only ``on demand''
-(e.g., Brandtner et al., 2021) is an unnecessary barrier when most
-journals offer the facility to share supplemental materials online. Then
-there are those papers that more closely comply with reproducibility
-standards, and share well-documented processes and data, as well as the
-code used in any analyses reported (Feyman et al., 2020; Paez et al.,
-2020; Stephens et al., 2021; Sy et al., 2021; White and Hébert-Dufresne,
-2020). Even in this case, the pressure to publish ``new findings''
-instead of replication studies can act as a deterrent, perhaps
-particularly for younger researchers\footnote{The present paper was desk
-  rejected by three journals that had previously published research on
-  population density and the spread of COVID-19; in one case, the paper
-  was too opinionated for the journal, in the other two cases, the paper
-  was not a ``good fit'' despite dealing with a nearly identical issue
-  as papers previously published in said journals. This does not inspire
-  much confidence in the commitment of journals to reproducibility in
-  research.}.
-
-In the following sections, the analysis of RWN is reproduced, some
-relevant questions from the perspective of an independent researcher
-with expertise in spatial analysis are asked, and the data are
-reanalyzed.
-
-\hypertarget{reproducing-swn}{%
-\section{Reproducing SWN}\label{reproducing-swn}}
-
-SWN examined the association between the basic reproductive number of
-COVID-19 and population density. The basic reproductive number \(R_0\)
-is a summary measure of contact rates, probability of transmission of a
-pathogen, and duration of infectiousness. In rough terms, \(R_0\)
-measures how many new infections each infections begets. Infectious
-disease outbreaks generally tend to die out when \(R_0<1\), and to grow
-when \(R_0>1\). Reliable calculation of \(R_0\) requires a minimum
-number of cases to be able to assume that there is community
-transmission of the pathogen. Accordingly, SWN based their analysis only
-on counties that had at least 25 cases or more at the end of the
-exponential growth phase (see Fig. \ref{fig:R0-map}). Their final sample
-included 1,151 counties in the US, including in Alaska, Hawaii, Puerto
-Rico, and island territories.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/R0-map-1} \caption{\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown).}\label{fig:R0-map}
-\end{figure}
-
-Table \ref{tab:swn-results} reproduces the first three models of SWN
-(the fourth model did not have any significant variables; see Table 1 in
-SWN). It is possible to verify that the results match, with only the
-minor (and irrelevant) exception of the magnitude of the coefficient for
-travel by private transportation, which is due to a difference in the
-input (here the variable is changed to one percent units, instead of the
-ten percent units used by SWN). The mixed linear model gives random
-intercepts (i.e., the intercept is a random variable), and the standard
-deviation is reported in the fourth row of Table \ref{tab:swn-results}.
-It is useful to map the random intercepts: as seen in Figure
-\ref{fig:random-terms-map}, other things being equal, counties in Texas
-tend to have somewhat lower values of \(R_0\) (i.e., a negative random
-intercept), whereas counties in South Dakota tend to have higher values
-of \(R_0\). The key of the analysis, after extensive sensitivity
-analysis, is a robust finding that population density has a positive
-association with the basic reproductive number. But does it?
-
-\begin{table}
-
-\caption{\label{tab:tabulate-swn-results}\label{tab:swn-results}Reproducing SWN: Models 1-3}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllllrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Model 1} & \multicolumn{2}{c}{Model 2} & \multicolumn{2}{c}{Model 3} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
-Variable & beta & 95\% CI & beta & 95\% CI & beta & 95\% CI\\
-\midrule
-Intercept & 2.274 & [2.167, 2.381] & 3.347 & [2.676, 4.018] & 3.386 & [2.614, 4.157]\\
-Log of population density & 0.162 & [0.133, 0.191] & 0.145 & [0.115, 0.176] & 0.147 & [0.113, 0.18]\\
-Percent of private transportation &  &  & -0.013 & [-0.02, -0.005] & -0.013 & [-0.021, -0.005]\\
-Median household income (\$10,000) &  &  &  &  & -0.003 & [-0.033, 0.026]\\
-Standard deviation (Intercept) & 0.166 & [0.108, 0.254] & 0.136 & [0.081, 0.229] & 0.137 & [0.081, 0.232]\\
-\addlinespace
-Within-group standard error & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.694]\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/random-terms-map-1} \caption{\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown).}\label{fig:random-terms-map}
-\end{figure}
-
-\hypertarget{expanding-on-swn}{%
-\section{Expanding on SWN}\label{expanding-on-swn}}
-
-The preceding section shows that thanks to the availability of code and
-data, it is possible to verify the results reported by SWN. As noted
-earlier, though, an independent researcher might have wondered about the
-implications of the spatial sampling procedure used by SWN. The decision
-to use a sample of counties with reliable basic reproductive numbers,
-although apparently sensible, results in a non-random spatial sampling
-scheme. Turning our attention back to Figure \ref{fig:R0-map}, we form
-the impression that many counties without reliable values of \(R_0\) are
-in more rural, less dense parts of the United States. This impression is
-reinforced when we overlay the boundaries of urban areas with population
-greater than 50,000 on the counties with valid values of \(R_0\) (see
-Figure \ref{fig:urban-areas-map}). The fact that \(R_0\) could not be
-accurately computed in many counties without large urban areas does not
-mean that there was no transmission of the virus: it simply means that
-we do not know with sufficient precision to what extent that was the
-case. The low number of cases may be related to low population and/or
-low population density. This is intriguing, to say the least: by
-excluding cases based on the ability to calculate \(R_0\) we are
-potentially \emph{selecting} the sample in a non-random way.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/urban-areas-map-1} \caption{\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown).}\label{fig:urban-areas-map}
-\end{figure}
-
-A problematic issue with non-random sample selection is that parameter
-estimates can become unreliable, and numerous techniques have been
-developed to address this. A model useful for sample selection problems
-is Heckman's selection model (see Maddala, 1983). The selection model is
-in fact a system of two equations, as follows: \[
-\begin{array}{c}
-y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
-y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
-\end{array}
-\] \noindent where \(y_i^{S*}\) is a latent variable for the sample
-selection process, and \(y_i^{O*}\) is the latent outcome. Vectors
-\(x_i^S\) and \(x_i^O\) are explanatory variables (with the possibility
-that \(x_i^S = x_i^S\)). Both equations include random terms (i.e.,
-\(\epsilon_i^S\) and \(\epsilon_i^O\)). The first equation is designed
-to model the \emph{probability} of sampling, and the second equation the
-outcome of interest (say \(R_0\)). The random terms are jointly
-distributed and correlated with parameter \(\rho\).
-
-What the analyst observes is the following: \[
-y_i^S =
-\begin{cases}
-0 & \text{if } y_i^{S*} < 0\\
-1 & \text{otherwise}
-\end{cases}
-\] \noindent and: \[
-y_i^O =
-\begin{cases}
-0 & \text{if } y_i^{S} = 0\\
-y_i^{O*} & \text{otherwise}
-\end{cases}
-\]
-
-In other words, the outcome of interest is observed \emph{only} for
-certain cases (\(y_i^S=1\), i.e., for sampled observations). The
-probability of sampling depends on \(x_i^S\). For the cases observed,
-the outcome \(y_i^O\) depends on \(x_i^O\).
-
-A sample selection model is estimated using the same selection of
-variables as SWN Model 3. This is Sample Selection Model 1 in Table
-\ref{tab:selection-results}. The first thing to notice about this model
-is that the sample selection process and the outcome are correlated
-(\(\rho\ne0\) with 5\% of confidence). The selection equation indicates
-that the probability of a county to be in the sample increases with
-population density (but at a decreasing rate due to the
-log-transformation), when travel by private modes is more prevalent, and
-as median household income in the county is higher. This is in line with
-the impression made by Figure \ref{fig:urban-areas-map} that counties
-with reliable values of \(R_0\) tended to be those with larger urban
-centers. Once that the selection probabilities are accounted for in the
-model, several things happen with the outcomes model. First, the
-coefficient for population density is still positive, but the magnitude
-changes: in effect, it appears that the effect of density is more
-pronounced than what SWN Model 3 indicated. The coefficient for percent
-of private transportation changes signs. And the coefficient for median
-household income is now significant.
-
-The second model in Table \ref{tab:selection-results} (Selection Model
-2) changes the way the variables are entered into the model. The
-log-transformation of density in SWN and Selection Model 1 assumes that
-the association between density and \(R_0\) is monotonically increasing
-(if the sign of the coefficient is positive) or decreasing (if the sign
-of the coefficient is negative). There are some indications that the
-relationship may actually not be monotonical. For example, Paez et al.
-(2020) found a positive (if non-significant) relationship between
-density and incidence of COVID-19 in the provinces of Spain at the
-beginning of the pandemic. This changed to a negative (and significant)
-relationship during the lockdown. In the case of the US, Fielding-Miller
-et al. (2020) found that the association between COVID-19 deaths and
-population density was positive in rural counties, but negative in urban
-counties. A variable transformation that allows for non-monotonic
-changes in the relationship is the square of the density.
-
-As seen in the table, Selection Model 2 replaces the log-transformation
-of population density with a quadratic expansion. The results of this
-analysis indicate that with this variable transformation, the selection
-and outcome processes are still correlated (\(\rho\ne0\) with 5\% of
-confidence). But a few other interesting things emerge. When we examine
-the outcomes model, we see that the quadratic expansion has a positive
-coefficient for the first order term, but a negative coefficient for the
-second order term. This indicates that \(R_0\) initially tends to
-increase as density grows, but only up to a point, after which the
-negative second term (which grows more rapidly due to the square),
-becomes increasingly dominant. Secondly, the sign of the coefficient for
-travel by private transportation becomes negative again. This, of
-course, makes more sense than the positive sign of Selection Model 1: if
-people tend to travel in private transportation, the potential for
-contact should be lower instead of higher. And finally median household
-income is no longer significant, similar to SWN Model 3.
-
-\begin{table}
-
-\caption{\label{tab:tabulate-sample-selection-results}\label{tab:selection-results}Estimation results of sample selection models}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lcccc}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Selection Model 1} & \multicolumn{2}{c}{Selection Model 2} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Variable & $\beta$ & 95\% CI & $\beta$ & 95\% CI\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{5}{l}{\textbf{Sample Selection Model}}\\
-\hspace{1em}Intercept & -2.237 & [-3.109, -1.365] & -7.339 & [-8.381, -6.297]\\
-\hspace{1em}Log of population density & 0.385 & [0.352, 0.418] &  & \\
-\hspace{1em}Density (1,000 per sq.km) &  &  & 2.484 & [2.13, 2.838]\\
-\hspace{1em}Density squared &  &  & -0.387 & [-0.473, -0.3]\\
-\hspace{1em}Percent of private transportation & 0.025 & [0.016, 0.034] & 0.057 & [0.046, 0.067]\\
-\hspace{1em}Median household income (10,000) & 0.202 & [0.168, 0.235] & 0.32 & [0.283, 0.357]\\
-\addlinespace[0.3em]
-\multicolumn{5}{l}{\textbf{Outcome Model}}\\
-\hspace{1em}Intercept & 0.605 & [-0.257, 1.466] & 2.784 & [1.652, 3.915]\\
-\hspace{1em}Log of population density & 0.39 & [0.354, 0.426] &  & \\
-\hspace{1em}Density (1,000 per sq.km) &  &  & 0.758 & [0.509, 1.008]\\
-\hspace{1em}Density squared &  &  & -0.132 & [-0.187, -0.077]\\
-\hspace{1em}Percent of private transportation & 0.01 & [0.001, 0.018] & -0.011 & [-0.021, -0.001]\\
-\hspace{1em}Median household income (\$10,000) & 0.126 & [0.094, 0.159] & 0.002 & [-0.033, 0.037]\\
-$\sigma$ & 0.954 & [0.904, 1.003] & 0.684 & [0.652, 0.716]\\
-$\rho$ & 0.971 & [0.961, 0.98] & -0.199 & [-0.377, -0.022]\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\hypertarget{proceed-with-caution-spatial-effects-ahead}{%
-\section{Proceed with caution: spatial effects
-ahead}\label{proceed-with-caution-spatial-effects-ahead}}
-
-The results of the selection models, in particular Selection Model 2,
-make us reassess the original conclusion that density has a positive
-association with the basic reproductive number of COVID-19. A spatial
-analyst might still wonder about spatial residual autocorrelation. A
-challenge here is that spatial models tend to be technically more
-demanding, and although spatial models for qualitative variables exist,
-a spatial implementation of the sample selection model does not appear
-to exist. It might be argued that a reproducible research project can
-also allow a researcher to be more adventurous with their modeling
-decisions: since data and code are shared, other researchers can
-promptly and with relative ease poke the methods and see if they appear
-to be sound.
-
-In the present case, it appears that an application of spatial filtering
-(see Getis and Griffith, 2002; Griffith, 2004; Paez, 2019) can help.
-Spatial filtering provides an elegant solution to regression problems
-that may have difficulties handling the spatial structures of spatial
-statistical and econometric models (Griffith, 2000). A key issue in the
-present example is the fact that there are numerous missing
-observations, which prevents the calculation of autocorrelation
-statistics, let alone the estimation of models with spatial components.
-
-The following is an unorthodox, but potentially effective use of filters
-in a sample selection model:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Estimate a sample selection model and retrieve the residuals of the
-  outcome. This will be a vector with missing values for locations that
-  were not sampled.
-\item
-  Fit a spatial filter to the residuals. This is done by regressing the
-  estimated residuals of the \emph{observed} data on the corresponding
-  values of the Moran eigenvectors.
-\item
-  The resulting filter will correlate highly with the known residuals,
-  and will provide information in non-sampled locations that is
-  consistent with the spatial pattern of the known residuals.
-\item
-  Test the filter for spatial autocorrelation:
-
-  4.1 If significant spatial autocorrelation is detected, this would be
-  indicative of residual spatial pattern. Introduce the filter as a
-  covariate in the outcome model of the sample selection model and
-  return to step 1.
-
-  4.2 If no significant spatial autocorrelation is detected, this would
-  be indicative of random residual pattern. Stop.
-\end{enumerate}
-
-This procedure is implemented using a stopping criterion whereby the
-search for the filter only stops when the p-value of Moran's Coefficient
-of the filter fitted to the residuals is greater than 0.25, which was
-chosen as a sufficiently conservative value for testing for
-autocorrelation. The correlation of the known residuals with the
-corresponding elements of the filter is consistently high (the
-correlation coefficient typically is greater than 0.9). The results of
-implementing this procedure appear in Table \ref{tab:selection3-results}
-as Selection Model 3. The results are consistent with Selection Model 2,
-with two intriguing differences: 1) the variance of Sample Model 3 is
-smaller; and 2) the sample and outcome processes are no longer
-correlated (the confidence interval of \(\rho\) includes zero). It
-appears that by capturing the spatial pattern of the residuals, which is
-likely strongly determined by the non-random sampling framework, the
-outcome model is not only substantially more precise, but also appears
-to be independent from the selection process.
-
-\begin{table}[!h]
-
-\caption{\label{tab:tabulate-sample-selection3-results}\label{tab:selection3-results}Estimation results of sample selection model with spatial filter}
-\centering
-\fontsize{8}{10}\selectfont
-\begin{tabular}[t]{lcc}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Selection Model 3} \\
-\cmidrule(l{3pt}r{3pt}){2-3}
-Variable & $\beta$ & 95\% CI\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{3}{l}{\textbf{Sample Selection Model}}\\
-\hspace{1em}Intercept & -7.304 & [-8.346, -6.262]\\
-\hspace{1em}Density (1,000 per sq.km) & 2.445 & [2.089, 2.802]\\
-\hspace{1em}Density squared & -0.380 & [-0.468, -0.292]\\
-\hspace{1em}Percent of private transportation & 0.056 & [0.046, 0.067]\\
-\hspace{1em}Median household income (10,000) & 0.318 & [0.28, 0.356]\\
-\addlinespace[0.3em]
-\multicolumn{3}{l}{\textbf{Outcome Model}}\\
-\hspace{1em}Intercept & 2.563 & [2.497, 2.629]\\
-\hspace{1em}Density (1,000 per sq.km) & 0.760 & [0.746, 0.774]\\
-\hspace{1em}Density squared & -0.133 & [-0.135, -0.13]\\
-\hspace{1em}Percent of private transportation & -0.011 & [-0.012, -0.011]\\
-\hspace{1em}Median household income (\$10,000) & 0.002 & [-0.001, 0.004]\\
-\hspace{1em}Spatial filter & 1.000 & [0.998, 1.001]\\
-$\sigma$ & 0.017 & [0.015, 0.019]\\
-$\rho$ & -0.304 & [-0.957, 0.349]\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-How relevant are the differences between the various model
-specifications presented above? Figure \ref{fig:comparison-results}
-shows the relationship between density and \(R_0\) implied by SWN Model
-3, Selection Model 2, and Selection Model 3. The left panel of the
-figure shows the non-linear but monotonic relationship implied by SWN
-Model 1. The conclusion is that at higher densities, \(R_0\) is
-\emph{always} higher. The two panels on the right, in contrast, shows
-that Selection Model 2 and Selection Model 3 coincide that \(R_0\) tends
-to increase as density grows. This continues until a density of
-approximately 2.9 (1,000 people per sq.km). At higher densities than
-that the relationship between density and \(R_0\) begins to weaken, and
-the relationship becomes negative at densities higher than approximately
-5.7 (1,000 people per sq.km).
-
-To put this into context, other things being equal, the effect of
-density in a county like Charlottesville in Virginia (density
-\textasciitilde1,639 people per sq.km) is roughly the same as that in a
-county like Philadelphia (density \textasciitilde4,127 people per
-sq.km). In contrast, the effect of density on \(R_0\) in a county like
-Arlington in Virginia (density \textasciitilde3,093 people per sq.km) is
-\emph{stronger} than either of the previous two examples. Lastly, the
-density of counties like San Francisco in California, or Queens and
-Bronx in NY, which are among the densest in the US, contributes even
-less to \(R_0\) than even the most rural counties in the country.
-
-It is worth at this point to recall Cressie's dictum about modelling:
-``{[}w{]}hat is one person's mean structure could be another person's
-correlation structure'' (Cressie, 1989, p. 201). There are almost always
-multiple ways to approach a modelling situation. In the present case, we
-would argue that spatial sampling is an important aspect of the modeling
-process, but one that perhaps required different technical skills than
-those available to SWN. There is nothing wrong with that. What matters
-is that, by adopting relatively high reproducibility standards, these
-researchers made a valuable and honest contribution to the collective
-enterprise of seeking knowledge. Their effort, and subsequent efforts to
-validate and expand on their work, can potentially contribute to provide
-clarity to ongoing conversations about the relevance of density and the
-spread of COVID-19.
-
-In particular, it is noteworthy that a sample selection model with a
-different variable transformation does not lend support to the thesis
-that higher density is \emph{always} associated with a greater risk of
-spread of the virus {[}as put by Wong and Li, ``\,`Density is destiny'
-is probably an overstatement''; (2020){]}. At the same time, this also
-stands in contrast to the findings of Hamidi et al., who found that
-higher density was either not significantly associated with the rate of
-the virus in a cross-sectional study (Hamidi et al., 2020b), or was
-negatively associated with it in a longitudinal setting {[}Hamidi et al.
-(2020a). In this sense, the conclusion that density does not aggravate
-the pandemic may have been somewhat premature; instead, reanalysis of
-the data of SWN suggests that Fielding-Miller et al. (2020) might be
-onto something with respect to the difference between rural and urban
-counties. More generally, there is no doubt that in population-level
-studies density is indicative of proximity, but it also potentially is a
-proxy for adaptive behavior. And it is possible that the determining
-factor during COVID-19, at least in the US, has been variations in
-perceptions of the risks associated with contagion (Chauhan et al.,
-2021), and subsequent compensations in behavior in more and less dense
-regions.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/comparison-results-1} \caption{\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2.}\label{fig:comparison-results}
-\end{figure}
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-The tension between the need to publish research potentially useful in
-dealing with a global pandemic, and a potential ``carnage of substandard
-research'' (Bramstedt, 2020), highlights the importance of efforts to
-maintain the quality of scientific outputs during COVID-19. An important
-part of quality control is the ability of independent researchers to
-verify and examine the results of materials published in the literature.
-As previous research illustrates, reproducibility in scientific research
-remains an important but elusive goal (Gustot, 2020; e.g., Iqbal et al.,
-2016; Stodden et al., 2018; Sumner et al., 2020). This idea is
-reinforced by the review conducted for this paper in the context of
-research about population density and the spread of COVID-19.
-
-Taking one recent example from the literature {[}Sy et al., Sy et al.
-(2021); SWN{]}, the present paper illustrates the importance of good
-reproducibility practices. Sharing data and code can catalyze research,
-by allowing independent verification of findings, as well as additional
-research. After verifying the results of SWN, experiments with sample
-selection models and variations in the definition of model inputs, lead
-to an important reappraisal of the conclusion that high density is
-associated with greater spread of the virus. Instead, the possibility of
-a non-monotonical relationship between population density and contagion
-is raised.
-
-In the spirit of openness, this paper is prepared as an R Markdown
-document, an a companion data package is provided. The data package
-contains the relevant documentation of the data, and all data
-pre-processing is fully documented. Hopefully this, and similar
-reproducible papers, will continue to encourage others to adopt
-reproducible standards in their research.
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Ahmad2020association}{}%
-Ahmad, K., Erqou, S., Shah, N., Nazir, U., Morrison, A.R., Choudhary,
-G., Wu, W.-C., 2020. Association of poor housing conditions with
-COVID-19 incidence and mortality across US counties. PLOS ONE 15,
-e0241327.
-doi:\href{https://doi.org/10.1371/journal.pone.0241327}{10.1371/journal.pone.0241327}
-
-\leavevmode\hypertarget{ref-Amadu2021assessing}{}%
-Amadu, I., Ahinkorah, B.O., Afitiri, A.-R., Seidu, A.-A., Ameyaw, E.K.,
-Hagan, J.E., Duku, E., Aram, S.A., 2021. Assessing sub-regional-specific
-strengths of healthcare systems associated with COVID-19 prevalence,
-deaths and recoveries in africa. PLOS ONE 16, e0247274.
-doi:\href{https://doi.org/10.1371/journal.pone.0247274}{10.1371/journal.pone.0247274}
-
-\leavevmode\hypertarget{ref-Anazco2021publication}{}%
-Añazco, D., Nicolalde, B., Espinosa, I., Camacho, J., Mushtaq, M.,
-Gimenez, J., Teran, E., 2021. Publication rate and citation counts for
-preprints released during the COVID-19 pandemic: The good, the bad and
-the ugly. PeerJ 9, e10927.
-doi:\href{https://doi.org/10.7717/peerj.10927}{10.7717/peerj.10927}
-
-\leavevmode\hypertarget{ref-Basu2017ten}{}%
-Basu, S., Carney, M.A., Kenworthy, N.J., 2017. Ten years after the
-financial crisis: The long reach of austerity and its global impacts on
-health. Social Science \& Medicine 187, 203--207.
-doi:\href{https://doi.org/10.1016/j.socscimed.2017.06.026}{10.1016/j.socscimed.2017.06.026}
-
-\leavevmode\hypertarget{ref-Bhadra2021impact}{}%
-Bhadra, A., Mukherjee, A., Sarkar, K., 2021. Impact of population
-density on covid-19 infected and mortality rate in india. Modeling Earth
-Systems and Environment 7, 623--629.
-doi:\href{https://doi.org/10.1007/s40808-020-00984-7}{10.1007/s40808-020-00984-7}
-
-\leavevmode\hypertarget{ref-Bramstedt2020carnage}{}%
-Bramstedt, K.A., 2020. The carnage of substandard research during the
-COVID-19 pandemic: A call for quality. Journal of Medical Ethics 46,
-803--807.
-doi:\href{https://doi.org/10.1136/medethics-2020-106494}{10.1136/medethics-2020-106494}
-
-\leavevmode\hypertarget{ref-Brandtner2021creatures}{}%
-Brandtner, C., Bettencourt, L.M.A., Berman, M.G., Stier, A.J., 2021.
-Creatures of the state? Metropolitan counties compensated for state
-inaction in initial u.s. Response to COVID-19 pandemic. PLOS ONE 16,
-e0246249.
-doi:\href{https://doi.org/10.1371/journal.pone.0246249}{10.1371/journal.pone.0246249}
-
-\leavevmode\hypertarget{ref-Broggini2017reproducible}{}%
-Broggini, F., Dellinger, J., Fomel, S., Liu, Y., 2017. Reproducible
-research: Geophysics papers of the future - introduction. Geophysics 82.
-doi:\href{https://doi.org/10.1190/geo2017-0918-spseintro.1}{10.1190/geo2017-0918-spseintro.1}
-
-\leavevmode\hypertarget{ref-Brunsdon2020opening}{}%
-Brunsdon, C., Comber, A., 2020. Opening practice: Supporting
-reproducibility and critical spatial data science. Journal of
-Geographical Systems.
-doi:\href{https://doi.org/10.1007/s10109-020-00334-2}{10.1007/s10109-020-00334-2}
-
-\leavevmode\hypertarget{ref-Chauhan2021covid}{}%
-Chauhan, R.S., Capasso da Silva, D., Salon, D., Shamshiripour, A.,
-Rahimi, E., Sutradhar, U., Khoeini, S., Mohammadian, A.(Kouros).,
-Derrible, S., Pendyala, R., 2021. COVID-19 related attitudes and risk
-perceptions across urban, rural, and suburban areas in the united
-states. Findings.
-doi:\href{https://doi.org/10.32866/001c.23714}{10.32866/001c.23714}
-
-\leavevmode\hypertarget{ref-Cressie1989geostatistics}{}%
-Cressie, N., 1989. Geostatistics. The American Statistician 43, 197.
-doi:\href{https://doi.org/10.2307/2685361}{10.2307/2685361}
-
-\leavevmode\hypertarget{ref-Cruz2020exploring}{}%
-Cruz, C.J.P., Ganly, R., Li, Z., Gietel-Basten, S., 2020. Exploring the
-young demographic profile of COVID-19 cases in hong kong: Evidence from
-migration and travel history data. PLOS ONE 15, e0235306.
-doi:\href{https://doi.org/10.1371/journal.pone.0235306}{10.1371/journal.pone.0235306}
-
-\leavevmode\hypertarget{ref-Farber2011running}{}%
-Farber, S., Páez, A., 2011. Running to stay in place: The time-use
-implications of automobile oriented land-use and travel. Journal of
-Transport Geography 19, 782--793.
-doi:\href{https://doi.org/10.1016/j.jtrangeo.2010.09.008}{10.1016/j.jtrangeo.2010.09.008}
-
-\leavevmode\hypertarget{ref-Feng2020spread}{}%
-Feng, Y., Li, Q., Tong, X., Wang, R., Zhai, S., Gao, C., Lei, Z., Chen,
-S., Zhou, Y., Wang, J., Yan, X., Xie, H., Chen, P., Liu, S., Xv, X.,
-Liu, S., Jin, Y., Wang, C., Hong, Z., Luan, K., Wei, C., Xu, J., Jiang,
-H., Xiao, C., Guo, Y., 2020. Spatiotemporal spread pattern of the
-COVID-19 cases in china. PLOS ONE 15, e0244351.
-doi:\href{https://doi.org/10.1371/journal.pone.0244351}{10.1371/journal.pone.0244351}
-
-\leavevmode\hypertarget{ref-Feyman2020effectiveness}{}%
-Feyman, Y., Bor, J., Raifman, J., Griffith, K.N., 2020. Effectiveness of
-COVID-19 shelter-in-place orders varied by state. PLOS ONE 15, e0245008.
-doi:\href{https://doi.org/10.1371/journal.pone.0245008}{10.1371/journal.pone.0245008}
-
-\leavevmode\hypertarget{ref-Fielding2020social}{}%
-Fielding-Miller, R.K., Sundaram, M.E., Brouwer, K., 2020. Social
-determinants of COVID-19 mortality at the county level. PLOS ONE 15,
-e0240151.
-doi:\href{https://doi.org/10.1371/journal.pone.0240151}{10.1371/journal.pone.0240151}
-
-\leavevmode\hypertarget{ref-Florida2020how}{}%
-Florida, R., Glaeser, E., Sharif, M., Bedi, K., Campanella, T., Chee,
-C., Doctoroff, D., Katz, B., Katz, R., Kotkin, J., 2020. How life in our
-cities will look after the coronavirus pandemic. Foreign Policy 1.
-
-\leavevmode\hypertarget{ref-Fraser2021evolving}{}%
-Fraser, N., Brierley, L., Dey, G., Polka, J.K., Pálfy, M., Nanni, F.,
-Coates, J.A., 2021. The evolving role of preprints in the dissemination
-of COVID-19 research and their impact on the science communication
-landscape. PLOS Biology 19, e3000959.
-doi:\href{https://doi.org/10.1371/journal.pbio.3000959}{10.1371/journal.pbio.3000959}
-
-\leavevmode\hypertarget{ref-Getis2002comparative}{}%
-Getis, A., Griffith, D.A., 2002. Comparative spatial filtering in
-regression analysis. Geographical Analysis 34, 130--140.
-
-\leavevmode\hypertarget{ref-Gomez2021infekta}{}%
-Gomez, J., Prieto, J., Leon, E., Rodríguez, A., 2021. INFEKTA---an
-agent-based model for transmission of infectious diseases: The COVID-19
-case in bogotá, colombia. PLOS ONE 16, e0245787.
-doi:\href{https://doi.org/10.1371/journal.pone.0245787}{10.1371/journal.pone.0245787}
-
-\leavevmode\hypertarget{ref-Griffith2000linear}{}%
-Griffith, D.A., 2000. A linear regression solution to the spatial
-autocorrelation problem. Journal of Geographical Systems 2, 141--156.
-
-\leavevmode\hypertarget{ref-Griffith2004spatial}{}%
-Griffith, D.A., 2004. A spatial filtering specification for the
-autologistic model. Environment and Planning A 36, 1791--1811.
-
-\leavevmode\hypertarget{ref-Gustot2020quality}{}%
-Gustot, T., 2020. Quality and reproducibility during the COVID-19
-pandemic. JHEP Rep 2, 100141.
-doi:\href{https://doi.org/10.1016/j.jhepr.2020.100141}{10.1016/j.jhepr.2020.100141}
-
-\leavevmode\hypertarget{ref-Hamidi2020longitudinal}{}%
-Hamidi, S., Ewing, R., Sabouri, S., 2020a. Longitudinal analyses of the
-relationship between development density and the COVID-19 morbidity and
-mortality rates: Early evidence from 1,165 metropolitan counties in the
-united states. Health \& Place 64, 102378.
-doi:\href{https://doi.org/10.1016/j.healthplace.2020.102378}{10.1016/j.healthplace.2020.102378}
-
-\leavevmode\hypertarget{ref-Hamidi2020density}{}%
-Hamidi, S., Sabouri, S., Ewing, R., 2020b. Does density aggravate the
-COVID-19 pandemic? Journal of the American Planning Association 86,
-495--509.
-doi:\href{https://doi.org/10.1080/01944363.2020.1777891}{10.1080/01944363.2020.1777891}
-
-\leavevmode\hypertarget{ref-Hamidi2021compact}{}%
-Hamidi, S., Zandiatashbar, A., 2021. Compact development and adherence
-to stay-at-home order during the COVID-19 pandemic: A longitudinal
-investigation in the united states. Landscape and Urban Planning 205,
-103952. doi:\url{https://doi.org/10.1016/j.landurbplan.2020.103952}
-
-\leavevmode\hypertarget{ref-Harris2021Changes}{}%
-Harris, M.A., Branion-Calles, M., 2021. Changes in commute mode
-attributed to COVID-19 risk in canadian national survey data. Findings.
-doi:\href{https://doi.org/10.32866/001c.19088}{10.32866/001c.19088}
-
-\leavevmode\hypertarget{ref-Herndon2014high}{}%
-Herndon, T., Ash, M., Pollin, R., 2014. Does high public debt
-consistently stifle economic growth? A critique of reinhart and rogoff.
-Cambridge Journal of Economics 38, 257--279.
-doi:\href{https://doi.org/10.1093/cje/bet075}{10.1093/cje/bet075}
-
-\leavevmode\hypertarget{ref-Inbaraj2021seroprevalence}{}%
-Inbaraj, L.R., George, C.E., Chandrasingh, S., 2021. Seroprevalence of
-COVID-19 infection in a rural district of south india: A
-population-based seroepidemiological study. PLOS ONE 16, e0249247.
-doi:\href{https://doi.org/10.1371/journal.pone.0249247}{10.1371/journal.pone.0249247}
-
-\leavevmode\hypertarget{ref-Ince2012case}{}%
-Ince, D.C., Hatton, L., Graham-Cumming, J., 2012. The case for open
-computer programs. Nature 482, 485--488.
-doi:\href{https://doi.org/10.1038/nature10836}{10.1038/nature10836}
-
-\leavevmode\hypertarget{ref-Ioannidis2014increasing}{}%
-Ioannidis, J.P.A., Greenland, S., Hlatky, M.A., Khoury, M.J., Macleod,
-M.R., Moher, D., Schulz, K.F., Tibshirani, R., 2014. Increasing value
-and reducing waste in research design, conduct, and analysis. Lancet
-383, 166--175.
-doi:\href{https://doi.org/10.1016/s0140-6736(13)62227-8}{10.1016/s0140-6736(13)62227-8}
-
-\leavevmode\hypertarget{ref-Iqbal2016reproducible}{}%
-Iqbal, S.A., Wallach, J.D., Khoury, M.J., Schully, S.D., Ioannidis,
-J.P.A., 2016. Reproducible research practices and transparency across
-the biomedical literature. Plos Biology 14.
-doi:\href{https://doi.org/10.1371/journal.pbio.1002333}{10.1371/journal.pbio.1002333}
-
-\leavevmode\hypertarget{ref-Jamal2020Changes}{}%
-Jamal, S., Paez, A., 2020. Changes in trip-making frequency by mode
-during COVID-19. Findings.
-doi:\href{https://doi.org/10.32866/001c.17977}{10.32866/001c.17977}
-
-\leavevmode\hypertarget{ref-Kadi2020population}{}%
-Kadi, N., Khelfaoui, M., 2020. Population density, a factor in the
-spread of COVID-19 in algeria: Statistic study. Bulletin of the National
-Research Centre 44.
-doi:\href{https://doi.org/10.1186/s42269-020-00393-x}{10.1186/s42269-020-00393-x}
-
-\leavevmode\hypertarget{ref-Khavarian2021high}{}%
-Khavarian-Garmsir, A.R., Sharifi, A., Moradpour, N., 2021. Are
-high-density districts more vulnerable to the COVID-19 pandemic?
-Sustainable Cities and Society 70, 102911.
-doi:\href{https://doi.org/10.1016/j.scs.2021.102911}{10.1016/j.scs.2021.102911}
-
-\leavevmode\hypertarget{ref-Konkol2019examination}{}%
-Konkol, M., Kray, C., 2019. In-depth examination of spatiotemporal
-figures in open reproducible research. Cartography and Geographic
-Information Science 46, 412--427.
-doi:\href{https://doi.org/10.1080/15230406.2018.1512421}{10.1080/15230406.2018.1512421}
-
-\leavevmode\hypertarget{ref-Konkol2019computational}{}%
-Konkol, M., Kray, C., Pfeiffer, M., 2019. Computational reproducibility
-in geoscientific papers: Insights from a series of studies with
-geoscientists and a reproduction study. International Journal of
-Geographical Information Science 33, 408--429.
-doi:\href{https://doi.org/10.1080/13658816.2018.1508687}{10.1080/13658816.2018.1508687}
-
-\leavevmode\hypertarget{ref-Kwon2021swamped}{}%
-Kwon, D., 2020. How swamped preprint servers are blocking bad
-coronavirus research. Nature 581, 130--132.
-
-\leavevmode\hypertarget{ref-Lee2020human}{}%
-Lee, M., Zhao, J., Sun, Q., Pan, Y., Zhou, W., Xiong, C., Zhang, L.,
-2020. Human mobility trends during the early stage of the COVID-19
-pandemic in the united states. PLOS ONE 15, e0241468.
-doi:\href{https://doi.org/10.1371/journal.pone.0241468}{10.1371/journal.pone.0241468}
-
-\leavevmode\hypertarget{ref-Li2018effect}{}%
-Li, R., Richmond, P., Roehner, B.M., 2018. Effect of population density
-on epidemics. Physica A: Statistical Mechanics and its Applications 510,
-713--724.
-doi:\href{https://doi.org/10.1016/j.physa.2018.07.025}{10.1016/j.physa.2018.07.025}
-
-\leavevmode\hypertarget{ref-Maddala1983limited}{}%
-Maddala, G.S., 1983. Limited-dependent and qualitative variables in
-econometrics. Cambridge University Press, Cambridge.
-
-\leavevmode\hypertarget{ref-Micallef2020first}{}%
-Micallef, S., Piscopo, T.V., Casha, R., Borg, D., Vella, C., Zammit,
-M.-A., Borg, J., Mallia, D., Farrugia, J., Vella, S.M., Xerri, T.,
-Portelli, A., Fenech, M., Fsadni, C., Mallia Azzopardi, C., 2020. The
-first wave of COVID-19 in malta; a national cross-sectional study. PLOS
-ONE 15, e0239389.
-doi:\href{https://doi.org/10.1371/journal.pone.0239389}{10.1371/journal.pone.0239389}
-
-\leavevmode\hypertarget{ref-Molloy2020Tracing}{}%
-Molloy, J., Tchervenkov, C., Hintermann, B., Axhausen, K.W., 2020.
-Tracing the sars-CoV-2 impact: The first month in switzerland. Findings.
-doi:\href{https://doi.org/10.32866/001c.12903}{10.32866/001c.12903}
-
-\leavevmode\hypertarget{ref-Moore1970some}{}%
-Moore, E.G., 1970. Some spatial properties of urban contact fields.
-Geographical Analysis 2, 376--386.
-
-\leavevmode\hypertarget{ref-Moore1970urban}{}%
-Moore, E.G., Brown, L.A., 1970. Urban acquaintance fields: An evaluation
-of a spatial model. Environment and Planning 2, 443--454.
-
-\leavevmode\hypertarget{ref-Noland1995perceived}{}%
-Noland, R.B., 1995. PERCEIVED RISK AND MODAL CHOICE - RISK COMPENSATION
-IN TRANSPORTATION SYSTEM. Accident Analysis and Prevention 27, 503--521.
-doi:\href{https://doi.org/10.1016/0001-4575(94)00087-3}{10.1016/0001-4575(94)00087-3}
-
-\leavevmode\hypertarget{ref-Noland2021mobility}{}%
-Noland, R.B., 2021. Mobility and the effective reproduction rate of
-COVID-19. Journal of Transport \& Health 20, 101016.
-doi:\url{https://doi.org/10.1016/j.jth.2021.101016}
-
-\leavevmode\hypertarget{ref-Noury2021how}{}%
-Noury, A., François, A., Gergaud, O., Garel, A., 2021. How does COVID-19
-affect electoral participation? Evidence from the french municipal
-elections. PLOS ONE 16, e0247026.
-doi:\href{https://doi.org/10.1371/journal.pone.0247026}{10.1371/journal.pone.0247026}
-
-\leavevmode\hypertarget{ref-Paez2019using}{}%
-Paez, A., 2019. Using spatial filters and exploratory data analysis to
-enhance regression models of spatial data. Geographical Analysis 51,
-314--338.
-doi:\href{https://doi.org/10.1111/gean.12180}{10.1111/gean.12180}
-
-\leavevmode\hypertarget{ref-Paez2020using}{}%
-Paez, A., 2020. Using google community mobility reports to investigate
-the incidence of COVID-19 in the united states. Findings.
-doi:\url{https://doi.org/10.32866/001c.12976}
-
-\leavevmode\hypertarget{ref-Paez2020spatio}{}%
-Paez, A., Lopez, F.A., Menezes, T., Cavalcanti, R., Pitta, M.G. da R.,
-2020. A spatio-temporal analysis of the environmental correlates of
-COVID-19 incidence in spain. Geographical Analysis n/a.
-doi:\href{https://doi.org/10.1111/gean.12241}{10.1111/gean.12241}
-
-\leavevmode\hypertarget{ref-Pequeno2020air}{}%
-Pequeno, P., Mendel, B., Rosa, C., Bosholn, M., Souza, J.L., Baccaro,
-F., Barbosa, R., Magnusson, W., 2020. Air transportation, population
-density and temperature predict the spread of COVID-19 in brazil. PeerJ
-8, e9322.
-doi:\href{https://doi.org/10.7717/peerj.9322}{10.7717/peerj.9322}
-
-\leavevmode\hypertarget{ref-Phillips2011risk}{}%
-Phillips, R.O., Fyhri, A., Sagberg, F., 2011. Risk compensation and
-bicycle helmets. Risk Analysis 31, 1187--1195.
-doi:\href{https://doi.org/10.1111/j.1539-6924.2011.01589.x}{10.1111/j.1539-6924.2011.01589.x}
-
-\leavevmode\hypertarget{ref-Praharaj2020Using}{}%
-Praharaj, S., King, D., Pettit, C., Wentz, E., 2020. Using aggregated
-mobility data to measure the effect of COVID-19 policies on mobility
-changes in sydney, london, phoenix, and pune. Findings.
-doi:\href{https://doi.org/10.32866/001c.17590}{10.32866/001c.17590}
-
-\leavevmode\hypertarget{ref-Richens2000condoms}{}%
-Richens, J., Imrie, J., Copas, A., 2000. Condoms and seat belts: The
-parallels and the lessons. Lancet 355, 400--403.
-doi:\href{https://doi.org/10.1016/s0140-6736(99)09109-6}{10.1016/s0140-6736(99)09109-6}
-
-\leavevmode\hypertarget{ref-Rocklov2020high}{}%
-Rocklöv, J., Sjödin, H., 2020. High population densities catalyse the
-spread of COVID-19. Journal of Travel Medicine 27.
-doi:\href{https://doi.org/10.1093/jtm/taaa038}{10.1093/jtm/taaa038}
-
-\leavevmode\hypertarget{ref-Roy2020factors}{}%
-Roy, S., Ghosh, P., 2020. Factors affecting COVID-19 infected and death
-rates inform lockdown-related policymaking. PLOS ONE 15, e0241165.
-doi:\href{https://doi.org/10.1371/journal.pone.0241165}{10.1371/journal.pone.0241165}
-
-\leavevmode\hypertarget{ref-Sharifi2020covid}{}%
-Sharifi, A., Khavarian-Garmsir, A.R., 2020. The COVID-19 pandemic:
-Impacts on cities and major lessons for urban planning, design, and
-management. Science of The Total Environment 749, 142391.
-doi:\url{https://doi.org/10.1016/j.scitotenv.2020.142391}
-
-\leavevmode\hypertarget{ref-Skorka2020macroecology}{}%
-Skórka, P., Grzywacz, B., Moroń, D., Lenda, M., 2020. The macroecology
-of the COVID-19 pandemic in the anthropocene. PLOS ONE 15, e0236856.
-doi:\href{https://doi.org/10.1371/journal.pone.0236856}{10.1371/journal.pone.0236856}
-
-\leavevmode\hypertarget{ref-Souris2020covid}{}%
-Souris, M., Gonzalez, J.-P., 2020. COVID-19: Spatial analysis of
-hospital case-fatality rate in france. PLOS ONE 15, e0243606.
-doi:\href{https://doi.org/10.1371/journal.pone.0243606}{10.1371/journal.pone.0243606}
-
-\leavevmode\hypertarget{ref-Stephens2021impact}{}%
-Stephens, K.E., Chernyavskiy, P., Bruns, D.R., 2021. Impact of altitude
-on COVID-19 infection and death in the united states: A modeling and
-observational study. PLOS ONE 16, e0245055.
-doi:\href{https://doi.org/10.1371/journal.pone.0245055}{10.1371/journal.pone.0245055}
-
-\leavevmode\hypertarget{ref-Stodden2018empirical}{}%
-Stodden, V., Seiler, J., Ma, Z.K., 2018. An empirical analysis of
-journal policy effectiveness for computational reproducibility.
-Proceedings of the National Academy of Sciences of the United States of
-America 115, 2584--2589.
-doi:\href{https://doi.org/10.1073/pnas.1708290115}{10.1073/pnas.1708290115}
-
-\leavevmode\hypertarget{ref-Sumner2020reproducibility}{}%
-Sumner, J., Haynes, L., Nathan, S., Hudson-Vitale, C., McIntosh, L.D.,
-2020. Reproducibility and reporting practices in COVID-19 preprint
-manuscripts. medRxiv 2020.03.24.20042796.
-doi:\href{https://doi.org/10.1101/2020.03.24.20042796}{10.1101/2020.03.24.20042796}
-
-\leavevmode\hypertarget{ref-Sun2020impacts}{}%
-Sun, Z., Zhang, H., Yang, Y., Wan, H., Wang, Y., 2020. Impacts of
-geographic factors and population density on the COVID-19 spreading
-under the lockdown policies of china. Science of The Total Environment
-746, 141347.
-doi:\href{https://doi.org/10.1016/j.scitotenv.2020.141347}{10.1016/j.scitotenv.2020.141347}
-
-\leavevmode\hypertarget{ref-Sy2021population}{}%
-Sy, K.T.L., White, L.F., Nichols, B.E., 2021. Population density and
-basic reproductive number of COVID-19 across united states counties.
-PLOS ONE 16, e0249271.
-doi:\href{https://doi.org/10.1371/journal.pone.0249271}{10.1371/journal.pone.0249271}
-
-\leavevmode\hypertarget{ref-Vlasschaert2020proliferation}{}%
-Vlasschaert, C., Topf, J.M., Hiremath, S., 2020. Proliferation of papers
-and preprints during the coronavirus disease 2019 pandemic: Progress or
-problems with peer review? Advances in Chronic Kidney Disease 27,
-418--426.
-doi:\href{https://doi.org/10.1053/j.ackd.2020.08.003}{10.1053/j.ackd.2020.08.003}
-
-\leavevmode\hypertarget{ref-Wang2021transmission}{}%
-Wang, F., Tan, Z., Yu, Z., Yao, S., Guo, C., 2021. Transmission and
-control pressure analysis of the COVID-19 epidemic situation using
-multisource spatio-temporal big data. PLOS ONE 16, e0249145.
-doi:\href{https://doi.org/10.1371/journal.pone.0249145}{10.1371/journal.pone.0249145}
-
-\leavevmode\hypertarget{ref-White2020state}{}%
-White, E.R., Hébert-Dufresne, L., 2020. State-level variation of initial
-COVID-19 dynamics in the united states. PLOS ONE 15, e0240648.
-doi:\href{https://doi.org/10.1371/journal.pone.0240648}{10.1371/journal.pone.0240648}
-
-\leavevmode\hypertarget{ref-Wong2020spreading}{}%
-Wong, D.W.S., Li, Y., 2020. Spreading of COVID-19: Density matters. PLOS
-ONE 15, e0242398.
-doi:\href{https://doi.org/10.1371/journal.pone.0242398}{10.1371/journal.pone.0242398}
-
-\end{CSLReferences}
-
 \end{document}

### 2021-11-18 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Reorganized the repository after paper's publication online at Geographical Analysis

@@
 author:
-  - name: Author
-    email: author@institution.edu
-    affiliation: University
+  - name: Antonio Paez
+    email: paezha@mcmaster.ca
+    affiliation: McMaster University
     footnote: 1
 address:
-  - code: University
-    address: Department, Street, City, State ZIP
+  - code: McMaster University
+    address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
 footnote:
@@
 
+  This paper is now published in Geographical Analysis (\url{https://doi.org/10.1111/gean.12307})
+
 journal: "Geographical Analysis"
 date: "`r Sys.Date()`"
-bibliography: [bibliography.bib, packages.bib]
+bibliography: bibliography.bib
 linenumbers: true

### 2021-11-18 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Reorganized the repository after paper's publication online at Geographical Analysis

--- 2021-09-16 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
@@
   shows, reproducibility is one of the keys to ensure this.
+
+  This paper is now published in Geographical Analysis
+  (\url{https://doi.org/10.1111/gean.12307})
   \end{abstract}
@@
 \midrule
-Intercept & 2.274 & [2.167, 2.381] & 3.347 & [2.676, 4.018] & 3.386 & [2.614, 4.157]\\
-Log of population density & 0.162 & [0.133, 0.191] & 0.145 & [0.115, 0.176] & 0.147 & [0.113, 0.18]\\
-Percent of private transportation &  &  & -0.013 & [-0.02, -0.005] & -0.013 & [-0.021, -0.005]\\
-Median household income (\$10,000) &  &  &  &  & -0.003 & [-0.033, 0.026]\\
-Standard deviation (Intercept) & 0.166 & [0.108, 0.254] & 0.136 & [0.081, 0.229] & 0.137 & [0.081, 0.232]\\
+Intercept & 2.274 & {}[2.167, 2.381] & 3.347 & {}[2.676, 4.018] & 3.386 & {}[2.614, 4.157]\\
+Log of population density & 0.162 & {}[0.133, 0.191] & 0.145 & {}[0.115, 0.176] & 0.147 & {}[0.113, 0.18]\\
+Percent of private transportation &  &  & -0.013 & {}[-0.02, -0.005] & -0.013 & {}[-0.021, -0.005]\\
+Median household income (\$10,000) &  &  &  &  & -0.003 & {}[-0.033, 0.026]\\
+Standard deviation (Intercept) & 0.166 & {}[0.108, 0.254] & 0.136 & {}[0.081, 0.229] & 0.137 & {}[0.081, 0.232]\\
 \addlinespace
-Within-group standard error & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.694]\\
+Within-group standard error & 0.665 & {}[0.638, 0.693] & 0.665 & {}[0.638, 0.693] & 0.665 & {}[0.638, 0.694]\\
 \bottomrule
@@
 \multicolumn{5}{l}{\textbf{Sample Selection Model}}\\
-\hspace{1em}Intercept & -2.237 & [-3.109, -1.365] & -7.339 & [-8.381, -6.297]\\
-\hspace{1em}Log of population density & 0.385 & [0.352, 0.418] &  & \\
-\hspace{1em}Density (1,000 per sq.km) &  &  & 2.484 & [2.13, 2.838]\\
-\hspace{1em}Density squared &  &  & -0.387 & [-0.473, -0.3]\\
-\hspace{1em}Percent of private transportation & 0.025 & [0.016, 0.034] & 0.057 & [0.046, 0.067]\\
-\hspace{1em}Median household income (10,000) & 0.202 & [0.168, 0.235] & 0.32 & [0.283, 0.357]\\
+\hspace{1em}Intercept & -2.237 & {}[-3.109, -1.365] & -7.339 & {}[-8.381, -6.297]\\
+\hspace{1em}Log of population density & 0.385 & {}[0.352, 0.418] &  & \\
+\hspace{1em}Density (1,000 per sq.km) &  &  & 2.484 & {}[2.13, 2.838]\\
+\hspace{1em}Density squared &  &  & -0.387 & {}[-0.473, -0.3]\\
+\hspace{1em}Percent of private transportation & 0.025 & {}[0.016, 0.034] & 0.057 & {}[0.046, 0.067]\\
+\hspace{1em}Median household income (10,000) & 0.202 & {}[0.168, 0.235] & 0.32 & {}[0.283, 0.357]\\
 \addlinespace[0.3em]
 \multicolumn{5}{l}{\textbf{Outcome Model}}\\
-\hspace{1em}Intercept & 0.605 & [-0.257, 1.466] & 2.784 & [1.652, 3.915]\\
-\hspace{1em}Log of population density & 0.39 & [0.354, 0.426] &  & \\
-\hspace{1em}Density (1,000 per sq.km) &  &  & 0.758 & [0.509, 1.008]\\
-\hspace{1em}Density squared &  &  & -0.132 & [-0.187, -0.077]\\
-\hspace{1em}Percent of private transportation & 0.01 & [0.001, 0.018] & -0.011 & [-0.021, -0.001]\\
-\hspace{1em}Median household income (\$10,000) & 0.126 & [0.094, 0.159] & 0.002 & [-0.033, 0.037]\\
-$\sigma$ & 0.954 & [0.904, 1.003] & 0.684 & [0.652, 0.716]\\
-$\rho$ & 0.971 & [0.961, 0.98] & -0.199 & [-0.377, -0.022]\\
+\hspace{1em}Intercept & 0.605 & {}[-0.257, 1.466] & 2.784 & {}[1.652, 3.915]\\
+\hspace{1em}Log of population density & 0.39 & {}[0.354, 0.426] &  & \\
+\hspace{1em}Density (1,000 per sq.km) &  &  & 0.758 & {}[0.509, 1.008]\\
+\hspace{1em}Density squared &  &  & -0.132 & {}[-0.187, -0.077]\\
+\hspace{1em}Percent of private transportation & 0.01 & {}[0.001, 0.018] & -0.011 & {}[-0.021, -0.001]\\
+\hspace{1em}Median household income (\$10,000) & 0.126 & {}[0.094, 0.159] & 0.002 & {}[-0.033, 0.037]\\
+$\sigma$ & 0.954 & {}[0.904, 1.003] & 0.684 & {}[0.652, 0.716]\\
+$\rho$ & 0.971 & {}[0.961, 0.98] & -0.199 & {}[-0.377, -0.022]\\
 \bottomrule
@@
 \multicolumn{3}{l}{\textbf{Sample Selection Model}}\\
-\hspace{1em}Intercept & -7.304 & [-8.346, -6.262]\\
-\hspace{1em}Density (1,000 per sq.km) & 2.445 & [2.089, 2.802]\\
-\hspace{1em}Density squared & -0.380 & [-0.468, -0.292]\\
-\hspace{1em}Percent of private transportation & 0.056 & [0.046, 0.067]\\
-\hspace{1em}Median household income (10,000) & 0.318 & [0.28, 0.356]\\
+\hspace{1em}Intercept & -7.249 & {}[-8.285, -6.214]\\
+\hspace{1em}Density (1,000 per sq.km) & 2.424 & {}[2.074, 2.774]\\
+\hspace{1em}Density squared & -0.373 & {}[-0.459, -0.288]\\
+\hspace{1em}Percent of private transportation & 0.056 & {}[0.045, 0.066]\\
+\hspace{1em}Median household income (10,000) & 0.319 & {}[0.282, 0.356]\\
 \addlinespace[0.3em]
 \multicolumn{3}{l}{\textbf{Outcome Model}}\\
-\hspace{1em}Intercept & 2.563 & [2.497, 2.629]\\
-\hspace{1em}Density (1,000 per sq.km) & 0.760 & [0.746, 0.774]\\
-\hspace{1em}Density squared & -0.133 & [-0.135, -0.13]\\
-\hspace{1em}Percent of private transportation & -0.011 & [-0.012, -0.011]\\
-\hspace{1em}Median household income (\$10,000) & 0.002 & [-0.001, 0.004]\\
-\hspace{1em}Spatial filter & 1.000 & [0.998, 1.001]\\
-$\sigma$ & 0.017 & [0.015, 0.019]\\
-$\rho$ & -0.304 & [-0.957, 0.349]\\
+\hspace{1em}Intercept & 2.290 & {}[2.026, 2.553]\\
+\hspace{1em}Density (1,000 per sq.km) & 0.843 & {}[0.786, 0.9]\\
+\hspace{1em}Density squared & -0.142 & {}[-0.153, -0.131]\\
+\hspace{1em}Percent of private transportation & -0.010 & {}[-0.012, -0.008]\\
+\hspace{1em}Median household income (\$10,000) & 0.011 & {}[0.003, 0.02]\\
+\hspace{1em}Spatial filter & 1.001 & {}[0.992, 1.011]\\
+$\sigma$ & 0.120 & {}[0.107, 0.133]\\
+$\rho$ & 0.495 & {}[0.218, 0.772]\\
 \bottomrule
@@
 research.
+
+\hypertarget{acknwledgments}{%
+\section*{Acknowledgments}\label{acknwledgments}}
+\addcontentsline{toc}{section}{Acknowledgments}
+
+The analysis reported in this paper was conducted in the \texttt{R}
+computing statistical language (\textbf{R-base?}). The source document
+is an Rmarkdown document (\textbf{rmarkdown2018?};
+\textbf{rmarkdown2020?}) processed using \texttt{knitr}
+(\textbf{knitr2015?}; \textbf{knitr2014?}). The following packages were
+used in the analysis, and I wish to acknowledge their creators for their
+generous efforts: \texttt{adespatial} (\textbf{R-adespatial?}),
+\texttt{censReg} (\textbf{R-censReg?}), \texttt{dplyr}
+(\textbf{R-dplyr?}), \texttt{forcats} (\textbf{R-forcats?}),
+\texttt{ggplot2} (\textbf{ggplot22016?}), \texttt{gmm}
+(\textbf{gmm2010?}), \texttt{kableExtra} (\textbf{R-kableExtra?}),
+\texttt{Matrix} (\textbf{R-Matrix?}), \texttt{maxLik}
+(\textbf{maxLik2011?}), \texttt{miscTools} (\textbf{R-miscTools?}),
+\texttt{mvtnorm} (\textbf{mvtnorm2009?}), \texttt{nlme}
+(\textbf{R-nlme?}), \texttt{patchwork} (\textbf{R-patchwork?}),
+\texttt{purrr} (\textbf{R-purrr?}), \texttt{readr} (\textbf{R-readr?}),
+\texttt{sampleSelection} (\textbf{sampleSelection2008?}),
+\texttt{sandwich} (\textbf{sandwich2020?}; \textbf{sandwich2004?};
+\textbf{sandwich2006?}), \texttt{scico} (\textbf{R-scico?}), \texttt{sf}
+(\textbf{sf2018?}), \texttt{sp} (\textbf{sp2005?}; \textbf{sp2013?}),
+\texttt{spatialprobit} (\textbf{R-spatialprobit?}), \texttt{spData}
+(\textbf{R-spData?}), \texttt{spdep} (\textbf{spdep2018?};
+\textbf{spdep2013?}), \texttt{stringr} {[}R-stringr{]}, \texttt{tibble}
+(\textbf{R-tibble?}), \texttt{tidycensus} (\textbf{R-tidycensus?}),
+\texttt{tidyr} (\textbf{R-tidyr?}), \texttt{tidyverse}
+(\textbf{tidyverse2019?}), \texttt{tmvtnorm} (\textbf{R-tmvtnorm?}),
+\texttt{units} (\textbf{units2016?}). This research was not supported by
+Canada's Research Councils.
 

### 2021-11-18 README.Rmd
Reorganized the repository after paper's publication online at Geographical Analysis

@@
 
-Paper desk-rejected by [~~PLoS ONE~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/PLOS ONE Decision_ PONE-D-21-16874.pdf) [~~Science of the Total Environment~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Your Submission STOTEN-D-21-12077.pdf) [~~Landscape and Urban Planning~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Decision on submission to Landscape and Urban Planning.pdf)
-
-Forthcoming in Geographical Analysis
+Paper desk-rejected by [~~PLoS ONE~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Editorial-Materials-and-Correspondence/PLOS ONE Decision_ PONE-D-21-16874.pdf) [~~Science of the Total Environment~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Editorial-Materials-and-Correspondence/Your Submission STOTEN-D-21-12077.pdf) [~~Landscape and Urban Planning~~](https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed/tree/main/R0-Density-Reanalysis/Editorial-Materials-and-Correspondence/Decision on submission to Landscape and Urban Planning.pdf)
+
+Geographical Analysis (2021) https://doi.org/10.1111/gean.12307
 

### 2022-10-08 Junk/Junk.Rmd
Updated with final bibliographical information and created a renv reproducible environment

---
title: "Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis"
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    correspondingauthor: true
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
abstract: |
  The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic in 2019 led to explosive growth in scientific research. Alas, much of the research in the literature lacks conditions to be reproducible, and recent publications on the association between population density and the basic reproductive number of SARS-CoV-2 are no exception. Relatively few papers share code and data sufficiently, which hinders not only verification but additional experimentation. In this paper, an example of reproducible research shows the potential of spatial analysis for epidemiology research during COVID-19. Transparency and openness means that independent researchers can, with only modest efforts, verify findings and use different approaches as appropriate. Given the high stakes of the situation, it is essential that scientific findings, on which good policy depends, are as robust as possible; as the empirical example shows, reproducibility is one of the keys to ensure this.

  This paper is now published in Geographical Analysis (\url{https://doi.org/10.1111/gean.12307})
keywords: 
  - keyword1
  - keyword2
journal: "Geographical Analysis"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: bibliography.bib
linenumbers: true
numbersections: false
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

Please make sure that your manuscript follows the guidelines in the 
Guide for Authors of the relevant journal. It is not necessary to 
typeset your manuscript in exactly the same way as an article, 
unless you are submitting to a camera-ready copy (CRC) journal.

For detailed instructions regarding the elsevier article class, see   <https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions>

# References {-}

### 2022-10-08 Junk/Junk.tex
Updated with final bibliographical information and created a renv reproducible environment

--- 2021-09-16 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
@@
-\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
+\documentclass[preprint, 3p,
+authoryear]{elsarticle} %review=doublespace preprint=single 5p=2 column
 %%% Begin My package additions %%%%%%%%%%%%%%%%%%%
+
 \usepackage[hyphens]{url}
@@
   \linenumbers % turns line numbering on
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
 \usepackage{graphicx}
-\usepackage{booktabs} % book-quality tables
 %%%%%%%%%%%%%%%% end my additions to header
@@
 \IfFileExists{microtype.sty}{\usepackage{microtype}}{}
-\bibliographystyle{elsarticle-harv}
+\usepackage[]{natbib}
+\bibliographystyle{plainnat}
+
 \ifxetex
@@
             pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
 
@@
 
-% Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-% Pandoc header
-\usepackage{setspace}\onehalfspacing
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
 \begin{document}
+
 \begin{frontmatter}
@@
 perspective of spatial analysis}
-    \author[University]{Author\corref{1}}
-   \ead{author@institution.edu} 
-      \address[University]{Department, Street, City, State ZIP}
-      \cortext[1]{Corresponding Author}
+    \author[McMaster University]{Antonio Paez%
+  \corref{cor1}%
+  }
+   \ead{paezha@mcmaster.ca} 
+      \affiliation[McMaster University]{School of Earth, Environment and
+Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada}
+    \cortext[cor1]{Corresponding author}
 
@@
   shows, reproducibility is one of the keys to ensure this.
+
+  This paper is now published in Geographical Analysis
+  (\url{https://doi.org/10.1111/gean.12307})
   \end{abstract}
+    \begin{keyword}
+    keyword1 \sep 
+    keyword2
+  \end{keyword}
 
@@
 
-\newpage
+Please make sure that your manuscript follows the guidelines in the
+Guide for Authors of the relevant journal. It is not necessary to
+typeset your manuscript in exactly the same way as an article, unless
+you are submitting to a camera-ready copy (CRC) journal.
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+For detailed instructions regarding the elsevier article class, see
+\url{https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions}
 
-The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the
-global pandemic that followed in its wake, led to an explosive growth of
-research around the globe. According to Fraser et al. (2021), over
-125,000 COVID-19-related papers were released in the first ten months
-from the first confirmed case of the disease. Of these, more than 30,000
-were shared in pre-print servers, the use of which also exploded in the
-past year (Añazco et al., 2021; Kwon, 2020; Vlasschaert et al., 2020).
-
-Given the ruinous human and economic cost of the pandemic, there has
-been a natural tension in the scientific community between the need to
-publish research results quickly and the imperative to maintain
-consistently high quality standards in scientific reporting; indeed, a
-call for maintaining the standards in published research termed the
-deluge of COVID-19 publications a ``carnage of substandard research''
-(Bramstedt, 2020). Part of the challenge of maintaining quality
-standards in published research is that, despite an abundance of
-recommendations and guidelines (e.g., Broggini et al., 2017; Brunsdon
-and Comber, 2020; Ince et al., 2012; Ioannidis et al., 2014), in
-practice reproducibility has remained a lofty and somewhat aspirational
-goal (Konkol et al., 2019; Konkol and Kray, 2019). As reported in the
-literature, only a woefully small proportion of published research was
-actually reproducible before the pandemic (Iqbal et al., 2016; Stodden
-et al., 2018), and the situation does not appear to have changed
-substantially since (Gustot, 2020; Sumner et al., 2020).
-
-The push for open software and data (Arribas-Bel et al., 2021; e.g.,
-Bivand, 2020), along with more strenuous efforts towards open,
-reproducible research, is simply a continuation of long-standing
-scientific practices of independent verification. Despite the (at times
-disproportionate) attention that high profile scandals in science tend
-to elicit in the media, science as a collective endeavor is remarkable
-for being a self-correcting enterprise, one with built-in mechanisms and
-incentives to weed out erroneous ideas. Over the long term, facts tend
-to prevail in science. At stake is the shorter-term impacts that
-research may have in other spheres of economic and social life. The case
-of economists Reinhart and Rogoff comes to mind: by the time the
-inaccuracies and errors in their research were uncovered (see Herndon et
-al., 2014), their claims about debt and economic growth had already been
-seized by policy-makers on both sides of the Atlantic to justify
-austerity policies in the aftermath of the Great Recession of
-2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
-  ``Reinhart--Rogoff may have had more immediate influence on public
-  debate than any previous paper in the history of economics''
-  \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
-As later research has demonstrated, those policies cast a long shadow,
-and their sequels continued to be felt for years (Basu et al., 2017).
-
-In the context of COVID-19, a topic that has grabbed the imagination of
-numerous thinkers has been the prospect of life in cities after the
-pandemic (e.g., Florida et al., 2020); as a result, the implications of
-the pandemic for urban planning, design, and management are the topic of
-ongoing research (e.g., Sharifi and Khavarian-Garmsir, 2020). The fact
-that the worst of the pandemic was initially felt in dense population
-centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent
-of research into the associations between density and the spread of the
-pandemic. The answers to some important questions hang on the results of
-these research efforts. For example, are lower density regions safer
-from the pandemic? Are de-densification policies warranted, even if just
-in the short term? In the longer term, will the risks of life in high
-density regions presage a flight from cities? And, what are the
-implications of the pandemic for future urban planning and practice?
-Over the past year, numerous papers have sought to throw light on the
-underlying issue of density and the pandemic; nonetheless the results,
-as will be detailed next, remain mixed. Further, to complicate matters,
-precious few of these studies appear to be sufficiently open to support
-independent verification.
-
-The objective of this paper is to illustrate the importance of
-reproducibility in research in the context of the flood of COVID-19
-papers. For this, I focus on a recent study by Sy et al. (2021) that
-examined the correlation between the basic reproductive number of
-COVID-19, \(R_0\), and population density. The basic reproductive number
-is a summary measure of contact rates, probability of transmission of a
-pathogen, and duration of infectiousness. In rough terms, it measures
-how many new infections each infections begets. The paper of Sy et al.
-(2021) was selected for being, in the literature examined, almost alone
-in supporting reproducible research. Accordingly, I wish to be clear
-that my objective in singling their work for discussion is not to malign
-their efforts, but rather to demonstrate how open and reproducible
-research efforts can greatly help to accelerate discovery. More
-concretely, open data and open code mean that an independent researcher
-can, with only modest efforts, not only verify the findings reported,
-but also examine the same data from a perspective which may not have
-been available to the original researchers due to differences in
-disciplinary perspectives, methodological traditions, and/or training,
-among other possible factors. The example, which shows consequential
-changes in the conclusions reached by different analyses, should serve
-as a call to researchers to redouble their efforts to increase
-transparency and reproducibility in their research. In this spirit, the
-present paper also aims to show how data can be packaged in
-well-documented, shareable units, and code can be embedded into
-self-contained documents suitable for review and independent
-verification. The source for this paper is an
-\href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
-with the data package, will be available in a public
-repository\footnote{For peer-review purposes, the data package and code
-  are currently in an anonymous Drive folder:
-  \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}}.
-
-\hypertarget{background-the-intuitive-relationship-between-density-and-spread-of-contagious-diseases}{%
-\section{Background: the intuitive relationship between density and
-spread of contagious
-diseases}\label{background-the-intuitive-relationship-between-density-and-spread-of-contagious-diseases}}
-
-The concern with population density and the spread of the virus during
-the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
-seen in real-time around the world from large urban centers such as
-Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
-believe that higher density could have a positive association with the
-transmission of a contagious virus. It has long been known that the
-potential for inter-personal contact is greater in regions with higher
-density (see for example the research on urban fields and
-time-geography, including Farber and Páez, 2011; Moore, 1970; Moore and
-Brown, 1970). Mathematically, models of exposure and contagion indicate
-that higher densities can catalyze the transmission of contagious
-diseases (Li et al., 2018; Rocklöv and Sjödin, 2020). The idea is
-intuitive and likely at the root of messages, by some figures in
-positions of authority, that regions with sparse population densities
-faced lower risks from the pandemic\footnote{Governor Kristi Noem of
-  South Dakota, for example, claimed that sparse population density
-  allowed her state to face the pandemic down without the need for
-  strict policy interventions
-  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.
-
-As Rocklöv and Sjödin (Rocklöv and Sjödin, 2020) note, however,
-mathematical models of contagion are valid at small-to-medium spaces
-(and presumably, smaller time intervals too, such as time spent in
-restaurants, concert halls, cruises), and the results do not necessarily
-transfer to larger spatial units and longer time periods. There are
-solid reasons for this: while in a restaurant, one can hardly avoid
-being in proximity to other customers. On the other hand, a person can
-choose to (or be forced to as a matter of policy) not go to a restaurant
-in the first place. Nonetheless, the idea that high density correlates
-with high transmission is so seemingly sensible that it is often taken
-for granted even at the scale of large spaces (e.g., Cruz et al., 2020;
-Micallef et al., 2020). In such conditions, however, there exists the
-possibility of behavioral adaptations, which are difficult to capture in
-the mechanistic framework of differential equations (or can be missing
-in agent-based models, e.g., Gomez et al., 2021); these adaptations, in
-fact, can be a key aspect of disease transmission.
-
-A plausible behavioral adaptation during a pandemic, especially one
-broadcast as widely and intensely as COVID-19, is risk compensation.
-Risk compensation is a process whereby people adjust their behavior in
-response to their \emph{perception} of risk (Noland, 1995; Phillips et
-al., 2011; Richens et al., 2000). In the case of COVID-19, Chauhan et
-al. (Chauhan et al., 2021) have found that perception of risks in the US
-varies between rural, suburban, and urban residents, with rural
-residents in general expressing less concern about the virus. It is
-possible that people who listened to the message of leaders saying that
-they were safe from the virus because of low density may not have taken
-adequate precautions. Conversely, people in dense places who could more
-directly observe the impact of the pandemic may have become overly
-cautious. Both Paez et al. (2020) and Hamidi et al. (2020b) posit this
-mechanism (i.e., greater compliance with social distancing in denser
-regions) to explain the results of their analyses. The evidence
-available does indeed show that there were important changes in behavior
-with respect to mobility during the pandemic (Harris and Branion-Calles,
-2021; Jamal and Paez, 2020; Molloy et al., 2020); furthermore, shelter
-in place orders may have had greater buy-in from the public in higher
-density regions (Feyman et al., 2020; Hamidi and Zandiatashbar, 2021),
-and the associated behavior may have persisted beyond the duration of
-official social-distancing policies (Praharaj et al., 2020). In
-addition, there is evidence that changes in mobility correlated with the
-trajectory of the pandemic (Noland, 2021; Paez, 2020). Given the
-potential for behavioral adaptation, the question of density becomes
-more nuanced: it is not just a matter of proximity, but also of human
-behavior, which is better studied using population-level data and
-models.
-
-\hypertarget{background-but-what-does-the-literature-say}{%
-\section{Background: but what does the literature
-say?}\label{background-but-what-does-the-literature-say}}
-
-When it comes to population density and the spread of COVID-19, the
-international literature to date remains inconclusive.
-
-On the one hand, there are studies that report positive associations
-between population density and various COVID-19-related outcomes. Bhadra
-(2021), for example, reported a moderate positive correlation between
-the spread of COVID-19 and population density at the district level in
-India, however their analysis was bivariate and did not control for
-other variables, such as income. Similarly, Kadi and Khelfaoui (2020)
-found a positive and significant correlation between number of cases and
-population density in cities in Algeria in a series of simple regression
-models (i.e., without other controls). A question in these relatively
-simple analyses is whether density is not a proxy for other factors.
-Other studies have included controls, such as Pequeno et al. (2020), a
-team that reported a positive association between density and cumulative
-counts of confirmed COVID-19 cases in state capitals in Brazil after
-controlling for covariates, including income, transport connectivity,
-and economic status. In a similar vein, Fielding-Miller et al. (2020)
-reported a positive relationship between the absolute number of COVID-19
-deaths and population density (rate) in rural counties in the US. Roy
-and Ghosh (2020) used a battery of machine learning techniques to find
-discriminatory factors, and a positive and significant association
-between COVID-19 infection and death rates in US states. Wong and Li
-(2020) also found a positive and significant association between
-population density and number of confirmed COVID-19 cases in US
-counties, using both univariate and multivariate regressions with
-spatial effects. More recently, Sy et al. (2021) reported that the basic
-reproductive number of COVID-19 in US counties tended to increase with
-population density, but at a decreasing rate at higher densities.
-
-On the flip side, a number of studies report non-significant or negative
-associations between population density and COVID-19 outcomes. This
-includes the research of Sun et al. (2020) who did not find evidence of
-significant correlation between population density and confirmed number
-of cases per day \emph{in conditions of lockdown} in China. This finding
-echoes the results of Paez et al. (2020), who in their study of
-provinces in Spain reported non-significant associations between
-population density and infection rates in the early days of the first
-wave of COVID-19, and negative significant associations in the later
-part of the first lockdown. Similarly, Skórka et al. (2020) found zero
-or negative associations between population density and infection
-numbers/deaths by country. Fielding-Miller et al. (2020) contrast their
-finding about rural counties with a negative relationship between
-COVID-19 deaths and population density in urban counties in the US. For
-their part, in their investigation of doubling time, White and
-Hébert-Dufresne (2020) identified a negative and significant correlation
-between population density and doubling time in US states. Likewise,
-Khavarian-Garmsir et al. (2021) found a small negative (and significant)
-association between population density and COVID-19 morbidity in
-districts in Tehran. Finally, two of the most complete studies in the
-US, by Hamidi et al. (2020a) and Hamidi et al. (2020b), used an
-extensive set of controls to find negative and significant correlations
-between density and COVID-19 cases and fatalities at the level of
-counties in the US.
-
-As can be seen, these studies are implemented at different scales in
-different regions of the world. They also use a range of techniques,
-from correlation analysis, to multivariate regression, spatial
-regressions, and machine learning techniques. This is natural and to be
-expected: individual researchers have only limited time and expertise.
-This is why reproducibility is important. To pick an example (which will
-be further elaborated in later sections of this paper), the study of Sy
-et al. (2021), hereafter referred to as SWN, would immediately grab the
-attention of a researcher with expertise in spatial analysis.
-
-\hypertarget{reproducibility-of-research}{%
-\section{Reproducibility of
-research}\label{reproducibility-of-research}}
-
-SWN investigated the basic reproductive number of COVID-19 in US
-counties, and its association with population density, median household
-income, and prevalence of private mobility. For their multivariate
-analysis, SWN used mixed linear models. This is an appropriate modelling
-choice: \(R_0\) is an interval-ratio variable that is suitably modeled
-using linear regression; further, as SWN note there is a likelihood that
-the process in not independent ``among counties within each state,
-potentially due to variable resource allocation and differing health
-systems across states'' (p.~3). A mixed linear model accounts for this
-by introducing random components; in the case of SWN, these are random
-intercepts at the state level. SWN estimated various models with
-different combinations of variables, including median household income
-and prevalence of travel by private transportation. These controls help
-to account for potential variations in behavior: people in more affluent
-counties may have greater opportunities to work from home, and use of
-private transportation reduces contact with strangers. Moreover, they
-also conducted various sensitivity analyses. After these efforts, SWN
-concluded that there is a positive association between the basic
-reproductive number and population density at the level of counties in
-the US.
-
-One salient aspect of the analysis in SWN is that the basic reproductive
-number can only be calculated reliably with a minimum number of cases,
-and a large number of counties did not meet such threshold. As
-researchers do, SWN made modelling decisions, in this case basing their
-analysis only on counties with valid observations. A modeler with
-expertise in spatial analysis would likely ask some of the following
-questions on reading SWN's paper: how were missing counties treated?
-What are the implications of the spatial sampling framework used in the
-analysis? Is it possible to spatially interpolate the missing
-observations? Was there spatial residual autocorrelation in the models,
-or was the use of mixed models sufficient to capture spatial
-dependencies? These questions are relevant and their implications
-important. Fortunately, SWN are an example of a reasonably open,
-reproducible research product: their paper is accompanied by (most of)
-the data and (most of) the code used in the analysis. This means that an
-independent researcher can, with only a moderate investment of time and
-effort, reproduce the results in the paper, as well as ask additional
-questions.
-
-Alas, reproducibility is not necessarily the norm in the relevant
-literature.
-
-There are various reasons why a project can fail to be reproducible. In
-some cases, there might be legitimate reasons to withhold the data,
-perhaps due to confidentiality and privacy reasons (e.g., Lee et al.,
-2020). But in many other cases the data are publicly available, which in
-fact has commonly been the case with population-level COVID-19
-information. Typically the provenance of the data is documented, but in
-numerous studies the data themselves are not shared (Amadu et al., 2021;
-Bhadra et al., 2021; Cruz et al., 2020; Feng et al., 2020;
-Fielding-Miller et al., 2020; Hamidi et al., 2020a, 2020b; Inbaraj et
-al., 2021; Souris and Gonzalez, 2020). As any researcher can attest,
-collecting, organizing, and preparing data for a project can take a
-substantial amount of time. Pointing to the sources of data, even when
-these sources are public, is a small step towards reproducibility-but
-only a very small one. Faced with the prospect of having to recreate a
-data set from raw sources is probably sufficient to dissuade all but the
-most dedicated (or stubborn) researcher from independent verification.
-This is true even if part of the data are shared (e.g., Wong and Li,
-2020). In other cases, data are shared, but the processes followed in
-the preparation of the data are not fully documented (Ahmad et al.,
-2020; Skórka et al., 2020). These processes matter, as shown by the
-errors in the spreadsheets of Reinhart and Rogoff (see Herndon et al.,
-2014 for the discovery of these errors), as well as by the data of
-biologist Jonathan Pruitt that led to an ``avalanche'' of paper
-retractions (see Viglione, 2020). Another situation is when papers share
-well-documented data, but fail to provide the code used in the analysis
-(Noury et al., 2021; Pequeno et al., 2020; Wang et al., 2021). Making
-code available only ``on demand'' (e.g., Brandtner et al., 2021) is an
-unnecessary barrier when most journals offer the facility to share
-supplemental materials online. Then there are those papers that more
-closely comply with reproducibility standards, and share well-documented
-processes and data, as well as the code used in any analyses reported
-(Feyman et al., 2020; Paez et al., 2020; Stephens et al., 2021; Sy et
-al., 2021; White and Hébert-Dufresne, 2020). Even in this case, the
-pressure to publish ``new findings'' instead of replication studies can
-act as a deterrent, perhaps particularly for younger
-researchers\footnote{The present paper was desk rejected by three
-  journals that had previously published research on population density
-  and the spread of COVID-19; in one case, the paper was too opinionated
-  for the journal, in the other two cases, the paper was not a ``good
-  fit'' despite dealing with a nearly identical issue as papers
-  previously published in said journals.}.
-
-In the following sections, the analysis of SWN is reproduced, some
-relevant questions from the perspective of an independent researcher
-with expertise in spatial analysis are asked, and the data are
-reanalyzed.
-
-\hypertarget{reproducing-swn}{%
-\section{Reproducing SWN}\label{reproducing-swn}}
-
-SWN examined the association between the basic reproductive number of
-COVID-19 and population density. The basic reproductive number \(R_0\)
-is a summary measure of contact rates, probability of transmission of a
-pathogen, and duration of infectiousness. In rough terms, \(R_0\)
-measures how many new infections each infections begets. Infectious
-disease outbreaks generally tend to die out when \(R_0<1\), and to grow
-when \(R_0>1\). Reliable calculation of \(R_0\) requires a minimum
-number of cases to be able to assume that there is community
-transmission of the pathogen. Accordingly, SWN based their analysis only
-on counties that had at least 25 cases or more at the end of the
-exponential growth phase (see Fig. \ref{fig:R0-map}). Their final sample
-included 1,151 counties in the US, including in Alaska, Hawaii, Puerto
-Rico, and island territories. SWN used COVID-19 data collected by the
-New York Times and made available (with versioning) in a GitHub
-repository\footnote{https://github.com/nytimes/covid-19-data}. For each
-county, SWN assumed that the exponential growth period began one week
-prior to the second daily increase in cases, and assumed that the period
-of exponential growth lasted approximately 18 days.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/R0-map-1} \caption{\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown).}\label{fig:R0-map}
-\end{figure}
-
-Table \ref{tab:swn-results} reproduces the first three models of SWN
-(the fourth model did not have any significant variables; see Table 1 in
-SWN). It is possible to verify that the results match, with only the
-minor (and irrelevant) exception of the magnitude of the coefficient for
-travel by private transportation, which is due to a difference in the
-input (here the variable is changed to one percent units, instead of the
-ten percent units used by SWN). The mixed linear model gives random
-intercepts (i.e., the intercept is a random variable), and the standard
-deviation is reported in the fifth row of Table \ref{tab:swn-results}.
-It is useful to map the random intercepts: as seen in Figure
-\ref{fig:random-terms-map}, other things being equal, counties in Texas
-tend to have somewhat lower values of \(R_0\) (i.e., a negative random
-intercept), whereas counties in South Dakota tend to have higher values
-of \(R_0\). The key of the analysis, after extensive sensitivity
-analysis, is a robust finding that population density has a positive
-association with the basic reproductive number. But does it?
-
-\begin{table}
-
-\caption{\label{tab:tabulate-swn-results}\label{tab:swn-results}Reproducing SWN: Models 1-3}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllllrl}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Model 1} & \multicolumn{2}{c}{Model 2} & \multicolumn{2}{c}{Model 3} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
-Variable & beta & 95\% CI & beta & 95\% CI & beta & 95\% CI\\
-\midrule
-Intercept & 2.274 & [2.167, 2.381] & 3.347 & [2.676, 4.018] & 3.386 & [2.614, 4.157]\\
-Log of population density & 0.162 & [0.133, 0.191] & 0.145 & [0.115, 0.176] & 0.147 & [0.113, 0.18]\\
-Percent of private transportation &  &  & -0.013 & [-0.02, -0.005] & -0.013 & [-0.021, -0.005]\\
-Median household income (\$10,000) &  &  &  &  & -0.003 & [-0.033, 0.026]\\
-Standard deviation (Intercept) & 0.166 & [0.108, 0.254] & 0.136 & [0.081, 0.229] & 0.137 & [0.081, 0.232]\\
-\addlinespace
-Within-group standard error & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.693] & 0.665 & [0.638, 0.694]\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/random-terms-map-1} \caption{\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown).}\label{fig:random-terms-map}
-\end{figure}
-
-\hypertarget{expanding-on-swn}{%
-\section{Expanding on SWN}\label{expanding-on-swn}}
-
-The preceding section shows that thanks to the availability of code and
-data, it is possible to verify the results reported by SWN. As noted
-earlier, though, an independent researcher might have wondered about the
-implications of the spatial sampling procedure used by SWN. The decision
-to use a sample of counties with reliable basic reproductive numbers,
-although apparently sensible, results in a non-random spatial sampling
-scheme. Turning our attention back to Figure \ref{fig:R0-map}, we form
-the impression that many counties without reliable values of \(R_0\) are
-in more rural, less dense parts of the United States. This impression is
-reinforced when we overlay the boundaries of urban areas with population
-greater than 50,000 on the counties with valid values of \(R_0\) (see
-Figure \ref{fig:urban-areas-map}). The fact that \(R_0\) could not be
-accurately computed in many counties without large urban areas does not
-mean that there was no transmission of the virus: it simply means that
-we do not know with sufficient precision to what extent that was the
-case. The low number of cases may be related to low population and/or
-low population density. This is intriguing, to say the least: by
-excluding cases based on the ability to calculate \(R_0\) we are
-potentially \emph{selecting} the sample in a non-random way.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/urban-areas-map-1} \caption{\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown).}\label{fig:urban-areas-map}
-\end{figure}
-
-A problematic issue with non-random sample selection is that parameter
-estimates can become unreliable, and numerous techniques have been
-developed to address this. A model useful for sample selection problems
-is Heckman's selection model (see Maddala, 1983). The selection model is
-in fact a system of two equations, as follows: \[
-\begin{array}{c}
-y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
-y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
-\end{array}
-\] \noindent where \(y_i^{S*}\) is a latent variable for the sample
-selection process, and \(y_i^{O*}\) is the latent outcome. Vectors
-\(x_i^S\) and \(x_i^O\) are explanatory variables (with the possibility
-that \(x_i^S = x_i^S\)). Both equations include random terms (i.e.,
-\(\epsilon_i^S\) and \(\epsilon_i^O\)). The first equation is designed
-to model the \emph{probability} of sampling, and the second equation the
-outcome of interest (say \(R_0\)). The random terms are jointly
-distributed and correlated with parameter \(\rho\).
-
-What the analyst observes is the following: \[
-y_i^S =
-\begin{cases}
-0 & \text{if } y_i^{S*} < 0\\
-1 & \text{otherwise}
-\end{cases}
-\] \noindent and: \[
-y_i^O =
-\begin{cases}
-0 & \text{if } y_i^{S} = 0\\
-y_i^{O*} & \text{otherwise}
-\end{cases}
-\]
-
-In other words, the outcome of interest is observed \emph{only} for
-certain cases (\(y_i^S=1\), i.e., for sampled observations). The
-probability of sampling depends on \(x_i^S\). For the cases observed,
-the outcome \(y_i^O\) depends on \(x_i^O\).
-
-A sample selection model is estimated using the same selection of
-variables as SWN Model 3. This is Sample Selection Model 1 in Table
-\ref{tab:selection-results}. The first thing to notice about this model
-is that the sample selection process and the outcome are correlated
-(\(\rho\ne0\) with 5\% of confidence). The selection equation indicates
-that the probability of a county to be in the sample increases with
-population density (but at a decreasing rate due to the
-log-transformation), when travel by private modes is more prevalent, and
-as median household income in the county is higher. This is in line with
-the impression made by Figure \ref{fig:urban-areas-map} that counties
-with reliable values of \(R_0\) tended to be those with larger urban
-centers. Once that the selection probabilities are accounted for in the
-model, several things happen with the outcomes model. First, the
-coefficient for population density is still positive, but the magnitude
-changes: in effect, it appears that the effect of density is more
-pronounced than what SWN Model 3 indicated. The coefficient for percent
-of private transportation changes signs. And the coefficient for median
-household income is now significant.
-
-The second model in Table \ref{tab:selection-results} (Selection Model
-2) changes the way the variables are entered into the model. The
-log-transformation of density in SWN and Selection Model 1 assumes that
-the association between density and \(R_0\) is monotonically increasing
-(if the sign of the coefficient is positive) or decreasing (if the sign
-of the coefficient is negative). There are some indications that the
-relationship may actually not be monotonical. For example, Paez et al.
-(2020) found a positive (if non-significant) relationship between
-density and incidence of COVID-19 in the provinces of Spain at the
-beginning of the pandemic. This changed to a negative (and significant)
-relationship during the lockdown. In the case of the US, Fielding-Miller
-et al. (2020) found that the association between COVID-19 deaths and
-population density was positive in rural counties, but negative in urban
-counties. A variable transformation that allows for non-monotonic
-changes in the relationship is the square of the density.
-
-As seen in the table, Selection Model 2 replaces the log-transformation
-of population density with a quadratic expansion. The results of this
-analysis indicate that with this variable transformation, the selection
-and outcome processes are still correlated (\(\rho\ne0\) with 5\% of
-confidence). But a few other interesting things emerge. When we examine
-the outcomes model, we see that the quadratic expansion has a positive
-coefficient for the first order term, but a negative coefficient for the
-second order term. This indicates that \(R_0\) initially tends to
-increase as density grows, but only up to a point, after which the
-negative second term (which grows more rapidly due to the square),
-becomes increasingly dominant. Secondly, the sign of the coefficient for
-travel by private transportation becomes negative again. This, of
-course, makes more sense than the positive sign of Selection Model 1: if
-people tend to travel in private transportation, the potential for
-contact should be lower instead of higher. And finally median household
-income is no longer significant, similar to SWN Model 3.
-
-\begin{table}
-
-\caption{\label{tab:tabulate-sample-selection-results}\label{tab:selection-results}Estimation results of sample selection models}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lcccc}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Selection Model 1} & \multicolumn{2}{c}{Selection Model 2} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Variable & $\beta$ & 95\% CI & $\beta$ & 95\% CI\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{5}{l}{\textbf{Sample Selection Model}}\\
-\hspace{1em}Intercept & -2.237 & [-3.109, -1.365] & -7.339 & [-8.381, -6.297]\\
-\hspace{1em}Log of population density & 0.385 & [0.352, 0.418] &  & \\
-\hspace{1em}Density (1,000 per sq.km) &  &  & 2.484 & [2.13, 2.838]\\
-\hspace{1em}Density squared &  &  & -0.387 & [-0.473, -0.3]\\
-\hspace{1em}Percent of private transportation & 0.025 & [0.016, 0.034] & 0.057 & [0.046, 0.067]\\
-\hspace{1em}Median household income (10,000) & 0.202 & [0.168, 0.235] & 0.32 & [0.283, 0.357]\\
-\addlinespace[0.3em]
-\multicolumn{5}{l}{\textbf{Outcome Model}}\\
-\hspace{1em}Intercept & 0.605 & [-0.257, 1.466] & 2.784 & [1.652, 3.915]\\
-\hspace{1em}Log of population density & 0.39 & [0.354, 0.426] &  & \\
-\hspace{1em}Density (1,000 per sq.km) &  &  & 0.758 & [0.509, 1.008]\\
-\hspace{1em}Density squared &  &  & -0.132 & [-0.187, -0.077]\\
-\hspace{1em}Percent of private transportation & 0.01 & [0.001, 0.018] & -0.011 & [-0.021, -0.001]\\
-\hspace{1em}Median household income (\$10,000) & 0.126 & [0.094, 0.159] & 0.002 & [-0.033, 0.037]\\
-$\sigma$ & 0.954 & [0.904, 1.003] & 0.684 & [0.652, 0.716]\\
-$\rho$ & 0.971 & [0.961, 0.98] & -0.199 & [-0.377, -0.022]\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\hypertarget{proceed-with-caution-spatial-effects-ahead}{%
-\section{Proceed with caution: spatial effects
-ahead}\label{proceed-with-caution-spatial-effects-ahead}}
-
-The results of the selection models, in particular Selection Model 2,
-make us reassess the original conclusion that density has a positive
-association with the basic reproductive number of COVID-19. A spatial
-analyst might still wonder about spatial residual autocorrelation. A
-challenge here is that spatial models tend to be technically more
-demanding, and although spatial models for qualitative variables exist,
-a spatial implementation of the sample selection model does not appear
-to exist. It might be argued that a reproducible research project can
-also allow a researcher to be more adventurous with their modeling
-decisions: since data and code are shared, other researchers can
-promptly and with relative ease poke the methods and see if they appear
-to be sound.
-
-In the present case, it appears that an application of spatial filtering
-(see Getis and Griffith, 2002; Griffith, 2004; Paez, 2019) can help.
-Spatial filtering provides an elegant solution to regression problems
-that may have difficulties handling the spatial structures of spatial
-statistical and econometric models (Griffith, 2000). A key issue in the
-present example is the fact that there are numerous missing
-observations, which prevents the calculation of autocorrelation
-statistics, let alone the estimation of models with spatial components.
-
-The following is an unorthodox, but potentially effective use of filters
-in a sample selection model:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Estimate a sample selection model and retrieve the residuals of the
-  outcome. This will be a vector with missing values for locations that
-  were not sampled.
-\item
-  Fit a spatial filter to the residuals. This is done by regressing the
-  estimated residuals of the \emph{observed} data on the corresponding
-  values of the Moran eigenvectors.
-\item
-  The resulting filter will correlate highly with the known residuals,
-  and will provide information in non-sampled locations that is
-  consistent with the spatial pattern of the known residuals.
-\item
-  Test the filter for spatial autocorrelation:
-
-  4.1 If significant spatial autocorrelation is detected, this would be
-  indicative of residual spatial pattern. Introduce the filter as a
-  covariate in the outcome model of the sample selection model and
-  return to step 1.
-
-  4.2 If no significant spatial autocorrelation is detected, this would
-  be indicative of random residual pattern. Stop.
-\end{enumerate}
-
-This procedure is implemented using a stopping criterion whereby the
-search for the filter only stops when the p-value of Moran's Coefficient
-of the filter fitted to the residuals is greater than 0.25, which was
-chosen as a sufficiently conservative value for testing for
-autocorrelation. The correlation of the known residuals with the
-corresponding elements of the filter is consistently high (the
-correlation coefficient typically is greater than 0.9). The results of
-implementing this procedure appear in Table \ref{tab:selection3-results}
-as Selection Model 3. The results are consistent with Selection Model 2,
-with two intriguing differences: 1) the variance of Sample Model 3 is
-smaller; and 2) the sample and outcome processes are no longer
-correlated (the confidence interval of \(\rho\) includes zero). It
-appears that by capturing the spatial pattern of the residuals, which is
-likely strongly determined by the non-random sampling framework, the
-outcome model is not only substantially more precise, but also appears
-to be independent from the selection process.
-
-\begin{table}[!h]
-
-\caption{\label{tab:tabulate-sample-selection3-results}\label{tab:selection3-results}Estimation results of sample selection model with spatial filter}
-\centering
-\fontsize{8}{10}\selectfont
-\begin{tabular}[t]{lcc}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Selection Model 3} \\
-\cmidrule(l{3pt}r{3pt}){2-3}
-Variable & $\beta$ & 95\% CI\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{3}{l}{\textbf{Sample Selection Model}}\\
-\hspace{1em}Intercept & -7.304 & [-8.346, -6.262]\\
-\hspace{1em}Density (1,000 per sq.km) & 2.445 & [2.089, 2.802]\\
-\hspace{1em}Density squared & -0.380 & [-0.468, -0.292]\\
-\hspace{1em}Percent of private transportation & 0.056 & [0.046, 0.067]\\
-\hspace{1em}Median household income (10,000) & 0.318 & [0.28, 0.356]\\
-\addlinespace[0.3em]
-\multicolumn{3}{l}{\textbf{Outcome Model}}\\
-\hspace{1em}Intercept & 2.563 & [2.497, 2.629]\\
-\hspace{1em}Density (1,000 per sq.km) & 0.760 & [0.746, 0.774]\\
-\hspace{1em}Density squared & -0.133 & [-0.135, -0.13]\\
-\hspace{1em}Percent of private transportation & -0.011 & [-0.012, -0.011]\\
-\hspace{1em}Median household income (\$10,000) & 0.002 & [-0.001, 0.004]\\
-\hspace{1em}Spatial filter & 1.000 & [0.998, 1.001]\\
-$\sigma$ & 0.017 & [0.015, 0.019]\\
-$\rho$ & -0.304 & [-0.957, 0.349]\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Clearly, the various models display some intriguing differences; but how
-relevant are said differences from a more substantive standpoint? Figure
-\ref{fig:comparison-results} shows the relationship between density and
-\(R_0\) implied by SWN Model 3, Selection Model 2, and Selection Model
-3. The left panel of the figure shows the non-linear but monotonic
-relationship implied by SWN Model 1. The conclusion is that at higher
-densities, \(R_0\) is \emph{always} higher. The two panels on the right,
-in contrast, shows that Selection Model 2 and Selection Model 3 coincide
-that \(R_0\) tends to increase as density grows. This continues until a
-density of approximately 2.9 (1,000 people per sq.km). At higher
-densities than that the relationship between density and \(R_0\) begins
-to weaken, and the relationship becomes negative at densities higher
-than approximately 5.7 (1,000 people per sq.km).
-
-To put this into context, other things being equal, the effect of
-density in a county like Charlottesville in Virginia (density
-\textasciitilde1,639 people per sq.km) is roughly the same as that in a
-county like Philadelphia (density \textasciitilde4,127 people per
-sq.km). In contrast, the effect of density on \(R_0\) in a county like
-Arlington in Virginia (density \textasciitilde3,093 people per sq.km) is
-\emph{stronger} than either of the previous two examples. Lastly, the
-density of counties like San Francisco in California, or Queens and
-Bronx in NY, which are among the densest in the US, contributes even
-less to \(R_0\) than even the most rural counties in the country.
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{R0-Density-Reanalysis_files/figure-latex/comparison-results-1} \caption{\label{fig:comparison-results}Effect of density according to SWN Model 3 and Sample Selection Model 2.}\label{fig:comparison-results}
-\end{figure}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-It is worth at this point to recall Cressie's dictum about modelling:
-``{[}w{]}hat is one person's mean structure could be another person's
-correlation structure'' (Cressie, 1989, p. 201). There are almost always
-multiple ways to approach a modelling situation, as lively illustrated
-by a recent paper that reports the results of a crowdsourced modelling
-experiment (Schweinsberg et al., 2021). In the present case, we would
-argue that spatial sampling is an important aspect of the modeling
-process. Importantly, by adopting high reproducibility standards, SWN
-made a valuable contribution to the collective enterprise of seeking
-knowledge. Their effort, and subsequent efforts to validate and expand
-on their work, can potentially contribute to provide clarity to ongoing
-conversations about the relevance of density and the spread of COVID-19.
-
-In particular, it is noteworthy that a sample selection model with a
-different variable transformation does not lend support to the thesis
-that higher density is \emph{always} associated with a greater risk of
-spread of the virus {[}in Wong and Li's words, ``\,`Density is destiny'
-is probably an overstatement''; (2020){]}. At the same time, the results
-presented here also stand in contrast to the findings of Hamidi et al.,
-who found that higher density was either not significantly associated
-with the rate of the virus in a cross-sectional study (Hamidi et al.,
-2020b), or was negatively associated with it in a longitudinal setting
-{[}Hamidi et al. (2020a). In this sense, the conclusion that density
-does not aggravate the pandemic may have been somewhat premature;
-instead, reanalysis of the data of SWN suggests that Fielding-Miller et
-al. (2020) might be onto something with respect to the difference
-between rural and urban counties. More generally, there is no doubt that
-in population-level studies density is indicative of proximity, but it
-also potentially is a proxy for adaptive behavior. And it is possible
-that the determining factor during COVID-19, at least in the US, has
-been variations in perceptions of the risks associated with contagion
-(Chauhan et al., 2021), and subsequent compensations in behavior in more
-and less dense regions.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-The tension between the need to publish research potentially useful in
-dealing with a global pandemic, and a potential ``carnage of substandard
-research'' (Bramstedt, 2020), highlights the importance of efforts to
-maintain the quality of scientific outputs during COVID-19. An important
-part of quality control is the ability of independent researchers to
-verify and examine the results of materials published in the literature.
-As previous research illustrates, reproducibility in scientific research
-remains an important but elusive goal (Gustot, 2020; e.g., Iqbal et al.,
-2016; Stodden et al., 2018; Sumner et al., 2020). This idea is
-reinforced by the review conducted for this paper in the context of
-research about population density and the spread of COVID-19.
-
-Taking one recent example from the literature {[}Sy et al., Sy et al.
-(2021); SWN{]}, the present paper illustrates the importance of good
-reproducibility practices. Sharing data and code can catalyze research,
-by allowing independent verification of findings, as well as additional
-research. After verifying the results of SWN, experiments with sample
-selection models and variations in the definition of model inputs, lead
-to an important reappraisal of the conclusion that high density is
-associated with greater spread of the virus. Instead, the possibility of
-a non-monotonical relationship between population density and contagion
-is raised. I do not claim that the analysis presented here is the last
-word on the topic of density and the spread of COVID-19, and there is
-always the possibility that someone else will be better equipped to
-analyze these data with greater competence. By opening up the analysis,
-documenting the way data were pre-processed, and by sharing analysis
-ready data, my hope would be that others will be able to discover the
-limitations of my own analysis and improve on it, as appropriate.
-
-More generally, my hope is that the research of Sy et al. (2021), the
-present paper, and similar reproducible publications, will continue to
-encourage others to adopt higher reproducibility standards in their
-research.
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Ahmad2020association}{}%
-Ahmad, K., Erqou, S., Shah, N., Nazir, U., Morrison, A.R., Choudhary,
-G., Wu, W.-C., 2020. Association of poor housing conditions with
-COVID-19 incidence and mortality across US counties. PLOS ONE 15,
-e0241327.
-doi:\href{https://doi.org/10.1371/journal.pone.0241327}{10.1371/journal.pone.0241327}
-
-\leavevmode\hypertarget{ref-Amadu2021assessing}{}%
-Amadu, I., Ahinkorah, B.O., Afitiri, A.-R., Seidu, A.-A., Ameyaw, E.K.,
-Hagan, J.E., Duku, E., Aram, S.A., 2021. Assessing sub-regional-specific
-strengths of healthcare systems associated with COVID-19 prevalence,
-deaths and recoveries in africa. PLOS ONE 16, e0247274.
-doi:\href{https://doi.org/10.1371/journal.pone.0247274}{10.1371/journal.pone.0247274}
-
-\leavevmode\hypertarget{ref-Anazco2021publication}{}%
-Añazco, D., Nicolalde, B., Espinosa, I., Camacho, J., Mushtaq, M.,
-Gimenez, J., Teran, E., 2021. Publication rate and citation counts for
-preprints released during the COVID-19 pandemic: The good, the bad and
-the ugly. PeerJ 9, e10927.
-doi:\href{https://doi.org/10.7717/peerj.10927}{10.7717/peerj.10927}
-
-\leavevmode\hypertarget{ref-Arribas2021open}{}%
-Arribas-Bel, D., Green, M., Rowe, F., Singleton, A., 2021. Open data
-products: A framework for creating valuable analysis-ready data. Journal
-of Geographical Systems.
-
-\leavevmode\hypertarget{ref-Basu2017ten}{}%
-Basu, S., Carney, M.A., Kenworthy, N.J., 2017. Ten years after the
-financial crisis: The long reach of austerity and its global impacts on
-health. Social Science \& Medicine 187, 203--207.
-doi:\href{https://doi.org/10.1016/j.socscimed.2017.06.026}{10.1016/j.socscimed.2017.06.026}
-
-\leavevmode\hypertarget{ref-Bhadra2021impact}{}%
-Bhadra, A., Mukherjee, A., Sarkar, K., 2021. Impact of population
-density on covid-19 infected and mortality rate in india. Modeling Earth
-Systems and Environment 7, 623--629.
-doi:\href{https://doi.org/10.1007/s40808-020-00984-7}{10.1007/s40808-020-00984-7}
-
-\leavevmode\hypertarget{ref-Bivand2020progress}{}%
-Bivand, R.S., 2020. Progress in the r ecosystem for representing and
-handling spatial data. Journal of Geographical Systems.
-doi:\href{https://doi.org/10.1007/s10109-020-00336-0}{10.1007/s10109-020-00336-0}
-
-\leavevmode\hypertarget{ref-Bramstedt2020carnage}{}%
-Bramstedt, K.A., 2020. The carnage of substandard research during the
-COVID-19 pandemic: A call for quality. Journal of Medical Ethics 46,
-803--807.
-doi:\href{https://doi.org/10.1136/medethics-2020-106494}{10.1136/medethics-2020-106494}
-
-\leavevmode\hypertarget{ref-Brandtner2021creatures}{}%
-Brandtner, C., Bettencourt, L.M.A., Berman, M.G., Stier, A.J., 2021.
-Creatures of the state? Metropolitan counties compensated for state
-inaction in initial u.s. Response to COVID-19 pandemic. PLOS ONE 16,
-e0246249.
-doi:\href{https://doi.org/10.1371/journal.pone.0246249}{10.1371/journal.pone.0246249}
-
-\leavevmode\hypertarget{ref-Broggini2017reproducible}{}%
-Broggini, F., Dellinger, J., Fomel, S., Liu, Y., 2017. Reproducible
-research: Geophysics papers of the future - introduction. Geophysics 82.
-doi:\href{https://doi.org/10.1190/geo2017-0918-spseintro.1}{10.1190/geo2017-0918-spseintro.1}
-
-\leavevmode\hypertarget{ref-Brunsdon2020opening}{}%
-Brunsdon, C., Comber, A., 2020. Opening practice: Supporting
-reproducibility and critical spatial data science. Journal of
-Geographical Systems.
-doi:\href{https://doi.org/10.1007/s10109-020-00334-2}{10.1007/s10109-020-00334-2}
-
-\leavevmode\hypertarget{ref-Chauhan2021covid}{}%
-Chauhan, R.S., Capasso da Silva, D., Salon, D., Shamshiripour, A.,
-Rahimi, E., Sutradhar, U., Khoeini, S., Mohammadian, A.(Kouros).,
-Derrible, S., Pendyala, R., 2021. COVID-19 related attitudes and risk
-perceptions across urban, rural, and suburban areas in the united
-states. Findings.
-doi:\href{https://doi.org/10.32866/001c.23714}{10.32866/001c.23714}
-
-\leavevmode\hypertarget{ref-Cressie1989geostatistics}{}%
-Cressie, N., 1989. Geostatistics. The American Statistician 43, 197.
-doi:\href{https://doi.org/10.2307/2685361}{10.2307/2685361}
-
-\leavevmode\hypertarget{ref-Cruz2020exploring}{}%
-Cruz, C.J.P., Ganly, R., Li, Z., Gietel-Basten, S., 2020. Exploring the
-young demographic profile of COVID-19 cases in hong kong: Evidence from
-migration and travel history data. PLOS ONE 15, e0235306.
-doi:\href{https://doi.org/10.1371/journal.pone.0235306}{10.1371/journal.pone.0235306}
-
-\leavevmode\hypertarget{ref-Farber2011running}{}%
-Farber, S., Páez, A., 2011. Running to stay in place: The time-use
-implications of automobile oriented land-use and travel. Journal of
-Transport Geography 19, 782--793.
-doi:\href{https://doi.org/10.1016/j.jtrangeo.2010.09.008}{10.1016/j.jtrangeo.2010.09.008}
-
-\leavevmode\hypertarget{ref-Feng2020spread}{}%
-Feng, Y., Li, Q., Tong, X., Wang, R., Zhai, S., Gao, C., Lei, Z., Chen,
-S., Zhou, Y., Wang, J., Yan, X., Xie, H., Chen, P., Liu, S., Xv, X.,
-Liu, S., Jin, Y., Wang, C., Hong, Z., Luan, K., Wei, C., Xu, J., Jiang,
-H., Xiao, C., Guo, Y., 2020. Spatiotemporal spread pattern of the
-COVID-19 cases in china. PLOS ONE 15, e0244351.
-doi:\href{https://doi.org/10.1371/journal.pone.0244351}{10.1371/journal.pone.0244351}
-
-\leavevmode\hypertarget{ref-Feyman2020effectiveness}{}%
-Feyman, Y., Bor, J., Raifman, J., Griffith, K.N., 2020. Effectiveness of
-COVID-19 shelter-in-place orders varied by state. PLOS ONE 15, e0245008.
-doi:\href{https://doi.org/10.1371/journal.pone.0245008}{10.1371/journal.pone.0245008}
-
-\leavevmode\hypertarget{ref-Fielding2020social}{}%
-Fielding-Miller, R.K., Sundaram, M.E., Brouwer, K., 2020. Social
-determinants of COVID-19 mortality at the county level. PLOS ONE 15,
-e0240151.
-doi:\href{https://doi.org/10.1371/journal.pone.0240151}{10.1371/journal.pone.0240151}
-
-\leavevmode\hypertarget{ref-Florida2020how}{}%
-Florida, R., Glaeser, E., Sharif, M., Bedi, K., Campanella, T., Chee,
-C., Doctoroff, D., Katz, B., Katz, R., Kotkin, J., 2020. How life in our
-cities will look after the coronavirus pandemic. Foreign Policy 1.
-
-\leavevmode\hypertarget{ref-Fraser2021evolving}{}%
-Fraser, N., Brierley, L., Dey, G., Polka, J.K., Pálfy, M., Nanni, F.,
-Coates, J.A., 2021. The evolving role of preprints in the dissemination
-of COVID-19 research and their impact on the science communication
-landscape. PLOS Biology 19, e3000959.
-doi:\href{https://doi.org/10.1371/journal.pbio.3000959}{10.1371/journal.pbio.3000959}
-
-\leavevmode\hypertarget{ref-Getis2002comparative}{}%
-Getis, A., Griffith, D.A., 2002. Comparative spatial filtering in
-regression analysis. Geographical Analysis 34, 130--140.
-
-\leavevmode\hypertarget{ref-Gomez2021infekta}{}%
-Gomez, J., Prieto, J., Leon, E., Rodríguez, A., 2021. INFEKTA---an
-agent-based model for transmission of infectious diseases: The COVID-19
-case in bogotá, colombia. PLOS ONE 16, e0245787.
-doi:\href{https://doi.org/10.1371/journal.pone.0245787}{10.1371/journal.pone.0245787}
-
-\leavevmode\hypertarget{ref-Griffith2000linear}{}%
-Griffith, D.A., 2000. A linear regression solution to the spatial
-autocorrelation problem. Journal of Geographical Systems 2, 141--156.
-
-\leavevmode\hypertarget{ref-Griffith2004spatial}{}%
-Griffith, D.A., 2004. A spatial filtering specification for the
-autologistic model. Environment and Planning A 36, 1791--1811.
-
-\leavevmode\hypertarget{ref-Gustot2020quality}{}%
-Gustot, T., 2020. Quality and reproducibility during the COVID-19
-pandemic. JHEP Rep 2, 100141.
-doi:\href{https://doi.org/10.1016/j.jhepr.2020.100141}{10.1016/j.jhepr.2020.100141}
-
-\leavevmode\hypertarget{ref-Hamidi2020longitudinal}{}%
-Hamidi, S., Ewing, R., Sabouri, S., 2020a. Longitudinal analyses of the
-relationship between development density and the COVID-19 morbidity and
-mortality rates: Early evidence from 1,165 metropolitan counties in the
-united states. Health \& Place 64, 102378.
-doi:\href{https://doi.org/10.1016/j.healthplace.2020.102378}{10.1016/j.healthplace.2020.102378}
-
-\leavevmode\hypertarget{ref-Hamidi2020density}{}%
-Hamidi, S., Sabouri, S., Ewing, R., 2020b. Does density aggravate the
-COVID-19 pandemic? Journal of the American Planning Association 86,
-495--509.
-doi:\href{https://doi.org/10.1080/01944363.2020.1777891}{10.1080/01944363.2020.1777891}
-
-\leavevmode\hypertarget{ref-Hamidi2021compact}{}%
-Hamidi, S., Zandiatashbar, A., 2021. Compact development and adherence
-to stay-at-home order during the COVID-19 pandemic: A longitudinal
-investigation in the united states. Landscape and Urban Planning 205,
-103952. doi:\url{https://doi.org/10.1016/j.landurbplan.2020.103952}
-
-\leavevmode\hypertarget{ref-Harris2021Changes}{}%
-Harris, M.A., Branion-Calles, M., 2021. Changes in commute mode
-attributed to COVID-19 risk in canadian national survey data. Findings.
-doi:\href{https://doi.org/10.32866/001c.19088}{10.32866/001c.19088}
-
-\leavevmode\hypertarget{ref-Herndon2014high}{}%
-Herndon, T., Ash, M., Pollin, R., 2014. Does high public debt
-consistently stifle economic growth? A critique of reinhart and rogoff.
-Cambridge Journal of Economics 38, 257--279.
-doi:\href{https://doi.org/10.1093/cje/bet075}{10.1093/cje/bet075}
-
-\leavevmode\hypertarget{ref-Inbaraj2021seroprevalence}{}%
-Inbaraj, L.R., George, C.E., Chandrasingh, S., 2021. Seroprevalence of
-COVID-19 infection in a rural district of south india: A
-population-based seroepidemiological study. PLOS ONE 16, e0249247.
-doi:\href{https://doi.org/10.1371/journal.pone.0249247}{10.1371/journal.pone.0249247}
-
-\leavevmode\hypertarget{ref-Ince2012case}{}%
-Ince, D.C., Hatton, L., Graham-Cumming, J., 2012. The case for open
-computer programs. Nature 482, 485--488.
-doi:\href{https://doi.org/10.1038/nature10836}{10.1038/nature10836}
-
-\leavevmode\hypertarget{ref-Ioannidis2014increasing}{}%
-Ioannidis, J.P.A., Greenland, S., Hlatky, M.A., Khoury, M.J., Macleod,
-M.R., Moher, D., Schulz, K.F., Tibshirani, R., 2014. Increasing value
-and reducing waste in research design, conduct, and analysis. Lancet
-383, 166--175.
-doi:\href{https://doi.org/10.1016/s0140-6736(13)62227-8}{10.1016/s0140-6736(13)62227-8}
-
-\leavevmode\hypertarget{ref-Iqbal2016reproducible}{}%
-Iqbal, S.A., Wallach, J.D., Khoury, M.J., Schully, S.D., Ioannidis,
-J.P.A., 2016. Reproducible research practices and transparency across
-the biomedical literature. Plos Biology 14.
-doi:\href{https://doi.org/10.1371/journal.pbio.1002333}{10.1371/journal.pbio.1002333}
-
-\leavevmode\hypertarget{ref-Jamal2020Changes}{}%
-Jamal, S., Paez, A., 2020. Changes in trip-making frequency by mode
-during COVID-19. Findings.
-doi:\href{https://doi.org/10.32866/001c.17977}{10.32866/001c.17977}
-
-\leavevmode\hypertarget{ref-Kadi2020population}{}%
-Kadi, N., Khelfaoui, M., 2020. Population density, a factor in the
-spread of COVID-19 in algeria: Statistic study. Bulletin of the National
-Research Centre 44.
-doi:\href{https://doi.org/10.1186/s42269-020-00393-x}{10.1186/s42269-020-00393-x}
-
-\leavevmode\hypertarget{ref-Khavarian2021high}{}%
-Khavarian-Garmsir, A.R., Sharifi, A., Moradpour, N., 2021. Are
-high-density districts more vulnerable to the COVID-19 pandemic?
-Sustainable Cities and Society 70, 102911.
-doi:\href{https://doi.org/10.1016/j.scs.2021.102911}{10.1016/j.scs.2021.102911}
-
-\leavevmode\hypertarget{ref-Konkol2019examination}{}%
-Konkol, M., Kray, C., 2019. In-depth examination of spatiotemporal
-figures in open reproducible research. Cartography and Geographic
-Information Science 46, 412--427.
-doi:\href{https://doi.org/10.1080/15230406.2018.1512421}{10.1080/15230406.2018.1512421}
-
-\leavevmode\hypertarget{ref-Konkol2019computational}{}%
-Konkol, M., Kray, C., Pfeiffer, M., 2019. Computational reproducibility
-in geoscientific papers: Insights from a series of studies with
-geoscientists and a reproduction study. International Journal of
-Geographical Information Science 33, 408--429.
-doi:\href{https://doi.org/10.1080/13658816.2018.1508687}{10.1080/13658816.2018.1508687}
-
-\leavevmode\hypertarget{ref-Kwon2021swamped}{}%
-Kwon, D., 2020. How swamped preprint servers are blocking bad
-coronavirus research. Nature 581, 130--132.
-
-\leavevmode\hypertarget{ref-Lee2020human}{}%
-Lee, M., Zhao, J., Sun, Q., Pan, Y., Zhou, W., Xiong, C., Zhang, L.,
-2020. Human mobility trends during the early stage of the COVID-19
-pandemic in the united states. PLOS ONE 15, e0241468.
-doi:\href{https://doi.org/10.1371/journal.pone.0241468}{10.1371/journal.pone.0241468}
-
-\leavevmode\hypertarget{ref-Li2018effect}{}%
-Li, R., Richmond, P., Roehner, B.M., 2018. Effect of population density
-on epidemics. Physica A: Statistical Mechanics and its Applications 510,
-713--724.
-doi:\href{https://doi.org/10.1016/j.physa.2018.07.025}{10.1016/j.physa.2018.07.025}
-
-\leavevmode\hypertarget{ref-Maddala1983limited}{}%
-Maddala, G.S., 1983. Limited-dependent and qualitative variables in
-econometrics. Cambridge University Press, Cambridge.
-
-\leavevmode\hypertarget{ref-Micallef2020first}{}%
-Micallef, S., Piscopo, T.V., Casha, R., Borg, D., Vella, C., Zammit,
-M.-A., Borg, J., Mallia, D., Farrugia, J., Vella, S.M., Xerri, T.,
-Portelli, A., Fenech, M., Fsadni, C., Mallia Azzopardi, C., 2020. The
-first wave of COVID-19 in malta; a national cross-sectional study. PLOS
-ONE 15, e0239389.
-doi:\href{https://doi.org/10.1371/journal.pone.0239389}{10.1371/journal.pone.0239389}
-
-\leavevmode\hypertarget{ref-Molloy2020Tracing}{}%
-Molloy, J., Tchervenkov, C., Hintermann, B., Axhausen, K.W., 2020.
-Tracing the sars-CoV-2 impact: The first month in switzerland. Findings.
-doi:\href{https://doi.org/10.32866/001c.12903}{10.32866/001c.12903}
-
-\leavevmode\hypertarget{ref-Moore1970some}{}%
-Moore, E.G., 1970. Some spatial properties of urban contact fields.
-Geographical Analysis 2, 376--386.
-
-\leavevmode\hypertarget{ref-Moore1970urban}{}%
-Moore, E.G., Brown, L.A., 1970. Urban acquaintance fields: An evaluation
-of a spatial model. Environment and Planning 2, 443--454.
-
-\leavevmode\hypertarget{ref-Noland1995perceived}{}%
-Noland, R.B., 1995. PERCEIVED RISK AND MODAL CHOICE - RISK COMPENSATION
-IN TRANSPORTATION SYSTEM. Accident Analysis and Prevention 27, 503--521.
-doi:\href{https://doi.org/10.1016/0001-4575(94)00087-3}{10.1016/0001-4575(94)00087-3}
-
-\leavevmode\hypertarget{ref-Noland2021mobility}{}%
-Noland, R.B., 2021. Mobility and the effective reproduction rate of
-COVID-19. Journal of Transport \& Health 20, 101016.
-doi:\url{https://doi.org/10.1016/j.jth.2021.101016}
-
-\leavevmode\hypertarget{ref-Noury2021how}{}%
-Noury, A., François, A., Gergaud, O., Garel, A., 2021. How does COVID-19
-affect electoral participation? Evidence from the french municipal
-elections. PLOS ONE 16, e0247026.
-doi:\href{https://doi.org/10.1371/journal.pone.0247026}{10.1371/journal.pone.0247026}
-
-\leavevmode\hypertarget{ref-Paez2019using}{}%
-Paez, A., 2019. Using spatial filters and exploratory data analysis to
-enhance regression models of spatial data. Geographical Analysis 51,
-314--338.
-doi:\href{https://doi.org/10.1111/gean.12180}{10.1111/gean.12180}
-
-\leavevmode\hypertarget{ref-Paez2020using}{}%
-Paez, A., 2020. Using google community mobility reports to investigate
-the incidence of COVID-19 in the united states. Findings.
-doi:\url{https://doi.org/10.32866/001c.12976}
-
-\leavevmode\hypertarget{ref-Paez2020spatio}{}%
-Paez, A., Lopez, F.A., Menezes, T., Cavalcanti, R., Pitta, M.G. da R.,
-2020. A spatio-temporal analysis of the environmental correlates of
-COVID-19 incidence in spain. Geographical Analysis n/a.
-doi:\href{https://doi.org/10.1111/gean.12241}{10.1111/gean.12241}
-
-\leavevmode\hypertarget{ref-Pequeno2020air}{}%
-Pequeno, P., Mendel, B., Rosa, C., Bosholn, M., Souza, J.L., Baccaro,
-F., Barbosa, R., Magnusson, W., 2020. Air transportation, population
-density and temperature predict the spread of COVID-19 in brazil. PeerJ
-8, e9322.
-doi:\href{https://doi.org/10.7717/peerj.9322}{10.7717/peerj.9322}
-
-\leavevmode\hypertarget{ref-Phillips2011risk}{}%
-Phillips, R.O., Fyhri, A., Sagberg, F., 2011. Risk compensation and
-bicycle helmets. Risk Analysis 31, 1187--1195.
-doi:\href{https://doi.org/10.1111/j.1539-6924.2011.01589.x}{10.1111/j.1539-6924.2011.01589.x}
-
-\leavevmode\hypertarget{ref-Praharaj2020Using}{}%
-Praharaj, S., King, D., Pettit, C., Wentz, E., 2020. Using aggregated
-mobility data to measure the effect of COVID-19 policies on mobility
-changes in sydney, london, phoenix, and pune. Findings.
-doi:\href{https://doi.org/10.32866/001c.17590}{10.32866/001c.17590}
-
-\leavevmode\hypertarget{ref-Richens2000condoms}{}%
-Richens, J., Imrie, J., Copas, A., 2000. Condoms and seat belts: The
-parallels and the lessons. Lancet 355, 400--403.
-doi:\href{https://doi.org/10.1016/s0140-6736(99)09109-6}{10.1016/s0140-6736(99)09109-6}
-
-\leavevmode\hypertarget{ref-Rocklov2020high}{}%
-Rocklöv, J., Sjödin, H., 2020. High population densities catalyse the
-spread of COVID-19. Journal of Travel Medicine 27.
-doi:\href{https://doi.org/10.1093/jtm/taaa038}{10.1093/jtm/taaa038}
-
-\leavevmode\hypertarget{ref-Roy2020factors}{}%
-Roy, S., Ghosh, P., 2020. Factors affecting COVID-19 infected and death
-rates inform lockdown-related policymaking. PLOS ONE 15, e0241165.
-doi:\href{https://doi.org/10.1371/journal.pone.0241165}{10.1371/journal.pone.0241165}
-
-\leavevmode\hypertarget{ref-Schweinsberg2021same}{}%
-Schweinsberg, M., Feldman, M., Staub, N., van den Akker, O.R., van Aert,
-R.C.M., van Assen, M.A.L.M., Liu, Y., Althoff, T., Heer, J., Kale, A.,
-Mohamed, Z., Amireh, H., Venkatesh Prasad, V., Bernstein, A., Robinson,
-E., Snellman, K., Amy Sommer, S., Otner, S.M.G., Robinson, D., Madan,
-N., Silberzahn, R., Goldstein, P., Tierney, W., Murase, T., Mandl, B.,
-Viganola, D., Strobl, C., Schaumans, C.B.C., Kelchtermans, S., Naseeb,
-C., Mason Garrison, S., Yarkoni, T., Richard Chan, C.S., Adie, P.,
-Alaburda, P., Albers, C., Alspaugh, S., Alstott, J., Nelson, A.A., Ariño
-de la Rubia, E., Arzi, A., Bahník, Š., Baik, J., Winther Balling, L.,
-Banker, S., AA Baranger, D., Barr, D.J., Barros-Rivera, B., Bauer, M.,
-Blaise, E., Boelen, L., Bohle Carbonell, K., Briers, R.A., Burkhard, O.,
-Canela, M.-A., Castrillo, L., Catlett, T., Chen, O., Clark, M., Cohn,
-B., Coppock, A., Cugueró-Escofet, N., Curran, P.G., Cyrus-Lai, W., Dai,
-D., Valentino Dalla Riva, G., Danielsson, H., F. S. M. Russo, R. de, de
-Silva, N., Derungs, C., Dondelinger, F., Duarte de Souza, C., Tyson
-Dube, B., Dubova, M., Mark Dunn, B., Adriaan Edelsbrunner, P., Finley,
-S., Fox, N., Gnambs, T., Gong, Y., Grand, E., Greenawalt, B., Han, D.,
-Hanel, P.H.P., Hong, A.B., Hood, D., Hsueh, J., Huang, L., Hui, K.N.,
-Hultman, K.A., Javaid, A., Ji Jiang, L., Jong, J., Kamdar, J., Kane, D.,
-Kappler, G., Kaszubowski, E., Kavanagh, C.M., Khabsa, M., Kleinberg, B.,
-Kouros, J., Krause, H., Krypotos, A.-M., Lavbič, D., Ling Lee, R.,
-Leffel, T., Yang Lim, W., Liverani, S., Loh, B., Lønsmann, D., Wei Low,
-J., Lu, A., MacDonald, K., Madan, C.R., Hjorth Madsen, L., Maimone, C.,
-Mangold, A., Marshall, A., Ester Matskewich, H., Mavon, K., McLain,
-K.L., McNamara, A.A., McNeill, M., Mertens, U., Miller, D., Moore, B.,
-Moore, A., Nantz, E., Nasrullah, Z., Nejkovic, V., Nell, C.S., Arthur
-Nelson, A., Nilsonne, G., Nolan, R., O'Brien, C.E., O'Neill, P., O'Shea,
-K., Olita, T., Otterbacher, J., Palsetia, D., Pereira, B., Pozdniakov,
-I., Protzko, J., Reyt, J.-N., Riddle, T., (Akmal) Ridhwan Omar Ali, A.,
-Ropovik, I., Rosenberg, J.M., Rothen, S., Schulte-Mecklenbeck, M.,
-Sharma, N., Shotwell, G., Skarzynski, M., Stedden, W., Stodden, V.,
-Stoffel, M.A., Stoltzman, S., Subbaiah, S., Tatman, R., Thibodeau, P.H.,
-Tomkins, S., Valdivia, A., Druijff-van de Woestijne, G.B., Viana, L.,
-Villesèche, F., Duncan Wadsworth, W., Wanders, F., Watts, K., Wells,
-J.D., Whelpley, C.E., Won, A., Wu, L., Yip, A., Youngflesh, C., Yu,
-J.-C., Zandian, A., Zhang, L., Zibman, C., Luis Uhlmann, E., 2021. Same
-data, different conclusions: Radical dispersion in empirical results
-when independent analysts operationalize and test the same hypothesis.
-Organizational Behavior and Human Decision Processes 165, 228--249.
-doi:\url{https://doi.org/10.1016/j.obhdp.2021.02.003}
-
-\leavevmode\hypertarget{ref-Sharifi2020covid}{}%
-Sharifi, A., Khavarian-Garmsir, A.R., 2020. The COVID-19 pandemic:
-Impacts on cities and major lessons for urban planning, design, and
-management. Science of The Total Environment 749, 142391.
-doi:\url{https://doi.org/10.1016/j.scitotenv.2020.142391}
-
-\leavevmode\hypertarget{ref-Skorka2020macroecology}{}%
-Skórka, P., Grzywacz, B., Moroń, D., Lenda, M., 2020. The macroecology
-of the COVID-19 pandemic in the anthropocene. PLOS ONE 15, e0236856.
-doi:\href{https://doi.org/10.1371/journal.pone.0236856}{10.1371/journal.pone.0236856}
-
-\leavevmode\hypertarget{ref-Souris2020covid}{}%
-Souris, M., Gonzalez, J.-P., 2020. COVID-19: Spatial analysis of
-hospital case-fatality rate in france. PLOS ONE 15, e0243606.
-doi:\href{https://doi.org/10.1371/journal.pone.0243606}{10.1371/journal.pone.0243606}
-
-\leavevmode\hypertarget{ref-Stephens2021impact}{}%
-Stephens, K.E., Chernyavskiy, P., Bruns, D.R., 2021. Impact of altitude
-on COVID-19 infection and death in the united states: A modeling and
-observational study. PLOS ONE 16, e0245055.
-doi:\href{https://doi.org/10.1371/journal.pone.0245055}{10.1371/journal.pone.0245055}
-
-\leavevmode\hypertarget{ref-Stodden2018empirical}{}%
-Stodden, V., Seiler, J., Ma, Z.K., 2018. An empirical analysis of
-journal policy effectiveness for computational reproducibility.
-Proceedings of the National Academy of Sciences of the United States of
-America 115, 2584--2589.
-doi:\href{https://doi.org/10.1073/pnas.1708290115}{10.1073/pnas.1708290115}
-
-\leavevmode\hypertarget{ref-Sumner2020reproducibility}{}%
-Sumner, J., Haynes, L., Nathan, S., Hudson-Vitale, C., McIntosh, L.D.,
-2020. Reproducibility and reporting practices in COVID-19 preprint
-manuscripts. medRxiv 2020.03.24.20042796.
-doi:\href{https://doi.org/10.1101/2020.03.24.20042796}{10.1101/2020.03.24.20042796}
-
-\leavevmode\hypertarget{ref-Sun2020impacts}{}%
-Sun, Z., Zhang, H., Yang, Y., Wan, H., Wang, Y., 2020. Impacts of
-geographic factors and population density on the COVID-19 spreading
-under the lockdown policies of china. Science of The Total Environment
-746, 141347.
-doi:\href{https://doi.org/10.1016/j.scitotenv.2020.141347}{10.1016/j.scitotenv.2020.141347}
-
-\leavevmode\hypertarget{ref-Sy2021population}{}%
-Sy, K.T.L., White, L.F., Nichols, B.E., 2021. Population density and
-basic reproductive number of COVID-19 across united states counties.
-PLOS ONE 16, e0249271.
-doi:\href{https://doi.org/10.1371/journal.pone.0249271}{10.1371/journal.pone.0249271}
-
-\leavevmode\hypertarget{ref-Viglione2020avalanche}{}%
-Viglione, G., 2020. 'Avalanche'of spider-paper retractions shakes
-behavioural-ecology community. Nature 578, 199--201.
-
-\leavevmode\hypertarget{ref-Vlasschaert2020proliferation}{}%
-Vlasschaert, C., Topf, J.M., Hiremath, S., 2020. Proliferation of papers
-and preprints during the coronavirus disease 2019 pandemic: Progress or
-problems with peer review? Advances in Chronic Kidney Disease 27,
-418--426.
-doi:\href{https://doi.org/10.1053/j.ackd.2020.08.003}{10.1053/j.ackd.2020.08.003}
-
-\leavevmode\hypertarget{ref-Wang2021transmission}{}%
-Wang, F., Tan, Z., Yu, Z., Yao, S., Guo, C., 2021. Transmission and
-control pressure analysis of the COVID-19 epidemic situation using
-multisource spatio-temporal big data. PLOS ONE 16, e0249145.
-doi:\href{https://doi.org/10.1371/journal.pone.0249145}{10.1371/journal.pone.0249145}
-
-\leavevmode\hypertarget{ref-White2020state}{}%
-White, E.R., Hébert-Dufresne, L., 2020. State-level variation of initial
-COVID-19 dynamics in the united states. PLOS ONE 15, e0240648.
-doi:\href{https://doi.org/10.1371/journal.pone.0240648}{10.1371/journal.pone.0240648}
-
-\leavevmode\hypertarget{ref-Wong2020spreading}{}%
-Wong, D.W.S., Li, Y., 2020. Spreading of COVID-19: Density matters. PLOS
-ONE 15, e0242398.
-doi:\href{https://doi.org/10.1371/journal.pone.0242398}{10.1371/journal.pone.0242398}
-
-\end{CSLReferences}
+\renewcommand\refname{References}
+\bibliography{bibliography.bib}
 

### 2022-10-08 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Updated with final bibliographical information and created a renv reproducible environment

--- 2021-11-01 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
@@
 author:
-  - name: Author
-    email: author@institution.edu
-    affiliation: University
-    footnote: 1
+  - name: Antonio Paez
+    email: paezha@mcmaster.ca
+    affiliation: McMaster University
+    correspondingauthor: true
 address:
-  - code: University
-    address: Department, Street, City, State ZIP
-footnote:
-  - code: 1
-    text: "Corresponding Author"
+  - code: McMaster University
+    address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
 abstract: |
@@
 
+  This paper is now published in Geographical Analysis (\url{https://doi.org/10.1111/gean.12307})
+keywords: 
+  - keyword1
+  - keyword2
 journal: "Geographical Analysis"
-date: "`r Sys.Date()`"
-bibliography: [bibliography.bib, packages.bib]
+date: "Final form published in Volume 54, Issue 4"
+classoption: preprint, 3p, authoryear
+bibliography: bibliography.bib
 linenumbers: true
-#numbersections: true
-csl: elsevier-harvard.csl
-output: rticles::elsevier_article
-header-includes:
-    - \usepackage{setspace}\onehalfspacing
+numbersections: false
+# Use a CSL with `citation_package = "default"`
+# csl: https://www.zotero.org/styles/elsevier-harvard
+output: 
+  rticles::elsevier_article:
+    keep_tex: true
+    citation_package: natbib
 ---
@@
 knitr::opts_chunk$set(cache = TRUE)
-```
-
-```{r install-data-package, include=FALSE}
-if(!require(r0density)){
-    remotes::install_github("paezha/Reproductive-Rate-and-Density-US-Reanalyzed", 
-                         subdir = "r0density")
-    library(r0density)
-}
 ```
@@
 library(censReg) # Package to estimate tobit model
+library(here) # A Simpler Way to Find Your Files
 library(kableExtra) # Package for creating tables
@@
 library(patchwork) # Package to organize ggplot2 plots
-library(r0density) # Data package
+library(rticles) # Article Formats for R Markdown
 library(sampleSelection) # Package to estimate sampling selection models
@@
 library(units) # Package to work with units of measurement
-#library(viridis) # Color palettes
-```
-
-```{r generate-package-bibliography, include=FALSE}
-knitr::write_bib(c(.packages(), 
-                   "knitr", 
-                   "rmarkdown"), 
-                 file = "packages.bib")
+```
+
+```{r install-data-package, include=FALSE}
+if(!require(r0density)){
+    install.packages(paste0(here(), "/r0density_0.1.0.tar.gz"), repos = NULL, type="source")
+    library(r0density) # not installed on this machine
+}
+```
+
+```{r load-data-package, include=FALSE, cache=FALSE}
+library(r0density) # Data package
 ```

### 2022-10-08 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Updated with final bibliographical information and created a renv reproducible environment

@@
 
-% Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-% Pandoc header
-\usepackage{setspace}\onehalfspacing
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
 \usepackage{booktabs}
@@
 \begin{document}
+
 \begin{frontmatter}
@@
 perspective of spatial analysis}
-    \author[McMaster University]{Antonio Paez\corref{1}}
+    \author[McMaster University]{Antonio Paez%
+  \corref{cor1}%
+  }
    \ead{paezha@mcmaster.ca} 
-      \address[McMaster University]{School of Earth, Environment and
+      \affiliation[McMaster University]{School of Earth, Environment and
 Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada}
-      \cortext[1]{Corresponding Author}
+    \cortext[cor1]{Corresponding author}
 
@@
   \end{abstract}
+    \begin{keyword}
+    keyword1 \sep 
+    keyword2
+  \end{keyword}
 
@@
 global pandemic that followed in its wake, led to an explosive growth of
-research around the globe. According to Fraser et al. (2021), over
-125,000 COVID-19-related papers were released in the first ten months
-from the first confirmed case of the disease. Of these, more than 30,000
-were shared in pre-print servers, the use of which also exploded in the
-past year (Añazco et al., 2021; Kwon, 2020; Vlasschaert et al., 2020).
+research around the globe. According to Fraser et al.
+\citeyearpar{Fraser2021evolving}, over 125,000 COVID-19-related papers
+were released in the first ten months from the first confirmed case of
+the disease. Of these, more than 30,000 were shared in pre-print
+servers, the use of which also exploded in the past year
+\citep{Kwon2021swamped, Vlasschaert2020proliferation, Anazco2021publication}.
 
@@
 deluge of COVID-19 publications a ``carnage of substandard research''
-(Bramstedt, 2020). Part of the challenge of maintaining quality
-standards in published research is that, despite an abundance of
-recommendations and guidelines (e.g., Broggini et al., 2017; Brunsdon
-and Comber, 2020; Ince et al., 2012; Ioannidis et al., 2014), in
-practice reproducibility has remained a lofty and somewhat aspirational
-goal (Konkol et al., 2019; Konkol and Kray, 2019). As reported in the
-literature, only a woefully small proportion of published research was
-actually reproducible before the pandemic (Iqbal et al., 2016; Stodden
-et al., 2018), and the situation does not appear to have changed
-substantially since (Gustot, 2020; Sumner et al., 2020).
-
-The push for open software and data (Arribas-Bel et al., 2021; e.g.,
-Bivand, 2020), along with more strenuous efforts towards open,
-reproducible research, is simply a continuation of long-standing
-scientific practices of independent verification. Despite the (at times
-disproportionate) attention that high profile scandals in science tend
-to elicit in the media, science as a collective endeavor is remarkable
-for being a self-correcting enterprise, one with built-in mechanisms and
-incentives to weed out erroneous ideas. Over the long term, facts tend
-to prevail in science. At stake is the shorter-term impacts that
-research may have in other spheres of economic and social life. The case
-of economists Reinhart and Rogoff comes to mind: by the time the
-inaccuracies and errors in their research were uncovered (see Herndon et
-al., 2014), their claims about debt and economic growth had already been
-seized by policy-makers on both sides of the Atlantic to justify
-austerity policies in the aftermath of the Great Recession of
-2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
-  ``Reinhart--Rogoff may have had more immediate influence on public
-  debate than any previous paper in the history of economics''
+\citep{Bramstedt2020carnage}. Part of the challenge of maintaining
+quality standards in published research is that, despite an abundance of
+recommendations and guidelines
+\citep[e.g.,][]{Broggini2017reproducible, Ince2012case, Ioannidis2014increasing, Brunsdon2020opening},
+in practice reproducibility has remained a lofty and somewhat
+aspirational goal
+\citep{Konkol2019examination, Konkol2019computational}. As reported in
+the literature, only a woefully small proportion of published research
+was actually reproducible before the pandemic
+\citep{Iqbal2016reproducible, Stodden2018empirical}, and the situation
+does not appear to have changed substantially since
+\citep{Sumner2020reproducibility, Gustot2020quality}.
+
+The push for open software and data
+\citep[e.g.,][]{Bivand2020progress, Arribas2021open}, along with more
+strenuous efforts towards open, reproducible research, is simply a
+continuation of long-standing scientific practices of independent
+verification. Despite the (at times disproportionate) attention that
+high profile scandals in science tend to elicit in the media, science as
+a collective endeavor is remarkable for being a self-correcting
+enterprise, one with built-in mechanisms and incentives to weed out
+erroneous ideas. Over the long term, facts tend to prevail in science.
+At stake is the shorter-term impacts that research may have in other
+spheres of economic and social life. The case of economists Reinhart and
+Rogoff comes to mind: by the time the inaccuracies and errors in their
+research were uncovered \citep[see][]{Herndon2014high}, their claims
+about debt and economic growth had already been seized by policy-makers
+on both sides of the Atlantic to justify austerity policies in the
+aftermath of the Great Recession of 2007-2009\footnote{Nobel Prize in
+  Economics Paul Krugman noted that ``Reinhart--Rogoff may have had more
+  immediate influence on public debate than any previous paper in the
+  history of economics''
   \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
 As later research has demonstrated, those policies cast a long shadow,
-and their sequels continued to be felt for years (Basu et al., 2017).
+and their sequels continued to be felt for years \citep{Basu2017ten}.
 
@@
 numerous thinkers has been the prospect of life in cities after the
-pandemic (e.g., Florida et al., 2020); as a result, the implications of
-the pandemic for urban planning, design, and management are the topic of
-ongoing research (e.g., Sharifi and Khavarian-Garmsir, 2020). The fact
-that the worst of the pandemic was initially felt in dense population
-centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent
-of research into the associations between density and the spread of the
+pandemic \citep[e.g.,][]{Florida2020how}; as a result, the implications
+of the pandemic for urban planning, design, and management are the topic
+of ongoing research \citep[e.g.,][]{Sharifi2020covid}. The fact that the
+worst of the pandemic was initially felt in dense population centers
+such as Wuhan, Milan, Madrid, and New York, unleashed a torrent of
+research into the associations between density and the spread of the
 pandemic. The answers to some important questions hang on the results of
@@
 reproducibility in research in the context of the flood of COVID-19
-papers. For this, I focus on a recent study by Sy et al. (2021) that
-examined the correlation between the basic reproductive number of
-COVID-19, \(R_0\), and population density. The basic reproductive number
-is a summary measure of contact rates, probability of transmission of a
-pathogen, and duration of infectiousness. In rough terms, it measures
-how many new infections each infections begets. The paper of Sy et al.
-(2021) was selected for being, in the literature examined, almost alone
-in supporting reproducible research. Accordingly, I wish to be clear
-that my objective in singling their work for discussion is not to malign
+papers. For this, I focus on a recent study by Sy et al.
+\citeyearpar{Sy2021population} that examined the correlation between the
+basic reproductive number of COVID-19, \(R_0\), and population density.
+The basic reproductive number is a summary measure of contact rates,
+probability of transmission of a pathogen, and duration of
+infectiousness. In rough terms, it measures how many new infections each
+infections begets. The paper of Sy et al. \citeyearpar{Sy2021population}
+was selected for being, in the literature examined, almost alone in
+supporting reproducible research. Accordingly, I wish to be clear that
+my objective in singling their work for discussion is not to malign
 their efforts, but rather to demonstrate how open and reproducible
@@
 potential for inter-personal contact is greater in regions with higher
-density (see for example the research on urban fields and
-time-geography, including Farber and Páez, 2011; Moore, 1970; Moore and
-Brown, 1970). Mathematically, models of exposure and contagion indicate
-that higher densities can catalyze the transmission of contagious
-diseases (Li et al., 2018; Rocklöv and Sjödin, 2020). The idea is
-intuitive and likely at the root of messages, by some figures in
-positions of authority, that regions with sparse population densities
-faced lower risks from the pandemic\footnote{Governor Kristi Noem of
-  South Dakota, for example, claimed that sparse population density
-  allowed her state to face the pandemic down without the need for
-  strict policy interventions
+density \citep[see for example the research on urban fields and
+time-geography,
+including][]{Farber2011running, Moore1970urban, Moore1970some}.
+Mathematically, models of exposure and contagion indicate that higher
+densities can catalyze the transmission of contagious diseases
+\citep{Rocklov2020high, Li2018effect}. The idea is intuitive and likely
+at the root of messages, by some figures in positions of authority, that
+regions with sparse population densities faced lower risks from the
+pandemic\footnote{Governor Kristi Noem of South Dakota, for example,
+  claimed that sparse population density allowed her state to face the
+  pandemic down without the need for strict policy interventions
   \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.
 
-As Rocklöv and Sjödin (Rocklöv and Sjödin, 2020) note, however,
+As Rocklöv and Sjödin \citep{Rocklov2020high} note, however,
 mathematical models of contagion are valid at small-to-medium spaces
@@
 with high transmission is so seemingly sensible that it is often taken
-for granted even at the scale of large spaces (e.g., Cruz et al., 2020;
-Micallef et al., 2020). In such conditions, however, there exists the
-possibility of behavioral adaptations, which are difficult to capture in
-the mechanistic framework of differential equations (or can be missing
-in agent-based models, e.g., Gomez et al., 2021); these adaptations, in
-fact, can be a key aspect of disease transmission.
+for granted even at the scale of large spaces
+\citep[e.g.,][]{Cruz2020exploring, Micallef2020first}. In such
+conditions, however, there exists the possibility of behavioral
+adaptations, which are difficult to capture in the mechanistic framework
+of differential equations \citep[or can be missing in agent-based
+models, e.g.,][]{Gomez2021infekta}; these adaptations, in fact, can be a
+key aspect of disease transmission.
 
@@
 Risk compensation is a process whereby people adjust their behavior in
-response to their \emph{perception} of risk (Noland, 1995; Phillips et
-al., 2011; Richens et al., 2000). In the case of COVID-19, Chauhan et
-al. (Chauhan et al., 2021) have found that perception of risks in the US
-varies between rural, suburban, and urban residents, with rural
-residents in general expressing less concern about the virus. It is
-possible that people who listened to the message of leaders saying that
-they were safe from the virus because of low density may not have taken
-adequate precautions. Conversely, people in dense places who could more
-directly observe the impact of the pandemic may have become overly
-cautious. Both Paez et al. (2020) and Hamidi et al. (2020b) posit this
-mechanism (i.e., greater compliance with social distancing in denser
-regions) to explain the results of their analyses. The evidence
-available does indeed show that there were important changes in behavior
-with respect to mobility during the pandemic (Harris and Branion-Calles,
-2021; Jamal and Paez, 2020; Molloy et al., 2020); furthermore, shelter
-in place orders may have had greater buy-in from the public in higher
-density regions (Feyman et al., 2020; Hamidi and Zandiatashbar, 2021),
-and the associated behavior may have persisted beyond the duration of
-official social-distancing policies (Praharaj et al., 2020). In
-addition, there is evidence that changes in mobility correlated with the
-trajectory of the pandemic (Noland, 2021; Paez, 2020). Given the
+response to their \emph{perception} of risk
+\citep{Noland1995perceived, Richens2000condoms, Phillips2011risk}. In
+the case of COVID-19, Chauhan et al. \citep{Chauhan2021covid} have found
+that perception of risks in the US varies between rural, suburban, and
+urban residents, with rural residents in general expressing less concern
+about the virus. It is possible that people who listened to the message
+of leaders saying that they were safe from the virus because of low
+density may not have taken adequate precautions. Conversely, people in
+dense places who could more directly observe the impact of the pandemic
+may have become overly cautious. Both Paez et al.
+\citeyearpar{Paez2020spatio} and Hamidi et al.
+\citeyearpar{Hamidi2020density} posit this mechanism (i.e., greater
+compliance with social distancing in denser regions) to explain the
+results of their analyses. The evidence available does indeed show that
+there were important changes in behavior with respect to mobility during
+the pandemic
+\citep{Jamal2020Changes, Harris2021Changes, Molloy2020Tracing};
+furthermore, shelter in place orders may have had greater buy-in from
+the public in higher density regions
+\citep{Feyman2020effectiveness, Hamidi2021compact}, and the associated
+behavior may have persisted beyond the duration of official
+social-distancing policies \citep{Praharaj2020Using}. In addition, there
+is evidence that changes in mobility correlated with the trajectory of
+the pandemic \citep{Paez2020using, Noland2021mobility}. Given the
 potential for behavioral adaptation, the question of density becomes
@@
 between population density and various COVID-19-related outcomes. Bhadra
-(2021), for example, reported a moderate positive correlation between
-the spread of COVID-19 and population density at the district level in
-India, however their analysis was bivariate and did not control for
-other variables, such as income. Similarly, Kadi and Khelfaoui (2020)
-found a positive and significant correlation between number of cases and
+\citeyearpar{Bhadra2021impact}, for example, reported a moderate
+positive correlation between the spread of COVID-19 and population
+density at the district level in India, however their analysis was
+bivariate and did not control for other variables, such as income.
+Similarly, Kadi and Khelfaoui \citeyearpar{Kadi2020population} found a
+positive and significant correlation between number of cases and
 population density in cities in Algeria in a series of simple regression
@@
 simple analyses is whether density is not a proxy for other factors.
-Other studies have included controls, such as Pequeno et al. (2020), a
-team that reported a positive association between density and cumulative
-counts of confirmed COVID-19 cases in state capitals in Brazil after
-controlling for covariates, including income, transport connectivity,
-and economic status. In a similar vein, Fielding-Miller et al. (2020)
+Other studies have included controls, such as Pequeno et al.
+\citeyearpar{Pequeno2020air}, a team that reported a positive
+association between density and cumulative counts of confirmed COVID-19
+cases in state capitals in Brazil after controlling for covariates,
+including income, transport connectivity, and economic status. In a
+similar vein, Fielding-Miller et al. \citeyearpar{Fielding2020social}
 reported a positive relationship between the absolute number of COVID-19
 deaths and population density (rate) in rural counties in the US. Roy
-and Ghosh (2020) used a battery of machine learning techniques to find
-discriminatory factors, and a positive and significant association
-between COVID-19 infection and death rates in US states. Wong and Li
-(2020) also found a positive and significant association between
-population density and number of confirmed COVID-19 cases in US
-counties, using both univariate and multivariate regressions with
-spatial effects. More recently, Sy et al. (2021) reported that the basic
-reproductive number of COVID-19 in US counties tended to increase with
-population density, but at a decreasing rate at higher densities.
+and Ghosh \citeyearpar{Roy2020factors} used a battery of machine
+learning techniques to find discriminatory factors, and a positive and
+significant association between COVID-19 infection and death rates in US
+states. Wong and Li \citeyearpar{Wong2020spreading} also found a
+positive and significant association between population density and
+number of confirmed COVID-19 cases in US counties, using both univariate
+and multivariate regressions with spatial effects. More recently, Sy et
+al. \citeyearpar{Sy2021population} reported that the basic reproductive
+number of COVID-19 in US counties tended to increase with population
+density, but at a decreasing rate at higher densities.
 
@@
 associations between population density and COVID-19 outcomes. This
-includes the research of Sun et al. (2020) who did not find evidence of
-significant correlation between population density and confirmed number
-of cases per day \emph{in conditions of lockdown} in China. This finding
-echoes the results of Paez et al. (2020), who in their study of
-provinces in Spain reported non-significant associations between
-population density and infection rates in the early days of the first
-wave of COVID-19, and negative significant associations in the later
-part of the first lockdown. Similarly, Skórka et al. (2020) found zero
-or negative associations between population density and infection
-numbers/deaths by country. Fielding-Miller et al. (2020) contrast their
-finding about rural counties with a negative relationship between
-COVID-19 deaths and population density in urban counties in the US. For
-their part, in their investigation of doubling time, White and
-Hébert-Dufresne (2020) identified a negative and significant correlation
-between population density and doubling time in US states. Likewise,
-Khavarian-Garmsir et al. (2021) found a small negative (and significant)
-association between population density and COVID-19 morbidity in
-districts in Tehran. Finally, two of the most complete studies in the
-US, by Hamidi et al. (2020a) and Hamidi et al. (2020b), used an
-extensive set of controls to find negative and significant correlations
-between density and COVID-19 cases and fatalities at the level of
-counties in the US.
+includes the research of Sun et al. \citeyearpar{Sun2020impacts} who did
+not find evidence of significant correlation between population density
+and confirmed number of cases per day \emph{in conditions of lockdown}
+in China. This finding echoes the results of Paez et al.
+\citeyearpar{Paez2020spatio}, who in their study of provinces in Spain
+reported non-significant associations between population density and
+infection rates in the early days of the first wave of COVID-19, and
+negative significant associations in the later part of the first
+lockdown. Similarly, \citet{Skorka2020macroecology} found zero or
+negative associations between population density and infection
+numbers/deaths by country. Fielding-Miller et al.
+\citeyearpar{Fielding2020social} contrast their finding about rural
+counties with a negative relationship between COVID-19 deaths and
+population density in urban counties in the US. For their part, in their
+investigation of doubling time, White and Hébert-Dufresne
+\citeyearpar{White2020state} identified a negative and significant
+correlation between population density and doubling time in US states.
+Likewise, \citet{Khavarian2021high} found a small negative (and
+significant) association between population density and COVID-19
+morbidity in districts in Tehran. Finally, two of the most complete
+studies in the US, by \citet{Hamidi2020longitudinal} and
+\citet{Hamidi2020density}, used an extensive set of controls to find
+negative and significant correlations between density and COVID-19 cases
+and fatalities at the level of counties in the US.
 
@@
 be further elaborated in later sections of this paper), the study of Sy
-et al. (2021), hereafter referred to as SWN, would immediately grab the
-attention of a researcher with expertise in spatial analysis.
+et al. \citeyearpar{Sy2021population}, hereafter referred to as SWN,
+would immediately grab the attention of a researcher with expertise in
+spatial analysis.
 
@@
 some cases, there might be legitimate reasons to withhold the data,
-perhaps due to confidentiality and privacy reasons (e.g., Lee et al.,
-2020). But in many other cases the data are publicly available, which in
-fact has commonly been the case with population-level COVID-19
-information. Typically the provenance of the data is documented, but in
-numerous studies the data themselves are not shared (Amadu et al., 2021;
-Bhadra et al., 2021; Cruz et al., 2020; Feng et al., 2020;
-Fielding-Miller et al., 2020; Hamidi et al., 2020a, 2020b; Inbaraj et
-al., 2021; Souris and Gonzalez, 2020). As any researcher can attest,
-collecting, organizing, and preparing data for a project can take a
-substantial amount of time. Pointing to the sources of data, even when
-these sources are public, is a small step towards reproducibility-but
-only a very small one. Faced with the prospect of having to recreate a
-data set from raw sources is probably sufficient to dissuade all but the
-most dedicated (or stubborn) researcher from independent verification.
-This is true even if part of the data are shared (e.g., Wong and Li,
-2020). In other cases, data are shared, but the processes followed in
-the preparation of the data are not fully documented (Ahmad et al.,
-2020; Skórka et al., 2020). These processes matter, as shown by the
-errors in the spreadsheets of Reinhart and Rogoff (see Herndon et al.,
-2014 for the discovery of these errors), as well as by the data of
-biologist Jonathan Pruitt that led to an ``avalanche'' of paper
-retractions (see Viglione, 2020). Another situation is when papers share
-well-documented data, but fail to provide the code used in the analysis
-(Noury et al., 2021; Pequeno et al., 2020; Wang et al., 2021). Making
-code available only ``on demand'' (e.g., Brandtner et al., 2021) is an
-unnecessary barrier when most journals offer the facility to share
+perhaps due to confidentiality and privacy reasons
+\citep[e.g.,][]{Lee2020human}. But in many other cases the data are
+publicly available, which in fact has commonly been the case with
+population-level COVID-19 information. Typically the provenance of the
+data is documented, but in numerous studies the data themselves are not
+shared
+\citep{Amadu2021assessing, Bhadra2021impact, Cruz2020exploring, Feng2020spread, Fielding2020social, Hamidi2020longitudinal, Hamidi2020density, Inbaraj2021seroprevalence, Souris2020covid}.
+As any researcher can attest, collecting, organizing, and preparing data
+for a project can take a substantial amount of time. Pointing to the
+sources of data, even when these sources are public, is a small step
+towards reproducibility-but only a very small one. Faced with the
+prospect of having to recreate a data set from raw sources is probably
+sufficient to dissuade all but the most dedicated (or stubborn)
+researcher from independent verification. This is true even if part of
+the data are shared \citep[e.g.,][]{Wong2020spreading}. In other cases,
+data are shared, but the processes followed in the preparation of the
+data are not fully documented
+\citep{Ahmad2020association, Skorka2020macroecology}. These processes
+matter, as shown by the errors in the spreadsheets of Reinhart and
+Rogoff \citep[see][ for the discovery of these errors]{Herndon2014high},
+as well as by the data of biologist Jonathan Pruitt that led to an
+``avalanche'' of paper retractions \citep[see][]{Viglione2020avalanche}.
+Another situation is when papers share well-documented data, but fail to
+provide the code used in the analysis
+\citep{Noury2021how, Pequeno2020air, Wang2021transmission}. Making code
+available only ``on demand'' \citep[e.g.,][]{Brandtner2021creatures} is
+an unnecessary barrier when most journals offer the facility to share
 supplemental materials online. Then there are those papers that more
@@
 processes and data, as well as the code used in any analyses reported
-(Feyman et al., 2020; Paez et al., 2020; Stephens et al., 2021; Sy et
-al., 2021; White and Hébert-Dufresne, 2020). Even in this case, the
-pressure to publish ``new findings'' instead of replication studies can
-act as a deterrent, perhaps particularly for younger
-researchers\footnote{The present paper was desk rejected by three
-  journals that had previously published research on population density
-  and the spread of COVID-19; in one case, the paper was too opinionated
-  for the journal, in the other two cases, the paper was not a ``good
-  fit'' despite dealing with a nearly identical issue as papers
+\citep{Paez2020spatio, Feyman2020effectiveness, Stephens2021impact, White2020state, Sy2021population}.
+Even in this case, the pressure to publish ``new findings'' instead of
+replication studies can act as a deterrent, perhaps particularly for
+younger researchers\footnote{The present paper was desk rejected by
+  three journals that had previously published research on population
+  density and the spread of COVID-19; in one case, the paper was too
+  opinionated for the journal, in the other two cases, the paper was not
+  a ``good fit'' despite dealing with a nearly identical issue as papers
   previously published in said journals.}.
@@
 developed to address this. A model useful for sample selection problems
-is Heckman's selection model (see Maddala, 1983). The selection model is
-in fact a system of two equations, as follows: \[
+is Heckman's selection model \citep[see][]{Maddala1983limited}. The
+selection model is in fact a system of two equations, as follows: \[
 \begin{array}{c}
@@
 relationship may actually not be monotonical. For example, Paez et al.
-(2020) found a positive (if non-significant) relationship between
-density and incidence of COVID-19 in the provinces of Spain at the
-beginning of the pandemic. This changed to a negative (and significant)
-relationship during the lockdown. In the case of the US, Fielding-Miller
-et al. (2020) found that the association between COVID-19 deaths and
-population density was positive in rural counties, but negative in urban
-counties. A variable transformation that allows for non-monotonic
-changes in the relationship is the square of the density.
+\citeyearpar{Paez2020spatio} found a positive (if non-significant)
+relationship between density and incidence of COVID-19 in the provinces
+of Spain at the beginning of the pandemic. This changed to a negative
+(and significant) relationship during the lockdown. In the case of the
+US, Fielding-Miller et al. \citeyearpar{Fielding2020social} found that
+the association between COVID-19 deaths and population density was
+positive in rural counties, but negative in urban counties. A variable
+transformation that allows for non-monotonic changes in the relationship
+is the square of the density.
 
@@
 In the present case, it appears that an application of spatial filtering
-(see Getis and Griffith, 2002; Griffith, 2004; Paez, 2019) can help.
-Spatial filtering provides an elegant solution to regression problems
-that may have difficulties handling the spatial structures of spatial
-statistical and econometric models (Griffith, 2000). A key issue in the
-present example is the fact that there are numerous missing
-observations, which prevents the calculation of autocorrelation
+\citep[see][]{Getis2002comparative, Paez2019using, Griffith2004spatial}
+can help. Spatial filtering provides an elegant solution to regression
+problems that may have difficulties handling the spatial structures of
+spatial statistical and econometric models \citep{Griffith2000linear}. A
+key issue in the present example is the fact that there are numerous
+missing observations, which prevents the calculation of autocorrelation
 statistics, let alone the estimation of models with spatial components.
@@
 ``{[}w{]}hat is one person's mean structure could be another person's
-correlation structure'' (Cressie, 1989, p. 201). There are almost always
-multiple ways to approach a modelling situation, as lively illustrated
-by a recent paper that reports the results of a crowdsourced modelling
-experiment (Schweinsberg et al., 2021). In the present case, we would
-argue that spatial sampling is an important aspect of the modeling
-process. Importantly, by adopting high reproducibility standards, SWN
-made a valuable contribution to the collective enterprise of seeking
-knowledge. Their effort, and subsequent efforts to validate and expand
-on their work, can potentially contribute to provide clarity to ongoing
-conversations about the relevance of density and the spread of COVID-19.
+correlation structure'' \citep[p.~201]{Cressie1989geostatistics}. There
+are almost always multiple ways to approach a modelling situation, as
+lively illustrated by a recent paper that reports the results of a
+crowdsourced modelling experiment \citep{Schweinsberg2021same}. In the
+present case, we would argue that spatial sampling is an important
+aspect of the modeling process. Importantly, by adopting high
+reproducibility standards, SWN made a valuable contribution to the
+collective enterprise of seeking knowledge. Their effort, and subsequent
+efforts to validate and expand on their work, can potentially contribute
+to provide clarity to ongoing conversations about the relevance of
+density and the spread of COVID-19.
 
@@
 spread of the virus {[}in Wong and Li's words, ``\,`Density is destiny'
-is probably an overstatement''; (2020){]}. At the same time, the results
-presented here also stand in contrast to the findings of Hamidi et al.,
-who found that higher density was either not significantly associated
-with the rate of the virus in a cross-sectional study (Hamidi et al.,
-2020b), or was negatively associated with it in a longitudinal setting
-{[}Hamidi et al. (2020a). In this sense, the conclusion that density
-does not aggravate the pandemic may have been somewhat premature;
-instead, reanalysis of the data of SWN suggests that Fielding-Miller et
-al. (2020) might be onto something with respect to the difference
-between rural and urban counties. More generally, there is no doubt that
-in population-level studies density is indicative of proximity, but it
-also potentially is a proxy for adaptive behavior. And it is possible
-that the determining factor during COVID-19, at least in the US, has
-been variations in perceptions of the risks associated with contagion
-(Chauhan et al., 2021), and subsequent compensations in behavior in more
-and less dense regions.
+is probably an overstatement''; -\citet{Wong2020spreading}{]}. At the
+same time, the results presented here also stand in contrast to the
+findings of Hamidi et al., who found that higher density was either not
+significantly associated with the rate of the virus in a cross-sectional
+study \citep{Hamidi2020density}, or was negatively associated with it in
+a longitudinal setting {[}\citet{Hamidi2020longitudinal}. In this sense,
+the conclusion that density does not aggravate the pandemic may have
+been somewhat premature; instead, reanalysis of the data of SWN suggests
+that Fielding-Miller et al. \citeyearpar{Fielding2020social} might be
+onto something with respect to the difference between rural and urban
+counties. More generally, there is no doubt that in population-level
+studies density is indicative of proximity, but it also potentially is a
+proxy for adaptive behavior. And it is possible that the determining
+factor during COVID-19, at least in the US, has been variations in
+perceptions of the risks associated with contagion
+\citep{Chauhan2021covid}, and subsequent compensations in behavior in
+more and less dense regions.
 
@@
 dealing with a global pandemic, and a potential ``carnage of substandard
-research'' (Bramstedt, 2020), highlights the importance of efforts to
-maintain the quality of scientific outputs during COVID-19. An important
-part of quality control is the ability of independent researchers to
-verify and examine the results of materials published in the literature.
-As previous research illustrates, reproducibility in scientific research
-remains an important but elusive goal (Gustot, 2020; e.g., Iqbal et al.,
-2016; Stodden et al., 2018; Sumner et al., 2020). This idea is
-reinforced by the review conducted for this paper in the context of
-research about population density and the spread of COVID-19.
-
-Taking one recent example from the literature {[}Sy et al., Sy et al.
-(2021); SWN{]}, the present paper illustrates the importance of good
-reproducibility practices. Sharing data and code can catalyze research,
-by allowing independent verification of findings, as well as additional
-research. After verifying the results of SWN, experiments with sample
-selection models and variations in the definition of model inputs, lead
-to an important reappraisal of the conclusion that high density is
-associated with greater spread of the virus. Instead, the possibility of
-a non-monotonical relationship between population density and contagion
-is raised. I do not claim that the analysis presented here is the last
-word on the topic of density and the spread of COVID-19, and there is
-always the possibility that someone else will be better equipped to
-analyze these data with greater competence. By opening up the analysis,
-documenting the way data were pre-processed, and by sharing analysis
-ready data, my hope would be that others will be able to discover the
-limitations of my own analysis and improve on it, as appropriate.
-
-More generally, my hope is that the research of Sy et al. (2021), the
-present paper, and similar reproducible publications, will continue to
-encourage others to adopt higher reproducibility standards in their
-research.
+research'' \citep{Bramstedt2020carnage}, highlights the importance of
+efforts to maintain the quality of scientific outputs during COVID-19.
+An important part of quality control is the ability of independent
+researchers to verify and examine the results of materials published in
+the literature. As previous research illustrates, reproducibility in
+scientific research remains an important but elusive goal
+\citep[e.g.,][]{Iqbal2016reproducible, Stodden2018empirical, Sumner2020reproducibility, Gustot2020quality}.
+This idea is reinforced by the review conducted for this paper in the
+context of research about population density and the spread of COVID-19.
+
+Taking one recent example from the literature {[}Sy et al.,
+\citet{Sy2021population}; SWN{]}, the present paper illustrates the
+importance of good reproducibility practices. Sharing data and code can
+catalyze research, by allowing independent verification of findings, as
+well as additional research. After verifying the results of SWN,
+experiments with sample selection models and variations in the
+definition of model inputs, lead to an important reappraisal of the
+conclusion that high density is associated with greater spread of the
+virus. Instead, the possibility of a non-monotonical relationship
+between population density and contagion is raised. I do not claim that
+the analysis presented here is the last word on the topic of density and
+the spread of COVID-19, and there is always the possibility that someone
+else will be better equipped to analyze these data with greater
+competence. By opening up the analysis, documenting the way data were
+pre-processed, and by sharing analysis ready data, my hope would be that
+others will be able to discover the limitations of my own analysis and
+improve on it, as appropriate.
+
+More generally, my hope is that the research of Sy et al.
+\citeyearpar{Sy2021population}, the present paper, and similar
+reproducible publications, will continue to encourage others to adopt
+higher reproducibility standards in their research.
 
@@
 The analysis reported in this paper was conducted in the \texttt{R}
-computing statistical language (\textbf{R-base?}). The source document
-is an Rmarkdown document (\textbf{rmarkdown2018?};
-\textbf{rmarkdown2020?}) processed using \texttt{knitr}
-(\textbf{knitr2015?}; \textbf{knitr2014?}). The following packages were
+computing statistical language \citep{R-base}. The source document is an
+Rmarkdown document \citep{rmarkdown2018, rmarkdown2020} processed using
+\texttt{knitr} \citep{knitr2015, knitr2014}. The following packages were
 used in the analysis, and I wish to acknowledge their creators for their
-generous efforts: \texttt{adespatial} (\textbf{R-adespatial?}),
-\texttt{censReg} (\textbf{R-censReg?}), \texttt{dplyr}
-(\textbf{R-dplyr?}), \texttt{forcats} (\textbf{R-forcats?}),
-\texttt{ggplot2} (\textbf{ggplot22016?}), \texttt{gmm}
-(\textbf{gmm2010?}), \texttt{kableExtra} (\textbf{R-kableExtra?}),
-\texttt{Matrix} (\textbf{R-Matrix?}), \texttt{maxLik}
-(\textbf{maxLik2011?}), \texttt{miscTools} (\textbf{R-miscTools?}),
-\texttt{mvtnorm} (\textbf{mvtnorm2009?}), \texttt{nlme}
-(\textbf{R-nlme?}), \texttt{patchwork} (\textbf{R-patchwork?}),
-\texttt{purrr} (\textbf{R-purrr?}), \texttt{readr} (\textbf{R-readr?}),
-\texttt{sampleSelection} (\textbf{sampleSelection2008?}),
-\texttt{sandwich} (\textbf{sandwich2020?}; \textbf{sandwich2004?};
-\textbf{sandwich2006?}), \texttt{scico} (\textbf{R-scico?}), \texttt{sf}
-(\textbf{sf2018?}), \texttt{sp} (\textbf{sp2005?}; \textbf{sp2013?}),
-\texttt{spatialprobit} (\textbf{R-spatialprobit?}), \texttt{spData}
-(\textbf{R-spData?}), \texttt{spdep} (\textbf{spdep2018?};
-\textbf{spdep2013?}), \texttt{stringr} {[}R-stringr{]}, \texttt{tibble}
-(\textbf{R-tibble?}), \texttt{tidycensus} (\textbf{R-tidycensus?}),
-\texttt{tidyr} (\textbf{R-tidyr?}), \texttt{tidyverse}
-(\textbf{tidyverse2019?}), \texttt{tmvtnorm} (\textbf{R-tmvtnorm?}),
-\texttt{units} (\textbf{units2016?}). This research was not supported by
+generous efforts: \texttt{adespatial} \citep{R-adespatial},
+\texttt{censReg} \citep{R-censReg}, \texttt{dplyr} \citep{R-dplyr},
+\texttt{forcats} \citep{R-forcats}, \texttt{ggplot2}
+\citep{ggplot22016}, \texttt{gmm} \citep{gmm2010}, \texttt{kableExtra}
+\citep{R-kableExtra}, \texttt{Matrix} \citep{R-Matrix}, \texttt{maxLik}
+\citep{maxLik2011}, \texttt{miscTools} \citep{R-miscTools},
+\texttt{mvtnorm} \citep{mvtnorm2009}, \texttt{nlme} \citep{R-nlme},
+\texttt{patchwork} \citep{R-patchwork}, \texttt{purrr} \citep{R-purrr},
+\texttt{readr} \citep{R-readr}, \texttt{sampleSelection}
+\citep{sampleSelection2008}, \texttt{sandwich}
+\citep{sandwich2020, sandwich2004, sandwich2006}, \texttt{scico}
+\citep{R-scico}, \texttt{sf} \citep{sf2018}, \texttt{sp}
+\citep{sp2005, sp2013}, \texttt{spatialprobit} \citep{R-spatialprobit},
+\texttt{spData} \citep{R-spData}, \texttt{spdep}
+\citep{spdep2018, spdep2013}, \texttt{stringr} {[}R-stringr{]},
+\texttt{tibble} \citep{R-tibble}, \texttt{tidycensus}
+\citep{R-tidycensus}, \texttt{tidyr} \citep{R-tidyr}, \texttt{tidyverse}
+\citep{tidyverse2019}, \texttt{tmvtnorm} \citep{R-tmvtnorm},
+\texttt{units} \citep{units2016}. This research was not supported by
 Canada's Research Councils.
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Ahmad2020association}{}%
-Ahmad, K., Erqou, S., Shah, N., Nazir, U., Morrison, A.R., Choudhary,
-G., Wu, W.-C., 2020. Association of poor housing conditions with
-COVID-19 incidence and mortality across US counties. PLOS ONE 15,
-e0241327.
-doi:\href{https://doi.org/10.1371/journal.pone.0241327}{10.1371/journal.pone.0241327}
-
-\leavevmode\hypertarget{ref-Amadu2021assessing}{}%
-Amadu, I., Ahinkorah, B.O., Afitiri, A.-R., Seidu, A.-A., Ameyaw, E.K.,
-Hagan, J.E., Duku, E., Aram, S.A., 2021. Assessing sub-regional-specific
-strengths of healthcare systems associated with COVID-19 prevalence,
-deaths and recoveries in africa. PLOS ONE 16, e0247274.
-doi:\href{https://doi.org/10.1371/journal.pone.0247274}{10.1371/journal.pone.0247274}
-
-\leavevmode\hypertarget{ref-Anazco2021publication}{}%
-Añazco, D., Nicolalde, B., Espinosa, I., Camacho, J., Mushtaq, M.,
-Gimenez, J., Teran, E., 2021. Publication rate and citation counts for
-preprints released during the COVID-19 pandemic: The good, the bad and
-the ugly. PeerJ 9, e10927.
-doi:\href{https://doi.org/10.7717/peerj.10927}{10.7717/peerj.10927}
-
-\leavevmode\hypertarget{ref-Arribas2021open}{}%
-Arribas-Bel, D., Green, M., Rowe, F., Singleton, A., 2021. Open data
-products: A framework for creating valuable analysis-ready data. Journal
-of Geographical Systems.
-
-\leavevmode\hypertarget{ref-Basu2017ten}{}%
-Basu, S., Carney, M.A., Kenworthy, N.J., 2017. Ten years after the
-financial crisis: The long reach of austerity and its global impacts on
-health. Social Science \& Medicine 187, 203--207.
-doi:\href{https://doi.org/10.1016/j.socscimed.2017.06.026}{10.1016/j.socscimed.2017.06.026}
-
-\leavevmode\hypertarget{ref-Bhadra2021impact}{}%
-Bhadra, A., Mukherjee, A., Sarkar, K., 2021. Impact of population
-density on covid-19 infected and mortality rate in india. Modeling Earth
-Systems and Environment 7, 623--629.
-doi:\href{https://doi.org/10.1007/s40808-020-00984-7}{10.1007/s40808-020-00984-7}
-
-\leavevmode\hypertarget{ref-Bivand2020progress}{}%
-Bivand, R.S., 2020. Progress in the r ecosystem for representing and
-handling spatial data. Journal of Geographical Systems.
-doi:\href{https://doi.org/10.1007/s10109-020-00336-0}{10.1007/s10109-020-00336-0}
-
-\leavevmode\hypertarget{ref-Bramstedt2020carnage}{}%
-Bramstedt, K.A., 2020. The carnage of substandard research during the
-COVID-19 pandemic: A call for quality. Journal of Medical Ethics 46,
-803--807.
-doi:\href{https://doi.org/10.1136/medethics-2020-106494}{10.1136/medethics-2020-106494}
-
-\leavevmode\hypertarget{ref-Brandtner2021creatures}{}%
-Brandtner, C., Bettencourt, L.M.A., Berman, M.G., Stier, A.J., 2021.
-Creatures of the state? Metropolitan counties compensated for state
-inaction in initial u.s. Response to COVID-19 pandemic. PLOS ONE 16,
-e0246249.
-doi:\href{https://doi.org/10.1371/journal.pone.0246249}{10.1371/journal.pone.0246249}
-
-\leavevmode\hypertarget{ref-Broggini2017reproducible}{}%
-Broggini, F., Dellinger, J., Fomel, S., Liu, Y., 2017. Reproducible
-research: Geophysics papers of the future - introduction. Geophysics 82.
-doi:\href{https://doi.org/10.1190/geo2017-0918-spseintro.1}{10.1190/geo2017-0918-spseintro.1}
-
-\leavevmode\hypertarget{ref-Brunsdon2020opening}{}%
-Brunsdon, C., Comber, A., 2020. Opening practice: Supporting
-reproducibility and critical spatial data science. Journal of
-Geographical Systems.
-doi:\href{https://doi.org/10.1007/s10109-020-00334-2}{10.1007/s10109-020-00334-2}
-
-\leavevmode\hypertarget{ref-Chauhan2021covid}{}%
-Chauhan, R.S., Capasso da Silva, D., Salon, D., Shamshiripour, A.,
-Rahimi, E., Sutradhar, U., Khoeini, S., Mohammadian, A.(Kouros).,
-Derrible, S., Pendyala, R., 2021. COVID-19 related attitudes and risk
-perceptions across urban, rural, and suburban areas in the united
-states. Findings.
-doi:\href{https://doi.org/10.32866/001c.23714}{10.32866/001c.23714}
-
-\leavevmode\hypertarget{ref-Cressie1989geostatistics}{}%
-Cressie, N., 1989. Geostatistics. The American Statistician 43, 197.
-doi:\href{https://doi.org/10.2307/2685361}{10.2307/2685361}
-
-\leavevmode\hypertarget{ref-Cruz2020exploring}{}%
-Cruz, C.J.P., Ganly, R., Li, Z., Gietel-Basten, S., 2020. Exploring the
-young demographic profile of COVID-19 cases in hong kong: Evidence from
-migration and travel history data. PLOS ONE 15, e0235306.
-doi:\href{https://doi.org/10.1371/journal.pone.0235306}{10.1371/journal.pone.0235306}
-
-\leavevmode\hypertarget{ref-Farber2011running}{}%
-Farber, S., Páez, A., 2011. Running to stay in place: The time-use
-implications of automobile oriented land-use and travel. Journal of
-Transport Geography 19, 782--793.
-doi:\href{https://doi.org/10.1016/j.jtrangeo.2010.09.008}{10.1016/j.jtrangeo.2010.09.008}
-
-\leavevmode\hypertarget{ref-Feng2020spread}{}%
-Feng, Y., Li, Q., Tong, X., Wang, R., Zhai, S., Gao, C., Lei, Z., Chen,
-S., Zhou, Y., Wang, J., Yan, X., Xie, H., Chen, P., Liu, S., Xv, X.,
-Liu, S., Jin, Y., Wang, C., Hong, Z., Luan, K., Wei, C., Xu, J., Jiang,
-H., Xiao, C., Guo, Y., 2020. Spatiotemporal spread pattern of the
-COVID-19 cases in china. PLOS ONE 15, e0244351.
-doi:\href{https://doi.org/10.1371/journal.pone.0244351}{10.1371/journal.pone.0244351}
-
-\leavevmode\hypertarget{ref-Feyman2020effectiveness}{}%
-Feyman, Y., Bor, J., Raifman, J., Griffith, K.N., 2020. Effectiveness of
-COVID-19 shelter-in-place orders varied by state. PLOS ONE 15, e0245008.
-doi:\href{https://doi.org/10.1371/journal.pone.0245008}{10.1371/journal.pone.0245008}
-
-\leavevmode\hypertarget{ref-Fielding2020social}{}%
-Fielding-Miller, R.K., Sundaram, M.E., Brouwer, K., 2020. Social
-determinants of COVID-19 mortality at the county level. PLOS ONE 15,
-e0240151.
-doi:\href{https://doi.org/10.1371/journal.pone.0240151}{10.1371/journal.pone.0240151}
-
-\leavevmode\hypertarget{ref-Florida2020how}{}%
-Florida, R., Glaeser, E., Sharif, M., Bedi, K., Campanella, T., Chee,
-C., Doctoroff, D., Katz, B., Katz, R., Kotkin, J., 2020. How life in our
-cities will look after the coronavirus pandemic. Foreign Policy 1.
-
-\leavevmode\hypertarget{ref-Fraser2021evolving}{}%
-Fraser, N., Brierley, L., Dey, G., Polka, J.K., Pálfy, M., Nanni, F.,
-Coates, J.A., 2021. The evolving role of preprints in the dissemination
-of COVID-19 research and their impact on the science communication
-landscape. PLOS Biology 19, e3000959.
-doi:\href{https://doi.org/10.1371/journal.pbio.3000959}{10.1371/journal.pbio.3000959}
-
-\leavevmode\hypertarget{ref-Getis2002comparative}{}%
-Getis, A., Griffith, D.A., 2002. Comparative spatial filtering in
-regression analysis. Geographical Analysis 34, 130--140.
-
-\leavevmode\hypertarget{ref-Gomez2021infekta}{}%
-Gomez, J., Prieto, J., Leon, E., Rodríguez, A., 2021. INFEKTA---an
-agent-based model for transmission of infectious diseases: The COVID-19
-case in bogotá, colombia. PLOS ONE 16, e0245787.
-doi:\href{https://doi.org/10.1371/journal.pone.0245787}{10.1371/journal.pone.0245787}
-
-\leavevmode\hypertarget{ref-Griffith2000linear}{}%
-Griffith, D.A., 2000. A linear regression solution to the spatial
-autocorrelation problem. Journal of Geographical Systems 2, 141--156.
-
-\leavevmode\hypertarget{ref-Griffith2004spatial}{}%
-Griffith, D.A., 2004. A spatial filtering specification for the
-autologistic model. Environment and Planning A 36, 1791--1811.
-
-\leavevmode\hypertarget{ref-Gustot2020quality}{}%
-Gustot, T., 2020. Quality and reproducibility during the COVID-19
-pandemic. JHEP Rep 2, 100141.
-doi:\href{https://doi.org/10.1016/j.jhepr.2020.100141}{10.1016/j.jhepr.2020.100141}
-
-\leavevmode\hypertarget{ref-Hamidi2020longitudinal}{}%
-Hamidi, S., Ewing, R., Sabouri, S., 2020a. Longitudinal analyses of the
-relationship between development density and the COVID-19 morbidity and
-mortality rates: Early evidence from 1,165 metropolitan counties in the
-united states. Health \& Place 64, 102378.
-doi:\href{https://doi.org/10.1016/j.healthplace.2020.102378}{10.1016/j.healthplace.2020.102378}
-
-\leavevmode\hypertarget{ref-Hamidi2020density}{}%
-Hamidi, S., Sabouri, S., Ewing, R., 2020b. Does density aggravate the
-COVID-19 pandemic? Journal of the American Planning Association 86,
-495--509.
-doi:\href{https://doi.org/10.1080/01944363.2020.1777891}{10.1080/01944363.2020.1777891}
-
-\leavevmode\hypertarget{ref-Hamidi2021compact}{}%
-Hamidi, S., Zandiatashbar, A., 2021. Compact development and adherence
-to stay-at-home order during the COVID-19 pandemic: A longitudinal
-investigation in the united states. Landscape and Urban Planning 205,
-103952. doi:\url{https://doi.org/10.1016/j.landurbplan.2020.103952}
-
-\leavevmode\hypertarget{ref-Harris2021Changes}{}%
-Harris, M.A., Branion-Calles, M., 2021. Changes in commute mode
-attributed to COVID-19 risk in canadian national survey data. Findings.
-doi:\href{https://doi.org/10.32866/001c.19088}{10.32866/001c.19088}
-
-\leavevmode\hypertarget{ref-Herndon2014high}{}%
-Herndon, T., Ash, M., Pollin, R., 2014. Does high public debt
-consistently stifle economic growth? A critique of reinhart and rogoff.
-Cambridge Journal of Economics 38, 257--279.
-doi:\href{https://doi.org/10.1093/cje/bet075}{10.1093/cje/bet075}
-
-\leavevmode\hypertarget{ref-Inbaraj2021seroprevalence}{}%
-Inbaraj, L.R., George, C.E., Chandrasingh, S., 2021. Seroprevalence of
-COVID-19 infection in a rural district of south india: A
-population-based seroepidemiological study. PLOS ONE 16, e0249247.
-doi:\href{https://doi.org/10.1371/journal.pone.0249247}{10.1371/journal.pone.0249247}
-
-\leavevmode\hypertarget{ref-Ince2012case}{}%
-Ince, D.C., Hatton, L., Graham-Cumming, J., 2012. The case for open
-computer programs. Nature 482, 485--488.
-doi:\href{https://doi.org/10.1038/nature10836}{10.1038/nature10836}
-
-\leavevmode\hypertarget{ref-Ioannidis2014increasing}{}%
-Ioannidis, J.P.A., Greenland, S., Hlatky, M.A., Khoury, M.J., Macleod,
-M.R., Moher, D., Schulz, K.F., Tibshirani, R., 2014. Increasing value
-and reducing waste in research design, conduct, and analysis. Lancet
-383, 166--175.
-doi:\href{https://doi.org/10.1016/s0140-6736(13)62227-8}{10.1016/s0140-6736(13)62227-8}
-
-\leavevmode\hypertarget{ref-Iqbal2016reproducible}{}%
-Iqbal, S.A., Wallach, J.D., Khoury, M.J., Schully, S.D., Ioannidis,
-J.P.A., 2016. Reproducible research practices and transparency across
-the biomedical literature. Plos Biology 14.
-doi:\href{https://doi.org/10.1371/journal.pbio.1002333}{10.1371/journal.pbio.1002333}
-
-\leavevmode\hypertarget{ref-Jamal2020Changes}{}%
-Jamal, S., Paez, A., 2020. Changes in trip-making frequency by mode
-during COVID-19. Findings.
-doi:\href{https://doi.org/10.32866/001c.17977}{10.32866/001c.17977}
-
-\leavevmode\hypertarget{ref-Kadi2020population}{}%
-Kadi, N., Khelfaoui, M., 2020. Population density, a factor in the
-spread of COVID-19 in algeria: Statistic study. Bulletin of the National
-Research Centre 44.
-doi:\href{https://doi.org/10.1186/s42269-020-00393-x}{10.1186/s42269-020-00393-x}
-
-\leavevmode\hypertarget{ref-Khavarian2021high}{}%
-Khavarian-Garmsir, A.R., Sharifi, A., Moradpour, N., 2021. Are
-high-density districts more vulnerable to the COVID-19 pandemic?
-Sustainable Cities and Society 70, 102911.
-doi:\href{https://doi.org/10.1016/j.scs.2021.102911}{10.1016/j.scs.2021.102911}
-
-\leavevmode\hypertarget{ref-Konkol2019examination}{}%
-Konkol, M., Kray, C., 2019. In-depth examination of spatiotemporal
-figures in open reproducible research. Cartography and Geographic
-Information Science 46, 412--427.
-doi:\href{https://doi.org/10.1080/15230406.2018.1512421}{10.1080/15230406.2018.1512421}
-
-\leavevmode\hypertarget{ref-Konkol2019computational}{}%
-Konkol, M., Kray, C., Pfeiffer, M., 2019. Computational reproducibility
-in geoscientific papers: Insights from a series of studies with
-geoscientists and a reproduction study. International Journal of
-Geographical Information Science 33, 408--429.
-doi:\href{https://doi.org/10.1080/13658816.2018.1508687}{10.1080/13658816.2018.1508687}
-
-\leavevmode\hypertarget{ref-Kwon2021swamped}{}%
-Kwon, D., 2020. How swamped preprint servers are blocking bad
-coronavirus research. Nature 581, 130--132.
-
-\leavevmode\hypertarget{ref-Lee2020human}{}%
-Lee, M., Zhao, J., Sun, Q., Pan, Y., Zhou, W., Xiong, C., Zhang, L.,
-2020. Human mobility trends during the early stage of the COVID-19
-pandemic in the united states. PLOS ONE 15, e0241468.
-doi:\href{https://doi.org/10.1371/journal.pone.0241468}{10.1371/journal.pone.0241468}
-
-\leavevmode\hypertarget{ref-Li2018effect}{}%
-Li, R., Richmond, P., Roehner, B.M., 2018. Effect of population density
-on epidemics. Physica A: Statistical Mechanics and its Applications 510,
-713--724.
-doi:\href{https://doi.org/10.1016/j.physa.2018.07.025}{10.1016/j.physa.2018.07.025}
-
-\leavevmode\hypertarget{ref-Maddala1983limited}{}%
-Maddala, G.S., 1983. Limited-dependent and qualitative variables in
-econometrics. Cambridge University Press, Cambridge.
-
-\leavevmode\hypertarget{ref-Micallef2020first}{}%
-Micallef, S., Piscopo, T.V., Casha, R., Borg, D., Vella, C., Zammit,
-M.-A., Borg, J., Mallia, D., Farrugia, J., Vella, S.M., Xerri, T.,
-Portelli, A., Fenech, M., Fsadni, C., Mallia Azzopardi, C., 2020. The
-first wave of COVID-19 in malta; a national cross-sectional study. PLOS
-ONE 15, e0239389.
-doi:\href{https://doi.org/10.1371/journal.pone.0239389}{10.1371/journal.pone.0239389}
-
-\leavevmode\hypertarget{ref-Molloy2020Tracing}{}%
-Molloy, J., Tchervenkov, C., Hintermann, B., Axhausen, K.W., 2020.
-Tracing the sars-CoV-2 impact: The first month in switzerland. Findings.
-doi:\href{https://doi.org/10.32866/001c.12903}{10.32866/001c.12903}
-
-\leavevmode\hypertarget{ref-Moore1970some}{}%
-Moore, E.G., 1970. Some spatial properties of urban contact fields.
-Geographical Analysis 2, 376--386.
-
-\leavevmode\hypertarget{ref-Moore1970urban}{}%
-Moore, E.G., Brown, L.A., 1970. Urban acquaintance fields: An evaluation
-of a spatial model. Environment and Planning 2, 443--454.
-
-\leavevmode\hypertarget{ref-Noland1995perceived}{}%
-Noland, R.B., 1995. PERCEIVED RISK AND MODAL CHOICE - RISK COMPENSATION
-IN TRANSPORTATION SYSTEM. Accident Analysis and Prevention 27, 503--521.
-doi:\href{https://doi.org/10.1016/0001-4575(94)00087-3}{10.1016/0001-4575(94)00087-3}
-
-\leavevmode\hypertarget{ref-Noland2021mobility}{}%
-Noland, R.B., 2021. Mobility and the effective reproduction rate of
-COVID-19. Journal of Transport \& Health 20, 101016.
-doi:\url{https://doi.org/10.1016/j.jth.2021.101016}
-
-\leavevmode\hypertarget{ref-Noury2021how}{}%
-Noury, A., François, A., Gergaud, O., Garel, A., 2021. How does COVID-19
-affect electoral participation? Evidence from the french municipal
-elections. PLOS ONE 16, e0247026.
-doi:\href{https://doi.org/10.1371/journal.pone.0247026}{10.1371/journal.pone.0247026}
-
-\leavevmode\hypertarget{ref-Paez2019using}{}%
-Paez, A., 2019. Using spatial filters and exploratory data analysis to
-enhance regression models of spatial data. Geographical Analysis 51,
-314--338.
-doi:\href{https://doi.org/10.1111/gean.12180}{10.1111/gean.12180}
-
-\leavevmode\hypertarget{ref-Paez2020using}{}%
-Paez, A., 2020. Using google community mobility reports to investigate
-the incidence of COVID-19 in the united states. Findings.
-doi:\url{https://doi.org/10.32866/001c.12976}
-
-\leavevmode\hypertarget{ref-Paez2020spatio}{}%
-Paez, A., Lopez, F.A., Menezes, T., Cavalcanti, R., Pitta, M.G. da R.,
-2020. A spatio-temporal analysis of the environmental correlates of
-COVID-19 incidence in spain. Geographical Analysis n/a.
-doi:\href{https://doi.org/10.1111/gean.12241}{10.1111/gean.12241}
-
-\leavevmode\hypertarget{ref-Pequeno2020air}{}%
-Pequeno, P., Mendel, B., Rosa, C., Bosholn, M., Souza, J.L., Baccaro,
-F., Barbosa, R., Magnusson, W., 2020. Air transportation, population
-density and temperature predict the spread of COVID-19 in brazil. PeerJ
-8, e9322.
-doi:\href{https://doi.org/10.7717/peerj.9322}{10.7717/peerj.9322}
-
-\leavevmode\hypertarget{ref-Phillips2011risk}{}%
-Phillips, R.O., Fyhri, A., Sagberg, F., 2011. Risk compensation and
-bicycle helmets. Risk Analysis 31, 1187--1195.
-doi:\href{https://doi.org/10.1111/j.1539-6924.2011.01589.x}{10.1111/j.1539-6924.2011.01589.x}
-
-\leavevmode\hypertarget{ref-Praharaj2020Using}{}%
-Praharaj, S., King, D., Pettit, C., Wentz, E., 2020. Using aggregated
-mobility data to measure the effect of COVID-19 policies on mobility
-changes in sydney, london, phoenix, and pune. Findings.
-doi:\href{https://doi.org/10.32866/001c.17590}{10.32866/001c.17590}
-
-\leavevmode\hypertarget{ref-Richens2000condoms}{}%
-Richens, J., Imrie, J., Copas, A., 2000. Condoms and seat belts: The
-parallels and the lessons. Lancet 355, 400--403.
-doi:\href{https://doi.org/10.1016/s0140-6736(99)09109-6}{10.1016/s0140-6736(99)09109-6}
-
-\leavevmode\hypertarget{ref-Rocklov2020high}{}%
-Rocklöv, J., Sjödin, H., 2020. High population densities catalyse the
-spread of COVID-19. Journal of Travel Medicine 27.
-doi:\href{https://doi.org/10.1093/jtm/taaa038}{10.1093/jtm/taaa038}
-
-\leavevmode\hypertarget{ref-Roy2020factors}{}%
-Roy, S., Ghosh, P., 2020. Factors affecting COVID-19 infected and death
-rates inform lockdown-related policymaking. PLOS ONE 15, e0241165.
-doi:\href{https://doi.org/10.1371/journal.pone.0241165}{10.1371/journal.pone.0241165}
-
-\leavevmode\hypertarget{ref-Schweinsberg2021same}{}%
-Schweinsberg, M., Feldman, M., Staub, N., van den Akker, O.R., van Aert,
-R.C.M., van Assen, M.A.L.M., Liu, Y., Althoff, T., Heer, J., Kale, A.,
-Mohamed, Z., Amireh, H., Venkatesh Prasad, V., Bernstein, A., Robinson,
-E., Snellman, K., Amy Sommer, S., Otner, S.M.G., Robinson, D., Madan,
-N., Silberzahn, R., Goldstein, P., Tierney, W., Murase, T., Mandl, B.,
-Viganola, D., Strobl, C., Schaumans, C.B.C., Kelchtermans, S., Naseeb,
-C., Mason Garrison, S., Yarkoni, T., Richard Chan, C.S., Adie, P.,
-Alaburda, P., Albers, C., Alspaugh, S., Alstott, J., Nelson, A.A., Ariño
-de la Rubia, E., Arzi, A., Bahník, Š., Baik, J., Winther Balling, L.,
-Banker, S., AA Baranger, D., Barr, D.J., Barros-Rivera, B., Bauer, M.,
-Blaise, E., Boelen, L., Bohle Carbonell, K., Briers, R.A., Burkhard, O.,
-Canela, M.-A., Castrillo, L., Catlett, T., Chen, O., Clark, M., Cohn,
-B., Coppock, A., Cugueró-Escofet, N., Curran, P.G., Cyrus-Lai, W., Dai,
-D., Valentino Dalla Riva, G., Danielsson, H., F. S. M. Russo, R. de, de
-Silva, N., Derungs, C., Dondelinger, F., Duarte de Souza, C., Tyson
-Dube, B., Dubova, M., Mark Dunn, B., Adriaan Edelsbrunner, P., Finley,
-S., Fox, N., Gnambs, T., Gong, Y., Grand, E., Greenawalt, B., Han, D.,
-Hanel, P.H.P., Hong, A.B., Hood, D., Hsueh, J., Huang, L., Hui, K.N.,
-Hultman, K.A., Javaid, A., Ji Jiang, L., Jong, J., Kamdar, J., Kane, D.,
-Kappler, G., Kaszubowski, E., Kavanagh, C.M., Khabsa, M., Kleinberg, B.,
-Kouros, J., Krause, H., Krypotos, A.-M., Lavbič, D., Ling Lee, R.,
-Leffel, T., Yang Lim, W., Liverani, S., Loh, B., Lønsmann, D., Wei Low,
-J., Lu, A., MacDonald, K., Madan, C.R., Hjorth Madsen, L., Maimone, C.,
-Mangold, A., Marshall, A., Ester Matskewich, H., Mavon, K., McLain,
-K.L., McNamara, A.A., McNeill, M., Mertens, U., Miller, D., Moore, B.,
-Moore, A., Nantz, E., Nasrullah, Z., Nejkovic, V., Nell, C.S., Arthur
-Nelson, A., Nilsonne, G., Nolan, R., O'Brien, C.E., O'Neill, P., O'Shea,
-K., Olita, T., Otterbacher, J., Palsetia, D., Pereira, B., Pozdniakov,
-I., Protzko, J., Reyt, J.-N., Riddle, T., (Akmal) Ridhwan Omar Ali, A.,
-Ropovik, I., Rosenberg, J.M., Rothen, S., Schulte-Mecklenbeck, M.,
-Sharma, N., Shotwell, G., Skarzynski, M., Stedden, W., Stodden, V.,
-Stoffel, M.A., Stoltzman, S., Subbaiah, S., Tatman, R., Thibodeau, P.H.,
-Tomkins, S., Valdivia, A., Druijff-van de Woestijne, G.B., Viana, L.,
-Villesèche, F., Duncan Wadsworth, W., Wanders, F., Watts, K., Wells,
-J.D., Whelpley, C.E., Won, A., Wu, L., Yip, A., Youngflesh, C., Yu,
-J.-C., Zandian, A., Zhang, L., Zibman, C., Luis Uhlmann, E., 2021. Same
-data, different conclusions: Radical dispersion in empirical results
-when independent analysts operationalize and test the same hypothesis.
-Organizational Behavior and Human Decision Processes 165, 228--249.
-doi:\url{https://doi.org/10.1016/j.obhdp.2021.02.003}
-
-\leavevmode\hypertarget{ref-Sharifi2020covid}{}%
-Sharifi, A., Khavarian-Garmsir, A.R., 2020. The COVID-19 pandemic:
-Impacts on cities and major lessons for urban planning, design, and
-management. Science of The Total Environment 749, 142391.
-doi:\url{https://doi.org/10.1016/j.scitotenv.2020.142391}
-
-\leavevmode\hypertarget{ref-Skorka2020macroecology}{}%
-Skórka, P., Grzywacz, B., Moroń, D., Lenda, M., 2020. The macroecology
-of the COVID-19 pandemic in the anthropocene. PLOS ONE 15, e0236856.
-doi:\href{https://doi.org/10.1371/journal.pone.0236856}{10.1371/journal.pone.0236856}
-
-\leavevmode\hypertarget{ref-Souris2020covid}{}%
-Souris, M., Gonzalez, J.-P., 2020. COVID-19: Spatial analysis of
-hospital case-fatality rate in france. PLOS ONE 15, e0243606.
-doi:\href{https://doi.org/10.1371/journal.pone.0243606}{10.1371/journal.pone.0243606}
-
-\leavevmode\hypertarget{ref-Stephens2021impact}{}%
-Stephens, K.E., Chernyavskiy, P., Bruns, D.R., 2021. Impact of altitude
-on COVID-19 infection and death in the united states: A modeling and
-observational study. PLOS ONE 16, e0245055.
-doi:\href{https://doi.org/10.1371/journal.pone.0245055}{10.1371/journal.pone.0245055}
-
-\leavevmode\hypertarget{ref-Stodden2018empirical}{}%
-Stodden, V., Seiler, J., Ma, Z.K., 2018. An empirical analysis of
-journal policy effectiveness for computational reproducibility.
-Proceedings of the National Academy of Sciences of the United States of
-America 115, 2584--2589.
-doi:\href{https://doi.org/10.1073/pnas.1708290115}{10.1073/pnas.1708290115}
-
-\leavevmode\hypertarget{ref-Sumner2020reproducibility}{}%
-Sumner, J., Haynes, L., Nathan, S., Hudson-Vitale, C., McIntosh, L.D.,
-2020. Reproducibility and reporting practices in COVID-19 preprint
-manuscripts. medRxiv 2020.03.24.20042796.
-doi:\href{https://doi.org/10.1101/2020.03.24.20042796}{10.1101/2020.03.24.20042796}
-
-\leavevmode\hypertarget{ref-Sun2020impacts}{}%
-Sun, Z., Zhang, H., Yang, Y., Wan, H., Wang, Y., 2020. Impacts of
-geographic factors and population density on the COVID-19 spreading
-under the lockdown policies of china. Science of The Total Environment
-746, 141347.
-doi:\href{https://doi.org/10.1016/j.scitotenv.2020.141347}{10.1016/j.scitotenv.2020.141347}
-
-\leavevmode\hypertarget{ref-Sy2021population}{}%
-Sy, K.T.L., White, L.F., Nichols, B.E., 2021. Population density and
-basic reproductive number of COVID-19 across united states counties.
-PLOS ONE 16, e0249271.
-doi:\href{https://doi.org/10.1371/journal.pone.0249271}{10.1371/journal.pone.0249271}
-
-\leavevmode\hypertarget{ref-Viglione2020avalanche}{}%
-Viglione, G., 2020. 'Avalanche'of spider-paper retractions shakes
-behavioural-ecology community. Nature 578, 199--201.
-
-\leavevmode\hypertarget{ref-Vlasschaert2020proliferation}{}%
-Vlasschaert, C., Topf, J.M., Hiremath, S., 2020. Proliferation of papers
-and preprints during the coronavirus disease 2019 pandemic: Progress or
-problems with peer review? Advances in Chronic Kidney Disease 27,
-418--426.
-doi:\href{https://doi.org/10.1053/j.ackd.2020.08.003}{10.1053/j.ackd.2020.08.003}
-
-\leavevmode\hypertarget{ref-Wang2021transmission}{}%
-Wang, F., Tan, Z., Yu, Z., Yao, S., Guo, C., 2021. Transmission and
-control pressure analysis of the COVID-19 epidemic situation using
-multisource spatio-temporal big data. PLOS ONE 16, e0249145.
-doi:\href{https://doi.org/10.1371/journal.pone.0249145}{10.1371/journal.pone.0249145}
-
-\leavevmode\hypertarget{ref-White2020state}{}%
-White, E.R., Hébert-Dufresne, L., 2020. State-level variation of initial
-COVID-19 dynamics in the united states. PLOS ONE 15, e0240648.
-doi:\href{https://doi.org/10.1371/journal.pone.0240648}{10.1371/journal.pone.0240648}
-
-\leavevmode\hypertarget{ref-Wong2020spreading}{}%
-Wong, D.W.S., Li, Y., 2020. Spreading of COVID-19: Density matters. PLOS
-ONE 15, e0242398.
-doi:\href{https://doi.org/10.1371/journal.pone.0242398}{10.1371/journal.pone.0242398}
-
-\end{CSLReferences}
+\renewcommand\refname{References}
+\bibliography{bibliography.bib}
 

### 2022-10-08 R0-Density-Reanalysis/_R0-Density-Reanalysis.Rmd
Updated with final bibliographical information and created a renv reproducible environment

--- 2021-11-18 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
@@
 
-```{r install-data-package, include=FALSE}
-if(!require(r0density)){
-    remotes::install_github("paezha/Reproductive-Rate-and-Density-US-Reanalyzed", 
-                         subdir = "r0density")
-    library(r0density)
-}
-```
-
 ```{r load-packages, include=FALSE, cache=FALSE}
@@
 library(patchwork) # Package to organize ggplot2 plots
-library(r0density) # Data package
 library(sampleSelection) # Package to estimate sampling selection models

### 2022-10-08 README.Rmd
Updated with final bibliographical information and created a renv reproducible environment

@@
 
-Geographical Analysis (2021) https://doi.org/10.1111/gean.12307
+Paez, A (2022) Reproducibility of Research During COVID-19: Examining the Case of Population Density and the Basic Reproductive Rate from the Perspective of Spatial Analysis.
+Geographical Analysis.
+54(4) 860-880 (https://doi.org/10.1111/gean.12307)
 
@@
 library(units) # Package to work with units of measurement
-#library(viridis) # Color palettes
+```
+
+```{r install-data-package, include=FALSE}
+if(!require(r0density)){
+    install.packages(paste0(here(), "/r0density_0.1.0.tar.gz"), repos = NULL, type="source")
+    library(r0density) # not installed on this machine
+}
+```
+
+```{r load-data-package, include=FALSE, cache=FALSE}
+library(r0density) # Data package
 ```
